### 2016-03-04 frontiers-latex-templates/frontiers.tex
push initial files

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.   
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.2 Generated 2016/02/09 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%
%%%  NB logo1.jpg is required in the path in order to correctly compile front page header %%%

\documentclass{frontiersSCNS} % for Science, Engineering and Humanities and Social Sciences articles
%\documentclass{frontiersHLTH} % for Health articles
%\documentclass{frontiersFPHY} % for Physics and Applied Mathematics and Statistics articles

%\setcitestyle{square}
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}
\linenumbers

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
\def\Authors{First Author\,$^{1,*}$, Co-Author\,$^{2}$ and Co-Author\,$^{1,2}$}
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X \\
$^{2}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X  }
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Corresponding Author}

\def\corrEmail{email@uni.edu}

\begin{document}
\onecolumn
\firstpage{1}

\title[Running Title]{Article Title}

\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle

\begin{abstract}

%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
\section{}
For full guidelines regarding your manuscript please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines}{Author Guidelines}.

As a primary goal, the abstract should render the general significance and conceptual advance of the work clearly accessible to a broad readership. References should not be cited in the abstract. Leave the Abstract empty if your article does not require one, please see \href{http://www.frontiersin.org/about/AuthorGuidelines#SummaryTable}{Summary Table} for details according to article type.

\tiny
 \keyFont{ \section{Keywords:} keyword, keyword, keyword, keyword, keyword, keyword, keyword, keyword} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\end{abstract}

\section{Introduction}

For Original Research Articles \citep{conference}, Clinical Trial Articles \citep{article}, and Technology Reports \citep{patent}, the introduction should be succinct, with no subheadings \citep{book}. For Case Reports the Introduction should include symptoms at presentation \citep{chapter}, physical exams and lab results.

\section{Article types}

For requirements for a specific article type please refer to the Article Types on any Frontiers journal page. Please also refer to  \href{http://home.frontiersin.org/about/author-guidelines#Sections}{Author Guidelines} for further information on how to organize your manuscript in the required sections or their equivalents for your field

% For Original Research articles, please note that the Material and Methods section can be placed in any of the following ways: before Results, before Discussion or after Discussion.

\section{Manuscript Formatting}

\subsection{Heading Levels}

%There are 5 heading levels

\subsection{Level 2}
\subsubsection{Level 3}
\paragraph{Level 4}
\subparagraph{Level 5}

\subsection{Equations}
Equations should be inserted in editable format from the equation editor.

\begin{equation}
\sum x+ y =Z\label{eq:01}
\end{equation}

\subsection{Figures}
Frontiers requires figures to be submitted individually, in the same order as they are referred to in the manuscript. Figures will then be automatically embedded at the bottom of the submitted manuscript. Kindly ensure that each table and figure is mentioned in the text and in numerical order. Figures must be of sufficient resolution for publication \href{http://home.frontiersin.org/about/author-guidelines#ResolutionRequirements}{see here for examples and minimum requirements}. Figures which are not according to the guidelines will cause substantial delay during the production process. Please see \href{http://home.frontiersin.org/about/author-guidelines#GeneralStyleGuidelinesforFigures}{here} for full Figure guidelines

\subsubsection{Permission to reuse and Copyright}
Permission must be obtained for use of copyrighted material from other sources (including the web). Please note that it is compulsory to follow figure instructions.

\subsection{Tables}
Tables should be inserted at the end of the manuscript. Please build your table directly in LaTeX.Tables provided as jpeg/tiff files will not be accepted. Please note that very large tables (covering several pages) cannot be included in the final PDF for reasons of space. These tables will be published as \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} on the online article page at the time of acceptance. The author will be notified during the typesetting of the final article if this is the case.

\section{Nomenclature}

\subsection{Resource Identification Initiative}
To take part in the Resource Identification Initiative, please use the corresponding catalog number and RRID in your current manuscript. For more information about the project and for steps on how to search for an RRID, please click \href{http://www.frontiersin.org/files/pdf/letter_to_author.pdf}{here}.

\subsection{Life Science Identifiers}
Life Science Identifiers (LSIDs) for ZOOBANK registered names or nomenclatural acts should be listed in the manuscript before the keywords. For more information on LSIDs please see \href{http://www.frontiersin.org/about/AuthorGuidelines#InclusionofZoologicalNomenclature}{Inclusion of Zoological Nomenclature} section of the guidelines.

\section{Additional Requirements}

For additional requirements for specific article types and further information please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines#AdditionalRequirements}{Author Guidelines}.

\section*{Conflict of Interest Statement}
%All financial, commercial or other relationships that might be perceived by the academic community as representing a potential conflict of interest must be disclosed. If no such relationship exists, authors will be asked to confirm the following statement:

The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.

\section*{Author Contributions}

The Author Contributions section is mandatory for all articles, including articles by sole authors. If an appropriate statement is not provided on submission, a standard one will be inserted during the production process. The Author Contributions statement must describe the contributions of individual authors referred to by their initials and, in doing so, all authors agree to be accountable for the content of the work. Please see  \href{http://home.frontiersin.org/about/author-guidelines#AuthorandContributors}{here} for full authorship criteria.

\section*{Funding}
Details of all funding sources should be provided, including grant numbers if applicable. Please ensure to add all necessary funding information, as after publication this is no longer possible.

\section*{Acknowledgments}
This is a short text to acknowledge the contributions of specific colleagues, institutions, or agencies that aided the efforts of the authors.

\section*{Supplemental Data}
 \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} should be uploaded separately on submission, if there are Supplementary Figures, please include the caption in the same file as the figure. LaTeX Supplementary Material templates can be found in the Frontiers LaTeX folder

\bibliographystyle{frontiersinSCNS_ENG_HUMS} % for Science, Engineering and Humanities and Social Sciences articles, for Humanities and Social Sciences articles please include page numbers in the in-text citations
%\bibliographystyle{frontiersinHLTH&FPHY} % for Health, Physics and Mathematics articles
\bibliography{test}

%%% Upload the *bib file along with the *tex file and PDF on submission if the bibliography is not in the main *tex file

%%% Please see the test.bib file for some examples of references

\section*{Figures}

%%% Use this if adding the figures directly in the mansucript, if so, please remember to also upload the files when submitting your article
%%% There is no need for adding the file termination, as long as you indicate where the file is saved. In the examples below the files (logo1.jpg and logo2.eps) are in the Frontiers LaTeX folder
%%% If using *.tif files convert them to .jpg or .png
%%% If using panelled/compiled figures with subcaptions, use the subcaption package as in the second example.  N.B. This package is incompatible with the subfigure and subfig packages
%%%  NB logo1.jpg is required in the path in order to correctly compile front page header %%%

\begin{figure}[h!]
\begin{center}
\includegraphics[width=10cm]{logo1}% This is a *.jpg file
\end{center}
\caption{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures}\label{fig:1}
\end{figure}

\begin{figure}[h!]
\begin{minipage}[b]{.5\linewidth}
\centering\includegraphics[width=6cm]{YM-logo}% This is a *.eps file
\subcaption{A subfigure}\label{fig:2a}
\end{minipage}%
\begin{minipage}[b]{.5\linewidth}
\centering\includegraphics[width=2cm]{logo2}% This is an *.eps file
\subcaption{Another subfigure}\label{fig:2b}
\end{minipage}
\caption{A figure}\label{fig:2b}
\end{figure}

%%% If you don't add the figures in the LaTeX files, please upload them when submitting the article.

%%% Frontiers will add the figures at the end of the provisional pdf automatically %%%

%%% The use of LaTeX coding to draw Diagrams/Figures/Structures should be avoided. They should be external callouts including graphics.

\end{document}

### 2016-03-04 manuscript.Rmd
push initial files

---

---

## Abstract

350 Words

## Introduction

### 2016-03-04 manuscript.Rmd
fill out the YAML

@@
 ---
-
+title: "Modelling palaeoenvironmental time series and feature detection"
+author:
+- name: Gavin L Simpson
+  affiliation: Insitute of Environmental CHange and Society, University of Regina, Regina, SK
+date: March 4, 2016
+output: 
+  pdf_document:
+    keep_tex: true
+    md_extensions: +header_attributes+superscript+subscript
+#    pandoc_args: [
+#      "--no-tex-ligatures"
+#    ]
+#    template: frontiers-template.tex
+# documentclass: frontiersSCNS # not yet ready for this; need to pandoc-ify the template
+bibliography: references.bib
 ---

### 2016-03-04 analysis/small-water/small-water-age-depth-model.R
initial data for examples

@@ -1,0 +1,86 @@
+## Simple Age depth model for Small Water
+
+## Should be run from base git repo
+
+## Monotonic spline model
+
+## function to make reproducing this easier below
+`monoSpline` <- function(depth, date, error = NULL, increasing = FALSE,
+                         bs = "cr", k = 10, fx = FALSE, ...) {
+    stopifnot(require("mgcv"))
+    if(is.null(error)) {
+        error <- rep(0, length(depth))
+        W <- rep(1, length(depth))
+    } else {
+        W <- 1 / error
+        W[1] <- W[2] * error[2]
+    }
+    df <- data.frame(Depth = depth, Date = date, error = error)
+    mod <- gam(Date ~ s(Depth, k = k, bs = bs, fx = fx), data = df,
+               fit = FALSE)
+    sm <- smoothCon(s(Depth, k = k, bs = bs, fx = fx), data = df,
+                    knots = NULL)[[1]]
+    ## Fm are the constraints to enforce monotonicity
+    Fm <- mono.con(sm$xp, up = increasing)
+    G <- list(X = sm$X, C = matrix(0,0,0), sp = mod$sp, p = -sm$xp,
+              y = df$Date, Ain = Fm$A, bin = Fm$b,
+              S = sm$S, off = 0, w = W) ## w = df$Date*0+1
+    ## fitted parameters
+    p <- pcls(G)
+    ## fitted values
+    pm <- Predict.matrix(sm, data = data.frame(Depth = df$Depth))
+    fit <- pm %*% p
+    ## set up return object
+    ## this needs all info for a predict() method to work, inc
+    ## sm - smoothing matrix
+    ## p  - a vector of fitted coefficients; the object returned from pcls
+    ##
+    ## also return fitted values
+    out <- list(y = df$Date,
+                fitted.values = drop(fit),
+                coefficients = p,
+                residuals = drop(df$Date - fit),
+                sm = sm)
+    class(out) <- "monoSpline"
+    out
+}
+
+## predict S3 method
+`predict.monoSpline` <- function(object, newdata, ...) {
+    if(missing(newdata))
+        yhat <- fitted(object)
+    else {
+        yhat <- Predict.matrix(object$sm, data = newdata) %*% coef(object)
+    }
+    yhat
+}
+
+## R Packages
+library("mgcv")                         # explicit but `monoSpline()` will check mgcv is available
+library("ggplot2")                      # for plotting
+
+## SMALL1 core
[…]

### 2016-03-04 analysis/small-water/small-water-analysis.R
initial small water analysis

@@ -1,0 +1,67 @@
+## Analyse the d15N data for Small Water, SMALL1 core
+
+## Packages
+library("mgcv")
+library("ggplot2")
+
+## load the simultaneous confidence interval derivatives code
+tmpf <- tempfile()
+download.file("https://gist.githubusercontent.com/gavinsimpson/ca18c9c789ef5237dbc6/raw/295fc5cf7366c831ab166efaee42093a80622fa8/derivSimulCI.R",
+              tmpf, method = "wget")
+source(tmpf)
+
+## Data
+small <- readRDS("./data/small-water/small-water-isotope-data.rds")
+head(small)
+
+## Generate a plot of the data
+plt <- ggplot(small, aes(x = Year, y = d15N)) +
+    geom_point() +
+        theme_bw() +
+            ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
+plt
+
+## model it using gamm()
+mod <- gamm(d15N ~ s(Year, k = 15), data = small, correlation = corCAR1(form = ~ Year), method = "REML")
+
+summary(mod$gam)
+intervals(mod$lme, which = "var-cov")
+plot(mod$gam)
+
+## Predict from model
+newYear <- with(small, data.frame(Year = seq(min(Year), max(Year), length.out = 100)))
+newYear <- cbind(newYear, data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
+crit <- qt(0.975, df = df.residual(mod$gam))
+newYear <- transform(newYear,
+                     upper = fit + (crit * se.fit),
+                     lower = fit - (crit * se.fit))
+
+## Draw the fitted spline on the data
+plt.fit <- ggplot(small, aes(x = Year, y = d15N)) +
+        geom_ribbon(data = newYear, aes(y = fit, ymax = upper, ymin = lower, x = Year),
+                    fill = "grey", alpha = 0.3) +
+            geom_line(data = newYear, aes(y = fit, x = Year), col = "darkgrey", size = 0.8) +
+                geom_point() +
+        theme_bw() +
+            ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
+plt.fit
+
+## Derivatives
+fd <- derivSimulCI(mod, samples = 10000, n = 100)
+CI <- apply(fd[[1]]$simulations, 1, quantile, probs = c(0.025, 0.975))
+
+## Need to do this manually, not through `fd` - write a function
+set.seed(1)
[…]

### 2016-03-06 functions/simulate.gamm.R
separate fodler for functions; add a simulate.gamm method

@@ -1,0 +1,23 @@
+`simulate.gamm` <- function(object, nsim = 1, seed = NULL, newdata,
+                            freq = FALSE, unconditional = FALSE, ...) {
+    stopifnot(require("MASS"))
+    if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE))
+        runif(1)
+    if (is.null(seed))
+        RNGstate <- get(".Random.seed", envir = .GlobalEnv)
+    else {
+        R.seed <- get(".Random.seed", envir = .GlobalEnv)
+        set.seed(seed)
+        RNGstate <- structure(seed, kind = as.list(RNGkind()))
+        on.exit(assign(".Random.seed", R.seed, envir = .GlobalEnv))
+    }
+
+    if (missing(newdata)) {
+        newdata <- object$gam$model
+    }
+    Rbeta <- mvrnorm(n = nsim, coef(object$gam),
+                     vcov(object$gam, freq = freq, unconditional = unconditional))
+    Xp <- predict(object$gam, newdata = newdata, type = "lpmatrix")
+    sims <- Xp %*% t(Rbeta)
+    sims
+}

### 2016-03-06 analysis/small-water/small-water-analysis.R
finish off the small water example

@@ -3,12 +3,16 @@
 ## Packages
 library("mgcv")
 library("ggplot2")
+library("cowplot")
 
 ## load the simultaneous confidence interval derivatives code
 tmpf <- tempfile()
 download.file("https://gist.githubusercontent.com/gavinsimpson/ca18c9c789ef5237dbc6/raw/295fc5cf7366c831ab166efaee42093a80622fa8/derivSimulCI.R",
               tmpf, method = "wget")
 source(tmpf)
+
+## Source other functions
+source("functions/simulate.gamm.R")
 
 ## Data
 small <- readRDS("./data/small-water/small-water-isotope-data.rds")
@@ -22,7 +26,8 @@ plt <- ggplot(small, aes(x = Year, y = d15N)) +
 plt
 
 ## model it using gamm()
-mod <- gamm(d15N ~ s(Year, k = 15), data = small, correlation = corCAR1(form = ~ Year), method = "REML")
+mod <- gamm(d15N ~ s(Year, k = 15), data = small,
+            correlation = corCAR1(form = ~ Year), method = "REML")
 
 summary(mod$gam)
 intervals(mod$lme, which = "var-cov")
@@ -38,30 +43,71 @@ newYear <- transform(newYear,
 
 ## Draw the fitted spline on the data
 plt.fit <- ggplot(small, aes(x = Year, y = d15N)) +
-        geom_ribbon(data = newYear, aes(y = fit, ymax = upper, ymin = lower, x = Year),
-                    fill = "grey", alpha = 0.3) +
-            geom_line(data = newYear, aes(y = fit, x = Year), col = "darkgrey", size = 0.8) +
-                geom_point() +
-        theme_bw() +
-            ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
+    geom_ribbon(data = newYear, aes(y = fit, ymax = upper, ymin = lower, x = Year),
+                fill = "grey", alpha = 0.3) +
+    geom_line(data = newYear, aes(y = fit, x = Year), col = "black", size = 0.8) +
+    geom_point(col = "darkgrey") +
+    theme_bw() +
+    ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
 plt.fit
+
+## Need to do this manually, not through `fd` - write a function
+set.seed(1)
+nsim <- 50
+randSims <- simulate(mod, nsim = nsim, newdata = newYear, unconditional = TRUE)
+colnames(randSims) <- paste0("sim", seq_len(nsim))
[…]

### 2016-03-07 analysis/small-water/small-water-analysis.R
drop cowplot as it doesn't align plots properly; use gtable instead

@@ -3,7 +3,8 @@
 ## Packages
 library("mgcv")
 library("ggplot2")
-library("cowplot")
+## library("cowplot")
+library("gtable")
 
 ## load the simultaneous confidence interval derivatives code
 tmpf <- tempfile()
@@ -45,7 +46,7 @@ newYear <- transform(newYear,
 plt.fit <- ggplot(small, aes(x = Year, y = d15N)) +
     geom_ribbon(data = newYear, aes(y = fit, ymax = upper, ymin = lower, x = Year),
                 fill = "grey", alpha = 0.3) +
-    geom_line(data = newYear, aes(y = fit, x = Year), col = "black", size = 0.8) +
+    geom_line(data = newYear, aes(y = fit, x = Year), col = "black") +
     geom_point(col = "darkgrey") +
     theme_bw() +
     ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
@@ -63,7 +64,7 @@ randSims <- transform(randSims, Year = rep(newYear$Year, nsim),
 plt.sim <- ggplot(newYear, aes(x = Year, y = fit)) +
     geom_line(data = randSims, mapping = aes(y = simulated, x = Year, group = run),
               col = "black", alpha = 0.1) +
-    geom_line(size = 0.8, col = "black") +
+    geom_line() +
     theme_bw() +
     ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter))) +
     xlab("Year")
@@ -87,7 +88,7 @@ derivPlt <- ggplot(newYear, aes(x = Year, y = derivative)) +
     geom_line() +
     geom_line(aes(y = increasing), col = "black", size = 1.5) +
     geom_line(aes(y = decreasing), col = "black", size = 1.5) +
-    ylab(expression(italic(hat(f)*"'"*(Year)))) +
+    ylab(expression(italic(hat(f) * "'") * (Year))) +
     xlab("Year") +
     theme_bw()
 derivPlt
@@ -106,8 +107,18 @@ plt.fit2 <- plt.fit +
     geom_line(data = newYear, mapping = aes(y = YearDecr), size = 1.5)
 plt.fit2
 
-## Cowplot
-displayPlt <- plot_grid(plt.fit, plt.sim, derivPlt, plt.fit2, labels = "AUTO")
-displayPlt
+## ## Cowplot
[…]

### 2016-03-07 analysis/small-water/small-water-analysis.R
switch back to cowplot as it can align the plots properly

@@ -3,8 +3,7 @@
 ## Packages
 library("mgcv")
 library("ggplot2")
-## library("cowplot")
-library("gtable")
+library("cowplot")
 
 ## load the simultaneous confidence interval derivatives code
 tmpf <- tempfile()
@@ -107,18 +106,18 @@ plt.fit2 <- plt.fit +
     geom_line(data = newYear, mapping = aes(y = YearDecr), size = 1.5)
 plt.fit2
 
-## ## Cowplot
-## displayPlt <- plot_grid(plt.fit, plt.sim, derivPlt, plt.fit2, labels = "AUTO", ncol = 2)
-## displayPlt
+## library("gtable")
+## c1 <- rbind(ggplotGrob(plt.fit), ggplotGrob(derivPlt), size = "last")
+## c2 <- rbind(ggplotGrob(plt.sim), ggplotGrob(plt.fit2), size = "last")
+## displayPlt <- cbind(c1, c2, size = "last")
+## plot(displayPlt)
 
-## ggsave("./figures/small-water-display.pdf", displayPlt)
+## pdf("./figures/small-water-display.pdf", height = 9, width = 9)
+## grid.draw(displayPlt)
+## dev.off()
 
+combinedPlt <- plot_grid(plt.fit, plt.sim, derivPlt, plt.fit2, labels = "AUTO",
+                         align = "vh")
+combinedPlt
 
-c1 <- rbind(ggplotGrob(plt.fit), ggplotGrob(derivPlt), size = "last")
-c2 <- rbind(ggplotGrob(plt.sim), ggplotGrob(plt.fit2), size = "last")
-displayPlt <- cbind(c1, c2, size = "last")
-plot(displayPlt)
-
-pdf("./figures/small-water-display.pdf", height = 9, width = 9)
-grid.draw(displayPlt)
-dev.off()
+ggsave("./figures/small-water-display.pdf", combinedPlt, width = 10, height = 9)

### 2016-03-07 functions/monoSpline.R
function for monotonic splines for age-depth models

@@ -1,0 +1,49 @@
+`monoSpline` <- function(depth, date, error = NULL, increasing = FALSE,
+                         bs = "cr", k = 10, fx = FALSE, ...) {
+    stopifnot(require("mgcv"))
+    if(is.null(error)) {
+        error <- rep(0, length(depth))
+        W <- rep(1, length(depth))
+    } else {
+        W <- 1 / error
+        W[1] <- W[2] * error[2]
+    }
+    df <- data.frame(Depth = depth, Date = date, error = error)
+    mod <- gam(Date ~ s(Depth, k = k, bs = bs, fx = fx), data = df,
+               fit = FALSE)
+    sm <- smoothCon(s(Depth, k = k, bs = bs, fx = fx), data = df,
+                    knots = NULL)[[1]]
+    ## Fm are the constraints to enforce monotonicity
+    Fm <- mono.con(sm$xp, up = increasing)
+    G <- list(X = sm$X, C = matrix(0,0,0), sp = mod$sp, p = -sm$xp,
+              y = df$Date, Ain = Fm$A, bin = Fm$b,
+              S = sm$S, off = 0, w = W) ## w = df$Date*0+1
+    ## fitted parameters
+    p <- pcls(G)
+    ## fitted values
+    pm <- Predict.matrix(sm, data = data.frame(Depth = df$Depth))
+    fit <- pm %*% p
+    ## set up return object
+    ## this needs all info for a predict() method to work, inc
+    ## sm - smoothing matrix
+    ## p  - a vector of fitted coefficients; the object returned from pcls
+    ##
+    ## also return fitted values
+    out <- list(y = df$Date,
+                fitted.values = drop(fit),
+                coefficients = p,
+                residuals = drop(df$Date - fit),
+                sm = sm)
+    class(out) <- "monoSpline"
+    out
+}
+
+## predict S3 method
+`predict.monoSpline` <- function(object, newdata, ...) {
+    if(missing(newdata))
+        yhat <- fitted(object)
+    else {
+        yhat <- Predict.matrix(object$sm, data = newdata) %*% coef(object)
+    }
+    yhat
+}

### 2016-03-07 functions/simulate.gam.R
add a simulate method for gam() models

@@ -1,0 +1,23 @@
+`simulate.gam` <- function(object, nsim = 1, seed = NULL, newdata,
+                            freq = FALSE, unconditional = FALSE, ...) {
+    stopifnot(require("MASS"))
+    if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE))
+        runif(1)
+    if (is.null(seed))
+        RNGstate <- get(".Random.seed", envir = .GlobalEnv)
+    else {
+        R.seed <- get(".Random.seed", envir = .GlobalEnv)
+        set.seed(seed)
+        RNGstate <- structure(seed, kind = as.list(RNGkind()))
+        on.exit(assign(".Random.seed", R.seed, envir = .GlobalEnv))
+    }
+
+    if (missing(newdata)) {
+        newdata <- object$model
+    }
+    Rbeta <- mvrnorm(n = nsim, coef(object),
+                     vcov(object, freq = freq, unconditional = unconditional))
+    Xp <- predict(object, newdata = newdata, type = "lpmatrix")
+    sims <- Xp %*% t(Rbeta)
+    sims
+}

### 2016-03-07 analysis/small-water/small-water-age-depth-model.R
updates to the small water example

@@ -4,60 +4,11 @@
 
 ## Monotonic spline model
 
-## function to make reproducing this easier below
-`monoSpline` <- function(depth, date, error = NULL, increasing = FALSE,
-                         bs = "cr", k = 10, fx = FALSE, ...) {
-    stopifnot(require("mgcv"))
-    if(is.null(error)) {
-        error <- rep(0, length(depth))
-        W <- rep(1, length(depth))
-    } else {
-        W <- 1 / error
-        W[1] <- W[2] * error[2]
-    }
-    df <- data.frame(Depth = depth, Date = date, error = error)
-    mod <- gam(Date ~ s(Depth, k = k, bs = bs, fx = fx), data = df,
-               fit = FALSE)
-    sm <- smoothCon(s(Depth, k = k, bs = bs, fx = fx), data = df,
-                    knots = NULL)[[1]]
-    ## Fm are the constraints to enforce monotonicity
-    Fm <- mono.con(sm$xp, up = increasing)
-    G <- list(X = sm$X, C = matrix(0,0,0), sp = mod$sp, p = -sm$xp,
-              y = df$Date, Ain = Fm$A, bin = Fm$b,
-              S = sm$S, off = 0, w = W) ## w = df$Date*0+1
-    ## fitted parameters
-    p <- pcls(G)
-    ## fitted values
-    pm <- Predict.matrix(sm, data = data.frame(Depth = df$Depth))
-    fit <- pm %*% p
-    ## set up return object
-    ## this needs all info for a predict() method to work, inc
-    ## sm - smoothing matrix
-    ## p  - a vector of fitted coefficients; the object returned from pcls
-    ##
-    ## also return fitted values
-    out <- list(y = df$Date,
-                fitted.values = drop(fit),
-                coefficients = p,
-                residuals = drop(df$Date - fit),
-                sm = sm)
-    class(out) <- "monoSpline"
-    out
-}
-
-## predict S3 method
-`predict.monoSpline` <- function(object, newdata, ...) {
-    if(missing(newdata))
-        yhat <- fitted(object)
-    else {
-        yhat <- Predict.matrix(object$sm, data = newdata) %*% coef(object)
-    }
-    yhat
-}
-
 ## R Packages
 library("mgcv")                         # explicit but `monoSpline()` will check mgcv is available
 library("ggplot2")                      # for plotting
+
+source("./functions/monoSpline.R")
 
 ## SMALL1 core
 small <- read.csv("./data/small-water/small1-dating.csv")

### 2016-03-07 analysis/small-water/small-water-analysis.R

@@ -20,9 +20,9 @@ head(small)
 
 ## Generate a plot of the data
 plt <- ggplot(small, aes(x = Year, y = d15N)) +
-    geom_point() +
-        theme_bw() +
-            ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
+    geom_point(col = "darkgrey") +
+    theme_bw() +
+    ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
 plt
 
 ## model it using gamm()
@@ -36,15 +36,9 @@ plot(mod$gam)
 ## Predict from model
 newYear <- with(small, data.frame(Year = seq(min(Year), max(Year), length.out = 100)))
 newYear <- cbind(newYear, data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
-crit <- qt(0.975, df = df.residual(mod$gam))
-newYear <- transform(newYear,
-                     upper = fit + (crit * se.fit),
-                     lower = fit - (crit * se.fit))
 
 ## Draw the fitted spline on the data
 plt.fit <- ggplot(small, aes(x = Year, y = d15N)) +
-    geom_ribbon(data = newYear, aes(y = fit, ymax = upper, ymin = lower, x = Year),
-                fill = "grey", alpha = 0.3) +
     geom_line(data = newYear, aes(y = fit, x = Year), col = "black") +
     geom_point(col = "darkgrey") +
     theme_bw() +
@@ -53,12 +47,18 @@ plt.fit
 
 ## Need to do this manually, not through `fd` - write a function
 set.seed(1)
-nsim <- 50
-randSims <- simulate(mod, nsim = nsim, newdata = newYear, unconditional = TRUE)
-colnames(randSims) <- paste0("sim", seq_len(nsim))
+nsim <- 10000
+take <- 50
+sims <- simulate(mod, nsim = nsim, newdata = newYear, unconditional = TRUE)
+randSims <- sims[, sample(nsim, take)]
+colnames(randSims) <- paste0("sim", seq_len(take))
 randSims <- setNames(stack(as.data.frame(randSims)), c("simulated", "run"))
-randSims <- transform(randSims, Year = rep(newYear$Year, nsim),
+randSims <- transform(randSims, Year = rep(newYear$Year, take),
                       simulated = simulated)
+
+## Add a CI to the fitted spline
[…]

### 2016-03-07 analysis/braya-so/braya-so-analysis.R
add braya-so UK37 example

@@ -1,0 +1,122 @@
+## Analyse the UK37 data for Braya So core
+
+## Packages
+library("mgcv")
+library("ggplot2")
+library("cowplot")
+
+## load the simultaneous confidence interval derivatives code
+tmpf <- tempfile()
+download.file("https://gist.githubusercontent.com/gavinsimpson/ca18c9c789ef5237dbc6/raw/295fc5cf7366c831ab166efaee42093a80622fa8/derivSimulCI.R",
+              tmpf, method = "wget")
+source(tmpf)
+
+## Source other functions
+source("functions/simulate.gam.R")
+
+## Data
+braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
+                    skip = 84)
+names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
+head(braya)
+
+## Generate a plot of the data
+ylabel <- expression(italic(U)[37]^{italic(k)})
+plt <- ggplot(braya, aes(x = Year, y = UK37)) +
+    geom_point(col = "darkgrey") +
+    geom_line(col = "darkgrey") +
+    theme_bw() +
+    ylab(ylabel)
+plt
+
+## model it using gam()
+mod <- gam(UK37 ~ s(Year, k = 50, bs = "gp", m = c(3, 100)), data = braya, method = "REML")
+summary(mod)
+plot(mod)
+
+## Predict from model
+N <- 300
+newYear <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = N)))
+newYear <- cbind(newYear, data.frame(predict(mod, newYear, se.fit = TRUE)))
+
+## Draw the fitted spline on the data
+plt.fit <- ggplot(braya, aes(x = Year, y = UK37)) +
+    geom_line(data = newYear, aes(y = fit, x = Year), col = "black") +
+    geom_point(col = "darkgrey") +
+    theme_bw() +
+    ylab(ylabel)
+plt.fit
+
+## Need to do this manually, not through `fd` - write a function
+set.seed(1)
+nsim <- 10000
+take <- 50
+sims <- simulate(mod, nsim = nsim, newdata = newYear, unconditional = TRUE)
+randSims <- sims[, sample(nsim, take)]
+colnames(randSims) <- paste0("sim", seq_len(take))
+randSims <- setNames(stack(as.data.frame(randSims)), c("simulated", "run"))
+randSims <- transform(randSims, Year = rep(newYear$Year, take),
[…]

### 2016-03-08 functions/correlation-functions.R
functions to generate correlation functions of power exponential and matern families

@@ -1,0 +1,13 @@
+## Implements the power exponential and matern covariance functions
+
+`powerExp` <- function(h, phi, k) {
+    exp(- abs(h / phi)^k)
+}
+
+`matern` <- function(h, phi, k) {
+    absh <- abs(h / phi)
+    addn <- ifelse(k == 1.5, 0,
+                   ifelse(k == 2.5, (1/3)*absh^2,
+                          ((2/5)*absh^2) + ((1/15) * absh^3)))
+    exp(- absh) * (1 + absh + addn)
+}

### 2016-03-08 analysis/other/correlation-functions.R
add code to generate examples of correlation functions implied by various gaussian processes

@@ -1,0 +1,46 @@
+## Packages
+library("ggplot2")
+library("cowplot")
+library("viridis")
+
+## load the correlation functions
+source("./functions/correlation-functions.R")
+
+pExpdf <- expand.grid(h = seq(0, 7, length = 100),
+                      phi = c(1,2),
+                      k = c(0.5, 1, 2))
+pExpdf <- transform(pExpdf, correlation = powerExp(h, phi, k),
+                    kfac = factor(k))
+
+pexp.plt <- ggplot(pExpdf, aes(x = h, y = correlation, group = kfac, colour = kfac)) +
+    geom_line(size = 1) +
+        facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
+        ylab(expression(rho)) +
+            xlab(expression(italic(h))) +
+                scale_color_viridis(discrete=TRUE) +
+                theme_bw() +
+                    guides(colour = guide_legend(title = expression(kappa)))
+pexp.plt
+
+pMaternDf <- expand.grid(h = seq(0, 10, length = 100),
+                         phi = c(1,2),
+                         k = c(1.5, 2.5, 3.5))
+pMaternDf <- transform(pMaternDf, correlation = matern(h, phi, k),
+                       kfac = factor(k))
+
+pmat.plt <- ggplot(pMaternDf, aes(x = h, y = correlation, group = kfac, colour = kfac)) +
+    geom_line(size = 1) +
+        facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
+        ylab(expression(rho)) +
+            xlab(expression(italic(h))) +
+                scale_color_viridis(discrete=TRUE) +
+                theme_bw() +
+                    guides(colour = guide_legend(title = expression(kappa)))
+pmat.plt
+
+correlFuns <- plot_grid(pexp.plt, pmat.plt, ncol = 1, align = "hv",
+                        labels = "AUTO")
+correlFuns
+
+ggsave("./figures/correlation-functions-plot.pdf", correlFuns, width = 9,
+       height = 9)

### 2016-03-08 analysis/other/correlation-functions.R
tweak the viridis scales to darken the yellow

@@ -14,12 +14,13 @@ pExpdf <- transform(pExpdf, correlation = powerExp(h, phi, k),
 
 pexp.plt <- ggplot(pExpdf, aes(x = h, y = correlation, group = kfac, colour = kfac)) +
     geom_line(size = 1) +
-        facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
-        ylab(expression(rho)) +
-            xlab(expression(italic(h))) +
-                scale_color_viridis(discrete=TRUE) +
-                theme_bw() +
-                    guides(colour = guide_legend(title = expression(kappa)))
+    facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
+    ylab(expression(rho)) +
+    xlab(expression(italic(h))) +
+    ##scale_color_viridis(option = "viridis", discrete = TRUE) +
+    scale_colour_manual(values = viridis(11)[c(2,5,10)]) +
+    theme_bw() +
+    guides(colour = guide_legend(title = expression(kappa)))
 pexp.plt
 
 pMaternDf <- expand.grid(h = seq(0, 10, length = 100),
@@ -30,12 +31,13 @@ pMaternDf <- transform(pMaternDf, correlation = matern(h, phi, k),
 
 pmat.plt <- ggplot(pMaternDf, aes(x = h, y = correlation, group = kfac, colour = kfac)) +
     geom_line(size = 1) +
-        facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
-        ylab(expression(rho)) +
-            xlab(expression(italic(h))) +
-                scale_color_viridis(discrete=TRUE) +
-                theme_bw() +
-                    guides(colour = guide_legend(title = expression(kappa)))
+    facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
+    ylab(expression(rho)) +
+    xlab(expression(italic(h))) +
+    ## scale_color_viridis(discrete = TRUE) +
+    scale_colour_manual(values = viridis(11)[c(2,5,10)]) +
+    theme_bw() +
+    guides(colour = guide_legend(title = expression(kappa)))
 pmat.plt
 
 correlFuns <- plot_grid(pexp.plt, pmat.plt, ncol = 1, align = "hv",

### 2016-11-03 analysis/braya-so/braya-so-induced-correlations.R
add code to look at induced correlations

@@ -1,0 +1,82 @@
+## Analyse the UK37 data for Braya So coreLook at induced correlations
+
+## Packages
+library("mgcv")
+library("ggplot2")
+library("cowplot")
+library("grid") # for unit.pmax(), unit.list()
+
+theme_set(theme_bw())
+
+## Source other functions
+source("functions/induced-correlations.R")
+
+## Data
+braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
+                    skip = 84)
+names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
+braya <- transform(braya, sampleInterval = YearYoung - YearOld)
+head(braya)
+
+## induced correlations....
+T <- 601
+t <- 1:T
+K <- 10
+set.seed(1)
+et <- rnorm(T, sd = 0.25)
+ft <- sin(2 * pi * ((t-1) / T))
+yt <- ft + et
+egData <- data.frame(yt = yt, t = t, ft = ft)
+
+egData.plt <- ggplot(egData, aes(x = t, y = yt)) +
+    geom_point(size = 0.9) +
+        geom_line(aes(y = ft), col = "red", lwd = 1) +
+            labs(x = expression(italic(t)), y = expression(y[italic(t)]))
+egData.plt
+
+m.ps <- gam(yt ~ s(t, bs = "ps", k = K), data = egData, method = "REML")
+m.tp <- gam(yt ~ s(t, bs = "tp", k = K), data = egData, method = "REML")
+
+tp <- c(100,225,350,500)
+
+egDataCor <- rbind(addInduced(t, tp, m.ps), addInduced(t, tp, m.tp))
+egDataCor <- cbind(egDataCor, Basis = rep(c("P Spline", "Thinplate Spline"), each = T * length(tp)))
+
+eg.plt <- ggplot(egDataCor, aes(x = t, y = correl, colour = tp)) +
+    geom_line() + facet_wrap( ~ Basis) + scale_colour_discrete(guide = FALSE) +
+            labs(x = expression(italic(t)), y = "Induced correlation")
+eg.plt
+
+uk.ps <- gam(UK37 ~ s(Year, k = 30, bs = "ps"), data = braya, method = "REML")
+uk.tp <- gam(UK37 ~ s(Year, k = 30, bs = "tp"), data = braya, method = "REML")
+
+ukTP <- c(20,40,60,80)
[…]

### 2016-11-03 functions/induced-correlations.R
functions to compute induced correlations of splines as random effects

@@ -1,0 +1,30 @@
+
+corFun <- function(V, t) {
+    top <- V[t, ]
+    bot <- sqrt(V[t,t] * diag(V))
+    top / bot
+}
+
+inducedCor <- function(m, i) {
+    ind <- rep(0,0)
+    for (j in 1:length(m$smooth))
+        ind <- c(ind, m$smooth[[j]]$first.para-1 + 1:m$smooth[[j]]$rank)
+    mmat <- model.matrix(m)
+    Z <- mmat[,ind]
+    ZZt <- tcrossprod(Z)
+    I <- diag(nrow(mmat))
+    tmp <- capture.output(vcomps <- gam.vcomp(m))
+    sigma <- vcomps[2,1]
+    tau <- vcomps[1,1]
+    V <- (sigma^2 * I) + (tau^2 * ZZt)
+    corFun(V, i)
+}
+
+addInduced <- function(t, tp, mod) {
+    cors <- unlist(lapply(tp, function(i, mod) inducedCor(mod, i), mod = mod),
+                   use.names = FALSE)
+    df <- data.frame(t  = rep(t, length(tp)),
+                     tp = factor(rep(tp, each = length(t))),
+                     correl = cors)
+    df
+}

### 2016-11-03 analysis/braya-so/braya-so-analysis.R
latest scripts to repeate analysis

@@ -7,6 +7,7 @@ library("cowplot")
 library("viridis")
 
 pal <- viridis(2)
+theme_set(theme_bw())
 
 ## load the simultaneous confidence interval derivatives code
 tmpf <- tempfile()
@@ -21,6 +22,7 @@ source("functions/simulate.gam.R")
 braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
                     skip = 84)
 names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
+braya <- transform(braya, sampleInterval = YearYoung - YearOld)
 head(braya)
 
 ## Generate a plot of the data
@@ -32,94 +34,194 @@ plt <- ggplot(braya, aes(x = Year, y = UK37)) +
     ylab(ylabel)
 plt
 
+## variography...
+library("gstat")
+library("sp")
+m.reml <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "REML")
+resi.reml <- data.frame(Residuals = resid(m.reml), Year = braya$Year, Dummy = 1)
+coordinates(resi.reml) <- ~ Year + Dummy
+hscat(Residuals ~ 1, data = resi.reml, breaks = c(0, 25, 50, 100,200, 500, 1000))
+resi.vgmCloud <- variogram(Residuals ~ 1, data = resi.reml, cloud = TRUE)
+resi.vgm <- variogram(Residuals ~ 1, data = resi.reml, width = 20, cressie = TRUE)
+
+ggplot(resi.vgm, aes(x = dist, y = gamma)) +
+    geom_smooth(se = FALSE, col = "grey", span = 0.75,
+                method.args = list(family = "gaussian")) +
+    geom_point() +
+    xlim(c(0, 1000)) + ylim(c(0, 1e-3)) +
+    ylab(expression(Semivariance ~~ gamma)) +
+    xlab("Temporal separation (years)")
+
 ## model it using gam()
-mod <- gam(UK37 ~ s(Year, k = 30, bs = "gp", m = c(3, 100)), data = braya, method = "REML")
+effRange <- 250
+mod <- gam(UK37 ~ s(Year, k = 30, bs = "gp", m = c(3, effRange)), data = braya,
+           method = "REML")
 summary(mod)
-plot(mod)
+plot(mod, residuals = TRUE, pch = 16, n = 300)
 
-## Predict from model
-N <- 300
[…]

### 2016-11-03 analysis/braya-so/braya-so-failure-analysis.R

@@ -45,15 +45,29 @@ summary(mod2)
 plot(mod2, n = N)
 gam.check(mod2)
 
-fit <- predict(mod2, newdata = newYear, se.fit = TRUE)
+fit2 <- predict(mod2, newdata = newYear, se.fit = TRUE)
 
-newYear <- rbind(newYear, newYear)
-newYear[seq(N+1, length.out = N, by = 1), ]$fit <- fit$fit
+mod3 <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "REML")
+summary(mod3)
+plot(mod3, n = N)
+gam.check(mod3)
+
+fit3 <- predict(mod3, newdata = newYear, se.fit = TRUE)
+
+newYear <- rbind(newYear, newYear, newYear)
+## GAM GCV
+newYear[seq(N+1, length.out = N, by = 1), ]$fit <- fit2$fit
 newYear[seq(N+1, length.out = N, by = 1), ]$upper <-
-    fit$fit + (qt(0.975, df.residual(mod2)) * fit$se.fit)
-newYear[seq(301, length.out = 300, by = 1), ]$lower <-
-    fit$fit - (qt(0.975, df.residual(mod2)) * fit$se.fit)
-newYear <- transform(newYear, Method = rep(c("GAMM (REML)", "GAM (GCV)"), each = N))
+    fit2$fit + (qt(0.975, df.residual(mod2)) * fit2$se.fit)
+newYear[seq(N+1, length.out = N, by = 1), ]$lower <-
+    fit2$fit - (qt(0.975, df.residual(mod2)) * fit2$se.fit)
+## GAM REML
+newYear[seq((N*2)+1, length.out = N, by = 1), ]$fit <- fit3$fit
+newYear[seq((N*2)+1, length.out = N, by = 1), ]$upper <-
+    fit3$fit + (qt(0.975, df.residual(mod2)) * fit3$se.fit)
+newYear[seq((N*2)+1, length.out = N, by = 1), ]$lower <-
+    fit3$fit - (qt(0.975, df.residual(mod2)) * fit3$se.fit)
+newYear <- transform(newYear, Method = rep(c("GAMM (REML)", "GAM (GCV)", "GAM (REML)"), each = N))
 
 p1 <- ggplot(braya, aes(y = UK37, x = Year)) +
     geom_point() +
@@ -63,25 +77,28 @@ p1 <- ggplot(braya, aes(y = UK37, x = Year)) +
[…]

### 2016-11-03 analysis/braya-so/braya-so-induced-correlations.R

@@ -1,4 +1,6 @@
-## Analyse the UK37 data for Braya So coreLook at induced correlations
+# ---- induced-correlations-plt ----
+
+## Analyse the UK37 data for Braya So core: Look at induced correlations
 
 ## Packages
 library("mgcv")
@@ -16,7 +18,6 @@ braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
                     skip = 84)
 names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
 braya <- transform(braya, sampleInterval = YearYoung - YearOld)
-head(braya)
 
 ## induced correlations....
 T <- 601
@@ -32,7 +33,7 @@ egData.plt <- ggplot(egData, aes(x = t, y = yt)) +
     geom_point(size = 0.9) +
         geom_line(aes(y = ft), col = "red", lwd = 1) +
             labs(x = expression(italic(t)), y = expression(y[italic(t)]))
-egData.plt
+## egData.plt
 
 m.ps <- gam(yt ~ s(t, bs = "ps", k = K), data = egData, method = "REML")
 m.tp <- gam(yt ~ s(t, bs = "tp", k = K), data = egData, method = "REML")
@@ -45,7 +46,7 @@ egDataCor <- cbind(egDataCor, Basis = rep(c("P Spline", "Thinplate Spline"), eac
 eg.plt <- ggplot(egDataCor, aes(x = t, y = correl, colour = tp)) +
     geom_line() + facet_wrap( ~ Basis) + scale_colour_discrete(guide = FALSE) +
             labs(x = expression(italic(t)), y = "Induced correlation")
-eg.plt
+## eg.plt
 
 uk.ps <- gam(UK37 ~ s(Year, k = 30, bs = "ps"), data = braya, method = "REML")
 uk.tp <- gam(UK37 ~ s(Year, k = 30, bs = "tp"), data = braya, method = "REML")
@@ -57,7 +58,7 @@ ukData <- cbind(ukData, Basis = rep(c("P Spline", "Thinplate Spline"), each = nr
 uk.plt <- ggplot(ukData, aes(x = t, y = correl, colour = tp)) +
     geom_line() + facet_wrap( ~ Basis) +
         labs(x = "Year BP", y = "Induced correlation") + scale_colour_discrete(guide = FALSE)
-uk.plt
+## uk.plt
 
 ## cowplot the panels together
[…]

### 2016-11-03 analysis/small-water/small-water-analysis.R
add code to plat the estimated CAR(1) correlation function

@@ -8,6 +8,8 @@ library("viridis")
 
 pal <- viridis(3)
 
+theme_set(theme_bw())
+
 ## load the simultaneous confidence interval derivatives code
 tmpf <- tempfile()
 download.file("https://gist.githubusercontent.com/gavinsimpson/ca18c9c789ef5237dbc6/raw/295fc5cf7366c831ab166efaee42093a80622fa8/derivSimulCI.R",
@@ -35,6 +37,22 @@ mod <- gamm(d15N ~ s(Year, k = 15), data = small,
 summary(mod$gam)
 intervals(mod$lme, which = "var-cov")
 plot(mod$gam)
+
+## plot CAR(1) process
+maxS <- with(small, diff(range(Year))) ## too large, truncate to 50
+S <- seq(0, 50, length = 100)
+phi <- intervals(mod$lme, which = "var-cov")$corStruct
+car1 <- setNames(as.data.frame(t(outer(phi, S, FUN = `^`)[1, , ])),
+                 c("Lower","Correlation","Upper"))
+car1 <- transform(car1, S = S)
+
+car1Plt <- ggplot(car1, aes(x = S, y = Correlation)) +
+    geom_ribbon(aes(ymax = Upper, ymin = Lower),
+                fill = "grey", alpha = 0.3) +
+    geom_line() +
+    ylab(expression(italic(h) * (list(Delta[t], phi))))
+car1Plt
+
 
 ## Predict from model
 newYear <- with(small, data.frame(Year = seq(min(Year), max(Year), length.out = 100)))

### 2016-11-03 functions/simulate.gam.R
updated simulation functions

@@ -1,6 +1,5 @@
 `simulate.gam` <- function(object, nsim = 1, seed = NULL, newdata,
                             freq = FALSE, unconditional = FALSE, ...) {
-    stopifnot(require("MASS"))
     if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE))
         runif(1)
     if (is.null(seed))
@@ -15,8 +14,20 @@
     if (missing(newdata)) {
         newdata <- object$model
     }
-    Rbeta <- mvrnorm(n = nsim, coef(object),
-                     vcov(object, freq = freq, unconditional = unconditional))
+
+    ## random multivariate Gaussian
+    ## From ?predict.gam in mgcv package
+    ## Copyright Simon N Wood
+    ## GPL >= 2
+    rmvn <- function(n, mu, sig) { ## MVN random deviates
+        L <- mroot(sig)
+        m <- ncol(L)
+        t(mu + L %*% matrix(rnorm(m * n), m, n))
+    }
+
+    Rbeta <- rmvn(n = nsim,
+                  mu = coef(object),
+                  sig = vcov(object, freq = freq, unconditional = unconditional))
     Xp <- predict(object, newdata = newdata, type = "lpmatrix")
     sims <- Xp %*% t(Rbeta)
     sims

### 2016-11-03 functions/simulate.gamm.R

@@ -1,6 +1,5 @@
 `simulate.gamm` <- function(object, nsim = 1, seed = NULL, newdata,
                             freq = FALSE, unconditional = FALSE, ...) {
-    stopifnot(require("MASS"))
     if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE))
         runif(1)
     if (is.null(seed))
@@ -15,8 +14,21 @@
     if (missing(newdata)) {
         newdata <- object$gam$model
     }
-    Rbeta <- mvrnorm(n = nsim, coef(object$gam),
-                     vcov(object$gam, freq = freq, unconditional = unconditional))
+
+    ## random multivariate Gaussian
+    ## From ?predict.gam in mgcv package
+    ## Copyright Simon N Wood
+    ## GPL >= 2
+    rmvn <- function(n, mu, sig) { ## MVN random deviates
+        L <- mroot(sig)
+        m <- ncol(L)
+        t(mu + L %*% matrix(rnorm(m * n), m, n))
+    }
+
+    Rbeta <- rmvn(n = nsim,
+                  mu = coef(object$gam),
+                  sig = vcov(object$gam, freq = freq,
+                             unconditional = unconditional))
     Xp <- predict(object$gam, newdata = newdata, type = "lpmatrix")
     sims <- Xp %*% t(Rbeta)
     sims

### 2016-11-03 manuscript.Rmd
minor tweaks
latest additions to manuscript

--- 2016-03-04 manuscript.Rmd
@@
+---
+title: "Modelling palaeoecological time series using generalised additive models"
+author: Gavin L. Simpson
+#author:
+#- name: "Gavin L Simpson"
+#  affiliation: Insitute of Environmental CHange and Society, University of Regina, Regina, SK
+date: November 3, 2016
+bibliography: references.bib
+output:
+  pdf_document:
+    fig_width: 10
+    fig_height: 7
+    fig_crop: false
+    keep_tex: true
+    latex_engine: xelatex
+    template: my.latex
+    md_extensions: +header_attributes+superscript+subscript
+    includes:
+      in_header: header.tex
+#    pandoc_args: [
+#      "--no-tex-ligatures"
+#    ]
+#    template: frontiers-template.tex
+# documentclass: frontiersSCNS # not yet ready for this; need to pandoc-ify the template
 ---
 
----
+```{r knitr-defaults, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
+knitr::opts_chunk$set(comment=NA, fig.align = "center", out.width = "0.8\\linewidth",
+                      echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
+knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
+```
+
+```{r load-packages, cache = FALSE}
+library("mgcv")
+library("ggplot2")
+library("cowplot")
+library("viridis")
+library("grid")                         # for unit.pmax(), unit.list()
+library("gstat")                        # for hscat & variogram
+library("sp")                           # for coordinates()
+```
+
+```{r common-components, cache = FALSE}
+## Set up a default palette
+pal <- viridis(3)
+
+## Default ggplot theme
+theme_set(theme_bw())
+
+## load the simultaneous confidence interval derivatives code
+tmpf <- tempfile()
+download.file("https://gist.githubusercontent.com/gavinsimpson/ca18c9c789ef5237dbc6/raw/295fc5cf7366c831ab166efaee42093a80622fa8/derivSimulCI.R",
+              tmpf, method = "wget")
+source(tmpf)
+
+## Source other functions
+source("functions/simulate.gamm.R")
+```
 
@@
 ## Introduction
+
+* intro about palaeo time series
+
+A linear model for a trend in a series of observations $y_t$ at observation times $x_t$ with $t = 1, 2, \ldots, T$ is
+
+\begin{equation} \label{eq:linear-model}
+y_t = \beta_0 + \beta_1 x_t + \varepsilon_t,
+\end{equation}
+
+where $\beta_0$ is a constant term, the model *intercept* representing the expected value of $y$ where $x$ is 0.
+$\beta_1$ is the *slope* of the best fit line through the data; it measures the rate of change in $y$ for a unit increase in $x$.
+The unknowns, the $\beta_j$ are commonly estimated using least squares by minimising the sum of squared errors, $\varepsilon^2$.
+If we want to ask if the estimated trend $\beta_1$ is statistically significance, a process commonly called *inference*, we make further assumptions about the data (conditional upon the fitted model) or the model errors (or residuals); $\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)$.
+This notation indicates that the residuals $\varepsilon_t$ are *independent* and *identically distributed* Gaussian random variables with mean equal to 0 and constant variance $\sigma^2$.
+In the time series setting, the assumption of independence of model residuals is often violated.
+
+The linear model described above is quite restrictive in terms of the types of trend it can fit; essentially linear increasing or decreasing trends, or trivially, a null trend of no change.
+This model can be extended to allow for non-linear trends, most notably via making $y_t$ depend on polynomials of $x_t$, for example
+
+\begin{align} \label{eq:polynomial-model}
+y_t &= \beta_0 + \beta_1 x_t + \beta_2 x_t^2 + \cdots + \beta_p x_t^p + \varepsilon_t \\
+    &= \beta_0 + \sum_{p = 1}^P \beta_p x_t^p  + \varepsilon_t\\
+    &= \boldsymbol{\beta} \mathbf{X} + \boldsymbol{\varepsilon}
+\end{align}
+
+where polynomials of $x_t$ up to order $P$ are used.
+This model allows for more complex fitted trends but it remains a fully parametric model and suffers from several problems, especially the behaviour of the fitted polynomial at the start and end of the observed series.
+
+## Generalised additive models
+
+A generalised additive models (GAM) is a generalisation of the generalised linear model (GLM), of which the linear model described above is a special case.
+In a GAM, the parametric terms, the $\beta_p$ are replace by a smooth function $f()$ of $x_t$
+
+\begin{equation} \label{eq:additive-model}
+y_t = \beta_0 + f(x_t) + \varepsilon_t
+\end{equation}
+
+The immediate advantage of the GAM is that we are no longer restricted to the types of trends that can be fitted via global polynomial models such as \eqref{eq:polynomial-model}.
+Instead, the form of the fitted trend is estimated from the data itself.
+
+For completeness, where the data are non-Gaussian, such as for where $y_t$ represents a time series of counts of species, we can write \eqref{eq:additive-model} as a GLM/GAM
+
+\begin{align} \label{eq:gam}
+g(\mu) &= \beta_0 + f(x_t) \\
+\mu    &= g^{-1}(\beta_0 + f(x_t)),
+\end{align}
+
+where $\mu$ is the expected value of the random variable $Y$ of which we have observations $y_t$.
+$g$ is the link function, an invertable, monotonic function, such as the natural logarithm, and $g^{-1}$ its inverse.
+The link function maps values from the linear predictor scale on to the scale of the response.
+For example, count data are strictly non-negative and are commonly modelled as a Poisson GLM/GAM using the natural log link function, the inverse of which, the exponential function, maps values to the interval 0--$\infty$.
+In \eqref{eq:gam}, we further assume that the observations are drawn from a member of the exponential family of distributions, such as the Poisson for count data, the binomial for presence/absence or counts from a total.
+Additionally, many software implementations of the above models now allow for the fitting of distributions that are not within the exponential family but which can be fitted using an algorithm superficially similar to the one used to fit GAMs to members of the exponential family.
+Common examples of such extended families include the negative binomial and beta distributions.
+
+### Basis functions
+
+### Smoothness selection
+
+### Smoothing time series data
+
+## δ^15^N time series from Small Water
+
+```{r small-water-example}
+small <- readRDS("./data/small-water/small-water-isotope-data.rds")
+
+## Generate a plot of the data - used as a base later
+plt <- ggplot(small, aes(x = Year, y = d15N)) +
+    geom_point(colour = pal[1]) +
+    theme_bw() +
+    ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
+
+mod <- gamm(d15N ~ s(Year, k = 15), data = small,
+            correlation = corCAR1(form = ~ Year), method = "REML")
+
+## Predict from the fitted model
+newYear <- with(small, data.frame(Year = seq(min(Year), max(Year), length.out = 100)))
+newYear <- cbind(newYear, data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
+
+## Need to do this manually, not through `fd` - write a function
+set.seed(1)
+nsim <- 10000
+take <- 50
+sims <- simulate(mod, nsim = nsim, newdata = newYear, unconditional = TRUE)
+randSims <- sims[, sample(nsim, take)]
+colnames(randSims) <- paste0("sim", seq_len(take))
+randSims <- setNames(stack(as.data.frame(randSims)), c("simulated", "run"))
+randSims <- transform(randSims, Year = rep(newYear$Year, take),
+                      simulated = simulated)
+
+## Add a CI to the fitted spline
+CI <- apply(sims, 1L, quantile, probs = c(0.025, 0.975))
+newYear <- transform(newYear, lower = CI[1,], upper = CI[2,])
+
+## Plot simulated trends
+plt.sim <- ggplot(newYear, aes(x = Year, y = fit)) +
+    geom_line(data = randSims, mapping = aes(y = simulated, x = Year, group = run),
+              colour = pal[1], alpha = 0.1) +
+    geom_line(colour = pal[1]) +
+    theme_bw() +
+    ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter))) +
+    xlab("Year")
+
+## Derivatives
+fd <- derivSimulCI(mod, samples = 10000, n = 100)
+CI <- apply(fd[[1]]$simulations, 1, quantile, probs = c(0.025, 0.975))
+sigD <- signifD(fd[["Year"]]$deriv, fd[["Year"]]$deriv, CI[2, ], CI[1, ],
+                eval = 0)
+newYear <- transform(newYear,
+                     derivative = fd[["Year"]]$deriv[, 1], # computed first derivative
+                     fdUpper = CI[2, ],                    # upper CI on first deriv
+                     fdLower = CI[1, ],                    # lower CI on first deriv
+                     increasing = sigD$incr,               # where is curve increasing?
+                     decreasing = sigD$decr)               # ... or decreasing?
+
+## Plot it
+derivPlt <- ggplot(newYear, aes(x = Year, y = derivative)) +
+    geom_ribbon(aes(ymax = fdUpper, ymin = fdLower), alpha = 0.3, fill = pal[1]) +
+    geom_line(colour = pal[1]) +
+    geom_line(aes(y = increasing), colour = pal[1], size = 1.5) +
+    geom_line(aes(y = decreasing), colour = pal[1], size = 1.5) +
+    ylab(expression(italic(hat(f) * "'") * (Year))) +
+    xlab("Year")
+
+## Plot of data, fitted spline, CI, and significant periods of change
+## Need the signifD helper function to generate the vectors for increasing
+## and decreasing periods
+sigDYear <- signifD(newYear$fit, fd[["Year"]]$deriv[,1], CI[2, ], CI[1, ],
+                    eval = 0)
+newYear <- transform(newYear,
+                     YearIncr = sigDYear$incr,
+                     YearDecr = sigDYear$decr)
+
+plt.fit2 <- ggplot(small, aes(x = Year, y = d15N)) +
+    geom_ribbon(data = newYear,
+                mapping = aes(ymax = upper, ymin = lower, y = fit, x = Year),
+                fill = pal[1], alpha = 0.3) +
+    geom_line(data = newYear, aes(y = fit, x = Year), colour = pal[1]) +
+    geom_point(colour = pal[1]) +
+    theme_bw() +
+    ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter))) +
+    geom_line(data = newYear, mapping = aes(y = YearIncr), size = 1.5) +
+    geom_line(data = newYear, mapping = aes(y = YearDecr), size = 1.5)
+
+## plot each in a 2x2 display
+smallWaterPlt <- plot_grid(plt, plt.sim, derivPlt, plt.fit2, labels = "AUTO",
+                           align = "vh")
+
+## summary object for use in document
+smallSumm <- summary(mod$gam)
+
+## estimate of phi and confidence interval
+smallPhi <- intervals(mod$lme, which = "var-cov")$corStruct
+
+## plot CAR(1) process
+maxS <- with(small, diff(range(Year))) ## too large, truncate to 50
+S <- seq(0, 50, length = 100)
+phi <- intervals(mod$lme, which = "var-cov")$corStruct
+car1 <- setNames(as.data.frame(t(outer(phi, S, FUN = `^`)[1, , ])),
+                 c("Lower","Correlation","Upper"))
+car1 <- transform(car1, S = S)
+
+car1Plt <- ggplot(car1, aes(x = S, y = Correlation)) +
+    geom_ribbon(aes(ymax = Upper, ymin = Lower),
+                fill = "grey", alpha = 0.3) +
+    geom_line() +
+    ylab(expression(italic(h) * (list(Delta[t], phi)))) +
+    xlab(expression(Delta[t]))
+```
+
+```{r small-water-display, fig = TRUE, fig.cap = "Stable isotopes"}
+smallWaterPlt
+```
+
+Figure \ref{fig:small-water-display}A shows `r nrow(small)` nitrogen stable isotope measurements from the bulk organic matter of a core collected from Small Water, a small corrie lake located in the English Lake District, UK.
+The data are shown on a ^210^Pb time scale.
+Common questions that might be asked about this series are; what is the trend in δ^15^N?, when do we first see evidence for a change in δ^15^N?, and is the reversal in δ^15^N values in the uppermost section of the core a real change?
+
+The trend in δ^15^N values is clearly non-linear and it would be difficult to suggest a suitable polynomial model that would allow for periods of relatively no change in δ^15^N as well as rapid change.
+Instead, a GAM is ideally suited to modelling such trends; the data suggest a smoothly varying change in δ^15^N between 1925 and 1975.
+It is reasonable to expect some autocorrelation in the model errors about the fitted trend.
+Therefore I fitted the following GAM to the δ^15^N time series.
+
+\begin{equation} \label{eq:small-gam}
+y_t = \beta_0 + f(x_t) + \varepsilon, \quad \varepsilon \sim \mathcal(0, \boldsymbol{\Lambda}\sigma^2)
+\end{equation}
+
+Note that now I have relaxed the i.i.d.
+assumption and have introduced $\boldsymbol{\Lambda}$, a corrleation matrix that is used to model any remaining autocorrelation in the residuals.
+The Small Water δ^15^N values are irregularly spaced in time and as a result a correlation structure that can handle the uneven spacing of the samples.
+A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the autoregressive process and simple stated states that the correlation between any two residuals is $\phi^d$, where $d$ is the amount of separation in time between the residuals.
+$d$ may be a real valued number in the CAR(1), which accomodates the irregular separation of samples.
+$\phi$ controls how quickly the correlation between any two residuals declines as a function of their separation in time.
+$\phi$ is an additional paramter that will be estimated during model fitting.
+The model in \ref{eq:small-gam} was fitted using the `gamm()` function in the **mgcv** package for R.
+
+The fitted trend is shown in Figure \ref{fig:small-water-display}B, and well-captures the strong pattern in the data.
+The trend is statistically significant (estimated degrees of freedom = `r round(smallSumm[["s.table"]][1,2], 3)`; $F$ = `r round(smallSumm[["s.table"]][1,3], 3)`, approximate $p$ value = $\ll$ 0.0001).
+However further analysis of the fitted model is required to answer the other questions posed earlier about the timing of change and whether features in the trend can be distinguished from random noise.
+
+### Simultaneous confidence intervals
+
+The first step in answer these and other questions is to form a confidence about the fitted trend.
+Standard theory can be used to compute point-wise intervals.
+Such intervals have the appropriate 1 - $\alpha$ confidence for a single point only and when considering multiple points, as we will here, suffer from the problem of multiple comparison; when considered over all observations or points, the confidence interval is much smaller than the required 1 - $\alpha$.
+
+Instead a 1 - $\alpha$ *simultaneous* confidence interval is required.
+One way to compute this interval is via posterior simulation from the fitted model.
+As described above, the GAM involves finding estimates for the coefficients of the spline basis functions given the value of the smoothing parameter.
+These spline coefficients form a multivariate normal distribution with mean vector and covariance matrix specified by the model estimates of the cofficients and their covariances.
+Having parameterised the multivariance normal distribution, random draws from the distribution can be taken.
+Each random draw represents a new trend that is consistent with the consistent with the fitted trend but it also includes the uncertainty in the estimated trend.
+Figure \ref{fig:small-water-display}B shows `r take` random trends drawn from the posterior distribution of the fitted model.
+
+The simultaneous confidence interval itself is based on `r format(nsim, big.mark = ",")` samples from the model posterior.
+The fitted value of the trend for each of the posterior samples was evaluated at `r nrow(newYear)` equally spaced time points over the period of observation.
+At each point, the 0.025 and 0.975 probability quantiles of the distribution of posterior samples was obtained; these extrema contain 95% of the simulated trends and thus give the 95% simultaneous confidence interval on the fitted trend.
+This interval is shown in Figure \ref{fig:small-water-display}D.
+
+### Identifying periods change
+
+In the simple linear trend model \eqref{eq:linear-model} whether the estimated trend constitutes evidence for or against a null hypothesis of no change rests on how large the estimated rate of change in $y_t$ is ($\hat{\beta}_1$) relative to its uncertainty.
+This is summarised in the $t$ statistic.
+As the rate of change in $y_t$ is constant over the fitted trend --- there is only a singe slope for the fitted trend $\hat{\beta}_1$ --- if the $t$ statistic of the test that $\hat{\beta}_1 = 0$ is unusually extreme we can conclude that the trend is statistically significant and this applies to the entire time series of data.
+The estimate $\hat{\beta}_1$ is also the first derivative of the fitted trend.
+
+In the GAM, the fitted trend is non-linear and hence the slope of the trend is potentially different at every point in the time series.
+As such we can reasonably ask where in the fitted trend is the response $y_t$ changing.
+Mirroring the simple linear model we can ask whether answer this question by determining whether or not the first derivative, the slope, of the fitted trend at any time point is consistent with a null hypothesis of no change; i.e.
+whether or not the first derivative is indistinguishable from a value of 0 given the uncertainty in the estimate of the derivative.
+
+Derivatives of the fitted spline are not easily available analytically, but they can be estimated using the method of finite differences.
+Two values, separated by a very small time-shift $\Delta_t$, of the fitted trend are predicted from the model; the difference between the fitted values for the two time points is an approximation of the true derivative of the trend.
+As $\Delta_t \rightarrow 0$ the approximation becomes more and more accurate.
+The derivative of the fitted trend is evaluated at large number of time points over the series.
+An approximate 1 - $\alpha$ pointwise confidence interval can be calculated for the derivative estimates using standard theory and the covariance matrix of the spline coefficients.
+
+Instead, a 1 - $\alpha$ simultaneous confidence interval can be computed using the posterior simulation method described above.
+This was done for the Small Water model for the `r format(nsim, big.mark = ",")` simulated trends by evaluating them at a large number of time points and computing the finite differences.
+The 0.025 and 0.975 probability quantiles of the distributions of derivative estimates at each evaluation point were computed and form the 95% simultaneous confidence for the first derivative of the fitted trend.
+The first derivative of the fitted trend and its 95% simultaneous confidence interval infor the δ^15^N time series are shown in Figure \ref{fig:small-water-display}C.
+Although the data and fitted trend suggest a slight increase in δ^15^N between 1850  and 1925, the estimated first derivative is consistent with a value of 0; hence there is little evidence against the null hypothesis of no change in δ^15^N over this period.
+
+From ~1925 the derivative of of the trend becomes negative, however it is not until ~1940 that the confidence interval doesn't include 0.
+At this point we have evidence to reject the null hypothesis of no change.
+This time point is taken as the first evidence for change in δ^15^N in the Small Water core.
+The confidence interval on the first derivatve of the trend in δ^15^N is bounded away from 0 between ~1940 and 1975, covering the major decline in values clearly evident in the observations.
+This period of significant change is indicated by bold sections on the fitted trend in Figure \ref{fig:small-water-display}C and D.
+The confidence interval includes 0 between 1975 and around the year 2000, after which time the estimated first derivative becomes positive, suggesting a small but significant increase in the δ^15^N over the last 10 years of the time series, although caution must exercised when interpreting N stable isotope values in the partly decomposed organic matter commonly encountered in the most recent period of the sedimentary record.
+
+Figure Figure \ref{fig:small-water-display}D summarises the main elements of the GAM analysis of the Small Water δ^15^N time series showing the i) fitted trend, ii) 95% simultaneous confidence interval, and iii) estimated periods of change.
+
+### Residual autocorrelation
+
+```{r car1-plot, fig = TRUE, fig.height = 7, fig.width = 10, fig.cap = "Estimated CAR(1) process from the GAM fitted to the Small Water $\\delta^{15}\\text{N}$ time series.
+$h(\\Delta_t, \\phi)$ is the correlation between residuals separated by $\\Delta_t$ years, where $\\hat{\\phi} = \\text{0.6}$.
+The shaded band is a 95\\% confidence interval on the estimated correlation $h$."}
+car1Plt
+```
+
+The estimated value of $\phi$ from the fitted model is `r round(smallPhi[2], 2)` with 95% confidence interval `r round(smallPhi[1], 2)`--`r round(smallPhi[3], 2)`, indicating moderate to strong residual autocorrelation about the fitted trend.
+Failure to account for the dependencies in the δ^15^N time series could lead have lead to the estimation of a more wiggly trend than the one shown in Figure \ref{fig:small-water-display}B which would negatively impact the confidence place on the inferences we might draw from the fitted model.
+Importantly, failing to account for the strong dependency in the residuals would lead to smaller uncertainties in the estimated spline coefficients, which would propagate through to narrower confidence intervals on the fitted trend and on the first derivatives, and ultimately to the identification of significant periods of change in the δ^15^N time series.
+
+## Braya-sø
+
+```{r braya-so-example}
+## Source other functions
+source("functions/simulate.gam.R")
+
+## Data
+braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt", skip = 84)
+names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
+braya <- transform(braya, sampleInterval = YearYoung - YearOld)
+
+## Generate a plot of the data
+ylabel <- expression(italic(U)[37]^{italic(k)})
+brayaData.plt <- ggplot(braya, aes(x = Year, y = UK37)) +
+    geom_point(colour = pal[1]) +
+    geom_line(colour = pal[1]) +
+    theme_bw() +
+    ylab(ylabel)
+
+## fit the car(1) model
+braya.car1 <- gamm(UK37 ~ s(Year, k = 20), data = braya, correlation = corCAR1(form = ~ Year),
+                   method = "REML")
+braya.gcv <- gam(UK37 ~ s(Year, k = 30), data = braya)
+
+## car(10 parameter
+brayaPhi <- intervals(braya.car1$lme)$corStruct
+
+## ggplot with data and fitted spline, then resids vs time in second panel
+ylabel <- expression(italic(U)[37]^{italic(k)})
+N <- 300
+newYear <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = N)))
+newYear <- cbind(newYear, data.frame(predict(braya.car1$gam, newYear, se.fit = TRUE)))
+crit.t <- qt(0.975, df = df.residual(braya.car1$gam))
+newYear <- transform(newYear,
+                     upper = fit + (crit.t * se.fit),
+                     lower = fit - (crit.t * se.fit))
+## add GAM GCV results
+fit.gcv <- predict(braya.gcv, newdata = newYear, se.fit = TRUE)
+newYear <- rbind(newYear, newYear)      # extend newYear to take GCV results
+newYear[seq(N+1, length.out = N, by = 1), ]$fit <- fit.gcv$fit
+newYear[seq(N+1, length.out = N, by = 1), ]$upper <-
+    fit.gcv$fit + (qt(0.975, df.residual(braya.gcv)) * fit.gcv$se.fit)
+newYear[seq(N+1, length.out = N, by = 1), ]$lower <-
+    fit.gcv$fit - (qt(0.975, df.residual(braya.gcv)) * fit.gcv$se.fit)
+newYear <- transform(newYear, Method = rep(c("GAMM (CAR(1))", "GAM (GCV)"), each = N))
+
+## plot CAR(1) and GCV fits
+braya.p1 <- ggplot(braya, aes(y = UK37, x = Year)) +
+    geom_point() +
+    geom_ribbon(data = newYear,
+                mapping = aes(y = fit, x = Year, ymax = upper, ymin = lower, fill = Method),
+                alpha = 0.4) +
+    geom_line(data = newYear, mapping = aes(y = fit, x = Year, colour = Method), size = 1) +
+    ylab(ylabel) +
+    theme_bw() +
+    scale_color_manual(values = viridis(11)[c(2, 6)]) +
+    scale_fill_manual(values = viridis(11)[c(2, 6)]) +
+    theme(legend.position = "top")
+
+## Generate GCV and REML traces
+lambda <- 1e-8
+gcv <- reml <- numeric(length = 100)
+
+for (i in seq_along(gcv)) {
+    mGCV <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "GCV.Cp", sp = lambda)
+    mREML <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "REML", sp = lambda)
+    gcv[i] <- mGCV$gcv.ubre
+    reml[i] <- mREML$gcv.ubre
+    lambda <- lambda * 1.5
+}
+
+## GCV and REML score traces
+result <- data.frame(lambda = 1e-8 * 1.5^{0:99}, score = c(gcv, reml),
+                     criterion = rep(c("GCV", "REML"), each = 100))
+
+## Full fits to get estimated models
+fullGCV  <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "GCV.Cp")
+fullREML <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "REML")
+obsSP <- data.frame(lambda = c(fullGCV$sp, fullREML$sp), criterion = c("GCV", "REML"))
+
+## Plot
+trace.plt <- ggplot(result, aes(x = lambda, y = score)) +
+    geom_line() +
+    facet_wrap(~ criterion, scales = "free_y") + theme_bw() + scale_x_log10() +
+    xlab(expression(lambda)) +
+    ylab("Score") +
+    geom_vline(aes(xintercept = lambda), obsSP, colour = viridis(3)[2])
+
+g.braya.p1 <- ggplotGrob(braya.p1) # convert to gtable
+g.trace.plt <- ggplotGrob(trace.plt) # convert to gtable
+
+if(getRversion() < "3.3.0"){
+  # We need to convert the widths to unit lists for the subsequent
+  # manipulations to be possible.
+  # Once R 3.3.0 is released, this will not be necessary anymore.
+  g.braya.p1$widths <- grid:::unit.list(g.braya.p1$widths)   
+  g.trace.plt$widths <-  grid:::unit.list(g.trace.plt$widths)
+}
+
+braya.p1.widths <- g.braya.p1$widths[1:3] # extract the first three widths, 
+                                          # corresponding to left margin, y lab, and y axis
+trace.plt.widths <- g.trace.plt$widths[1:3] # same for trace.plt plot
+max.widths <- unit.pmax(braya.p1.widths, trace.plt.widths) # calculate maximum widths
+g.braya.p1$widths[1:3] <- max.widths # assign max widths to braya.p1 gtable
+g.trace.plt$widths[1:3] <- max.widths # assign max widths to trace.plt gtable
+
+# plot_grid() can work directly with gtables, so this works
+brayaFail.plt <- plot_grid(g.braya.p1, g.trace.plt, labels = "AUTO", ncol = 1, rel_heights = c(0.55, 0.45))
+```
+
+The second example time series is a 6,000 year record of alkenone unsaturation, \uk, from Braya-sø, a meromictic lake in  West Greenland [@DAndrea2011-oa].
+Alkenones are long-chained unsaturated organic compounds that are produced by a small number of planktonic organisms known as haptophytes.
+The \uk{} unsaturation index is
+
+\begin{equation*}
+\uk = \frac{[C_{37:2}] - [C_{37:4}]}{[C_{37:2}] + [C_{37:3}]+ [C_{37:4}]}
+\end{equation*}
+
+where $[C_{37:x}]$ is the concentration of the alkenone with 37 carbon atoms and $x$ double carbon bonds.
+The relative abundance of these alkenones is known to vary with changes in water temperature, and as a result \uk{} is used as a proxy for lake- and sea-surface temperatures.
+For further details on the Braya-sø \uk{} record and age model see @DAndrea2011-oa.
+Here I use the 3,000 year \uk{} record from the PAGES 2K database [@PAGES_2k_Consortium2013-fm].
+The data are presented in Figures \ref{fig:braya-so-car1-gcv-fits}A and \ref{fig:braya-display}A.
+
+```{r braya-so-car1-gcv-fits, cache = FALSE, fig = TRUE, fig.height = 7, fig.width = 10, fig.cap = "Fitted trends using CAR(1) and GCV models"}
+brayaFail.plt
+```
+
+The \uk{} data present a more difficult data analysis challenge than the δ^15^N time series from Small Water because of the much more complex variations present.
+Fitting the same model as the Small Water example, \ref{eq:small-gam}, to the \uk{} data resulted in the unsatisfactory fit shown as the very smooth line in Figure \ref{fig:braya-so-car1-gcv-fits}A.
+Further problems were evident with this model fit --- the covariance matrix of the variance-covariance terms in the model was non-positive definite, a sure sign of problems with the fitted model.
+Refitting with a smaller basis dimension (*k* = 20) for the trend term resulted in a model with a positive-definite covariance matrix for the model variance-covraiance terms, but the estimate of the CAR(1) parameter $\phi$ = `r round(brayaPhi[2], 2)` was exceedingly uncertain (95% confidence interval `r round(brayaPhi[1], 2)` -- `r round(brayaPhi[3], 2)`!)
+
+Fitting this model as a standard GAM with REML smoothness selection resulted in the same fitted trend as the GAM with CAR(1) errors (not shown), whilst using GCV smoothness selection resulted in a much more satisfactory fitted trend, yet normally, we would prefer REML smoothness selection.
+There are two potential problems with the GCV-selected trend: i) GCV is sensitive to the profile of the GCV score and has been shown to badly undersmooth data in situations where the profile is flat around the minimum GCV score, and ii) the model fitted assumes that the observations are independent, an assumption that is certainly violated in the \uk{} time series.
+
+The GCV and REML scores for an increasing sequence of values of the smoothness penalty $\lambda$ were evaluated for for the standard GAM fit to the \uk{} time series.
+The resulting profiles are shown in Figure \ref{fig:braya-so-car1-gcv-fits}B, with the selected value of the penalty shown by the vertical line.
+The GCV score profile suggests that the potential for undersmoothing identified by Wood is unlikely to apply here.
+Addressing the violation of the independence assumption is more difficult; the standard errors of the model coefficients are most likely anti-conservative, which presents a significant problem if we wish to determine which of the wiggles in the fitted \uk{} trend can be identified as real.
+
+To understand the reason why the GAM plus CAR(1) and the GAM with REML smoothness selection performed poorly with the \uk{} time series we need to delve a little deeper into what is happening when we are fitting these two models.
+
+### Identifiability and induced correlations
+
+Consider again the basic GAM for a smooth trend, \eqref{eq:additive-model}.
+In that equation the correlation matrix $\boldsymbol{\Lambda}$ was ommitted for the sake of simplicity.
+Here I reintroduce it and restate the distributional assumptions of this model
+
+\begin{equation}
+y_t = \beta_0 + f(x_t) + \varepsilon_t, \quad \varepsilon \sim \mathcal(0, \boldsymbol{\Lambda}\sigma^2)
+\end{equation}
+
+In the basic GAM, $\boldsymbol{\Lambda} \equiv \mathbf{I}$ is an identity matrix, a matrix with 1s on the diagonal and 0s elsewhere.
+This is where the independence assumption of the model comes from; a model residual is perfectly correlated with itself (the 1s on the diagonal), but uncorrelated with any other residual (the off-diagonal 0s).
+In the GAM plus CAR(1) model, I specified an alternative correlation function for $\boldsymbol{\Lambda}$ --- the CAR(1) with correlation parameter $\phi$.
+@Fahrmeir2008-si show that where the stochastic structure of $f$ and $\boldsymbol{\Lambda}$ approach one another, where we have a potentially wiggly trend or strong autocorrelation as $\phi \rightarrow 1$, the two processes can quickly become unidentifiable [see also @Fahrmeir2013-xu **give pages**].
+By unidentifiable, I mean that it becomes increasingly difficult to distinguish between a wiggly trend or strong autocorrelation because these two processes are very similar to one another in appearance.
+
+Recall that the estimate of the autocorrelation parameter in the CAR(1) for the \uk{} series was $\hat{\phi}$ = `r round(brayaPhi[2], 2)`.
+It seems unlikely therefore that the CAR(1) process and the trend are unidentifiable in this instance.
+The large degree of uncertainty in $\hat{\phi}$ suggests that the fitted model is too complex, containing too many similar terms.
+That the GAM plus CAR(1) model and the GAM with REML smoothness selection, which lacks the CAR(1) process, both resulted in the same fitted trend suggests that the problem lies elsewhere.
+
+When fitting GAMs using REML smoothness selection, the splines are treated as random effects.
+More specifically, the entirely smooth components of the spline basis enter the model as fixed effects and the wiggly or non-smooth basis functions enter as random effects terms.
+Choosing the smoothness parameter $\lambda$ is then a matter of estimating variance of the random effect terms.
+If you are familiar with mixed effects models, you will know that random effects incude correlations between observations.
+In a mixed effects moedl with a simple random intercept term for a grouping variable, the induced correlation is
+
+$$\rho = \sqrt{\sigma^2_{B} / \sigma^2_{T}}$$
+
+where $\sigma^2_{B}$ is the between group variance (or the variance of the distribution of random intercepts), and $\sigma^2_{T}$ is the total variance in the data, and $\rho$ is the correlation between any two observations in the same group.
+
+Treating the wiggly parts of a spline basis as random effects results in similar but more complex induced correlations [@Fahrmeir2008-si].
+It seems likely that these induced correlations between observations are modelling components of the data which we might rightly wish to consider as trend.
+These induced correlations provide a complex structure to $\boldsymbol{\Lambda}$, and the identifiability issue discussed previously again applies, but from a slightly different source.
+
+```{r induced-correlations, cache = FALSE}
+knitr::read_chunk("./analysis/braya-so/braya-so-induced-correlations.R")
+```
+```{r induced-correlations-plt, fig = TRUE, fig.cap = "Induced correlations"}
+```
+
+Ideally, we would use a method account for autocorrelation in the data and yet not suffer from the same identifiability problems described here.
+In the next section I look at one such model, which goes some way to addressing this issue.
+
+### Gaussian process smooths
+
+```{r fit-braya-example-gp-smooths}
+m.reml <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "REML")
+resi.reml <- data.frame(Residuals = resid(m.reml), Year = braya$Year, Dummy = 1)
+coordinates(resi.reml) <- ~ Year + Dummy
+lagScatter <- hscat(Residuals ~ 1, data = resi.reml, breaks = c(0, 25, 50, 100,200, 500, 1000))
+resi.vgmCloud <- variogram(Residuals ~ 1, data = resi.reml, cloud = TRUE)
+resi.vgm <- variogram(Residuals ~ 1, data = resi.reml, width = 20, cressie = TRUE)
+
+variogPlt <- ggplot(resi.vgm, aes(x = dist, y = gamma)) +
+    geom_smooth(se = FALSE, col = "grey", span = 0.75,
+                method.args = list(family = "gaussian")) +
+    geom_point() + 
+    xlim(c(0, 1000)) + ylim(c(0, 1e-3)) +
+    ylab(expression(Semivariance ~~ gamma)) +
+    xlab("Temporal separation (years)")
+
+## Fit gaussian process spline models
+effRange <- 250                         # sets effective range in years for Matern correl
+mod.gp <- gam(UK37 ~ s(Year, k = 30, bs = "gp", m = c(3, effRange)), data = braya,
+              method = "REML")
+## Add time interval of each sample as prior weights; effectively averaging more time per sample
+mod2.gp <- gam(UK37 ~ s(Year, k = 30, bs = "gp", m = c(3, effRange)), data = braya,
+               method = "REML", weights = sampleInterval)
+
+## Predict from model
+N <- 500
+newBraya <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = N)))
+newBraya <- cbind(newBraya, data.frame(predict(mod.gp, newBraya, se.fit = TRUE)))
+
+## Draw the fitted spline on the data
+ylabel <- expression(italic(U)[37]^{italic(k)})
+brayaFit.plt <- ggplot(braya, aes(x = Year, y = UK37)) +
+    geom_line(data = newYear, aes(y = fit, x = Year), colour = pal[1]) +
+    geom_point(colour = pal[1]) +
+    theme_bw() +
+    ylab(ylabel)
+
+## posterior simulation
+set.seed(1)
+nsim <- 10000
+take <- 50
+sims <- simulate(mod.gp, nsim = nsim, newdata = newBraya, unconditional = TRUE)
+randSims <- sims[, sample(nsim, take)]
+colnames(randSims) <- paste0("sim", seq_len(take))
+randSims <- setNames(stack(as.data.frame(randSims)), c("simulated", "run"))
+randSims <- transform(randSims, Year = rep(newBraya$Year, take),
+                      simulated = simulated)
+
+## Add a CI to the fitted spline
+CI <- apply(sims, 1L, quantile, probs = c(0.025, 0.975))
+newBraya <- transform(newBraya, lower = CI[1,], upper = CI[2,])
+
+brayaSim.plt <- ggplot(newBraya, aes(x = Year, y = fit)) +
+    geom_line(data = randSims, mapping = aes(y = simulated, x = Year, group = run),
+              colour = pal[1], alpha = 0.1) +
+    geom_line(colour = pal[1]) +
+    theme_bw() +
+    ylab(ylabel) +
+    xlab("Year")
+
+## Derivatives
+fd <- derivSimulCI(mod.gp, samples = 10000, n = N)
+CI <- apply(fd[[1]]$simulations, 1, quantile, probs = c(0.025, 0.975))
+sigD <- signifD(fd[["Year"]]$deriv, fd[["Year"]]$deriv, CI[2, ], CI[1, ],
+                eval = 0)
+newBraya <- transform(newBraya,
+                     derivative = fd[["Year"]]$deriv[, 1], # computed first derivative
+                     fdUpper = CI[2, ],                    # upper CI on first deriv
+                     fdLower = CI[1, ],                    # lower CI on first deriv
+                     increasing = sigD$incr,               # where is curve increasing?
+                     decreasing = sigD$decr)               # ... or decreasing?
+
+## Plot it
+braya.derivPlt <- ggplot(newBraya, aes(x = Year, y = derivative)) +
+    geom_ribbon(aes(ymax = fdUpper, ymin = fdLower), alpha = 0.3, fill = pal[1]) +
+    geom_line(colour = pal[1]) +
+    geom_line(aes(y = increasing), colour = pal[1], size = 1.5) +
+    geom_line(aes(y = decreasing), colour = pal[1], size = 1.5) +
+    ylab(expression(italic(hat(f) * "'") * (Year))) +
+    xlab("Year") +
+    theme_bw()
+
+## Plot of data, fitted spline, CI, and significant periods of change
+## Need the signifD helper function to generate the vectors for increasing
+## and decreasing periods
+sigDYear <- signifD(newBraya$fit, fd[["Year"]]$deriv[,1], CI[2, ], CI[1, ],
+                    eval = 0)
+newBraya <- transform(newBraya,
+                     YearIncr = sigDYear$incr,
+                     YearDecr = sigDYear$decr)
+
+brayaFit2.plt <- ggplot(braya, aes(x = Year, y = UK37)) +
+    geom_ribbon(data = newBraya,
+                mapping = aes(ymax = upper, ymin = lower, y = fit, x = Year),
+                fill = pal[1], alpha = 0.3) +
+    geom_line(data = newBraya, aes(y = fit, x = Year), colour = pal[1]) +
+    geom_point(colour = pal[1]) +
+    theme_bw() +
+    ylab(ylabel) +
+    geom_line(data = newBraya, mapping = aes(y = YearIncr), size = 1.5, colour = pal[1]) +
+    geom_line(data = newBraya, mapping = aes(y = YearDecr), size = 1.5, colour = pal[1])
+
+brayaPlt <- plot_grid(brayaData.plt, brayaSim.plt, braya.derivPlt, brayaFit2.plt, labels = "AUTO",
+                      align = "vh")
+```
+
+```{r variogram-braya, fig = TRUE, fig.cap = "Variogram", cache = FALSE}
+variogPlt
+```
+
+```{r braya-display, fig = TRUE, fig.cap = "Braya-sø", cache = FALSE}
+brayaPlt
+```
+
+### Sediment compaction & time averaging
+
+## Further considerations
+
+### Multivariate data
+
+## Discussion
+
+## References

### 2018-04-27 manuscript.Rmd
latest version of the manuscript

--- 2016-03-04 manuscript.Rmd
@@
+---
+title: "Modelling palaeoecological time series using generalised additive models"
+author: Gavin L. Simpson
+#author:
+#- name: "Gavin L Simpson"
+#  affiliation: Insitute of Environmental Change and Society, University of Regina, Regina, SK
+date: April 26, 2018
+bibliography: references.bib
+geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
+fontsize: 12pt
+fontfamily: stix
+mainfont: XITS
+output:
+  pdf_document:
+    fig_width: 8
+    fig_height: 5
+    fig_crop: false
+    keep_tex: true
+    latex_engine: xelatex
+    template: my.latex
+    md_extensions: +header_attributes+superscript+subscript
+    includes:
+      in_header: header.tex
+#    pandoc_args: [
+#      "--no-tex-ligatures"
+#    ]
+#    template: frontiers-template.tex
+# documentclass: frontiersSCNS # not yet ready for this; need to pandoc-ify the template
 ---
 
----
+```{r knitr-defaults, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
+knitr::opts_chunk$set(comment=NA, fig.align = "center", out.width = "0.8\\linewidth",
+                      echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
+knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
+```
+
+```{r load-packages, cache = FALSE}
+library("mgcv")
+library("scam")
+library("ggplot2")
+library("cowplot")
+library("grid")                         # for unit.pmax(), unit.list()
+## library("sp")                           # for coordinates()
+library("schoenberg")
+library("tidyr")
+```
+
+```{r common-components, cache = FALSE}
+## Default ggplot theme
+theme_set(theme_bw())
+
+## source Small Water data
+small <- readRDS("./data/small-water/small-water-isotope-data.rds")
+
+## load braya so data set
+braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt", skip = 84)
+names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
+braya <- transform(braya, sampleInterval = YearYoung - YearOld)
+
+## plot labels
+d15n_label <- expression(delta^{15}*N)
+braya_ylabel <- expression(italic(U)[37]^{italic(k)})
+```
 
@@
 ## Introduction
+
+* intro about palaeo time series
+
+### Data
+
+intro on the two data sets goes here.
+
+#### δ^15^N time series from Small Water
+
+Figure \ref{fig:data-figure}a shows `r nrow(small)` nitrogen stable isotope measurements on the bulk organic matter of a sediment core collected from Small Water, a small corrie lake located in the English Lake District, UK.
+The data were collected to investigate disturbance of nitrogen (N) cycling in remote, oligotrophic lakes by N deposited from the atmosphere (Simpson, unpublished data, REF Curtis et al FUMBLE report).
+The data are shown on a ^210^Pb time scale.
+Questions that might be asked about this series are; what is the trend in δ^15^N?, when do we first see evidence for a change in δ^15^N?, and is the reversal in δ^15^N values in the uppermost section of the core a real change?
+
+#### Braya-sø alkenone time series
+
+The second example time series is a 6,000 year record of alkenone unsaturation, \uk, from Braya-sø, a meromictic lake in  West Greenland [@DAndrea2011-oa].
+Alkenones are long-chained unsaturated organic compounds that are produced by a small number of planktonic organisms known as haptophytes.
+The \uk{} unsaturation index is
+
+\begin{equation*}
+\uk{} = \frac{[C_{37:2}] - [C_{37:4}]}{[C_{37:2}] + [C_{37:3}]+ [C_{37:4}]}
+\end{equation*}
+
+where $[C_{37:x}]$ is the concentration of the alkenone with 37 carbon atoms and $x$ double carbon bonds.
+The relative abundance of these alkenones is known to vary with changes in water temperature, and as a result \uk{} is used as a proxy for lake- and sea-surface temperatures.
+For further details on the Braya-sø \uk{} record and age model see @DAndrea2011-oa.
+Here I use the 3,000 year \uk{} record from the PAGES 2K database [@PAGES_2k_Consortium2013-fm].
+The data are presented in Figure \ref{fig:data-figure}b.
+
+```{r data-figure, fig = TRUE, fig.width = 8, fig.height = 5, fig.cap = "data plot"}
+## Generate a plot of the data - used as a base later
+small_plt <- ggplot(small, aes(x = Year, y = d15N)) +
+    geom_point() +
+    labs(y = d15n_label, x = "Year CE")
+
+## Generate a plot of the data
+braya_plt <- ggplot(braya, aes(x = Year, y = UK37)) +
+    geom_line(colour = "grey") +
+    geom_point() +
+    labs(y = braya_ylabel, x = "Years BP")
+
+plot_grid(small_plt, braya_plt, ncol = 1, labels = "auto", align = "hv", axis = "lr")
+```
+
+### Regression models for palaeoenvironmental time series
+
+A linear model for a trend in a series of $T$ observations $y_t$ at observation times $x_t$ with $t = 1, 2, \ldots, T$ is
+
+\begin{equation} \label{eq:linear-model}
+y_t = \beta_0 + \beta_1 x_t + \varepsilon_t,
+\end{equation}
+
+where $\beta_0$ is a constant term, the model *intercept*, representing the expected value of $y_t$ where $x_t$ is 0.
+$\beta_1$ is the *slope* of the best fit line through the data; it measures the rate of change in $y$ for a unit increase in $x$.
+The unknowns, the $\beta_j$ are commonly estimated using least squares by minimising the sum of squared errors, $\sum_t \varepsilon_t^2$.
+If we want to ask if the estimated trend $\beta_1$ is statistically significant, a process called *inference*, we must make further assumptions about the data (conditional upon the fitted model) or the model errors (residuals); $\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)$.
+This notation indicates that the residuals $\varepsilon_t$ are *independent* and *identically distributed* Gaussian random variables with mean equal to $0$ and constant variance $\sigma^2$.
+In the time series setting, the assumption of independence of model residuals is often violated.
+
+The linear model described above is quite restrictive in terms of the types of trend it can fit; essentially linear increasing or decreasing trends, or trivially, a null trend of no change.
+This model can be extended to allow for non-linear trends, most notably via making $y_t$ depend on polynomials of $x_t$, for example
+
+\begin{align} \label{eq:polynomial-model}
+y_t &= \beta_0 + \beta_1 x_t + \beta_2 x_t^2 + \cdots + \beta_P x_t^P + \varepsilon_t \\
+    &= \beta_0 + \sum_{p = 1}^P \beta_p x_t^p  + \varepsilon_t \nonumber %\\
+    %&= \boldsymbol{\beta} \mathbf{X} + \boldsymbol{\varepsilon}
+\end{align}
+
+where polynomials of $x_t$ up to order $P$ are used.
+This model allows for more complex fitted trends but it remains a fully parametric model and suffers from several problems, especially the behaviour of the fitted polynomial at the start and end of the observed series.
+
+```{r polynomial-example}
+p <- c(1,3,5,10)
+N <- 300
+newd <- with(small, data.frame(Year = seq(min(Year), max(Year), length = N)))
+polyFun <- function(i, data = data) {
+    lm(d15N ~ poly(Year, degree = i), data = data)
+}
+mods <- lapply(p, polyFun, data = small)
+pred <- vapply(mods, predict, numeric(N), newdata = newd)
+colnames(pred) <- p
+newd <- cbind(newd, pred)
+polyDat <- gather(newd, Degree, Fitted, - Year)
+polyDat <- transform(polyDat, Degree = ordered(Degree, levels = p))
+```
+
+Linear models employing a range of polynomials of the covariate Year for the Small Water data set are shown in Figure \ref{fig:polynomial-example-plot}.
+The low-order models ($P \in \{1, 3\}$) result in very poor fit to the data.
+The model with $P = 5$ does a reasonable job of capturing the gross pattern in the data, but it fails to adapt quickly enough to the decrease in δ^15^N that begins ~1940 CE, and the resulting trend is quite biased as a result.
+The $P = 10$th-order polynomial model is well able to capture this period of rapid change, but it does so at the expense of increased complexity in the estimated trend prior to ~1940.
+Additionally, this model ($P = 10$) has undesirable behaviour at the ends of the series, significantly overfitting the data, a commonly observed in polynomial models such as these.
+Finally, the choice of what order of polynomial to fit is an additional choice left for the analyst to specify; choosing the optimal $p$ is not a trivial task when the data are a time series and residual autocorrelation is likely present.
+
+```{r polynomial-example-plot, fig.width = 8, fig.height = 2.5, fig.cap = "Linear models with various orders of polynomial of the covariate Year as covariates.", dependson = c(-1)}
+small_plt + geom_line(data = polyDat, mapping = aes(x = Year, y = Fitted, colour = Degree)) +
+    scale_color_brewer(name = "Degree", palette = "PuOr") +
+    theme(legend.position = "right")
+```
+
+Can we do better than these polynomial fits?
+In the remainder , I hope to demonstrate that the answer to that question is emphatically "yes".
+Below I describe a coherent and consistent approach to modelling palaeoenvironmental time series using generalised additive models (GAMs), which builds upon the linear regression framework.
+
+## Generalised additive models
+
+GAMs extend the generalised linear model (GLM) --- of which the linear model described above is a special case --- where some or all of the parametric terms, the $\beta_p$, are replace by smooth functions $f_j$ of the covariates $x_j$.
+The GAM version of \eqref{eq:linear-model} is
+
+\begin{equation} \label{eq:additive-model}
+y_t = \beta_0 + f(x_t) + \varepsilon_t
+\end{equation}
+
+The immediate advantage of the GAM is that we are no longer restricted to the types of trends that can be fitted via global polynomial functions such as \eqref{eq:polynomial-model}.
+Instead, the form of the fitted trend is estimated from the data itself.
+
+For completeness, where the data are non-Gaussian, such as for where $y_t$ represents a time series of counts of a particular species, we can write \eqref{eq:additive-model} as a GLM/GAM
+
+\begin{align} \label{eq:gam}
+y_t &\sim \text{EF}(\mu_t, \Theta) \\
+g(\mu_t) &= \beta_0 + f(x_t) \\
+\mu_t    &= g^{-1}(\beta_0 + f(x_t)),
+\end{align}
+
+where $\mu_t$ is the expected value of the random variable $Y_t$ ($\mu_t \equiv \mathbb{E}(Y_t)$) of which we have observations $y_t$.
+$g$ is the link function, an invertible, monotonic function, such as the natural logarithm, and $g^{-1}$ its inverse.
+The link function maps values from the linear predictor scale on to the scale of the response.
+For example, count data are strictly non-negative integer values and are commonly modelled as a Poisson GLM/GAM using the natural log link function.
+On the log scale, the response can take any real value between $-\infty$ and $+\infty$, and it is on this scale that model fitting actually occurs.
+However we need to map these unbounded back on to the non-negative response scale the.
+The inverse of the log link function is used, the exponential function, maps values to the interval 0--$\infty$.
+
+In \eqref{eq:gam}, we further assume that the observations are drawn from a member of the exponential family of distributions --- such as the Poisson for count data, the binomial for presence/absence or counts from a total --- with expected value $\mu_t$ and possibly some additional parameters $\Theta$ ($y_t \sim \text{EF}(\mu_t, \Theta)$).
+Additionally, many software implementations of the above models now allow for the fitting of distributions that are not within the exponential family but which can be fitted using an algorithm superficially similar to the one used to fit GAMs to members of the exponential family.
+Common examples of such extended families include the negative binomial distribution (for overdispersed counts) and the beta distribution (for true proportions or other interval-bounded data).
+
+### Basis functions
+
+We want the fitted trends for the Small Water δ^15^N and Braya-sø \uk{} time series to be wiggly functions, whose shape is informed by the data themselves.
+These wiggly functions are smooths, which we denote by $f(x_t)$.
+We need to represent the smooths in a way that (\ref{eq:gam}) is a linear model, which is achieved by using a *basis*.
+A basis is a set of functions that collectively span a space of functions that, we hope, contains the true $f(x_t)$ or a close approximation to it.
+The functions in the basis are known as *basis functions*, and arise from a *basis expansion* of a covariate.
+Writing $b_j(x_t)$ as the $j$th basis function of $x_t$, the smooth $f(x_t)$ can be represented as a weighted sum of basis functions
+
+$$
+f(x_t) = \sum_{j = 1}^{k} b_j(x_t) \beta_j
+$$
+
+where $\beta_j$ is the weight applied to the $j$th basis function. Note that here we concern ourselves only with univariate smooths involving a single covariate, the time variable $x_t$.
+
+The polynomial model we encountered earlier is an example of a statistical model that uses a basis expansion.
+For the cubic polynomial ($P = 3$) fit shown in Figure \ref{fig:polynomial-example-plot} there are in fact 4 basis functions: $b_1(x_t) = x_t^0 = 1$, $b_2(x_t) = x_t$, $b_3(x_t) = x^2_t$, $b_4(x_t) = x_t^3$.
+Note that here $b_1(x_t)$ is constant and is in fact the model intercept, $\beta_0$, in (\ref{eq:polynomial-model}), and the weights are the estimated coefficients in the model, the $\beta_j$.
+
+As we have already seen, polynomial basis expansions do not necessarily lead to well-fitting models unless the true function $f$ is itself a polynomial.
+One of the primary criticisms is that polynomial basis functions are global; the value of $f$ at time point $x_t$ affects the value of $f$ at time point $x_{t+n}$ even if the two time points are at opposite ends of the series.
+There are many other bases we could use; here I discuss one such set of bases, that of splines.
+
+There are a bewildering array of different types of spline.
+In the models discussed below we will largely restrict ourselves to cubic regression splines (CRS) and thin plate regression splines (TPRS).
+In addition, I also discuss two special types of spline basis, an adaptive spline basis and a Gaussian process spline basis.
+I begin with the cubic regression spline basis.
+
+A cubic spline is a smooth curve comprised of sections of cubic polynomials ($P = 3$), where the sections are joined together at some specified locations --- known as *knots* --- in such a way that at the joins, the two sections of cubic polynomial that meet have the same value as well as the same first and second derivative.
+These properties mean that the sections join smoothly and differentiably at the knots.
+
+```{r basis-function-example-plot, fig.width = 8, fig.height = 5, fig.cap = "basis functions plot"}
+## set up
+k <- 7
+df <- with(small, data.frame(Year = seq(min(Year), max(Year), length = 200)))
+knots <- with(small, list(Year = seq(min(Year), max(Year), length = k)))
+sm <- smoothCon(s(Year, k = k, bs = "cr"), data = df, knots = knots)[[1]]$X
+colnames(sm) <- levs <- paste0("F", seq_len(k))
+basis <- gather(cbind(sm, df), Fun, Value, -Year)
+basis <- transform(basis, Fun = factor(Fun, levels = levs))
+
+sm2 <- smoothCon(s(Year, k = k, bs = "cr"), data = small, knots = knots)[[1]]$X
+beta <- coef(lm(d15N ~ sm2 - 1, data = small))
+scbasis <- sweep(sm, 2L, beta, FUN = "*")
+colnames(scbasis) <- levs <- paste0("F", seq_len(k))
+fit <- cbind(df, fitted = rowSums(scbasis))
+scbasis <- gather(cbind(scbasis, df), Fun, Value, -Year)
+scbasis <- transform(scbasis, Fun = factor(Fun, levels = levs))
+
+ylims <- range(basis$Value, scbasis$Value, small$d15N)
+
+p1 <- ggplot(basis, aes(x = Year, y = Value, group = Fun, colour = Fun)) +
+    geom_path() +
+    scale_x_continuous(breaks = knots$Year, labels = NULL, minor_breaks = NULL) +
+    scale_y_continuous(limits = ylims) +
+    scale_colour_discrete(name = "Basis Function") +
+    theme(legend.position = "none") +
+    geom_point(data = small, mapping = aes(x = Year, y = d15N), inherit.aes = FALSE, size = 2, colour = "grey70") +
+    labs(y = d15n_label, x = "Year CE (Knots)")
+
+p2 <- ggplot(scbasis, aes(x = Year, y = Value, group = Fun, colour = Fun)) +
+    geom_path() +
+    scale_x_continuous(breaks = knots$Year, labels = NULL, minor_breaks = NULL) +
+    scale_y_continuous(limits = ylims) +
+    scale_colour_discrete(name = "Basis Function") +
+    theme(legend.position = "none") +
+    geom_point(data = small, mapping = aes(x = Year, y = d15N), inherit.aes = FALSE, size = 2, colour = "grey70") +
+    geom_line(data = fit, mapping = aes(x = Year, y = fitted), inherit.aes = FALSE,
+              size = 0.75, colour = "black") +
+    labs(y = d15n_label, x = "Year CE (Knots)")
+
+tp <- smoothCon(s(Year, k = k, bs = "tp"), data = df)[[1]]$X
+colnames(tp) <- levs <- paste0("F", seq_len(k))
+tpbasis <- gather(cbind(tp, df), Fun, Value, -Year)
+tpbasis <- transform(tpbasis, Fun = factor(Fun, levels = levs))
+
+p3 <- ggplot(tpbasis, aes(x = Year, y = Value, group = Fun, colour = Fun)) +
+    geom_path() +
+    scale_colour_discrete(name = "Basis Function") +
+    theme(legend.position = "none") +
+    labs(y = d15n_label, x = "Year CE")
+
+pbasis <- plot_grid(p1, p2, p3, ncol = 2, align = "hv", labels = "auto")
+pbasis
+```
+
+A CRS can be parameterized in a number of different ways.
+One requires a knot at each unique data value in $x_t$, which is computationally inefficient.
+One approach to specifying a CRS basis is to parameterize in terms of the value of the spline at the knots.
+Typically in this parametrization there are many fewer knots than in alternative parameterizations, and these knots are either distributed evenly over the range of $x_t$ or at the quantiles of $x_t$.
+Placing knots at the quantiles of $x_t$ has the effect of placing a greater number of knots where the data is densest.
+
+A CRS basis expansion comprised of `r k` basis functions for the time covariate in the Small Water series, is shown in Figure \ref{fig:basis-function-example-plot}a.
+The tick marks on the x-axis show the locations of the knots, which are located at the ends of the series and evenly in between.
+Notice that in this particular parametrization, the $j$th basis function takes a value of 1 at the $j$th knot and at all other knots a value of 0.
+
+To estimate a model using this basis expansion each basis function forms a column in the model matrix $\boldsymbol{X}$ and the weights $\beta_j$ can be found using least squares regression (assuming a Gaussian response).
+Note that in order to estimate a coefficient for each basis function the model has to be fitted without an intercept term.
+In practice we would include an intercept term in the model and therefore the basis functions are modified via an identifiability constraint.
+This has the effect of making the basis orthogonal to the intercept but results in more complicated basis functions than those shown in in Figure \ref{fig:basis-function-example-plot}a.
+
+Having estimated the weight for each basis function, the $j$th basis function $b_j$ is scaled (weighted) by its coefficient $\beta_j$.
+The scaled CRS basis functions for the Small Water time series are shown in Figure \ref{fig:basis-function-example-plot}b.
+The solid line passing through the data points is formed by summing up the values of the scaled basis functions ($b_j(x_t) \beta_j$) at any value of $x_t$ (time).
+
+Cubic regression splines, as well as many other types of spline, require the analyst to choose the number and location of the knots that parametrise the basis.
+Thin plate regression splines (TPRS) remove this element of subjectivity when fitting GAMs.
+Thin plate splines were introduced by @Duchon1977-jr and, as well as solving the knot selection problem, have several additional attractive properties in terms of optimality and their ability to estimate a smooth function of two or more variables, leading to the inclusion of smooth interactions between covariates.
+However, thin plate splines have one key disadvantage over CRS; thin plate splines have as many unknown parameters as there are unique combinations of covariate values in a data set.
+It is unlikely in the extreme that any real data problem would require functions of such complexity that they require so many basis functions.
+It is much more likely that the true functions that we attempt to estimate are far simpler than the set of functions able to be represented by 1 basis function per unique data value.
+From a practical point of view, it is highly inefficient to carry around all these basis functions whilst model fitting, and the available computational resources would become quickly exhausted for large time series with many observations.
+
+To address this issue, thin plate regression splines (TPRS) have been suggested which truncate the space of the thin plate spline basis to some lower number of basis functions whilst preserving much of the advantage of the original basis as an optimally-fitting spline [@Wood2003-qy].
+A rank `r k` TPRS basis is shown in Figure \ref{fig:basis-function-example-plot}c for the Small Water time series.
+The truncation is achieved by performing an eigen-decomposition of the basis functions and retaining the eigenvectors associated with the $k$ largest eigenvalues.
+This is similar to the way principal components analysis decomposes a data set into axes of variation (eigenvectors) in decreasing order of variance explained.
+By concentrating the "wiggliness" information in those eigenvectors with large eigenvalues, the truncated basis can preserve much of the space of functions spanned by the original basis but at the cost of using far fewer basis functions.
+
+The truncation suggested by @Wood2003-qy is not without cost; the eigen-decomposition and related steps can be relatively costly to set up for large data sets.
+For data sets of similar size to the two examples used here, the additional computational effort required to set up the thin plate regression spline basis over say the cubic regression spline basis will not be noticeable.
+For highly resolved data sets containing more than 500--1000 observations the truncation may be too costly computationally.
+In such instances, little is lost by moving to the cubic regression spline basis, with the same number of knots as the rank of the desired thin plate spline, either distributed evenly throughout the data or at the quantiles of the data, with the benefit of considerably reduced set up time for the basis.
+
+With the two regression spline bases described above, to fit a GAM the analyst is generally only required to the specify the size (rank, number of knots) of the basis expansion required to represent or closely approximate the true function $f$.
+With practice and some knowledge of the system from which the observations arise, it can be relatively easy to put an upper limit on the expected complexity of the true trend in the data.
+Additionally, the number available data points places a constraint on the upper limit of the size of basis expansion that can be used.
+
+In practice, the size of the basis is an upper limit on the expected complexity of the trend, and a simple test is available to check if the basis used was sufficiently large [@Pya2016-rk].
+This test is available via the `gam.check()` function in **mgcv** for example, and essentially looks at whether there is any additional nonlinearity or structure in the residuals that can be explained by $x_t$.
+Should a smooth term in the fitted model fail this test the model can be refitted using a larger basis expansion, say by doubling the value of `k` used to fit the original.
+Note also that a smooth might fail this test whilst using fewer effective degrees of freedom than the maximum possible for the dimension of basis used.
+This may happen when the true function lies at the upper limit of the set of functions encompassed by the size of basis used.
+Additionally, a basis of size $2k$ encompasses a larger space of functions of a given complexity than a basis of size $a$ [@Wood2017-qi]; increasing the basis dimension used to fit the model may unlock this additional function space resulting in a better fitting model whilst using a similar number of effective degrees of freedom.
+
+### Smoothness selection
+
+Having identified regression splines as a useful way to represent $f$, we next need a way to decide how wiggly the fitted trend should be.
+A backwards elimination approach to sequentially remove knots or basis functions might seem appropriate, however such an approach would likely fail as the resulting sequence of models would not strictly nested, precluding many forms of statistical comparison.
+Alternatively, we could keep the basis dimension at a fixed size but gaurd against fitting very complex models through the use of a wiggliness penalty.
+
+The default wiggliness penalty used in GAMs is on the second derivative of the spline, which measures the rate of change of the slope, the curvature, of the spline at any infinitesimal point in the interval spanned by $x_t$.
+The actual penalty used is the integrated squared second derivative of the spline
+
+$$\int_{\mathbb{R}} [f^{\prime\prime}]^2 dx$$
+
+which can be written in quadratic form as a function of the estimated coefficients of $f(x_t)$
+
+$$\int_{\mathbb{R}} [f^{\prime\prime}]^2 dx = \boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}$$
+
+where $\mathbf{S}$ is known as the penalty matrix.
+
+Now that we have a convenient way to measure wiggliness, it needs to be incorporated into the objective function that will be minimised to fit the GAM.
+The likelihood of the model given the parameter estimates $\mathcal{L}(\boldsymbol{\beta})$ is combined with the penalty to create the penalized likelihood $\mathcal{L}_p(\boldsymbol{\beta})$
+
+$$\mathcal{L}_p(\boldsymbol{\beta}) = \mathcal{L}(\boldsymbol{\beta}) - \frac{1}{2} \lambda\boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}$$
+
+The fraction of a half is there simply to make the penalised likelihood equal the penalised sum of squares in the case of a Gaussian model.
+$\lambda$, the smoothess parameter, controls the extent to which the penalty contributes to the likelihood of the model.
+In the extreme case, if $\lambda = 0$ then the penalty has no effect and the penalized likelihood equals the likelihood of the model given the parameters.
+As $\lambda \rightarrow \infty$ the penalty comes to dominate $\mathcal{L}_p(\boldsymbol{\beta})$ and the wiggliness of $f(x_t)$ tends to $0$ resulting in an infinitely smooth function, which, in the case of a second derivative penalty, is a straight line.
+If a penalty was on the integrated squared *first* derivative of $f(x_t)$, the fully penalised spline would be a flat, horizontal line.
+
+```{r penalty-example}
+K <- 40
+lambda <- c(10000, 1, 0.01, 0.00001)
+N <- 300
+newd <- with(small, data.frame(Year = seq(min(Year), max(Year), length = N)))
+fits <- lapply(lambda, function(lambda) gam(d15N ~ s(Year, k = K, sp = lambda), data = small))
+pred <- vapply(fits, predict, numeric(N), newdata = newd)
+op <- options(scipen = 100)
+colnames(pred) <- lambda
+newd <- cbind(newd, pred)
+lambdaDat <- gather(newd, Lambda, Fitted, - Year)
+lambdaDat <- transform(lambdaDat, Lambda = factor(paste("lambda ==", as.character(Lambda)), levels = paste("lambda ==", as.character(lambda))))
+options(op)
+```
+
+```{r penalty-example-plot, fig.width = 8, fig.height = 5, fig.cap = "the effect of different smoothness", dependson = -1}
+op <- options(scipen = 100)
+small_plt + geom_line(data = lambdaDat, mapping = aes(x = Year, y = Fitted, group = Lambda),
+                      size = .75, colour = "#e66101") +
+    facet_wrap( ~ Lambda, ncol = 2, labeller = label_parsed)
+options(op)
+```
+
+Figure \ref{fig:penalty-example-plot} illustrates how the smoothness parameter $\lambda$ controls the degree of wiggliness in the fitted spline.
+Four models are shown each fitted with a fixed value of $\lambda$; 10000, 1, 0.01, and 0.00001.
+At $\lambda = 10000$ the model effectively fits a linear model through the data.
+As the value of $\lambda$ decreases, the fitted spline becomes increasingly wiggly.
+As $\lambda$ becomes very small, the resulting spline passes through most of the δ^15^N observations resulting in a model that is clearly over fitted to the data.
+
+To fully automate smoothness selection for $f(x_t)$ we need to estimate $\lambda$.
+There are two main ways that $\lambda$ can be automatically chosen during model fitting.
+The first way is to choose $\lambda$ such that it minimises the prediction error of the model.
+This can be achieved by choosing $\lambda$ to minimise Akaike's information criterion (AIC) or via cross-validation (CV) or generalized cross-validation (GCV).
+GCV avoids the computational overhead inherent to CV of having to repeatedly refit the model with one or more observations left out as a test set.
+Minimising the GCV score will, with a sufficiently large data set, find a model with the minimal prediction error.
+The second approach is to treat the smooth as a random effect, in which $\lambda$ is now a variance parameter to be estimated using maximum likelihood (ML) or restricted maximum likelihood (REML).
+
+Several recent results have shown that GCV, under certain circumstances, has a tendency to under smooth, resulting in fitted splines that are overly wiggly [@Reiss2009-fk].
+Much better behaviour has been observed for REML and ML smoothness selection, in that order [@Wood2011-kn].
+REML is therefore the recommended means of fitting GAMs, though, where models have different fixed effects (covariates) they cannot be compared using REML, and ML selection should be used instead.
+In the sorts of data examples considered here there is only a single covariate $x_t$ as our models contain a single estimated trend so REML smoothness selection is used throughout unless otherwise stated.
+
+### Smoothing time series data
+
+**Do I need this as a separate section?** Or just cover the points as they occur?
+
+## Fitting GAMs
+
+### Small Water
+
+The trend in δ^15^N values is clearly non-linear and it would be difficult to suggest a suitable polynomial model that would allow for periods of relatively no change in δ^15^N as well as rapid change.
+Instead, a GAM is ideally suited to modelling such trends; the data suggest a smoothly varying change in δ^15^N between 1925 and 1975.
+It is reasonable to expect some autocorrelation in the model errors about the fitted trend.
+Therefore I fitted the following GAM to the δ^15^N time series.
+
+\begin{equation} \label{eq:small-gam}
+y_t = \beta_0 + f(x_t) + \varepsilon, \quad \varepsilon_t \sim \mathcal(0, \boldsymbol{\Lambda}\sigma^2)
+\end{equation}
+
+Note that now I have relaxed the i.i.d.
+assumption and have introduced $\boldsymbol{\Lambda}$, a correlation matrix that is used to model autocorrelation in the residuals.
+The δ^15^N values are irregularly spaced in time and as a result a correlation structure that can handle the uneven spacing of the samples is needed.
+A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the first-order autoregressive process (AR(1)) and, simply stated, models the correlation between any two residuals as $\phi^h$, where $h$ is the amount of separation in time between the residuals.
+$h$ may be a real valued number in the CAR(1), which accommodates the irregular separation of samples.
+$\phi$ controls how quickly the correlation between any two residuals declines as a function of their separation in time.
+$\phi$ is an additional parameter that will be estimated during model fitting.
+The model in \ref{eq:small-gam} was fitted using the `gamm()` function [@Wood2004-zv] in the **mgcv** package [@Wood2017-qi] for R [@R-core-team].
+
+```{r fit-small-water-gamm}
+mod <- gamm(d15N ~ s(Year, k = 15), data = small,
+            correlation = corCAR1(form = ~ Year), method = "REML")
+
+## estimate of phi and confidence interval
+smallPhi <- intervals(mod$lme, which = "var-cov")$corStruct
+
+## summary object for use in document
+smallSumm <- summary(mod$gam)
+```
+
+```{r fit-braya-so-car1-and-gcv-models}
+## fit the car(1) model --- needs optim as this is not a stable fit!
+braya.car1 <- gamm(UK37 ~ s(Year, k = 10), data = braya, correlation = corCAR1(form = ~ Year),
+                   method = "REML",
+		   control = list(niterEM = 0, optimMethod = "BFGS", opt = "optim"))
+braya.gcv <- gam(UK37 ~ s(Year, k = 30), data = braya)
+
+## car(1) parameter
+## optima-estimated version of the model gives phi = 0.200016
+##           lower      est. upper
+## Phi 2.58185e-16 0.2000162     1
+brayaPhi <- intervals(braya.car1$lme)$corStruct
+```
+
+```{r plot-fitted-models, fig.width = 8, fig.height = 5, fig.cap = "Fitted GAMs"}
+N <- 300                                # number of points at which to evaluate the splines
+## Predict from the fitted model
+newYear <- with(small, data.frame(Year = seq(min(Year), max(Year), length.out = 200)))
+newYear <- cbind(newYear, data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
+newYear <- transform(newYear, upper = fit + (2 * se.fit), lower = fit - (2 * se.fit))
+
+## Plot simulated trends
+small_fitted <- ggplot(newYear, aes(x = Year, y = fit)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper, x = Year), alpha = 0.2, inherit.aes = FALSE, fill = "black") +
+    geom_point(data = small, mapping = aes(x = Year, y = d15N), inherit.aes = FALSE) +
+    geom_line() +
+    labs(y = d15n_label, x = "Year CE")
+
+## ggplot with data and fitted spline, then resids vs time in second panel
+newBraya <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = N)))
+newBraya <- cbind(newBraya, data.frame(predict(braya.car1$gam, newBraya, se.fit = TRUE)))
+crit.t <- qt(0.975, df = df.residual(braya.car1$gam))
+newBraya <- transform(newBraya,
+                      upper = fit + (crit.t * se.fit),
+                      lower = fit - (crit.t * se.fit))
+## add GAM GCV results
+fit.gcv <- predict(braya.gcv, newdata = newBraya, se.fit = TRUE)
+newBraya <- rbind(newBraya, newBraya)      # extend newBraya to take GCV results
+newBraya[seq(N+1, length.out = N, by = 1), ]$fit <- fit.gcv$fit
+newBraya[seq(N+1, length.out = N, by = 1), ]$upper <-
+    fit.gcv$fit + (qt(0.975, df.residual(braya.gcv)) * fit.gcv$se.fit)
+newBraya[seq(N+1, length.out = N, by = 1), ]$lower <-
+    fit.gcv$fit - (qt(0.975, df.residual(braya.gcv)) * fit.gcv$se.fit)
+newBraya <- transform(newBraya, Method = rep(c("GAMM (CAR(1))", "GAM (GCV)"), each = N))
+
+## plot CAR(1) and GCV fits
+braya_fitted <- ggplot(braya, aes(y = UK37, x = Year)) +
+    geom_point() +
+    geom_ribbon(data = newBraya,
+                mapping = aes(x = Year, ymax = upper, ymin = lower, fill = Method),
+                alpha = 0.3, inherit.aes = FALSE) +
+    geom_line(data = newBraya, mapping = aes(y = fit, x = Year, colour = Method)) +
+    labs(y = braya_ylabel, x = "Year CE") +
+    scale_color_manual(values = c("#5e3c99", "#e66101")) + ## values = viridis(11)[c(2, 6)]) +
+    scale_fill_manual(values = c("#5e3c99", "#e66101")) + ## values = viridis(11)[c(2, 6)]) +
+    theme(legend.position = "right")
+
+## convert to gtabl to align heights
+g_small_fitted <- ggplotGrob(small_fitted) # convert to gtable
+g_braya_fitted <- ggplotGrob(braya_fitted) # convert to gtable
+
+height_take <- c(4,5)
+small_heights <- g_small_fitted$heights[height_take]
+braya_heights <- g_braya_fitted$heights[height_take]
+max_heights <- unit.pmax(small_heights, braya_heights) # calculate maximum heights
+g_small_fitted$heights[height_take] <- max_heights # assign max heightss to small gtable
+g_braya_fitted$heights[height_take] <- max_heights
+
+plot_grid(g_small_fitted, g_braya_fitted, ncol = 1, labels = "auto", align = "hv", axis = "lr")
+```
+
+The fitted trend is shown in Figure \ref{fig:plot-fitted-models}a, and well-captures the strong pattern in the data.
+The trend is statistically significant (estimated degrees of freedom = `r round(smallSumm[["s.table"]][1,2], 3)`; $F$ = `r round(smallSumm[["s.table"]][1,3], 3)`, approximate $p$ value = $\ll$ 0.0001).
+However further analysis of the fitted model is required to answer the other questions posed earlier about the timing of change and whether features in the trend can be distinguished from random noise.
+
+### Braya-sø
+
+The \uk{} data present a more difficult data analysis challenge than the δ^15^N time series because of the much more complex variation present.
+Fitting the same model as the Small Water example, \ref{eq:small-gam}, to the \uk{} data resulted in the unsatisfactory fit shown as the very smooth line in Figure \ref{fig:plot-fitted-models}b (labelled GAMM (CAR(1))).
+Further problems were evident with this model fit --- the covariance matrix of the model was non-positive definite, a sure sign of problems with the fitted model.
+Refitting with a smaller basis dimension (*k* = 20) for the trend term resulted in a model with a positive-definite covariance matrix for the model variance-covariance terms, but the estimated value of of the CAR(1) parameter $\phi$ = `r round(brayaPhi[2], 2)` was exceedingly uncertain (95% confidence interval `r round(brayaPhi[1], 2)` -- `r round(brayaPhi[3], 2)`!)
+
+Fitting this model as a standard GAM with REML smoothness selection resulted in the same fitted trend as the GAM with CAR(1) errors (not shown), whilst using GCV smoothness selection resulted in a much more satisfactory fitted trend.
+There are two potential problems with the GCV-selected trend: i) GCV is sensitive to the profile of the GCV score and has been shown to badly under smooth data in situations where the profile is flat around the minimum GCV score, and ii) the model fitted assumes that the observations are independent, an assumption that is certainly violated in the \uk{} time series.
+
+To investigate the first issue the GCV and REML scores for an increasing sequence of values of the smoothness penalty $\lambda$ were evaluated for the standard GAM (equation \ref{eq:gam}) fit to the \uk{} time series.
+The resulting profiles are shown in Figure \ref{fig:trace-smoothness-parameters}, with the selected value of the penalty shown by the vertical line.
+The GCV score profile suggests that the potential for under smoothing identified by @Reiss2009-fk is unlikely to apply here.
+Addressing the violation of the independence assumption is more difficult; the standard errors of the model coefficients are most likely anti-conservative, which presents a significant problem if we wish to determine which of the wiggles in the fitted \uk{} trend can be identified as real.
+
+To understand the reason why the GAM plus CAR(1) and the GAM with REML smoothness selection performed poorly with the \uk{} time series we need to delve a little deeper into what is happening when we are fitting these two models.
+
+```{r trace-smoothness-parameters, fig.width = 8, fig.height = 2.5, fig.cap = "GCV and REML scores as a function of the smoothness parameter λ."}
+## Generate GCV and REML traces
+lambda <- 1e-8
+gcv <- reml <- reml2 <- numeric(length = 100)
+
+for (i in seq_along(gcv)) {
+    mGCV <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "GCV.Cp", sp = lambda)
+    mREML <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "REML", sp = lambda)
+    mREML2 <- gam(UK37 ~ s(Year, k = 40), data = braya, method = "REML", sp = lambda,
+                  weights = sampleInterval / mean(sampleInterval))
+    gcv[i] <- mGCV$gcv.ubre
+    reml[i] <- mREML$gcv.ubre
+    reml2[i] <- mREML2$gcv.ubre
+    lambda <- lambda * 1.5
+}
+
+## GCV and REML score traces
+result <- data.frame(lambda = 1e-8 * 1.5^{0:99}, score = c(gcv, reml, reml2),
+                     criterion = rep(c("GCV", "REML", "REML (k=40, weights)"), each = 100))
+
+## Full fits to get estimated models
+fullGCV  <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "GCV.Cp")
+fullREML <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "REML")
+fullREML2 <- gam(UK37 ~ s(Year, k = 40), data = braya, method = "REML",
+                 weights = sampleInterval / mean(sampleInterval))
+obsSP <- data.frame(lambda = c(fullGCV$sp, fullREML$sp, fullREML2$sp), criterion = c("GCV", "REML", "REML (k=40, weights)"))
+
+## summary object for use in document
+brayaSumm <- summary(fullREML2)
+
+## Plot
+trace.plt <- ggplot(result, aes(x = lambda, y = score)) +
+    geom_line() +
+    facet_wrap(~ criterion, scales = "free_y") + scale_x_log10() +
+    xlab(expression(lambda)) +
+    ylab("Score") +
+    geom_vline(aes(xintercept = lambda), obsSP, colour = "grey")
+trace.plt
+```
+
+The primary issue leading to poor fit in both the GAM plus CAR(1) and the GAM with REML smoothness selection is that neither model accounts for the different variance of each observation in the \uk{} record.
+The sediments in Braya-sø are not annually laminated and therefore the core was sliced at regular depth intervals.
+Owing to compaction of older sediments and variation in accumulation rates over time, each sediment slice represents a different number of "lake years".
+We can think of older samples as representing some average of many years of lake sediment, whilst younger samples are representative of fewer lake years.
+The average of a larger set of numbers is estimated more precisely than the average of a smaller set, all things equal.
+A direct result of this variable averaging of lake years it that some samples are more precise and therefore have lower variance than others.
+
+The models fitted thus far fail to take account of this feature of the data, but including this information in the model is relatively simple via the use of observational weights.
+The number of lake years represented by each slice is estimated by assigning a date to the top and bottom of each sediment slice.
+The variance of each observation should be proportional to the inverse of the number of lake years each sample represents.
+In the `gam()` function used here, weights should be specified as the number of lake years each sample represents.
+Other software may require the weights to be specified in a different way.
+
+A secondary problem is the size of the basis dimension used for the time variable.
+The main user selectable option when fitting a GAM in the penalised likelihood framework of @Wood2004-zv is how large a basis to.
+As described above, the basis should be large enough to contain the true, but unknown, function, or a close approximation to it.
+For GCV selection the basis contained 29 basis functions, whilst the CAR(1) model with REML smoothness selection would only converge with a basis containing 20 functions.
+The size of the basis appears to be sufficient for GCV smoothness selection, but following @Wood2011-kn REML smoothness selection is preferred.
+Using the test of @Pya2016-rk, the basis dimension for the models with REML smoothness selection was too small.
+To proceed therefore, we must drop the CAR(1) term and increase the basis dimension to 39 functions (by setting `k = 40`).
+
+With the larger basis dimension and accounting for the non-constant variance of the observations via weights, the model fitted using REML is indistinguishable from that obtained using GCV (Figure \ref{fig:plot-fitted-models}b).
+The trace of the REML score for this model (Figure \ref{fig:trace-smoothness-parameters}c) shows a pronounced minimum at a much smaller value of $\lambda$ than the original REML fit (Figure \ref{fig:trace-smoothness-parameters}b), indicating that a more wiggly trend provides a better fit to the Braya-sø time series.
+This example illustrates that some care and understanding of the underlying principles of GAMs is required to diagnose potential issues with the estimated model.
+After standard modelling choices (size of basis to use, correct selection of response distribution and link function, etc.) are checked, it often pays to think carefully about the properties of the data and ensure that the assumptions of the model are met.
+Here, despite increasing the basis size, it was the failure to appreciate the magnitude of the effect on fitting of the non-constant variance that lead to the initially poor fit and the problems associated with the estimation of the CAR(1) process.
+
+### Confidence intervals
+
+If we want to ask whether either of the estimated trends is statistically interesting or proceed to identifying periods of significant change, we must address the issue of uncertainty in the estimated model.
+What uncertainty is associated with the trend estimates?
+One way to visualise this is through a 1 - $\alpha$ confidence interval around the fitted trend, where $\alpha$ is say 0.05 for a 95% interval.
+A 95% interval would be drawn at $\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))$, where $m_{1-\alpha} = 1.96$, the 0.95 probability quantile of a standard normal distribution, and $\text{SE}(\hat{y}_t)$ is the standard error of the estimated value at time $x_t$.
+This type of confidence interval would normally be described as *pointwise*; the coverage properties of the interval would be correct at a single point on the fitted trend, but, if we were to consider additional points on the trend, the coverage would logically be lower than 1 - $\alpha$.
+This would be the traditional frequentist interpretation of a confidence interval.
+However, the approach to fitting GAMs described here is an empirical Bayesian one and therefore the typical frequentist interpretation is not applicable.
+@Nychka1988-rz investigated the properties of a confidence interval created as described above using standard errors derived from the Bayesian posterior covariance matrix for the estimated mode parameters.
+Such intervals have the interesting property that they have good *across-the-function* coverage when considered from a frequentist perspective.
+This means that, when averaged over the range of the function, the Bayesian credible intervals shown in Figure \ref{fig:plot-fitted-models} have close to the expected 95% coverage.
+However, to achieve this some parts of the function may have more or less than 95%-coverage.
+@Marra2012-bq recently explained Nychka's [-@Nychka1988-rz] suprising results and extended them to the case of generalised models (non-Gaussian responses).
+
+```{r posterior-simulation, fig.height = 5, fig.width = 8, fig.cap = "posterior simulation"}
+set.seed(1)
+nsim <- 20
+take <- 20
+sims <- simulate(mod, nsim = nsim, newdata = newYear, unconditional = TRUE)
+randSims <- sims[, sample(nsim, take)]
+colnames(randSims) <- paste0("sim", seq_len(take))
+randSims <- setNames(stack(as.data.frame(randSims)), c("simulated", "run"))
+randSims <- transform(randSims, Year = rep(newYear$Year, take),
+                      simulated = simulated)
+
+## Plot simulated trends
+smallSim.plt <- ggplot(newYear, aes(x = Year, y = fit)) +
+    geom_line(data = randSims, mapping = aes(y = simulated, x = Year, group = run),
+              colour = "#fdb863", alpha = 0.5) +
+    geom_line() +
+    labs(y = d15n_label, x = "Year CE")
+
+## posterior simulation
+## need to reset-up newBraya
+newBraya <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = N)))
+brayaREML2 <- cbind(newBraya, data.frame(predict(fullREML2, newBraya, se.fit = TRUE)))
+## simulate
+set.seed(1)
+sims2 <- simulate(fullREML2, nsim = nsim, newdata = newBraya, unconditional = TRUE)
+randSims2 <- sims2[, sample(nsim, take)]
+colnames(randSims2) <- paste0("sim", seq_len(take))
+randSims2 <- setNames(stack(as.data.frame(randSims2)), c("simulated", "run"))
+randSims2 <- transform(randSims2, Year = rep(newBraya$Year, take),
+                       simulated = simulated)
+
+brayaSim.plt <- ggplot(brayaREML2, aes(x = Year, y = fit)) +
+    geom_line(data = randSims2, mapping = aes(y = simulated, x = Year, group = run),
+              colour = "#fdb863", alpha = 0.5) +
+    geom_line() +
+    labs(y = braya_ylabel, x = "Year")
+
+plot_grid(smallSim.plt, brayaSim.plt, ncol = 1, labels = "auto", align = "hv", axis = "lr")
+```
+
+Whilst the *across-the-function* or average coverage frequentist interpretation of the Bayesian credible intervals is useful, if may be important to have an interval that contains the entirety of the true function with state probability (1 - $\alpha$).
+Such an interval is known as a *simultaneous* interval.
+A 1 - $\alpha$ *simultaneous* confidence interval contains *in their entirety* 1 - $\alpha$ of all random draws from the posterior distribution of the fitted model.
+
+Fitting a GAM involves finding estimates for coefficients of the basis functions.
+Together, these coefficients are distributed multivariate normal with mean vector and covariance matrix specified by the model estimates of the coefficients and their covariances.
+Random draws from this distribution can be taken, where each random draw represents a new trend that is consistent with the fitted trend but it also includes the uncertainty in the estimated trend.
+This process is known as *posterior simulation*.
+
+Figure \ref{fig:posterior-simulation} shows `r take` such random draws from the posterior distribution of the GAMs fitted to the Small Water and Braya-sø data sets.
+For the random draws illustrated in Figure \ref{fig:posterior-simulation}, a simultaneous interval should contain the entire function for on average `r round(take * 0.95, 0)` of the `r take` draws.
+In the early period of the δ^15^N time series many of the posterior simulations exhibit short periods of increasing and decreasing trend, balancing out to the relatively flat trend estimated by the GAM (Fig.\ \ref{fig:posterior-simulation}a).
+Reflecting this uncertainty, we might expect relatively wide simultaneous intervals during this period in order to contain the vast majority of the simulated trends.
+Conversely, the decreasing δ^15^N trend starting at ~1945 is consistently reproduced in the posterior simulations, suggesting both that this feature of the time series is real and statistically significant, and that the rate of change in δ^15^N is relatively precisely estimated.
+We see a similar pattern in Figure\_\ref{fig:posterior-simulation}b for the Braya-sø record; the large peak in \uk{} at ~250CE and the strong decline at ~1200CE are well defined in the posterior simulations, whereas most of the localised trends that are smaller magnitude changes in $y_t$ are associated with posterior simulations that are less well constrained with the ends of the record in particular showing considerable variation in the strength, timing, and even sign of simulated trends, reflecting the greater uncertainty in estimated trend during these periods.
+
+There are a number of ways in which a simultaneous interval can be computed.
+Here I follow the simulation approach described by @Ruppert2009-ig and present only the basic detail; a fuller description is contained in Appendix **X**.
+The general idea is that if we want to create an interval that contains the whole of the true function with 1 - $\alpha$ probability, we need to increase the standard Bayesian credible interval by some amount.
+We could simulate a large number of functions from the posterior distribution of the model and then search for the value of $m_{1-\alpha}$ that when multiplied by $\text{SE}(\hat{f}(x_t))$ yielded an interval that contained the whole function for $(1-\alpha)$ 100% of the functions simulated.
+In practice, the simulation method of @Ruppert2009-ig does not involve a direct search, but in a practical sense results in the critical value $m_{1-\alpha}$ required.
+
+```{r compare-intervals, fig.height = 5, fig.width = 8, fig.cap = "simultaneous intervals illustration"}
+## small water
+sw.cint <- confint(mod, parm = "Year", newdata = newYear, type = "confidence")
+sw.sint <- confint(mod, parm = "Year", newdata = newYear, type = "simultaneous")
+
+## braya so
+bs.cint <- confint(fullREML2, parm = "Year", newdata = newBraya, type = "confidence")
+bs.sint <- confint(fullREML2, parm = "Year", newdata = newBraya, type = "simultaneous")
+
+smallInt.plt <- ggplot(sw.cint, aes(x = Year, y = est)) +
+    geom_ribbon(data = sw.sint, mapping = aes(ymin = lower, ymax = upper, x = Year),
+                fill = "#fdb863", alpha = 0.5, inherit.aes = FALSE) +
+    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
+                fill = "#e66101", alpha = 0.5, inherit.aes = FALSE) +
+    geom_line() +
+    labs(y = d15n_label, x = "Year CE")
+
+brayaInt.plt <- ggplot(bs.cint, aes(x = Year, y = est)) +
+    geom_ribbon(data = bs.sint, mapping = aes(ymin = lower, ymax = upper, x = Year),
+                fill = "#fdb863", alpha = 0.5, inherit.aes = FALSE) +
+    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
+                fill = "#e66101", alpha = 0.5, inherit.aes = FALSE) +
+    geom_line() +
+    labs(y = braya_ylabel, x = "Year")
+
+plot_grid(smallInt.plt, brayaInt.plt, ncol = 1, labels = "auto", align = "hv", axis = "lr")
+```
+
+Simultaneous intervals computed using the method described are show in Figure \ref{fig:compare-intervals} alongside the *across-the-function* confidence intervals for the trends fitted to both example data sets.
+As expected, the simultaneous interval in somewhat wider than the *across-the-function* interval.
+The critical value $m_{1-\alpha}$ for the simultaneous interval of the estimated trend in δ^15^N is `r with(sw.sint, round(unique(crit), 2))`, whilst the same value for the \uk{} series is `r with(bs.sint, round(unique(crit), 2))`, leading to intervals that are approximately ±50% and ±75% wider than the equivalent across-the-function intervals.
+
+### Identifying periods change
+
+In the simple linear trend model \eqref{eq:linear-model} whether the estimated trend constitutes evidence for or against a null hypothesis of no change rests on how large the estimated rate of change in $y_t$ is ($\hat{\beta}_1$) relative to its uncertainty.
+This is summarised in the $t$ statistic.
+As the rate of change in $y_t$ is constant over the fitted trend --- there is only a singe slope for the fitted trend $\hat{\beta}_1$ --- if the $t$ statistic of the test that $\hat{\beta}_1 = 0$ is unusually extreme this would be evidence against the null hypothesis of no change.
+Importantly, this applies to the whole time series as the linear model implies a constant rate of change throughout.
+The estimate $\hat{\beta}_1$ is the first derivative of the fitted trend.
+
+In the GAM, the fitted trend need not be linear; the slope of the trend is potentially different at every point in the time series.
+As such we might reasonably ask *where* in the series the response $y_t$ is changing?
+Mirroring the linear model we can answer this question by determining whether or not the first derivative --- the slope --- of the fitted trend at any time point is consistent with a null hypothesis of no change.
+We want to know whether or not the first derivative is indistinguishable from a value of $0$ --- no trend --- given the uncertainty in the estimate of the derivative.
+
+Derivatives of the fitted spline are not easily available analytically, but they can be estimated using the method of finite differences.
+Two values of the fitted trend, separated by a very small time-shift ($\Delta_t$), are predicted from the model; the difference between the fitted values for the two time points is an approximation of the true first derivative of the trend.
+As $\Delta_t \rightarrow 0$ the approximation becomes increasingly accurate.
+In practice, the first derivative of the fitted trend is evaluated via finite differences at a large number of points in the time series.
+An approximate $(1 - \alpha)$ 100% pointwise confidence interval can be calculated for the derivative estimates using standard theory (i.e.\ $\pm 1.96 \times \text{SE}(y_t)$) and the covariance matrix of the spline coefficients.
+A $(1 - \alpha)$ 100% simultaneous interval can be computed using the method described above.
+
+Figure \ref{fig:derivatives} shows the estimated first derivative of the fitted trend in the Small Water (\ref{fig:derivatives}a) and Braya-sø (\ref{fig:derivatives}b) time series.
+Although the estimated trend suggests a slight increase in δ^15^N from the start of the record to ~1940, the estimated trend is sufficiently uncertain that the simultaneous interval on the first derivative includes 0 throughout this period.
+We can understand why this is so by looking at the posterior simulations in Figure\ \ref{fig:posterior-simulation}a; there is considerable variation in the shape of the simulated trends throughout this period.
+From ~1925 the derivative of the trend becomes negative, however it is not until ~1940 that the simultaneous interval doesn't include $0$.
+At this point we have evidence to reject the null hypothesis of no change.
+This time point may be taken as the first evidence for change in δ^15^N in the Small Water core.
+The simultaneous interval on the first derivative of the trend in δ^15^N is bounded away from $0$ between ~1940 and ~1975, covering the major decline in values clearly evident in the observations.
+The simultaneous interval includes $0$ from ~1975 onward, suggesting that, whilst quite pronounced, the recent increase in δ^15^N is not statistically significant.
+To determine whether the recent increase is real or not, we would require considerably more samples with which to (hopefully) more-precisely estimate the trend during this period.
+Alternatively, we might just have to wait until sufficient additional sedimentation has occurred to warrant recoring Small Water and reestimating the trend in δ^15^N.
+
+The estimated trend at Braya-sø exhibited a number of oscillations in in \uk{} .
+As we saw previously in Figures\ \ref{fig:posterior-simulation}b and \ref{fig:compare-intervals}b, many of these are subject to significant uncertainty and it is important therefore to discern which, if any, of the oscillations in the response can be identified given the model uncertainty.
+In Figure\ \ref{fig:derivatives}b only two features of the estimated trend are considered significant based on the derivatives of the smooth; one centred on ~250CE and a second at ~1150CE.
+In both these periods, the simultaneous interval for the first derivative of the trend excludes zero.
+In the first case we detect the large peak and subsequent decline in \uk{} at ~250CE, whilst at ~1150CE the large trough is identified, but not the increasing trend immediately prior to this excursion to lower \uk.
+Recall that these intervals are simultaneous in nature, strongly guarding against false positives, and as such we can be confident in the estimation of these two features, whilst care must be taken to not over-interpret the remaining variations in the estimated trend.
+
+```{r derivatives, fig.height = 5, fig.width = 8, fig.cap = "first derivatives and 95\\% simultaneous interval"}
+small.d <- fderiv(mod, newdata = newYear, n = N)
+small.sint <- with(newYear, cbind(confint(small.d, nsim = nsim, type = "simultaneous"), Year = Year))
+
+small_deriv_plt <- ggplot(small.sint, aes(x = Year, y = est)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "black") +
+    geom_line() +
+    labs(x = "Year CE", y = "First derivative")
+
+braya.d <- fderiv(fullREML2, newdata = newBraya, n = N)
+braya.sint <- with(newBraya, cbind(confint(braya.d, nsim = nsim, type = "simultaneous"), Year = Year))
+
+braya_deriv_plt <- ggplot(braya.sint, aes(x = Year, y = est)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "black") +
+    geom_line() +
+    labs(x = "Year CE", y = "First derivative")
+
+plot_grid(small_deriv_plt, braya_deriv_plt, ncol = 1, labels = "auto", align = "hv", axis = "lr")
+```
+
+### Residual autocorrelation
+
+The estimated value of $\phi$ from the fitted model is `r round(smallPhi[2], 2)` with 95% confidence interval `r round(smallPhi[1], 2)`--`r round(smallPhi[3], 2)`, indicating moderate to strong residual autocorrelation about the fitted trend.
+Failure to account for the dependencies in the δ^15^N time series could lead have lead to the estimation of a more wiggly trend than the one shown in Figure \ref{fig:plot-fitted-models}a which would negatively impact the confidence place on the inferences we might draw from the fitted model.
+Importantly, failing to account for the strong dependency in the residuals would lead to smaller uncertainties in the estimated spline coefficients, which would propagate through to narrower confidence intervals on the fitted trend and on the first derivatives, and ultimately to the identification of significant periods of change in the δ^15^N time series.
+
+```{r car1-plot, fig.height = 2.5, fig.width = 4, out.width = "0.5\\linewidth", fig.cap = "Estimated CAR(1) process from the GAM fitted to the Small Water $\\delta^{15}\\text{N}$ time series.
+$h(\\Delta_t, \\phi)$ is the correlation between residuals separated by $\\Delta_t$ years, where $\\hat{\\phi} = \\text{0.6}$.
+The shaded band is a 95\\% confidence interval on the estimated correlation $h$."}
+## plot CAR(1) process
+maxS <- with(small, diff(range(Year))) ## too large, truncate to 50
+S <- seq(0, 50, length = 100)
+
+car1 <- setNames(as.data.frame(t(outer(smallPhi, S, FUN = `^`)[1, , ])),
+                 c("Lower","Correlation","Upper"))
+car1 <- transform(car1, S = S)
+
+car1Plt <- ggplot(car1, aes(x = S, y = Correlation)) +
+    geom_ribbon(aes(ymax = Upper, ymin = Lower),
+                fill = "black", alpha = 0.2) +
+    geom_line() +
+    ylab(expression(italic(h) * (list(Delta[t], varphi)))) +
+    xlab(expression(Delta[t]))
+car1Plt
+```
+
+### Identifiability
+
+Consider again the basic GAM for a smooth trend, \eqref{eq:additive-model}.
+In that equation the correlation matrix $\boldsymbol{\Lambda}$ was ommitted for the sake of simplicity.
+Here I reintroduce it and restate the distributional assumptions of this model
+
+\begin{equation}
+y_t = \beta_0 + f(x_t) + \varepsilon_t, \quad \varepsilon \sim \mathcal(0, \boldsymbol{\Lambda}\sigma^2)
+\end{equation}
+
+In the basic GAM, $\boldsymbol{\Lambda} \equiv \mathbf{I}$ is an identity matrix, a matrix with 1s on the diagonal and 0s elsewhere.
+This is where the independence assumption of the model comes from; a model residual is perfectly correlated with itself (the 1s on the diagonal), but uncorrelated with any other residual (the off-diagonal 0s).
+In the GAM plus CAR(1) model, I specified an alternative correlation function for $\boldsymbol{\Lambda}$ --- the CAR(1) with correlation parameter $\phi$.
+@Fahrmeir2008-si show that where the stochastic structure of $f$ and $\boldsymbol{\Lambda}$ approach one another, where we have a potentially wiggly trend or strong autocorrelation as $\phi \rightarrow 1$, the two processes can quickly become unidentifiable [see also @Fahrmeir2013-xu **give pages**].
+By unidentifiable, we mean that it becomes increasingly difficult to distinguish between a wiggly trend or strong autocorrelation because these two processes are very similar to one another in appearance.
+
+Why might this be so?
+Autocorrelation is the tendency for a large (small) value of $y_t$ at time $x_t$ to be followed by a likewise large (small) value at time $x_{t+1}$.
+This leads to runs of values that are consistently greater (less) than the overall mean.
+Short runs would indicate weaker autocorrelation whilst longer runs are associated with stronger autocorrelation, and long runs of values greater (less) than the mean would evident as non-linear trends in the time series.
+As a result, a wiggly trend and an autocorrelation function with large $\phi$ are two ways to describe the same pattern of values in a time series, and without any further information to constrain either the model is unable to distinguish both components uniquely.
+
+A situation where it may be possible to uniquely identify separate wiggly trends and autocorrelation is exemplified in Small Water.
+The non-linear trend and the autocorrelation operate at very different scales; the trend represents decadal-scale variation in the mean δ^15^N, whilst the CAR(1) process represents the much smaller-scale tendency for values of the response to be followed in time by similar values.
+That such a pattern is observed in the Small Water core is the result of the high resolution of the sampling in time relative to the long-term trend.
+In contrast, the Braya-sø record is sampled at far lower resolution, and consequently the data do not contain sufficient information to separate trend and autocorrelation.
+
+### Gaussian process smooths
+
+In the world of machine learning, Gaussian processes are a widely-used method for fitting smooth non-parametric regression models.
+A Gaussian process is a distribution over all possible smooth functions $f(x)$.
+In the field of spatial statisics, Gaussian processes are known by name *kriging*.
+
+With a Gaussian process we are interested in fitting a smooth temporal or spatial trend by modelling the way the correlation between pairs of observations, $x_i$ and $x_j$, varies as a function of the distance in space or time that separates ($h$) the observations.
+The correlation between pairs of samples decreases with increasing separation, which is modelled using a correlation function, $c(h)$.
+Figure \ref{fig:gp-correlation-functions} shows examples of two different correlation functions; the *power exponential* (Figure \ref{fig:gp-correlation-functions}a), and the Matérn (Figure \ref{fig:gp-correlation-functions}b) correlation functions.
+These functions are smooth and monotonic-decreasing, meaning that the value of the correlation function decreases with increasing separation ($h$).
+When $h$ = 0, the correlation is equal to 1 ($c(0) = 1$); two samples take from exactly the same location or time point are perfectly correlated.
+As $h \rightarrow \infty$, the correlation tends to zero ($c(h) \rightarrow 0$); two samples separated by a large distance or amount of time tend to be uncorrelated.
+
+```{r gp-correlation-functions, fig.width = 8, fig.height = 5, fig.cap = "correlation functions"}
+`powerExp` <- function(h, phi, k) {
+    exp(- abs(h / phi)^k)
+}
+
+`matern` <- function(h, phi, k) {
+    absh <- abs(h / phi)
+    addn <- ifelse(k == 1.5, 0,
+                   ifelse(k == 2.5, (1/3)*absh^2,
+                          ((2/5)*absh^2) + ((1/15) * absh^3)))
+    exp(- absh) * (1 + absh + addn)
+}
+
+pExpdf <- expand.grid(h = seq(0, 7, length = 100), phi = c(1,2), k = c(0.5, 1, 2))
+pExpdf <- transform(pExpdf, correlation = powerExp(h, phi, k), kfac = factor(k))
+
+pexp.plt <- ggplot(pExpdf, aes(x = h, y = correlation, group = kfac, colour = kfac)) +
+    geom_line(size = 1) +
+    facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
+    ylab(expression(rho)) +
+    xlab(expression(italic(h))) +
+    ## scale_color_viridis(option = "viridis", discrete = TRUE) +
+    ## scale_colour_manual(values = viridis(11)[c(2,5,10)]) +
+    scale_colour_manual(name = expression(kappa), values = c("#e66101","#fdb863","#5e3c99"))
+
+pMaternDf <- expand.grid(h = seq(0, 10, length = 100),
+                         phi = c(1,2),
+                         k = c(1.5, 2.5, 3.5))
+pMaternDf <- transform(pMaternDf, correlation = matern(h, phi, k),
+                       kfac = factor(k))
+
+pmat.plt <- ggplot(pMaternDf, aes(x = h, y = correlation, group = kfac, colour = kfac)) +
+    geom_line(size = 1) +
+    facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
+    ylab(expression(rho)) +
+    xlab(expression(italic(h))) +
+    scale_colour_manual(name = expression(kappa), values = c("#e66101","#fdb863","#5e3c99"))
+
+correlFuns <- plot_grid(pexp.plt, pmat.plt, ncol = 1, align = "hv",
+                        labels = c("a","b"), axis = "lr")
+correlFuns
+```
+
+The power exponential function at separation distance $h$ is
+
+$$c(h) = exp\{(-h/\phi)^{\kappa}\}$$
+
+where $0 < \kappa \leq 2$.
+The Matérn correlation function is actually a family of functions with closed-forms only available for a subset of the family, distinguished by $\kappa$.
+When $\kappa = 1.5$, the Matérn correlation function is
+
+$$c(h) = (1 + h/\phi) exp(-h/\phi)$$
+
+whilst for $\kappa = 2.5$ it is
+
+$$c(h) = \{1 + h/\phi + (h/\phi)^2/3\} exp(-h/\phi)$$
+
+and for $\kappa = 3.5$
+
+$$c(h) = \{1 + h/\phi + 2(h/\phi)^2/5 + (h/\phi)^3/15\} exp(-h/\phi)$$
+
+In both cases, $\phi$ is the effective range parameter, which sets the distance beyond which the correlation function is effectively zero.
+
+```{r profile-range-parameter-in-gp-smooth-model}
+nn <- 200                               # number of points at which to evaluate profile likelihood
+dseq <- seq(15, 500, length.out = nn)  # effective ranges to fit at
+## Mat <- SExp <- vector("list", length = nn)     # object to hold model fits
+Mat <- SEx <- numeric(length = nn)     # object to hold model fits
+for (i in seq_along(dseq)) {            # iterate over dseq, fit GP GAM w Matérn covariance
+    Mat[i] <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, dseq[i])),
+                  weights = sampleInterval / mean(sampleInterval),
+                  data = braya, method = "REML", family = gaussian())[["gcv.ubre"]]
+    SEx[i] <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(2, dseq[i], 1)),
+                  weights = sampleInterval / mean(sampleInterval),
+                  data = braya, method = "REML", family = gaussian())[["gcv.ubre"]]
+}
+## extract the REML score into ggplot-friendly object
+## reml.scr <- data.frame(effrange = dseq, reml = sapply(mods, `[[`, "gcv.ubre"), aic = sapply(mods, AIC))
+reml.scr <- data.frame(cor = rep(c("Matérn","Exponential"), each = nn),
+                       effrange = rep(dseq, 2),
+                       reml = c(Mat, SEx))
+```
+
+```{r fit-example-gp-smooths, dependsons = "profile-range-parameter-in-gp-smooth-model"}
+effRange1 <- 250      # sets effective range in years for Matérn correl
+effRange2 <- with(subset(reml.scr, cor == "Matérn"), dseq[which.min(reml)])
+effRange3 <- with(subset(reml.scr, cor == "Exponential"), dseq[which.min(reml)])
+## Add time interval of each sample as prior weights; effectively averaging more time per sample
+## gp1 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, effRange1)), data = braya,
+##            method = "REML", weights = sampleInterval / mean(sampleInterval))
+gp2 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, effRange2)), data = braya,
+           method = "REML", weights = sampleInterval / mean(sampleInterval))
+gp3 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(2, effRange3, 1)), data = braya,
+           method = "REML", weights = sampleInterval / mean(sampleInterval))
+newd <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = 1000)))
+## p.gp1 <- transform(newd, fitted = predict(gp1, newdata = newd, type = "response"),
+##                    effRange = effRange1)
+p.gp2 <- transform(newd, fitted = predict(gp2, newdata = newd, type = "response"),
+                   effRange = round(effRange2))
+p.gp3 <- transform(newd, fitted = predict(gp3, newdata = newd, type = "response"),
+                   effRange = round(effRange3))
+## pred <- rbind(p.gp1, p.gp2, p.gp3)
+pred <- rbind(p.gp2, p.gp3)
+pred <- transform(pred, effRange = factor(effRange),
+                  cor = rep(c("Matérn", "Exponential"), each = nrow(newd)))
+```
+
+Gaussian processes and GAMs share many similarities and we can fit a Gaussian process using the techniques already described above for splines [@Handcock1994-lj].
+It can be shown [e.g.
+@Fahrmeir2013-xu section/pages] that the Gaussian process (kriging) model has the same penalised likelihood form as the GAM that we discussed earlier.
+Here, the observations are the knots of the smoother and each has a basis function in the form of a correlation function.
+The equivalence is only true if the basis functions do not depend on any other parameters of the model, which is only achievable if the value of $\phi$ is fixed and known.
+In general, however, we would like to estimate $\phi$ as part of model fitting.
+To achieve this we can maximise the profile likelihood or score statistic of the model over a range of values of $\phi$.
+This involves proposing a value of $\phi$ for the effective range of the correlation function and then estimating the resulting GAM by minimising the penalised log-likehood conditional upon this value of $\phi$.
+The model, and its corresponding value of $\phi$, with lowest penalised log-likelihood or score statistic is then retained as the estimated GAM.
+Figure\ \ref{fig:gp-gam-detail-plot}a shows the REML score for models estimated using a Gaussian process smooth with a Matérn correlation function, $\kappa$ = 1.5, for a values of $\phi$ between 15 and 1000.
+There is a clear minimum around 40 years separation, with the minimum REML score being observed at $\phi = `r round(effRange2, 2)`)$.
+Also shown are REML score for models using the power exponential function with $\kappa$ = 1, the minimum score observed at a somewhat higher effective range of $\phi = `r round(effRange3, 2)`$.
+
+```{r gp-gam-detail-plot, fig.height = 5, fig.width = 8, fig.cap = "profile likelihood", dependson="fit-example-gp-smooths"}
+## profile-likelihood plot
+proflik.plt <- ggplot(reml.scr, aes(x = effrange, y = reml, colour = cor)) +
+    geom_line() +
+    scale_colour_manual(name = "", values = c("#e66101","#5e3c99")) +
+    labs(y = "REML score", x = expression(Effective ~ range ~ (italic(h))))
+## plot at two values of h
+gp.plt2 <- ggplot(pred, aes(x = Year, y = fitted, colour = cor)) +
+    geom_line() + theme(legend.position = "right") +
+    geom_point(aes(x = Year, y = UK37), data = braya, inherit.aes = FALSE) +
+    scale_colour_manual(name = "", values = c("#e66101","#5e3c99")) +
+    labs(y = ylabel, x = "Year CE")
+plot_grid(proflik.plt, gp.plt2, ncol = 1, labels = c("a","b"), align = "hv", axis = "lr")
+```
+
+Figure \ref{fig:gp-gam-detail-plot}b shows the estimated trends for the \uk{} time series using Gaussian process smooths with exponential and Matérn correlations functions, both using effective range values at their respective optimal value as assessed using the REML score.
+The estimated trends are very similar to one another, although differences in the more-sparsely sampled sections of the time series, there is a noticeable difference in behaviour, with the power exponential ($\kappa = 1$) version being noticeably less-smooth than the Matérn version.
+This difference is attributable to the shapes of the respective correlation functions; the Matérn approaches a correlation of 1 smoothly as $h$ approaches 0, whilst the power exponential with $\kappa$ = 1 approaches a correlation of 1 increasingly quickly with decreasing $h$.
+The power exponential with $\kappa$ = 2, like the Matérn, approaches $\phi$ = 1 smoothly, and consequently the trend estimated using this correlation function is qualitatively similar to that estimated using the Matérn correlation function.
+
+### Adaptive smoothing
+
+```{r braya-so-model-comparisons, fig.width = 8, fig.height = 2.5, fig.cap = "compare fits", dependson = "fit-example-gp-smooths"}
+## model it using gam()
+effRange <- effRange2
+
+## Gaussian process, Matern, kappa = 1.5, weights as sampleInterval
+mod_gp <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, effRange)), data = braya,
+              method = "REML", weights = sampleInterval / mean(sampleInterval))
+
+## Adaptive spline, weights as sampleInterval
+mod_ad <- gam(UK37 ~ s(Year, k = 45, bs = "ad"), data = braya,
+              method = "REML", weights = sampleInterval / mean(sampleInterval))
+
+## TPRS, weights as sampleInterval, k = needs to be higher
+mod_tprs <- gam(UK37 ~ s(Year, k = 45, bs = "tp"), data = braya, method = "REML",
+                weights = sampleInterval / mean(sampleInterval))
+
+## wrap this in a function that will return all the plots and derived objects
+processGAM <- function(mod) {
+    ## Predict from model
+    N <- 500
+    newYear <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = N)))
+    newYear <- cbind(newYear, data.frame(predict(mod, newYear, se.fit = TRUE)))
+
+    out <- list(#plots = list(combinedPlt, plt.fit, plt.fit2, plt.sim, derivPlt),
+                objects = newYear)
+    out
+}
+
+plts_gp   <- processGAM(mod = mod_gp)      # Gaussian process smooth with weights
+plts_ad   <- processGAM(mod = mod_ad)      # Adaptive smooth with weights
+plts_tprs <- processGAM(mod = mod_tprs)    # TPRS with weights
+
+pltData <- do.call("rbind", lapply(list(plts_gp, plts_ad, plts_tprs),
+                                   `[[`, "objects"))
+pltData <- transform(pltData, Model = rep(c("GP", "Adaptive", "TPRS"),
+                              each = nrow(plts_gp$objects)))
+
+allFits <- ggplot(pltData, aes(x = Year, y = fit)) +
+    geom_point(aes(x = Year, y = UK37), data = braya) +
+    geom_line(aes(colour = Model)) + labs(y = ylabel, x = "Year") +
+    theme(legend.position = "right") +
+    scale_colour_manual(name = "", values = c("#e66101", "#fdb863", "#5e3c99"))
+allFits
+```
+
+Each of the spline types that I have discussed so far shares a common feature; the degree of wiggliness over the time series is constant.
+This is due to the wiggliness being controlled by a single smoothness parameter.
+The definition of wiggliness, as the integrated squared second derivative of the spline, ensures that fitted smoother does not jump about wildly.
+This assumes that the data themselves are well described by a smoothly varying function.
+If we anticipate abrupt change or step-like responses to environmental forcing this underlying assumption of the GAM would suggest it is ill-suited to modelling palaeo time series in which features are evident or expected.
+
+While there is not much we can do within the GAM framework to model a series that contains both smooth trends and step-like responses, an adaptive smoother can help address problems where the time series consists of periods of very rapid change in the mean combined with periods of complacency or relatively little change.
+As suggested by their name, adaptive smoothers can adjust to changes in the wiggliness of the time series.
+This adaptive behaviour is achieved by making the smoothness parameter $\lambda itself depend smoothly on $x_t$; in other words, the adaptive smoother allows the wiggliness of the estimated trend to vary smoothly over time.
+This is achieved by using multiple penalties and smoothness parameters, which have the effect of letting the strength of the wiggliness vary smoothly over time.
+Whilst this does allow the estimated trend to adapt to periods of rapid change in the response, adaptive smoothers make significant demands on the data; if we used $m$ smoothness penalties to allow the wiggliness to vary over a time series, it would be like estimating $m$ separate smooths from chunks of the original series each of length $n/m$.
+In a practical sense, this limits the use of adaptive splines in palaeoecology to proxies that are readily enumerated, such as the biogeochemical proxies used in the two example data sets.
+
+Figure \ref{fig:braya-so-model-comparisons} compares trends for the Braya-sø time series estimated using GAMs with a TPRS and a Gaussian process smooths with an adaptive smoother using a basis dimension, *k*, of 45 and 5 smoothing parameters.
+There is a clear difference in behaviour of the adaptive and non-adaptive smoothers for the first 1000 years of the record, with the adaptive smooth exhibiting much less variation compared to the TPRS and Gaussian process spline.
+Over the remaining two thirds of the series, there is much closer agreement in the three trends.
+
+The behaviour of the TPRS and Gaussian process spline for these data is the result of requiring a large amount of wiggliness to adapt to the large oscillations in \uk{} present around year 250CE and again at ~900--1500CE.
+This large degree of wiggliness allows the splines to chase individual data points much earlier in the record.
+Because the adaptive smoother, in contrast, can adapt to these periods of rapid change in the response it is much less susceptible to this "chasing" behaviour.
+
+The behaviour of the TPRS- and Gaussian process-based models is undesirable, yet if we recall Figure \ref{fig:derivatives} and the discussion around the use of the first derivative to identify periods of significant change, we would not interpret the oscillations in the early part of the \uk{} record as being statistically significant.
+Owing to the paucity of data in this part of the series the trends fitted using the non-adaptive smoothers, the estimated trends are subject to such a large degree of uncertainty that the alternative of no trend through the first 1000 years of the record is also a plausible explanation of the data.
+Should we conclude that there is no trend in \uk{} and thence climate in this period?
+I believe that to be too-strong a statement; those oscillations in \uk{} may be real responses to climate forcing but we may simply lack the statistical power to distinguish them from the null hypothesis of no trend through this period.
+The adaptive smoother is only adjusting to the data available to it; simply because it does not detect a trend during this period does not lend itself to an interpretation of stable climate forcing or complacency in the lake's response to forcing.
+If there were particular interest in the climate of this particular period we might take from the Braya-sø record that there is potential variation in climate forcing, but that additional data from this or other sites is required before any definitive conclusion can be drawn.
+
+### Accounting for age model uncertainty
+
+```{r small-scam-fit}
+knots <- with(small, list(Year = seq(min(Year), max(Year), length = 14)))
+mod <- gamm(d15N ~ s(Year, k = 15), data = small, method = "REML",
+            correlation = corCAR1(form = ~ Year),
+            knots = knots)
+
+swAge <- read.csv("./data/small-water/small1-dating.csv")
+
+## monotonic spline age-depth model
+swAge$Error[1] <- 1.1
+swAgeMod <- scam(Date ~ s(Depth, k = 5, bs = "mpd"), data = swAge, weights = 1 / swAge$Error, gamma = 1.4)
+
+## predict from the age model for a smooth set of points in `Depth`
+newAge <- with(swAge, data.frame(Depth = seq(min(Depth), max(Depth), length.out = 200)))
+newAge <- transform(newAge, fitted = predict(swAgeMod, newdata = newAge, type = "response"))
+newSims <- as.data.frame(simulate(swAgeMod, nsim = 25, newdata = newAge))
+newSims <- cbind(Depth = newAge$Depth, newSims)
+newSims <- gather(newSims, Simulation, Age, -Depth)
+
+## simulate from age model; each column is a simulation
+ageSims <- simulate(swAgeMod, nsim = 100, newdata = small, seed = 42)
+ageSims <- as.data.frame(ageSims)
+
+fitSWModels <- function(x, y, knots) {
+    dat <- data.frame(d15N = y, Year = x)
+    m <- gamm(d15N ~ s(Year, k = 15), data = dat, method = "REML",
+              correlation = corCAR1(form = ~ Year), knots = knots)
+}
+
+predSWModels <- function(mod, newdata) {
+    predict(mod$gam, newdata = newdata, type = "response")
+}
+
+simulateSWModels <- function(mod, newdata, nsim, seed = 42) {
+    sims <- simulate(mod, nsim = nsim, newdata = newdata, seed = seed)
+    as.vector(sims)
+}
+
+simTrendMods <- lapply(ageSims, fitSWModels, y = small$d15N, knots = knots)
+simTrends <- lapply(simTrendMods, predSWModels, newdata = newYear)
+simTrends <- data.frame(Year  = with(newYear, rep(Year, length(simTrends))),
+                        Trend = unlist(simTrends),
+                        Group = rep(seq_along(simTrends), times = lengths(simTrends)))
+
+NSIM <- 50
+simSimulate <- lapply(simTrendMods, simulateSWModels, newdata = newYear, nsim = NSIM, seed = 42)
+simSimulate <- data.frame(Year  = with(newYear, rep(Year, times = NSIM * length(simSimulate))),
+                          Trend = unlist(simSimulate),
+                          Group = rep(seq_len(NSIM * length(simSimulate)), each = nrow(newYear)))
+```
+
+Thus far, the trend models that I have described and illustrated assumed that the time covariate was fixed and known.
+In both examples, and more generally for most palaeoecological records, this assumption is violated.
+Unless the record is annually laminated, assigning an age to a sediment interval requires the development of an age model from observations of the relationship between depth down the sediment core and estimates of the age of the sample arrived at using any of a number of techniques, for example ^210^Pb or ^14^C radiometric dating.
+This age-depth relationship is itself uncertain, usually being derived from a mathematical or statistical model applied to point age estimates.
+Incorporating this additional component of uncertainty complicates the estimation of statistical models from palaeoenvironmental data.
+In this section I illustrate a simulation based approach to quantify and account for age-model uncertainty as part of the trend estimation using a GAM.
+
+```{r small-scam-fit-plots, fig.width = 8, fig.height = 7.5, fig.cap = "foo", dependson = -1}
+plt1 <- ggplot(swAge, aes(y = Date, x = Depth)) +
+    geom_errorbar(aes(ymin = Date - Error, ymax = Date + Error, width = 0)) +
+    geom_point() +
+    geom_line(data = newSims, mapping = aes(y = Age, x = Depth, group = Simulation),
+              alpha = 0.1, colour = "#e66101") +
+    geom_line(data = newAge, mapping = aes(y = fitted, x = Depth)) +
+    labs(y = "Year CE", x = "Depth (cm)")
+
+plt2 <- ggplot(simTrends, aes(x = Year, y = Trend, group = Group)) +
+    geom_line(alpha = 0.1, colour = "#e66101") +
+    geom_line(data = newYear, mapping = aes(x = Year, y = fit), inherit.aes = FALSE) +
+    geom_point(data = small, mapping = aes(x = Year, y = d15N), inherit.aes = FALSE, size = 0.5) +
+    labs(x = "Year", y = d15n_label)
+
+plt3 <- ggplot(simSimulate, aes(x = Year, y = Trend, group = Group)) +
+    geom_line(alpha = 0.1, colour = "#e66101") +
+    geom_point(data = small, mapping = aes(x = Year, y = d15N), inherit.aes = FALSE,
+               size = 0.5) +
+    geom_line(data = newYear, mapping = aes(x = Year, y = fit), inherit.aes = FALSE) +
+    labs(x = "Year", y = d15n_label)
+
+plot_grid(plt1, plt2, plt3, ncol = 1, labels = "auto", align = "hv", axis = "lrtb",
+          rel_widths = c(0.5, 1, 1))
+```
+
+Figure \ref{fig:small-scam-fit-plots}a shows the estimated dates (in Years CE) for `r nrow(swAge)` levels in the Small Water core dated using ^210^Pb.
+The vertical bars show the estimated age uncertainty of each level.
+The solid line through the data points is an additive model fitted to the observations, with prior weights given by the estimated age uncertainties.
+The fitted age-depth model is constrained to be monotonically decreasing with increasing depth, following the method of [@Pya2015-wr] using the **scam** package [@pya-scam-pkg].
+Also shown are 25 simulations from the model posterior of the montonically constrained GAM.
+Each simulation from the posterior distribution of the age-model is itself a potential age-depth model which can be used to assign dates to the Small Water core.
+The trend model in \ref{eq:gam} can be fitted to the δ^15^N data using these new dates as a covariate, and the whole process repeated for a large number of simulations.
+
+Figure \ref{fig:small-scam-fit-plots}b shows the trend in δ^15^N for the observed age-depth model, plus trends estimated via the same model using 100 draws from the posterior distribution of the age model.
+In this case, the age-depth model is relatively simple with little variation in the posterior draws, resulting in trends that match closely that obtained from the estimated age-depth relationship.
+Even so, this additional uncertainty accounts suggests that the timing of the obvious decline in δ^15^N covers the interval ~1935--1945.
+
+The uncertainty in the trend estimates illustrated in Figure \ref{fig:small-scam-fit-plots}b only reflects the variation in trends fitted to the uncertain dates of the sediment samples.
+To fully visualise the uncertainty in the trend estimates, incorporating both age model uncertainty *and* uncertainty in the estimated model coefficients themselves, 50 simulations from the posterior distributions of each of the 100 estimated trends shown in Figure \ref{fig:small-scam-fit-plots}b we performed, resulting in 5,000 trend estimates for the δ^15^N series.
+These are shown in Figure \ref{fig:small-scam-fit-plots}c, where the two obvious changes over the same simulations (Figure \ref{fig:posterior-simulation}a) are that the uncertainty band traced out by the simulations is approximately 50% wider and, not suprisingly, the uncertainty in the estimated trend is most pronounced in the least accurately-dated section of the core.
+Despite this additional uncertainty however, the main result holds; a marked decline of ~1.5‰ that occurred between approximately 1930 and 1945, with mild evidence of a small increase in δ^15^N post 2000 CE.
+
+Although beyond the scope of this analysis, it is anticipated that uncertainties in the age model for Braya-sø will induce significant variation in the trend estimation process such that some of the features tentatively identified using the first differences would no-longer met the required level of significance to claim identification.
+This is not to say that the features in the estimated trend are not real, just that if they are real, we lack the statistical power to identify them unequivocally given the data and the signal to noise ratio therein.
+
+## Further considerations
+
+### Multivariate data
+
+## Discussion
+
+## References
+
+<div id = "refs"></div>
+
+## Appendix
+
+## Title
+
+For a set of $M$ locations over the time period of interest, which we denote by $\mathbf{g} = (g_1, g_2, \ldots, g_M)$ the 1 - $\alpha$ *simultaneous* confidence interval is
+
+$$
+\begin{aligned}
+\mathbf{\hat{f}_g} & \pm m_{1 - \alpha}
+  \begin{bmatrix} \widehat{\mathrm{st.dev}} (\hat{f}(g_1) - f(g_1)) \\
+	\widehat{\mathrm{st.dev}} (\hat{f}(g_2) - f(g_2)) \\
+	\vdots \\
+	\widehat{\mathrm{st.dev}} (\hat{f}(g_M) - f(g_M)) \\
+	\end{bmatrix}
+\end{aligned}
+$$
+
+where $\mathbf{f_g}$ is the true, but unknown, function and $\mathbf{\hat{f}_g}$ the estimated function at $\mathbf{g}$.
+The values in the square brackets are the standard error of the difference between the true and estimated functions at each location in $\mathbf{g}$; these are simply the standard errors of predictions from the fitted GAM.
+$m_{1 - \alpha}$ is the critical value by which the standard errors are multiplied to form the upper and lower bounds of the $1 - \alpha$ simultaneous interval.
+$m_{1 - \alpha}$ is the $1 - \alpha$ quantile of the random variable
+
+$$
+\sup_{x \in \mathcal{X}} \left | \frac{\hat{f}(x) - f(x)}{\widehat{\mathrm{st.dev}} (\hat{f}(x) - f(x))} \right |
+$$
+
+where $\sup_{x \in \mathcal{X}}$ is the *supremum* or the *least upper bound* of the absolute values of the standardised deviation of the fitted function from the true function.
+The supremum is the least value of $\mathcal{X}$, the set of all values of which we observed subset $x$, that is *greater* than all of the values in the subset.
+We approximate $\sup_{x \in \mathcal{X}}$ as the maximum absolute standardised deviation of the estimated function from the true function.
+Additionally, we don't know the distribution of $\sup_{x \in \mathcal{X}}$, so we approximate it via posterior simulation.

### 2018-04-28 manuscript.Rmd
updated manuscript

@@
 
-### Data
-
-intro on the two data sets goes here.
+### Example time series
+
+To illustrate the estimation of trends in palaeoenvironmental data using GAMs, I use two proxy time series; a 150-year bulk organic matter δ^15^N time series from Small Water, and a 3000-year alkenone record from Braya-sø.
+Between them, the two examples, combine many of the features of interest to palaeoecologists that motivate the use of GAMs; non-linear trends and the question of when changes in the measured proxy occurred.
+The example analyses were all performed using the *mgcv* package and R, and the supplementary material contains a fully annotated document showing the R code used to replicate all the analyses described in the remainder of the paper.
 
@@
 The \uk{} unsaturation index is
-
 \begin{equation*}
@@
 \end{equation*}
-
 where $[C_{37:x}]$ is the concentration of the alkenone with 37 carbon atoms and $x$ double carbon bonds.
@@
 
-### Regression models for palaeoenvironmental time series
+In the subsequent sections, I present regression models that can be used to estimate trends in palaeoenvironmental time series and give an introduction to GAMs and how they estimate smooth, non-linear trends from time series.
+Next the issue of uncertainty in model-estimated trends is discussed, as is the topic of posterior simulation from a regression model.
+The issue of identifying periods of significant environmental change is considered and I show how this can be achieved using the first derivative of the estimated trend.
+Two non-standard types of spline --- adaptive smoothers and Gaussian process splines --- that are especially applicable to GAMs in the palaeoenvironmental setting are described, followed by an assessment of the the impact of age-model uncertainty on trends estimation via GAMs.
+Finally, I briefly discuss the application of GAM trend analysis to multivariate species abundance and compositional data.
+
+## Regression models for palaeoenvironmental time series
 
 A linear model for a trend in a series of $T$ observations $y_t$ at observation times $x_t$ with $t = 1, 2, \ldots, T$ is
-
 \begin{equation} \label{eq:linear-model}
-y_t = \beta_0 + \beta_1 x_t + \varepsilon_t,
+y_t = \beta_0 + \beta_1 x_t + \varepsilon_t \,,
 \end{equation}
-
 where $\beta_0$ is a constant term, the model *intercept*, representing the expected value of $y_t$ where $x_t$ is 0.
@@
 
-The linear model described above is quite restrictive in terms of the types of trend it can fit; essentially linear increasing or decreasing trends, or trivially, a null trend of no change.
-This model can be extended to allow for non-linear trends, most notably via making $y_t$ depend on polynomials of $x_t$, for example
-
+The linear model described above is quite restrictive in terms of the types of trend it can fit; essentially linear increasing or decreasing trends, or, trivially, a null trend of no change.
+This model can be extended to allow for non-linear trends, most notably by making $y_t$ depend on polynomials of $x_t$, for example
 \begin{align} \label{eq:polynomial-model}
 y_t &= \beta_0 + \beta_1 x_t + \beta_2 x_t^2 + \cdots + \beta_P x_t^P + \varepsilon_t \\
-    &= \beta_0 + \sum_{p = 1}^P \beta_p x_t^p  + \varepsilon_t \nonumber %\\
-    %&= \boldsymbol{\beta} \mathbf{X} + \boldsymbol{\varepsilon}
+    &= \beta_0 + \sum_{p = 1}^P \beta_p x_t^p  + \varepsilon_t \,, \nonumber
 \end{align}
-
 where polynomials of $x_t$ up to order $P$ are used.
-This model allows for more complex fitted trends but it remains a fully parametric model and suffers from several problems, especially the behaviour of the fitted polynomial at the start and end of the observed series.
+This model allows for more complex trends but it remains a fully parametric model and suffers from several problems, especially the behaviour of the fitted polynomial at the start and end of the observed series.
 
@@
 
-Linear models employing a range of polynomials of the covariate Year for the Small Water data set are shown in Figure \ref{fig:polynomial-example-plot}.
+Linear models using a range of polynomials for the Small Water data set are shown in Figure \ref{fig:polynomial-example-plot}.
 The low-order models ($P \in \{1, 3\}$) result in very poor fit to the data.
@@
 The $P = 10$th-order polynomial model is well able to capture this period of rapid change, but it does so at the expense of increased complexity in the estimated trend prior to ~1940.
-Additionally, this model ($P = 10$) has undesirable behaviour at the ends of the series, significantly overfitting the data, a commonly observed in polynomial models such as these.
-Finally, the choice of what order of polynomial to fit is an additional choice left for the analyst to specify; choosing the optimal $p$ is not a trivial task when the data are a time series and residual autocorrelation is likely present.
+Additionally, this model ($P = 10$) has undesirable behaviour at the ends of the series, significantly overfitting the data, a commonly observed problem in polynomial models such as these.
+Finally, the choice of what order of polynomial to fit is an additional choice left for the analyst to specify; choosing the optimal $P$ is not a trivial task when the data are a time series and residual autocorrelation is likely present.
 
@@
 Can we do better than these polynomial fits?
-In the remainder , I hope to demonstrate that the answer to that question is emphatically "yes".
+In the remainder , I hope to demonstrate that the answer to that question is emphatically "yes"!
 Below I describe a coherent and consistent approach to modelling palaeoenvironmental time series using generalised additive models (GAMs), which builds upon the linear regression framework.
@@
 GAMs extend the generalised linear model (GLM) --- of which the linear model described above is a special case --- where some or all of the parametric terms, the $\beta_p$, are replace by smooth functions $f_j$ of the covariates $x_j$.
-The GAM version of \eqref{eq:linear-model} is
+The GAM version of the linear model \eqref{eq:linear-model} is
 
@@
 
-where $\mu_t$ is the expected value of the random variable $Y_t$ ($\mu_t \equiv \mathbb{E}(Y_t)$) of which we have observations $y_t$.
-$g$ is the link function, an invertible, monotonic function, such as the natural logarithm, and $g^{-1}$ its inverse.
+where $\mu_t$ is the expected value (e.g.\ the mean count or the probability of occurrence) of the random variable $Y_t$ ($\mu_t \equiv \mathbb{E}(Y_t)$) of which we have observations $y_t$.
+$g$ is the link function, an invertible, monotonic function, such as the natural logarithm, and $g^{-1}$ is its inverse.
 The link function maps values from the linear predictor scale on to the scale of the response.
@@
 On the log scale, the response can take any real value between $-\infty$ and $+\infty$, and it is on this scale that model fitting actually occurs.
-However we need to map these unbounded back on to the non-negative response scale the.
-The inverse of the log link function is used, the exponential function, maps values to the interval 0--$\infty$.
+However we need to map these unbounded values back on to the non-negative response scale the.
+The inverse of the log link function, the exponential function, achieves this and maps values to the interval 0--$\infty$.
 
 In \eqref{eq:gam}, we further assume that the observations are drawn from a member of the exponential family of distributions --- such as the Poisson for count data, the binomial for presence/absence or counts from a total --- with expected value $\mu_t$ and possibly some additional parameters $\Theta$ ($y_t \sim \text{EF}(\mu_t, \Theta)$).
-Additionally, many software implementations of the above models now allow for the fitting of distributions that are not within the exponential family but which can be fitted using an algorithm superficially similar to the one used to fit GAMs to members of the exponential family.
+Additionally, many software implementations of the above model also allow for distributions that are not within the exponential family but which can be fitted using an algorithm superficially similar to the one used to fit GAMs to members of the exponential family.
 Common examples of such extended families include the negative binomial distribution (for overdispersed counts) and the beta distribution (for true proportions or other interval-bounded data).
@@
 These wiggly functions are smooths, which we denote by $f(x_t)$.
-We need to represent the smooths in a way that (\ref{eq:gam}) is a linear model, which is achieved by using a *basis*.
+We would like to represent the smooths in a way that \eqref{eq:gam} is a linear model so that we can estimate the model within the well-studied GLM framework, which is achieved by using a *basis*.
 A basis is a set of functions that collectively span a space of functions that, we hope, contains the true $f(x_t)$ or a close approximation to it.
@@
 $$
-f(x_t) = \sum_{j = 1}^{k} b_j(x_t) \beta_j
+f(x_t) = \sum_{j = 1}^{k} b_j(x_t) \beta_j \,,
 $$
@@
 The polynomial model we encountered earlier is an example of a statistical model that uses a basis expansion.
-For the cubic polynomial ($P = 3$) fit shown in Figure \ref{fig:polynomial-example-plot} there are in fact 4 basis functions: $b_1(x_t) = x_t^0 = 1$, $b_2(x_t) = x_t$, $b_3(x_t) = x^2_t$, $b_4(x_t) = x_t^3$.
-Note that here $b_1(x_t)$ is constant and is in fact the model intercept, $\beta_0$, in (\ref{eq:polynomial-model}), and the weights are the estimated coefficients in the model, the $\beta_j$.
+For the cubic polynomial ($P = 3$) fit shown in Figure \ref{fig:polynomial-example-plot} there are in fact 4 basis functions: $b_1(x_t) = x_t^0 = 1$, $b_2(x_t) = x_t$, $b_3(x_t) = x^2_t$, and $b_4(x_t) = x_t^3$.
+Note that here $b_1(x_t)$ is constant and is in fact the model intercept, $\beta_0$, in the linear model \eqref{eq:polynomial-model}, and the weights are the estimated coefficients in the model, the $\beta_j$.
 
 As we have already seen, polynomial basis expansions do not necessarily lead to well-fitting models unless the true function $f$ is itself a polynomial.
-One of the primary criticisms is that polynomial basis functions are global; the value of $f$ at time point $x_t$ affects the value of $f$ at time point $x_{t+n}$ even if the two time points are at opposite ends of the series.
+One of the primary criticisms is that polynomial basis functions are global; the value of $f$ at time point $x_t$ affects the value of $f$ at time point $x_{t+s}$ even if the two time points are at opposite ends of the series.
 There are many other bases we could use; here I discuss one such set of bases, that of splines.
@@
 
-A CRS can be parameterized in a number of different ways.
+The CRS can be parameterized in a number of different ways.
 One requires a knot at each unique data value in $x_t$, which is computationally inefficient.
-One approach to specifying a CRS basis is to parameterize in terms of the value of the spline at the knots.
-Typically in this parametrization there are many fewer knots than in alternative parameterizations, and these knots are either distributed evenly over the range of $x_t$ or at the quantiles of $x_t$.
+Another way of specifying a CRS basis is to parameterize in terms of the value of the spline at the knots.
+Typically in this parametrization there are many fewer knots than unique data, with the knots distributed evenly over the range of $x_t$ or at the quantiles of $x_t$.
 Placing knots at the quantiles of $x_t$ has the effect of placing a greater number of knots where the data is densest.
@@
 Thin plate regression splines (TPRS) remove this element of subjectivity when fitting GAMs.
-Thin plate splines were introduced by @Duchon1977-jr and, as well as solving the knot selection problem, have several additional attractive properties in terms of optimality and their ability to estimate a smooth function of two or more variables, leading to the inclusion of smooth interactions between covariates.
+Thin plate splines were introduced by @Duchon1977-jr and, as well as solving the knot selection problem, have several additional attractive properties in terms of optimality and their ability to estimate a smooth function of two or more variables, leading to smooth interactions between covariates.
 However, thin plate splines have one key disadvantage over CRS; thin plate splines have as many unknown parameters as there are unique combinations of covariate values in a data set.
-It is unlikely in the extreme that any real data problem would require functions of such complexity that they require so many basis functions.
-It is much more likely that the true functions that we attempt to estimate are far simpler than the set of functions able to be represented by 1 basis function per unique data value.
-From a practical point of view, it is highly inefficient to carry around all these basis functions whilst model fitting, and the available computational resources would become quickly exhausted for large time series with many observations.
+It is unlikely in the extreme that any real data problem would involve functions of such complexity that they require as many basis functions as data.
+It is much more likely that the true functions that we attempt to estimate are far simpler than the set of functions representable by 1 basis function per unique data value.
+From a practical point of view, it is also highly inefficient to carry around all these basis functions whilst model fitting, and the available computational resources would become quickly exhausted for large time series with many observations.
 
@@
 The truncation suggested by @Wood2003-qy is not without cost; the eigen-decomposition and related steps can be relatively costly to set up for large data sets.
-For data sets of similar size to the two examples used here, the additional computational effort required to set up the thin plate regression spline basis over say the cubic regression spline basis will not be noticeable.
-For highly resolved data sets containing more than 500--1000 observations the truncation may be too costly computationally.
-In such instances, little is lost by moving to the cubic regression spline basis, with the same number of knots as the rank of the desired thin plate spline, either distributed evenly throughout the data or at the quantiles of the data, with the benefit of considerably reduced set up time for the basis.
-
-With the two regression spline bases described above, to fit a GAM the analyst is generally only required to the specify the size (rank, number of knots) of the basis expansion required to represent or closely approximate the true function $f$.
+For data sets of similar size to the two examples used here, the additional computational effort required to set up the TPRS basis over the CRS basis will not be noticeable.
+For highly resolved series containing more than 500--1000 observations the truncation may be too costly computationally.
+In such instances, little is lost by moving to the CRS basis, with the same number of knots as the rank of the desired thin plate spline, either distributed evenly throughout the data or at the quantiles of the data, with the benefit of considerably reduced set up time for the basis.
+
+To fit a GAM using either of the two regression spline bases described above, the analyst is generally only required to the specify the size (rank) of the basis expansion required to represent or closely approximate the true function $f$.
 With practice and some knowledge of the system from which the observations arise, it can be relatively easy to put an upper limit on the expected complexity of the true trend in the data.
@@
 In practice, the size of the basis is an upper limit on the expected complexity of the trend, and a simple test is available to check if the basis used was sufficiently large [@Pya2016-rk].
-This test is available via the `gam.check()` function in **mgcv** for example, and essentially looks at whether there is any additional nonlinearity or structure in the residuals that can be explained by $x_t$.
-Should a smooth term in the fitted model fail this test the model can be refitted using a larger basis expansion, say by doubling the value of `k` used to fit the original.
+This test is available via the `gam.check()` function in *mgcv* for example, and essentially looks at whether there is any additional nonlinearity or structure in the residuals that can be explained by $x_t$.
+Should a smooth term in the fitted model fail this test the model can be refitted using a larger basis expansion, say by doubling the value of `k` (the rank) used to fit the original.
 Note also that a smooth might fail this test whilst using fewer effective degrees of freedom than the maximum possible for the dimension of basis used.
 This may happen when the true function lies at the upper limit of the set of functions encompassed by the size of basis used.
-Additionally, a basis of size $2k$ encompasses a larger space of functions of a given complexity than a basis of size $a$ [@Wood2017-qi]; increasing the basis dimension used to fit the model may unlock this additional function space resulting in a better fitting model whilst using a similar number of effective degrees of freedom.
+Additionally, a basis of size $2k$ encompasses a richer space of functions of a given complexity than a basis of size $k$ [@Wood2017-qi]; increasing the basis dimension used to fit the model may unlock this additional function space resulting in a better fitting model whilst using a similar number of effective degrees of freedom.
 
@@
 Having identified regression splines as a useful way to represent $f$, we next need a way to decide how wiggly the fitted trend should be.
-A backwards elimination approach to sequentially remove knots or basis functions might seem appropriate, however such an approach would likely fail as the resulting sequence of models would not strictly nested, precluding many forms of statistical comparison.
-Alternatively, we could keep the basis dimension at a fixed size but gaurd against fitting very complex models through the use of a wiggliness penalty.
+A backwards elimination approach to sequentially remove knots or basis functions might seem appropriate, however such an approach would likely fail as the resulting sequence of models would not be strictly nested, precluding many forms of statistical comparison.
+Alternatively, we could keep the basis dimension at a fixed size but guard against fitting very complex models through the use of a wiggliness penalty.
 
@@
 
-$$\mathcal{L}_p(\boldsymbol{\beta}) = \mathcal{L}(\boldsymbol{\beta}) - \frac{1}{2} \lambda\boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}$$
+$$\mathcal{L}_p(\boldsymbol{\beta}) = \mathcal{L}(\boldsymbol{\beta}) - \frac{1}{2} \lambda\boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}\, .
+$$
 
@@
 GCV avoids the computational overhead inherent to CV of having to repeatedly refit the model with one or more observations left out as a test set.
-Minimising the GCV score will, with a sufficiently large data set, find a model with the minimal prediction error.
-The second approach is to treat the smooth as a random effect, in which $\lambda$ is now a variance parameter to be estimated using maximum likelihood (ML) or restricted maximum likelihood (REML).
+Minimising the GCV score will, with a sufficiently large data set, find a model with the minimal prediction error [@Wood2017-qi].
+The second approach is to treat the smooth as a random effect, in which $\lambda$ is now a variance parameter to be estimated using maximum likelihood (ML) or restricted maximum likelihood (REML) [@Wood2011-kn; @Wood2016-fx].
 
@@
 
-### Smoothing time series data
-
-**Do I need this as a separate section?** Or just cover the points as they occur?
-
-## Fitting GAMs
+## Fitting GAMs (Need a better title)
 
@@
 The δ^15^N values are irregularly spaced in time and as a result a correlation structure that can handle the uneven spacing of the samples is needed.
-A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the first-order autoregressive process (AR(1)) and, simply stated, models the correlation between any two residuals as $\phi^h$, where $h$ is the amount of separation in time between the residuals.
-$h$ may be a real valued number in the CAR(1), which accommodates the irregular separation of samples.
-$\phi$ controls how quickly the correlation between any two residuals declines as a function of their separation in time.
-$\phi$ is an additional parameter that will be estimated during model fitting.
-The model in \ref{eq:small-gam} was fitted using the `gamm()` function [@Wood2004-zv] in the **mgcv** package [@Wood2017-qi] for R [@R-core-team].
+A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the first-order autoregressive process (AR(1)) and, simply stated, models the correlation between any two residuals as and exponentially decreasing function of $h$ ($\phi^h$), where $h$ is the amount of separation in time between the residuals.
+$h$ may be a real valued number in the CAR(1), which is how it can accommodate the irregular separation of samples in time.
+$\phi$ controls how quickly the correlation between any two residuals declines as a function of their separation in time and is an additional parameter that will be estimated during model fitting.
+The model in \eqref{eq:small-gam} was fitted using the `gamm()` function [@Wood2004-zv] in the *mgcv* package [@Wood2017-qi] for R [@R-core-team].
 
@@
 
-## convert to gtabl to align heights
-g_small_fitted <- ggplotGrob(small_fitted) # convert to gtable
-g_braya_fitted <- ggplotGrob(braya_fitted) # convert to gtable
-
-height_take <- c(4,5)
-small_heights <- g_small_fitted$heights[height_take]
-braya_heights <- g_braya_fitted$heights[height_take]
-max_heights <- unit.pmax(small_heights, braya_heights) # calculate maximum heights
-g_small_fitted$heights[height_take] <- max_heights # assign max heightss to small gtable
-g_braya_fitted$heights[height_take] <- max_heights
-
-plot_grid(g_small_fitted, g_braya_fitted, ncol = 1, labels = "auto", align = "hv", axis = "lr")
+plot_grid(small_fitted, braya_fitted, ncol = 1, labels = "auto", align = "hv", axis = "lr")
 ```
@@
 However further analysis of the fitted model is required to answer the other questions posed earlier about the timing of change and whether features in the trend can be distinguished from random noise.
+I discuss these issues shortly.
 
@@
 The \uk{} data present a more difficult data analysis challenge than the δ^15^N time series because of the much more complex variation present.
-Fitting the same model as the Small Water example, \ref{eq:small-gam}, to the \uk{} data resulted in the unsatisfactory fit shown as the very smooth line in Figure \ref{fig:plot-fitted-models}b (labelled GAMM (CAR(1))).
+Fitting the same model as the Small Water example, \eqref{eq:small-gam}, to the \uk{} data resulted in the unsatisfactory fit shown as the very smooth line in Figure \ref{fig:plot-fitted-models}b (labelled GAMM (CAR(1))).
 Further problems were evident with this model fit --- the covariance matrix of the model was non-positive definite, a sure sign of problems with the fitted model.
@@
 
-To investigate the first issue the GCV and REML scores for an increasing sequence of values of the smoothness penalty $\lambda$ were evaluated for the standard GAM (equation \ref{eq:gam}) fit to the \uk{} time series.
+To investigate the first issue the GCV and REML scores for an increasing sequence of values of the smoothness penalty $\lambda$ were evaluated for the standard GAM (equation \eqref{eq:gam}) fit to the \uk{} time series.
 The resulting profiles are shown in Figure \ref{fig:trace-smoothness-parameters}, with the selected value of the penalty shown by the vertical line.
@@
 As suggested by their name, adaptive smoothers can adjust to changes in the wiggliness of the time series.
-This adaptive behaviour is achieved by making the smoothness parameter $\lambda itself depend smoothly on $x_t$; in other words, the adaptive smoother allows the wiggliness of the estimated trend to vary smoothly over time.
+This adaptive behaviour is achieved by making the smoothness parameter $\lambda$ itself depend smoothly on $x_t$; in other words, the adaptive smoother allows the wiggliness of the estimated trend to vary smoothly over time.
 This is achieved by using multiple penalties and smoothness parameters, which have the effect of letting the strength of the wiggliness vary smoothly over time.
@@
 The solid line through the data points is an additive model fitted to the observations, with prior weights given by the estimated age uncertainties.
-The fitted age-depth model is constrained to be monotonically decreasing with increasing depth, following the method of [@Pya2015-wr] using the **scam** package [@pya-scam-pkg].
+The fitted age-depth model is constrained to be monotonically decreasing with increasing depth, following the method of [@Pya2015-wr] using the *scam* package [@pya-scam-pkg].
 Also shown are 25 simulations from the model posterior of the montonically constrained GAM.
 Each simulation from the posterior distribution of the age-model is itself a potential age-depth model which can be used to assign dates to the Small Water core.
-The trend model in \ref{eq:gam} can be fitted to the δ^15^N data using these new dates as a covariate, and the whole process repeated for a large number of simulations.
+The trend model in \eqref{eq:gam} can be fitted to the δ^15^N data using these new dates as a covariate, and the whole process repeated for a large number of simulations.
 
@@
 
-## Further considerations
-
 ### Multivariate data
 
-## Discussion
+A large proportion of the palaeoenvironmental data generated today is multivariate species or biological abundance in nature and yet the two examples that have been used here to illustrate the use of GAMs to model palaeoenvironmental time series were univariate.
+Can the approach described here be used for multivariate data?
+Yes and no.
+With one main exception it is not possible to directly apply the GAM methodology to multivariate abundance data, where the aim is to model all species at once, with the methods that I have described here and illustrate in the supplementary materials.
+The available software is not able to estimate the penalized GAM for multiple non-Gaussian responses.
+The exception is for a small number of correlated Gaussian responses; these could be modelled as being distributed multivariate normal conditional upon the covariates.
+Such a model would estimate the expected values of each response and the correlation between them.
+For example, we could model δ^15^N and δ^13^C series using this approach.
+
+Formal multivariate versions of GLM or GAMs are currently an important area research within ecology (see [@Warton2015-ge] for a recent review), where they go by the name joint species distribution models (JSDMs).
+Whilst undoubtedly powerful, our knowledge regarding JSDMs and their availability in software are still in their relative infancy and they require considerable expertise to implement.
+As such, JSDMs are currently beyond the reach of most palaeoecologists.
+Despite this, we should be watching JSDM research as developments are ongoing and a degree of method maturation is occurring.
+
+One currently available avenue for fitting a multivariate GAM is via regularized sandwich estimators with GLMs [@Warton2011-vp], which involves fitting separate GLMs (or GAMs) to each response variable and subsequently using resampling-based hypothesis tests to determine which covariates are related to variation at the community level and for individual taxa [@Warton2011-vp @Warton2012-fg @Wang2012-uu].
+The *mvabund* package [@Wang2012-uu] implements this approach within R.
+
+A pragmatic although inelegant approach that has been used to estimate trends in multivariate palaeoecological data is to first summarise the response data using an unconstrained ordination via a PCA, CA, or principal curve and then fit separate GAM models to the site (sample) scores of the first few ordination axes or principal curve.
+Whilst this two-step approach is relatively easy to implement and builds on existing approaches palaeoecologists use to summarise multivariate stratigraphic data, it is best though of as modelling changes in abundance or relative composition at the community level.
+It is less well suited to unpicking taxon-specific trends however, because the ordination step combines individual species information into latent variables (axes) that are linear combinations of *all* species and it is these latent variables that are modelled using GAM.
+
+## Conclusions
 

### 2018-05-11 manuscript.Rmd
final draft

@@
 ---
-title: "Modelling palaeoecological time series using generalised additive models"
+title: "Modelling palaeoecological time series using generalized additive models"
 author: Gavin L. Simpson
-#author:
-#- name: "Gavin L Simpson"
-#  affiliation: Insitute of Environmental Change and Society, University of Regina, Regina, SK
-date: April 26, 2018
+affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
+mail: "gavin.simpson@uregina.ca"
+date: May 11, 2018
+keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
 bibliography: references.bib
+csl: frontiers-in-ecology-and-evolution.csl
 geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
 fontsize: 12pt
-fontfamily: stix
-mainfont: XITS
+mainfont: TeX Gyre Pagella
 output:
   pdf_document:
+    number_sections: true
     fig_width: 8
@@
       in_header: header.tex
-#    pandoc_args: [
-#      "--no-tex-ligatures"
-#    ]
-#    template: frontiers-template.tex
-# documentclass: frontiersSCNS # not yet ready for this; need to pandoc-ify the template
 ---
@@
 library("grid")                         # for unit.pmax(), unit.list()
-## library("sp")                           # for coordinates()
 library("schoenberg")
 library("tidyr")
-```
-
-```{r common-components, cache = FALSE}
+
 ## Default ggplot theme
 theme_set(theme_bw())
-
+```
+
+```{r common-components, cache = FALSE}
 ## source Small Water data
@@
 
-## Abstract
-
-350 Words
-
-## Introduction
-
-* intro about palaeo time series
-
-### Example time series
-
-To illustrate the estimation of trends in palaeoenvironmental data using GAMs, I use two proxy time series; a 150-year bulk organic matter δ^15^N time series from Small Water, and a 3000-year alkenone record from Braya-sø.
+# Abstract {.unnumbered}
+
+In the absence of annual laminations, time series generated from lake sediments or other similar stratigraphic sequences are irregularly spaced in time, which complicates formal analysis using classical statistical time series models.
+In lieu, statistical analyses of trends in palaeoenvironmental time series, if done at all, have typically used simpler linear regressions or (non-) parametric correlations with little regard for the violation of assumptions that almost surely occurs due to temporal dependencies in the data or that correlations do not provide estimates of the magnitude of change, just whether or not there is a linear or monotonic trend.
+Alternative approaches have used \loess{}-estimated trends to justify data interpretations or test hypotheses as to the causal factors without considering the inherent subjectivity of the choice of parameters used to achieve the \loess{} fit (e.g.\ span width, degree of polynomial).
+Generalized additive models (GAMs) are statistical models that can be used to estimate trends as smooth functions of time.
+Unlike \loess{}, GAMs use automatic smoothness selection methods to objectively determine the complexity of the fitted trend, and as formal statistical models, GAMs, allow for potentially complex, non-linear trends, a proper accounting of model uncertainty, and the identification of periods of significant temporal change.
+Here, I present a consistent and modern approach to the estimation of trends in palaeoenvironmental time series using GAMs, illustrating features of the methodology with two example time series of contrasting complexity; a 150-year bulk organic matter δ^15^N time series from Small Water, UK, and a 3000-year alkenone record from Braya-Sø, Greenland.
+I discuss the underlying mechanics of GAMs that allow them to learn the shape of the trend from the data themselves and how simultaneous confidence intervals and the first derivatives of the trend are used to properly account for model uncertainty and identify periods of change.
+It is hoped that by using GAMs greater attention is paid to the statistical estimation of trends in palaeoenvironmental time series leading to more a robust and reproducible palaeoscience.
+
+# Introduction
+
+The vast majority of data produced by palaeoecologists and palaeolimnologists is in the form of time-ordered observations on one or more proxies or biological taxa [@Smol2008-ld; @Smol2012-xt; @Birks2012-fa].
+Typically these data are arranged irregularly in time; in the absence of annual laminae or varves, the sediment core is sectioned at regular depth intervals [@Glew2001-co], which, owing to variation in accumulation rates over time and compaction by overlying sediments, results in an uneven sampling in time.
+An under-appreciated secondary feature of such sampling is that younger sediments often have larger variance than older sediments; each section of core represents fewer lake years in newer samples, relative to older samples.
+This variable time averaging acts as a variable low-pass (high-cut) filter of the raw, annual depositional signal.
+
+The irregular intervals between samples means that the customary time-series analysis methods of autoregressive or moving average processes, in the form of autoregressive integrated moving average (ARIMA) models, are practically impossible to apply because software implementing these methods typically requires even spacing of observations in time.
+@Dutilleul2012-su and @Birks2012-op, eschewing the term *time series*, prefer to call such data *temporal series* on account of the irregular spacing of samples, a distinction that I find unnecessary.
+
+Where statistical approaches have been applied to trend estimation in palaeoenvironmental time series, the most commonly used has been \loess{} [@Cleveland1979-vf; @Juggins2012-uu; @Birks2012-op; @Birks1998-po].
+\loess{}, or locally weighted scatterplot smoother, as it's name suggests, was developed to smooth x-y scatterplot data, ostensibly as part of exploratory data analysis [@Cleveland1979-vf].
+The method fits a smooth line through data by fitting weighted least squares (WLS) models to observations within a particular, user-specified window of the target point, whose width is typically expressed as a proportion $\alpha$ of the $n$ data points.
+Weights are determined by how close (in the x-axis only) an observation in the window is to the focal point, with a tricubic function, by default, giving greatest weight to points closest to the focal point.
+The interim \loess{}-smoothed value for the focal point is the predicted value from the weighted regression at the focal point.
+The interim values are updated using weights based on how far in the y-axis direction, the interim smoothed value lies from the observed value plus the x-axis distance weights; this has the effect of down-weighting the effects of outlier observations.
+The final \loess{} is obtained by joining up the smoothed (fitted) values.
+The user has to choose how large a window to use, whether to fit degree 1 (linear) or degree 2 (quadratic) polynomials in the WLS model, and how to weight points in the x-axis.
+When used in an exploratory mode, the user has considerable freedom to choose the detail of the \loess{} fit; the window width, for example, can be infinitely tweaked to give as close a fit to the data, as assessed by eye, as is desired.
+Using cross-validation (CV) to choose $\alpha$ or the degree of polynomial in the WLS model is complicated for a number of reasons, not least that the CV scheme used must involve the time ordering of the data [e.g.
+@Bergmeir2018-pu].
+This level of subjectivity is problematic however once we wish to move beyond exploratory data analysis to statistically identify trends and to test hypotheses involving those trend estimates.
+
+Running means or other types of filter [@Juggins2012-uu] have also been used extensively to smooth palaeoenvironmental time series, but, as with \loess{}, their behaviour depends on a number of factors, including the filter width.
+Furthermore, the width of the filter causes boundary issues; with a centred filter, of width five data points, the filtered time series would be two data points shorter at both ends of the series because the filter weights are not defined for the first and last two observations of the original series as these time points were not observed.
+Considerable research effort has been expended to identify ways to pad the original time series at one or both ends to maintain the original length in the filtered series, without introducing bias because of the method of padding [e.g.
+@Mann2004-kd; @Mann2008-yk; @Mills2006-rb; @Mills2007-pq; @Mills2010-oe].
+
+These are by no means the only methods that have been used to estimated trends in stratigraphic series.
+Another common approach involves fitting a simple linear trend using ordinary least squares regression and use the resulting *t* statistic as evidence against the null hypothesis of no trend despite the statistical assumptions of the method being almost surely violated due to dependence among observations arising from temporal autocorrelation in the series.
+The Pearson correlation coefficient, *r*, is also often used to detect trends in palaeo time series [@Birks2012-op], despite the fact that *r* provides no information as to the magnitude of the estimated trend, and the same temporal autocorrelation problem that dogs ordinary least squares similarly plagues significance testing for *r* [@Tian2011-ip].
+Additionally, both the simple least squares trend line and *r* are tests for *linear* trends only, and yet we typically face data sets with potentially far more complex trends than can be identified by these methods.
+Instead, non-parametric rank correlation coefficients have been used [@Birks2012-op; @Gautheir2001-ee], and whilst these do allow for the detection of non-linear trends, trends are restricted to be monotonic, no magnitude of the trend is provided, and the theory underlying significance testing of Spearman's $\rho$ and Kendall's $\tau$ assumes independent observations.
+
+Palaeoecology and palaeolimnology have moved away from being descriptive disciplines, rapidly adopting new statistical developments in the 1990s and beyond [@@Smol2012-xt].
+Less development has been observed in the area of trend estimation in palaeoenvironmental time series.
+Here, I describe the use of generalized additive models [GAMs; @Hastie1986-xs; @Hastie1990-bx; @Ruppert2003-pt; @Wood2017-qi; @Yee1991-mg] for trend estimation.
+GAMs, like simple linear regression, are a formal regression-based method for estimating trends, yet they are also, superficially at least, similar to \loess{}.
+GAMs and \loess{} estimate smooth, non-linear trends in time series and both can handle the irregular spacing of samples in time, yet GAMs do not suffer from the subjectivity that plagues \loess{} as a method of formal statistical inference.
+
+In the subsequent sections, I present an introduction to GAMs and how they are able to estimate smooth, non-linear trends from time series.
+Next the issue of uncertainty in model-estimated trends is discussed, as is the topic of posterior simulation from a regression model.
+The issue of identifying periods of significant environmental change is considered and I show how this can be achieved using the first derivative of the estimated trend.
+Two non-standard types of spline --- adaptive smoothers and Gaussian process splines --- that are especially applicable to GAMs in the palaeoenvironmental setting are subsequently described, followed by an assessment of the the impact of age-model uncertainty on trend estimation via GAMs.
+Finally, I briefly discuss the application of GAM trend analysis to multivariate species abundance and compositional data.
+
+## Example time series
+
+To illustrate the estimation of trends in palaeoenvironmental data using GAMs, I use two proxy time series; a 150-year bulk organic matter δ^15^N time series from Small Water, and a 3000-year alkenone record from Braya-Sø.
 Between them, the two examples, combine many of the features of interest to palaeoecologists that motivate the use of GAMs; non-linear trends and the question of when changes in the measured proxy occurred.
@@
 
-#### δ^15^N time series from Small Water
+### δ^15^N time series from Small Water
 
 Figure \ref{fig:data-figure}a shows `r nrow(small)` nitrogen stable isotope measurements on the bulk organic matter of a sediment core collected from Small Water, a small corrie lake located in the English Lake District, UK.
-The data were collected to investigate disturbance of nitrogen (N) cycling in remote, oligotrophic lakes by N deposited from the atmosphere (Simpson, unpublished data, REF Curtis et al FUMBLE report).
+The data were collected to investigate disturbance of nitrogen (N) cycling in remote, oligotrophic lakes by N deposited from the atmosphere (Simpson, unpublished data).
 The data are shown on a ^210^Pb time scale.
@@
 
-#### Braya-sø alkenone time series
-
-The second example time series is a 6,000 year record of alkenone unsaturation, \uk, from Braya-sø, a meromictic lake in  West Greenland [@DAndrea2011-oa].
+### Braya-Sø alkenone time series
+
+The second example time series is a 3,000 year record of alkenone unsaturation, \uk, from Braya-Sø, a meromictic lake in  West Greenland [@DAndrea2011-oa].
 Alkenones are long-chained unsaturated organic compounds that are produced by a small number of planktonic organisms known as haptophytes.
-The \uk{} unsaturation index is
+The \uk{} unsaturation index is [@Brassell1993-fw]
 \begin{equation*}
@@
 where $[C_{37:x}]$ is the concentration of the alkenone with 37 carbon atoms and $x$ double carbon bonds.
-The relative abundance of these alkenones is known to vary with changes in water temperature, and as a result \uk{} is used as a proxy for lake- and sea-surface temperatures.
-For further details on the Braya-sø \uk{} record and age model see @DAndrea2011-oa.
+The relative abundance of these alkenones is known to vary with changes in water temperature [@Brassell1993-fw; @Zink2001-cr; @Chu2005-es; @Toney2010-ba], and as a result \uk{} is used as a proxy for lake- and sea-surface temperatures.
+For further details on the Braya-Sø \uk{} record and age model see @DAndrea2011-oa.
 Here I use the 3,000 year \uk{} record from the PAGES 2K database [@PAGES_2k_Consortium2013-fm].
@@
 
-```{r data-figure, fig = TRUE, fig.width = 8, fig.height = 5, fig.cap = "data plot"}
+```{r data-figure, fig = TRUE, fig.width = 8, fig.height = 5, fig.cap = "Example time series; a) Small Water bulk organic matter $\\delta^{15}\\text{N}$ time series on a ${}^{210}\\text{Pb}$ time scale, and b) Braya-Sø \\uk{} time series on a calibrated ${}^{14}\\text{C}$ time scale.
+The observations \\uk{} time series have been joined by lines purely as a visual aid to highlight potential trends."}
 ## Generate a plot of the data - used as a base later
@@
     geom_point() +
-    labs(y = braya_ylabel, x = "Years BP")
+    labs(y = braya_ylabel, x = "Year CE")
 
@@
 
-In the subsequent sections, I present regression models that can be used to estimate trends in palaeoenvironmental time series and give an introduction to GAMs and how they estimate smooth, non-linear trends from time series.
-Next the issue of uncertainty in model-estimated trends is discussed, as is the topic of posterior simulation from a regression model.
-The issue of identifying periods of significant environmental change is considered and I show how this can be achieved using the first derivative of the estimated trend.
-Two non-standard types of spline --- adaptive smoothers and Gaussian process splines --- that are especially applicable to GAMs in the palaeoenvironmental setting are described, followed by an assessment of the the impact of age-model uncertainty on trends estimation via GAMs.
-Finally, I briefly discuss the application of GAM trend analysis to multivariate species abundance and compositional data.
-
-## Regression models for palaeoenvironmental time series
+# Regression models for palaeoenvironmental time series
 
@@
 
-Linear models using a range of polynomials for the Small Water data set are shown in Figure \ref{fig:polynomial-example-plot}.
+Linear models using a range of polynomials fitted to the Small Water data set are shown in Figure \ref{fig:polynomial-example-plot}.
 The low-order models ($P \in \{1, 3\}$) result in very poor fit to the data.
-The model with $P = 5$ does a reasonable job of capturing the gross pattern in the data, but it fails to adapt quickly enough to the decrease in δ^15^N that begins ~1940 CE, and the resulting trend is quite biased as a result.
+The model with $P = 5$ does a reasonable job of capturing the gross pattern in the time series, but fails to adapt quickly enough to the decrease in δ^15^N that begins ~1940 CE, and the estimated trend is quite biased as a result.
 The $P = 10$th-order polynomial model is well able to capture this period of rapid change, but it does so at the expense of increased complexity in the estimated trend prior to ~1940.
-Additionally, this model ($P = 10$) has undesirable behaviour at the ends of the series, significantly overfitting the data, a commonly observed problem in polynomial models such as these.
+Additionally, this model ($P = 10$) has undesirable behaviour at the ends of the series, significantly overfitting the data, a commonly observed problem in polynomial models such as these [@Epperson1987-ui; @Runge1901-tc].
 Finally, the choice of what order of polynomial to fit is an additional choice left for the analyst to specify; choosing the optimal $P$ is not a trivial task when the data are a time series and residual autocorrelation is likely present.
 
-```{r polynomial-example-plot, fig.width = 8, fig.height = 2.5, fig.cap = "Linear models with various orders of polynomial of the covariate Year as covariates.", dependson = c(-1)}
+```{r polynomial-example-plot, fig.width = 8, fig.height = 2.5, fig.cap = "Linear models with various orders of polynomial of the covariate Year fitted using ordinary least squares to the $\\delta^{15}\\text{N}$ time series from Small Water.
+The degree of polynomial is indicated, with the degree 1 line equal to a simple linear regression model.", dependson = c(-1)}
 small_plt + geom_line(data = polyDat, mapping = aes(x = Year, y = Fitted, colour = Degree)) +
@@
 Can we do better than these polynomial fits?
-In the remainder , I hope to demonstrate that the answer to that question is emphatically "yes"!
-Below I describe a coherent and consistent approach to modelling palaeoenvironmental time series using generalised additive models (GAMs), which builds upon the linear regression framework.
-
-## Generalised additive models
-
-GAMs extend the generalised linear model (GLM) --- of which the linear model described above is a special case --- where some or all of the parametric terms, the $\beta_p$, are replace by smooth functions $f_j$ of the covariates $x_j$.
+In the remainder, I hope to demonstrate that the answer to that question is emphatically "yes"!
+Below I describe a coherent and consistent approach to modelling palaeoenvironmental time series using generalized additive models, which builds upon the linear regression framework.
+
+# Generalized additive models
+
 The GAM version of the linear model \eqref{eq:linear-model} is
-
 \begin{equation} \label{eq:additive-model}
-y_t = \beta_0 + f(x_t) + \varepsilon_t
+y_t = \beta_0 + f(x_t) + \varepsilon_t \, ,
 \end{equation}
-
-The immediate advantage of the GAM is that we are no longer restricted to the types of trends that can be fitted via global polynomial functions such as \eqref{eq:polynomial-model}.
-Instead, the form of the fitted trend is estimated from the data itself.
-
-For completeness, where the data are non-Gaussian, such as for where $y_t$ represents a time series of counts of a particular species, we can write \eqref{eq:additive-model} as a GLM/GAM
-
-\begin{align} \label{eq:gam}
-y_t &\sim \text{EF}(\mu_t, \Theta) \\
-g(\mu_t) &= \beta_0 + f(x_t) \\
-\mu_t    &= g^{-1}(\beta_0 + f(x_t)),
+where the linear effect of time(the $\beta_1 x_t$ part) has been replaced by a smooth function of time, $f(x_t)$.
+The immediate advantage of the GAM is that we are no longer restricted to the shapes of trends that can be fitted via global polynomial functions such as \eqref{eq:polynomial-model}.
+Instead, the shape of the fitted trend will be estimated from the data itself.
+
+The linear model is a special case of a broader class know as the generalized linear model [GLM; @McCullagh1989-gv].
+The GLM provides a common framework for modelling a wide range of types of data, such as count, proportions, or binary (presence/absence) data, that are not conditionally distributed Gaussian.
+GLMs are, like the linear model, parametric in nature; the types of trends that we can fit using a GLM are the linear or polynomial models.
+GAMs extend the GLM by relaxing this parametric assumption; in a GAM some, or all, of the parametric terms, the $\beta_p$, are replace by smooth functions $f_j$ of the covariates $x_j$.
+For completeness then, we can write \eqref{eq:additive-model} as a GLM/GAM
+\begin{subequations}
+\label{eq:gam}
+\begin{align}
+y_t &\sim \text{EF}(\mu_t, \Theta) \label{eq:gam-ef} \\
+g(\mu_t) &= \beta_0 + f(x_t) \label{eq:gam-linear-pred} \\
+\mu_t    &= g^{-1}(\beta_0 + f(x_t)), \label{eq:gam-inverse} 
 \end{align}
-
+\end{subequations}
 where $\mu_t$ is the expected value (e.g.\ the mean count or the probability of occurrence) of the random variable $Y_t$ ($\mu_t \equiv \mathbb{E}(Y_t)$) of which we have observations $y_t$.
 $g$ is the link function, an invertible, monotonic function, such as the natural logarithm, and $g^{-1}$ is its inverse.
-The link function maps values from the linear predictor scale on to the scale of the response.
+The link function maps values from the response scale on to the scale of the linear predictor, whilst the inverse of thew link function provides the reverse mapping.
 For example, count data are strictly non-negative integer values and are commonly modelled as a Poisson GLM/GAM using the natural log link function.
-On the log scale, the response can take any real value between $-\infty$ and $+\infty$, and it is on this scale that model fitting actually occurs.
-However we need to map these unbounded values back on to the non-negative response scale the.
-The inverse of the log link function, the exponential function, achieves this and maps values to the interval 0--$\infty$.
-
-In \eqref{eq:gam}, we further assume that the observations are drawn from a member of the exponential family of distributions --- such as the Poisson for count data, the binomial for presence/absence or counts from a total --- with expected value $\mu_t$ and possibly some additional parameters $\Theta$ ($y_t \sim \text{EF}(\mu_t, \Theta)$).
-Additionally, many software implementations of the above model also allow for distributions that are not within the exponential family but which can be fitted using an algorithm superficially similar to the one used to fit GAMs to members of the exponential family.
+On the log scale, the response can take any real value between $-\infty$ and $+\infty$, and it is on this scale that model fitting actually occurs (i.e.
+using equation \eqref{eq:gam-linear-pred}).
+However we need to map these unbounded values back on to the non-negative response scale.
+The inverse of the log link function, the exponential function, achieves this and maps values to the interval 0--$\infty$ (equation \eqref{eq:gam-inverse}).
+
+In \eqref{eq:gam-ef}, we further assume that the observations are drawn from a member of the exponential family of distributions --- such as the Poisson for count data, the binomial for presence/absence or counts from a total --- with expected value $\mu_t$ and possibly some additional parameters $\Theta$ ($y_t \sim \text{EF}(\mu_t, \Theta)$).
+Additionally, many software implementations of the above model also allow for distributions that are not within the exponential family but which can be fitted using an algorithm superficially similar to the one used to fit GAMs to members of the exponential family [e.g.
+@Wood2016-fx].
 Common examples of such extended families include the negative binomial distribution (for overdispersed counts) and the beta distribution (for true proportions or other interval-bounded data).
 
-### Basis functions
-
-We want the fitted trends for the Small Water δ^15^N and Braya-sø \uk{} time series to be wiggly functions, whose shape is informed by the data themselves.
-These wiggly functions are smooths, which we denote by $f(x_t)$.
-We would like to represent the smooths in a way that \eqref{eq:gam} is a linear model so that we can estimate the model within the well-studied GLM framework, which is achieved by using a *basis*.
-A basis is a set of functions that collectively span a space of functions that, we hope, contains the true $f(x_t)$ or a close approximation to it.
+## Basis functions
+
+It is clear from the plots of the data that we require the fitted trends for the Small Water δ^15^N and Braya-Sø \uk{} time series to be non-linear functions, but it is less clear how to specify the actual shape require.
+Ideally, we'd like to learn the shape of the trend from the data themselves.
+We will refer to these non-linear functions as *smooth functions*, or just *smooths* for short, and we will denote a smooth using $f(x_t)$.
+Further, we would like to represent the smooths in a way that \eqref{eq:gam} is represented parametrically so that it can be estimate within the well-studied GLM framework.
+This is achieved by representing the smooth using a *basis*.
+A basis is a set of functions that collectively span a space of smooths that, we hope, contains the true $f(x_t)$ or a close approximation to it.
 The functions in the basis are known as *basis functions*, and arise from a *basis expansion* of a covariate.
 Writing $b_j(x_t)$ as the $j$th basis function of $x_t$, the smooth $f(x_t)$ can be represented as a weighted sum of basis functions
-
-$$
+\begin{equation*}
 f(x_t) = \sum_{j = 1}^{k} b_j(x_t) \beta_j \,,
-$$
-
+\end{equation*}
 where $\beta_j$ is the weight applied to the $j$th basis function. Note that here we concern ourselves only with univariate smooths involving a single covariate, the time variable $x_t$.
@@
 For the cubic polynomial ($P = 3$) fit shown in Figure \ref{fig:polynomial-example-plot} there are in fact 4 basis functions: $b_1(x_t) = x_t^0 = 1$, $b_2(x_t) = x_t$, $b_3(x_t) = x^2_t$, and $b_4(x_t) = x_t^3$.
-Note that here $b_1(x_t)$ is constant and is in fact the model intercept, $\beta_0$, in the linear model \eqref{eq:polynomial-model}, and the weights are the estimated coefficients in the model, the $\beta_j$.
+Note that $b_1(x_t)$ is constant and is linked to the model intercept, $\beta_0$, in the linear model \eqref{eq:polynomial-model}, and further, that the weights are the estimated coefficients in the model, the $\beta_j$.
 
 As we have already seen, polynomial basis expansions do not necessarily lead to well-fitting models unless the true function $f$ is itself a polynomial.
-One of the primary criticisms is that polynomial basis functions are global; the value of $f$ at time point $x_t$ affects the value of $f$ at time point $x_{t+s}$ even if the two time points are at opposite ends of the series.
+One of the primary criticisms is that polynomial basis functions are global [@Magee1998-oq]; the value of $f$ at time point $x_t$ affects the value of $f$ at time point $x_{t+s}$ even if the two time points are at opposite ends of the series.
 There are many other bases we could use; here I discuss one such set of bases, that of splines.
@@
 In addition, I also discuss two special types of spline basis, an adaptive spline basis and a Gaussian process spline basis.
-I begin with the cubic regression spline basis.
 
 A cubic spline is a smooth curve comprised of sections of cubic polynomials ($P = 3$), where the sections are joined together at some specified locations --- known as *knots* --- in such a way that at the joins, the two sections of cubic polynomial that meet have the same value as well as the same first and second derivative.
-These properties mean that the sections join smoothly and differentiably at the knots.
-
-```{r basis-function-example-plot, fig.width = 8, fig.height = 5, fig.cap = "basis functions plot"}
+These properties mean that the sections join smoothly and differentiably at the knots [@Wood2017-qi, sec.
+5.3.1].
+
+```{r basis-function-example-plot, fig.width = 8, fig.height = 5, fig.cap = "Basis functions for the time covariate and the Small Water $\\delta^{15}\\text{N}$ time series.
+A rank (size) 7 cubic regression spline (CRS) basis expansion is show in a), with knots, indicated by tick marks on the x-axis, spread evenly through the rang of the data.
+b) shows the same CRS basis functions weighted by the estimated coefficients $\\beta_j$, plus the resulting GAM trend line (black line drawn through the data).
+The grey points in both panels are the observed $\\delta^{15}\\text{N}$ values.
+c) A rank 7 thin plate regression spline basis for the same data."}
 ## set up
@@
 Thin plate splines were introduced by @Duchon1977-jr and, as well as solving the knot selection problem, have several additional attractive properties in terms of optimality and their ability to estimate a smooth function of two or more variables, leading to smooth interactions between covariates.
-However, thin plate splines have one key disadvantage over CRS; thin plate splines have as many unknown parameters as there are unique combinations of covariate values in a data set.
+However, thin plate splines have one key disadvantage over CRS; thin plate splines have as many unknown parameters as there are unique combinations of covariate values in a data set [@Wood2017-qi, sec.
+5.5.1].
 It is unlikely in the extreme that any real data problem would involve functions of such complexity that they require as many basis functions as data.
@@
 To address this issue, thin plate regression splines (TPRS) have been suggested which truncate the space of the thin plate spline basis to some lower number of basis functions whilst preserving much of the advantage of the original basis as an optimally-fitting spline [@Wood2003-qy].
-A rank `r k` TPRS basis is shown in Figure \ref{fig:basis-function-example-plot}c for the Small Water time series.
+A rank `r k` TPRS basis (i.e.
+one containing `r k` basis functions) is shown in Figure \ref{fig:basis-function-example-plot}c for the Small Water time series.
 The truncation is achieved by performing an eigen-decomposition of the basis functions and retaining the eigenvectors associated with the $k$ largest eigenvalues.
 This is similar to the way principal components analysis decomposes a data set into axes of variation (eigenvectors) in decreasing order of variance explained.
-By concentrating the "wiggliness" information in those eigenvectors with large eigenvalues, the truncated basis can preserve much of the space of functions spanned by the original basis but at the cost of using far fewer basis functions.
-
-The truncation suggested by @Wood2003-qy is not without cost; the eigen-decomposition and related steps can be relatively costly to set up for large data sets.
+The truncated basis can preserve much of the space of functions spanned by the original basis but at the cost of using far fewer basis functions [@Wood2003-qy; @Wood2017-qi sec.
+5.5.1].
+Note the horizontal TPRS basis function (at δ^15^N = 1) in Figure \ref{fig:basis-function-example-plot}c; this basis function is confounded with the intercept term and, after the application of identifiability constraints, ends up being removed from the set of basis functions used to fit the model.
+
+The truncation suggested by @Wood2003-qy is not without cost; the eigen-decomposition and related steps can be relatively costly for large data sets.
 For data sets of similar size to the two examples used here, the additional computational effort required to set up the TPRS basis over the CRS basis will not be noticeable.
-For highly resolved series containing more than 500--1000 observations the truncation may be too costly computationally.
-In such instances, little is lost by moving to the CRS basis, with the same number of knots as the rank of the desired thin plate spline, either distributed evenly throughout the data or at the quantiles of the data, with the benefit of considerably reduced set up time for the basis.
+For highly resolved series containing more than ~1000 observations the truncation may be too costly computationally.
+In such instances, little is lost by moving to the CRS basis, with the same number of knots as the rank of the desired TPRS, with the benefit of considerably reduced set up time for the basis.
 
@@
 In practice, the size of the basis is an upper limit on the expected complexity of the trend, and a simple test is available to check if the basis used was sufficiently large [@Pya2016-rk].
-This test is available via the `gam.check()` function in *mgcv* for example, and essentially looks at whether there is any additional nonlinearity or structure in the residuals that can be explained by $x_t$.
-Should a smooth term in the fitted model fail this test the model can be refitted using a larger basis expansion, say by doubling the value of `k` (the rank) used to fit the original.
+This test is available via the `gam.check()` function in *mgcv* for example, which essentially looks at whether there is any additional nonlinearity or structure in the residuals that can be explained by a further smooth of $x_t$.
+Should a smooth term in the fitted model fail this test the model can be refitted using a larger basis expansion, say by doubling the value of *k* (the rank) used to fit the original.
 Note also that a smooth might fail this test whilst using fewer effective degrees of freedom than the maximum possible for the dimension of basis used.
@@
 
-### Smoothness selection
+## Smoothness selection
 
 Having identified regression splines as a useful way to represent $f$, we next need a way to decide how wiggly the fitted trend should be.
-A backwards elimination approach to sequentially remove knots or basis functions might seem appropriate, however such an approach would likely fail as the resulting sequence of models would not be strictly nested, precluding many forms of statistical comparison.
+A backwards elimination approach to sequentially remove knots or basis functions might seem appropriate, however such an approach would likely fail as the resulting sequence of models would not be strictly nested, precluding many forms of statistical comparison [@Wood2017-qi].
 Alternatively, we could keep the basis dimension at a fixed size but guard against fitting very complex models through the use of a wiggliness penalty.
 
-The default wiggliness penalty used in GAMs is on the second derivative of the spline, which measures the rate of change of the slope, the curvature, of the spline at any infinitesimal point in the interval spanned by $x_t$.
+The default wiggliness penalty used in GAMs is on the second derivative of the spline, which measures the rate of change of the slope, or the curvature, of the spline at any infinitesimal point in the interval spanned by $x_t$.
 The actual penalty used is the integrated squared second derivative of the spline
-
-$$\int_{\mathbb{R}} [f^{\prime\prime}]^2 dx$$
-
-which can be written in quadratic form as a function of the estimated coefficients of $f(x_t)$
-
-$$\int_{\mathbb{R}} [f^{\prime\prime}]^2 dx = \boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}$$
-
-where $\mathbf{S}$ is known as the penalty matrix.
+\begin{equation} \label{eq:quadratic-penalty}
+\int_{\mathbb{R}} [f^{\prime\prime}]^2 dx = \boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}\, .
+\end{equation}
+The right hand side of \eqref{eq:quadratic-penalty} is the penalty in quadratic form.
+The convenience of the quadratic form is that it is a function of the estimated coefficients of $f(x_t)$ where $\mathbf{S}$ is known as the penalty matrix.
+Notice that now both the weights for the basis functions and the wiggliness penalty are expressed as functions of the model coefficients.
 
 Now that we have a convenient way to measure wiggliness, it needs to be incorporated into the objective function that will be minimised to fit the GAM.
-The likelihood of the model given the parameter estimates $\mathcal{L}(\boldsymbol{\beta})$ is combined with the penalty to create the penalized likelihood $\mathcal{L}_p(\boldsymbol{\beta})$
-
-$$\mathcal{L}_p(\boldsymbol{\beta}) = \mathcal{L}(\boldsymbol{\beta}) - \frac{1}{2} \lambda\boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}\, .
-$$
-
+The likelihood of the model given the parameter estimates $\mathcal{L}(\boldsymbol{\beta})$ is combined with the penalty to create the penalized likelihood $\mathcal{L}_p(\boldsymbol{\beta})$:
+\begin{equation*}
+\mathcal{L}_p(\boldsymbol{\beta}) = \mathcal{L}(\boldsymbol{\beta}) - \frac{1}{2} \lambda\boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}\, .
+\end{equation*}
 The fraction of a half is there simply to make the penalised likelihood equal the penalised sum of squares in the case of a Gaussian model.
-$\lambda$, the smoothess parameter, controls the extent to which the penalty contributes to the likelihood of the model.
-In the extreme case, if $\lambda = 0$ then the penalty has no effect and the penalized likelihood equals the likelihood of the model given the parameters.
-As $\lambda \rightarrow \infty$ the penalty comes to dominate $\mathcal{L}_p(\boldsymbol{\beta})$ and the wiggliness of $f(x_t)$ tends to $0$ resulting in an infinitely smooth function, which, in the case of a second derivative penalty, is a straight line.
-If a penalty was on the integrated squared *first* derivative of $f(x_t)$, the fully penalised spline would be a flat, horizontal line.
+$\lambda$ is known as the smoothness parameter and controls the extent to which the penalty contributes to the likelihood of the model.
+In the extreme case of $\lambda = 0$ the penalty has no effect and the penalized likelihood equals the likelihood of the model given the parameters.
+At the other extreme, as $\lambda \rightarrow \infty$ the penalty comes to dominate $\mathcal{L}_p(\boldsymbol{\beta})$ and the wiggliness of $f(x_t)$ tends to $0$ resulting in an infinitely smooth function.
+In the case of a second derivative penalty, this is a straight line, and we recover the simple linear trend from \eqref{eq:linear-model} when assuming a Gaussian response.
 
@@
 
-```{r penalty-example-plot, fig.width = 8, fig.height = 5, fig.cap = "the effect of different smoothness", dependson = -1}
+```{r penalty-example-plot, fig.width = 8, fig.height = 5, fig.cap = "The effect of the smoothness parameter $\\lambda$ on the resulting wiggliness of the estimated spline.
+Large values of $\\lambda$ penalize wiggliness strongly, resulting in smooth trends (upper row), while smaller values allow increasingly wiggly trends.
+The aim of automatic smoothness selection is to find an optimal value of $\\lambda$ that balances the fit of the model with model complexity to avoid overfitting.", dependson = -1}
 op <- options(scipen = 100)
@@
 Figure \ref{fig:penalty-example-plot} illustrates how the smoothness parameter $\lambda$ controls the degree of wiggliness in the fitted spline.
-Four models are shown each fitted with a fixed value of $\lambda$; 10000, 1, 0.01, and 0.00001.
+Four models are shown, each fitted with a fixed value of $\lambda$; 10000, 1, 0.01, and 0.00001.
 At $\lambda = 10000$ the model effectively fits a linear model through the data.
@@
 The first way is to choose $\lambda$ such that it minimises the prediction error of the model.
-This can be achieved by choosing $\lambda$ to minimise Akaike's information criterion (AIC) or via cross-validation (CV) or generalized cross-validation (GCV).
+This can be achieved by choosing $\lambda$ to minimise Akaike's information criterion (AIC) or via cross-validation (CV) or generalized cross-validation [GCV; @Craven1978-nz].
 GCV avoids the computational overhead inherent to CV of having to repeatedly refit the model with one or more observations left out as a test set.
 Minimising the GCV score will, with a sufficiently large data set, find a model with the minimal prediction error [@Wood2017-qi].
-The second approach is to treat the smooth as a random effect, in which $\lambda$ is now a variance parameter to be estimated using maximum likelihood (ML) or restricted maximum likelihood (REML) [@Wood2011-kn; @Wood2016-fx].
+The second approach is to treat the smooth as a random effect, in which $\lambda$ is now a variance parameter to be estimated using maximum likelihood (ML) or restricted maximum likelihood [REML; @Wood2011-kn; @Wood2016-fx].
 
@@
 
-## Fitting GAMs (Need a better title)
-
-### Small Water
+# Fitting GAMs
+
+## Small Water
 
@@
 Therefore I fitted the following GAM to the δ^15^N time series.
-
 \begin{equation} \label{eq:small-gam}
@@
 \end{equation}
-
 Note that now I have relaxed the i.i.d.
 assumption and have introduced $\boldsymbol{\Lambda}$, a correlation matrix that is used to model autocorrelation in the residuals.
-The δ^15^N values are irregularly spaced in time and as a result a correlation structure that can handle the uneven spacing of the samples is needed.
-A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the first-order autoregressive process (AR(1)) and, simply stated, models the correlation between any two residuals as and exponentially decreasing function of $h$ ($\phi^h$), where $h$ is the amount of separation in time between the residuals.
+The δ^15^N values are irregularly spaced in time and a correlation structure that can handle the uneven spacing of the samples is needed [@Pinheiro2000-kx].
+A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the first-order autoregressive process (AR(1)) and, simply stated, models the correlation between any two residuals as and exponentially decreasing function of $h$ ($\phi^h$), where $h$ is the amount of separation in time between the residuals [@Pinheiro2000-kx].
 $h$ may be a real valued number in the CAR(1), which is how it can accommodate the irregular separation of samples in time.
@@
 
-```{r plot-fitted-models, fig.width = 8, fig.height = 5, fig.cap = "Fitted GAMs"}
+```{r plot-fitted-models, fig.width = 8, fig.height = 5, fig.cap = "GAM-based trends fitted to the Small Water $\\delta^{15}\\text{N}$ (a) and Braya-Sø \\uk{} (b) time series.
+The shaded bands surrounding the estimated trends are approximate 95\\% across-the-function confidence intervals.
+For the \\uk{} series, two models are shown; the orange fit is the result of a GAM with a continuous-time AR(1) process estimated using REML smoothness selection, while the blue fit is that of a simple GAM with GCV-based smoothness selection.
+The REML-based fit significantly oversmooths the \\uk{} time series."}
 N <- 300                                # number of points at which to evaluate the splines
@@
 The fitted trend is shown in Figure \ref{fig:plot-fitted-models}a, and well-captures the strong pattern in the data.
-The trend is statistically significant (estimated degrees of freedom = `r round(smallSumm[["s.table"]][1,2], 3)`; $F$ = `r round(smallSumm[["s.table"]][1,3], 3)`, approximate $p$ value = $\ll$ 0.0001).
+The trend is statistically significant (estimated degrees of freedom = `r round(smallSumm[["s.table"]][1,2], 2)`; $F$ = `r round(smallSumm[["s.table"]][1,3], 2)`, approximate $p$ value = $\ll$ 0.0001).
 However further analysis of the fitted model is required to answer the other questions posed earlier about the timing of change and whether features in the trend can be distinguished from random noise.
@@
 
-### Braya-sø
+## Braya-Sø
 
@@
 
-To investigate the first issue the GCV and REML scores for an increasing sequence of values of the smoothness penalty $\lambda$ were evaluated for the standard GAM (equation \eqref{eq:gam}) fit to the \uk{} time series.
-The resulting profiles are shown in Figure \ref{fig:trace-smoothness-parameters}, with the selected value of the penalty shown by the vertical line.
-The GCV score profile suggests that the potential for under smoothing identified by @Reiss2009-fk is unlikely to apply here.
+To investigate the first issue, the GCV and REML scores for an increasing sequence of values of the smoothness parameter ($\lambda$) were evaluated for the standard GAM (equation \eqref{eq:gam}) fit to the \uk{} time series.
+The resulting profiles are shown in Figure \ref{fig:trace-smoothness-parameters}, with the optimal value of the parameter shown by the vertical line.
+The GCV score profile suggests that the potential for under smoothing identified by @Reiss2009-fk is unlikely to apply here as there is a well-defined minimum in profile.
 Addressing the violation of the independence assumption is more difficult; the standard errors of the model coefficients are most likely anti-conservative, which presents a significant problem if we wish to determine which of the wiggles in the fitted \uk{} trend can be identified as real.
 
-To understand the reason why the GAM plus CAR(1) and the GAM with REML smoothness selection performed poorly with the \uk{} time series we need to delve a little deeper into what is happening when we are fitting these two models.
-
-```{r trace-smoothness-parameters, fig.width = 8, fig.height = 2.5, fig.cap = "GCV and REML scores as a function of the smoothness parameter λ."}
+To understand the reason why the GAM plus CAR(1) and the simple GAM with REML smoothness selection performed poorly with the \uk{} time series we need to delve a little deeper into what is happening when we are fitting these two models.
+
+```{r trace-smoothness-parameters, fig.width = 8, fig.height = 2.5, fig.cap = "GCV and REML scores as a function of the smoothness parameter $\\lambda$.
+From left to right, GAMs were estimated using GCV and REML smoothness selection, and REML using a basis dimension of 40 and observational weights to account for heterogeneity in the \\uk{} times series.
+The selected value of $\\lambda$ for each model is indicated by the vertical grey line."}
 ## Generate GCV and REML traces
@@
 
-The primary issue leading to poor fit in both the GAM plus CAR(1) and the GAM with REML smoothness selection is that neither model accounts for the different variance of each observation in the \uk{} record.
-The sediments in Braya-sø are not annually laminated and therefore the core was sliced at regular depth intervals.
+The primary issue leading to poor fit is that neither model accounts for the different variance (known as (heteroscedasticity) of each observation in the \uk{} record.
+The sediments in Braya-Sø are not annually laminated and therefore the core was sliced at regular depth intervals.
 Owing to compaction of older sediments and variation in accumulation rates over time, each sediment slice represents a different number of "lake years".
-We can think of older samples as representing some average of many years of lake sediment, whilst younger samples are representative of fewer lake years.
+We can think of older samples as representing some average of many years of sediment deposition, whilst younger samples are representative of fewer of these lake years.
 The average of a larger set of numbers is estimated more precisely than the average of a smaller set, all things equal.
-A direct result of this variable averaging of lake years it that some samples are more precise and therefore have lower variance than others.
-
-The models fitted thus far fail to take account of this feature of the data, but including this information in the model is relatively simple via the use of observational weights.
+A direct result of this variable averaging of lake years it that some samples are more precise and therefore have lower variance than other samples and yet the model assumed that the variance was constant across samples.
+
+Accounting for heteroscedasticity within the model is relatively simple via the use of observational weights.
 The number of lake years represented by each slice is estimated by assigning a date to the top and bottom of each sediment slice.
@@
 A secondary problem is the size of the basis dimension used for the time variable.
-The main user selectable option when fitting a GAM in the penalised likelihood framework of @Wood2004-zv is how large a basis to.
+The main user selectable option when fitting a GAM in the penalised likelihood framework of @Wood2004-zv is how many basis functions to use.
 As described above, the basis should be large enough to contain the true, but unknown, function, or a close approximation to it.
-For GCV selection the basis contained 29 basis functions, whilst the CAR(1) model with REML smoothness selection would only converge with a basis containing 20 functions.
+For GCV selection the basis used contained 29 basis functions, whilst the CAR(1) model with REML smoothness selection would only converge with a basis containing 20 functions.
 The size of the basis appears to be sufficient for GCV smoothness selection, but following @Wood2011-kn REML smoothness selection is preferred.
 Using the test of @Pya2016-rk, the basis dimension for the models with REML smoothness selection was too small.
-To proceed therefore, we must drop the CAR(1) term and increase the basis dimension to 39 functions (by setting `k = 40`).
+To proceed therefore, we must drop the CAR(1) term and increase the basis dimension to 39 functions (by setting `k = 40`; I set it larger than expected wiggliness because the larger basis contains a richer family of functions and the excess complexity is reduced because of the smoothness penalty in play.)
 
 With the larger basis dimension and accounting for the non-constant variance of the observations via weights, the model fitted using REML is indistinguishable from that obtained using GCV (Figure \ref{fig:plot-fitted-models}b).
-The trace of the REML score for this model (Figure \ref{fig:trace-smoothness-parameters}c) shows a pronounced minimum at a much smaller value of $\lambda$ than the original REML fit (Figure \ref{fig:trace-smoothness-parameters}b), indicating that a more wiggly trend provides a better fit to the Braya-sø time series.
+The trace of the REML score for this model (Figure \ref{fig:trace-smoothness-parameters}c) shows a pronounced minimum at a much smaller value of $\lambda$ than the original REML fit (Figure \ref{fig:trace-smoothness-parameters}b), indicating that a more wiggly trend provides a better fit to the Braya-Sø time series.
 This example illustrates that some care and understanding of the underlying principles of GAMs is required to diagnose potential issues with the estimated model.
@@
 Here, despite increasing the basis size, it was the failure to appreciate the magnitude of the effect on fitting of the non-constant variance that lead to the initially poor fit and the problems associated with the estimation of the CAR(1) process.
-
-### Confidence intervals
+I return to the issue of why the GAM plus CAR(1) model encountered problems during fitting later ([see Identifiability](#identifiability)).
+
+## Confidence intervals and uncertainty estimation {#confints}
 
@@
 What uncertainty is associated with the trend estimates?
-One way to visualise this is through a 1 - $\alpha$ confidence interval around the fitted trend, where $\alpha$ is say 0.05 for a 95% interval.
-A 95% interval would be drawn at $\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))$, where $m_{1-\alpha} = 1.96$, the 0.95 probability quantile of a standard normal distribution, and $\text{SE}(\hat{y}_t)$ is the standard error of the estimated value at time $x_t$.
-This type of confidence interval would normally be described as *pointwise*; the coverage properties of the interval would be correct at a single point on the fitted trend, but, if we were to consider additional points on the trend, the coverage would logically be lower than 1 - $\alpha$.
-This would be the traditional frequentist interpretation of a confidence interval.
-However, the approach to fitting GAMs described here is an empirical Bayesian one and therefore the typical frequentist interpretation is not applicable.
+One way to visualise this is through a 1 - $\alpha$ confidence interval around the fitted trend, where $\alpha$ is say 0.05 leading to a 95% interval.
+A 95% interval would be drawn at $\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))$, with $m_{1-\alpha} = 1.96$, the 0.95 probability quantile of a standard normal distribution, and $\text{SE}(\hat{y}_t)$ is the standard error of the estimated trend at time $x_t$.
+This type of confidence interval would normally be described as *pointwise*; the coverage properties of the interval are correct for a single point on the fitted trend, but, if we were to consider additional points on the trend, the coverage would logically be lower than 1 - $\alpha$.
+This is the traditional frequentist interpretation of a confidence interval.
+However, the approach to fitting GAMs described here has a Bayesian interpretation [@Kimeldorf1970-cn; @Wahba1990-ag; @Wahba1983-eh; @Silverman1985-kw] and therefore the typical frequentist interpretation does not apply.
 @Nychka1988-rz investigated the properties of a confidence interval created as described above using standard errors derived from the Bayesian posterior covariance matrix for the estimated mode parameters.
@@
 However, to achieve this some parts of the function may have more or less than 95%-coverage.
-@Marra2012-bq recently explained Nychka's [-@Nychka1988-rz] suprising results and extended them to the case of generalised models (non-Gaussian responses).
-
-```{r posterior-simulation, fig.height = 5, fig.width = 8, fig.cap = "posterior simulation"}
+@Marra2012-bq recently explained Nychka's [-@Nychka1988-rz] surprising results and extended them to the case of generalized models (non-Gaussian responses).
+
+```{r posterior-simulation, fig.height = 5, fig.width = 8, fig.cap = "Estimated trends (thick black lines) and 20 random draws (grey lines) from the posterior distribution of the GAM fitted to the Small Water $\\delta^{15}\\text{N}$ (a) and Braya-Sø \\uk{} (b) time series."}
 set.seed(1)
@@
     geom_line(data = randSims, mapping = aes(y = simulated, x = Year, group = run),
-              colour = "#fdb863", alpha = 0.5) +
-    geom_line() +
+              colour = "grey80") +
+    geom_line(lwd = 1) +
     labs(y = d15n_label, x = "Year CE")
@@
     geom_line(data = randSims2, mapping = aes(y = simulated, x = Year, group = run),
-              colour = "#fdb863", alpha = 0.5) +
-    geom_line() +
-    labs(y = braya_ylabel, x = "Year")
+              colour = "grey80") +
+    geom_line(lwd = 1) +
+    labs(y = braya_ylabel, x = "Year CE")
 
@@
 
-Whilst the *across-the-function* or average coverage frequentist interpretation of the Bayesian credible intervals is useful, if may be important to have an interval that contains the entirety of the true function with state probability (1 - $\alpha$).
+Whilst the *across-the-function* or average coverage frequentist interpretation of the Bayesian credible intervals is useful, if may be important to have an interval that contains the entirety of the true function with some state probability (1 - $\alpha$).
 Such an interval is known as a *simultaneous* interval.
-A 1 - $\alpha$ *simultaneous* confidence interval contains *in their entirety* 1 - $\alpha$ of all random draws from the posterior distribution of the fitted model.
+A (1 - $\alpha$)100% *simultaneous* confidence interval contains *in their entirety* 1 - $\alpha$ of all random draws from the posterior distribution of the fitted model.
 
 Fitting a GAM involves finding estimates for coefficients of the basis functions.
-Together, these coefficients are distributed multivariate normal with mean vector and covariance matrix specified by the model estimates of the coefficients and their covariances.
-Random draws from this distribution can be taken, where each random draw represents a new trend that is consistent with the fitted trend but it also includes the uncertainty in the estimated trend.
+Together, these coefficients are distributed multivariate normal with mean vector and covariance matrix specified by the model estimates of the coefficients and their covariances respectively.
+Random draws from this distribution can be taken, where each random draw represents a new trend that is consistent with the fitted trend but also includes the uncertainty in the estimated trend.
 This process is known as *posterior simulation*.
 
-Figure \ref{fig:posterior-simulation} shows `r take` such random draws from the posterior distribution of the GAMs fitted to the Small Water and Braya-sø data sets.
-For the random draws illustrated in Figure \ref{fig:posterior-simulation}, a simultaneous interval should contain the entire function for on average `r round(take * 0.95, 0)` of the `r take` draws.
+Figure \ref{fig:posterior-simulation} shows `r take` random draws from the posterior distribution of the GAMs fitted to the Small Water and Braya-Sø data sets.
 In the early period of the δ^15^N time series many of the posterior simulations exhibit short periods of increasing and decreasing trend, balancing out to the relatively flat trend estimated by the GAM (Fig.\ \ref{fig:posterior-simulation}a).
 Reflecting this uncertainty, we might expect relatively wide simultaneous intervals during this period in order to contain the vast majority of the simulated trends.
-Conversely, the decreasing δ^15^N trend starting at ~1945 is consistently reproduced in the posterior simulations, suggesting both that this feature of the time series is real and statistically significant, and that the rate of change in δ^15^N is relatively precisely estimated.
-We see a similar pattern in Figure\_\ref{fig:posterior-simulation}b for the Braya-sø record; the large peak in \uk{} at ~250CE and the strong decline at ~1200CE are well defined in the posterior simulations, whereas most of the localised trends that are smaller magnitude changes in $y_t$ are associated with posterior simulations that are less well constrained with the ends of the record in particular showing considerable variation in the strength, timing, and even sign of simulated trends, reflecting the greater uncertainty in estimated trend during these periods.
+Conversely, the decreasing δ^15^N trend starting at ~1945 is consistently reproduced in the posterior simulations, suggesting that this feature of the time series is both real and statistically significant, and that the rate of change in δ^15^N is relatively precisely estimated.
+We see a similar pattern in Figure\ \ref{fig:posterior-simulation}b for the Braya-Sø record; the large peak in \uk{} at ~250CE and the strong decline at ~1200CE are well defined in the posterior simulations, whereas most of the localised trends that are smaller magnitude changes in $y_t$ are associated with posterior simulations that are less well constrained with the ends of the record in particular showing considerable variation in the strength, timing, and even sign of simulated trends, reflecting the greater uncertainty in estimated trend during these periods.
+For the random draws illustrated in Figure \ref{fig:posterior-simulation}, a simultaneous interval should contain the entire function for on average `r round(take * 0.95, 0)` of the `r take` draws.
 
 There are a number of ways in which a simultaneous interval can be computed.
-Here I follow the simulation approach described by @Ruppert2009-ig and present only the basic detail; a fuller description is contained in Appendix **X**.
+Here I follow the simulation approach described by @Ruppert2003-pt and present only the basic detail; a fuller description is contained in Appendix [1](#appendix).
 The general idea is that if we want to create an interval that contains the whole of the true function with 1 - $\alpha$ probability, we need to increase the standard Bayesian credible interval by some amount.
 We could simulate a large number of functions from the posterior distribution of the model and then search for the value of $m_{1-\alpha}$ that when multiplied by $\text{SE}(\hat{f}(x_t))$ yielded an interval that contained the whole function for $(1-\alpha)$ 100% of the functions simulated.
-In practice, the simulation method of @Ruppert2009-ig does not involve a direct search, but in a practical sense results in the critical value $m_{1-\alpha}$ required.
-
-```{r compare-intervals, fig.height = 5, fig.width = 8, fig.cap = "simultaneous intervals illustration"}
+In practice, the simulation method of @Ruppert2003-pt does not involve a direct search, but yields the critical value $m_{1-\alpha}$ required.
+
+```{r compare-intervals, fig.height = 5, fig.width = 8, fig.cap = "95\\% simultaneous confidence intervals (light grey bands) and across-the-function confidence intervals (dark grey bands) on the estimated trends (black lines) for the Small Water $\\delta^{15}\\text{N}$ (a) and Braya-Sø \\uk{} (b) time series."}
 ## small water
@@
     geom_ribbon(data = sw.sint, mapping = aes(ymin = lower, ymax = upper, x = Year),
-                fill = "#fdb863", alpha = 0.5, inherit.aes = FALSE) +
+                fill = "grey80", inherit.aes = FALSE) +
     geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
-                fill = "#e66101", alpha = 0.5, inherit.aes = FALSE) +
-    geom_line() +
+                fill = "grey60", inherit.aes = FALSE) +
+    geom_line(lwd = 1) +
     labs(y = d15n_label, x = "Year CE")
@@
     geom_ribbon(data = bs.sint, mapping = aes(ymin = lower, ymax = upper, x = Year),
-                fill = "#fdb863", alpha = 0.5, inherit.aes = FALSE) +
+                fill = "grey80", inherit.aes = FALSE) +
     geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
-                fill = "#e66101", alpha = 0.5, inherit.aes = FALSE) +
-    geom_line() +
-    labs(y = braya_ylabel, x = "Year")
+                fill = "grey60", inherit.aes = FALSE) +
+    geom_line(lwd = 1) +
+    labs(y = braya_ylabel, x = "Year CE")
 
@@
 Simultaneous intervals computed using the method described are show in Figure \ref{fig:compare-intervals} alongside the *across-the-function* confidence intervals for the trends fitted to both example data sets.
-As expected, the simultaneous interval in somewhat wider than the *across-the-function* interval.
+As expected, the simultaneous interval is somewhat wider than the *across-the-function* interval.
 The critical value $m_{1-\alpha}$ for the simultaneous interval of the estimated trend in δ^15^N is `r with(sw.sint, round(unique(crit), 2))`, whilst the same value for the \uk{} series is `r with(bs.sint, round(unique(crit), 2))`, leading to intervals that are approximately ±50% and ±75% wider than the equivalent across-the-function intervals.
 
-### Identifying periods change
+## Identifying periods change
 
@@
 Importantly, this applies to the whole time series as the linear model implies a constant rate of change throughout.
-The estimate $\hat{\beta}_1$ is the first derivative of the fitted trend.
+More formally, the estimate $\hat{\beta}_1$ is the first derivative of the fitted trend.
 
 In the GAM, the fitted trend need not be linear; the slope of the trend is potentially different at every point in the time series.
-As such we might reasonably ask *where* in the series the response $y_t$ is changing?
-Mirroring the linear model we can answer this question by determining whether or not the first derivative --- the slope --- of the fitted trend at any time point is consistent with a null hypothesis of no change.
+As such we might reasonably ask *where* in the series the response $y_t$ is changing, if at all?
+Mirroring the linear model we can answer this question by determining whether or not the first derivative at any time point $x_t$ of the fitted trend at any time point is consistent with a null hypothesis of no change.
 We want to know whether or not the first derivative is indistinguishable from a value of $0$ --- no trend --- given the uncertainty in the estimate of the derivative.
@@
 Derivatives of the fitted spline are not easily available analytically, but they can be estimated using the method of finite differences.
-Two values of the fitted trend, separated by a very small time-shift ($\Delta_t$), are predicted from the model; the difference between the fitted values for the two time points is an approximation of the true first derivative of the trend.
+Two values of the estimated trend, separated by a very small time-shift ($\Delta_t$), are predicted from the model; the difference between the estimated values for the two time points is an approximation of the true first derivative of the trend.
 As $\Delta_t \rightarrow 0$ the approximation becomes increasingly accurate.
-In practice, the first derivative of the fitted trend is evaluated via finite differences at a large number of points in the time series.
-An approximate $(1 - \alpha)$ 100% pointwise confidence interval can be calculated for the derivative estimates using standard theory (i.e.\ $\pm 1.96 \times \text{SE}(y_t)$) and the covariance matrix of the spline coefficients.
-A $(1 - \alpha)$ 100% simultaneous interval can be computed using the method described above.
-
-Figure \ref{fig:derivatives} shows the estimated first derivative of the fitted trend in the Small Water (\ref{fig:derivatives}a) and Braya-sø (\ref{fig:derivatives}b) time series.
+In practice, the first derivative of the fitted trend is evaluated using finite differences at a large number of points in the time series.
+An approximate (1 - $\alpha$)100% pointwise confidence interval can be calculated for the derivative estimates using standard theory (i.e.\ $\pm 1.96 \times \text{SE}(\hat{y}_t)$ for a 85% interval) and the covariance matrix of the spline coefficients.
+A (1 - $\alpha$)100% simultaneous interval for the derivatives can also be computed using the method described [earlier](#confints).
+Periods of significant change are identified as those time points where the (simultaneous) confidence interval on the first derivative does not include zero.
+
+Figure \ref{fig:derivatives} shows the estimated first derivative of the fitted trend in the Small Water (\ref{fig:derivatives}a) and Braya-Sø (\ref{fig:derivatives}b) time series.
 Although the estimated trend suggests a slight increase in δ^15^N from the start of the record to ~1940, the estimated trend is sufficiently uncertain that the simultaneous interval on the first derivative includes 0 throughout this period.
@@
 This time point may be taken as the first evidence for change in δ^15^N in the Small Water core.
-The simultaneous interval on the first derivative of the trend in δ^15^N is bounded away from $0$ between ~1940 and ~1975, covering the major decline in values clearly evident in the observations.
+The simultaneous interval on the first derivative of the trend in δ^15^N is bounded away from $0$ between ~1940 and ~1975, covering the major decline in values evident in the observations.
 The simultaneous interval includes $0$ from ~1975 onward, suggesting that, whilst quite pronounced, the recent increase in δ^15^N is not statistically significant.
-To determine whether the recent increase is real or not, we would require considerably more samples with which to (hopefully) more-precisely estimate the trend during this period.
+To determine whether or not the recent increase is real, we would require considerably more samples with which to (hopefully) more-precisely estimate the trend during this period.
 Alternatively, we might just have to wait until sufficient additional sedimentation has occurred to warrant recoring Small Water and reestimating the trend in δ^15^N.
 
-The estimated trend at Braya-sø exhibited a number of oscillations in in \uk{} .
+The estimated trend at Braya-Sø exhibited a number of oscillations in \uk{}.
 As we saw previously in Figures\ \ref{fig:posterior-simulation}b and \ref{fig:compare-intervals}b, many of these are subject to significant uncertainty and it is important therefore to discern which, if any, of the oscillations in the response can be identified given the model uncertainty.
 In Figure\ \ref{fig:derivatives}b only two features of the estimated trend are considered significant based on the derivatives of the smooth; one centred on ~250CE and a second at ~1150CE.
-In both these periods, the simultaneous interval for the first derivative of the trend excludes zero.
+In both these periods, the simultaneous interval for the first derivative of the trend does not include zero.
 In the first case we detect the large peak and subsequent decline in \uk{} at ~250CE, whilst at ~1150CE the large trough is identified, but not the increasing trend immediately prior to this excursion to lower \uk.
@@
 
-```{r derivatives, fig.height = 5, fig.width = 8, fig.cap = "first derivatives and 95\\% simultaneous interval"}
+```{r derivatives, fig.height = 5, fig.width = 8, fig.cap = "Estimated first derivatives (black lines) and 95\\% simultaneous confidence intervals of the GAM trends fitted to the Small Water $\\delta^{15}\\text{N}$ (a) and Braya-Sø \\uk{} (b) time series.
+Where the simultaneous interval does not include 0, the models detect significant temporal change in the response."}
 small.d <- fderiv(mod, newdata = newYear, n = N)
@@
 
-### Residual autocorrelation
-
-The estimated value of $\phi$ from the fitted model is `r round(smallPhi[2], 2)` with 95% confidence interval `r round(smallPhi[1], 2)`--`r round(smallPhi[3], 2)`, indicating moderate to strong residual autocorrelation about the fitted trend.
-Failure to account for the dependencies in the δ^15^N time series could lead have lead to the estimation of a more wiggly trend than the one shown in Figure \ref{fig:plot-fitted-models}a which would negatively impact the confidence place on the inferences we might draw from the fitted model.
-Importantly, failing to account for the strong dependency in the residuals would lead to smaller uncertainties in the estimated spline coefficients, which would propagate through to narrower confidence intervals on the fitted trend and on the first derivatives, and ultimately to the identification of significant periods of change in the δ^15^N time series.
+## Residual autocorrelation and model identification {#identifiability}
+
+The GAM fitted to the δ^15^N time series contained a CAR(1) process to model residual temporal autocorrelation in the residuals.
+The estimated magnitude of the autocorrelation is given by the parameter $\phi$.
+The estimated value of $\phi$ for the δ^15^N series is `r round(smallPhi[2], 2)` with 95% confidence interval `r round(smallPhi[1], 2)`--`r round(smallPhi[3], 2)`, indicating moderate to strong residual autocorrelation about the fitted trend.
+The correlation function is an exponentially decreasing function of temporal separation ($\Delta_t$), and whilst observations that are a few year apart are quite strongly dependent on one another, this dependence drops off rapidly as $\Delta_t$ increases and is effectively zero when samples are separated by a decade or more (Figure\ \ref{fig:car1-plot}).
+
+Failure to account for the dependencies in the δ^15^N time series could lead to the estimation of a more wiggly trend than the one shown in Figure \ref{fig:plot-fitted-models}a which would negatively impact the confidence place on the inferences we might draw from the fitted model.
+Importantly, failing to account for the strong dependency in the residuals would lead to smaller uncertainties in the estimated spline coefficients, which would propagate through to narrower confidence intervals on the fitted trend and on the first derivatives, and ultimately to the identification of significant periods of change.
+The end result would be a tendency toward anti-conservative identification of periods of change; the coverage probability would be lower than the anticipated $1 - \alpha$, leading to a greater chance of false positive results.
 
@@
 $h(\\Delta_t, \\phi)$ is the correlation between residuals separated by $\\Delta_t$ years, where $\\hat{\\phi} = \\text{0.6}$.
-The shaded band is a 95\\% confidence interval on the estimated correlation $h$."}
+The shaded band is a 95\\% pointwise confidence interval on the estimated correlation $h$."}
 ## plot CAR(1) process
@@
     ylab(expression(italic(h) * (list(Delta[t], varphi)))) +
-    xlab(expression(Delta[t]))
+    xlab(expression(Delta[t] ~ (years)))
 car1Plt
@@
 
-### Identifiability
+Problems estimating the GAM plus CAR(1) model were encountered when this was fitted to the \uk{} time series; both a smooth trend in the mean \uk{} and a CAR(1) process in the residuals lead to an unidentifiable model.
+What makes a model with a spline-based trend and an autocorrelation process like the CAR(1) potentially unidentifiable?
 
 Consider again the basic GAM for a smooth trend, \eqref{eq:additive-model}.
-In that equation the correlation matrix $\boldsymbol{\Lambda}$ was ommitted for the sake of simplicity.
-Here I reintroduce it and restate the distributional assumptions of this model
-
+In that equation the correlation matrix $\boldsymbol{\Lambda}$ was omitted for the sake of simplicity.
+As I did in \eqref{eq:small-gam}, I reintroduce it and restate the distributional assumptions of this model
 \begin{equation}
@@
 \end{equation}
-
-In the basic GAM, $\boldsymbol{\Lambda} \equiv \mathbf{I}$ is an identity matrix, a matrix with 1s on the diagonal and 0s elsewhere.
-This is where the independence assumption of the model comes from; a model residual is perfectly correlated with itself (the 1s on the diagonal), but uncorrelated with any other residual (the off-diagonal 0s).
-In the GAM plus CAR(1) model, I specified an alternative correlation function for $\boldsymbol{\Lambda}$ --- the CAR(1) with correlation parameter $\phi$.
-@Fahrmeir2008-si show that where the stochastic structure of $f$ and $\boldsymbol{\Lambda}$ approach one another, where we have a potentially wiggly trend or strong autocorrelation as $\phi \rightarrow 1$, the two processes can quickly become unidentifiable [see also @Fahrmeir2013-xu **give pages**].
+In the basic GAM, $\boldsymbol{\Lambda} \equiv \mathbf{I}$ is an identity matrix, a matrix with 1s on the diagonal and 0s elsewhere
+\begin{equation*}
+\begin{bmatrix}
+1      & 0 & 0 & \dots & 0 \\
+0      & 1 & 0 & \dots & 0 \\
+0      & 0 & 1 & \dots & 0 \\
+\vdots & \vdots & \vdots & \ddots & \vdots \\
+0      & 0 & 0 & \dots & 1
+\end{bmatrix}\, ,
+\end{equation*}
+which is where the independence assumption of the model comes from; a model residual is perfectly correlated with itself (the 1s on the diagonal), but uncorrelated with any other residual (the off-diagonal 0s).
+In the GAM plus CAR(1) model, an alternative correlation function for $\boldsymbol{\Lambda}$ was used --- the CAR(1) with correlation parameter $\phi$.
+@Fahrmeir2008-si show that where the stochastic structure of $f$ and $\boldsymbol{\Lambda}$ approach one another, i.e.\ where we have a potentially wiggly trend or strong autocorrelation as $\phi \rightarrow 1$, the two processes can quickly become unidentifiable [see also @Fahrmeir2013-xu].
 By unidentifiable, we mean that it becomes increasingly difficult to distinguish between a wiggly trend or strong autocorrelation because these two processes are very similar to one another in appearance.
+This leads to model estimation problems of the sort encountered with fitting the GAM plus CAR(1) model to the Braya-sø \uk{} series.
 
@@
 
-A situation where it may be possible to uniquely identify separate wiggly trends and autocorrelation is exemplified in Small Water.
-The non-linear trend and the autocorrelation operate at very different scales; the trend represents decadal-scale variation in the mean δ^15^N, whilst the CAR(1) process represents the much smaller-scale tendency for values of the response to be followed in time by similar values.
+Situations where it may be possible to uniquely identify separate wiggly trends and autocorrelation are exemplified by the Small Water δ^15^N time series.
+The non-linear trend and the autocorrelation operate at very different scales; the trend represents decadal-scale variation in mean δ^15^N, whilst the CAR(1) process represents the much smaller-scale tendency for values of the response to be followed in time by similar values.
 That such a pattern is observed in the Small Water core is the result of the high resolution of the sampling in time relative to the long-term trend.
-In contrast, the Braya-sø record is sampled at far lower resolution, and consequently the data do not contain sufficient information to separate trend and autocorrelation.
-
-### Gaussian process smooths
-
-In the world of machine learning, Gaussian processes are a widely-used method for fitting smooth non-parametric regression models.
+In contrast, the Braya-Sø record is sampled at far lower resolution relative to the fluctuations in the mean response, and consequently the data do not contain sufficient information to separate trend and autocorrelation.
+
+## Gaussian process smooths
+
+In the world of machine learning, Gaussian processes [@Rasmussen2006-xk; @Golding2016-bb] are a widely-used method for fitting smooth non-parametric regression models.
 A Gaussian process is a distribution over all possible smooth functions $f(x)$.
-In the field of spatial statisics, Gaussian processes are known by name *kriging*.
-
-With a Gaussian process we are interested in fitting a smooth temporal or spatial trend by modelling the way the correlation between pairs of observations, $x_i$ and $x_j$, varies as a function of the distance in space or time that separates ($h$) the observations.
-The correlation between pairs of samples decreases with increasing separation, which is modelled using a correlation function, $c(h)$.
+In the field of spatial statistics, Gaussian processes are known by name *kriging*.
+
+With a Gaussian process we are interested in fitting a smooth temporal trend by modelling the way the correlation between pairs of observations varies as a function of the distance, $h$, in time that separates ($h$) the observations.
+The correlation between pairs of observations decreases with increasing separation, which is modelled using a correlation function, $c(h)$.
 Figure \ref{fig:gp-correlation-functions} shows examples of two different correlation functions; the *power exponential* (Figure \ref{fig:gp-correlation-functions}a), and the Matérn (Figure \ref{fig:gp-correlation-functions}b) correlation functions.
 These functions are smooth and monotonic-decreasing, meaning that the value of the correlation function decreases with increasing separation ($h$).
-When $h$ = 0, the correlation is equal to 1 ($c(0) = 1$); two samples take from exactly the same location or time point are perfectly correlated.
-As $h \rightarrow \infty$, the correlation tends to zero ($c(h) \rightarrow 0$); two samples separated by a large distance or amount of time tend to be uncorrelated.
-
-```{r gp-correlation-functions, fig.width = 8, fig.height = 5, fig.cap = "correlation functions"}
+When $h$ = 0, the correlation is equal to 1 ($c(0) = 1$); two samples taken at exactly the same time point are perfectly correlated.
+As $h \rightarrow \infty$, the correlation tends to zero ($c(h) \rightarrow 0$); two samples separated by a large amount of time tend to be uncorrelated.
+Often we are interested in learning how large the separation in time needs to be before, on average, a pair of observations is effectively uncorrelated (i.e.
+where $c(h)$ is sufficiently close to zero).
+
+```{r gp-correlation-functions, fig.width = 8, fig.height = 5, fig.cap = "Power exponential (a) and Matérn (b) correlation functions for observation separation distance $h$.
+Two values of the effective range parameter ($\\phi$)) are shown for each function.
+For the power exponential function, $\\kappa$ is the power in the power exponential function.
+For the Matérn correlation function, $\\kappa$ distinguishes the member of the Matérn family."}
 `powerExp` <- function(h, phi, k) {
@@
 
+Several functions can be used to represent $c(h)$.
+Two common ones are the power exponential function and the Matérn family of correlation functions.
 The power exponential function at separation distance $h$ is
-
-$$c(h) = exp\{(-h/\phi)^{\kappa}\}$$
-
+$$c(h) = \exp\{(-h/\phi)^{\kappa}\}$$
 where $0 < \kappa \leq 2$.
@@
 When $\kappa = 1.5$, the Matérn correlation function is
-
-$$c(h) = (1 + h/\phi) exp(-h/\phi)$$
-
+$$c(h) = (1 + h/\phi) \exp(-h/\phi)$$
 whilst for $\kappa = 2.5$ it is
-
-$$c(h) = \{1 + h/\phi + (h/\phi)^2/3\} exp(-h/\phi)$$
-
+$$c(h) = \{1 + h/\phi + (h/\phi)^2/3\} \exp(-h/\phi)$$
 and for $\kappa = 3.5$
-
-$$c(h) = \{1 + h/\phi + 2(h/\phi)^2/5 + (h/\phi)^3/15\} exp(-h/\phi)$$
-
-In both cases, $\phi$ is the effective range parameter, which sets the distance beyond which the correlation function is effectively zero.
+$$c(h) = \{1 + h/\phi + 2(h/\phi)^2/5 + (h/\phi)^3/15\} \exp(-h/\phi)\, .$$
+
+In all cases, $\phi$ is the effective range parameter, which sets the distance beyond which the correlation function is effectively zero.
 
@@
 
-Gaussian processes and GAMs share many similarities and we can fit a Gaussian process using the techniques already described above for splines [@Handcock1994-lj].
-It can be shown [e.g.
-@Fahrmeir2013-xu section/pages] that the Gaussian process (kriging) model has the same penalised likelihood form as the GAM that we discussed earlier.
-Here, the observations are the knots of the smoother and each has a basis function in the form of a correlation function.
-The equivalence is only true if the basis functions do not depend on any other parameters of the model, which is only achievable if the value of $\phi$ is fixed and known.
-In general, however, we would like to estimate $\phi$ as part of model fitting.
-To achieve this we can maximise the profile likelihood or score statistic of the model over a range of values of $\phi$.
-This involves proposing a value of $\phi$ for the effective range of the correlation function and then estimating the resulting GAM by minimising the penalised log-likehood conditional upon this value of $\phi$.
-The model, and its corresponding value of $\phi$, with lowest penalised log-likelihood or score statistic is then retained as the estimated GAM.
-Figure\ \ref{fig:gp-gam-detail-plot}a shows the REML score for models estimated using a Gaussian process smooth with a Matérn correlation function, $\kappa$ = 1.5, for a values of $\phi$ between 15 and 1000.
-There is a clear minimum around 40 years separation, with the minimum REML score being observed at $\phi = `r round(effRange2, 2)`)$.
-Also shown are REML score for models using the power exponential function with $\kappa$ = 1, the minimum score observed at a somewhat higher effective range of $\phi = `r round(effRange3, 2)`$.
-
-```{r gp-gam-detail-plot, fig.height = 5, fig.width = 8, fig.cap = "profile likelihood", dependson="fit-example-gp-smooths"}
+```{r gp-gam-detail-plot, fig.height = 5, fig.width = 8, fig.cap = "Gaussian process smooths fitted to the \\uk{} time series.
+REML score traces for GAMs fitted using power exponential ($\\kappa = 1$) or Matérn ($\\kappa = 1.5$) correlation functions as a function of the effective range parameter ($\\phi$) are shown (a).
+The optimal model for each function is that with the lowest REML score.
+b) shows the resulting trends estimated using the respective correlation function with the value of $\\phi$ set to the optimal value.", dependson="fit-example-gp-smooths"}
 ## profile-likelihood plot
@@
     scale_colour_manual(name = "", values = c("#e66101","#5e3c99")) +
-    labs(y = "REML score", x = expression(Effective ~ range ~ (italic(h))))
+    labs(y = "REML score", x = expression(Effective ~ range ~ (varphi)))
 ## plot at two values of h
@@
     scale_colour_manual(name = "", values = c("#e66101","#5e3c99")) +
-    labs(y = ylabel, x = "Year CE")
+    labs(y = braya_ylabel, x = "Year CE")
 plot_grid(proflik.plt, gp.plt2, ncol = 1, labels = c("a","b"), align = "hv", axis = "lr")
@@
 
-Figure \ref{fig:gp-gam-detail-plot}b shows the estimated trends for the \uk{} time series using Gaussian process smooths with exponential and Matérn correlations functions, both using effective range values at their respective optimal value as assessed using the REML score.
-The estimated trends are very similar to one another, although differences in the more-sparsely sampled sections of the time series, there is a noticeable difference in behaviour, with the power exponential ($\kappa = 1$) version being noticeably less-smooth than the Matérn version.
+Gaussian processes and GAMs share many similarities and we can fit a Gaussian process using the techniques already described above for splines [@Handcock1994-lj; @Kammann2003-pi].
+It can be shown [e.g.
+@Fahrmeir2013-xu section/pages] that the Gaussian process model has the same penalised likelihood form as the GAM that we discussed earlier; the observations are the knots of the smoother and each has a basis function in the form of a correlation function.
+The equivalence is only true if the basis functions do not depend on any other parameters of the model, which is only achievable if the value of $\phi$ is fixed and known [@Fahrmeir2013-xu].
+In general, however, we would like to estimate $\phi$ as part of model fitting.
+To achieve this we can maximise the profile likelihood or score statistic of the model over a range of values of $\phi$ [@Wood2017-qi, pp.
+362--363].
+This involves proposing a value of $\phi$ for the effective range of the correlation function and then estimating the resulting GAM by minimising the penalised log-likehood conditional upon this value of $\phi$.
+The model, and its corresponding value of $\phi$, with lowest penalised log-likelihood or score statistic is then retained as the estimated GAM.
+Figure\ \ref{fig:gp-gam-detail-plot}a shows the REML score for models estimated using a Gaussian process smooth with a Matérn correlation function ($\kappa$ = 1.5) for a sequence of values of $\phi$ between 15 and 1000 years.
+There is a clear minimum around 40 years separation, with the minimum REML score being observed at $\phi = `r round(effRange2, 2)`)$.
+Also shown are the REML scores for models using the power exponential function ($\kappa$ = 1) with the minimum score observed at a somewhat higher effective range of $\phi = `r round(effRange3, 2)`$.
+
+Figure \ref{fig:gp-gam-detail-plot}b shows the estimated trends for the \uk{} time series using Gaussian process smooths with exponential and Matérn correlations functions, both using $\phi$ values at their respective optimal value as assessed using the REML score.
+The estimated trends are very similar to one another, although there is a noticeable difference in behaviour, with the power exponential ($\kappa = 1$) version being noticeably less-smooth than the Matérn version.
 This difference is attributable to the shapes of the respective correlation functions; the Matérn approaches a correlation of 1 smoothly as $h$ approaches 0, whilst the power exponential with $\kappa$ = 1 approaches a correlation of 1 increasingly quickly with decreasing $h$.
@@
 
-### Adaptive smoothing
-
-```{r braya-so-model-comparisons, fig.width = 8, fig.height = 2.5, fig.cap = "compare fits", dependson = "fit-example-gp-smooths"}
+## Adaptive smoothing
+
+```{r braya-so-model-comparisons, fig.width = 8, fig.height = 2.5, fig.cap = "Comparison of trends estimated using i) adaptive smooth, ii) Gaussian process, and iii) thin plate regression spline bases for the \\uk{} time series.", dependson = "fit-example-gp-smooths"}
 ## model it using gam()
@@
     geom_point(aes(x = Year, y = UK37), data = braya) +
-    geom_line(aes(colour = Model)) + labs(y = ylabel, x = "Year") +
+    geom_line(aes(colour = Model)) + labs(y = braya_ylabel, x = "Year") +
     theme(legend.position = "right") +
@@
 
-Each of the spline types that I have discussed so far shares a common feature; the degree of wiggliness over the time series is constant.
-This is due to the wiggliness being controlled by a single smoothness parameter.
+Each of the spline types that I have discussed so far share a common feature; the degree of wiggliness over the time series is fixed due to the use of a single smoothness parameter, $\lambda$.
 The definition of wiggliness, as the integrated squared second derivative of the spline, ensures that fitted smoother does not jump about wildly.
-This assumes that the data themselves are well described by a smoothly varying function.
-If we anticipate abrupt change or step-like responses to environmental forcing this underlying assumption of the GAM would suggest it is ill-suited to modelling palaeo time series in which features are evident or expected.
-
-While there is not much we can do within the GAM framework to model a series that contains both smooth trends and step-like responses, an adaptive smoother can help address problems where the time series consists of periods of very rapid change in the mean combined with periods of complacency or relatively little change.
+This assumes that the data themselves are well described by a smoothly varying trend.
+If we anticipate abrupt change or step-like responses to environmental forcing this underlying assumption of the GAM would suggest that the method is ill-suited to modelling palaeo time series in which such features are evident or expected.
+
+While there is not much we can do within the GAM framework to model a series that contains both smooth trends and step-like responses, an adaptive smoother can help address problems where the time series consists of periods of rapid change in the mean combined with periods of complacency or relatively little change.
 As suggested by their name, adaptive smoothers can adjust to changes in the wiggliness of the time series.
-This adaptive behaviour is achieved by making the smoothness parameter $\lambda$ itself depend smoothly on $x_t$; in other words, the adaptive smoother allows the wiggliness of the estimated trend to vary smoothly over time.
-This is achieved by using multiple penalties and smoothness parameters, which have the effect of letting the strength of the wiggliness vary smoothly over time.
-Whilst this does allow the estimated trend to adapt to periods of rapid change in the response, adaptive smoothers make significant demands on the data; if we used $m$ smoothness penalties to allow the wiggliness to vary over a time series, it would be like estimating $m$ separate smooths from chunks of the original series each of length $n/m$.
+This adaptive behaviour is achieved by making the smoothness parameter $\lambda$ itself depend smoothly on $x_t$ [@Ruppert2003-pt, chapter 17; @Wood2017-qi, sec.
+5.3.5]; in other words, the adaptive smoother allows the wiggliness of the estimated trend to vary smoothly over time.
+Whilst this does allow the estimated trend to adapt to periods of rapid change in the response, adaptive smoothers make significant demands on the data [@Wood2017-qi, sec.
+5.3.5]; if we used $m$ smoothness penalties to allow the wiggliness to vary over a time series, it would be like estimating $m$ separate smooths from chunks of the original series each of length $n/m$.
 In a practical sense, this limits the use of adaptive splines in palaeoecology to proxies that are readily enumerated, such as the biogeochemical proxies used in the two example data sets.
 
-Figure \ref{fig:braya-so-model-comparisons} compares trends for the Braya-sø time series estimated using GAMs with a TPRS and a Gaussian process smooths with an adaptive smoother using a basis dimension, *k*, of 45 and 5 smoothing parameters.
-There is a clear difference in behaviour of the adaptive and non-adaptive smoothers for the first 1000 years of the record, with the adaptive smooth exhibiting much less variation compared to the TPRS and Gaussian process spline.
-Over the remaining two thirds of the series, there is much closer agreement in the three trends.
-
-The behaviour of the TPRS and Gaussian process spline for these data is the result of requiring a large amount of wiggliness to adapt to the large oscillations in \uk{} present around year 250CE and again at ~900--1500CE.
-This large degree of wiggliness allows the splines to chase individual data points much earlier in the record.
-Because the adaptive smoother, in contrast, can adapt to these periods of rapid change in the response it is much less susceptible to this "chasing" behaviour.
-
-The behaviour of the TPRS- and Gaussian process-based models is undesirable, yet if we recall Figure \ref{fig:derivatives} and the discussion around the use of the first derivative to identify periods of significant change, we would not interpret the oscillations in the early part of the \uk{} record as being statistically significant.
-Owing to the paucity of data in this part of the series the trends fitted using the non-adaptive smoothers, the estimated trends are subject to such a large degree of uncertainty that the alternative of no trend through the first 1000 years of the record is also a plausible explanation of the data.
-Should we conclude that there is no trend in \uk{} and thence climate in this period?
+Figure \ref{fig:braya-so-model-comparisons} compares trends for the Braya-Sø time series estimated using GAMs with the three main types of spline discussed; i) TPRS, ii) Gaussian process smooths, and iii) an adaptive smoother using 45 basis functions and 5 smoothing parameters.
+There is a clear difference in the behaviour of the adaptive and non-adaptive smoothers for the first 1000 years of the record, with the adaptive smooth exhibiting much less variation compared with either the TPRS or Gaussian process splines.
+Over the remaining two thirds of the series, there is much closer agreement in the three smooths.
+
+The behaviour of the TPRS and Gaussian process splines for these data is the result of requiring a large amount of wiggliness (a small $\lambda$) to adapt to the large oscillations in \uk{} present around year 250CE and again at ~900--1500CE.
+This large degree of wiggliness allows the splines to potentially over-fit individual data points much earlier in the record.
+Because the adaptive smoother, in contrast, can adapt to these periods of rapid change in the response it is much less susceptible to this "chasing" behaviour --- we don't need to waste effective degrees of freedom in periods with little or no change just to be able to fit the data well when there is a lot of change.
+
+This potential for over-fitting in such situations is undesirable, yet if we recall Figure \ref{fig:derivatives} and the discussion around the use of the first derivative to identify periods of significant change, we would not interpret the oscillations in the early part of the \uk{} record as being statistically significant.
+Owing to the paucity of data in this part of the series the trends fitted using the non-adaptive smoothers are subject to such a large degree of uncertainty that the alternative of no trend through the first 1000 years of the record is also a plausible explanation of the data.
+The trend estimated using the adaptive smooth reflects this.
+Therefore, should we conclude that there is no trend in \uk{} and thence climate in this period?
 I believe that to be too-strong a statement; those oscillations in \uk{} may be real responses to climate forcing but we may simply lack the statistical power to distinguish them from the null hypothesis of no trend through this period.
-The adaptive smoother is only adjusting to the data available to it; simply because it does not detect a trend during this period does not lend itself to an interpretation of stable climate forcing or complacency in the lake's response to forcing.
-If there were particular interest in the climate of this particular period we might take from the Braya-sø record that there is potential variation in climate forcing, but that additional data from this or other sites is required before any definitive conclusion can be drawn.
-
-### Accounting for age model uncertainty
+The adaptive smoother is only adjusting to the data available to it; just because it does not detect a trend during this period does not lend itself to an interpretation of stable climate forcing or complacency in the lake's response to forcing.
+If there were particular interest in the climate of this particular period we might take from the Braya-Sø record that there is potential early variation in climate forcing, but that additional data from this or other sites is required before any definitive conclusion can be drawn.
+
+## Accounting for age model uncertainty
 
@@
 
-Thus far, the trend models that I have described and illustrated assumed that the time covariate was fixed and known.
+Thus far, the trend models that I have described and illustrated assumed that the time covariate ($x_t$) was fixed and known.
 In both examples, and more generally for most palaeoecological records, this assumption is violated.
 Unless the record is annually laminated, assigning an age to a sediment interval requires the development of an age model from observations of the relationship between depth down the sediment core and estimates of the age of the sample arrived at using any of a number of techniques, for example ^210^Pb or ^14^C radiometric dating.
-This age-depth relationship is itself uncertain, usually being derived from a mathematical or statistical model applied to point age estimates.
+This age-depth relationship is itself uncertain, usually being derived from a mathematical or statistical model applied to point age estimates [e.g.\ @Blaauw2012-yq].
 Incorporating this additional component of uncertainty complicates the estimation of statistical models from palaeoenvironmental data.
-In this section I illustrate a simulation based approach to quantify and account for age-model uncertainty as part of the trend estimation using a GAM.
-
-```{r small-scam-fit-plots, fig.width = 8, fig.height = 7.5, fig.cap = "foo", dependson = -1}
+In this section I illustrate a simulation based approach to quantify and account for age-model uncertainty as part of the trend estimation using a GAM (see @Anchukaitis2013-la for a similar, non-GAM related idea).
+
+```{r small-scam-fit-plots, fig.width = 8, fig.height = 7.5, fig.cap = "Accounting for uncertainty in age estimates whilst fitting a smooth trend to the Small Water $\\delta^{15}\\text{N}$ time series.
+(a) Estimated age model using a monotonically-constrained spline fitted to ${}^{210}\\text{Pb}$ inferred ages for selected depths in the sediment core (red points).
+The uncertainty in the ${}^{210}\\text{Pb}$ inferred age is show by the red vertical bars.
+The fitted age model is illustrated by the solid black line.
+The faint grey lines are 25 random draws from the posterior distribution of the monotonically constrained GAM.
+The effect of age uncertainty on trend estimation is shown in b); for 100 simulations from the posterior distribution of the age model in a) a trend was estimated using a GAM with a thin plate regression spline basis and a CAR(1) process in the residuals.
+These trends are shown as grey lines.
+The combined effect of age model and fitted GAM uncertainty on the trends for the $\\delta^{15}\\text{N}$ time series is shown in c).
+The grey lines in c) are based on 50 random draws from the model posterior distribution for each of the 100 trends shown in b).
+For both b) and c) the black line shows the trend estimated assuming the ages of each sediment sample are known and fixed.", dependson = -1}
 plt1 <- ggplot(swAge, aes(y = Date, x = Depth)) +
-    geom_errorbar(aes(ymin = Date - Error, ymax = Date + Error, width = 0)) +
-    geom_point() +
     geom_line(data = newSims, mapping = aes(y = Age, x = Depth, group = Simulation),
-              alpha = 0.1, colour = "#e66101") +
+              alpha = 1, colour = "grey80") +
     geom_line(data = newAge, mapping = aes(y = fitted, x = Depth)) +
+    geom_point(size = 1.5, colour = "red") +
+    geom_errorbar(aes(ymin = Date - Error, ymax = Date + Error, width = 0), colour = "red") +
     labs(y = "Year CE", x = "Depth (cm)")
@@
 plt2 <- ggplot(simTrends, aes(x = Year, y = Trend, group = Group)) +
-    geom_line(alpha = 0.1, colour = "#e66101") +
+    geom_line(alpha = 0.1, colour = "grey80") +
     geom_line(data = newYear, mapping = aes(x = Year, y = fit), inherit.aes = FALSE) +
-    geom_point(data = small, mapping = aes(x = Year, y = d15N), inherit.aes = FALSE, size = 0.5) +
+    geom_point(data = small, mapping = aes(x = Year, y = d15N), inherit.aes = FALSE, size = 0.7) +
     labs(x = "Year", y = d15n_label)
@@
 plt3 <- ggplot(simSimulate, aes(x = Year, y = Trend, group = Group)) +
-    geom_line(alpha = 0.1, colour = "#e66101") +
+    geom_line(alpha = 0.2, colour = "grey80") +
     geom_point(data = small, mapping = aes(x = Year, y = d15N), inherit.aes = FALSE,
-               size = 0.5) +
+               size = 0.7) +
     geom_line(data = newYear, mapping = aes(x = Year, y = fit), inherit.aes = FALSE) +
@@
 The fitted age-depth model is constrained to be monotonically decreasing with increasing depth, following the method of [@Pya2015-wr] using the *scam* package [@pya-scam-pkg].
-Also shown are 25 simulations from the model posterior of the montonically constrained GAM.
-Each simulation from the posterior distribution of the age-model is itself a potential age-depth model which can be used to assign dates to the Small Water core.
-The trend model in \eqref{eq:gam} can be fitted to the δ^15^N data using these new dates as a covariate, and the whole process repeated for a large number of simulations.
+Also shown are 25 simulations from the model posterior of the monotonically-constrained GAM.
+Each simulation from the posterior distribution of the age-model is itself a potential age-depth model, which can be used to assign dates to the Small Water core.
+The trend model in \eqref{eq:gam} can be fitted to the δ^15^N data using these new dates as $x_t$, and the whole process repeated for a large number of simulations from the age model.
 
@@
 In this case, the age-depth model is relatively simple with little variation in the posterior draws, resulting in trends that match closely that obtained from the estimated age-depth relationship.
-Even so, this additional uncertainty accounts suggests that the timing of the obvious decline in δ^15^N covers the interval ~1935--1945.
+Even so, this additional uncertainty suggests that the timing of the decline in δ^15^N covers the interval ~1935--1945.
 
 The uncertainty in the trend estimates illustrated in Figure \ref{fig:small-scam-fit-plots}b only reflects the variation in trends fitted to the uncertain dates of the sediment samples.
-To fully visualise the uncertainty in the trend estimates, incorporating both age model uncertainty *and* uncertainty in the estimated model coefficients themselves, 50 simulations from the posterior distributions of each of the 100 estimated trends shown in Figure \ref{fig:small-scam-fit-plots}b we performed, resulting in 5,000 trend estimates for the δ^15^N series.
-These are shown in Figure \ref{fig:small-scam-fit-plots}c, where the two obvious changes over the same simulations (Figure \ref{fig:posterior-simulation}a) are that the uncertainty band traced out by the simulations is approximately 50% wider and, not suprisingly, the uncertainty in the estimated trend is most pronounced in the least accurately-dated section of the core.
+To fully visualise the uncertainty in the trend estimates, incorporating both age model uncertainty *and* uncertainty in the estimated model coefficients themselves, 50 simulations from the posterior distribution of each of the 100 estimated trends shown in Figure \ref{fig:small-scam-fit-plots}b were performed, resulting in 5,000 trend estimates for the δ^15^N series.
+These are shown in Figure \ref{fig:small-scam-fit-plots}c, where the two obvious changes over the same simulations without accounting for uncertainty in $x_t$ (Figure \ref{fig:posterior-simulation}a) are that the uncertainty band traced out by the simulations is approximately 50% wider and, not surprisingly, the uncertainty in the estimated trend is most pronounced in the least accurately-dated section of the core.
 Despite this additional uncertainty however, the main result holds; a marked decline of ~1.5‰ that occurred between approximately 1930 and 1945, with mild evidence of a small increase in δ^15^N post 2000 CE.
 
-Although beyond the scope of this analysis, it is anticipated that uncertainties in the age model for Braya-sø will induce significant variation in the trend estimation process such that some of the features tentatively identified using the first differences would no-longer met the required level of significance to claim identification.
-This is not to say that the features in the estimated trend are not real, just that if they are real, we lack the statistical power to identify them unequivocally given the data and the signal to noise ratio therein.
-
-### Multivariate data
-
-A large proportion of the palaeoenvironmental data generated today is multivariate species or biological abundance in nature and yet the two examples that have been used here to illustrate the use of GAMs to model palaeoenvironmental time series were univariate.
+## Multivariate data
+
+A large proportion of the palaeoenvironmental data generated today is multivariate in nature and yet the two examples used to illustrate GAMs were univariate.
 Can the approach described here be used for multivariate data?
-Yes and no.
-With one main exception it is not possible to directly apply the GAM methodology to multivariate abundance data, where the aim is to model all species at once, with the methods that I have described here and illustrate in the supplementary materials.
-The available software is not able to estimate the penalized GAM for multiple non-Gaussian responses.
+Yes, and no.
+With one main exception it is not possible to directly apply the GAM methodology described here to multivariate abundance data, where the aim is to model all species at once.
+The *mgcv* software for example, is not able to estimate the penalized GAM for multiple non-Gaussian responses.
 The exception is for a small number of correlated Gaussian responses; these could be modelled as being distributed multivariate normal conditional upon the covariates.
-Such a model would estimate the expected values of each response and the correlation between them.
-For example, we could model δ^15^N and δ^13^C series using this approach.
-
-Formal multivariate versions of GLM or GAMs are currently an important area research within ecology (see [@Warton2015-ge] for a recent review), where they go by the name joint species distribution models (JSDMs).
+Such a model would estimate the expected values of each response and the correlations between them.
+For example, we could jointly model δ^15^N and δ^13^C series using this approach.
+
+Formal multivariate versions of GLM or GAMs are currently an important area of research within ecology (see @Warton2015-ge for a recent review), where they go by the name joint species distribution models (JSDMs).
 Whilst undoubtedly powerful, our knowledge regarding JSDMs and their availability in software are still in their relative infancy and they require considerable expertise to implement.
 As such, JSDMs are currently beyond the reach of most palaeoecologists.
-Despite this, we should be watching JSDM research as developments are ongoing and a degree of method maturation is occurring.
-
-One currently available avenue for fitting a multivariate GAM is via regularized sandwich estimators with GLMs [@Warton2011-vp], which involves fitting separate GLMs (or GAMs) to each response variable and subsequently using resampling-based hypothesis tests to determine which covariates are related to variation at the community level and for individual taxa [@Warton2011-vp @Warton2012-fg @Wang2012-uu].
-The *mvabund* package [@Wang2012-uu] implements this approach within R.
-
-A pragmatic although inelegant approach that has been used to estimate trends in multivariate palaeoecological data is to first summarise the response data using an unconstrained ordination via a PCA, CA, or principal curve and then fit separate GAM models to the site (sample) scores of the first few ordination axes or principal curve.
-Whilst this two-step approach is relatively easy to implement and builds on existing approaches palaeoecologists use to summarise multivariate stratigraphic data, it is best though of as modelling changes in abundance or relative composition at the community level.
-It is less well suited to unpicking taxon-specific trends however, because the ordination step combines individual species information into latent variables (axes) that are linear combinations of *all* species and it is these latent variables that are modelled using GAM.
-
-## Conclusions
-
-## References
+Despite this, we should be watching JSDM research as developments are ongoing and a degree of method maturation occurring.
+
+One currently available avenue for fitting a multivariate GAM is via regularized sandwich estimators and  GLMs [@Warton2011-vp], which involves fitting separate GLMs (or GAMs) to each response variable and subsequently using resampling-based hypothesis tests to determine which covariates are related to variation at the community level and for individual taxa [@Warton2011-vp; @Warton2012-fg; @Wang2012-uu].
+The *mvabund* package [@Wang2012-uu] implements this approach within R and can use *mgcv* to fit GAMs to each species.
+
+A pragmatic although inelegant approach that has been used to estimate trends in multivariate palaeoecological data is to first summarise the response data using an unconstrained ordination via a PCA, CA, or principal curve and then fit separate GAM models to the site (sample) scores of the first few ordination axes or principal curve [@Bennion2015-ph; @Beck2018-zm].
+Whilst this two-step approach is relatively easy to implement and builds on approaches that palaeoecologists already use to summarise multivariate stratigraphic data, it is best thought of as modelling changes in abundance or relative composition at the community level.
+It is less well suited to unpicking taxon-specific trends however, because the ordination step combines individual species information into latent variables (axes) that are linear combinations of *all* species and it is these latent variables that are then modelled using GAM.
+
+# Conclusions
+
+Formal statistical estimation of trends in palaeoenvironmental data has been hampered by the nature of the data that comprise the time series; the uneven spacing of samples in time makes it, if not impossible, difficult to fit classical statistical time series models like ARIMA.
+This has lead palaeoecologists and palaeolimnologists to fall back on basic statistical methods such as linear parametric and non-parametric correlations or simple linear regression models, where the assumptions of the method are often grossly violated by the dependencies inherent to time series data.
+GAMs, whilst similar to the popular \loess{} smoother, provide a superior alternative approach to trend estimation in palaeoenvironmental time series.
+They can estimate non-linear trends, provide estimates of the magnitude of change as well as allow the identification of periods of change, can account for the lack of independence (either via autocorrelation processes or via the fitting of a wiggly trend), and provide a formal framework for statistical inference on each of these features.
+
+In presenting the GAM with specific palaeoenvironmental examples and addressing the issues that arise in palaeoenvironmental time series, it is hoped that palaeoecologists and palaeolimnologists will be motivated to give greater consideration to the estimation of trends and the identification of change.
+
+# Conflict of interest statement {.unnumbered}
+
+The author declares that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
+
+# Acknowledgements {.unnumbered}
+
+The ideas expressed in this paper are the result of many fruitful conversations with colleagues at the Environmental Change Research Centre, UCL, and the University of Regina.
+In particular I am indebted to Helen Bennion, Rick Battarbee, and Peter Leavitt for their collaborations on projects over many years.
+Without Simon Wood's *mgcv* software and his research on GAMs, the application of these models to palaeo time series would not be as straight forward.
+I also thank David Miller, Eric Pedersen, and Noam Ross, my GAM workshop partners in crime.
+This work was supported by a Natural Sciences and Engineering Council of Canada (NSERC) Discovery Grant to the author (RGPIN-2014-04032).
+
+# References {.unnumbered}
 
@@
 
-## Appendix
-
-## Title
-
-For a set of $M$ locations over the time period of interest, which we denote by $\mathbf{g} = (g_1, g_2, \ldots, g_M)$ the 1 - $\alpha$ *simultaneous* confidence interval is
+# Appendix 1 --- Simultaneous intervals {#appendix .unnumbered}
+
+We proceed by considering a simultaneous confidence interval for a function $f(x)$ at a set of $M$ locations in $x$; we'll refer to these locations, following the notation of @Ruppert2003-pt by
+
+$$
+\mathbf{g} = (g_1, g_2, \ldots, g_M)
+$$
+
+The true function over $\mathbf{g}$, $\mathbf{f_g}$, is defined as the vector of evaluations of $f$ at each of the $M$ locations
 
@@
 \begin{aligned}
-\mathbf{\hat{f}_g} & \pm m_{1 - \alpha}
-  \begin{bmatrix} \widehat{\mathrm{st.dev}} (\hat{f}(g_1) - f(g_1)) \\
+    \mathbf{f_g} &\equiv \begin{bmatrix}
+           f(g_1) \\
+		   f(g_2) \\
+           \vdots \\
+           f({g_M}) \\
+         \end{bmatrix}
+\end{aligned}
+$$
+
+and the corresponding estimate of the true function given by the fitted GAM denoted by $\hat{\mathbf{f}}_{\mathbf{g}}$. The difference between the true function and our unbiased estimator is given by
+
+$$
+\begin{aligned}
+    \hat{\mathbf{f}}_{\mathbf{g}} - \mathbf{f_g} &= \mathbf{C_g} \begin{bmatrix}
+           \boldsymbol{\hat{\beta}} - \boldsymbol{\beta} \\
+		   \mathbf{\hat{u}} - \mathbf{u} \\
+         \end{bmatrix}\, ,
+\end{aligned}
+$$
+where $\mathbf{C_g}$ is a matrix formed by the evaluation of the basis functions at locations $\mathbf{g}$, and the part in square brackets is the bias in the estimated model coefficients, which we assume to be mean 0 and distributed, approximately, multivariate normal with mean vector $\mathbf{0}$ and covariance matrix $\mathbf{V_b}$
+
+$$
+\begin{bmatrix}
+	\boldsymbol{\hat{\beta}} - \boldsymbol{\beta} \\
+	 \mathbf{\hat{u}} - \mathbf{u} \\
+\end{bmatrix} \stackrel{\text{approx.}}{\sim} N \left (\mathbf{0}, \mathbf{V_b} \right )\, ,
+$$
+where $\mathbf{V_b}$ is the Bayesian covariance matrix of the GAM coefficients.
+
+Now, the (1 - $\alpha$)100% simultaneous confidence interval is
+
+$$
+\begin{aligned}
+    \hat{\mathbf{f}}_{\mathbf{g}} &\pm m_{1 - \alpha} \begin{bmatrix}
+           \widehat{\mathrm{st.dev}} (\hat{f}(g_1) - f(g_1)) \\
 		   \widehat{\mathrm{st.dev}} (\hat{f}(g_2) - f(g_2)) \\
@@
            \widehat{\mathrm{st.dev}} (\hat{f}(g_M) - f(g_M)) \\
-	\end{bmatrix}
+         \end{bmatrix}\, ,
 \end{aligned}
 $$
-
-where $\mathbf{f_g}$ is the true, but unknown, function and $\mathbf{\hat{f}_g}$ the estimated function at $\mathbf{g}$.
-The values in the square brackets are the standard error of the difference between the true and estimated functions at each location in $\mathbf{g}$; these are simply the standard errors of predictions from the fitted GAM.
-$m_{1 - \alpha}$ is the critical value by which the standard errors are multiplied to form the upper and lower bounds of the $1 - \alpha$ simultaneous interval.
-$m_{1 - \alpha}$ is the $1 - \alpha$ quantile of the random variable
+where $m_{1 - \alpha}$ is the 1 - $\alpha$ quantile of the random variable
 
 $$
-\sup_{x \in \mathcal{X}} \left | \frac{\hat{f}(x) - f(x)}{\widehat{\mathrm{st.dev}} (\hat{f}(x) - f(x))} \right |
+\sup_{x \in \mathcal{x}} \left | \frac{\hat{f}(x) - f(x)}{\widehat{\mathrm{st.dev}} (\hat{f}(x) - f(x))} \right | \approx \max_{1 \leq \ell \leq M} \left | \frac{\left ( \mathbf{C_g} \begin{bmatrix}
+           \boldsymbol{\hat{\beta}} - \boldsymbol{\beta} \\
+		   \mathbf{\hat{u}} - \mathbf{u} \\
+         \end{bmatrix} \right )_\ell}{\widehat{\mathrm{st.dev}} (\hat{f}(g_{\ell}) - f(g_{\ell}))} \right |
 $$
 
-where $\sup_{x \in \mathcal{X}}$ is the *supremum* or the *least upper bound* of the absolute values of the standardised deviation of the fitted function from the true function.
-The supremum is the least value of $\mathcal{X}$, the set of all values of which we observed subset $x$, that is *greater* than all of the values in the subset.
-We approximate $\sup_{x \in \mathcal{X}}$ as the maximum absolute standardised deviation of the estimated function from the true function.
-Additionally, we don't know the distribution of $\sup_{x \in \mathcal{X}}$, so we approximate it via posterior simulation.
+The $\sup$ refers to the *supremum* or the *least upper bound*; this is the least value of $\mathcal{X}$, the set of all values of which we observed subset $x$, that is *greater* than all of the values in the subset.
+Often this is the maximum value of the subset.
+This is what is indicated by the right-hand side of the equation; we want the maximum (absolute) value of the ratio over all values in $\mathbf{g}$.
+
+The fractions in both sides of the equation correspond to the standardized deviation between the true function and the model estimate, and we consider the *maximum absolute* standardized deviation.
+We don't usually know the distribution of the maximum absolute standardized deviation but we need this to access its quantiles.
+However, we can closely approximate the distribution via simulation.
+The difference here is that rather than simulating from the posterior of the model as we did earlier [see section *Confidence intervals*](#confint), this time we simulate from the multivariate normal distribution with mean vector $\mathbf{0}$ and covariance matrix $\mathbf{V_{b}}$.
+For each simulation we find the maximum absolute standardized deviation of the fitted function from the true function over the grid of $x$ values we are considering.
+Then we collect all these maxima, sort them and either take the 1 - $\alpha$ probability quantile of the maxima, or the maximum with rank $\lceil (1 - \alpha) / N \rceil$.

### 2018-05-11 analysis/braya-so/braya-so-analysis.R
clean up

@@ -1,227 +1,0 @@
-## Analyse the UK37 data for Braya So core
-
-## Packages
-library("mgcv")
-library("ggplot2")
-library("cowplot")
-library("viridis")
-
-pal <- viridis(2)
-theme_set(theme_bw())
-
-## load the simultaneous confidence interval derivatives code
-tmpf <- tempfile()
-download.file("https://gist.githubusercontent.com/gavinsimpson/ca18c9c789ef5237dbc6/raw/295fc5cf7366c831ab166efaee42093a80622fa8/derivSimulCI.R",
-              tmpf, method = "wget")
-source(tmpf)
-
-## Source other functions
-source("functions/simulate.gam.R")
-
-## Data
-braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
-                    skip = 84)
-names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
-braya <- transform(braya, sampleInterval = YearYoung - YearOld)
-head(braya)
-
-## Generate a plot of the data
-ylabel <- expression(italic(U)[37]^{italic(k)})
-plt <- ggplot(braya, aes(x = Year, y = UK37)) +
-    geom_point(colour = pal[1]) +
-    geom_line(colour = pal[1]) +
-    theme_bw() +
-    ylab(ylabel)
-plt
-
-## variography...
-library("gstat")
-library("sp")
-m.reml <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "REML")
-resi.reml <- data.frame(Residuals = resid(m.reml), Year = braya$Year, Dummy = 1)
-coordinates(resi.reml) <- ~ Year + Dummy
-hscat(Residuals ~ 1, data = resi.reml, breaks = c(0, 25, 50, 100,200, 500, 1000))
-resi.vgmCloud <- variogram(Residuals ~ 1, data = resi.reml, cloud = TRUE)
-resi.vgm <- variogram(Residuals ~ 1, data = resi.reml, width = 20, cressie = TRUE)
-
-ggplot(resi.vgm, aes(x = dist, y = gamma)) +
-    geom_smooth(se = FALSE, col = "grey", span = 0.75,
-                method.args = list(family = "gaussian")) +
-    geom_point() +
-    xlim(c(0, 1000)) + ylim(c(0, 1e-3)) +
-    ylab(expression(Semivariance ~~ gamma)) +
[…]

### 2018-05-11 analysis/braya-so/braya-so-failure-analysis.R

@@ -1,104 +1,0 @@
-## Analyse the UK37 data for Braya So core
-##
-## Illustrates how f(x) + CAR(1) + e model fails
-
-## Packages
-library("mgcv")
-library("ggplot2")
-library("cowplot")
-library("viridis")
-
-## load the simultaneous confidence interval derivatives code
-tmpf <- tempfile()
-download.file("https://gist.githubusercontent.com/gavinsimpson/ca18c9c789ef5237dbc6/raw/295fc5cf7366c831ab166efaee42093a80622fa8/derivSimulCI.R",
-              tmpf, method = "wget")
-source(tmpf)
-
-## Source other functions
-source("functions/simulate.gamm.R")
-
-## Data
-braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
-                    skip = 84)
-names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
-head(braya)
-
-mod <- gamm(UK37 ~ s(Year, k = 20), data = braya,
-            correlation = corCAR1(form = ~ Year),
-            method = "REML")
-summary(mod$gam)
-intervals(mod$lme)
-plot(mod$gam, resid = TRUE, pch = 16)
-
-## ggplot with data and fitted spline, then resids vs time in second panel
-ylabel <- expression(italic(U)[37]^{italic(k)})
-N <- 300
-newYear <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = N)))
-newYear <- cbind(newYear, data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
-crit.t <- qt(0.975, df = df.residual(mod$gam))
-newYear <- transform(newYear,
-                     upper = fit + (crit.t * se.fit),
-                     lower = fit - (crit.t * se.fit))
-
-mod2 <- gam(UK37 ~ s(Year, k = 30), data = braya)
-summary(mod2)
-plot(mod2, n = N)
-gam.check(mod2)
-
-fit2 <- predict(mod2, newdata = newYear, se.fit = TRUE)
-
-mod3 <- gam(UK37 ~ s(Year, k = 30), data = braya, method = "REML")
-summary(mod3)
-plot(mod3, n = N)
-gam.check(mod3)
-
-fit3 <- predict(mod3, newdata = newYear, se.fit = TRUE)
-
[…]

### 2018-05-11 analysis/braya-so/braya-so-induced-correlations.R

@@ -1,82 +1,0 @@
-# ---- induced-correlations-plt ----
-
-## Analyse the UK37 data for Braya So core: Look at induced correlations
-
-## Packages
-library("mgcv")
-library("ggplot2")
-library("cowplot")
-library("grid") # for unit.pmax(), unit.list()
-
-theme_set(theme_bw())
-
-## Source other functions
-source("functions/induced-correlations.R")
-
-## Data
-braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
-                    skip = 84)
-names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
-braya <- transform(braya, sampleInterval = YearYoung - YearOld)
-
-## induced correlations....
-T <- 601
-t <- 1:T
-K <- 10
-set.seed(1)
-et <- rnorm(T, sd = 0.25)
-ft <- sin(2 * pi * ((t-1) / T))
-yt <- ft + et
-egData <- data.frame(yt = yt, t = t, ft = ft)
-
-egData.plt <- ggplot(egData, aes(x = t, y = yt)) +
-    geom_point(size = 0.9) +
-        geom_line(aes(y = ft), col = "red", lwd = 1) +
-            labs(x = expression(italic(t)), y = expression(y[italic(t)]))
-## egData.plt
-
-m.ps <- gam(yt ~ s(t, bs = "ps", k = K), data = egData, method = "REML")
-m.tp <- gam(yt ~ s(t, bs = "tp", k = K), data = egData, method = "REML")
-
-tp <- c(100,225,350,500)
-
-egDataCor <- rbind(addInduced(t, tp, m.ps), addInduced(t, tp, m.tp))
-egDataCor <- cbind(egDataCor, Basis = rep(c("P Spline", "Thinplate Spline"), each = T * length(tp)))
-
-eg.plt <- ggplot(egDataCor, aes(x = t, y = correl, colour = tp)) +
-    geom_line() + facet_wrap( ~ Basis) + scale_colour_discrete(guide = FALSE) +
-            labs(x = expression(italic(t)), y = "Induced correlation")
-## eg.plt
-
-uk.ps <- gam(UK37 ~ s(Year, k = 30, bs = "ps"), data = braya, method = "REML")
-uk.tp <- gam(UK37 ~ s(Year, k = 30, bs = "tp"), data = braya, method = "REML")
-
-ukTP <- c(20,40,60,80)
[…]

### 2018-05-11 analysis/other/correlation-functions.R

@@ -1,48 +1,0 @@
-## Packages
-library("ggplot2")
-library("cowplot")
-library("viridis")
-
-## load the correlation functions
-source("./functions/correlation-functions.R")
-
-pExpdf <- expand.grid(h = seq(0, 7, length = 100),
-                      phi = c(1,2),
-                      k = c(0.5, 1, 2))
-pExpdf <- transform(pExpdf, correlation = powerExp(h, phi, k),
-                    kfac = factor(k))
-
-pexp.plt <- ggplot(pExpdf, aes(x = h, y = correlation, group = kfac, colour = kfac)) +
-    geom_line(size = 1) +
-    facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
-    ylab(expression(rho)) +
-    xlab(expression(italic(h))) +
-    ##scale_color_viridis(option = "viridis", discrete = TRUE) +
-    scale_colour_manual(values = viridis(11)[c(2,5,10)]) +
-    theme_bw() +
-    guides(colour = guide_legend(title = expression(kappa)))
-pexp.plt
-
-pMaternDf <- expand.grid(h = seq(0, 10, length = 100),
-                         phi = c(1,2),
-                         k = c(1.5, 2.5, 3.5))
-pMaternDf <- transform(pMaternDf, correlation = matern(h, phi, k),
-                       kfac = factor(k))
-
-pmat.plt <- ggplot(pMaternDf, aes(x = h, y = correlation, group = kfac, colour = kfac)) +
-    geom_line(size = 1) +
-    facet_wrap( ~ phi, labeller = label_bquote(phi == .(phi))) +
-    ylab(expression(rho)) +
-    xlab(expression(italic(h))) +
-    ## scale_color_viridis(discrete = TRUE) +
-    scale_colour_manual(values = viridis(11)[c(2,5,10)]) +
-    theme_bw() +
-    guides(colour = guide_legend(title = expression(kappa)))
-pmat.plt
-
-correlFuns <- plot_grid(pexp.plt, pmat.plt, ncol = 1, align = "hv",
-                        labels = "AUTO")
-correlFuns
-
-ggsave("./figures/correlation-functions-plot.pdf", correlFuns, width = 9,
-       height = 9)

### 2018-05-11 analysis/small-water/small-water-age-depth-model.R

@@ -1,37 +1,0 @@
-## Simple Age depth model for Small Water
-
-## Should be run from base git repo
-
-## Monotonic spline model
-
-## R Packages
-library("mgcv")                         # explicit but `monoSpline()` will check mgcv is available
-library("ggplot2")                      # for plotting
-
-source("./functions/monoSpline.R")
-
-## SMALL1 core
-small <- read.csv("./data/small-water/small1-dating.csv")
-## monotonic spline age-depth model
-smallAgeMod <- with(small, monoSpline(Depth, Date, Error))
-
-## Predict from fitted monotonic spline
-newDepths <- data.frame(Depth = seq(0, max(small$Depth), length.out = 100))
-newDepths <- transform(newDepths, Fitted = predict(smallAgeMod, newdata = newDepths))
-
-## Plot observed and interpolated ages
-plt <- ggplot(small, aes(x = Date, y = Depth)) +
-    geom_errorbarh(aes(xmin = Date - Error, xmax = Date + Error, height = 0)) +
-        geom_point() +
-            geom_line(data = newDepths, mapping = aes(x = Fitted, y = Depth), colour = "red") +
-                scale_y_reverse() + scale_x_reverse() +
-                    theme_bw() +
-                        xlab("Year") + ylab("Sediment depth (cm)")
-plt
-
-## Add age info the Small Water Isotope Data
-smallIso <- read.csv("./data/small-water/small-water-small1-isotope-data.csv")
-smallIso <- transform(smallIso, Year = predict(smallAgeMod, newdata = smallIso))
-
-## Save updated Small Water isotope data
-saveRDS(smallIso, "./data/small-water/small-water-isotope-data.rds")

### 2018-05-11 analysis/small-water/small-water-analysis.R

@@ -1,151 +1,0 @@
-## Analyse the d15N data for Small Water, SMALL1 core
-
-## Packages
-library("mgcv")
-library("ggplot2")
-library("cowplot")
-library("viridis")
-
-pal <- viridis(3)
-
-theme_set(theme_bw())
-
-## load the simultaneous confidence interval derivatives code
-tmpf <- tempfile()
-download.file("https://gist.githubusercontent.com/gavinsimpson/ca18c9c789ef5237dbc6/raw/295fc5cf7366c831ab166efaee42093a80622fa8/derivSimulCI.R",
-              tmpf, method = "wget")
-source(tmpf)
-
-## Source other functions
-source("functions/simulate.gamm.R")
-
-## Data
-small <- readRDS("./data/small-water/small-water-isotope-data.rds")
-head(small)
-
-## Generate a plot of the data
-plt <- ggplot(small, aes(x = Year, y = d15N)) +
-    geom_point(colour = pal[1]) +
-    theme_bw() +
-    ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
-plt
-
-## model it using gamm()
-mod <- gamm(d15N ~ s(Year, k = 15), data = small,
-            correlation = corCAR1(form = ~ Year), method = "REML")
-
-summary(mod$gam)
-intervals(mod$lme, which = "var-cov")
-plot(mod$gam)
-
-## plot CAR(1) process
-maxS <- with(small, diff(range(Year))) ## too large, truncate to 50
-S <- seq(0, 50, length = 100)
-phi <- intervals(mod$lme, which = "var-cov")$corStruct
-car1 <- setNames(as.data.frame(t(outer(phi, S, FUN = `^`)[1, , ])),
-                 c("Lower","Correlation","Upper"))
-car1 <- transform(car1, S = S)
-
-car1Plt <- ggplot(car1, aes(x = S, y = Correlation)) +
-    geom_ribbon(aes(ymax = Upper, ymin = Lower),
-                fill = "grey", alpha = 0.3) +
-    geom_line() +
-    ylab(expression(italic(h) * (list(Delta[t], phi))))
-car1Plt
-
-
-## Predict from model
-newYear <- with(small, data.frame(Year = seq(min(Year), max(Year), length.out = 100)))
-newYear <- cbind(newYear, data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
-
-## Draw the fitted spline on the data
-plt.fit <- ggplot(small, aes(x = Year, y = d15N)) +
[…]

### 2018-05-11 functions/correlation-functions.R

@@ -1,13 +1,0 @@
-## Implements the power exponential and matern covariance functions
-
-`powerExp` <- function(h, phi, k) {
-    exp(- abs(h / phi)^k)
-}
-
-`matern` <- function(h, phi, k) {
-    absh <- abs(h / phi)
-    addn <- ifelse(k == 1.5, 0,
-                   ifelse(k == 2.5, (1/3)*absh^2,
-                          ((2/5)*absh^2) + ((1/15) * absh^3)))
-    exp(- absh) * (1 + absh + addn)
-}

### 2018-05-11 functions/induced-correlations.R

@@ -1,30 +1,0 @@
-
-corFun <- function(V, t) {
-    top <- V[t, ]
-    bot <- sqrt(V[t,t] * diag(V))
-    top / bot
-}
-
-inducedCor <- function(m, i) {
-    ind <- rep(0,0)
-    for (j in 1:length(m$smooth))
-        ind <- c(ind, m$smooth[[j]]$first.para-1 + 1:m$smooth[[j]]$rank)
-    mmat <- model.matrix(m)
-    Z <- mmat[,ind]
-    ZZt <- tcrossprod(Z)
-    I <- diag(nrow(mmat))
-    tmp <- capture.output(vcomps <- gam.vcomp(m))
-    sigma <- vcomps[2,1]
-    tau <- vcomps[1,1]
-    V <- (sigma^2 * I) + (tau^2 * ZZt)
-    corFun(V, i)
-}
-
-addInduced <- function(t, tp, mod) {
-    cors <- unlist(lapply(tp, function(i, mod) inducedCor(mod, i), mod = mod),
-                   use.names = FALSE)
-    df <- data.frame(t  = rep(t, length(tp)),
-                     tp = factor(rep(tp, each = length(t))),
-                     correl = cors)
-    df
-}

### 2018-05-11 functions/monoSpline.R

@@ -1,49 +1,0 @@
-`monoSpline` <- function(depth, date, error = NULL, increasing = FALSE,
-                         bs = "cr", k = 10, fx = FALSE, ...) {
-    stopifnot(require("mgcv"))
-    if(is.null(error)) {
-        error <- rep(0, length(depth))
-        W <- rep(1, length(depth))
-    } else {
-        W <- 1 / error
-        W[1] <- W[2] * error[2]
-    }
-    df <- data.frame(Depth = depth, Date = date, error = error)
-    mod <- gam(Date ~ s(Depth, k = k, bs = bs, fx = fx), data = df,
-               fit = FALSE)
-    sm <- smoothCon(s(Depth, k = k, bs = bs, fx = fx), data = df,
-                    knots = NULL)[[1]]
-    ## Fm are the constraints to enforce monotonicity
-    Fm <- mono.con(sm$xp, up = increasing)
-    G <- list(X = sm$X, C = matrix(0,0,0), sp = mod$sp, p = -sm$xp,
-              y = df$Date, Ain = Fm$A, bin = Fm$b,
-              S = sm$S, off = 0, w = W) ## w = df$Date*0+1
-    ## fitted parameters
-    p <- pcls(G)
-    ## fitted values
-    pm <- Predict.matrix(sm, data = data.frame(Depth = df$Depth))
-    fit <- pm %*% p
-    ## set up return object
-    ## this needs all info for a predict() method to work, inc
-    ## sm - smoothing matrix
-    ## p  - a vector of fitted coefficients; the object returned from pcls
-    ##
-    ## also return fitted values
-    out <- list(y = df$Date,
-                fitted.values = drop(fit),
-                coefficients = p,
-                residuals = drop(df$Date - fit),
-                sm = sm)
-    class(out) <- "monoSpline"
-    out
-}
-
-## predict S3 method
-`predict.monoSpline` <- function(object, newdata, ...) {
-    if(missing(newdata))
-        yhat <- fitted(object)
-    else {
-        yhat <- Predict.matrix(object$sm, data = newdata) %*% coef(object)
-    }
-    yhat
-}

### 2018-05-11 functions/simulate.gam.R

@@ -1,34 +1,0 @@
-`simulate.gam` <- function(object, nsim = 1, seed = NULL, newdata,
-                            freq = FALSE, unconditional = FALSE, ...) {
-    if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE))
-        runif(1)
-    if (is.null(seed))
-        RNGstate <- get(".Random.seed", envir = .GlobalEnv)
-    else {
-        R.seed <- get(".Random.seed", envir = .GlobalEnv)
-        set.seed(seed)
-        RNGstate <- structure(seed, kind = as.list(RNGkind()))
-        on.exit(assign(".Random.seed", R.seed, envir = .GlobalEnv))
-    }
-
-    if (missing(newdata)) {
-        newdata <- object$model
-    }
-
-    ## random multivariate Gaussian
-    ## From ?predict.gam in mgcv package
-    ## Copyright Simon N Wood
-    ## GPL >= 2
-    rmvn <- function(n, mu, sig) { ## MVN random deviates
-        L <- mroot(sig)
-        m <- ncol(L)
-        t(mu + L %*% matrix(rnorm(m * n), m, n))
-    }
-
-    Rbeta <- rmvn(n = nsim,
-                  mu = coef(object),
-                  sig = vcov(object, freq = freq, unconditional = unconditional))
-    Xp <- predict(object, newdata = newdata, type = "lpmatrix")
-    sims <- Xp %*% t(Rbeta)
-    sims
-}

### 2018-05-11 functions/simulate.gamm.R

@@ -1,35 +1,0 @@
-`simulate.gamm` <- function(object, nsim = 1, seed = NULL, newdata,
-                            freq = FALSE, unconditional = FALSE, ...) {
-    if (!exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE))
-        runif(1)
-    if (is.null(seed))
-        RNGstate <- get(".Random.seed", envir = .GlobalEnv)
-    else {
-        R.seed <- get(".Random.seed", envir = .GlobalEnv)
-        set.seed(seed)
-        RNGstate <- structure(seed, kind = as.list(RNGkind()))
-        on.exit(assign(".Random.seed", R.seed, envir = .GlobalEnv))
-    }
-
-    if (missing(newdata)) {
-        newdata <- object$gam$model
-    }
-
-    ## random multivariate Gaussian
-    ## From ?predict.gam in mgcv package
-    ## Copyright Simon N Wood
-    ## GPL >= 2
-    rmvn <- function(n, mu, sig) { ## MVN random deviates
-        L <- mroot(sig)
-        m <- ncol(L)
-        t(mu + L %*% matrix(rnorm(m * n), m, n))
-    }
-
-    Rbeta <- rmvn(n = nsim,
-                  mu = coef(object$gam),
-                  sig = vcov(object$gam, freq = freq,
-                             unconditional = unconditional))
-    Xp <- predict(object$gam, newdata = newdata, type = "lpmatrix")
-    sims <- Xp %*% t(Rbeta)
-    sims
-}

### 2018-05-11 manuscript.R
extract R code

@@ -1,0 +1,564 @@
+## ----knitr-defaults, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE----
+knitr::opts_chunk$set(comment=NA, fig.align = "center", out.width = "0.8\\linewidth",
+                      echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
+knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
+
+## ----load-packages, cache = FALSE----------------------------------------
+library("mgcv")
+library("scam")
+library("ggplot2")
+library("cowplot")
+library("grid")                         # for unit.pmax(), unit.list()
+library("schoenberg")
+library("tidyr")
+
+## Default ggplot theme
+theme_set(theme_bw())
+
+## ----common-components, cache = FALSE------------------------------------
+## source Small Water data
+small <- readRDS("./data/small-water/small-water-isotope-data.rds")
+
+## load braya so data set
+braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt", skip = 84)
+names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
+braya <- transform(braya, sampleInterval = YearYoung - YearOld)
+
+## plot labels
+d15n_label <- expression(delta^{15}*N)
+braya_ylabel <- expression(italic(U)[37]^{italic(k)})
+
+## ----data-figure, fig = TRUE, fig.width = 8, fig.height = 5, fig.cap = "Example time series; a) Small Water bulk organic matter $\\delta^{15}\\text{N}$ time series on a ${}^{210}\\text{Pb}$ time scale, and b) Braya-Sø \\uk{} time series on a calibrated ${}^{14}\\text{C}$ time scale. The observations \\uk{} time series have been joined by lines purely as a visual aid to highlight potential trends."----
+## Generate a plot of the data - used as a base later
+small_plt <- ggplot(small, aes(x = Year, y = d15N)) +
+    geom_point() +
+    labs(y = d15n_label, x = "Year CE")
+
+## Generate a plot of the data
+braya_plt <- ggplot(braya, aes(x = Year, y = UK37)) +
+    geom_line(colour = "grey") +
+    geom_point() +
+    labs(y = braya_ylabel, x = "Year CE")
+
[…]

### 2018-05-11 manuscript.tex
the rendered tex

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    %\usepackage{mathspec}
    \usepackage{unicode-math}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
    \setmainfont{TeX Gyre Pagella}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Gavin L. Simpson},
            pdftitle={Modelling palaeoecological time series using generalized additive models},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

\title{Modelling palaeoecological time series using generalized additive models}

%%% %%% \author{Gavin L. Simpson}
%%% 
\usepackage{authblk}
\author[1]{Gavin L. Simpson\\ \href{mailto:gavin.simpson@uregina.ca}{gavin.simpson@uregina.ca}}
\affil[1]{Institute of Environmental Change and Society, ~University of Regina,
Regina, Saskatchewan, Canada, S4S 0A2}

\date{May 11, 2018}

%\setromanfont{Constantia}
%\setromanfont{texgyrepagella.otf}
%\setmainfont{XITS}
%\setmathfont{XITS Math}
\setmathfont{TeX Gyre Pagella Math}
\setmathfont[range={\mathcal,\mathbfcal},StylisticSet=1]{XITS Math}

%\renewcommand\familydefault{\rmdefault} %comment to see the difference
%\DeclareMathAlphabet      {\mathup}{OT1}{\familydefault}{m}{n}

\newcommand{\uk}{\ensuremath{\mathit{U}^{\mathit{K}}_{\mathup{37}}}}

\newcommand{\loess}{\textsc{Loess}}

\usepackage{lineno}
\linenumbers

\begin{document}
\maketitle

\newcommand*{\keywords}[1]{\textbf{\textit{Keywords---}} #1}
    \keywords{time series; generalized additive model; simultaneous interval; spline;
environmental change}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{tbp}
\makeatother

\section*{Abstract}\label{abstract}
\addcontentsline{toc}{section}{Abstract}

In the absence of annual laminations, time series generated from lake
sediments or other similar stratigraphic sequences are irregularly
spaced in time, which complicates formal analysis using classical
statistical time series models. In lieu, statistical analyses of trends
in palaeoenvironmental time series, if done at all, have typically used
simpler linear regressions or (non-) parametric correlations with little
regard for the violation of assumptions that almost surely occurs due to
temporal dependencies in the data or that correlations do not provide
estimates of the magnitude of change, just whether or not there is a
linear or monotonic trend. Alternative approaches have used
\loess{}-estimated trends to justify data interpretations or test
hypotheses as to the causal factors without considering the inherent
subjectivity of the choice of parameters used to achieve the \loess{}
fit (e.g.~span width, degree of polynomial). Generalized additive models
(GAMs) are statistical models that can be used to estimate trends as
smooth functions of time. Unlike \loess{}, GAMs use automatic smoothness
selection methods to objectively determine the complexity of the fitted
trend, and as formal statistical models, GAMs, allow for potentially
complex, non-linear trends, a proper accounting of model uncertainty,
and the identification of periods of significant temporal change. Here,
I present a consistent and modern approach to the estimation of trends
in palaeoenvironmental time series using GAMs, illustrating features of
the methodology with two example time series of contrasting complexity;
a 150-year bulk organic matter δ\textsuperscript{15}N time series from
Small Water, UK, and a 3000-year alkenone record from Braya-Sø,
Greenland. I discuss the underlying mechanics of GAMs that allow them to
learn the shape of the trend from the data themselves and how
simultaneous confidence intervals and the first derivatives of the trend
are used to properly account for model uncertainty and identify periods
of change. It is hoped that by using GAMs greater attention is paid to
the statistical estimation of trends in palaeoenvironmental time series
leading to more a robust and reproducible palaeoscience.

\section{Introduction}\label{introduction}

The vast majority of data produced by palaeoecologists and
palaeolimnologists is in the form of time-ordered observations on one or
more proxies or biological taxa (Birks, 2012b; Smol, 2008; Smol et al.,
2012). Typically these data are arranged irregularly in time; in the
absence of annual laminae or varves, the sediment core is sectioned at
regular depth intervals (Glew et al., 2001), which, owing to variation
in accumulation rates over time and compaction by overlying sediments,
results in an uneven sampling in time. An under-appreciated secondary
feature of such sampling is that younger sediments often have larger
variance than older sediments; each section of core represents fewer
lake years in newer samples, relative to older samples. This variable
time averaging acts as a variable low-pass (high-cut) filter of the raw,
annual depositional signal.

The irregular intervals between samples means that the customary
time-series analysis methods of autoregressive or moving average
processes, in the form of autoregressive integrated moving average
(ARIMA) models, are practically impossible to apply because software
implementing these methods typically requires even spacing of
observations in time. Dutilleul et al. (2012) and Birks (2012a),
eschewing the term \emph{time series}, prefer to call such data
\emph{temporal series} on account of the irregular spacing of samples, a
distinction that I find unnecessary.

Where statistical approaches have been applied to trend estimation in
palaeoenvironmental time series, the most commonly used has been
\loess{} (Birks, 1998, 2012a; Cleveland, 1979; Juggins and Telford,
2012). \loess{}, or locally weighted scatterplot smoother, as it's name
suggests, was developed to smooth x-y scatterplot data, ostensibly as
part of exploratory data analysis (Cleveland, 1979). The method fits a
smooth line through data by fitting weighted least squares (WLS) models
to observations within a particular, user-specified window of the target
point, whose width is typically expressed as a proportion \(\alpha\) of
the \(n\) data points. Weights are determined by how close (in the
x-axis only) an observation in the window is to the focal point, with a
tricubic function, by default, giving greatest weight to points closest
to the focal point. The interim \loess{}-smoothed value for the focal
point is the predicted value from the weighted regression at the focal
point. The interim values are updated using weights based on how far in
the y-axis direction, the interim smoothed value lies from the observed
value plus the x-axis distance weights; this has the effect of
down-weighting the effects of outlier observations. The final \loess{}
is obtained by joining up the smoothed (fitted) values. The user has to
choose how large a window to use, whether to fit degree 1 (linear) or
degree 2 (quadratic) polynomials in the WLS model, and how to weight
points in the x-axis. When used in an exploratory mode, the user has
considerable freedom to choose the detail of the \loess{} fit; the
window width, for example, can be infinitely tweaked to give as close a
fit to the data, as assessed by eye, as is desired. Using
cross-validation (CV) to choose \(\alpha\) or the degree of polynomial
in the WLS model is complicated for a number of reasons, not least that
the CV scheme used must involve the time ordering of the data (e.g.
Bergmeir et al., 2018). This level of subjectivity is problematic
however once we wish to move beyond exploratory data analysis to
statistically identify trends and to test hypotheses involving those
trend estimates.

Running means or other types of filter (Juggins and Telford, 2012) have
also been used extensively to smooth palaeoenvironmental time series,
but, as with \loess{}, their behaviour depends on a number of factors,
including the filter width. Furthermore, the width of the filter causes
boundary issues; with a centred filter, of width five data points, the
filtered time series would be two data points shorter at both ends of
the series because the filter weights are not defined for the first and
last two observations of the original series as these time points were
not observed. Considerable research effort has been expended to identify
ways to pad the original time series at one or both ends to maintain the
original length in the filtered series, without introducing bias because
of the method of padding (e.g. Mann, 2004, 2008; Mills, 2006, 2007,
2010).

These are by no means the only methods that have been used to estimated
trends in stratigraphic series. Another common approach involves fitting
a simple linear trend using ordinary least squares regression and use
the resulting \emph{t} statistic as evidence against the null hypothesis
of no trend despite the statistical assumptions of the method being
almost surely violated due to dependence among observations arising from
temporal autocorrelation in the series. The Pearson correlation
coefficient, \emph{r}, is also often used to detect trends in palaeo
time series (Birks, 2012a), despite the fact that \emph{r} provides no
information as to the magnitude of the estimated trend, and the same
temporal autocorrelation problem that dogs ordinary least squares
similarly plagues significance testing for \emph{r} (Tian et al., 2011).
Additionally, both the simple least squares trend line and \emph{r} are
tests for \emph{linear} trends only, and yet we typically face data sets
with potentially far more complex trends than can be identified by these
methods. Instead, non-parametric rank correlation coefficients have been
used (Birks, 2012a; Gautheir, 2001), and whilst these do allow for the
detection of non-linear trends, trends are restricted to be monotonic,
no magnitude of the trend is provided, and the theory underlying
significance testing of Spearman's \(\rho\) and Kendall's \(\tau\)
assumes independent observations.

Palaeoecology and palaeolimnology have moved away from being descriptive
disciplines, rapidly adopting new statistical developments in the 1990s
and beyond (@ Smol et al., 2012). Less development has been observed in
the area of trend estimation in palaeoenvironmental time series. Here, I
describe the use of generalized additive models (GAMs; Hastie and
Tibshirani, 1986, 1990; Ruppert et al., 2003; Wood, 2017; Yee and
Mitchell, 1991) for trend estimation. GAMs, like simple linear
regression, are a formal regression-based method for estimating trends,
yet they are also, superficially at least, similar to \loess{}. GAMs and
\loess{} estimate smooth, non-linear trends in time series and both can
handle the irregular spacing of samples in time, yet GAMs do not suffer
from the subjectivity that plagues \loess{} as a method of formal
statistical inference.

In the subsequent sections, I present an introduction to GAMs and how
they are able to estimate smooth, non-linear trends from time series.
Next the issue of uncertainty in model-estimated trends is discussed, as
is the topic of posterior simulation from a regression model. The issue
of identifying periods of significant environmental change is considered
and I show how this can be achieved using the first derivative of the
estimated trend. Two non-standard types of spline --- adaptive smoothers
and Gaussian process splines --- that are especially applicable to GAMs
in the palaeoenvironmental setting are subsequently described, followed
by an assessment of the the impact of age-model uncertainty on trend
estimation via GAMs. Finally, I briefly discuss the application of GAM
trend analysis to multivariate species abundance and compositional data.

\subsection{Example time series}\label{example-time-series}

To illustrate the estimation of trends in palaeoenvironmental data using
GAMs, I use two proxy time series; a 150-year bulk organic matter
δ\textsuperscript{15}N time series from Small Water, and a 3000-year
alkenone record from Braya-Sø. Between them, the two examples, combine
many of the features of interest to palaeoecologists that motivate the
use of GAMs; non-linear trends and the question of when changes in the
measured proxy occurred. The example analyses were all performed using
the \emph{mgcv} package and R, and the supplementary material contains a
fully annotated document showing the R code used to replicate all the
analyses described in the remainder of the paper.

\subsubsection{\texorpdfstring{δ\textsuperscript{15}N time series from
Small
Water}{δ15N time series from Small Water}}\label{15n-time-series-from-small-water}

Figure \ref{fig:data-figure}a shows 48 nitrogen stable isotope
measurements on the bulk organic matter of a sediment core collected
from Small Water, a small corrie lake located in the English Lake
District, UK. The data were collected to investigate disturbance of
nitrogen (N) cycling in remote, oligotrophic lakes by N deposited from
the atmosphere (Simpson, unpublished data). The data are shown on a
\textsuperscript{210}Pb time scale. Questions that might be asked about
this series are; what is the trend in δ\textsuperscript{15}N?, when do
we first see evidence for a change in δ\textsuperscript{15}N?, and is
the reversal in δ\textsuperscript{15}N values in the uppermost section
of the core a real change?

\subsubsection{Braya-Sø alkenone time
series}\label{braya-s-alkenone-time-series}

The second example time series is a 3,000 year record of alkenone
unsaturation, \uk, from Braya-Sø, a meromictic lake in West Greenland
(D'Andrea et al., 2011). Alkenones are long-chained unsaturated organic
compounds that are produced by a small number of planktonic organisms
known as haptophytes. The \uk{} unsaturation index is (Brassell, 1993)

\begin{equation*}
\uk{} = \frac{[C_{37:2}] - [C_{37:4}]}{[C_{37:2}] + [C_{37:3}]+ [C_{37:4}]}
\end{equation*}

where \([C_{37:x}]\) is the concentration of the alkenone with 37 carbon
atoms and \(x\) double carbon bonds. The relative abundance of these
alkenones is known to vary with changes in water temperature (Brassell,
1993; Chu et al., 2005; Toney et al., 2010; Zink et al., 2001), and as a
result \uk{} is used as a proxy for lake- and sea-surface temperatures.
For further details on the Braya-Sø \uk{} record and age model see
D'Andrea et al. (2011). Here I use the 3,000 year \uk{} record from the
PAGES 2K database (PAGES 2K Consortium, 2013). The data are presented in
Figure \ref{fig:data-figure}b.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/data-figure-1}

}

\caption{Example time series; a) Small Water bulk organic matter $\delta^{15}\text{N}$ time series on a ${}^{210}\text{Pb}$ time scale, and b) Braya-Sø \uk{} time series on a calibrated ${}^{14}\text{C}$ time scale. The observations \uk{} time series have been joined by lines purely as a visual aid to highlight potential trends.}\label{fig:data-figure}
\end{figure}

\section{Regression models for palaeoenvironmental time
series}\label{regression-models-for-palaeoenvironmental-time-series}

A linear model for a trend in a series of \(T\) observations \(y_t\) at
observation times \(x_t\) with \(t = 1, 2, \ldots, T\) is

\begin{equation} \label{eq:linear-model}
y_t = \beta_0 + \beta_1 x_t + \varepsilon_t \,,
\end{equation}

where \(\beta_0\) is a constant term, the model \emph{intercept},
representing the expected value of \(y_t\) where \(x_t\) is 0.
\(\beta_1\) is the \emph{slope} of the best fit line through the data;
it measures the rate of change in \(y\) for a unit increase in \(x\).
The unknowns, the \(\beta_j\) are commonly estimated using least squares
by minimising the sum of squared errors, \(\sum_t \varepsilon_t^2\). If
we want to ask if the estimated trend \(\beta_1\) is statistically
significant, a process called \emph{inference}, we must make further
assumptions about the data (conditional upon the fitted model) or the
model errors (residuals);
\(\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)\). This
notation indicates that the residuals \(\varepsilon_t\) are
\emph{independent} and \emph{identically distributed} Gaussian random
variables with mean equal to \(0\) and constant variance \(\sigma^2\).
In the time series setting, the assumption of independence of model
residuals is often violated.

The linear model described above is quite restrictive in terms of the
types of trend it can fit; essentially linear increasing or decreasing
trends, or, trivially, a null trend of no change. This model can be
extended to allow for non-linear trends, most notably by making \(y_t\)
depend on polynomials of \(x_t\), for example

\begin{align} \label{eq:polynomial-model}
y_t &= \beta_0 + \beta_1 x_t + \beta_2 x_t^2 + \cdots + \beta_P x_t^P + \varepsilon_t \\
    &= \beta_0 + \sum_{p = 1}^P \beta_p x_t^p  + \varepsilon_t \,, \nonumber
\end{align}

where polynomials of \(x_t\) up to order \(P\) are used. This model
allows for more complex trends but it remains a fully parametric model
and suffers from several problems, especially the behaviour of the
fitted polynomial at the start and end of the observed series.

Linear models using a range of polynomials fitted to the Small Water
data set are shown in Figure \ref{fig:polynomial-example-plot}. The
low-order models (\(P \in \{1, 3\}\)) result in very poor fit to the
data. The model with \(P = 5\) does a reasonable job of capturing the
gross pattern in the time series, but fails to adapt quickly enough to
the decrease in δ\textsuperscript{15}N that begins \textasciitilde{}1940
CE, and the estimated trend is quite biased as a result. The
\(P = 10\)th-order polynomial model is well able to capture this period
of rapid change, but it does so at the expense of increased complexity
in the estimated trend prior to \textasciitilde{}1940. Additionally,
this model (\(P = 10\)) has undesirable behaviour at the ends of the
series, significantly overfitting the data, a commonly observed problem
in polynomial models such as these (Epperson, 1987; Runge, 1901).
Finally, the choice of what order of polynomial to fit is an additional
choice left for the analyst to specify; choosing the optimal \(P\) is
not a trivial task when the data are a time series and residual
autocorrelation is likely present.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/polynomial-example-plot-1}

}

\caption{Linear models with various orders of polynomial of the covariate Year fitted using ordinary least squares to the $\delta^{15}\text{N}$ time series from Small Water. The degree of polynomial is indicated, with the degree 1 line equal to a simple linear regression model.}\label{fig:polynomial-example-plot}
\end{figure}

Can we do better than these polynomial fits? In the remainder, I hope to
demonstrate that the answer to that question is emphatically ``yes''!
Below I describe a coherent and consistent approach to modelling
palaeoenvironmental time series using generalized additive models, which
builds upon the linear regression framework.

\section{Generalized additive models}\label{generalized-additive-models}

The GAM version of the linear model \eqref{eq:linear-model} is

\begin{equation} \label{eq:additive-model}
y_t = \beta_0 + f(x_t) + \varepsilon_t \, ,
\end{equation}

where the linear effect of time(the \(\beta_1 x_t\) part) has been
replaced by a smooth function of time, \(f(x_t)\). The immediate
advantage of the GAM is that we are no longer restricted to the shapes
of trends that can be fitted via global polynomial functions such as
\eqref{eq:polynomial-model}. Instead, the shape of the fitted trend will
be estimated from the data itself.

The linear model is a special case of a broader class know as the
generalized linear model (GLM; McCullagh and Nelder, 1989). The GLM
provides a common framework for modelling a wide range of types of data,
such as count, proportions, or binary (presence/absence) data, that are
not conditionally distributed Gaussian. GLMs are, like the linear model,
parametric in nature; the types of trends that we can fit using a GLM
are the linear or polynomial models. GAMs extend the GLM by relaxing
this parametric assumption; in a GAM some, or all, of the parametric
terms, the \(\beta_p\), are replace by smooth functions \(f_j\) of the
covariates \(x_j\). For completeness then, we can write
\eqref{eq:additive-model} as a GLM/GAM

\begin{subequations}
\label{eq:gam}
\begin{align}
y_t &\sim \text{EF}(\mu_t, \Theta) \label{eq:gam-ef} \\
g(\mu_t) &= \beta_0 + f(x_t) \label{eq:gam-linear-pred} \\
\mu_t    &= g^{-1}(\beta_0 + f(x_t)), \label{eq:gam-inverse} 
\end{align}
\end{subequations}

where \(\mu_t\) is the expected value (e.g.~the mean count or the
probability of occurrence) of the random variable \(Y_t\)
(\(\mu_t \equiv \mathbb{E}(Y_t)\)) of which we have observations
\(y_t\). \(g\) is the link function, an invertible, monotonic function,
such as the natural logarithm, and \(g^{-1}\) is its inverse. The link
function maps values from the response scale on to the scale of the
linear predictor, whilst the inverse of thew link function provides the
reverse mapping. For example, count data are strictly non-negative
integer values and are commonly modelled as a Poisson GLM/GAM using the
natural log link function. On the log scale, the response can take any
real value between \(-\infty\) and \(+\infty\), and it is on this scale
that model fitting actually occurs (i.e.~using equation
\eqref{eq:gam-linear-pred}). However we need to map these unbounded
values back on to the non-negative response scale. The inverse of the
log link function, the exponential function, achieves this and maps
values to the interval 0--\(\infty\) (equation \eqref{eq:gam-inverse}).

In \eqref{eq:gam-ef}, we further assume that the observations are drawn
from a member of the exponential family of distributions --- such as the
Poisson for count data, the binomial for presence/absence or counts from
a total --- with expected value \(\mu_t\) and possibly some additional
parameters \(\Theta\) (\(y_t \sim \text{EF}(\mu_t, \Theta)\)).
Additionally, many software implementations of the above model also
allow for distributions that are not within the exponential family but
which can be fitted using an algorithm superficially similar to the one
used to fit GAMs to members of the exponential family (e.g. Wood et al.,
2016). Common examples of such extended families include the negative
binomial distribution (for overdispersed counts) and the beta
distribution (for true proportions or other interval-bounded data).

\subsection{Basis functions}\label{basis-functions}

It is clear from the plots of the data that we require the fitted trends
for the Small Water δ\textsuperscript{15}N and Braya-Sø \uk{} time
series to be non-linear functions, but it is less clear how to specify
the actual shape require. Ideally, we'd like to learn the shape of the
trend from the data themselves. We will refer to these non-linear
functions as \emph{smooth functions}, or just \emph{smooths} for short,
and we will denote a smooth using \(f(x_t)\). Further, we would like to
represent the smooths in a way that \eqref{eq:gam} is represented
parametrically so that it can be estimate within the well-studied GLM
framework. This is achieved by representing the smooth using a
\emph{basis}. A basis is a set of functions that collectively span a
space of smooths that, we hope, contains the true \(f(x_t)\) or a close
approximation to it. The functions in the basis are known as \emph{basis
functions}, and arise from a \emph{basis expansion} of a covariate.
Writing \(b_j(x_t)\) as the \(j\)th basis function of \(x_t\), the
smooth \(f(x_t)\) can be represented as a weighted sum of basis
functions

\begin{equation*}
f(x_t) = \sum_{j = 1}^{k} b_j(x_t) \beta_j \,,
\end{equation*}

where \(\beta_j\) is the weight applied to the \(j\)th basis function.
Note that here we concern ourselves only with univariate smooths
involving a single covariate, the time variable \(x_t\).

The polynomial model we encountered earlier is an example of a
statistical model that uses a basis expansion. For the cubic polynomial
(\(P = 3\)) fit shown in Figure \ref{fig:polynomial-example-plot} there
are in fact 4 basis functions: \(b_1(x_t) = x_t^0 = 1\),
\(b_2(x_t) = x_t\), \(b_3(x_t) = x^2_t\), and \(b_4(x_t) = x_t^3\). Note
that \(b_1(x_t)\) is constant and is linked to the model intercept,
\(\beta_0\), in the linear model \eqref{eq:polynomial-model}, and
further, that the weights are the estimated coefficients in the model,
the \(\beta_j\).

As we have already seen, polynomial basis expansions do not necessarily
lead to well-fitting models unless the true function \(f\) is itself a
polynomial. One of the primary criticisms is that polynomial basis
functions are global (Magee, 1998); the value of \(f\) at time point
\(x_t\) affects the value of \(f\) at time point \(x_{t+s}\) even if the
two time points are at opposite ends of the series. There are many other
bases we could use; here I discuss one such set of bases, that of
splines.

There are a bewildering array of different types of spline. In the
models discussed below we will largely restrict ourselves to cubic
regression splines (CRS) and thin plate regression splines (TPRS). In
addition, I also discuss two special types of spline basis, an adaptive
spline basis and a Gaussian process spline basis.

A cubic spline is a smooth curve comprised of sections of cubic
polynomials (\(P = 3\)), where the sections are joined together at some
specified locations --- known as \emph{knots} --- in such a way that at
the joins, the two sections of cubic polynomial that meet have the same
value as well as the same first and second derivative. These properties
mean that the sections join smoothly and differentiably at the knots
(Wood, 2017, 5.3.1).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/basis-function-example-plot-1}

}

\caption{Basis functions for the time covariate and the Small Water $\delta^{15}\text{N}$ time series. A rank (size) 7 cubic regression spline (CRS) basis expansion is show in a), with knots, indicated by tick marks on the x-axis, spread evenly through the rang of the data. b) shows the same CRS basis functions weighted by the estimated coefficients $\beta_j$, plus the resulting GAM trend line (black line drawn through the data). The grey points in both panels are the observed $\delta^{15}\text{N}$ values. c) A rank 7 thin plate regression spline basis for the same data.}\label{fig:basis-function-example-plot}
\end{figure}

The CRS can be parameterized in a number of different ways. One requires
a knot at each unique data value in \(x_t\), which is computationally
inefficient. Another way of specifying a CRS basis is to parameterize in
terms of the value of the spline at the knots. Typically in this
parametrization there are many fewer knots than unique data, with the
knots distributed evenly over the range of \(x_t\) or at the quantiles
of \(x_t\). Placing knots at the quantiles of \(x_t\) has the effect of
placing a greater number of knots where the data is densest.

A CRS basis expansion comprised of 7 basis functions for the time
covariate in the Small Water series, is shown in Figure
\ref{fig:basis-function-example-plot}a. The tick marks on the x-axis
show the locations of the knots, which are located at the ends of the
series and evenly in between. Notice that in this particular
parametrization, the \(j\)th basis function takes a value of 1 at the
\(j\)th knot and at all other knots a value of 0.

To estimate a model using this basis expansion each basis function forms
a column in the model matrix \(\boldsymbol{X}\) and the weights
\(\beta_j\) can be found using least squares regression (assuming a
Gaussian response). Note that in order to estimate a coefficient for
each basis function the model has to be fitted without an intercept
term. In practice we would include an intercept term in the model and
therefore the basis functions are modified via an identifiability
constraint. This has the effect of making the basis orthogonal to the
intercept but results in more complicated basis functions than those
shown in in Figure \ref{fig:basis-function-example-plot}a.

Having estimated the weight for each basis function, the \(j\)th basis
function \(b_j\) is scaled (weighted) by its coefficient \(\beta_j\).
The scaled CRS basis functions for the Small Water time series are shown
in Figure \ref{fig:basis-function-example-plot}b. The solid line passing
through the data points is formed by summing up the values of the scaled
basis functions (\(b_j(x_t) \beta_j\)) at any value of \(x_t\) (time).

Cubic regression splines, as well as many other types of spline, require
the analyst to choose the number and location of the knots that
parametrise the basis. Thin plate regression splines (TPRS) remove this
element of subjectivity when fitting GAMs. Thin plate splines were
introduced by Duchon (1977) and, as well as solving the knot selection
problem, have several additional attractive properties in terms of
optimality and their ability to estimate a smooth function of two or
more variables, leading to smooth interactions between covariates.
However, thin plate splines have one key disadvantage over CRS; thin
plate splines have as many unknown parameters as there are unique
combinations of covariate values in a data set (Wood, 2017, 5.5.1). It
is unlikely in the extreme that any real data problem would involve
functions of such complexity that they require as many basis functions
as data. It is much more likely that the true functions that we attempt
to estimate are far simpler than the set of functions representable by 1
basis function per unique data value. From a practical point of view, it
is also highly inefficient to carry around all these basis functions
whilst model fitting, and the available computational resources would
become quickly exhausted for large time series with many observations.

To address this issue, thin plate regression splines (TPRS) have been
suggested which truncate the space of the thin plate spline basis to
some lower number of basis functions whilst preserving much of the
advantage of the original basis as an optimally-fitting spline (Wood,
2003). A rank 7 TPRS basis (i.e.~one containing 7 basis functions) is
shown in Figure \ref{fig:basis-function-example-plot}c for the Small
Water time series. The truncation is achieved by performing an
eigen-decomposition of the basis functions and retaining the
eigenvectors associated with the \(k\) largest eigenvalues. This is
similar to the way principal components analysis decomposes a data set
into axes of variation (eigenvectors) in decreasing order of variance
explained. The truncated basis can preserve much of the space of
functions spanned by the original basis but at the cost of using far
fewer basis functions (Wood, 2003, 2017, 5.5.1). Note the horizontal
TPRS basis function (at δ\textsuperscript{15}N = 1) in Figure
\ref{fig:basis-function-example-plot}c; this basis function is
confounded with the intercept term and, after the application of
identifiability constraints, ends up being removed from the set of basis
functions used to fit the model.

The truncation suggested by Wood (2003) is not without cost; the
eigen-decomposition and related steps can be relatively costly for large
data sets. For data sets of similar size to the two examples used here,
the additional computational effort required to set up the TPRS basis
over the CRS basis will not be noticeable. For highly resolved series
containing more than \textasciitilde{}1000 observations the truncation
may be too costly computationally. In such instances, little is lost by
moving to the CRS basis, with the same number of knots as the rank of
the desired TPRS, with the benefit of considerably reduced set up time
for the basis.

To fit a GAM using either of the two regression spline bases described
above, the analyst is generally only required to the specify the size
(rank) of the basis expansion required to represent or closely
approximate the true function \(f\). With practice and some knowledge of
the system from which the observations arise, it can be relatively easy
to put an upper limit on the expected complexity of the true trend in
the data. Additionally, the number available data points places a
constraint on the upper limit of the size of basis expansion that can be
used.

In practice, the size of the basis is an upper limit on the expected
complexity of the trend, and a simple test is available to check if the
basis used was sufficiently large (Pya and Wood, 2016). This test is
available via the \texttt{gam.check()} function in \emph{mgcv} for
example, which essentially looks at whether there is any additional
nonlinearity or structure in the residuals that can be explained by a
further smooth of \(x_t\). Should a smooth term in the fitted model fail
this test the model can be refitted using a larger basis expansion, say
by doubling the value of \emph{k} (the rank) used to fit the original.
Note also that a smooth might fail this test whilst using fewer
effective degrees of freedom than the maximum possible for the dimension
of basis used. This may happen when the true function lies at the upper
limit of the set of functions encompassed by the size of basis used.
Additionally, a basis of size \(2k\) encompasses a richer space of
functions of a given complexity than a basis of size \(k\) (Wood, 2017);
increasing the basis dimension used to fit the model may unlock this
additional function space resulting in a better fitting model whilst
using a similar number of effective degrees of freedom.

\subsection{Smoothness selection}\label{smoothness-selection}

Having identified regression splines as a useful way to represent \(f\),
we next need a way to decide how wiggly the fitted trend should be. A
backwards elimination approach to sequentially remove knots or basis
functions might seem appropriate, however such an approach would likely
fail as the resulting sequence of models would not be strictly nested,
precluding many forms of statistical comparison (Wood, 2017).
Alternatively, we could keep the basis dimension at a fixed size but
guard against fitting very complex models through the use of a
wiggliness penalty.

The default wiggliness penalty used in GAMs is on the second derivative
of the spline, which measures the rate of change of the slope, or the
curvature, of the spline at any infinitesimal point in the interval
spanned by \(x_t\). The actual penalty used is the integrated squared
second derivative of the spline

\begin{equation} \label{eq:quadratic-penalty}
\int_{\mathbb{R}} [f^{\prime\prime}]^2 dx = \boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}\, .
\end{equation}

The right hand side of \eqref{eq:quadratic-penalty} is the penalty in
quadratic form. The convenience of the quadratic form is that it is a
function of the estimated coefficients of \(f(x_t)\) where
\(\mathbf{S}\) is known as the penalty matrix. Notice that now both the
weights for the basis functions and the wiggliness penalty are expressed
as functions of the model coefficients.

Now that we have a convenient way to measure wiggliness, it needs to be
incorporated into the objective function that will be minimised to fit
the GAM. The likelihood of the model given the parameter estimates
\(\mathcal{L}(\boldsymbol{\beta})\) is combined with the penalty to
create the penalized likelihood \(\mathcal{L}_p(\boldsymbol{\beta})\):

\begin{equation*}
\mathcal{L}_p(\boldsymbol{\beta}) = \mathcal{L}(\boldsymbol{\beta}) - \frac{1}{2} \lambda\boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}\, .
\end{equation*}

The fraction of a half is there simply to make the penalised likelihood
equal the penalised sum of squares in the case of a Gaussian model.
\(\lambda\) is known as the smoothness parameter and controls the extent
to which the penalty contributes to the likelihood of the model. In the
extreme case of \(\lambda = 0\) the penalty has no effect and the
penalized likelihood equals the likelihood of the model given the
parameters. At the other extreme, as \(\lambda \rightarrow \infty\) the
penalty comes to dominate \(\mathcal{L}_p(\boldsymbol{\beta})\) and the
wiggliness of \(f(x_t)\) tends to \(0\) resulting in an infinitely
smooth function. In the case of a second derivative penalty, this is a
straight line, and we recover the simple linear trend from
\eqref{eq:linear-model} when assuming a Gaussian response.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/penalty-example-plot-1}

}

\caption{The effect of the smoothness parameter $\lambda$ on the resulting wiggliness of the estimated spline. Large values of $\lambda$ penalize wiggliness strongly, resulting in smooth trends (upper row), while smaller values allow increasingly wiggly trends. The aim of automatic smoothness selection is to find an optimal value of $\lambda$ that balances the fit of the model with model complexity to avoid overfitting.}\label{fig:penalty-example-plot}
\end{figure}

Figure \ref{fig:penalty-example-plot} illustrates how the smoothness
parameter \(\lambda\) controls the degree of wiggliness in the fitted
spline. Four models are shown, each fitted with a fixed value of
\(\lambda\); 10000, 1, 0.01, and 0.00001. At \(\lambda = 10000\) the
model effectively fits a linear model through the data. As the value of
\(\lambda\) decreases, the fitted spline becomes increasingly wiggly. As
\(\lambda\) becomes very small, the resulting spline passes through most
of the δ\textsuperscript{15}N observations resulting in a model that is
clearly over fitted to the data.

To fully automate smoothness selection for \(f(x_t)\) we need to
estimate \(\lambda\). There are two main ways that \(\lambda\) can be
automatically chosen during model fitting. The first way is to choose
\(\lambda\) such that it minimises the prediction error of the model.
This can be achieved by choosing \(\lambda\) to minimise Akaike's
information criterion (AIC) or via cross-validation (CV) or generalized
cross-validation (GCV; Craven and Wahba, 1978). GCV avoids the
computational overhead inherent to CV of having to repeatedly refit the
model with one or more observations left out as a test set. Minimising
the GCV score will, with a sufficiently large data set, find a model
with the minimal prediction error (Wood, 2017). The second approach is
to treat the smooth as a random effect, in which \(\lambda\) is now a
variance parameter to be estimated using maximum likelihood (ML) or
restricted maximum likelihood (REML; Wood, 2011; Wood et al., 2016).

Several recent results have shown that GCV, under certain circumstances,
has a tendency to under smooth, resulting in fitted splines that are
overly wiggly (Reiss and Ogden, 2009). Much better behaviour has been
observed for REML and ML smoothness selection, in that order (Wood,
2011). REML is therefore the recommended means of fitting GAMs, though,
where models have different fixed effects (covariates) they cannot be
compared using REML, and ML selection should be used instead. In the
sorts of data examples considered here there is only a single covariate
\(x_t\) as our models contain a single estimated trend so REML
smoothness selection is used throughout unless otherwise stated.

\section{Fitting GAMs}\label{fitting-gams}

\subsection{Small Water}\label{small-water}

The trend in δ\textsuperscript{15}N values is clearly non-linear and it
would be difficult to suggest a suitable polynomial model that would
allow for periods of relatively no change in δ\textsuperscript{15}N as
well as rapid change. Instead, a GAM is ideally suited to modelling such
trends; the data suggest a smoothly varying change in
δ\textsuperscript{15}N between 1925 and 1975. It is reasonable to expect
some autocorrelation in the model errors about the fitted trend.
Therefore I fitted the following GAM to the δ\textsuperscript{15}N time
series.

\begin{equation} \label{eq:small-gam}
y_t = \beta_0 + f(x_t) + \varepsilon, \quad \varepsilon_t \sim \mathcal(0, \boldsymbol{\Lambda}\sigma^2)
\end{equation}

Note that now I have relaxed the i.i.d. assumption and have introduced
\(\boldsymbol{\Lambda}\), a correlation matrix that is used to model
autocorrelation in the residuals. The δ\textsuperscript{15}N values are
irregularly spaced in time and a correlation structure that can handle
the uneven spacing of the samples is needed (Pinheiro and Bates, 2000).
A continuous time first-order autoregressive process (CAR(1)) is a
reasonable choice; it is the continuous-time equivalent of the
first-order autoregressive process (AR(1)) and, simply stated, models
the correlation between any two residuals as and exponentially
decreasing function of \(h\) (\(\phi^h\)), where \(h\) is the amount of
separation in time between the residuals (Pinheiro and Bates, 2000).
\(h\) may be a real valued number in the CAR(1), which is how it can
accommodate the irregular separation of samples in time. \(\phi\)
controls how quickly the correlation between any two residuals declines
as a function of their separation in time and is an additional parameter
that will be estimated during model fitting. The model in
\eqref{eq:small-gam} was fitted using the \texttt{gamm()} function
(Wood, 2004) in the \emph{mgcv} package (Wood, 2017) for R (R Core Team,
2017).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/plot-fitted-models-1}

}

\caption{GAM-based trends fitted to the Small Water $\delta^{15}\text{N}$ (a) and Braya-Sø \uk{} (b) time series. The shaded bands surrounding the estimated trends are approximate 95\% across-the-function confidence intervals. For the \uk{} series, two models are shown; the orange fit is the result of a GAM with a continuous-time AR(1) process estimated using REML smoothness selection, while the blue fit is that of a simple GAM with GCV-based smoothness selection. The REML-based fit significantly oversmooths the \uk{} time series.}\label{fig:plot-fitted-models}
\end{figure}

The fitted trend is shown in Figure \ref{fig:plot-fitted-models}a, and
well-captures the strong pattern in the data. The trend is statistically
significant (estimated degrees of freedom = 7.95; \(F\) = 47.44,
approximate \(p\) value = \(\ll\) 0.0001). However further analysis of
the fitted model is required to answer the other questions posed earlier
about the timing of change and whether features in the trend can be
distinguished from random noise. I discuss these issues shortly.

\subsection{Braya-Sø}\label{braya-s}

The \uk{} data present a more difficult data analysis challenge than the
δ\textsuperscript{15}N time series because of the much more complex
variation present. Fitting the same model as the Small Water example,
\eqref{eq:small-gam}, to the \uk{} data resulted in the unsatisfactory
fit shown as the very smooth line in Figure
\ref{fig:plot-fitted-models}b (labelled GAMM (CAR(1))). Further problems
were evident with this model fit --- the covariance matrix of the model
was non-positive definite, a sure sign of problems with the fitted
model. Refitting with a smaller basis dimension (\emph{k} = 20) for the
trend term resulted in a model with a positive-definite covariance
matrix for the model variance-covariance terms, but the estimated value
of of the CAR(1) parameter \(\phi\) = 0.2 was exceedingly uncertain
(95\% confidence interval 0 -- 1!)

Fitting this model as a standard GAM with REML smoothness selection
resulted in the same fitted trend as the GAM with CAR(1) errors (not
shown), whilst using GCV smoothness selection resulted in a much more
satisfactory fitted trend. There are two potential problems with the
GCV-selected trend: i) GCV is sensitive to the profile of the GCV score
and has been shown to badly under smooth data in situations where the
profile is flat around the minimum GCV score, and ii) the model fitted
assumes that the observations are independent, an assumption that is
certainly violated in the \uk{} time series.

To investigate the first issue, the GCV and REML scores for an
increasing sequence of values of the smoothness parameter (\(\lambda\))
were evaluated for the standard GAM (equation \eqref{eq:gam}) fit to the
\uk{} time series. The resulting profiles are shown in Figure
\ref{fig:trace-smoothness-parameters}, with the optimal value of the
parameter shown by the vertical line. The GCV score profile suggests
that the potential for under smoothing identified by Reiss and Ogden
(2009) is unlikely to apply here as there is a well-defined minimum in
profile. Addressing the violation of the independence assumption is more
difficult; the standard errors of the model coefficients are most likely
anti-conservative, which presents a significant problem if we wish to
determine which of the wiggles in the fitted \uk{} trend can be
identified as real.

To understand the reason why the GAM plus CAR(1) and the simple GAM with
REML smoothness selection performed poorly with the \uk{} time series we
need to delve a little deeper into what is happening when we are fitting
these two models.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/trace-smoothness-parameters-1}

}

\caption{GCV and REML scores as a function of the smoothness parameter $\lambda$. From left to right, GAMs were estimated using GCV and REML smoothness selection, and REML using a basis dimension of 40 and observational weights to account for heterogeneity in the \uk{} times series. The selected value of $\lambda$ for each model is indicated by the vertical grey line.}\label{fig:trace-smoothness-parameters}
\end{figure}

The primary issue leading to poor fit is that neither model accounts for
the different variance (known as (heteroscedasticity) of each
observation in the \uk{} record. The sediments in Braya-Sø are not
annually laminated and therefore the core was sliced at regular depth
intervals. Owing to compaction of older sediments and variation in
accumulation rates over time, each sediment slice represents a different
number of ``lake years''. We can think of older samples as representing
some average of many years of sediment deposition, whilst younger
samples are representative of fewer of these lake years. The average of
a larger set of numbers is estimated more precisely than the average of
a smaller set, all things equal. A direct result of this variable
averaging of lake years it that some samples are more precise and
therefore have lower variance than other samples and yet the model
assumed that the variance was constant across samples.

Accounting for heteroscedasticity within the model is relatively simple
via the use of observational weights. The number of lake years
represented by each slice is estimated by assigning a date to the top
and bottom of each sediment slice. The variance of each observation
should be proportional to the inverse of the number of lake years each
sample represents. In the \texttt{gam()} function used here, weights
should be specified as the number of lake years each sample represents.
Other software may require the weights to be specified in a different
way.

A secondary problem is the size of the basis dimension used for the time
variable. The main user selectable option when fitting a GAM in the
penalised likelihood framework of Wood (2004) is how many basis
functions to use. As described above, the basis should be large enough
to contain the true, but unknown, function, or a close approximation to
it. For GCV selection the basis used contained 29 basis functions,
whilst the CAR(1) model with REML smoothness selection would only
converge with a basis containing 20 functions. The size of the basis
appears to be sufficient for GCV smoothness selection, but following
Wood (2011) REML smoothness selection is preferred. Using the test of
Pya and Wood (2016), the basis dimension for the models with REML
smoothness selection was too small. To proceed therefore, we must drop
the CAR(1) term and increase the basis dimension to 39 functions (by
setting \texttt{k\ =\ 40}; I set it larger than expected wiggliness
because the larger basis contains a richer family of functions and the
excess complexity is reduced because of the smoothness penalty in play.)

With the larger basis dimension and accounting for the non-constant
variance of the observations via weights, the model fitted using REML is
indistinguishable from that obtained using GCV (Figure
\ref{fig:plot-fitted-models}b). The trace of the REML score for this
model (Figure \ref{fig:trace-smoothness-parameters}c) shows a pronounced
minimum at a much smaller value of \(\lambda\) than the original REML
fit (Figure \ref{fig:trace-smoothness-parameters}b), indicating that a
more wiggly trend provides a better fit to the Braya-Sø time series.
This example illustrates that some care and understanding of the
underlying principles of GAMs is required to diagnose potential issues
with the estimated model. After standard modelling choices (size of
basis to use, correct selection of response distribution and link
function, etc.) are checked, it often pays to think carefully about the
properties of the data and ensure that the assumptions of the model are
met. Here, despite increasing the basis size, it was the failure to
appreciate the magnitude of the effect on fitting of the non-constant
variance that lead to the initially poor fit and the problems associated
with the estimation of the CAR(1) process. I return to the issue of why
the GAM plus CAR(1) model encountered problems during fitting later
(\protect\hyperlink{identifiability}{see Identifiability}).

\hypertarget{confints}{\subsection{Confidence intervals and uncertainty
estimation}\label{confints}}

If we want to ask whether either of the estimated trends is
statistically interesting or proceed to identifying periods of
significant change, we must address the issue of uncertainty in the
estimated model. What uncertainty is associated with the trend
estimates? One way to visualise this is through a 1 - \(\alpha\)
confidence interval around the fitted trend, where \(\alpha\) is say
0.05 leading to a 95\% interval. A 95\% interval would be drawn at
\(\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))\), with
\(m_{1-\alpha} = 1.96\), the 0.95 probability quantile of a standard
normal distribution, and \(\text{SE}(\hat{y}_t)\) is the standard error
of the estimated trend at time \(x_t\). This type of confidence interval
would normally be described as \emph{pointwise}; the coverage properties
of the interval are correct for a single point on the fitted trend, but,
if we were to consider additional points on the trend, the coverage
would logically be lower than 1 - \(\alpha\). This is the traditional
frequentist interpretation of a confidence interval. However, the
approach to fitting GAMs described here has a Bayesian interpretation
(Kimeldorf and Wahba, 1970; Silverman, 1985; Wahba, 1983, 1990) and
therefore the typical frequentist interpretation does not apply. Nychka
(1988) investigated the properties of a confidence interval created as
described above using standard errors derived from the Bayesian
posterior covariance matrix for the estimated mode parameters. Such
intervals have the interesting property that they have good
\emph{across-the-function} coverage when considered from a frequentist
perspective. This means that, when averaged over the range of the
function, the Bayesian credible intervals shown in Figure
\ref{fig:plot-fitted-models} have close to the expected 95\% coverage.
However, to achieve this some parts of the function may have more or
less than 95\%-coverage. Marra and Wood (2012) recently explained
Nychka's (1988) surprising results and extended them to the case of
generalized models (non-Gaussian responses).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/posterior-simulation-1}

}

\caption{Estimated trends (thick black lines) and 20 random draws (grey lines) from the posterior distribution of the GAM fitted to the Small Water $\delta^{15}\text{N}$ (a) and Braya-Sø \uk{} (b) time series.}\label{fig:posterior-simulation}
\end{figure}

Whilst the \emph{across-the-function} or average coverage frequentist
interpretation of the Bayesian credible intervals is useful, if may be
important to have an interval that contains the entirety of the true
function with some state probability (1 - \(\alpha\)). Such an interval
is known as a \emph{simultaneous} interval. A (1 - \(\alpha\))100\%
\emph{simultaneous} confidence interval contains \emph{in their
entirety} 1 - \(\alpha\) of all random draws from the posterior
distribution of the fitted model.

Fitting a GAM involves finding estimates for coefficients of the basis
functions. Together, these coefficients are distributed multivariate
normal with mean vector and covariance matrix specified by the model
estimates of the coefficients and their covariances respectively. Random
draws from this distribution can be taken, where each random draw
represents a new trend that is consistent with the fitted trend but also
includes the uncertainty in the estimated trend. This process is known
as \emph{posterior simulation}.

Figure \ref{fig:posterior-simulation} shows 20 random draws from the
posterior distribution of the GAMs fitted to the Small Water and
Braya-Sø data sets. In the early period of the δ\textsuperscript{15}N
time series many of the posterior simulations exhibit short periods of
increasing and decreasing trend, balancing out to the relatively flat
trend estimated by the GAM (Fig.~\ref{fig:posterior-simulation}a).
Reflecting this uncertainty, we might expect relatively wide
simultaneous intervals during this period in order to contain the vast
majority of the simulated trends. Conversely, the decreasing
δ\textsuperscript{15}N trend starting at \textasciitilde{}1945 is
consistently reproduced in the posterior simulations, suggesting that
this feature of the time series is both real and statistically
significant, and that the rate of change in δ\textsuperscript{15}N is
relatively precisely estimated. We see a similar pattern in
Figure~\ref{fig:posterior-simulation}b for the Braya-Sø record; the
large peak in \uk{} at \textasciitilde{}250CE and the strong decline at
\textasciitilde{}1200CE are well defined in the posterior simulations,
whereas most of the localised trends that are smaller magnitude changes
in \(y_t\) are associated with posterior simulations that are less well
constrained with the ends of the record in particular showing
considerable variation in the strength, timing, and even sign of
simulated trends, reflecting the greater uncertainty in estimated trend
during these periods. For the random draws illustrated in Figure
\ref{fig:posterior-simulation}, a simultaneous interval should contain
the entire function for on average 19 of the 20 draws.

There are a number of ways in which a simultaneous interval can be
computed. Here I follow the simulation approach described by Ruppert et
al. (2003) and present only the basic detail; a fuller description is
contained in Appendix \protect\hyperlink{appendix}{1}. The general idea
is that if we want to create an interval that contains the whole of the
true function with 1 - \(\alpha\) probability, we need to increase the
standard Bayesian credible interval by some amount. We could simulate a
large number of functions from the posterior distribution of the model
and then search for the value of \(m_{1-\alpha}\) that when multiplied
by \(\text{SE}(\hat{f}(x_t))\) yielded an interval that contained the
whole function for \((1-\alpha)\) 100\% of the functions simulated. In
practice, the simulation method of Ruppert et al. (2003) does not
involve a direct search, but yields the critical value \(m_{1-\alpha}\)
required.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/compare-intervals-1}

}

\caption{95\% simultaneous confidence intervals (light grey bands) and across-the-function confidence intervals (dark grey bands) on the estimated trends (black lines) for the Small Water $\delta^{15}\text{N}$ (a) and Braya-Sø \uk{} (b) time series.}\label{fig:compare-intervals}
\end{figure}

Simultaneous intervals computed using the method described are show in
Figure \ref{fig:compare-intervals} alongside the
\emph{across-the-function} confidence intervals for the trends fitted to
both example data sets. As expected, the simultaneous interval is
somewhat wider than the \emph{across-the-function} interval. The
critical value \(m_{1-\alpha}\) for the simultaneous interval of the
estimated trend in δ\textsuperscript{15}N is 3.08, whilst the same value
for the \uk{} series is 3.42, leading to intervals that are
approximately ±50\% and ±75\% wider than the equivalent
across-the-function intervals.

\subsection{Identifying periods
change}\label{identifying-periods-change}

In the simple linear trend model \eqref{eq:linear-model} whether the
estimated trend constitutes evidence for or against a null hypothesis of
no change rests on how large the estimated rate of change in \(y_t\) is
(\(\hat{\beta}_1\)) relative to its uncertainty. This is summarised in
the \(t\) statistic. As the rate of change in \(y_t\) is constant over
the fitted trend --- there is only a singe slope for the fitted trend
\(\hat{\beta}_1\) --- if the \(t\) statistic of the test that
\(\hat{\beta}_1 = 0\) is unusually extreme this would be evidence
against the null hypothesis of no change. Importantly, this applies to
the whole time series as the linear model implies a constant rate of
change throughout. More formally, the estimate \(\hat{\beta}_1\) is the
first derivative of the fitted trend.

In the GAM, the fitted trend need not be linear; the slope of the trend
is potentially different at every point in the time series. As such we
might reasonably ask \emph{where} in the series the response \(y_t\) is
changing, if at all? Mirroring the linear model we can answer this
question by determining whether or not the first derivative at any time
point \(x_t\) of the fitted trend at any time point is consistent with a
null hypothesis of no change. We want to know whether or not the first
derivative is indistinguishable from a value of \(0\) --- no trend ---
given the uncertainty in the estimate of the derivative.

Derivatives of the fitted spline are not easily available analytically,
but they can be estimated using the method of finite differences. Two
values of the estimated trend, separated by a very small time-shift
(\(\Delta_t\)), are predicted from the model; the difference between the
estimated values for the two time points is an approximation of the true
first derivative of the trend. As \(\Delta_t \rightarrow 0\) the
approximation becomes increasingly accurate. In practice, the first
derivative of the fitted trend is evaluated using finite differences at
a large number of points in the time series. An approximate (1 -
\(\alpha\))100\% pointwise confidence interval can be calculated for the
derivative estimates using standard theory
(i.e.~\(\pm 1.96 \times \text{SE}(\hat{y}_t)\) for a 85\% interval) and
the covariance matrix of the spline coefficients. A (1 -
\(\alpha\))100\% simultaneous interval for the derivatives can also be
computed using the method described
\protect\hyperlink{confints}{earlier}. Periods of significant change are
identified as those time points where the (simultaneous) confidence
interval on the first derivative does not include zero.

Figure \ref{fig:derivatives} shows the estimated first derivative of the
fitted trend in the Small Water (\ref{fig:derivatives}a) and Braya-Sø
(\ref{fig:derivatives}b) time series. Although the estimated trend
suggests a slight increase in δ\textsuperscript{15}N from the start of
the record to \textasciitilde{}1940, the estimated trend is sufficiently
uncertain that the simultaneous interval on the first derivative
includes 0 throughout this period. We can understand why this is so by
looking at the posterior simulations in
Figure~\ref{fig:posterior-simulation}a; there is considerable variation
in the shape of the simulated trends throughout this period. From
\textasciitilde{}1925 the derivative of the trend becomes negative,
however it is not until \textasciitilde{}1940 that the simultaneous
interval doesn't include \(0\). At this point we have evidence to reject
the null hypothesis of no change. This time point may be taken as the
first evidence for change in δ\textsuperscript{15}N in the Small Water
core. The simultaneous interval on the first derivative of the trend in
δ\textsuperscript{15}N is bounded away from \(0\) between
\textasciitilde{}1940 and \textasciitilde{}1975, covering the major
decline in values evident in the observations. The simultaneous interval
includes \(0\) from \textasciitilde{}1975 onward, suggesting that,
whilst quite pronounced, the recent increase in δ\textsuperscript{15}N
is not statistically significant. To determine whether or not the recent
increase is real, we would require considerably more samples with which
to (hopefully) more-precisely estimate the trend during this period.
Alternatively, we might just have to wait until sufficient additional
sedimentation has occurred to warrant recoring Small Water and
reestimating the trend in δ\textsuperscript{15}N.

The estimated trend at Braya-Sø exhibited a number of oscillations in
\uk{}. As we saw previously in Figures~\ref{fig:posterior-simulation}b
and \ref{fig:compare-intervals}b, many of these are subject to
significant uncertainty and it is important therefore to discern which,
if any, of the oscillations in the response can be identified given the
model uncertainty. In Figure~\ref{fig:derivatives}b only two features of
the estimated trend are considered significant based on the derivatives
of the smooth; one centred on \textasciitilde{}250CE and a second at
\textasciitilde{}1150CE. In both these periods, the simultaneous
interval for the first derivative of the trend does not include zero. In
the first case we detect the large peak and subsequent decline in \uk{}
at \textasciitilde{}250CE, whilst at \textasciitilde{}1150CE the large
trough is identified, but not the increasing trend immediately prior to
this excursion to lower \uk. Recall that these intervals are
simultaneous in nature, strongly guarding against false positives, and
as such we can be confident in the estimation of these two features,
whilst care must be taken to not over-interpret the remaining variations
in the estimated trend.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/derivatives-1}

}

\caption{Estimated first derivatives (black lines) and 95\% simultaneous confidence intervals of the GAM trends fitted to the Small Water $\delta^{15}\text{N}$ (a) and Braya-Sø \uk{} (b) time series. Where the simultaneous interval does not include 0, the models detect significant temporal change in the response.}\label{fig:derivatives}
\end{figure}

\hypertarget{identifiability}{\subsection{Residual autocorrelation and
model identification}\label{identifiability}}

The GAM fitted to the δ\textsuperscript{15}N time series contained a
CAR(1) process to model residual temporal autocorrelation in the
residuals. The estimated magnitude of the autocorrelation is given by
the parameter \(\phi\). The estimated value of \(\phi\) for the
δ\textsuperscript{15}N series is 0.6 with 95\% confidence interval
0.28--0.85, indicating moderate to strong residual autocorrelation about
the fitted trend. The correlation function is an exponentially
decreasing function of temporal separation (\(\Delta_t\)), and whilst
observations that are a few year apart are quite strongly dependent on
one another, this dependence drops off rapidly as \(\Delta_t\) increases
and is effectively zero when samples are separated by a decade or more
(Figure~\ref{fig:car1-plot}).

Failure to account for the dependencies in the δ\textsuperscript{15}N
time series could lead to the estimation of a more wiggly trend than the
one shown in Figure \ref{fig:plot-fitted-models}a which would negatively
impact the confidence place on the inferences we might draw from the
fitted model. Importantly, failing to account for the strong dependency
in the residuals would lead to smaller uncertainties in the estimated
spline coefficients, which would propagate through to narrower
confidence intervals on the fitted trend and on the first derivatives,
and ultimately to the identification of significant periods of change.
The end result would be a tendency toward anti-conservative
identification of periods of change; the coverage probability would be
lower than the anticipated \(1 - \alpha\), leading to a greater chance
of false positive results.

\begin{figure}

{\centering \includegraphics[width=0.5\linewidth]{manuscript_files/figure-latex/car1-plot-1}

}

\caption{Estimated CAR(1) process from the GAM fitted to the Small Water $\delta^{15}\text{N}$ time series. $h(\Delta_t, \phi)$ is the correlation between residuals separated by $\Delta_t$ years, where $\hat{\phi} = \text{0.6}$. The shaded band is a 95\% pointwise confidence interval on the estimated correlation $h$.}\label{fig:car1-plot}
\end{figure}

Problems estimating the GAM plus CAR(1) model were encountered when this
was fitted to the \uk{} time series; both a smooth trend in the mean
\uk{} and a CAR(1) process in the residuals lead to an unidentifiable
model. What makes a model with a spline-based trend and an
autocorrelation process like the CAR(1) potentially unidentifiable?

Consider again the basic GAM for a smooth trend,
\eqref{eq:additive-model}. In that equation the correlation matrix
\(\boldsymbol{\Lambda}\) was omitted for the sake of simplicity. As I
did in \eqref{eq:small-gam}, I reintroduce it and restate the
distributional assumptions of this model

\begin{equation}
y_t = \beta_0 + f(x_t) + \varepsilon_t, \quad \varepsilon \sim \mathcal(0, \boldsymbol{\Lambda}\sigma^2)
\end{equation}

In the basic GAM, \(\boldsymbol{\Lambda} \equiv \mathbf{I}\) is an
identity matrix, a matrix with 1s on the diagonal and 0s elsewhere

\begin{equation*}
\begin{bmatrix}
1      & 0 & 0 & \dots & 0 \\
0      & 1 & 0 & \dots & 0 \\
0      & 0 & 1 & \dots & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
0      & 0 & 0 & \dots & 1
\end{bmatrix}\, ,
\end{equation*}

which is where the independence assumption of the model comes from; a
model residual is perfectly correlated with itself (the 1s on the
diagonal), but uncorrelated with any other residual (the off-diagonal
0s). In the GAM plus CAR(1) model, an alternative correlation function
for \(\boldsymbol{\Lambda}\) was used --- the CAR(1) with correlation
parameter \(\phi\). Fahrmeir and Kneib (2008) show that where the
stochastic structure of \(f\) and \(\boldsymbol{\Lambda}\) approach one
another, i.e.~where we have a potentially wiggly trend or strong
autocorrelation as \(\phi \rightarrow 1\), the two processes can quickly
become unidentifiable (see also Fahrmeir et al., 2013). By
unidentifiable, we mean that it becomes increasingly difficult to
distinguish between a wiggly trend or strong autocorrelation because
these two processes are very similar to one another in appearance. This
leads to model estimation problems of the sort encountered with fitting
the GAM plus CAR(1) model to the Braya-sø \uk{} series.

Why might this be so? Autocorrelation is the tendency for a large
(small) value of \(y_t\) at time \(x_t\) to be followed by a likewise
large (small) value at time \(x_{t+1}\). This leads to runs of values
that are consistently greater (less) than the overall mean. Short runs
would indicate weaker autocorrelation whilst longer runs are associated
with stronger autocorrelation, and long runs of values greater (less)
than the mean would evident as non-linear trends in the time series. As
a result, a wiggly trend and an autocorrelation function with large
\(\phi\) are two ways to describe the same pattern of values in a time
series, and without any further information to constrain either the
model is unable to distinguish both components uniquely.

Situations where it may be possible to uniquely identify separate wiggly
trends and autocorrelation are exemplified by the Small Water
δ\textsuperscript{15}N time series. The non-linear trend and the
autocorrelation operate at very different scales; the trend represents
decadal-scale variation in mean δ\textsuperscript{15}N, whilst the
CAR(1) process represents the much smaller-scale tendency for values of
the response to be followed in time by similar values. That such a
pattern is observed in the Small Water core is the result of the high
resolution of the sampling in time relative to the long-term trend. In
contrast, the Braya-Sø record is sampled at far lower resolution
relative to the fluctuations in the mean response, and consequently the
data do not contain sufficient information to separate trend and
autocorrelation.

\subsection{Gaussian process smooths}\label{gaussian-process-smooths}

In the world of machine learning, Gaussian processes (Golding and Purse,
2016; Rasmussen and Williams, 2006) are a widely-used method for fitting
smooth non-parametric regression models. A Gaussian process is a
distribution over all possible smooth functions \(f(x)\). In the field
of spatial statistics, Gaussian processes are known by name
\emph{kriging}.

With a Gaussian process we are interested in fitting a smooth temporal
trend by modelling the way the correlation between pairs of observations
varies as a function of the distance, \(h\), in time that separates
(\(h\)) the observations. The correlation between pairs of observations
decreases with increasing separation, which is modelled using a
correlation function, \(c(h)\). Figure
\ref{fig:gp-correlation-functions} shows examples of two different
correlation functions; the \emph{power exponential} (Figure
\ref{fig:gp-correlation-functions}a), and the Matérn (Figure
\ref{fig:gp-correlation-functions}b) correlation functions. These
functions are smooth and monotonic-decreasing, meaning that the value of
the correlation function decreases with increasing separation (\(h\)).
When \(h\) = 0, the correlation is equal to 1 (\(c(0) = 1\)); two
samples taken at exactly the same time point are perfectly correlated.
As \(h \rightarrow \infty\), the correlation tends to zero
(\(c(h) \rightarrow 0\)); two samples separated by a large amount of
time tend to be uncorrelated. Often we are interested in learning how
large the separation in time needs to be before, on average, a pair of
observations is effectively uncorrelated (i.e.~where \(c(h)\) is
sufficiently close to zero).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/gp-correlation-functions-1}

}

\caption{Power exponential (a) and Matérn (b) correlation functions for observation separation distance $h$. Two values of the effective range parameter ($\phi$)) are shown for each function. For the power exponential function, $\kappa$ is the power in the power exponential function. For the Matérn correlation function, $\kappa$ distinguishes the member of the Matérn family.}\label{fig:gp-correlation-functions}
\end{figure}

Several functions can be used to represent \(c(h)\). Two common ones are
the power exponential function and the Matérn family of correlation
functions. The power exponential function at separation distance \(h\)
is \[c(h) = \exp\{(-h/\phi)^{\kappa}\}\] where \(0 < \kappa \leq 2\).
The Matérn correlation function is actually a family of functions with
closed-forms only available for a subset of the family, distinguished by
\(\kappa\). When \(\kappa = 1.5\), the Matérn correlation function is
\[c(h) = (1 + h/\phi) \exp(-h/\phi)\] whilst for \(\kappa = 2.5\) it is
\[c(h) = \{1 + h/\phi + (h/\phi)^2/3\} \exp(-h/\phi)\] and for
\(\kappa = 3.5\)
\[c(h) = \{1 + h/\phi + 2(h/\phi)^2/5 + (h/\phi)^3/15\} \exp(-h/\phi)\, .\]

In all cases, \(\phi\) is the effective range parameter, which sets the
distance beyond which the correlation function is effectively zero.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/gp-gam-detail-plot-1}

}

\caption{Gaussian process smooths fitted to the \uk{} time series. REML score traces for GAMs fitted using power exponential ($\kappa = 1$) or Matérn ($\kappa = 1.5$) correlation functions as a function of the effective range parameter ($\phi$) are shown (a). The optimal model for each function is that with the lowest REML score. b) shows the resulting trends estimated using the respective correlation function with the value of $\phi$ set to the optimal value.}\label{fig:gp-gam-detail-plot}
\end{figure}

Gaussian processes and GAMs share many similarities and we can fit a
Gaussian process using the techniques already described above for
splines (Handcock et al., 1994; Kammann and Wand, 2003). It can be shown
(e.g. Fahrmeir et al., 2013 section/pages) that the Gaussian process
model has the same penalised likelihood form as the GAM that we
discussed earlier; the observations are the knots of the smoother and
each has a basis function in the form of a correlation function. The
equivalence is only true if the basis functions do not depend on any
other parameters of the model, which is only achievable if the value of
\(\phi\) is fixed and known (Fahrmeir et al., 2013). In general,
however, we would like to estimate \(\phi\) as part of model fitting. To
achieve this we can maximise the profile likelihood or score statistic
of the model over a range of values of \(\phi\) (Wood, 2017, 362--363).
This involves proposing a value of \(\phi\) for the effective range of
the correlation function and then estimating the resulting GAM by
minimising the penalised log-likehood conditional upon this value of
\(\phi\). The model, and its corresponding value of \(\phi\), with
lowest penalised log-likelihood or score statistic is then retained as
the estimated GAM. Figure~\ref{fig:gp-gam-detail-plot}a shows the REML
score for models estimated using a Gaussian process smooth with a Matérn
correlation function (\(\kappa\) = 1.5) for a sequence of values of
\(\phi\) between 15 and 1000 years. There is a clear minimum around 40
years separation, with the minimum REML score being observed at
\(\phi = 41.81)\). Also shown are the REML scores for models using the
power exponential function (\(\kappa\) = 1) with the minimum score
observed at a somewhat higher effective range of \(\phi = 71.06\).

Figure \ref{fig:gp-gam-detail-plot}b shows the estimated trends for the
\uk{} time series using Gaussian process smooths with exponential and
Matérn correlations functions, both using \(\phi\) values at their
respective optimal value as assessed using the REML score. The estimated
trends are very similar to one another, although there is a noticeable
difference in behaviour, with the power exponential (\(\kappa = 1\))
version being noticeably less-smooth than the Matérn version. This
difference is attributable to the shapes of the respective correlation
functions; the Matérn approaches a correlation of 1 smoothly as \(h\)
approaches 0, whilst the power exponential with \(\kappa\) = 1
approaches a correlation of 1 increasingly quickly with decreasing
\(h\). The power exponential with \(\kappa\) = 2, like the Matérn,
approaches \(\phi\) = 1 smoothly, and consequently the trend estimated
using this correlation function is qualitatively similar to that
estimated using the Matérn correlation function.

\subsection{Adaptive smoothing}\label{adaptive-smoothing}

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/braya-so-model-comparisons-1}

}

\caption{Comparison of trends estimated using i) adaptive smooth, ii) Gaussian process, and iii) thin plate regression spline bases for the \uk{} time series.}\label{fig:braya-so-model-comparisons}
\end{figure}

Each of the spline types that I have discussed so far share a common
feature; the degree of wiggliness over the time series is fixed due to
the use of a single smoothness parameter, \(\lambda\). The definition of
wiggliness, as the integrated squared second derivative of the spline,
ensures that fitted smoother does not jump about wildly. This assumes
that the data themselves are well described by a smoothly varying trend.
If we anticipate abrupt change or step-like responses to environmental
forcing this underlying assumption of the GAM would suggest that the
method is ill-suited to modelling palaeo time series in which such
features are evident or expected.

While there is not much we can do within the GAM framework to model a
series that contains both smooth trends and step-like responses, an
adaptive smoother can help address problems where the time series
consists of periods of rapid change in the mean combined with periods of
complacency or relatively little change. As suggested by their name,
adaptive smoothers can adjust to changes in the wiggliness of the time
series. This adaptive behaviour is achieved by making the smoothness
parameter \(\lambda\) itself depend smoothly on \(x_t\) (Ruppert et al.,
2003, 17; Wood, 2017, 5.3.5); in other words, the adaptive smoother
allows the wiggliness of the estimated trend to vary smoothly over time.
Whilst this does allow the estimated trend to adapt to periods of rapid
change in the response, adaptive smoothers make significant demands on
the data (Wood, 2017, 5.3.5); if we used \(m\) smoothness penalties to
allow the wiggliness to vary over a time series, it would be like
estimating \(m\) separate smooths from chunks of the original series
each of length \(n/m\). In a practical sense, this limits the use of
adaptive splines in palaeoecology to proxies that are readily
enumerated, such as the biogeochemical proxies used in the two example
data sets.

Figure \ref{fig:braya-so-model-comparisons} compares trends for the
Braya-Sø time series estimated using GAMs with the three main types of
spline discussed; i) TPRS, ii) Gaussian process smooths, and iii) an
adaptive smoother using 45 basis functions and 5 smoothing parameters.
There is a clear difference in the behaviour of the adaptive and
non-adaptive smoothers for the first 1000 years of the record, with the
adaptive smooth exhibiting much less variation compared with either the
TPRS or Gaussian process splines. Over the remaining two thirds of the
series, there is much closer agreement in the three smooths.

The behaviour of the TPRS and Gaussian process splines for these data is
the result of requiring a large amount of wiggliness (a small
\(\lambda\)) to adapt to the large oscillations in \uk{} present around
year 250CE and again at \textasciitilde{}900--1500CE. This large degree
of wiggliness allows the splines to potentially over-fit individual data
points much earlier in the record. Because the adaptive smoother, in
contrast, can adapt to these periods of rapid change in the response it
is much less susceptible to this ``chasing'' behaviour --- we don't need
to waste effective degrees of freedom in periods with little or no
change just to be able to fit the data well when there is a lot of
change.

This potential for over-fitting in such situations is undesirable, yet
if we recall Figure \ref{fig:derivatives} and the discussion around the
use of the first derivative to identify periods of significant change,
we would not interpret the oscillations in the early part of the \uk{}
record as being statistically significant. Owing to the paucity of data
in this part of the series the trends fitted using the non-adaptive
smoothers are subject to such a large degree of uncertainty that the
alternative of no trend through the first 1000 years of the record is
also a plausible explanation of the data. The trend estimated using the
adaptive smooth reflects this. Therefore, should we conclude that there
is no trend in \uk{} and thence climate in this period? I believe that
to be too-strong a statement; those oscillations in \uk{} may be real
responses to climate forcing but we may simply lack the statistical
power to distinguish them from the null hypothesis of no trend through
this period. The adaptive smoother is only adjusting to the data
available to it; just because it does not detect a trend during this
period does not lend itself to an interpretation of stable climate
forcing or complacency in the lake's response to forcing. If there were
particular interest in the climate of this particular period we might
take from the Braya-Sø record that there is potential early variation in
climate forcing, but that additional data from this or other sites is
required before any definitive conclusion can be drawn.

\subsection{Accounting for age model
uncertainty}\label{accounting-for-age-model-uncertainty}

Thus far, the trend models that I have described and illustrated assumed
that the time covariate (\(x_t\)) was fixed and known. In both examples,
and more generally for most palaeoecological records, this assumption is
violated. Unless the record is annually laminated, assigning an age to a
sediment interval requires the development of an age model from
observations of the relationship between depth down the sediment core
and estimates of the age of the sample arrived at using any of a number
of techniques, for example \textsuperscript{210}Pb or
\textsuperscript{14}C radiometric dating. This age-depth relationship is
itself uncertain, usually being derived from a mathematical or
statistical model applied to point age estimates (e.g.~ Blaauw and
Heegaard, 2012). Incorporating this additional component of uncertainty
complicates the estimation of statistical models from
palaeoenvironmental data. In this section I illustrate a simulation
based approach to quantify and account for age-model uncertainty as part
of the trend estimation using a GAM (see Anchukaitis and Tierney (2013)
for a similar, non-GAM related idea).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/small-scam-fit-plots-1}

}

\caption{Accounting for uncertainty in age estimates whilst fitting a smooth trend to the Small Water $\delta^{15}\text{N}$ time series. (a) Estimated age model using a monotonically-constrained spline fitted to ${}^{210}\text{Pb}$ inferred ages for selected depths in the sediment core (red points). The uncertainty in the ${}^{210}\text{Pb}$ inferred age is show by the red vertical bars. The fitted age model is illustrated by the solid black line. The faint grey lines are 25 random draws from the posterior distribution of the monotonically constrained GAM. The effect of age uncertainty on trend estimation is shown in b); for 100 simulations from the posterior distribution of the age model in a) a trend was estimated using a GAM with a thin plate regression spline basis and a CAR(1) process in the residuals. These trends are shown as grey lines. The combined effect of age model and fitted GAM uncertainty on the trends for the $\delta^{15}\text{N}$ time series is shown in c). The grey lines in c) are based on 50 random draws from the model posterior distribution for each of the 100 trends shown in b). For both b) and c) the black line shows the trend estimated assuming the ages of each sediment sample are known and fixed.}\label{fig:small-scam-fit-plots}
\end{figure}

Figure \ref{fig:small-scam-fit-plots}a shows the estimated dates (in
Years CE) for 12 levels in the Small Water core dated using
\textsuperscript{210}Pb. The vertical bars show the estimated age
uncertainty of each level. The solid line through the data points is an
additive model fitted to the observations, with prior weights given by
the estimated age uncertainties. The fitted age-depth model is
constrained to be monotonically decreasing with increasing depth,
following the method of (Pya and Wood, 2015) using the \emph{scam}
package (Pya, 2017). Also shown are 25 simulations from the model
posterior of the monotonically-constrained GAM. Each simulation from the
posterior distribution of the age-model is itself a potential age-depth
model, which can be used to assign dates to the Small Water core. The
trend model in \eqref{eq:gam} can be fitted to the
δ\textsuperscript{15}N data using these new dates as \(x_t\), and the
whole process repeated for a large number of simulations from the age
model.

Figure \ref{fig:small-scam-fit-plots}b shows the trend in
δ\textsuperscript{15}N for the observed age-depth model, plus trends
estimated via the same model using 100 draws from the posterior
distribution of the age model. In this case, the age-depth model is
relatively simple with little variation in the posterior draws,
resulting in trends that match closely that obtained from the estimated
age-depth relationship. Even so, this additional uncertainty suggests
that the timing of the decline in δ\textsuperscript{15}N covers the
interval \textasciitilde{}1935--1945.

The uncertainty in the trend estimates illustrated in Figure
\ref{fig:small-scam-fit-plots}b only reflects the variation in trends
fitted to the uncertain dates of the sediment samples. To fully
visualise the uncertainty in the trend estimates, incorporating both age
model uncertainty \emph{and} uncertainty in the estimated model
coefficients themselves, 50 simulations from the posterior distribution
of each of the 100 estimated trends shown in Figure
\ref{fig:small-scam-fit-plots}b were performed, resulting in 5,000 trend
estimates for the δ\textsuperscript{15}N series. These are shown in
Figure \ref{fig:small-scam-fit-plots}c, where the two obvious changes
over the same simulations without accounting for uncertainty in \(x_t\)
(Figure \ref{fig:posterior-simulation}a) are that the uncertainty band
traced out by the simulations is approximately 50\% wider and, not
surprisingly, the uncertainty in the estimated trend is most pronounced
in the least accurately-dated section of the core. Despite this
additional uncertainty however, the main result holds; a marked decline
of \textasciitilde{}1.5‰ that occurred between approximately 1930 and
1945, with mild evidence of a small increase in δ\textsuperscript{15}N
post 2000 CE.

\subsection{Multivariate data}\label{multivariate-data}

A large proportion of the palaeoenvironmental data generated today is
multivariate in nature and yet the two examples used to illustrate GAMs
were univariate. Can the approach described here be used for
multivariate data? Yes, and no. With one main exception it is not
possible to directly apply the GAM methodology described here to
multivariate abundance data, where the aim is to model all species at
once. The \emph{mgcv} software for example, is not able to estimate the
penalized GAM for multiple non-Gaussian responses. The exception is for
a small number of correlated Gaussian responses; these could be modelled
as being distributed multivariate normal conditional upon the
covariates. Such a model would estimate the expected values of each
response and the correlations between them. For example, we could
jointly model δ\textsuperscript{15}N and δ\textsuperscript{13}C series
using this approach.

Formal multivariate versions of GLM or GAMs are currently an important
area of research within ecology (see Warton et al. (2015) for a recent
review), where they go by the name joint species distribution models
(JSDMs). Whilst undoubtedly powerful, our knowledge regarding JSDMs and
their availability in software are still in their relative infancy and
they require considerable expertise to implement. As such, JSDMs are
currently beyond the reach of most palaeoecologists. Despite this, we
should be watching JSDM research as developments are ongoing and a
degree of method maturation occurring.

One currently available avenue for fitting a multivariate GAM is via
regularized sandwich estimators and GLMs (Warton, 2011), which involves
fitting separate GLMs (or GAMs) to each response variable and
subsequently using resampling-based hypothesis tests to determine which
covariates are related to variation at the community level and for
individual taxa (Wang et al., 2012; Warton, 2011; Warton et al., 2012).
The \emph{mvabund} package (Wang et al., 2012) implements this approach
within R and can use \emph{mgcv} to fit GAMs to each species.

A pragmatic although inelegant approach that has been used to estimate
trends in multivariate palaeoecological data is to first summarise the
response data using an unconstrained ordination via a PCA, CA, or
principal curve and then fit separate GAM models to the site (sample)
scores of the first few ordination axes or principal curve (Beck et al.,
2018; Bennion et al., 2015). Whilst this two-step approach is relatively
easy to implement and builds on approaches that palaeoecologists already
use to summarise multivariate stratigraphic data, it is best thought of
as modelling changes in abundance or relative composition at the
community level. It is less well suited to unpicking taxon-specific
trends however, because the ordination step combines individual species
information into latent variables (axes) that are linear combinations of
\emph{all} species and it is these latent variables that are then
modelled using GAM.

\section{Conclusions}\label{conclusions}

Formal statistical estimation of trends in palaeoenvironmental data has
been hampered by the nature of the data that comprise the time series;
the uneven spacing of samples in time makes it, if not impossible,
difficult to fit classical statistical time series models like ARIMA.
This has lead palaeoecologists and palaeolimnologists to fall back on
basic statistical methods such as linear parametric and non-parametric
correlations or simple linear regression models, where the assumptions
of the method are often grossly violated by the dependencies inherent to
time series data. GAMs, whilst similar to the popular \loess{} smoother,
provide a superior alternative approach to trend estimation in
palaeoenvironmental time series. They can estimate non-linear trends,
provide estimates of the magnitude of change as well as allow the
identification of periods of change, can account for the lack of
independence (either via autocorrelation processes or via the fitting of
a wiggly trend), and provide a formal framework for statistical
inference on each of these features.

In presenting the GAM with specific palaeoenvironmental examples and
addressing the issues that arise in palaeoenvironmental time series, it
is hoped that palaeoecologists and palaeolimnologists will be motivated
to give greater consideration to the estimation of trends and the
identification of change.

\section*{Conflict of interest
statement}\label{conflict-of-interest-statement}
\addcontentsline{toc}{section}{Conflict of interest statement}

The author declares that the research was conducted in the absence of
any commercial or financial relationships that could be construed as a
potential conflict of interest.

\section*{Acknowledgements}\label{acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

The ideas expressed in this paper are the result of many fruitful
conversations with colleagues at the Environmental Change Research
Centre, UCL, and the University of Regina. In particular I am indebted
to Helen Bennion, Rick Battarbee, and Peter Leavitt for their
collaborations on projects over many years. Without Simon Wood's
\emph{mgcv} software and his research on GAMs, the application of these
models to palaeo time series would not be as straight forward. I also
thank David Miller, Eric Pedersen, and Noam Ross, my GAM workshop
partners in crime. This work was supported by a Natural Sciences and
Engineering Council of Canada (NSERC) Discovery Grant to the author
(RGPIN-2014-04032).

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-Anchukaitis2013-la}{}
Anchukaitis, K. J., and Tierney, J. E. (2013). Identifying coherent
spatiotemporal modes in time-uncertain proxy paleoclimate records.
\emph{Climate Dynamics} 41, 1291--1306.
doi:\href{https://doi.org/10.1007/s00382-012-1483-0}{10.1007/s00382-012-1483-0}.

\hypertarget{ref-Beck2018-zm}{}
Beck, K. K., Fletcher, M.-S., Gadd, P. S., Heijnis, H., Saunders, K. M.,
Simpson, G. L., et al. (2018). Variance and Rate-of-Change as early
warning signals for a critical transition in an aquatic ecosystem state:
A test case from tasmania, australia. \emph{Journal of Geophysical
Research: Biogeosciences} 123, 2017JG004135.
doi:\href{https://doi.org/10.1002/2017JG004135}{10.1002/2017JG004135}.

\hypertarget{ref-Bennion2015-ph}{}
Bennion, H., Simpson, G. L., and Goldsmith, B. J. (2015). Assessing
degradation and recovery pathways in lakes impacted by eutrophication
using the sediment record. \emph{Frontiers in Ecology and Evolution} 3.
doi:\href{https://doi.org/10.3389/fevo.2015.00094}{10.3389/fevo.2015.00094}.

\hypertarget{ref-Bergmeir2018-pu}{}
Bergmeir, C., Hyndman, R. J., and Koo, B. (2018). A note on the validity
of cross-validation for evaluating autoregressive time series
prediction. \emph{Computational Statistics \& Data Analysis} 120,
70--83.
doi:\href{https://doi.org/10.1016/j.csda.2017.11.003}{10.1016/j.csda.2017.11.003}.

\hypertarget{ref-Birks1998-po}{}
Birks, H. J. B. (1998). Numerical tools in palaeolimnology --- progress,
potentialities, and problems. \emph{Journal of Paleolimnology} 20,
307--332.
doi:\href{https://doi.org/10.1023/A:1008038808690}{10.1023/A:1008038808690}.

\hypertarget{ref-Birks2012-op}{}
Birks, H. J. B. (2012a). ``Introduction and overview of part III,'' in
\emph{Tracking environmental change using lake sediments} (Springer,
Dordrecht), 331--353.
doi:\href{https://doi.org/10.1007/978-94-007-2745-8/_10}{10.1007/978-94-007-2745-8\textbackslash{}\_10}.

\hypertarget{ref-Birks2012-fa}{}
Birks, H. J. B. (2012b). ``Overview of numerical methods in
palaeolimnology,'' in \emph{Tracking environmental change using lake
sediments} (Springer, Dordrecht), 19--92.
doi:\href{https://doi.org/10.1007/978-94-007-2745-8/_2}{10.1007/978-94-007-2745-8\textbackslash{}\_2}.

\hypertarget{ref-Blaauw2012-yq}{}
Blaauw, M., and Heegaard, E. (2012). ``Estimation of Age-Depth
relationships,'' in \emph{Tracking environmental change using lake
sediments} (Springer, Dordrecht), 379--413.
doi:\href{https://doi.org/10.1007/978-94-007-2745-8/_12}{10.1007/978-94-007-2745-8\textbackslash{}\_12}.

\hypertarget{ref-Brassell1993-fw}{}
Brassell, S. C. (1993). ``Applications of biomarkers for delineating
marine paleoclimatic fluctuations during the pleistocene,'' in
\emph{Organic geochemistry: Principles and applications}, eds. M. H.
Engel and S. A. Macko (Boston, MA: Springer US), 699--738.
doi:\href{https://doi.org/10.1007/978-1-4615-2890-6/_34}{10.1007/978-1-4615-2890-6\textbackslash{}\_34}.

\hypertarget{ref-Chu2005-es}{}
Chu, G., Sun, Q., Li, S., Zheng, M., Jia, X., Lu, C., et al. (2005).
Long-chain alkenone distributions and temperature dependence in
lacustrine surface sediments from china. \emph{Geochimica et
Cosmochimica Acta} 69, 4985--5003.
doi:\href{https://doi.org/10.1016/j.gca.2005.04.008}{10.1016/j.gca.2005.04.008}.

\hypertarget{ref-Cleveland1979-vf}{}
Cleveland, W. S. (1979). Robust locally weighted regression and
smoothing scatterplots. \emph{Journal of the American Statistical
Association} 74, 829--836.
doi:\href{https://doi.org/10.1080/01621459.1979.10481038}{10.1080/01621459.1979.10481038}.

\hypertarget{ref-Craven1978-nz}{}
Craven, P., and Wahba, G. (1978). Smoothing noisy data with spline
functions. \emph{Numerische Mathematik} 31, 377--403.
doi:\href{https://doi.org/10.1007/BF01404567}{10.1007/BF01404567}.

\hypertarget{ref-Duchon1977-jr}{}
Duchon, J. (1977). ``Splines minimizing rotation-invariant semi-norms in
sobolev spaces,'' in \emph{Constructive theory of functions of several
variables} (Springer, Berlin, Heidelberg), 85--100.
doi:\href{https://doi.org/10.1007/BFb0086566}{10.1007/BFb0086566}.

\hypertarget{ref-Dutilleul2012-su}{}
Dutilleul, P., Cumming, B. F., and Lontoc-Roy, M. (2012).
``Autocorrelogram and periodogram analyses of palaeolimnological
Temporal-Series from lakes in central and western north america to
assess shifts in drought conditions,'' in \emph{Tracking environmental
change using lake sediments} (Springer, Dordrecht), 523--548.
doi:\href{https://doi.org/10.1007/978-94-007-2745-8/_16}{10.1007/978-94-007-2745-8\textbackslash{}\_16}.

\hypertarget{ref-DAndrea2011-oa}{}
D'Andrea, W. J., Huang, Y., Fritz, S. C., and Anderson, N. J. (2011).
Abrupt holocene climate change as an important factor for human
migration in west greenland. \emph{Proceedings of the National Academy
of Sciences} 108, 9765--9769.
doi:\href{https://doi.org/10.1073/pnas.1101708108}{10.1073/pnas.1101708108}.

\hypertarget{ref-Epperson1987-ui}{}
Epperson, J. F. (1987). On the Runge example. \emph{The American
Mathematical Monthly} 94, 329--341.
doi:\href{https://doi.org/10.2307/2323093}{10.2307/2323093}.

\hypertarget{ref-Fahrmeir2008-si}{}
Fahrmeir, L., and Kneib, T. (2008). ``On the identification of trend and
correlation in temporal and spatial regression,'' in \emph{Recent
advances in linear models and related areas} (Physica-Verlag HD), 1--27.
doi:\href{https://doi.org/10.1007/978-3-7908-2064-5/_1}{10.1007/978-3-7908-2064-5\textbackslash{}\_1}.

\hypertarget{ref-Fahrmeir2013-xu}{}
Fahrmeir, L., Kneib, T., Lang, S., and Marx, B. (2013).
\emph{Regression: Models, methods and applications}. Springer Berlin
Heidelberg
doi:\href{https://doi.org/10.1007/978-3-642-34333-9}{10.1007/978-3-642-34333-9}.

\hypertarget{ref-Gautheir2001-ee}{}
Gautheir, T. D. (2001). Detecting trends using spearman's rank
correlation coefficient. \emph{Environmental Forensics} 2, 359--362.
doi:\href{https://doi.org/10.1080/713848278}{10.1080/713848278}.

\hypertarget{ref-Glew2001-co}{}
Glew, J. R., Smol, J. P., and Last, W. M. (2001). ``Sediment core
collection and extrusion,'' in \emph{Tracking environmental change using
lake sediments: Basin analysis, coring, and chronological techniques},
eds. W. M. Last and J. P. Smol (Dordrecht: Springer Netherlands),
73--105.
doi:\href{https://doi.org/10.1007/0-306-47669-X/_5}{10.1007/0-306-47669-X\textbackslash{}\_5}.

\hypertarget{ref-Golding2016-bb}{}
Golding, N., and Purse, B. V. (2016). Fast and flexible bayesian species
distribution modelling using gaussian processes. \emph{Methods in
Ecology and Evolution}.
doi:\href{https://doi.org/10.1111/2041-210X.12523}{10.1111/2041-210X.12523}.

\hypertarget{ref-Handcock1994-lj}{}
Handcock, M. S., Meier, K., and Nychka, D. (1994). Kriging and splines:
An empirical comparison of their predictive performance in some
applications: Comment. \emph{Journal of the American Statistical
Association} 89, 401--403.
doi:\href{https://doi.org/10.2307/2290838}{10.2307/2290838}.

\hypertarget{ref-Hastie1990-bx}{}
Hastie, T. J., and Tibshirani, R. J. (1990). \emph{Generalized additive
models}. Boca Raton, Fl.: Chapman \& Hall / CRC.

\hypertarget{ref-Hastie1986-xs}{}
Hastie, T., and Tibshirani, R. (1986). Generalized additive models.
\emph{Statistical Science} 1, 297--310.

\hypertarget{ref-Juggins2012-uu}{}
Juggins, S., and Telford, R. J. (2012). ``Exploratory data analysis and
data display,'' in \emph{Tracking environmental change using lake
sediments} (Springer, Dordrecht), 123--141.
doi:\href{https://doi.org/10.1007/978-94-007-2745-8/_5}{10.1007/978-94-007-2745-8\textbackslash{}\_5}.

\hypertarget{ref-Kammann2003-pi}{}
Kammann, E. E., and Wand, M. P. (2003). Geoadditive models.
\emph{Journal of the Royal Statistical Society. Series C, Applied
statistics} 52, 1--18.
doi:\href{https://doi.org/10.1111/1467-9876.00385}{10.1111/1467-9876.00385}.

\hypertarget{ref-Kimeldorf1970-cn}{}
Kimeldorf, G. S., and Wahba, G. (1970). A correspondence between
bayesian estimation on stochastic processes and smoothing by splines.
\emph{Annals of Mathematical Statistics} 41, 495--502.

\hypertarget{ref-Magee1998-oq}{}
Magee, L. (1998). Nonlocal behavior in polynomial regressions. \emph{The
American Statistician} 52, 20--22.
doi:\href{https://doi.org/10.2307/2685560}{10.2307/2685560}.

\hypertarget{ref-Mann2004-kd}{}
Mann, M. E. (2004). On smoothing potentially non-stationary climate time
series. \emph{Geophysical Research Letters} 31, L07214.
doi:\href{https://doi.org/10.1029/2004GL019569}{10.1029/2004GL019569}.

\hypertarget{ref-Mann2008-yk}{}
Mann, M. E. (2008). Smoothing of climate time series revisited.
\emph{Geophysical Research Letters} 35, L16708.
doi:\href{https://doi.org/10.1029/2008GL034716}{10.1029/2008GL034716}.

\hypertarget{ref-Marra2012-bq}{}
Marra, G., and Wood, S. N. (2012). Coverage properties of confidence
intervals for generalized additive model components. \emph{Scandinavian
Journal of Statistics, Theory and Applications} 39, 53--74.
doi:\href{https://doi.org/10.1111/j.1467-9469.2011.00760.x}{10.1111/j.1467-9469.2011.00760.x}.

\hypertarget{ref-McCullagh1989-gv}{}
McCullagh, P., and Nelder, J. A. (1989). \emph{Generalized linear
models, second edition}. CRC Press.

\hypertarget{ref-Mills2006-rb}{}
Mills, T. C. (2006). Modelling current trends in Northern Hemisphere
temperatures. \emph{International Journal of Climatology} 26, 867--884.
doi:\href{https://doi.org/10.1002/joc.1286}{10.1002/joc.1286}.

\hypertarget{ref-Mills2007-pq}{}
Mills, T. C. (2007). A note on trend decomposition: The ``classical''
approach revisited with an application to surface temperature trends.
\emph{Journal of applied statistics} 34, 963--972.
doi:\href{https://doi.org/10.1080/02664760701590418}{10.1080/02664760701590418}.

\hypertarget{ref-Mills2010-oe}{}
Mills, T. C. (2010). ``Skinning a cat'': Alternative models of
representing temperature trends. \emph{Climatic Change} 101, 415--426.

\hypertarget{ref-Nychka1988-rz}{}
Nychka, D. (1988). Bayesian confidence intervals for smoothing splines.
\emph{Journal of the American Statistical Association} 83, 1134--1143.
doi:\href{https://doi.org/10.1080/01621459.1988.10478711}{10.1080/01621459.1988.10478711}.

\hypertarget{ref-PAGES_2k_Consortium2013-fm}{}
PAGES 2K Consortium (2013). Continental-scale temperature variability
during the past two millennia. \emph{Nature Geoscience} 6, 339--346.
doi:\href{https://doi.org/10.1038/ngeo1797}{10.1038/ngeo1797}.

\hypertarget{ref-Pinheiro2000-kx}{}
Pinheiro, J. C., and Bates, D. M. (2000). \emph{Mixed-Effects models in
S and S-PLUS}. Springer Science \& Business Media.

\hypertarget{ref-pya-scam-pkg}{}
Pya, N. (2017). \emph{Scam: Shape constrained additive models}.
Available at: \url{https://CRAN.R-project.org/package=scam}.

\hypertarget{ref-Pya2015-wr}{}
Pya, N., and Wood, S. N. (2015). Shape constrained additive models.
\emph{Statistics and Computing} 25, 543--559.
doi:\href{https://doi.org/10.1007/s11222-013-9448-7}{10.1007/s11222-013-9448-7}.

\hypertarget{ref-Pya2016-rk}{}
Pya, N., and Wood, S. N. (2016). A note on basis dimension selection in
generalized additive modelling. \emph{ArXiv e-prints}.

\hypertarget{ref-R-core-team}{}
R Core Team (2017). \emph{R: A language and environment for statistical
computing}. Vienna, Austria: R Foundation for Statistical Computing
Available at: \url{https://www.R-project.org/}.

\hypertarget{ref-Rasmussen2006-xk}{}
Rasmussen, C. E., and Williams, C. K. I. (2006). \emph{Gaussian
processes for machine learning}. MIT Press.

\hypertarget{ref-Reiss2009-fk}{}
Reiss, P. T., and Ogden, R. T. (2009). Smoothing parameter selection for
a class of semiparametric linear models. \emph{Journal of the Royal
Statistical Society. Series B, Statistical methodology} 71, 505--523.
doi:\href{https://doi.org/10.1111/j.1467-9868.2008.00695.x}{10.1111/j.1467-9868.2008.00695.x}.

\hypertarget{ref-Runge1901-tc}{}
Runge, C. (1901). Über empirische funktionen und die interpolation
zwischen äquidistanten ordinaten. \emph{Zeitschrift fur Angewandte
Mathematik und Physik} 46, 224--243.

\hypertarget{ref-Ruppert2003-pt}{}
Ruppert, D., Wand, M. P., and Carroll, R. J. (2003).
\emph{Semiparametric Regression}. Cambridge University Press.

\hypertarget{ref-Silverman1985-kw}{}
Silverman, B. W. (1985). Some aspects of the spline smoothing approach
to non-parametric regression curve fitting. \emph{Journal of the Royal
Statistical Society. Series B, Statistical methodology} 47, 1--52.

\hypertarget{ref-Smol2008-ld}{}
Smol, J. P. (2008). \emph{Pollution of lakes and rivers: A
paleoenvironmental perspective}. Blackwell Pub.

\hypertarget{ref-Smol2012-xt}{}
Smol, J. P., Birks, H. J. B., Lotter, A. F., and Juggins, S. (2012).
``The march towards the quantitative analysis of palaeolimnological
data,'' in \emph{Tracking environmental change using lake sediments}
(Springer, Dordrecht), 3--17.
doi:\href{https://doi.org/10.1007/978-94-007-2745-8/_1}{10.1007/978-94-007-2745-8\textbackslash{}\_1}.

\hypertarget{ref-Tian2011-ip}{}
Tian, J., Nelson, D. M., and Hu, F. S. (2011). How well do sediment
indicators record past climate? An evaluation using annually laminated
sediments. \emph{Journal of Paleolimnology} 45, 73--84.
doi:\href{https://doi.org/10.1007/s10933-010-9481-x}{10.1007/s10933-010-9481-x}.

\hypertarget{ref-Toney2010-ba}{}
Toney, J. L., Huang, Y., Fritz, S. C., Baker, P. A., Grimm, E., and
Nyren, P. (2010). Climatic and environmental controls on the occurrence
and distributions of long chain alkenones in lakes of the interior
united states. \emph{Geochimica et Cosmochimica Acta} 74, 1563--1578.
doi:\href{https://doi.org/10.1016/j.gca.2009.11.021}{10.1016/j.gca.2009.11.021}.

\hypertarget{ref-Wahba1983-eh}{}
Wahba, G. (1983). Bayesian ``confidence intervals'' for the
Cross-Validated smoothing spline. \emph{Journal of the Royal Statistical
Society. Series B, Statistical methodology} 45, 133--150.

\hypertarget{ref-Wahba1990-ag}{}
Wahba, G. (1990). \emph{Spline models for observational data}. SIAM.

\hypertarget{ref-Wang2012-uu}{}
Wang, Y., Naumann, U., Wright, S. T., and Warton, D. I. (2012).
Mvabund-- an R package for model-based analysis of multivariate
abundance data. \emph{Methods in Ecology and Evolution} 3, 471--474.
doi:\href{https://doi.org/10.1111/j.2041-210X.2012.00190.x}{10.1111/j.2041-210X.2012.00190.x}.

\hypertarget{ref-Warton2011-vp}{}
Warton, D. I. (2011). Regularized sandwich estimators for analysis of
high-dimensional data using generalized estimating equations.
\emph{Biometrics} 67, 116--123.
doi:\href{https://doi.org/10.1111/j.1541-0420.2010.01438.x}{10.1111/j.1541-0420.2010.01438.x}.

\hypertarget{ref-Warton2015-ge}{}
Warton, D. I., Blanchet, F. G., O'Hara, R. B., Ovaskainen, O., Taskinen,
S., Walker, S. C., et al. (2015). So many variables: Joint modeling in
community ecology. \emph{Trends in Ecology \& Evolution}.
doi:\href{https://doi.org/10.1016/j.tree.2015.09.007}{10.1016/j.tree.2015.09.007}.

\hypertarget{ref-Warton2012-fg}{}
Warton, D. I., Wright, S. T., and Wang, Y. (2012). Distance-based
multivariate analyses confound location and dispersion effects.
\emph{Methods in Ecology and Evolution} 3, 89--101.
doi:\href{https://doi.org/10.1111/j.2041-210X.2011.00127.x}{10.1111/j.2041-210X.2011.00127.x}.

\hypertarget{ref-Wood2003-qy}{}
Wood, S. N. (2003). Thin plate regression splines. \emph{Journal of the
Royal Statistical Society. Series B, Statistical methodology} 65,
95--114.
doi:\href{https://doi.org/10.1111/1467-9868.00374}{10.1111/1467-9868.00374}.

\hypertarget{ref-Wood2004-zv}{}
Wood, S. N. (2004). Stable and efficient multiple smoothing parameter
estimation for generalized additive models. \emph{Journal of the
American Statistical Association} 99, 673--686.
doi:\href{https://doi.org/10.2307/27590439}{10.2307/27590439}.

\hypertarget{ref-Wood2011-kn}{}
Wood, S. N. (2011). Fast stable restricted maximum likelihood and
marginal likelihood estimation of semiparametric generalized linear
models. \emph{Journal of the Royal Statistical Society. Series B,
Statistical methodology} 73, 3--36.
doi:\href{https://doi.org/10.1111/j.1467-9868.2010.00749.x}{10.1111/j.1467-9868.2010.00749.x}.

\hypertarget{ref-Wood2017-qi}{}
Wood, S. N. (2017). \emph{Generalized Additive Models: An Introduction
with R, Second edition}. CRC Press.

\hypertarget{ref-Wood2016-fx}{}
Wood, S. N., Pya, N., and Säfken, B. (2016). Smoothing parameter and
model selection for general smooth models. \emph{Journal of the American
Statistical Association} 111, 1548--1563.
doi:\href{https://doi.org/10.1080/01621459.2016.1180986}{10.1080/01621459.2016.1180986}.

\hypertarget{ref-Yee1991-mg}{}
Yee, T. W., and Mitchell, N. D. (1991). Generalized additive models in
plant ecology. \emph{Journal of Vegetation Science} 2, 587--602.
doi:\href{https://doi.org/10.2307/3236170}{10.2307/3236170}.

\hypertarget{ref-Zink2001-cr}{}
Zink, K.-G., Leythaeuser, D., Melkonian, M., and Schwark, L. (2001).
Temperature dependency of long-chain alkenone distributions in recent to
fossil limnic sediments and in lake waters11Associate editor: J. b.
fein. \emph{Geochimica et Cosmochimica Acta} 65, 253--265.
doi:\href{https://doi.org/10.1016/S0016-7037(00)00509-3}{10.1016/S0016-7037(00)00509-3}.

\hypertarget{appendix}{\section*{Appendix 1 --- Simultaneous
intervals}\label{appendix}}
\addcontentsline{toc}{section}{Appendix 1 --- Simultaneous intervals}

We proceed by considering a simultaneous confidence interval for a
function \(f(x)\) at a set of \(M\) locations in \(x\); we'll refer to
these locations, following the notation of Ruppert et al. (2003) by

\[
\mathbf{g} = (g_1, g_2, \ldots, g_M)
\]

The true function over \(\mathbf{g}\), \(\mathbf{f_g}\), is defined as
the vector of evaluations of \(f\) at each of the \(M\) locations

\[
\begin{aligned}
    \mathbf{f_g} &\equiv \begin{bmatrix}
           f(g_1) \\
           f(g_2) \\
           \vdots \\
           f({g_M}) \\
         \end{bmatrix}
\end{aligned}
\]

and the corresponding estimate of the true function given by the fitted
GAM denoted by \(\hat{\mathbf{f}}_{\mathbf{g}}\). The difference between
the true function and our unbiased estimator is given by

\[
\begin{aligned}
    \hat{\mathbf{f}}_{\mathbf{g}} - \mathbf{f_g} &= \mathbf{C_g} \begin{bmatrix}
           \boldsymbol{\hat{\beta}} - \boldsymbol{\beta} \\
           \mathbf{\hat{u}} - \mathbf{u} \\
         \end{bmatrix}\, ,
\end{aligned}
\] where \(\mathbf{C_g}\) is a matrix formed by the evaluation of the
basis functions at locations \(\mathbf{g}\), and the part in square
brackets is the bias in the estimated model coefficients, which we
assume to be mean 0 and distributed, approximately, multivariate normal
with mean vector \(\mathbf{0}\) and covariance matrix \(\mathbf{V_b}\)

\[
\begin{bmatrix}
    \boldsymbol{\hat{\beta}} - \boldsymbol{\beta} \\
     \mathbf{\hat{u}} - \mathbf{u} \\
\end{bmatrix} \stackrel{\text{approx.}}{\sim} N \left (\mathbf{0}, \mathbf{V_b} \right )\, ,
\] where \(\mathbf{V_b}\) is the Bayesian covariance matrix of the GAM
coefficients.

Now, the (1 - \(\alpha\))100\% simultaneous confidence interval is

\[
\begin{aligned}
    \hat{\mathbf{f}}_{\mathbf{g}} &\pm m_{1 - \alpha} \begin{bmatrix}
           \widehat{\mathrm{st.dev}} (\hat{f}(g_1) - f(g_1)) \\
           \widehat{\mathrm{st.dev}} (\hat{f}(g_2) - f(g_2)) \\
           \vdots \\
           \widehat{\mathrm{st.dev}} (\hat{f}(g_M) - f(g_M)) \\
         \end{bmatrix}\, ,
\end{aligned}
\] where \(m_{1 - \alpha}\) is the 1 - \(\alpha\) quantile of the random
variable

\[
\sup_{x \in \mathcal{x}} \left | \frac{\hat{f}(x) - f(x)}{\widehat{\mathrm{st.dev}} (\hat{f}(x) - f(x))} \right | \approx \max_{1 \leq \ell \leq M} \left | \frac{\left ( \mathbf{C_g} \begin{bmatrix}
           \boldsymbol{\hat{\beta}} - \boldsymbol{\beta} \\
           \mathbf{\hat{u}} - \mathbf{u} \\
         \end{bmatrix} \right )_\ell}{\widehat{\mathrm{st.dev}} (\hat{f}(g_{\ell}) - f(g_{\ell}))} \right |
\]

The \(\sup\) refers to the \emph{supremum} or the \emph{least upper
bound}; this is the least value of \(\mathcal{X}\), the set of all
values of which we observed subset \(x\), that is \emph{greater} than
all of the values in the subset. Often this is the maximum value of the
subset. This is what is indicated by the right-hand side of the
equation; we want the maximum (absolute) value of the ratio over all
values in \(\mathbf{g}\).

The fractions in both sides of the equation correspond to the
standardized deviation between the true function and the model estimate,
and we consider the \emph{maximum absolute} standardized deviation. We
don't usually know the distribution of the maximum absolute standardized
deviation but we need this to access its quantiles. However, we can
closely approximate the distribution via simulation. The difference here
is that rather than simulating from the posterior of the model as we did
earlier \protect\hyperlink{confint}{see section \emph{Confidence
intervals}}, this time we simulate from the multivariate normal
distribution with mean vector \(\mathbf{0}\) and covariance matrix
\(\mathbf{V_{b}}\). For each simulation we find the maximum absolute
standardized deviation of the fitted function from the true function
over the grid of \(x\) values we are considering. Then we collect all
these maxima, sort them and either take the 1 - \(\alpha\) probability
quantile of the maxima, or the maximum with rank
\(\lceil (1 - \alpha) / N \rceil\).

\end{document}

### 2018-05-14 manuscript.Rmd
submitted version of the manuscript

@@
       in_header: header.tex
+  md_document:
+    variant: markdown_github
 ---
@@
 
+Palaeoecology and palaeolimnology have moved away from being descriptive disciplines, rapidly adopting new statistical developments in the 1990s and beyond [@@Smol2012-xt].
+Less development has been observed in the area of trend estimation in palaeoenvironmental time series.
 The vast majority of data produced by palaeoecologists and palaeolimnologists is in the form of time-ordered observations on one or more proxies or biological taxa [@Smol2008-ld; @Smol2012-xt; @Birks2012-fa].
 Typically these data are arranged irregularly in time; in the absence of annual laminae or varves, the sediment core is sectioned at regular depth intervals [@Glew2001-co], which, owing to variation in accumulation rates over time and compaction by overlying sediments, results in an uneven sampling in time.
-An under-appreciated secondary feature of such sampling is that younger sediments often have larger variance than older sediments; each section of core represents fewer lake years in newer samples, relative to older samples.
-This variable time averaging acts as a variable low-pass (high-cut) filter of the raw, annual depositional signal.
-
-The irregular intervals between samples means that the customary time-series analysis methods of autoregressive or moving average processes, in the form of autoregressive integrated moving average (ARIMA) models, are practically impossible to apply because software implementing these methods typically requires even spacing of observations in time.
+An under-appreciated feature of such sampling is that younger sediments often have larger variance than older sediments; each section of core represents fewer lake years in newer samples, relative to older samples.
+This variable averaging acts as a time-varying low-pass (high-cut) filter of the annual depositional signal.
+
+Irregular intervals between samples means that the time-series analysis methods of autoregressive or moving average processes, in the form of autoregressive integrated moving average (ARIMA) models, are practically impossible to apply because software typically requires even spacing of observations in time.
 @Dutilleul2012-su and @Birks2012-op, eschewing the term *time series*, prefer to call such data *temporal series* on account of the irregular spacing of samples, a distinction that I find unnecessary.
 
-Where statistical approaches have been applied to trend estimation in palaeoenvironmental time series, the most commonly used has been \loess{} [@Cleveland1979-vf; @Juggins2012-uu; @Birks2012-op; @Birks1998-po].
-\loess{}, or locally weighted scatterplot smoother, as it's name suggests, was developed to smooth x-y scatterplot data, ostensibly as part of exploratory data analysis [@Cleveland1979-vf].
-The method fits a smooth line through data by fitting weighted least squares (WLS) models to observations within a particular, user-specified window of the target point, whose width is typically expressed as a proportion $\alpha$ of the $n$ data points.
-Weights are determined by how close (in the x-axis only) an observation in the window is to the focal point, with a tricubic function, by default, giving greatest weight to points closest to the focal point.
+Where statistical approaches have been applied to trend estimation in palaeoenvironmental time series, a commonly-used method is \loess{} [@Cleveland1979-vf; @Juggins2012-uu; @Birks2012-op; @Birks1998-po].
+\loess{}, locally weighted scatterplot smoother, as it's name suggests, was developed to smooth x-y scatterplot data [@Cleveland1979-vf].
+The method fits a smooth line through data by fitting weighted least squares (WLS) models to observations within a particular, user-specified window of the focal point, whose width is typically expressed as a proportion $\alpha$ of the $n$ data points.
+Weights are determined by how close (in the x-axis only) an observation in the window is to the focal point giving greatest weight given to points closest to the focal point.
 The interim \loess{}-smoothed value for the focal point is the predicted value from the weighted regression at the focal point.
-The interim values are updated using weights based on how far in the y-axis direction, the interim smoothed value lies from the observed value plus the x-axis distance weights; this has the effect of down-weighting the effects of outlier observations.
-The final \loess{} is obtained by joining up the smoothed (fitted) values.
+The interim values are updated using weights based on how far in the y-axis direction the interim smoothed value lies from the observed value plus the x-axis distance weights; this has the effect of down-weighting outlier observations.
+The final \loess{} is obtained by joining the smoothed values.
 The user has to choose how large a window to use, whether to fit degree 1 (linear) or degree 2 (quadratic) polynomials in the WLS model, and how to weight points in the x-axis.
 When used in an exploratory mode, the user has considerable freedom to choose the detail of the \loess{} fit; the window width, for example, can be infinitely tweaked to give as close a fit to the data, as assessed by eye, as is desired.
-Using cross-validation (CV) to choose $\alpha$ or the degree of polynomial in the WLS model is complicated for a number of reasons, not least that the CV scheme used must involve the time ordering of the data [e.g.
+Using cross-validation (CV) to choose $\alpha$ or the degree of polynomial in the WLS model is complicated for a number of reasons, not least because the CV scheme used must involve the time ordering of the data [e.g.
 @Bergmeir2018-pu].
-This level of subjectivity is problematic however once we wish to move beyond exploratory data analysis to statistically identify trends and to test hypotheses involving those trend estimates.
+This subjectivity is problematic however once we wish to move beyond exploratory analysis and statistically identify trends to test hypotheses involving those trend estimates.
 
 Running means or other types of filter [@Juggins2012-uu] have also been used extensively to smooth palaeoenvironmental time series, but, as with \loess{}, their behaviour depends on a number of factors, including the filter width.
-Furthermore, the width of the filter causes boundary issues; with a centred filter, of width five data points, the filtered time series would be two data points shorter at both ends of the series because the filter weights are not defined for the first and last two observations of the original series as these time points were not observed.
-Considerable research effort has been expended to identify ways to pad the original time series at one or both ends to maintain the original length in the filtered series, without introducing bias because of the method of padding [e.g.
+Furthermore, the width of the filter causes boundary issues; with a centred filter, of width five, the filtered time series would be two data points shorter at both ends of the series because the filter values are not defined for the first and last two observations of the original series as these extra time points were not observed.
+Considerable research effort has been expended to identify ways to pad the original time series at one or both ends to maintain the original length in the filtered series, without introducing bias due to the padding [e.g.
 @Mann2004-kd; @Mann2008-yk; @Mills2006-rb; @Mills2007-pq; @Mills2010-oe].
 
-These are by no means the only methods that have been used to estimated trends in stratigraphic series.
-Another common approach involves fitting a simple linear trend using ordinary least squares regression and use the resulting *t* statistic as evidence against the null hypothesis of no trend despite the statistical assumptions of the method being almost surely violated due to dependence among observations arising from temporal autocorrelation in the series.
+These are not the only methods that have been used to estimated trends in stratigraphic series.
+Another common approach involves fitting a simple linear trend using ordinary least squares regression and use the resulting *t* statistic as evidence against the null hypothesis of no trend despite the statistical assumptions being almost surely violated due to dependence among observations.
 The Pearson correlation coefficient, *r*, is also often used to detect trends in palaeo time series [@Birks2012-op], despite the fact that *r* provides no information as to the magnitude of the estimated trend, and the same temporal autocorrelation problem that dogs ordinary least squares similarly plagues significance testing for *r* [@Tian2011-ip].
@@
 
-Palaeoecology and palaeolimnology have moved away from being descriptive disciplines, rapidly adopting new statistical developments in the 1990s and beyond [@@Smol2012-xt].
-Less development has been observed in the area of trend estimation in palaeoenvironmental time series.
-Here, I describe the use of generalized additive models [GAMs; @Hastie1986-xs; @Hastie1990-bx; @Ruppert2003-pt; @Wood2017-qi; @Yee1991-mg] for trend estimation.
-GAMs, like simple linear regression, are a formal regression-based method for estimating trends, yet they are also, superficially at least, similar to \loess{}.
+Here, I describe generalized additive models [GAMs; @Hastie1986-xs; @Hastie1990-bx; @Ruppert2003-pt; @Wood2017-qi; @Yee1991-mg] for trend estimation.
+GAMs, like simple linear regression, are a regression-based method for estimating trends, yet they are also, superficially at least, similar to \loess{}.
 GAMs and \loess{} estimate smooth, non-linear trends in time series and both can handle the irregular spacing of samples in time, yet GAMs do not suffer from the subjectivity that plagues \loess{} as a method of formal statistical inference.
 
-In the subsequent sections, I present an introduction to GAMs and how they are able to estimate smooth, non-linear trends from time series.
-Next the issue of uncertainty in model-estimated trends is discussed, as is the topic of posterior simulation from a regression model.
-The issue of identifying periods of significant environmental change is considered and I show how this can be achieved using the first derivative of the estimated trend.
+In the subsequent sections, I present an introduction to GAMs and discuss the issue of uncertainty in model-estimated trends, the topic of posterior simulation from a regression model and how to identify periods of significant environmental change using the first derivative of the estimated trend.
 Two non-standard types of spline --- adaptive smoothers and Gaussian process splines --- that are especially applicable to GAMs in the palaeoenvironmental setting are subsequently described, followed by an assessment of the the impact of age-model uncertainty on trend estimation via GAMs.
@@
 
-To illustrate the estimation of trends in palaeoenvironmental data using GAMs, I use two proxy time series; a 150-year bulk organic matter δ^15^N time series from Small Water, and a 3000-year alkenone record from Braya-Sø.
+To illustrate trend estimation in palaeoenvironmental data using GAMs, I use two proxy time series; a 150-year bulk organic matter δ^15^N record from Small Water, and a 3000-year alkenone record from Braya-Sø.
 Between them, the two examples, combine many of the features of interest to palaeoecologists that motivate the use of GAMs; non-linear trends and the question of when changes in the measured proxy occurred.
-The example analyses were all performed using the *mgcv* package and R, and the supplementary material contains a fully annotated document showing the R code used to replicate all the analyses described in the remainder of the paper.
+The example analyses were all performed using the *mgcv* package [version 1.8.23; @Wood2017-qi] and R [version 3.4.4; @Rcore], and the supplementary material contains a fully annotated document showing the R code used to replicate all the analyses described in the remainder of the paper.
 
@@
 $\beta_1$ is the *slope* of the best fit line through the data; it measures the rate of change in $y$ for a unit increase in $x$.
-The unknowns, the $\beta_j$ are commonly estimated using least squares by minimising the sum of squared errors, $\sum_t \varepsilon_t^2$.
-If we want to ask if the estimated trend $\beta_1$ is statistically significant, a process called *inference*, we must make further assumptions about the data (conditional upon the fitted model) or the model errors (residuals); $\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)$.
+The unknowns, the $\beta_j$, are commonly estimated using least squares by minimising the sum of squared errors, $\sum_t \varepsilon_t^2$.
+If we want to ask if the estimated trend $\beta_1$ is statistically significant we must make further assumptions about the data (conditional upon the fitted model) or the model errors (residuals); $\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)$.
 This notation indicates that the residuals $\varepsilon_t$ are *independent* and *identically distributed* Gaussian random variables with mean equal to $0$ and constant variance $\sigma^2$.
@@
 The linear model described above is quite restrictive in terms of the types of trend it can fit; essentially linear increasing or decreasing trends, or, trivially, a null trend of no change.
-This model can be extended to allow for non-linear trends, most notably by making $y_t$ depend on polynomials of $x_t$, for example
+This model can be extended to allow for non-linear trends by making $y_t$ depend on polynomials of $x_t$, for example
 \begin{align} \label{eq:polynomial-model}
@@
 where polynomials of $x_t$ up to order $P$ are used.
-This model allows for more complex trends but it remains a fully parametric model and suffers from several problems, especially the behaviour of the fitted polynomial at the start and end of the observed series.
+This model allows for more complex trends but it remains a fully parametric model and suffers from several problems, especially the behaviour of the fitted trend at the start and end of the observed series.
 
@@
 In the remainder, I hope to demonstrate that the answer to that question is emphatically "yes"!
-Below I describe a coherent and consistent approach to modelling palaeoenvironmental time series using generalized additive models, which builds upon the linear regression framework.
+Below I describe a coherent and consistent approach to modelling palaeoenvironmental time series using generalized additive models that builds upon the linear regression framework.
 
@@
 \end{equation}
-where the linear effect of time(the $\beta_1 x_t$ part) has been replaced by a smooth function of time, $f(x_t)$.
+where the linear effect of time (the $\beta_1 x_t$ part) has been replaced by a smooth function of time, $f(x_t)$.
 The immediate advantage of the GAM is that we are no longer restricted to the shapes of trends that can be fitted via global polynomial functions such as \eqref{eq:polynomial-model}.
@@
 
-The linear model is a special case of a broader class know as the generalized linear model [GLM; @McCullagh1989-gv].
+The linear model is a special case of a broader class, known as the generalized linear model [GLM; @McCullagh1989-gv].
 The GLM provides a common framework for modelling a wide range of types of data, such as count, proportions, or binary (presence/absence) data, that are not conditionally distributed Gaussian.
@@
 $g$ is the link function, an invertible, monotonic function, such as the natural logarithm, and $g^{-1}$ is its inverse.
-The link function maps values from the response scale on to the scale of the linear predictor, whilst the inverse of thew link function provides the reverse mapping.
+The link function maps values from the response scale on to the scale of the linear predictor, whilst the inverse of the link function provides the reverse mapping.
 For example, count data are strictly non-negative integer values and are commonly modelled as a Poisson GLM/GAM using the natural log link function.
@@
 
-It is clear from the plots of the data that we require the fitted trends for the Small Water δ^15^N and Braya-Sø \uk{} time series to be non-linear functions, but it is less clear how to specify the actual shape require.
+It is clear from the plots of the data (Figure\ \ref{fig:data-figure}) that we require the fitted trends for the Small Water δ^15^N and Braya-Sø \uk{} time series to be non-linear functions, but it is less clear how to specify the actual shape require.
 Ideally, we'd like to learn the shape of the trend from the data themselves.
@@
 \end{equation*}
-where $\beta_j$ is the weight applied to the $j$th basis function. Note that here we concern ourselves only with univariate smooths involving a single covariate, the time variable $x_t$.
-
-The polynomial model we encountered earlier is an example of a statistical model that uses a basis expansion.
+where $\beta_j$ is the weight applied to the $j$th basis function.
+
+The polynomial model is an example of a statistical model that uses a basis expansion.
 For the cubic polynomial ($P = 3$) fit shown in Figure \ref{fig:polynomial-example-plot} there are in fact 4 basis functions: $b_1(x_t) = x_t^0 = 1$, $b_2(x_t) = x_t$, $b_3(x_t) = x^2_t$, and $b_4(x_t) = x_t^3$.
-Note that $b_1(x_t)$ is constant and is linked to the model intercept, $\beta_0$, in the linear model \eqref{eq:polynomial-model}, and further, that the weights are the estimated coefficients in the model, the $\beta_j$.
+Note that $b_1(x_t)$ is constant and is linked to the model intercept, $\beta_0$, in the linear model \eqref{eq:polynomial-model}, and further, that the basis function weights are the estimated coefficients in the model, the $\beta_j$.
 
@@
 
-A cubic spline is a smooth curve comprised of sections of cubic polynomials ($P = 3$), where the sections are joined together at some specified locations --- known as *knots* --- in such a way that at the joins, the two sections of cubic polynomial that meet have the same value as well as the same first and second derivative.
+A cubic spline is a smooth curve comprised of sections of cubic polynomials, where the sections are joined together at some specified locations --- known as *knots* --- in such a way that at the joins, the two sections of cubic polynomial that meet have the same value as well as the same first and second derivative.
 These properties mean that the sections join smoothly and differentiably at the knots [@Wood2017-qi, sec.
@@
 
-To estimate a model using this basis expansion each basis function forms a column in the model matrix $\boldsymbol{X}$ and the weights $\beta_j$ can be found using least squares regression (assuming a Gaussian response).
+To estimate a model using this basis expansion each basis function forms a column in the model matrix **X** and the weights $\beta_j$ can be found using least squares regression (assuming a Gaussian response).
 Note that in order to estimate a coefficient for each basis function the model has to be fitted without an intercept term.
-In practice we would include an intercept term in the model and therefore the basis functions are modified via an identifiability constraint.
+In practice we would include an intercept term in the model and therefore the basis functions are modified via an identifiability constraint [@@Wood2017-qi].
 This has the effect of making the basis orthogonal to the intercept but results in more complicated basis functions than those shown in in Figure \ref{fig:basis-function-example-plot}a.
@@
 5.5.1].
-It is unlikely in the extreme that any real data problem would involve functions of such complexity that they require as many basis functions as data.
+It is unlikely that any real data problem would involve functions of such complexity that they require as many basis functions as data.
 It is much more likely that the true functions that we attempt to estimate are far simpler than the set of functions representable by 1 basis function per unique data value.
@@
 
-To address this issue, thin plate regression splines (TPRS) have been suggested which truncate the space of the thin plate spline basis to some lower number of basis functions whilst preserving much of the advantage of the original basis as an optimally-fitting spline [@Wood2003-qy].
+To address this issue low rank thin plate regression splines (TPRS) have been suggested which truncate the space of the thin plate spline basis to some lower number of basis functions whilst preserving much of the advantage of the original basis as an optimally-fitting spline [@Wood2003-qy].
 A rank `r k` TPRS basis (i.e.
@@
 For data sets of similar size to the two examples used here, the additional computational effort required to set up the TPRS basis over the CRS basis will not be noticeable.
-For highly resolved series containing more than ~1000 observations the truncation may be too costly computationally.
-In such instances, little is lost by moving to the CRS basis, with the same number of knots as the rank of the desired TPRS, with the benefit of considerably reduced set up time for the basis.
+For highly resolved series containing more than ~1000 observations the truncation may be costly computationally.
+In such instances, little is lost by moving to the CRS basis with the same number of knots as the rank of the desired TPRS, with the benefit of considerably reduced set up time for the basis.
 
@@
 
-In practice, the size of the basis is an upper limit on the expected complexity of the trend, and a simple test is available to check if the basis used was sufficiently large [@Pya2016-rk].
+In practice, the size of the basis is an upper limit on the expected complexity of the trend, and a simple test to check if the basis used was sufficiently large [@Pya2016-rk].
 This test is available via the `gam.check()` function in *mgcv* for example, which essentially looks at whether there is any additional nonlinearity or structure in the residuals that can be explained by a further smooth of $x_t$.
@@
 
-Having identified regression splines as a useful way to represent $f$, we next need a way to decide how wiggly the fitted trend should be.
+Having identified low rank regression splines as a useful way to represent $f$, we next need a way to decide how wiggly the fitted trend should be.
 A backwards elimination approach to sequentially remove knots or basis functions might seem appropriate, however such an approach would likely fail as the resulting sequence of models would not be strictly nested, precluding many forms of statistical comparison [@Wood2017-qi].
@@
 
-The trend in δ^15^N values is clearly non-linear and it would be difficult to suggest a suitable polynomial model that would allow for periods of relatively no change in δ^15^N as well as rapid change.
+The trend in δ^15^N values is clearly non-linear but it would be difficult to suggest a suitable polynomial model that would allow for periods of relatively no change in δ^15^N as well as rapid change.
 Instead, a GAM is ideally suited to modelling such trends; the data suggest a smoothly varying change in δ^15^N between 1925 and 1975.
@@
 \end{equation}
-Note that now I have relaxed the i.i.d.
-assumption and have introduced $\boldsymbol{\Lambda}$, a correlation matrix that is used to model autocorrelation in the residuals.
-The δ^15^N values are irregularly spaced in time and a correlation structure that can handle the uneven spacing of the samples is needed [@Pinheiro2000-kx].
-A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the first-order autoregressive process (AR(1)) and, simply stated, models the correlation between any two residuals as and exponentially decreasing function of $h$ ($\phi^h$), where $h$ is the amount of separation in time between the residuals [@Pinheiro2000-kx].
+Now the i.i.d.
+assumption has been relaxed and a correlation matrix, $\boldsymbol{\Lambda}$, has been introduced that is used to model autocorrelation in the residuals.
+The δ^15^N values are irregularly spaced in time and a correlation structure that can handle the uneven spacing is needed [@Pinheiro2000-kx].
+A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the first-order autoregressive process (AR(1)) and, simply stated, models the correlation between any two residuals as an exponentially decreasing function of $h$ ($\phi^h$), where $h$ is the amount of separation in time between the residuals [@Pinheiro2000-kx].
 $h$ may be a real valued number in the CAR(1), which is how it can accommodate the irregular separation of samples in time.
@@
 Fitting this model as a standard GAM with REML smoothness selection resulted in the same fitted trend as the GAM with CAR(1) errors (not shown), whilst using GCV smoothness selection resulted in a much more satisfactory fitted trend.
-There are two potential problems with the GCV-selected trend: i) GCV is sensitive to the profile of the GCV score and has been shown to badly under smooth data in situations where the profile is flat around the minimum GCV score, and ii) the model fitted assumes that the observations are independent, an assumption that is certainly violated in the \uk{} time series.
+There are two potential problems with the GCV-selected trend: i) GCV is sensitive to the profile of the GCV score and has been shown to under smooth data in situations where the profile is flat around the minimum GCV score, and ii) the model fitted assumes that the observations are independent, an assumption that is certainly violated in the \uk{} time series.
 
@@
 The GCV score profile suggests that the potential for under smoothing identified by @Reiss2009-fk is unlikely to apply here as there is a well-defined minimum in profile.
-Addressing the violation of the independence assumption is more difficult; the standard errors of the model coefficients are most likely anti-conservative, which presents a significant problem if we wish to determine which of the wiggles in the fitted \uk{} trend can be identified as real.
 
@@
 The primary issue leading to poor fit is that neither model accounts for the different variance (known as (heteroscedasticity) of each observation in the \uk{} record.
+This seemingly isn't a problem for GCV which minimizes prediction error.
 The sediments in Braya-Sø are not annually laminated and therefore the core was sliced at regular depth intervals.
@@
 The main user selectable option when fitting a GAM in the penalised likelihood framework of @Wood2004-zv is how many basis functions to use.
-As described above, the basis should be large enough to contain the true, but unknown, function, or a close approximation to it.
+As described above, the basis should be large enough to contain the true, but unknown, function or a close approximation to it.
 For GCV selection the basis used contained 29 basis functions, whilst the CAR(1) model with REML smoothness selection would only converge with a basis containing 20 functions.
@@
 Using the test of @Pya2016-rk, the basis dimension for the models with REML smoothness selection was too small.
-To proceed therefore, we must drop the CAR(1) term and increase the basis dimension to 39 functions (by setting `k = 40`; I set it larger than expected wiggliness because the larger basis contains a richer family of functions and the excess complexity is reduced because of the smoothness penalty in play.)
+To proceed therefore, we must drop the CAR(1) term and increase the basis dimension to 39 functions (by setting `k = 40`; I set it larger than expected because the larger basis contains a richer family of functions and the excess complexity is reduced because of the smoothness penalty.)
 
 With the larger basis dimension and accounting for the non-constant variance of the observations via weights, the model fitted using REML is indistinguishable from that obtained using GCV (Figure \ref{fig:plot-fitted-models}b).
-The trace of the REML score for this model (Figure \ref{fig:trace-smoothness-parameters}c) shows a pronounced minimum at a much smaller value of $\lambda$ than the original REML fit (Figure \ref{fig:trace-smoothness-parameters}b), indicating that a more wiggly trend provides a better fit to the Braya-Sø time series.
+The trace of the REML score for this model shows a pronounced minimum at a much smaller value of $\lambda$ than the original REML fit (Figure \ref{fig:trace-smoothness-parameters}), indicating that a more wiggly trend provides a better fit to the Braya-Sø time series.
 This example illustrates that some care and understanding of the underlying principles of GAMs is required to diagnose potential issues with the estimated model.
@@
 A 95% interval would be drawn at $\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))$, with $m_{1-\alpha} = 1.96$, the 0.95 probability quantile of a standard normal distribution, and $\text{SE}(\hat{y}_t)$ is the standard error of the estimated trend at time $x_t$.
-This type of confidence interval would normally be described as *pointwise*; the coverage properties of the interval are correct for a single point on the fitted trend, but, if we were to consider additional points on the trend, the coverage would logically be lower than 1 - $\alpha$.
+This type of confidence interval would normally be described as *pointwise*; the coverage properties of the interval being correct for a single point on the fitted trend, but, if we were to consider additional points on the trend, the coverage would logically be lower than 1 - $\alpha$.
 This is the traditional frequentist interpretation of a confidence interval.
-However, the approach to fitting GAMs described here has a Bayesian interpretation [@Kimeldorf1970-cn; @Wahba1990-ag; @Wahba1983-eh; @Silverman1985-kw] and therefore the typical frequentist interpretation does not apply.
+However, the GAM described here has a Bayesian interpretation [@Kimeldorf1970-cn; @Wahba1990-ag; @Wahba1983-eh; @Silverman1985-kw] and therefore the typical frequentist interpretation does not apply.
 @Nychka1988-rz investigated the properties of a confidence interval created as described above using standard errors derived from the Bayesian posterior covariance matrix for the estimated mode parameters.
@@
 
-Whilst the *across-the-function* or average coverage frequentist interpretation of the Bayesian credible intervals is useful, if may be important to have an interval that contains the entirety of the true function with some state probability (1 - $\alpha$).
+Whilst the *across-the-function* frequentist interpretation of the Bayesian credible intervals is useful, if may be important to have an interval that contains the entirety of the true function with some state probability (1 - $\alpha$).
 Such an interval is known as a *simultaneous* interval.
-A (1 - $\alpha$)100% *simultaneous* confidence interval contains *in their entirety* 1 - $\alpha$ of all random draws from the posterior distribution of the fitted model.
+A (1 - $\alpha$)100% simultaneous confidence interval contains *in their entirety* 1 - $\alpha$ of all random draws from the posterior distribution of the fitted model.
 
@@
 Together, these coefficients are distributed multivariate normal with mean vector and covariance matrix specified by the model estimates of the coefficients and their covariances respectively.
-Random draws from this distribution can be taken, where each random draw represents a new trend that is consistent with the fitted trend but also includes the uncertainty in the estimated trend.
+Random draws from this distribution can be taken, where each random draw represents a new trend that is consistent with the fitted trend but also reflects the uncertainty in the estimated trend.
 This process is known as *posterior simulation*.
@@
 We see a similar pattern in Figure\ \ref{fig:posterior-simulation}b for the Braya-Sø record; the large peak in \uk{} at ~250CE and the strong decline at ~1200CE are well defined in the posterior simulations, whereas most of the localised trends that are smaller magnitude changes in $y_t$ are associated with posterior simulations that are less well constrained with the ends of the record in particular showing considerable variation in the strength, timing, and even sign of simulated trends, reflecting the greater uncertainty in estimated trend during these periods.
-For the random draws illustrated in Figure \ref{fig:posterior-simulation}, a simultaneous interval should contain the entire function for on average `r round(take * 0.95, 0)` of the `r take` draws.
+For the random draws illustrated in Figure \ref{fig:posterior-simulation}, a (1 - $\alpha$)100% simultaneous interval should contain the entire function for on average `r round(take * 0.95, 0)` of the `r take` draws.
 
@@
 Figure \ref{fig:derivatives} shows the estimated first derivative of the fitted trend in the Small Water (\ref{fig:derivatives}a) and Braya-Sø (\ref{fig:derivatives}b) time series.
-Although the estimated trend suggests a slight increase in δ^15^N from the start of the record to ~1940, the estimated trend is sufficiently uncertain that the simultaneous interval on the first derivative includes 0 throughout this period.
+Although the estimated trend suggests a slight increase in δ^15^N from the start of the record to ~1940, the estimated trend is sufficiently uncertain that the simultaneous interval on the first derivative includes 0 throughout.
 We can understand why this is so by looking at the posterior simulations in Figure\ \ref{fig:posterior-simulation}a; there is considerable variation in the shape of the simulated trends throughout this period.
@@
 The estimated value of $\phi$ for the δ^15^N series is `r round(smallPhi[2], 2)` with 95% confidence interval `r round(smallPhi[1], 2)`--`r round(smallPhi[3], 2)`, indicating moderate to strong residual autocorrelation about the fitted trend.
-The correlation function is an exponentially decreasing function of temporal separation ($\Delta_t$), and whilst observations that are a few year apart are quite strongly dependent on one another, this dependence drops off rapidly as $\Delta_t$ increases and is effectively zero when samples are separated by a decade or more (Figure\ \ref{fig:car1-plot}).
-
-Failure to account for the dependencies in the δ^15^N time series could lead to the estimation of a more wiggly trend than the one shown in Figure \ref{fig:plot-fitted-models}a which would negatively impact the confidence place on the inferences we might draw from the fitted model.
+The correlation function is an exponentially decreasing function of temporal separation ($\Delta_t$), and whilst observations that are a few years apart are quite strongly dependent on one another, this dependence drops off rapidly as $\Delta_t$ increases and is effectively zero when samples are separated by a decade or more (Figure\ \ref{fig:car1-plot}).
+
+Failure to account for the dependencies in the δ^15^N time series could lead to the estimation of a more wiggly trend than the one shown in Figure \ref{fig:plot-fitted-models}a which would negatively impact the confidence placed on the inferences we might draw from the fitted model.
 Importantly, failing to account for the strong dependency in the residuals would lead to smaller uncertainties in the estimated spline coefficients, which would propagate through to narrower confidence intervals on the fitted trend and on the first derivatives, and ultimately to the identification of significant periods of change.
@@
 
-Problems estimating the GAM plus CAR(1) model were encountered when this was fitted to the \uk{} time series; both a smooth trend in the mean \uk{} and a CAR(1) process in the residuals lead to an unidentifiable model.
+Problems estimating the GAM plus CAR(1) model were encountered when this was fitted to the \uk{} time series; including both a smooth trend in the mean \uk{} and a CAR(1) process in the residuals lead to an unidentifiable model.
 What makes a model with a spline-based trend and an autocorrelation process like the CAR(1) potentially unidentifiable?
@@
 This leads to runs of values that are consistently greater (less) than the overall mean.
-Short runs would indicate weaker autocorrelation whilst longer runs are associated with stronger autocorrelation, and long runs of values greater (less) than the mean would evident as non-linear trends in the time series.
+Short runs would indicate weaker autocorrelation whilst longer runs are associated with stronger autocorrelation, and long runs of values greater (less) than the mean would be evident as non-linear trends in the time series.
 As a result, a wiggly trend and an autocorrelation function with large $\phi$ are two ways to describe the same pattern of values in a time series, and without any further information to constrain either the model is unable to distinguish both components uniquely.
@@
 
-With a Gaussian process we are interested in fitting a smooth temporal trend by modelling the way the correlation between pairs of observations varies as a function of the distance, $h$, in time that separates ($h$) the observations.
+With a Gaussian process we are interested in fitting a smooth temporal trend by modelling the way the correlation between pairs of observations varies as a function of the distance, $h$, in time that separates the observations.
 The correlation between pairs of observations decreases with increasing separation, which is modelled using a correlation function, $c(h)$.
-Figure \ref{fig:gp-correlation-functions} shows examples of two different correlation functions; the *power exponential* (Figure \ref{fig:gp-correlation-functions}a), and the Matérn (Figure \ref{fig:gp-correlation-functions}b) correlation functions.
-These functions are smooth and monotonic-decreasing, meaning that the value of the correlation function decreases with increasing separation ($h$).
-When $h$ = 0, the correlation is equal to 1 ($c(0) = 1$); two samples taken at exactly the same time point are perfectly correlated.
-As $h \rightarrow \infty$, the correlation tends to zero ($c(h) \rightarrow 0$); two samples separated by a large amount of time tend to be uncorrelated.
-Often we are interested in learning how large the separation in time needs to be before, on average, a pair of observations is effectively uncorrelated (i.e.
-where $c(h)$ is sufficiently close to zero).
 
@@
 ```
+Figure \ref{fig:gp-correlation-functions} shows examples of two different correlation functions; the *power exponential* (Figure \ref{fig:gp-correlation-functions}a), and the Matérn (Figure \ref{fig:gp-correlation-functions}b) correlation functions.
+These functions are smooth and monotonic-decreasing, meaning that the value of the correlation function decreases with increasing separation ($h$).
+When $h$ = 0, the correlation is equal to 1 ($c(0) = 1$); two samples taken at exactly the same time point are perfectly correlated.
+As $h \rightarrow \infty$, the correlation tends to zero ($c(h) \rightarrow 0$); two samples separated by a large amount of time tend to be uncorrelated.
+Often we are interested in learning how large the separation in time needs to be before, on average, a pair of observations is effectively uncorrelated (i.e.
+where $c(h)$ is sufficiently close to zero).
 
@@
 It can be shown [e.g.
-@Fahrmeir2013-xu section/pages] that the Gaussian process model has the same penalised likelihood form as the GAM that we discussed earlier; the observations are the knots of the smoother and each has a basis function in the form of a correlation function.
+@Fahrmeir2013-xu] that the Gaussian process model has the same penalised likelihood form as the GAM that we discussed earlier; the observations are the knots of the smoother and each has a basis function in the form of a correlation function.
 The equivalence is only true if the basis functions do not depend on any other parameters of the model, which is only achievable if the value of $\phi$ is fixed and known [@Fahrmeir2013-xu].
@@
 362--363].
-This involves proposing a value of $\phi$ for the effective range of the correlation function and then estimating the resulting GAM by minimising the penalised log-likehood conditional upon this value of $\phi$.
+This involves proposing a value of $\phi$ for the effective range of the correlation function and then estimating the resulting GAM by minimising the penalised log-likehood conditional upon this value of $\phi$ and repeating for a range of values for $\phi$.
 The model, and its corresponding value of $\phi$, with lowest penalised log-likelihood or score statistic is then retained as the estimated GAM.
@@
 Each of the spline types that I have discussed so far share a common feature; the degree of wiggliness over the time series is fixed due to the use of a single smoothness parameter, $\lambda$.
-The definition of wiggliness, as the integrated squared second derivative of the spline, ensures that fitted smoother does not jump about wildly.
+The definition of wiggliness, as the integrated squared second derivative of the spline, ensures that the fitted smoother does not jump about wildly.
 This assumes that the data themselves are well described by a smoothly varying trend.
@@
 5.3.5]; in other words, the adaptive smoother allows the wiggliness of the estimated trend to vary smoothly over time.
-Whilst this does allow the estimated trend to adapt to periods of rapid change in the response, adaptive smoothers make significant demands on the data [@Wood2017-qi, sec.
+Whilst this allows the estimated trend to adapt to periods of rapid change in the response, adaptive smoothers make significant demands on the data [@Wood2017-qi, sec.
 5.3.5]; if we used $m$ smoothness penalties to allow the wiggliness to vary over a time series, it would be like estimating $m$ separate smooths from chunks of the original series each of length $n/m$.
@@
 Therefore, should we conclude that there is no trend in \uk{} and thence climate in this period?
-I believe that to be too-strong a statement; those oscillations in \uk{} may be real responses to climate forcing but we may simply lack the statistical power to distinguish them from the null hypothesis of no trend through this period.
+I believe that to be too-strong a statement; those oscillations in \uk{} may be real responses to climate forcing but may simply lack the statistical power to distinguish them from the null hypothesis of no trend through this period.
 The adaptive smoother is only adjusting to the data available to it; just because it does not detect a trend during this period does not lend itself to an interpretation of stable climate forcing or complacency in the lake's response to forcing.
@@
 The fitted age-depth model is constrained to be monotonically decreasing with increasing depth, following the method of [@Pya2015-wr] using the *scam* package [@pya-scam-pkg].
-Also shown are 25 simulations from the model posterior of the monotonically-constrained GAM.
+Also shown are 25 simulations from the posterior distribution of the monotonically-constrained GAM.
 Each simulation from the posterior distribution of the age-model is itself a potential age-depth model, which can be used to assign dates to the Small Water core.
@@
 With one main exception it is not possible to directly apply the GAM methodology described here to multivariate abundance data, where the aim is to model all species at once.
-The *mgcv* software for example, is not able to estimate the penalized GAM for multiple non-Gaussian responses.
+The *mgcv* software, for example, is not able to estimate the penalized GAM for multiple non-Gaussian responses.
 The exception is for a small number of correlated Gaussian responses; these could be modelled as being distributed multivariate normal conditional upon the covariates.
@@
 GAMs, whilst similar to the popular \loess{} smoother, provide a superior alternative approach to trend estimation in palaeoenvironmental time series.
-They can estimate non-linear trends, provide estimates of the magnitude of change as well as allow the identification of periods of change, can account for the lack of independence (either via autocorrelation processes or via the fitting of a wiggly trend), and provide a formal framework for statistical inference on each of these features.
-
-In presenting the GAM with specific palaeoenvironmental examples and addressing the issues that arise in palaeoenvironmental time series, it is hoped that palaeoecologists and palaeolimnologists will be motivated to give greater consideration to the estimation of trends and the identification of change.
+GAMs can estimate non-linear trends, provide estimates of the magnitude of change as well as allow the identification of periods of change, can account for the lack of independence (either via autocorrelation processes or via the fitting of a wiggly trend), and provide a formal framework for statistical inference on each of these features.
+
+In presenting the GAM with specific palaeoenvironmental examples and addressing the issues that arise in palaeoenvironmental time series, it is hoped that palaeoecologists and palaeolimnologists will be motivated to give greater consideration to the estimation of trends and the identification of change in stratigraphic time series.
 
@@
 The ideas expressed in this paper are the result of many fruitful conversations with colleagues at the Environmental Change Research Centre, UCL, and the University of Regina.
-In particular I am indebted to Helen Bennion, Rick Battarbee, and Peter Leavitt for their collaborations on projects over many years.
+In particular I am indebted to Helen Bennion, Rick Battarbee, and Peter Leavitt for their collaborations on projects over many years, and to David Miller, Eric Pedersen, and Noam Ross, my GAM workshop partners in crime.
 Without Simon Wood's *mgcv* software and his research on GAMs, the application of these models to palaeo time series would not be as straight forward.
-I also thank David Miller, Eric Pedersen, and Noam Ross, my GAM workshop partners in crime.
 This work was supported by a Natural Sciences and Engineering Council of Canada (NSERC) Discovery Grant to the author (RGPIN-2014-04032).

### 2018-05-14 manuscript.tex
submitted version of the manuscript

@@
 
-The vast majority of data produced by palaeoecologists and
+Palaeoecology and palaeolimnology have moved away from being descriptive
+disciplines, rapidly adopting new statistical developments in the 1990s
+and beyond (@ Smol et al., 2012). Less development has been observed in
+the area of trend estimation in palaeoenvironmental time series. The
+vast majority of data produced by palaeoecologists and
 palaeolimnologists is in the form of time-ordered observations on one or
@@
 in accumulation rates over time and compaction by overlying sediments,
-results in an uneven sampling in time. An under-appreciated secondary
-feature of such sampling is that younger sediments often have larger
-variance than older sediments; each section of core represents fewer
-lake years in newer samples, relative to older samples. This variable
-time averaging acts as a variable low-pass (high-cut) filter of the raw,
-annual depositional signal.
-
-The irregular intervals between samples means that the customary
-time-series analysis methods of autoregressive or moving average
-processes, in the form of autoregressive integrated moving average
-(ARIMA) models, are practically impossible to apply because software
-implementing these methods typically requires even spacing of
+results in an uneven sampling in time. An under-appreciated feature of
+such sampling is that younger sediments often have larger variance than
+older sediments; each section of core represents fewer lake years in
+newer samples, relative to older samples. This variable averaging acts
+as a time-varying low-pass (high-cut) filter of the annual depositional
+signal.
+
+Irregular intervals between samples means that the time-series analysis
+methods of autoregressive or moving average processes, in the form of
+autoregressive integrated moving average (ARIMA) models, are practically
+impossible to apply because software typically requires even spacing of
 observations in time. Dutilleul et al. (2012) and Birks (2012a),
@@
 Where statistical approaches have been applied to trend estimation in
-palaeoenvironmental time series, the most commonly used has been
-\loess{} (Birks, 1998, 2012a; Cleveland, 1979; Juggins and Telford,
-2012). \loess{}, or locally weighted scatterplot smoother, as it's name
-suggests, was developed to smooth x-y scatterplot data, ostensibly as
-part of exploratory data analysis (Cleveland, 1979). The method fits a
-smooth line through data by fitting weighted least squares (WLS) models
-to observations within a particular, user-specified window of the target
-point, whose width is typically expressed as a proportion \(\alpha\) of
-the \(n\) data points. Weights are determined by how close (in the
-x-axis only) an observation in the window is to the focal point, with a
-tricubic function, by default, giving greatest weight to points closest
-to the focal point. The interim \loess{}-smoothed value for the focal
-point is the predicted value from the weighted regression at the focal
-point. The interim values are updated using weights based on how far in
-the y-axis direction, the interim smoothed value lies from the observed
-value plus the x-axis distance weights; this has the effect of
-down-weighting the effects of outlier observations. The final \loess{}
-is obtained by joining up the smoothed (fitted) values. The user has to
-choose how large a window to use, whether to fit degree 1 (linear) or
-degree 2 (quadratic) polynomials in the WLS model, and how to weight
-points in the x-axis. When used in an exploratory mode, the user has
-considerable freedom to choose the detail of the \loess{} fit; the
-window width, for example, can be infinitely tweaked to give as close a
-fit to the data, as assessed by eye, as is desired. Using
-cross-validation (CV) to choose \(\alpha\) or the degree of polynomial
-in the WLS model is complicated for a number of reasons, not least that
-the CV scheme used must involve the time ordering of the data (e.g.
-Bergmeir et al., 2018). This level of subjectivity is problematic
-however once we wish to move beyond exploratory data analysis to
-statistically identify trends and to test hypotheses involving those
-trend estimates.
+palaeoenvironmental time series, a commonly-used method is \loess{}
+(Birks, 1998, 2012a; Cleveland, 1979; Juggins and Telford, 2012).
+\loess{}, locally weighted scatterplot smoother, as it's name suggests,
+was developed to smooth x-y scatterplot data (Cleveland, 1979). The
+method fits a smooth line through data by fitting weighted least squares
+(WLS) models to observations within a particular, user-specified window
+of the focal point, whose width is typically expressed as a proportion
+\(\alpha\) of the \(n\) data points. Weights are determined by how close
+(in the x-axis only) an observation in the window is to the focal point
+giving greatest weight given to points closest to the focal point. The
+interim \loess{}-smoothed value for the focal point is the predicted
+value from the weighted regression at the focal point. The interim
+values are updated using weights based on how far in the y-axis
+direction the interim smoothed value lies from the observed value plus
+the x-axis distance weights; this has the effect of down-weighting
+outlier observations. The final \loess{} is obtained by joining the
+smoothed values. The user has to choose how large a window to use,
+whether to fit degree 1 (linear) or degree 2 (quadratic) polynomials in
+the WLS model, and how to weight points in the x-axis. When used in an
+exploratory mode, the user has considerable freedom to choose the detail
+of the \loess{} fit; the window width, for example, can be infinitely
+tweaked to give as close a fit to the data, as assessed by eye, as is
+desired. Using cross-validation (CV) to choose \(\alpha\) or the degree
+of polynomial in the WLS model is complicated for a number of reasons,
+not least because the CV scheme used must involve the time ordering of
+the data (e.g. Bergmeir et al., 2018). This subjectivity is problematic
+however once we wish to move beyond exploratory analysis and
+statistically identify trends to test hypotheses involving those trend
+estimates.
 
@@
 including the filter width. Furthermore, the width of the filter causes
-boundary issues; with a centred filter, of width five data points, the
-filtered time series would be two data points shorter at both ends of
-the series because the filter weights are not defined for the first and
-last two observations of the original series as these time points were
-not observed. Considerable research effort has been expended to identify
+boundary issues; with a centred filter, of width five, the filtered time
+series would be two data points shorter at both ends of the series
+because the filter values are not defined for the first and last two
+observations of the original series as these extra time points were not
+observed. Considerable research effort has been expended to identify
 ways to pad the original time series at one or both ends to maintain the
-original length in the filtered series, without introducing bias because
-of the method of padding (e.g. Mann, 2004, 2008; Mills, 2006, 2007,
-2010).
-
-These are by no means the only methods that have been used to estimated
-trends in stratigraphic series. Another common approach involves fitting
-a simple linear trend using ordinary least squares regression and use
-the resulting \emph{t} statistic as evidence against the null hypothesis
-of no trend despite the statistical assumptions of the method being
-almost surely violated due to dependence among observations arising from
-temporal autocorrelation in the series. The Pearson correlation
+original length in the filtered series, without introducing bias due to
+the padding (e.g. Mann, 2004, 2008; Mills, 2006, 2007, 2010).
+
+These are not the only methods that have been used to estimated trends
+in stratigraphic series. Another common approach involves fitting a
+simple linear trend using ordinary least squares regression and use the
+resulting \emph{t} statistic as evidence against the null hypothesis of
+no trend despite the statistical assumptions being almost surely
+violated due to dependence among observations. The Pearson correlation
 coefficient, \emph{r}, is also often used to detect trends in palaeo
@@
 
-Palaeoecology and palaeolimnology have moved away from being descriptive
-disciplines, rapidly adopting new statistical developments in the 1990s
-and beyond (@ Smol et al., 2012). Less development has been observed in
-the area of trend estimation in palaeoenvironmental time series. Here, I
-describe the use of generalized additive models (GAMs; Hastie and
+Here, I describe generalized additive models (GAMs; Hastie and
 Tibshirani, 1986, 1990; Ruppert et al., 2003; Wood, 2017; Yee and
 Mitchell, 1991) for trend estimation. GAMs, like simple linear
-regression, are a formal regression-based method for estimating trends,
-yet they are also, superficially at least, similar to \loess{}. GAMs and
+regression, are a regression-based method for estimating trends, yet
+they are also, superficially at least, similar to \loess{}. GAMs and
 \loess{} estimate smooth, non-linear trends in time series and both can
@@
 
-In the subsequent sections, I present an introduction to GAMs and how
-they are able to estimate smooth, non-linear trends from time series.
-Next the issue of uncertainty in model-estimated trends is discussed, as
-is the topic of posterior simulation from a regression model. The issue
-of identifying periods of significant environmental change is considered
-and I show how this can be achieved using the first derivative of the
+In the subsequent sections, I present an introduction to GAMs and
+discuss the issue of uncertainty in model-estimated trends, the topic of
+posterior simulation from a regression model and how to identify periods
+of significant environmental change using the first derivative of the
 estimated trend. Two non-standard types of spline --- adaptive smoothers
@@
 
-To illustrate the estimation of trends in palaeoenvironmental data using
-GAMs, I use two proxy time series; a 150-year bulk organic matter
-δ\textsuperscript{15}N time series from Small Water, and a 3000-year
-alkenone record from Braya-Sø. Between them, the two examples, combine
-many of the features of interest to palaeoecologists that motivate the
-use of GAMs; non-linear trends and the question of when changes in the
-measured proxy occurred. The example analyses were all performed using
-the \emph{mgcv} package and R, and the supplementary material contains a
-fully annotated document showing the R code used to replicate all the
-analyses described in the remainder of the paper.
+To illustrate trend estimation in palaeoenvironmental data using GAMs, I
+use two proxy time series; a 150-year bulk organic matter
+δ\textsuperscript{15}N record from Small Water, and a 3000-year alkenone
+record from Braya-Sø. Between them, the two examples, combine many of
+the features of interest to palaeoecologists that motivate the use of
+GAMs; non-linear trends and the question of when changes in the measured
+proxy occurred. The example analyses were all performed using the
+\emph{mgcv} package (version 1.8.23; Wood, 2017) and R (version 3.4.4; R
+Core Team, 2018), and the supplementary material contains a fully
+annotated document showing the R code used to replicate all the analyses
+described in the remainder of the paper.
 
@@
 it measures the rate of change in \(y\) for a unit increase in \(x\).
-The unknowns, the \(\beta_j\) are commonly estimated using least squares
-by minimising the sum of squared errors, \(\sum_t \varepsilon_t^2\). If
-we want to ask if the estimated trend \(\beta_1\) is statistically
-significant, a process called \emph{inference}, we must make further
+The unknowns, the \(\beta_j\), are commonly estimated using least
+squares by minimising the sum of squared errors,
+\(\sum_t \varepsilon_t^2\). If we want to ask if the estimated trend
+\(\beta_1\) is statistically significant we must make further
 assumptions about the data (conditional upon the fitted model) or the
@@
 trends, or, trivially, a null trend of no change. This model can be
-extended to allow for non-linear trends, most notably by making \(y_t\)
-depend on polynomials of \(x_t\), for example
+extended to allow for non-linear trends by making \(y_t\) depend on
+polynomials of \(x_t\), for example
 
@@
 and suffers from several problems, especially the behaviour of the
-fitted polynomial at the start and end of the observed series.
+fitted trend at the start and end of the observed series.
 
@@
 Below I describe a coherent and consistent approach to modelling
-palaeoenvironmental time series using generalized additive models, which
+palaeoenvironmental time series using generalized additive models that
 builds upon the linear regression framework.
@@
 
-where the linear effect of time(the \(\beta_1 x_t\) part) has been
+where the linear effect of time (the \(\beta_1 x_t\) part) has been
 replaced by a smooth function of time, \(f(x_t)\). The immediate
@@
 
-The linear model is a special case of a broader class know as the
+The linear model is a special case of a broader class, known as the
 generalized linear model (GLM; McCullagh and Nelder, 1989). The GLM
@@
 function maps values from the response scale on to the scale of the
-linear predictor, whilst the inverse of thew link function provides the
+linear predictor, whilst the inverse of the link function provides the
 reverse mapping. For example, count data are strictly non-negative
@@
 
-It is clear from the plots of the data that we require the fitted trends
-for the Small Water δ\textsuperscript{15}N and Braya-Sø \uk{} time
-series to be non-linear functions, but it is less clear how to specify
-the actual shape require. Ideally, we'd like to learn the shape of the
-trend from the data themselves. We will refer to these non-linear
-functions as \emph{smooth functions}, or just \emph{smooths} for short,
-and we will denote a smooth using \(f(x_t)\). Further, we would like to
-represent the smooths in a way that \eqref{eq:gam} is represented
-parametrically so that it can be estimate within the well-studied GLM
-framework. This is achieved by representing the smooth using a
-\emph{basis}. A basis is a set of functions that collectively span a
-space of smooths that, we hope, contains the true \(f(x_t)\) or a close
-approximation to it. The functions in the basis are known as \emph{basis
-functions}, and arise from a \emph{basis expansion} of a covariate.
-Writing \(b_j(x_t)\) as the \(j\)th basis function of \(x_t\), the
-smooth \(f(x_t)\) can be represented as a weighted sum of basis
-functions
+It is clear from the plots of the data (Figure~\ref{fig:data-figure})
+that we require the fitted trends for the Small Water
+δ\textsuperscript{15}N and Braya-Sø \uk{} time series to be non-linear
+functions, but it is less clear how to specify the actual shape require.
+Ideally, we'd like to learn the shape of the trend from the data
+themselves. We will refer to these non-linear functions as \emph{smooth
+functions}, or just \emph{smooths} for short, and we will denote a
+smooth using \(f(x_t)\). Further, we would like to represent the smooths
+in a way that \eqref{eq:gam} is represented parametrically so that it
+can be estimate within the well-studied GLM framework. This is achieved
+by representing the smooth using a \emph{basis}. A basis is a set of
+functions that collectively span a space of smooths that, we hope,
+contains the true \(f(x_t)\) or a close approximation to it. The
+functions in the basis are known as \emph{basis functions}, and arise
+from a \emph{basis expansion} of a covariate. Writing \(b_j(x_t)\) as
+the \(j\)th basis function of \(x_t\), the smooth \(f(x_t)\) can be
+represented as a weighted sum of basis functions
 
@@
 where \(\beta_j\) is the weight applied to the \(j\)th basis function.
-Note that here we concern ourselves only with univariate smooths
-involving a single covariate, the time variable \(x_t\).
-
-The polynomial model we encountered earlier is an example of a
-statistical model that uses a basis expansion. For the cubic polynomial
-(\(P = 3\)) fit shown in Figure \ref{fig:polynomial-example-plot} there
-are in fact 4 basis functions: \(b_1(x_t) = x_t^0 = 1\),
-\(b_2(x_t) = x_t\), \(b_3(x_t) = x^2_t\), and \(b_4(x_t) = x_t^3\). Note
-that \(b_1(x_t)\) is constant and is linked to the model intercept,
-\(\beta_0\), in the linear model \eqref{eq:polynomial-model}, and
-further, that the weights are the estimated coefficients in the model,
-the \(\beta_j\).
+
+The polynomial model is an example of a statistical model that uses a
+basis expansion. For the cubic polynomial (\(P = 3\)) fit shown in
+Figure \ref{fig:polynomial-example-plot} there are in fact 4 basis
+functions: \(b_1(x_t) = x_t^0 = 1\), \(b_2(x_t) = x_t\),
+\(b_3(x_t) = x^2_t\), and \(b_4(x_t) = x_t^3\). Note that \(b_1(x_t)\)
+is constant and is linked to the model intercept, \(\beta_0\), in the
+linear model \eqref{eq:polynomial-model}, and further, that the basis
+function weights are the estimated coefficients in the model, the
+\(\beta_j\).
 
@@
 A cubic spline is a smooth curve comprised of sections of cubic
-polynomials (\(P = 3\)), where the sections are joined together at some
-specified locations --- known as \emph{knots} --- in such a way that at
-the joins, the two sections of cubic polynomial that meet have the same
-value as well as the same first and second derivative. These properties
-mean that the sections join smoothly and differentiably at the knots
-(Wood, 2017, 5.3.1).
+polynomials, where the sections are joined together at some specified
+locations --- known as \emph{knots} --- in such a way that at the joins,
+the two sections of cubic polynomial that meet have the same value as
+well as the same first and second derivative. These properties mean that
+the sections join smoothly and differentiably at the knots (Wood, 2017,
+5.3.1).
 
@@
 To estimate a model using this basis expansion each basis function forms
-a column in the model matrix \(\boldsymbol{X}\) and the weights
-\(\beta_j\) can be found using least squares regression (assuming a
-Gaussian response). Note that in order to estimate a coefficient for
-each basis function the model has to be fitted without an intercept
-term. In practice we would include an intercept term in the model and
-therefore the basis functions are modified via an identifiability
-constraint. This has the effect of making the basis orthogonal to the
-intercept but results in more complicated basis functions than those
-shown in in Figure \ref{fig:basis-function-example-plot}a.
+a column in the model matrix \textbf{X} and the weights \(\beta_j\) can
+be found using least squares regression (assuming a Gaussian response).
+Note that in order to estimate a coefficient for each basis function the
+model has to be fitted without an intercept term. In practice we would
+include an intercept term in the model and therefore the basis functions
+are modified via an identifiability constraint (@ Wood, 2017). This has
+the effect of making the basis orthogonal to the intercept but results
+in more complicated basis functions than those shown in in Figure
+\ref{fig:basis-function-example-plot}a.
 
@@
 combinations of covariate values in a data set (Wood, 2017, 5.5.1). It
-is unlikely in the extreme that any real data problem would involve
-functions of such complexity that they require as many basis functions
-as data. It is much more likely that the true functions that we attempt
-to estimate are far simpler than the set of functions representable by 1
-basis function per unique data value. From a practical point of view, it
-is also highly inefficient to carry around all these basis functions
-whilst model fitting, and the available computational resources would
-become quickly exhausted for large time series with many observations.
-
-To address this issue, thin plate regression splines (TPRS) have been
-suggested which truncate the space of the thin plate spline basis to
-some lower number of basis functions whilst preserving much of the
+is unlikely that any real data problem would involve functions of such
+complexity that they require as many basis functions as data. It is much
+more likely that the true functions that we attempt to estimate are far
+simpler than the set of functions representable by 1 basis function per
+unique data value. From a practical point of view, it is also highly
+inefficient to carry around all these basis functions whilst model
+fitting, and the available computational resources would become quickly
+exhausted for large time series with many observations.
+
+To address this issue low rank thin plate regression splines (TPRS) have
+been suggested which truncate the space of the thin plate spline basis
+to some lower number of basis functions whilst preserving much of the
 advantage of the original basis as an optimally-fitting spline (Wood,
@@
 containing more than \textasciitilde{}1000 observations the truncation
-may be too costly computationally. In such instances, little is lost by
-moving to the CRS basis, with the same number of knots as the rank of
-the desired TPRS, with the benefit of considerably reduced set up time
-for the basis.
+may be costly computationally. In such instances, little is lost by
+moving to the CRS basis with the same number of knots as the rank of the
+desired TPRS, with the benefit of considerably reduced set up time for
+the basis.
 
@@
 In practice, the size of the basis is an upper limit on the expected
-complexity of the trend, and a simple test is available to check if the
-basis used was sufficiently large (Pya and Wood, 2016). This test is
-available via the \texttt{gam.check()} function in \emph{mgcv} for
-example, which essentially looks at whether there is any additional
-nonlinearity or structure in the residuals that can be explained by a
-further smooth of \(x_t\). Should a smooth term in the fitted model fail
-this test the model can be refitted using a larger basis expansion, say
-by doubling the value of \emph{k} (the rank) used to fit the original.
-Note also that a smooth might fail this test whilst using fewer
-effective degrees of freedom than the maximum possible for the dimension
-of basis used. This may happen when the true function lies at the upper
-limit of the set of functions encompassed by the size of basis used.
-Additionally, a basis of size \(2k\) encompasses a richer space of
-functions of a given complexity than a basis of size \(k\) (Wood, 2017);
-increasing the basis dimension used to fit the model may unlock this
-additional function space resulting in a better fitting model whilst
-using a similar number of effective degrees of freedom.
+complexity of the trend, and a simple test to check if the basis used
+was sufficiently large (Pya and Wood, 2016). This test is available via
+the \texttt{gam.check()} function in \emph{mgcv} for example, which
+essentially looks at whether there is any additional nonlinearity or
+structure in the residuals that can be explained by a further smooth of
+\(x_t\). Should a smooth term in the fitted model fail this test the
+model can be refitted using a larger basis expansion, say by doubling
+the value of \emph{k} (the rank) used to fit the original. Note also
+that a smooth might fail this test whilst using fewer effective degrees
+of freedom than the maximum possible for the dimension of basis used.
+This may happen when the true function lies at the upper limit of the
+set of functions encompassed by the size of basis used. Additionally, a
+basis of size \(2k\) encompasses a richer space of functions of a given
+complexity than a basis of size \(k\) (Wood, 2017); increasing the basis
+dimension used to fit the model may unlock this additional function
+space resulting in a better fitting model whilst using a similar number
+of effective degrees of freedom.
 
@@
 
-Having identified regression splines as a useful way to represent \(f\),
-we next need a way to decide how wiggly the fitted trend should be. A
-backwards elimination approach to sequentially remove knots or basis
-functions might seem appropriate, however such an approach would likely
-fail as the resulting sequence of models would not be strictly nested,
-precluding many forms of statistical comparison (Wood, 2017).
-Alternatively, we could keep the basis dimension at a fixed size but
-guard against fitting very complex models through the use of a
-wiggliness penalty.
+Having identified low rank regression splines as a useful way to
+represent \(f\), we next need a way to decide how wiggly the fitted
+trend should be. A backwards elimination approach to sequentially remove
+knots or basis functions might seem appropriate, however such an
+approach would likely fail as the resulting sequence of models would not
+be strictly nested, precluding many forms of statistical comparison
+(Wood, 2017). Alternatively, we could keep the basis dimension at a
+fixed size but guard against fitting very complex models through the use
+of a wiggliness penalty.
 
@@
 
-The trend in δ\textsuperscript{15}N values is clearly non-linear and it
+The trend in δ\textsuperscript{15}N values is clearly non-linear but it
 would be difficult to suggest a suitable polynomial model that would
@@
 
-Note that now I have relaxed the i.i.d. assumption and have introduced
-\(\boldsymbol{\Lambda}\), a correlation matrix that is used to model
+Now the i.i.d. assumption has been relaxed and a correlation matrix,
+\(\boldsymbol{\Lambda}\), has been introduced that is used to model
 autocorrelation in the residuals. The δ\textsuperscript{15}N values are
 irregularly spaced in time and a correlation structure that can handle
-the uneven spacing of the samples is needed (Pinheiro and Bates, 2000).
-A continuous time first-order autoregressive process (CAR(1)) is a
-reasonable choice; it is the continuous-time equivalent of the
-first-order autoregressive process (AR(1)) and, simply stated, models
-the correlation between any two residuals as and exponentially
-decreasing function of \(h\) (\(\phi^h\)), where \(h\) is the amount of
-separation in time between the residuals (Pinheiro and Bates, 2000).
-\(h\) may be a real valued number in the CAR(1), which is how it can
-accommodate the irregular separation of samples in time. \(\phi\)
-controls how quickly the correlation between any two residuals declines
-as a function of their separation in time and is an additional parameter
-that will be estimated during model fitting. The model in
-\eqref{eq:small-gam} was fitted using the \texttt{gamm()} function
-(Wood, 2004) in the \emph{mgcv} package (Wood, 2017) for R (R Core Team,
-2017).
+the uneven spacing is needed (Pinheiro and Bates, 2000). A continuous
+time first-order autoregressive process (CAR(1)) is a reasonable choice;
+it is the continuous-time equivalent of the first-order autoregressive
+process (AR(1)) and, simply stated, models the correlation between any
+two residuals as an exponentially decreasing function of \(h\)
+(\(\phi^h\)), where \(h\) is the amount of separation in time between
+the residuals (Pinheiro and Bates, 2000). \(h\) may be a real valued
+number in the CAR(1), which is how it can accommodate the irregular
+separation of samples in time. \(\phi\) controls how quickly the
+correlation between any two residuals declines as a function of their
+separation in time and is an additional parameter that will be estimated
+during model fitting. The model in \eqref{eq:small-gam} was fitted using
+the \texttt{gamm()} function (Wood, 2004) in the \emph{mgcv} package
+(Wood, 2017) for R (R Core Team, 2017).
 
@@
 GCV-selected trend: i) GCV is sensitive to the profile of the GCV score
-and has been shown to badly under smooth data in situations where the
-profile is flat around the minimum GCV score, and ii) the model fitted
-assumes that the observations are independent, an assumption that is
-certainly violated in the \uk{} time series.
+and has been shown to under smooth data in situations where the profile
+is flat around the minimum GCV score, and ii) the model fitted assumes
+that the observations are independent, an assumption that is certainly
+violated in the \uk{} time series.
 
@@
 (2009) is unlikely to apply here as there is a well-defined minimum in
-profile. Addressing the violation of the independence assumption is more
-difficult; the standard errors of the model coefficients are most likely
-anti-conservative, which presents a significant problem if we wish to
-determine which of the wiggles in the fitted \uk{} trend can be
-identified as real.
+profile.
 
@@
 the different variance (known as (heteroscedasticity) of each
-observation in the \uk{} record. The sediments in Braya-Sø are not
+observation in the \uk{} record. This seemingly isn't a problem for GCV
+which minimizes prediction error. The sediments in Braya-Sø are not
 annually laminated and therefore the core was sliced at regular depth
@@
 functions to use. As described above, the basis should be large enough
-to contain the true, but unknown, function, or a close approximation to
+to contain the true, but unknown, function or a close approximation to
 it. For GCV selection the basis used contained 29 basis functions,
@@
 the CAR(1) term and increase the basis dimension to 39 functions (by
-setting \texttt{k\ =\ 40}; I set it larger than expected wiggliness
-because the larger basis contains a richer family of functions and the
-excess complexity is reduced because of the smoothness penalty in play.)
+setting \texttt{k\ =\ 40}; I set it larger than expected because the
+larger basis contains a richer family of functions and the excess
+complexity is reduced because of the smoothness penalty.)
 
@@
 \ref{fig:plot-fitted-models}b). The trace of the REML score for this
-model (Figure \ref{fig:trace-smoothness-parameters}c) shows a pronounced
-minimum at a much smaller value of \(\lambda\) than the original REML
-fit (Figure \ref{fig:trace-smoothness-parameters}b), indicating that a
-more wiggly trend provides a better fit to the Braya-Sø time series.
-This example illustrates that some care and understanding of the
-underlying principles of GAMs is required to diagnose potential issues
-with the estimated model. After standard modelling choices (size of
-basis to use, correct selection of response distribution and link
-function, etc.) are checked, it often pays to think carefully about the
-properties of the data and ensure that the assumptions of the model are
-met. Here, despite increasing the basis size, it was the failure to
-appreciate the magnitude of the effect on fitting of the non-constant
-variance that lead to the initially poor fit and the problems associated
-with the estimation of the CAR(1) process. I return to the issue of why
-the GAM plus CAR(1) model encountered problems during fitting later
+model shows a pronounced minimum at a much smaller value of \(\lambda\)
+than the original REML fit (Figure
+\ref{fig:trace-smoothness-parameters}), indicating that a more wiggly
+trend provides a better fit to the Braya-Sø time series. This example
+illustrates that some care and understanding of the underlying
+principles of GAMs is required to diagnose potential issues with the
+estimated model. After standard modelling choices (size of basis to use,
+correct selection of response distribution and link function, etc.) are
+checked, it often pays to think carefully about the properties of the
+data and ensure that the assumptions of the model are met. Here, despite
+increasing the basis size, it was the failure to appreciate the
+magnitude of the effect on fitting of the non-constant variance that
+lead to the initially poor fit and the problems associated with the
+estimation of the CAR(1) process. I return to the issue of why the GAM
+plus CAR(1) model encountered problems during fitting later
 (\protect\hyperlink{identifiability}{see Identifiability}).
@@
 would normally be described as \emph{pointwise}; the coverage properties
-of the interval are correct for a single point on the fitted trend, but,
-if we were to consider additional points on the trend, the coverage
+of the interval being correct for a single point on the fitted trend,
+but, if we were to consider additional points on the trend, the coverage
 would logically be lower than 1 - \(\alpha\). This is the traditional
-frequentist interpretation of a confidence interval. However, the
-approach to fitting GAMs described here has a Bayesian interpretation
-(Kimeldorf and Wahba, 1970; Silverman, 1985; Wahba, 1983, 1990) and
-therefore the typical frequentist interpretation does not apply. Nychka
-(1988) investigated the properties of a confidence interval created as
-described above using standard errors derived from the Bayesian
-posterior covariance matrix for the estimated mode parameters. Such
-intervals have the interesting property that they have good
-\emph{across-the-function} coverage when considered from a frequentist
-perspective. This means that, when averaged over the range of the
-function, the Bayesian credible intervals shown in Figure
-\ref{fig:plot-fitted-models} have close to the expected 95\% coverage.
-However, to achieve this some parts of the function may have more or
-less than 95\%-coverage. Marra and Wood (2012) recently explained
-Nychka's (1988) surprising results and extended them to the case of
-generalized models (non-Gaussian responses).
+frequentist interpretation of a confidence interval. However, the GAM
+described here has a Bayesian interpretation (Kimeldorf and Wahba, 1970;
+Silverman, 1985; Wahba, 1983, 1990) and therefore the typical
+frequentist interpretation does not apply. Nychka (1988) investigated
+the properties of a confidence interval created as described above using
+standard errors derived from the Bayesian posterior covariance matrix
+for the estimated mode parameters. Such intervals have the interesting
+property that they have good \emph{across-the-function} coverage when
+considered from a frequentist perspective. This means that, when
+averaged over the range of the function, the Bayesian credible intervals
+shown in Figure \ref{fig:plot-fitted-models} have close to the expected
+95\% coverage. However, to achieve this some parts of the function may
+have more or less than 95\%-coverage. Marra and Wood (2012) recently
+explained Nychka's (1988) surprising results and extended them to the
+case of generalized models (non-Gaussian responses).
 
@@
 
-Whilst the \emph{across-the-function} or average coverage frequentist
-interpretation of the Bayesian credible intervals is useful, if may be
-important to have an interval that contains the entirety of the true
-function with some state probability (1 - \(\alpha\)). Such an interval
-is known as a \emph{simultaneous} interval. A (1 - \(\alpha\))100\%
-\emph{simultaneous} confidence interval contains \emph{in their
-entirety} 1 - \(\alpha\) of all random draws from the posterior
-distribution of the fitted model.
+Whilst the \emph{across-the-function} frequentist interpretation of the
+Bayesian credible intervals is useful, if may be important to have an
+interval that contains the entirety of the true function with some state
+probability (1 - \(\alpha\)). Such an interval is known as a
+\emph{simultaneous} interval. A (1 - \(\alpha\))100\% simultaneous
+confidence interval contains \emph{in their entirety} 1 - \(\alpha\) of
+all random draws from the posterior distribution of the fitted model.
 
@@
 represents a new trend that is consistent with the fitted trend but also
-includes the uncertainty in the estimated trend. This process is known
+reflects the uncertainty in the estimated trend. This process is known
 as \emph{posterior simulation}.
@@
 during these periods. For the random draws illustrated in Figure
-\ref{fig:posterior-simulation}, a simultaneous interval should contain
-the entire function for on average 19 of the 20 draws.
+\ref{fig:posterior-simulation}, a (1 - \(\alpha\))100\% simultaneous
+interval should contain the entire function for on average 19 of the 20
+draws.
 
@@
 uncertain that the simultaneous interval on the first derivative
-includes 0 throughout this period. We can understand why this is so by
-looking at the posterior simulations in
-Figure~\ref{fig:posterior-simulation}a; there is considerable variation
-in the shape of the simulated trends throughout this period. From
-\textasciitilde{}1925 the derivative of the trend becomes negative,
-however it is not until \textasciitilde{}1940 that the simultaneous
-interval doesn't include \(0\). At this point we have evidence to reject
-the null hypothesis of no change. This time point may be taken as the
-first evidence for change in δ\textsuperscript{15}N in the Small Water
-core. The simultaneous interval on the first derivative of the trend in
-δ\textsuperscript{15}N is bounded away from \(0\) between
-\textasciitilde{}1940 and \textasciitilde{}1975, covering the major
-decline in values evident in the observations. The simultaneous interval
-includes \(0\) from \textasciitilde{}1975 onward, suggesting that,
-whilst quite pronounced, the recent increase in δ\textsuperscript{15}N
-is not statistically significant. To determine whether or not the recent
-increase is real, we would require considerably more samples with which
-to (hopefully) more-precisely estimate the trend during this period.
-Alternatively, we might just have to wait until sufficient additional
-sedimentation has occurred to warrant recoring Small Water and
-reestimating the trend in δ\textsuperscript{15}N.
+includes 0 throughout. We can understand why this is so by looking at
+the posterior simulations in Figure~\ref{fig:posterior-simulation}a;
+there is considerable variation in the shape of the simulated trends
+throughout this period. From \textasciitilde{}1925 the derivative of the
+trend becomes negative, however it is not until \textasciitilde{}1940
+that the simultaneous interval doesn't include \(0\). At this point we
+have evidence to reject the null hypothesis of no change. This time
+point may be taken as the first evidence for change in
+δ\textsuperscript{15}N in the Small Water core. The simultaneous
+interval on the first derivative of the trend in δ\textsuperscript{15}N
+is bounded away from \(0\) between \textasciitilde{}1940 and
+\textasciitilde{}1975, covering the major decline in values evident in
+the observations. The simultaneous interval includes \(0\) from
+\textasciitilde{}1975 onward, suggesting that, whilst quite pronounced,
+the recent increase in δ\textsuperscript{15}N is not statistically
+significant. To determine whether or not the recent increase is real, we
+would require considerably more samples with which to (hopefully)
+more-precisely estimate the trend during this period. Alternatively, we
+might just have to wait until sufficient additional sedimentation has
+occurred to warrant recoring Small Water and reestimating the trend in
+δ\textsuperscript{15}N.
 
@@
 decreasing function of temporal separation (\(\Delta_t\)), and whilst
-observations that are a few year apart are quite strongly dependent on
+observations that are a few years apart are quite strongly dependent on
 one another, this dependence drops off rapidly as \(\Delta_t\) increases
@@
 one shown in Figure \ref{fig:plot-fitted-models}a which would negatively
-impact the confidence place on the inferences we might draw from the
+impact the confidence placed on the inferences we might draw from the
 fitted model. Importantly, failing to account for the strong dependency
@@
 Problems estimating the GAM plus CAR(1) model were encountered when this
-was fitted to the \uk{} time series; both a smooth trend in the mean
-\uk{} and a CAR(1) process in the residuals lead to an unidentifiable
-model. What makes a model with a spline-based trend and an
-autocorrelation process like the CAR(1) potentially unidentifiable?
+was fitted to the \uk{} time series; including both a smooth trend in
+the mean \uk{} and a CAR(1) process in the residuals lead to an
+unidentifiable model. What makes a model with a spline-based trend and
+an autocorrelation process like the CAR(1) potentially unidentifiable?
 
@@
 with stronger autocorrelation, and long runs of values greater (less)
-than the mean would evident as non-linear trends in the time series. As
-a result, a wiggly trend and an autocorrelation function with large
+than the mean would be evident as non-linear trends in the time series.
+As a result, a wiggly trend and an autocorrelation function with large
 \(\phi\) are two ways to describe the same pattern of values in a time
@@
 trend by modelling the way the correlation between pairs of observations
-varies as a function of the distance, \(h\), in time that separates
-(\(h\)) the observations. The correlation between pairs of observations
-decreases with increasing separation, which is modelled using a
-correlation function, \(c(h)\). Figure
-\ref{fig:gp-correlation-functions} shows examples of two different
-correlation functions; the \emph{power exponential} (Figure
+varies as a function of the distance, \(h\), in time that separates the
+observations. The correlation between pairs of observations decreases
+with increasing separation, which is modelled using a correlation
+function, \(c(h)\).
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/gp-correlation-functions-1}
+
+}
+
+\caption{Power exponential (a) and Matérn (b) correlation functions for observation separation distance $h$.
+Two values of the effective range parameter ($\phi$)) are shown for each function.
+For the power exponential function, $\kappa$ is the power in the power exponential function.
+For the Matérn correlation function, $\kappa$ distinguishes the member of the Matérn family.}\label{fig:gp-correlation-functions}
+\end{figure}
+
+Several functions can be used to represent \(c(h)\). Two common ones are
+the power exponential function and the Matérn family of correlation
+functions. The power exponential function at separation distance \(h\)
+is \[c(h) = \exp\{(-h/\phi)^{\kappa}\}\] where \(0 < \kappa \leq 2\).
+The Matérn correlation function is actually a family of functions with
+closed-forms only available for a subset of the family, distinguished by
+\(\kappa\). When \(\kappa = 1.5\), the Matérn correlation function is
+\[c(h) = (1 + h/\phi) \exp(-h/\phi)\] whilst for \(\kappa = 2.5\) it is
+\[c(h) = \{1 + h/\phi + (h/\phi)^2/3\} \exp(-h/\phi)\] and for
+\(\kappa = 3.5\)
+\[c(h) = \{1 + h/\phi + 2(h/\phi)^2/5 + (h/\phi)^3/15\} \exp(-h/\phi)\, .\]
+
+In all cases, \(\phi\) is the effective range parameter, which sets the
+distance beyond which the correlation function is effectively zero.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/gp-gam-detail-plot-1}
+
+}
+
+\caption{Gaussian process smooths fitted to the \uk{} time series.
+REML score traces for GAMs fitted using power exponential ($\kappa = 1$) or Matérn ($\kappa = 1.5$) correlation functions as a function of the effective range parameter ($\phi$) are shown (a).
+The optimal model for each function is that with the lowest REML score.
+b) shows the resulting trends estimated using the respective correlation function with the value of $\phi$ set to the optimal value.}\label{fig:gp-gam-detail-plot}
+\end{figure}
+
+Figure \ref{fig:gp-correlation-functions} shows examples of two
+different correlation functions; the \emph{power exponential} (Figure
 \ref{fig:gp-correlation-functions}a), and the Matérn (Figure
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/gp-correlation-functions-1}
-
-}
-
-\caption{Power exponential (a) and Matérn (b) correlation functions for observation separation distance $h$.
-Two values of the effective range parameter ($\phi$)) are shown for each function.
-For the power exponential function, $\kappa$ is the power in the power exponential function.
-For the Matérn correlation function, $\kappa$ distinguishes the member of the Matérn family.}\label{fig:gp-correlation-functions}
-\end{figure}
-
-Several functions can be used to represent \(c(h)\). Two common ones are
-the power exponential function and the Matérn family of correlation
-functions. The power exponential function at separation distance \(h\)
-is \[c(h) = \exp\{(-h/\phi)^{\kappa}\}\] where \(0 < \kappa \leq 2\).
-The Matérn correlation function is actually a family of functions with
-closed-forms only available for a subset of the family, distinguished by
-\(\kappa\). When \(\kappa = 1.5\), the Matérn correlation function is
-\[c(h) = (1 + h/\phi) \exp(-h/\phi)\] whilst for \(\kappa = 2.5\) it is
-\[c(h) = \{1 + h/\phi + (h/\phi)^2/3\} \exp(-h/\phi)\] and for
-\(\kappa = 3.5\)
-\[c(h) = \{1 + h/\phi + 2(h/\phi)^2/5 + (h/\phi)^3/15\} \exp(-h/\phi)\, .\]
-
-In all cases, \(\phi\) is the effective range parameter, which sets the
-distance beyond which the correlation function is effectively zero.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{manuscript_files/figure-latex/gp-gam-detail-plot-1}
-
-}
-
-\caption{Gaussian process smooths fitted to the \uk{} time series.
-REML score traces for GAMs fitted using power exponential ($\kappa = 1$) or Matérn ($\kappa = 1.5$) correlation functions as a function of the effective range parameter ($\phi$) are shown (a).
-The optimal model for each function is that with the lowest REML score.
-b) shows the resulting trends estimated using the respective correlation function with the value of $\phi$ set to the optimal value.}\label{fig:gp-gam-detail-plot}
-\end{figure}
-
 Gaussian processes and GAMs share many similarities and we can fit a
@@
 splines (Handcock et al., 1994; Kammann and Wand, 2003). It can be shown
-(e.g. Fahrmeir et al., 2013 section/pages) that the Gaussian process
-model has the same penalised likelihood form as the GAM that we
-discussed earlier; the observations are the knots of the smoother and
-each has a basis function in the form of a correlation function. The
-equivalence is only true if the basis functions do not depend on any
-other parameters of the model, which is only achievable if the value of
-\(\phi\) is fixed and known (Fahrmeir et al., 2013). In general,
-however, we would like to estimate \(\phi\) as part of model fitting. To
-achieve this we can maximise the profile likelihood or score statistic
-of the model over a range of values of \(\phi\) (Wood, 2017, 362--363).
-This involves proposing a value of \(\phi\) for the effective range of
-the correlation function and then estimating the resulting GAM by
-minimising the penalised log-likehood conditional upon this value of
-\(\phi\). The model, and its corresponding value of \(\phi\), with
-lowest penalised log-likelihood or score statistic is then retained as
-the estimated GAM. Figure~\ref{fig:gp-gam-detail-plot}a shows the REML
-score for models estimated using a Gaussian process smooth with a Matérn
-correlation function (\(\kappa\) = 1.5) for a sequence of values of
-\(\phi\) between 15 and 1000 years. There is a clear minimum around 40
-years separation, with the minimum REML score being observed at
-\(\phi = 41.81)\). Also shown are the REML scores for models using the
-power exponential function (\(\kappa\) = 1) with the minimum score
-observed at a somewhat higher effective range of \(\phi = 71.06\).
+(e.g. Fahrmeir et al., 2013) that the Gaussian process model has the
+same penalised likelihood form as the GAM that we discussed earlier; the
+observations are the knots of the smoother and each has a basis function
+in the form of a correlation function. The equivalence is only true if
+the basis functions do not depend on any other parameters of the model,
+which is only achievable if the value of \(\phi\) is fixed and known
+(Fahrmeir et al., 2013). In general, however, we would like to estimate
+\(\phi\) as part of model fitting. To achieve this we can maximise the
+profile likelihood or score statistic of the model over a range of
+values of \(\phi\) (Wood, 2017, 362--363). This involves proposing a
+value of \(\phi\) for the effective range of the correlation function
+and then estimating the resulting GAM by minimising the penalised
+log-likehood conditional upon this value of \(\phi\) and repeating for a
+range of values for \(\phi\). The model, and its corresponding value of
+\(\phi\), with lowest penalised log-likelihood or score statistic is
+then retained as the estimated GAM. Figure~\ref{fig:gp-gam-detail-plot}a
+shows the REML score for models estimated using a Gaussian process
+smooth with a Matérn correlation function (\(\kappa\) = 1.5) for a
+sequence of values of \(\phi\) between 15 and 1000 years. There is a
+clear minimum around 40 years separation, with the minimum REML score
+being observed at \(\phi = 41.81)\). Also shown are the REML scores for
+models using the power exponential function (\(\kappa\) = 1) with the
+minimum score observed at a somewhat higher effective range of
+\(\phi = 71.06\).
 
@@
 wiggliness, as the integrated squared second derivative of the spline,
-ensures that fitted smoother does not jump about wildly. This assumes
-that the data themselves are well described by a smoothly varying trend.
-If we anticipate abrupt change or step-like responses to environmental
-forcing this underlying assumption of the GAM would suggest that the
-method is ill-suited to modelling palaeo time series in which such
-features are evident or expected.
+ensures that the fitted smoother does not jump about wildly. This
+assumes that the data themselves are well described by a smoothly
+varying trend. If we anticipate abrupt change or step-like responses to
+environmental forcing this underlying assumption of the GAM would
+suggest that the method is ill-suited to modelling palaeo time series in
+which such features are evident or expected.
 
@@
 allows the wiggliness of the estimated trend to vary smoothly over time.
-Whilst this does allow the estimated trend to adapt to periods of rapid
+Whilst this allows the estimated trend to adapt to periods of rapid
 change in the response, adaptive smoothers make significant demands on
@@
 to be too-strong a statement; those oscillations in \uk{} may be real
-responses to climate forcing but we may simply lack the statistical
-power to distinguish them from the null hypothesis of no trend through
-this period. The adaptive smoother is only adjusting to the data
-available to it; just because it does not detect a trend during this
-period does not lend itself to an interpretation of stable climate
-forcing or complacency in the lake's response to forcing. If there were
-particular interest in the climate of this particular period we might
-take from the Braya-Sø record that there is potential early variation in
-climate forcing, but that additional data from this or other sites is
-required before any definitive conclusion can be drawn.
+responses to climate forcing but may simply lack the statistical power
+to distinguish them from the null hypothesis of no trend through this
+period. The adaptive smoother is only adjusting to the data available to
+it; just because it does not detect a trend during this period does not
+lend itself to an interpretation of stable climate forcing or
+complacency in the lake's response to forcing. If there were particular
+interest in the climate of this particular period we might take from the
+Braya-Sø record that there is potential early variation in climate
+forcing, but that additional data from this or other sites is required
+before any definitive conclusion can be drawn.
 
@@
 following the method of (Pya and Wood, 2015) using the \emph{scam}
-package (Pya, 2017). Also shown are 25 simulations from the model
-posterior of the monotonically-constrained GAM. Each simulation from the
-posterior distribution of the age-model is itself a potential age-depth
-model, which can be used to assign dates to the Small Water core. The
-trend model in \eqref{eq:gam} can be fitted to the
+package (Pya, 2017). Also shown are 25 simulations from the posterior
+distribution of the monotonically-constrained GAM. Each simulation from
+the posterior distribution of the age-model is itself a potential
+age-depth model, which can be used to assign dates to the Small Water
+core. The trend model in \eqref{eq:gam} can be fitted to the
 δ\textsuperscript{15}N data using these new dates as \(x_t\), and the
@@
 multivariate abundance data, where the aim is to model all species at
-once. The \emph{mgcv} software for example, is not able to estimate the
+once. The \emph{mgcv} software, for example, is not able to estimate the
 penalized GAM for multiple non-Gaussian responses. The exception is for
@@
 provide a superior alternative approach to trend estimation in
-palaeoenvironmental time series. They can estimate non-linear trends,
+palaeoenvironmental time series. GAMs can estimate non-linear trends,
 provide estimates of the magnitude of change as well as allow the
@@
 to give greater consideration to the estimation of trends and the
-identification of change.
+identification of change in stratigraphic time series.
 
@@
 to Helen Bennion, Rick Battarbee, and Peter Leavitt for their
-collaborations on projects over many years. Without Simon Wood's
-\emph{mgcv} software and his research on GAMs, the application of these
-models to palaeo time series would not be as straight forward. I also
-thank David Miller, Eric Pedersen, and Noam Ross, my GAM workshop
-partners in crime. This work was supported by a Natural Sciences and
+collaborations on projects over many years, and to David Miller, Eric
+Pedersen, and Noam Ross, my GAM workshop partners in crime. Without
+Simon Wood's \emph{mgcv} software and his research on GAMs, the
+application of these models to palaeo time series would not be as
+straight forward. This work was supported by a Natural Sciences and
 Engineering Council of Canada (NSERC) Discovery Grant to the author
@@
 R Core Team (2017). \emph{R: A language and environment for statistical
+computing}. Vienna, Austria: R Foundation for Statistical Computing
+Available at: \url{https://www.R-project.org/}.
+
+\hypertarget{ref-Rcore}{}
+R Core Team (2018). \emph{R: A language and environment for statistical
 computing}. Vienna, Austria: R Foundation for Statistical Computing

### 2018-05-14 supplementary-materials.Rmd
add supplementary materials

---
title: "Supplementary materials for: Modelling palaeoecological time series using generalized additive models"
author: Gavin L. Simpson
affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
mail: "gavin.simpson@uregina.ca"
date: May 13, 2018
keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
bibliography: references.bib
csl: frontiers-in-ecology-and-evolution.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
fontsize: 12pt
mainfont: TeX Gyre Pagella
output:
  pdf_document:
    number_sections: true
    fig_width: 8
    fig_height: 5
    fig_crop: false
    keep_tex: true
    latex_engine: xelatex
    #template: my.latex
    md_extensions: +header_attributes+superscript+subscript
    includes:
      in_header: header.tex
  md_document:
    variant: markdown_github
---

```{r knitr-defaults, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = "#>", fig.align = "center", out.width = "0.8\\linewidth",
                      echo = TRUE, message = TRUE, warning = TRUE, cache = TRUE)
knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
```

# Introduction

This document is an annotated version of the R code used to fit the GAMs and related analyses to the Small Water and Braya-Sø example data sets.

The following packages are required: *mgcv*, *scam*, *ggplot2*, *cowplot*, and *tidyr*. Also, the *schoenberg* package is required; it is not on CRAN but can be installed from Github.

```{r load-packages, cache = FALSE}
library("mgcv")
library("scam")
library("ggplot2")
library("cowplot")

## schoenberg is not on CRAN, install from github:
## install.packages("devtools")
## devtools::install_github("gavinsimpson/schoenberg")
library("schoenberg")
library("tidyr")

## Default ggplot theme
theme_set(theme_bw())
```

The example data sets are also stored on github; https://github.com/gavinsimpson/frontiers-palaeo-additive-modelling. Once downloaded the data are read in and processed a little

```{r common-components, cache = FALSE}
## source Small Water data
small <- readRDS("./data/small-water/small-water-isotope-data.rds")
head(small)

## load braya so data set
braya <- read.table("./data/braya-so/DAndrea.2011.Lake Braya So.txt",
                    skip = 84)
names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung",
                  "YearOld", "UK37")
braya <- transform(braya, sampleInterval = YearYoung - YearOld)
head(braya)

## plot labels
d15n_label <- expression(delta^{15}*N)
braya_ylabel <- expression(italic(U)[37]^{italic(k)})
```

Plots of the two data sets are prepared using *ggplot2*

```{r data-figure}
## plot Small Water data
small_plt <- ggplot(small, aes(x = Year, y = d15N)) +
    geom_point() +
    labs(y = d15n_label, x = "Year CE")

## Generate a plot of the data
braya_plt <- ggplot(braya, aes(x = Year, y = UK37)) +
    geom_line(colour = "grey") +
    geom_point() +
    labs(y = braya_ylabel, x = "Year CE")

## Recreate plot from manuscript
plot_grid(small_plt, braya_plt, ncol = 1, labels = "auto", align = "hv",
          axis = "lr")
```

# Fitting GAMs

The GAM plus CAR(1) process is fitted to the Small Water data set using the `gamm()` function. This fits GAMs as mixed effects models via the *nlme* package, which allows the use of correlation structures in the model residuals via the `correlation` argument. Here, the `corCAR1()` function is used to select the CAR(1) process and we specify the ordering of samples via the `Year` variable in `small`.

```{r fit-small-water-gamm}
## fit small water GAM usin gamm() with a CAR(1)
mod <- gamm(d15N ~ s(Year, k = 15), data = small,
            correlation = corCAR1(form = ~ Year), method = "REML")
```

The estimated value of $\phi$ for the CAR(1) can be extracted from the fitted model via the `$lme` component. Here we just extract the correlation structure component.

```{r extract-phi}
## estimate of phi and confidence interval
smallPhi <- intervals(mod$lme, which = "var-cov")$corStruct
smallPhi
```

The model summary is prepared from the `$gam` component of the fitted model

```{r small-summary}
## summary object
summary(mod$gam)
```

The output shows the estimated complexity of the fitted smooth, in terms of the effective degrees of freedom of the spline. An associated $F$ statistic and test of the null hypothesis of no trend (effect). Here the estimated trend provides strong evidence against this null.

The CAR(1) process plotted in Figure 10 of the manuscript was prepared using

```{r car1-plot}
## plot CAR(1) process
maxS <- with(small, diff(range(Year))) ## too large, truncate to 50
S <- seq(0, 50, length = 100)

car1 <- setNames(as.data.frame(t(outer(smallPhi, S, FUN = `^`)[1, , ])),
                 c("Lower","Correlation","Upper"))
car1 <- transform(car1, S = S)

car1Plt <- ggplot(car1, aes(x = S, y = Correlation)) +
    geom_ribbon(aes(ymax = Upper, ymin = Lower),
                fill = "black", alpha = 0.2) +
    geom_line() +
    ylab(expression(italic(h) * (list(Delta[t], varphi)))) +
    xlab(expression(Delta[t] ~ (years)))
car1Plt
```
The exponential decline in correlation with increasing separation is evident here; once samples are ~10 years apart, there is little estimated dependence between them.

The same model is fitted to the Braya-Sø data set. Note however that in order to even fit the model with both a smooth and the CAR(1) process, I have had to change the default optimiser used to fit the model, and reduce the basis dimension to a small number. We also fit the model using GCV, which is the defaut, hence no `method` argument

```{r fit-braya-so-car1-and-gcv-models}
## fit the car(1) model --- needs optim as this is not a stable fit!
## also needs k setting lower than default
braya.car1 <- gamm(UK37 ~ s(Year, k = 5), data = braya, 
                   correlation = corCAR1(form = ~ Year),
                   method = "REML",
		           control = list(niterEM = 0, optimMethod = "BFGS", 
                                  opt = "optim"))

## fit model using GCV
braya.gcv <- gam(UK37 ~ s(Year, k = 30), data = braya)

## estimate of phi and confidence interval
brayaPhi <- intervals(braya.car1$lme)$corStruct
brayaPhi
```

Note the wide confidence interval --- effectively 0--1 --- on $\phi$. If you were to increase the value of `k` to be `k = 10` in the `s(Year)` above, the model will fit but a warning message will be emitted when trying to extract $\phi$ due to a non-positive definite model covariance matrix, indicating problems with the model.

The next couple of code chunks prepare plots of the fitted GAMS. The general idea is to predict fom the fitted model for a fine grid of points over the range of the time variable. First we plot the trend for Small Water with an approximate 95% confidence interval assuming asymptotic normality

```{r small-plot-fitted-models}
N <- 300   # number of points at which to evaluate the splines
## Predict from the fitted model
newYear <- with(small, data.frame(Year = seq(min(Year), max(Year),
                                             length.out = 200)))
newYear <- cbind(newYear,
                 data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
newYear <- transform(newYear,
                     upper = fit + (2 * se.fit),
                     lower = fit - (2 * se.fit))

## Plot simulated trends
small_fitted <- ggplot(newYear, aes(x = Year, y = fit)) +
    geom_ribbon(aes(ymin = lower, ymax = upper, x = Year), alpha = 0.2,
                inherit.aes = FALSE, fill = "black") +
    geom_point(data = small, mapping = aes(x = Year, y = d15N),
               inherit.aes = FALSE) +
    geom_line() +
    labs(y = d15n_label, x = "Year CE")
small_fitted
```

For Braya-Sø, we repeat the process, but we do so for both models (GAMM + CAR(1) and GCV), and use a critical value from the $t$ distribution to form the confidence interval

```{r braya-plot-fitted-models}
## ggplot with data and fitted spline, then resids vs time in second panel
newBraya <- with(braya, data.frame(Year = seq(min(Year), max(Year),
                                              length.out = N)))
newBraya <- cbind(newBraya,
                  data.frame(predict(braya.car1$gam, newBraya,
                                     se.fit = TRUE)))
crit.t <- qt(0.975, df = df.residual(braya.car1$gam))
newBraya <- transform(newBraya,
                      upper = fit + (crit.t * se.fit),
                      lower = fit - (crit.t * se.fit))
## add GAM GCV results
fit_gcv <- predict(braya.gcv, newdata = newBraya, se.fit = TRUE)
newBraya <- rbind(newBraya, newBraya) # extend newBraya to take GCV results
newBraya[seq(N+1, length.out = N, by = 1), ]$fit <- fit_gcv$fit
newBraya[seq(N+1, length.out = N, by = 1), ]$upper <-
    fit_gcv$fit + (qt(0.975, df.residual(braya.gcv)) * fit_gcv$se.fit)
newBraya[seq(N+1, length.out = N, by = 1), ]$lower <-
    fit_gcv$fit - (qt(0.975, df.residual(braya.gcv)) * fit_gcv$se.fit)
newBraya <- transform(newBraya,
                      Method = rep(c("GAMM (CAR(1))", "GAM (GCV)"), 
                                   each = N))

## plot CAR(1) and GCV fits
braya_fitted <- ggplot(braya, aes(y = UK37, x = Year)) +
    geom_point() +
    geom_ribbon(data = newBraya,
                mapping = aes(x = Year, ymax = upper, ymin = lower,
                              fill = Method),
                alpha = 0.3, inherit.aes = FALSE) +
    geom_line(data = newBraya,
              mapping = aes(y = fit, x = Year, colour = Method)) +
    labs(y = braya_ylabel, x = "Year CE") +
    scale_color_manual(values = c("#5e3c99", "#e66101")) +
    scale_fill_manual(values = c("#5e3c99", "#e66101")) +
    theme(legend.position = "right")
braya_fitted
```

Figure 5 in the manuscript was produced using:

```{r manuscript-fig-2}
plot_grid(small_fitted, braya_fitted, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

To proceed with the Braya-Sø example, we need to increase the basis dimension (`k = 45`), fit using `method = "REML"`, and use observational weights. Here I use the `sampleInterval` variable as the measure of lake years per sample, and to avoid changing the model likelihood, the weights are actually the values of `sampleInterval` divided by the mean of `sampleInterval`:

```{r final-braya-reml-fit}
## TPRS, weights as sampleInterval
braya_reml <- gam(UK37 ~ s(Year, k = 45, bs = "tp"), data = braya,
                  method = "REML",
                  weights = sampleInterval / mean(sampleInterval))
```

# Posterior simulation

Samples from the posterior distribution of a GAM can be drawn using the `simulate()` methods from the *schoenberg* package.

```{r small-posterior-simulation}
set.seed(1) # set the random seed to make this reproducible
nsim <- 20  # how many simulations to draw

## do the simulations
sims <- simulate(mod, nsim = nsim, newdata = newYear, unconditional = TRUE)

## rearrange the output into a long/tidy format
colnames(sims) <- paste0("sim", seq_len(nsim))
sims <- setNames(stack(as.data.frame(sims)), c("simulated", "run"))
sims <- transform(sims, Year = rep(newYear$Year, nsim),
                  simulated = simulated)

## Plot simulated trends
smallSim.plt <- ggplot(newYear, aes(x = Year, y = fit)) +
    geom_line(data = sims,
              mapping = aes(y = simulated, x = Year, group = run),
              colour = "grey80") +
    geom_line(lwd = 1) +
    labs(y = d15n_label, x = "Year CE")
smallSim.plt
```

We repeat the same simulation for Braya-Sø

```{r braya-posterior-simulation}
## posterior simulation
## need to reset-up newBraya
newBraya <- with(braya,
                 data.frame(Year = seq(min(Year), max(Year),
                                       length.out = N)))
braya_pred <- cbind(newBraya,
                    data.frame(predict(braya_reml, newBraya,
                                       se.fit = TRUE)))

## simulate
set.seed(1)
sims2 <- simulate(braya_reml, nsim = nsim, newdata = newBraya,
                  unconditional = TRUE)
colnames(sims2) <- paste0("sim", seq_len(nsim))
sims2 <- setNames(stack(as.data.frame(sims2)),
                      c("simulated", "run"))
sims2 <- transform(sims2, Year = rep(newBraya$Year, nsim),
                       simulated = simulated)

brayaSim.plt <- ggplot(braya_pred, aes(x = Year, y = fit)) +
    geom_line(data = sims2,
              mapping = aes(y = simulated, x = Year, group = run),
              colour = "grey80") +
    geom_line(lwd = 1) +
    labs(y = braya_ylabel, x = "Year CE")
brayaSim.plt
```

Figure 7 in the manuscript was prepared using

```{r figure-7}
plot_grid(smallSim.plt, brayaSim.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

# Confidence and simultaneous intervals

Across-the-function and simultaneous confidence intervals are computed using the `confint()` method. The type of interval required is given via the `type` argument with options `"confidence"` and `"simultaneous"`.

```{r small-compare-intervals}
## small water
sw.cint <- confint(mod, parm = "Year", newdata = newYear,
                   type = "confidence")
sw.sint <- confint(mod, parm = "Year", newdata = newYear,
                   type = "simultaneous")

smallInt.plt <- ggplot(sw.cint, aes(x = Year, y = est)) +
    geom_ribbon(data = sw.sint,
                mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = d15n_label, x = "Year CE")
smallInt.plt
```

```{r braya-compare-intervals}
## braya so
bs.cint <- confint(braya_reml, parm = "Year", newdata = newBraya,
                   type = "confidence")
bs.sint <- confint(braya_reml, parm = "Year", newdata = newBraya,
                   type = "simultaneous")

brayaInt.plt <- ggplot(bs.cint, aes(x = Year, y = est)) +
    geom_ribbon(data = bs.sint,
                mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = braya_ylabel, x = "Year CE")
brayaInt.plt
```

Figure 8 in the manuscript was prepared using

```{r figure-8}
plot_grid(smallInt.plt, brayaInt.plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

# Derivatives of the estimated trend

The first derivative of the estimated trend is calculated using finite differences using the `fderiv()` function. There is also a `confint()` method for objects produced by `fderiv()`. The first derivatives and a 95% simultaneous confidence interval for the Small Water trend we computed and plotted using

```{r small-derivatives}
small.d <- fderiv(mod, newdata = newYear, n = N)
small.sint <- with(newYear,
                   cbind(confint(small.d, nsim = nsim,
                                 type = "simultaneous"),
                         Year = Year))

small_deriv_plt <- ggplot(small.sint, aes(x = Year, y = est)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2,
                fill = "black") +
    geom_line() +
    labs(x = "Year CE", y = "First derivative")
```
whilst for Braya-Sø, the following was used
```{r braya-derivatives}
braya.d <- fderiv(braya_reml, newdata = newBraya, n = N)
braya.sint <- with(newBraya,
                   cbind(confint(braya.d, nsim = nsim,
                                 type = "simultaneous"),
                         Year = Year))

braya_deriv_plt <- ggplot(braya.sint, aes(x = Year, y = est)) +
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.2, fill = "black") +
    geom_line() +
    labs(x = "Year CE", y = "First derivative")
```

Figure 9 in the manuscript was prepared using

```{r figure 9}
plot_grid(small_deriv_plt, braya_deriv_plt, ncol = 1, labels = "auto",
          align = "hv", axis = "lr")
```

# Gaussian process smooths

For the Gaussian process smooth to fit within the GAM framework described in the manuscript, we need to supply the value of $\phi$ for the effective range of the correlation function. To estimate $\phi$, we need to repeatedly fit the required GAM using a range of plausible values for $\phi$, which we do using a loop. In the chunk below I fit 200 models with $\phi$ in the range 15--500. For value of $\phi$ I fit the required GAM and extract the REML score (from component `gcv.ubre`) and store it in the numeric vectors `Mat` or `SEx`, for Matérn and Squared Exponential correlation functions, respectively. The final line of the chunk prepares the REML scores for the two correlation functions in long or tidy format suitable for plotting with *ggplot2*.

```{r profile-range-parameter-in-gp-smooth-model}
nn <- 200    # number of points at which to evaluate profile likelihood
dseq <- seq(15, 500, length.out = nn)  # effective ranges to fit at
Mat <- SEx <- numeric(length = nn)     # object to hold model fits
for (i in seq_along(dseq)) { 
    ## iterate over dseq, fit GP GAM w Matérn covariance
    Mat[i] <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, dseq[i])),
                  weights = sampleInterval / mean(sampleInterval),
                  data = braya, method = "REML",
                  family = gaussian())[["gcv.ubre"]]
    ## fit squared exponential
    SEx[i] <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(2, dseq[i], 1)),
                  weights = sampleInterval / mean(sampleInterval),
                  data = braya, method = "REML",
                  family = gaussian())[["gcv.ubre"]]
}

## extract the REML score into ggplot-friendly object
reml.scr <- data.frame(cor = rep(c("Matérn","Exponential"), each = nn),
                       effrange = rep(dseq, 2),
                       reml = c(Mat, SEx))
```

The REML scores for the models are plotted with *ggplot2* using

```{r gp-gam-detail-plot}
## profile-likelihood plot
proflik.plt <- ggplot(reml.scr, aes(x = effrange, y = reml, colour = cor)) +
    geom_line() +
    scale_colour_manual(name = "", values = c("#e66101","#5e3c99")) +
    labs(y = "REML score", x = expression(Effective ~ range ~ (varphi)))
proflik.plt
```

Next we extract the minimum of the REML scores for the two correlation functions and refit those models (we threw away all the models in the `for ()` loop earlier to avoid storing lots of model objects). Then we fit GAMs with Gaussian process smooths using the values of $\phi$ that produced the minimum REML scores, and predict using the fitted models to visualize the trends.

```{r fit-example-gp-smooths, dependsons = "profile-range-parameter-in-gp-smooth-model"}
effRange1 <- 250      # sets effective range in years for Matérn correl
## minima from profile likelihood
effRange2 <- with(subset(reml.scr, cor == "Matérn"), dseq[which.min(reml)])
effRange3 <- with(subset(reml.scr, cor == "Exponential"), dseq[which.min(reml)])

## Matern
gp2 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, effRange2)),
           data = braya,
           method = "REML", weights = sampleInterval / mean(sampleInterval))
## Power exponential
gp3 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(2, effRange3, 1)),
           data = braya,
           method = "REML", weights = sampleInterval / mean(sampleInterval))

newd <- with(braya, data.frame(Year = seq(min(Year), max(Year),
                               length.out = 1000)))
p.gp2 <- transform(newd,
                   fitted = predict(gp2, newdata = newd, type = "response"),
                   effRange = round(effRange2))
p.gp3 <- transform(newd,
                   fitted = predict(gp3, newdata = newd, type = "response"),
                   effRange = round(effRange3))
## pred <- rbind(p.gp1, p.gp2, p.gp3)
pred <- rbind(p.gp2, p.gp3)
pred <- transform(pred, effRange = factor(effRange),
                  cor = rep(c("Matérn", "Exponential"), each = nrow(newd)))
```

The estimated trends are plotted using

```{r plot-both-gp-smooth-trends}
## plot at two values of h
gp.plt2 <- ggplot(pred, aes(x = Year, y = fitted, colour = cor)) +
    geom_line() + theme(legend.position = "right") +
    geom_point(aes(x = Year, y = UK37), data = braya, inherit.aes = FALSE) +
    scale_colour_manual(name = "", values = c("#e66101","#5e3c99")) +
    labs(y = braya_ylabel, x = "Year CE")
```

whilst Figure 12 in the manuscript was prepared using

```{r figure-12}
plot_grid(proflik.plt, gp.plt2, ncol = 1, labels = c("a","b"),
          align = "hv", axis = "lr")
```

# Adaptive smooths

The adaptive smooth was fitted to the Braya-Sø data by adding `bs = "ad"` to the `s()` term in the model formula. The other aspects of the fit are as previously used for the other models, REML smoothness selection and observational weights:

```{r adaptove-smooth}
## Adaptive spline, weights as sampleInterval
mod_ad <- gam(UK37 ~ s(Year, k = 45, bs = "ad"), data = braya,
              method = "REML",
              weights = sampleInterval / mean(sampleInterval))
```

# Compare various trends

For the model comparison, I refitted all the models for consistency; the code to fit each of the

1. Thin plate regression spline,
2. Gaussian process spline (Matérn correlation functions), and
3. Adaptive smoother

is shown below.

```{r braya-so-model-comparisons, dependson = "fit-example-gp-smooths"}
## model it using gam()
effRange <- effRange2

## TPRS, weights as sampleInterval, k = needs to be higher
mod_tprs <- gam(UK37 ~ s(Year, k = 45, bs = "tp"), data = braya,
                method = "REML",
                weights = sampleInterval / mean(sampleInterval))

## Gaussian process, Matern, kappa = 1.5, weights as sampleInterval
mod_gp <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, effRange)),
              data = braya,
              method = "REML",
              weights = sampleInterval / mean(sampleInterval))

## Adaptive spline, weights as sampleInterval
mod_ad <- gam(UK37 ~ s(Year, k = 45, bs = "ad"), data = braya,
              method = "REML",
              weights = sampleInterval / mean(sampleInterval))
```

We write a small function to predict from each model over the range of `Year` and return the data in tidy format for plotting. The plot produce reproduces figure 13 in the manuscript.

```{r process-models, dependson = -1}
## wrap this in a function that will return all the plots & derived objects
processGAM <- function(mod) {
    ## Predict from model
    N <- 500
    newYear <- with(braya,
                    data.frame(Year = seq(min(Year), max(Year),
                                          length.out = N)))
    newYear <- cbind(newYear,
                     data.frame(predict(mod, newYear, se.fit = TRUE)))

    out <- list(objects = newYear)
    out
}

plts_gp   <- processGAM(mod = mod_gp) # Gaussian process smooth with weights
plts_ad   <- processGAM(mod = mod_ad) # Adaptive smooth with weights
plts_tprs <- processGAM(mod = mod_tprs) # TPRS with weights

pltData <- do.call("rbind", lapply(list(plts_gp, plts_ad, plts_tprs),
                                   `[[`, "objects"))
pltData <- transform(pltData, Model = rep(c("GP", "Adaptive", "TPRS"),
                              each = nrow(plts_gp$objects)))

allFits <- ggplot(pltData, aes(x = Year, y = fit)) +
    geom_point(aes(x = Year, y = UK37), data = braya) +
    geom_line(aes(colour = Model)) + labs(y = braya_ylabel, x = "Year") +
    theme(legend.position = "right") +
    scale_colour_manual(name = "",
                        values = c("#e66101", "#fdb863", "#5e3c99"))
allFits
```

# Accounting for age-model uncertainty

The manuscript proposed to simulate from the posterior distribution of the fitted age model as a way to account for age-model uncertainty. The first step in the process is to fit the age model from which to simulate new age models. This was done using the *scam* package for a *shape-constrained GAM*, with the age-model spline constrained to be monotonic decreasing (`bs = "mpd"`).

To make this section self-contained, I refitted the Small Water GAM plus CAR(1) model

```{r small-scam-fit}
knots <- with(small, list(Year = seq(min(Year), max(Year), length = 14)))
mod <- gamm(d15N ~ s(Year, k = 15), data = small, method = "REML",
            correlation = corCAR1(form = ~ Year),
            knots = knots)
```

and then we load the ^210^Pb dating results for the dated core sections.

```{r}
swAge <- read.csv("./data/small-water/small1-dating.csv")
```

before fitting the shape-constrained GAM. Currently, *scam* can only fit models using GCV smoothness selection. I used the `gamma` argument here to add a larger penalty for more-complex models. Each effective degree of freedom used by the spline is counted as 1.4 degrees of freedom in the GCV score.

```{r}
## monotonic spline age-depth model
swAge$Error[1] <- 1.1
swAgeMod <- scam(Date ~ s(Depth, k = 5, bs = "mpd"), data = swAge,
                 weights = 1 / swAge$Error, gamma = 1.4)
```

Note that I added a small amount of error to the surface sample age as the model cannot be fitted if an observation has `0` weight.

Next, predict from the estimated age model, and draw 25 samples from the posterior distribution using `simulate()`. The results are tidied into a format suitable for further processing and plotting. Note that the posterior samples here are only used for plotting.

```{r, dependson = -1}
## predict from the age model for a smooth set of points in `Depth`
newAge <- with(swAge, data.frame(Depth = seq(min(Depth), max(Depth),
                                             length.out = 200)))
newAge <- transform(newAge,
                    fitted = predict(swAgeMod, newdata = newAge, 
                                     type = "response"))
newSims <- as.data.frame(simulate(swAgeMod, nsim = 25, newdata = newAge))
newSims <- cbind(Depth = newAge$Depth, newSims)
newSims <- gather(newSims, Simulation, Age, -Depth)
```

In the next code chunk, I draw 100 samples from the posterior distribution of the age model, but notice that I pass in the `small` data to `newdata` in the call to `simulate()` as the locations I want new age estimates for are the depths for which we have δ^15^N values. A small function (`fitSWModels`) is written to prepare each simulation for fitting and then actually fit the GAM plus CAR(1) model using the updated age information.

```{r, dependson = -1}
## simulate from age model; each column is a simulation
ageSims <- simulate(swAgeMod, nsim = 100, newdata = small, seed = 42)
ageSims <- as.data.frame(ageSims)

fitSWModels <- function(x, y, knots) {
    dat <- data.frame(d15N = y, Year = x)
    m <- gamm(d15N ~ s(Year, k = 15), data = dat, method = "REML",
              correlation = corCAR1(form = ~ Year), knots = knots)
}

## generate new trends using draws from age-model posterior
simTrendMods <- lapply(ageSims, fitSWModels, y = small$d15N, knots = knots)

## function wrapper to predict new trends at locations over the
## range of `Year`
predSWModels <- function(mod, newdata) {
    predict(mod$gam, newdata = newdata, type = "response")
}

## predict from fitted model to produce a smooth trend for each posterior
## sample
simTrends <- lapply(simTrendMods, predSWModels, newdata = newYear)

## arrange in a tidy format form plottings
simTrends <- data.frame(Year  = with(newYear, rep(Year, length(simTrends))),
                        Trend = unlist(simTrends),
                        Group = rep(seq_along(simTrends),
                                    times = lengths(simTrends)))
```

The next chunk does the final step in the process. For each of the models we just fitted to include age model uncertainty, we simulate 50 draws from the model posterior distribution. We start with a wrapper function around the `simulate()` code we want to run on each model, then do the actual posterior draws for each model using `lapply()`. The final step just arranges data for plotting.

```{r, dependson = -1}
## wrapper to simulate from a fitted GAM with the arguments/settings
## I want
simulateSWModels <- function(mod, newdata, nsim, seed = 42) {
    sims <- simulate(mod, nsim = nsim, newdata = newdata, seed = seed)
    as.vector(sims)
}

## now do the posterior simulation
NSIM <- 50     # number of posterior samples *per* model
simSimulate <- lapply(simTrendMods, simulateSWModels, newdata = newYear,
                      nsim = NSIM, seed = 42)

## arrange in a tidy format
simSimulate <-
  data.frame(Year  = with(newYear,
                          rep(Year, times = NSIM * length(simSimulate))),
             Trend = unlist(simSimulate),
             Group = rep(seq_len(NSIM * length(simSimulate)),
                         each = nrow(newYear)))
```

Each of the steps is visualized using the plot code shown below.

```{r small-scam-fit-plots, dependson = -1}
plt1 <- ggplot(swAge, aes(y = Date, x = Depth)) +
    geom_line(data = newSims,
              mapping = aes(y = Age, x = Depth, group = Simulation),
              alpha = 1, colour = "grey80") +
    geom_line(data = newAge, mapping = aes(y = fitted, x = Depth)) +
    geom_point(size = 1.5, colour = "red") +
    geom_errorbar(aes(ymin = Date - Error, ymax = Date + Error, width = 0),
                  colour = "red") +
    labs(y = "Year CE", x = "Depth (cm)")

plt2 <- ggplot(simTrends, aes(x = Year, y = Trend, group = Group)) +
    geom_line(alpha = 0.1, colour = "grey80") +
    geom_line(data = newYear,
              mapping = aes(x = Year, y = fit), inherit.aes = FALSE) +
    geom_point(data = small,
               mapping = aes(x = Year, y = d15N),
               inherit.aes = FALSE, size = 0.7) +
    labs(x = "Year", y = d15n_label)

plt3 <- ggplot(simSimulate, aes(x = Year, y = Trend, group = Group)) +
    geom_line(alpha = 0.2, colour = "grey80") +
    geom_point(data = small,
               mapping = aes(x = Year, y = d15N),
               inherit.aes = FALSE,
               size = 0.7) +
    geom_line(data = newYear,
              mapping = aes(x = Year, y = fit),
              inherit.aes = FALSE) +
    labs(x = "Year", y = d15n_label)

plot_grid(plt1, plt2, plt3, ncol = 1, labels = "auto", align = "hv",
          axis = "lrtb", rel_widths = c(0.5, 1, 1))
```

This reproduces figure 14 from the manuscript.

# Session information

```{r session_info, include=TRUE, echo=TRUE, results='markup'}
devtools::session_info()
```

### 2018-05-14 supplementary-materials.tex
add supplementary materials

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[]{TeX Gyre Pagella}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Supplementary materials for: Modelling palaeoecological time series using generalized additive models},
            pdfauthor={Gavin L. Simpson},
            pdfkeywords={time series; generalized additive model; simultaneous interval; spline;
environmental change},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Supplementary materials for: Modelling palaeoecological time series
using generalized additive models}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Gavin L. Simpson}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{May 13, 2018}

%\setromanfont{Constantia}
%\setromanfont{texgyrepagella.otf}
%\setmainfont{XITS}
%\setmathfont{XITS Math}
\setmathfont{TeX Gyre Pagella Math}
\setmathfont[range={\mathcal,\mathbfcal},StylisticSet=1]{XITS Math}

%\renewcommand\familydefault{\rmdefault} %comment to see the difference
%\DeclareMathAlphabet      {\mathup}{OT1}{\familydefault}{m}{n}

\newcommand{\uk}{\ensuremath{\mathit{U}^{\mathit{K}}_{\mathup{37}}}}

\newcommand{\loess}{\textsc{Loess}}

\usepackage{lineno}
\linenumbers

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

This document is an annotated version of the R code used to fit the GAMs
and related analyses to the Small Water and Braya-Sø example data sets.

The following packages are required: \emph{mgcv}, \emph{scam},
\emph{ggplot2}, \emph{cowplot}, and \emph{tidyr}. Also, the
\emph{schoenberg} package is required; it is not on CRAN but can be
installed from Github.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"mgcv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> Loading required package: nlme
\end{verbatim}

\begin{verbatim}
#> This is mgcv 1.8-23. For overview type 'help("mgcv-package")'.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"scam"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> This is scam 1.2-2.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"ggplot2"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"cowplot"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> 
#> Attaching package: 'cowplot'
\end{verbatim}

\begin{verbatim}
#> The following object is masked from 'package:ggplot2':
#> 
#>     ggsave
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## schoenberg is not on CRAN, install from github:}
\NormalTok{## install.packages("devtools")}
\NormalTok{## devtools::install_github("gavinsimpson/schoenberg")}
\KeywordTok{library}\NormalTok{(}\StringTok{"schoenberg"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"tidyr"}\NormalTok{)}

\NormalTok{## Default ggplot theme}
\KeywordTok{theme_set}\NormalTok{(}\KeywordTok{theme_bw}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

The example data sets are also stored on github;
\url{https://github.com/gavinsimpson/frontiers-palaeo-additive-modelling}.
Once downloaded the data are read in and processed a little

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## source Small Water data}
\NormalTok{small <-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"./data/small-water/small-water-isotope-data.rds"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(small)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>   Depth   d13C  TotalC d15N TotalN DryWeight     Year
#> 1   0.2 -27.57  806.49 3.05  64.21       8.2 2007.254
#> 2   0.4 -27.67  949.33 3.01  73.26       7.6 2006.510
#> 3   0.8 -27.63 1305.52 2.93  93.25      11.6 2004.941
#> 4   1.2 -27.62 1136.04 2.33  86.09       9.6 2003.269
#> 5   1.6 -27.48 1028.27 2.09  93.80      10.9 2001.496
#> 6   2.0 -27.39  809.91 2.66  79.98       9.9 1999.626
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## load braya so data set}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"./data/braya-so/DAndrea.2011.Lake Braya So.txt"}\NormalTok{,}
                    \DataTypeTok{skip =} \DecValTok{84}\NormalTok{)}
\KeywordTok{names}\NormalTok{(braya) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Depth"}\NormalTok{, }\StringTok{"DepthUpper"}\NormalTok{, }\StringTok{"DepthLower"}\NormalTok{, }\StringTok{"Year"}\NormalTok{, }\StringTok{"YearYoung"}\NormalTok{,}
                  \StringTok{"YearOld"}\NormalTok{, }\StringTok{"UK37"}\NormalTok{)}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(braya, }\DataTypeTok{sampleInterval =} \NormalTok{YearYoung -}\StringTok{ }\NormalTok{YearOld)}
\KeywordTok{head}\NormalTok{(braya)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>   Depth DepthUpper DepthLower     Year YearYoung YearOld   UK37
#> 1  0.25        0.0        0.5 1999.125   2006.00 1992.25 -0.640
#> 2  0.75        0.5        1.0 1985.375   1992.25 1978.50 -0.637
#> 3  1.25        1.0        1.5 1971.525   1978.50 1964.55 -0.614
#> 4  1.75        1.5        2.0 1957.575   1964.55 1950.60 -0.627
#> 5  2.25        2.0        2.5 1943.150   1950.60 1935.70 -0.633
#> 6  2.75        2.5        3.0 1928.250   1935.70 1920.80 -0.616
#>   sampleInterval
#> 1          13.75
#> 2          13.75
#> 3          13.95
#> 4          13.95
#> 5          14.90
#> 6          14.90
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## plot labels}
\NormalTok{d15n_label <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(delta^\{}\DecValTok{15}\NormalTok{\}*N)}
\NormalTok{braya_ylabel <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(U)[}\DecValTok{37}\NormalTok{]^\{}\KeywordTok{italic}\NormalTok{(k)\})}
\end{Highlighting}
\end{Shaded}

Plots of the two data sets are prepared using \emph{ggplot2}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## plot Small Water data}
\NormalTok{small_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(small, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N)) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}

\NormalTok{## Generate a plot of the data}
\NormalTok{braya_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{colour =} \StringTok{"grey"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}

\NormalTok{## Recreate plot from manuscript}
\KeywordTok{plot_grid}\NormalTok{(small_plt, braya_plt, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{, }\DataTypeTok{align =} \StringTok{"hv"}\NormalTok{,}
          \DataTypeTok{axis =} \StringTok{"lr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/data-figure-1} \end{center}

\section{Fitting GAMs}\label{fitting-gams}

The GAM plus CAR(1) process is fitted to the Small Water data set using
the \texttt{gamm()} function. This fits GAMs as mixed effects models via
the \emph{nlme} package, which allows the use of correlation structures
in the model residuals via the \texttt{correlation} argument. Here, the
\texttt{corCAR1()} function is used to select the CAR(1) process and we
specify the ordering of samples via the \texttt{Year} variable in
\texttt{small}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## fit small water GAM usin gamm() with a CAR(1)}
\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small,}
            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

The estimated value of \(\phi\) for the CAR(1) can be extracted from the
fitted model via the \texttt{\$lme} component. Here we just extract the
correlation structure component.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## estimate of phi and confidence interval}
\NormalTok{smallPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(mod$lme, }\DataTypeTok{which =} \StringTok{"var-cov"}\NormalTok{)$corStruct}
\NormalTok{smallPhi}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>         lower      est.     upper
#> Phi 0.2811107 0.6026967 0.8547543
#> attr(,"label")
#> [1] "Correlation structure:"
\end{verbatim}

The model summary is prepared from the \texttt{\$gam} component of the
fitted model

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## summary object}
\KeywordTok{summary}\NormalTok{(mod$gam)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> 
#> Family: gaussian 
#> Link function: identity 
#> 
#> Formula:
#> d15N ~ s(Year, k = 15)
#> 
#> Parametric coefficients:
#>             Estimate Std. Error t value Pr(>|t|)    
#> (Intercept)  3.30909    0.03489   94.84   <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> Approximate significance of smooth terms:
#>           edf Ref.df     F p-value    
#> s(Year) 7.954  7.954 47.44  <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> R-sq.(adj) =  0.929   
#>   Scale est. = 0.037268  n = 48
\end{verbatim}

The output shows the estimated complexity of the fitted smooth, in terms
of the effective degrees of freedom of the spline. An associated \(F\)
statistic and test of the null hypothesis of no trend (effect). Here the
estimated trend provides strong evidence against this null.

The CAR(1) process plotted in Figure 10 of the manuscript was prepared
using

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## plot CAR(1) process}
\NormalTok{maxS <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(Year))) ## too large, truncate to 50}
\NormalTok{S <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DataTypeTok{length =} \DecValTok{100}\NormalTok{)}

\NormalTok{car1 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{t}\NormalTok{(}\KeywordTok{outer}\NormalTok{(smallPhi, S, }\DataTypeTok{FUN =} \StringTok{`}\DataTypeTok{^}\StringTok{`}\NormalTok{)[}\DecValTok{1}\NormalTok{, , ])),}
                 \KeywordTok{c}\NormalTok{(}\StringTok{"Lower"}\NormalTok{,}\StringTok{"Correlation"}\NormalTok{,}\StringTok{"Upper"}\NormalTok{))}
\NormalTok{car1 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(car1, }\DataTypeTok{S =} \NormalTok{S)}

\NormalTok{car1Plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(car1, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{S, }\DataTypeTok{y =} \NormalTok{Correlation)) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymax =} \NormalTok{Upper, }\DataTypeTok{ymin =} \NormalTok{Lower),}
                \DataTypeTok{fill =} \StringTok{"black"}\NormalTok{, }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
\StringTok{    }\KeywordTok{ylab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(h) *}\StringTok{ }\NormalTok{(}\KeywordTok{list}\NormalTok{(Delta[t], varphi)))) +}
\StringTok{    }\KeywordTok{xlab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(Delta[t] ~}\StringTok{ }\NormalTok{(years)))}
\NormalTok{car1Plt}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/car1-plot-1} \end{center}

The exponential decline in correlation with increasing separation is
evident here; once samples are \textasciitilde{}10 years apart, there is
little estimated dependence between them.

The same model is fitted to the Braya-Sø data set. Note however that in
order to even fit the model with both a smooth and the CAR(1) process, I
have had to change the default optimiser used to fit the model, and
reduce the basis dimension to a small number. We also fit the model
using GCV, which is the defaut, hence no \texttt{method} argument

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## fit the car(1) model --- needs optim as this is not a stable fit!}
\NormalTok{## also needs k setting lower than default}
\NormalTok{braya.car1 <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya, }
                   \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year),}
                   \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                   \DataTypeTok{control =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{niterEM =} \DecValTok{0}\NormalTok{, }\DataTypeTok{optimMethod =} \StringTok{"BFGS"}\NormalTok{, }
                                  \DataTypeTok{opt =} \StringTok{"optim"}\NormalTok{))}

\NormalTok{## fit model using GCV}
\NormalTok{braya.gcv <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{30}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya)}

\NormalTok{## estimate of phi and confidence interval}
\NormalTok{brayaPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(braya.car1$lme)$corStruct}
\NormalTok{brayaPhi}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>            lower      est. upper
#> Phi 1.611049e-22 0.2000156     1
#> attr(,"label")
#> [1] "Correlation structure:"
\end{verbatim}

Note the wide confidence interval --- effectively 0--1 --- on \(\phi\).
If you were to increase the value of \texttt{k} to be \texttt{k\ =\ 10}
in the \texttt{s(Year)} above, the model will fit but a warning message
will be emitted when trying to extract \(\phi\) due to a non-positive
definite model covariance matrix, indicating problems with the model.

The next couple of code chunks prepare plots of the fitted GAMS. The
general idea is to predict fom the fitted model for a fine grid of
points over the range of the time variable. First we plot the trend for
Small Water with an approximate 95\% confidence interval assuming
asymptotic normality

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{N <-}\StringTok{ }\DecValTok{300}   \CommentTok{# number of points at which to evaluate the splines}
\NormalTok{## Predict from the fitted model}
\NormalTok{newYear <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                             \DataTypeTok{length.out =} \DecValTok{200}\NormalTok{)))}
\NormalTok{newYear <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newYear,}
                 \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(mod$gam, newYear, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
\NormalTok{newYear <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newYear,}
                     \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(}\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{se.fit),}
                     \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(}\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{se.fit))}

\NormalTok{## Plot simulated trends}
\NormalTok{small_fitted <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(newYear, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year), }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{,}
                \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N),}
               \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
\NormalTok{small_fitted}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/small-plot-fitted-models-1} \end{center}

For Braya-Sø, we repeat the process, but we do so for both models (GAMM
+ CAR(1) and GCV), and use a critical value from the \(t\) distribution
to form the confidence interval

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## ggplot with data and fitted spline, then resids vs time in second panel}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                              \DataTypeTok{length.out =} \NormalTok{N)))}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newBraya,}
                  \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(braya.car1$gam, newBraya,}
                                     \DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\DataTypeTok{df =} \KeywordTok{df.residual}\NormalTok{(braya.car1$gam))}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newBraya,}
                      \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
                      \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}
\NormalTok{## add GAM GCV results}
\NormalTok{fit_gcv <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(braya.gcv, }\DataTypeTok{newdata =} \NormalTok{newBraya, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(newBraya, newBraya) }\CommentTok{# extend newBraya to take GCV results}
\NormalTok{newBraya[}\KeywordTok{seq}\NormalTok{(N}\DecValTok{+1}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{N, }\DataTypeTok{by =} \DecValTok{1}\NormalTok{), ]$fit <-}\StringTok{ }\NormalTok{fit_gcv$fit}
\NormalTok{newBraya[}\KeywordTok{seq}\NormalTok{(N}\DecValTok{+1}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{N, }\DataTypeTok{by =} \DecValTok{1}\NormalTok{), ]$upper <-}
\StringTok{    }\NormalTok{fit_gcv$fit +}\StringTok{ }\NormalTok{(}\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\KeywordTok{df.residual}\NormalTok{(braya.gcv)) *}\StringTok{ }\NormalTok{fit_gcv$se.fit)}
\NormalTok{newBraya[}\KeywordTok{seq}\NormalTok{(N}\DecValTok{+1}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{N, }\DataTypeTok{by =} \DecValTok{1}\NormalTok{), ]$lower <-}
\StringTok{    }\NormalTok{fit_gcv$fit -}\StringTok{ }\NormalTok{(}\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\KeywordTok{df.residual}\NormalTok{(braya.gcv)) *}\StringTok{ }\NormalTok{fit_gcv$se.fit)}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newBraya,}
                      \DataTypeTok{Method =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GAMM (CAR(1))"}\NormalTok{, }\StringTok{"GAM (GCV)"}\NormalTok{), }
                                   \DataTypeTok{each =} \NormalTok{N))}

\NormalTok{## plot CAR(1) and GCV fits}
\NormalTok{braya_fitted <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{UK37, }\DataTypeTok{x =} \NormalTok{Year)) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newBraya,}
                \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{ymin =} \NormalTok{lower,}
                              \DataTypeTok{fill =} \NormalTok{Method),}
                \DataTypeTok{alpha =} \FloatTok{0.3}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newBraya,}
              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{fit, }\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{colour =} \NormalTok{Method)) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{) +}
\StringTok{    }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#5e3c99"}\NormalTok{, }\StringTok{"#e66101"}\NormalTok{)) +}
\StringTok{    }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#5e3c99"}\NormalTok{, }\StringTok{"#e66101"}\NormalTok{)) +}
\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"right"}\NormalTok{)}
\NormalTok{braya_fitted}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/braya-plot-fitted-models-1} \end{center}

Figure 5 in the manuscript was produced using:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_grid}\NormalTok{(small_fitted, braya_fitted, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{,}
          \DataTypeTok{align =} \StringTok{"hv"}\NormalTok{, }\DataTypeTok{axis =} \StringTok{"lr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/manuscript-fig-2-1} \end{center}

To proceed with the Braya-Sø example, we need to increase the basis
dimension (\texttt{k\ =\ 45}), fit using \texttt{method\ =\ "REML"}, and
use observational weights. Here I use the \texttt{sampleInterval}
variable as the measure of lake years per sample, and to avoid changing
the model likelihood, the weights are actually the values of
\texttt{sampleInterval} divided by the mean of \texttt{sampleInterval}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## TPRS, weights as sampleInterval}
\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"tp"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
                  \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
\end{Highlighting}
\end{Shaded}

\section{Posterior simulation}\label{posterior-simulation}

Samples from the posterior distribution of a GAM can be drawn using the
\texttt{simulate()} methods from the \emph{schoenberg} package.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{) }\CommentTok{# set the random seed to make this reproducible}
\NormalTok{nsim <-}\StringTok{ }\DecValTok{20}  \CommentTok{# how many simulations to draw}

\NormalTok{## do the simulations}
\NormalTok{sims <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(mod, }\DataTypeTok{nsim =} \NormalTok{nsim, }\DataTypeTok{newdata =} \NormalTok{newYear, }\DataTypeTok{unconditional =} \OtherTok{TRUE}\NormalTok{)}

\NormalTok{## rearrange the output into a long/tidy format}
\KeywordTok{colnames}\NormalTok{(sims) <-}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"sim"}\NormalTok{, }\KeywordTok{seq_len}\NormalTok{(nsim))}
\NormalTok{sims <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{stack}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(sims)), }\KeywordTok{c}\NormalTok{(}\StringTok{"simulated"}\NormalTok{, }\StringTok{"run"}\NormalTok{))}
\NormalTok{sims <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(sims, }\DataTypeTok{Year =} \KeywordTok{rep}\NormalTok{(newYear$Year, nsim),}
                  \DataTypeTok{simulated =} \NormalTok{simulated)}

\NormalTok{## Plot simulated trends}
\NormalTok{smallSim.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(newYear, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{sims,}
              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{simulated, }\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{group =} \NormalTok{run),}
              \DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
\NormalTok{smallSim.plt}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/small-posterior-simulation-1} \end{center}

We repeat the same simulation for Braya-Sø

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## posterior simulation}
\NormalTok{## need to reset-up newBraya}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya,}
                 \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                       \DataTypeTok{length.out =} \NormalTok{N)))}
\NormalTok{braya_pred <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newBraya,}
                    \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(braya_reml, newBraya,}
                                       \DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}

\NormalTok{## simulate}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(braya_reml, }\DataTypeTok{nsim =} \NormalTok{nsim, }\DataTypeTok{newdata =} \NormalTok{newBraya,}
                  \DataTypeTok{unconditional =} \OtherTok{TRUE}\NormalTok{)}
\KeywordTok{colnames}\NormalTok{(sims2) <-}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"sim"}\NormalTok{, }\KeywordTok{seq_len}\NormalTok{(nsim))}
\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{stack}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(sims2)),}
                      \KeywordTok{c}\NormalTok{(}\StringTok{"simulated"}\NormalTok{, }\StringTok{"run"}\NormalTok{))}
\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(sims2, }\DataTypeTok{Year =} \KeywordTok{rep}\NormalTok{(newBraya$Year, nsim),}
                       \DataTypeTok{simulated =} \NormalTok{simulated)}

\NormalTok{brayaSim.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya_pred, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{sims2,}
              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{simulated, }\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{group =} \NormalTok{run),}
              \DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
\NormalTok{brayaSim.plt}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/braya-posterior-simulation-1} \end{center}

Figure 7 in the manuscript was prepared using

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_grid}\NormalTok{(smallSim.plt, brayaSim.plt, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{,}
          \DataTypeTok{align =} \StringTok{"hv"}\NormalTok{, }\DataTypeTok{axis =} \StringTok{"lr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure-7-1} \end{center}

\section{Confidence and simultaneous
intervals}\label{confidence-and-simultaneous-intervals}

Across-the-function and simultaneous confidence intervals are computed
using the \texttt{confint()} method. The type of interval required is
given via the \texttt{type} argument with options \texttt{"confidence"}
and \texttt{"simultaneous"}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## small water}
\NormalTok{sw.cint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(mod, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newYear,}
                   \DataTypeTok{type =} \StringTok{"confidence"}\NormalTok{)}
\NormalTok{sw.sint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(mod, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newYear,}
                   \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{)}

\NormalTok{smallInt.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(sw.cint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{data =} \NormalTok{sw.sint,}
                \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year),}
                \DataTypeTok{fill =} \StringTok{"grey80"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year),}
                \DataTypeTok{fill =} \StringTok{"grey60"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
\NormalTok{smallInt.plt}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/small-compare-intervals-1} \end{center}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## braya so}
\NormalTok{bs.cint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(braya_reml, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newBraya,}
                   \DataTypeTok{type =} \StringTok{"confidence"}\NormalTok{)}
\NormalTok{bs.sint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(braya_reml, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newBraya,}
                   \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{)}

\NormalTok{brayaInt.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(bs.cint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{data =} \NormalTok{bs.sint,}
                \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year),}
                \DataTypeTok{fill =} \StringTok{"grey80"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year),}
                \DataTypeTok{fill =} \StringTok{"grey60"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
\NormalTok{brayaInt.plt}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/braya-compare-intervals-1} \end{center}

Figure 8 in the manuscript was prepared using

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_grid}\NormalTok{(smallInt.plt, brayaInt.plt, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{,}
          \DataTypeTok{align =} \StringTok{"hv"}\NormalTok{, }\DataTypeTok{axis =} \StringTok{"lr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure-8-1} \end{center}

\section{Derivatives of the estimated
trend}\label{derivatives-of-the-estimated-trend}

The first derivative of the estimated trend is calculated using finite
differences using the \texttt{fderiv()} function. There is also a
\texttt{confint()} method for objects produced by \texttt{fderiv()}. The
first derivatives and a 95\% simultaneous confidence interval for the
Small Water trend we computed and plotted using

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{small.d <-}\StringTok{ }\KeywordTok{fderiv}\NormalTok{(mod, }\DataTypeTok{newdata =} \NormalTok{newYear, }\DataTypeTok{n =} \NormalTok{N)}
\NormalTok{small.sint <-}\StringTok{ }\KeywordTok{with}\NormalTok{(newYear,}
                   \KeywordTok{cbind}\NormalTok{(}\KeywordTok{confint}\NormalTok{(small.d, }\DataTypeTok{nsim =} \NormalTok{nsim,}
                                 \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{),}
                         \DataTypeTok{Year =} \NormalTok{Year))}

\NormalTok{small_deriv_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(small.sint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper), }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{,}
                \DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"First derivative"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

whilst for Braya-Sø, the following was used

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{braya.d <-}\StringTok{ }\KeywordTok{fderiv}\NormalTok{(braya_reml, }\DataTypeTok{newdata =} \NormalTok{newBraya, }\DataTypeTok{n =} \NormalTok{N)}
\NormalTok{braya.sint <-}\StringTok{ }\KeywordTok{with}\NormalTok{(newBraya,}
                   \KeywordTok{cbind}\NormalTok{(}\KeywordTok{confint}\NormalTok{(braya.d, }\DataTypeTok{nsim =} \NormalTok{nsim,}
                                 \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{),}
                         \DataTypeTok{Year =} \NormalTok{Year))}

\NormalTok{braya_deriv_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya.sint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper),}
                \DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"First derivative"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Figure 9 in the manuscript was prepared using

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_grid}\NormalTok{(small_deriv_plt, braya_deriv_plt, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{,}
          \DataTypeTok{align =} \StringTok{"hv"}\NormalTok{, }\DataTypeTok{axis =} \StringTok{"lr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure 9-1} \end{center}

\section{Gaussian process smooths}\label{gaussian-process-smooths}

For the Gaussian process smooth to fit within the GAM framework
described in the manuscript, we need to supply the value of \(\phi\) for
the effective range of the correlation function. To estimate \(\phi\),
we need to repeatedly fit the required GAM using a range of plausible
values for \(\phi\), which we do using a loop. In the chunk below I fit
200 models with \(\phi\) in the range 15--500. For value of \(\phi\) I
fit the required GAM and extract the REML score (from component
\texttt{gcv.ubre}) and store it in the numeric vectors \texttt{Mat} or
\texttt{SEx}, for Matérn and Squared Exponential correlation functions,
respectively. The final line of the chunk prepares the REML scores for
the two correlation functions in long or tidy format suitable for
plotting with \emph{ggplot2}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nn <-}\StringTok{ }\DecValTok{200}    \CommentTok{# number of points at which to evaluate profile likelihood}
\NormalTok{dseq <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{15}\NormalTok{, }\DecValTok{500}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{nn)  }\CommentTok{# effective ranges to fit at}
\NormalTok{Mat <-}\StringTok{ }\NormalTok{SEx <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DataTypeTok{length =} \NormalTok{nn)     }\CommentTok{# object to hold model fits}
\NormalTok{for (i in }\KeywordTok{seq_along}\NormalTok{(dseq)) \{ }
    \NormalTok{## iterate over dseq, fit GP GAM w Matérn covariance}
    \NormalTok{Mat[i] <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, dseq[i])),}
                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval),}
                  \DataTypeTok{data =} \NormalTok{braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                  \DataTypeTok{family =} \KeywordTok{gaussian}\NormalTok{())[[}\StringTok{"gcv.ubre"}\NormalTok{]]}
    \NormalTok{## fit squared exponential}
    \NormalTok{SEx[i] <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, dseq[i], }\DecValTok{1}\NormalTok{)),}
                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval),}
                  \DataTypeTok{data =} \NormalTok{braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                  \DataTypeTok{family =} \KeywordTok{gaussian}\NormalTok{())[[}\StringTok{"gcv.ubre"}\NormalTok{]]}
\NormalTok{\}}

\NormalTok{## extract the REML score into ggplot-friendly object}
\NormalTok{reml.scr <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{cor =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"Matérn"}\NormalTok{,}\StringTok{"Exponential"}\NormalTok{), }\DataTypeTok{each =} \NormalTok{nn),}
                       \DataTypeTok{effrange =} \KeywordTok{rep}\NormalTok{(dseq, }\DecValTok{2}\NormalTok{),}
                       \DataTypeTok{reml =} \KeywordTok{c}\NormalTok{(Mat, SEx))}
\end{Highlighting}
\end{Shaded}

The REML scores for the models are plotted with \emph{ggplot2} using

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## profile-likelihood plot}
\NormalTok{proflik.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(reml.scr, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{effrange, }\DataTypeTok{y =} \NormalTok{reml, }\DataTypeTok{colour =} \NormalTok{cor)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
\StringTok{    }\KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{""}\NormalTok{, }\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#e66101"}\NormalTok{,}\StringTok{"#5e3c99"}\NormalTok{)) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \StringTok{"REML score"}\NormalTok{, }\DataTypeTok{x =} \KeywordTok{expression}\NormalTok{(Effective ~}\StringTok{ }\NormalTok{range ~}\StringTok{ }\NormalTok{(varphi)))}
\NormalTok{proflik.plt}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/gp-gam-detail-plot-1} \end{center}

Next we extract the minimum of the REML scores for the two correlation
functions and refit those models (we threw away all the models in the
\texttt{for\ ()} loop earlier to avoid storing lots of model objects).
Then we fit GAMs with Gaussian process smooths using the values of
\(\phi\) that produced the minimum REML scores, and predict using the
fitted models to visualize the trends.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{effRange1 <-}\StringTok{ }\DecValTok{250}      \CommentTok{# sets effective range in years for Matérn correl}
\NormalTok{## minima from profile likelihood}
\NormalTok{effRange2 <-}\StringTok{ }\KeywordTok{with}\NormalTok{(}\KeywordTok{subset}\NormalTok{(reml.scr, cor ==}\StringTok{ "Matérn"}\NormalTok{), dseq[}\KeywordTok{which.min}\NormalTok{(reml)])}
\NormalTok{effRange3 <-}\StringTok{ }\KeywordTok{with}\NormalTok{(}\KeywordTok{subset}\NormalTok{(reml.scr, cor ==}\StringTok{ "Exponential"}\NormalTok{), dseq[}\KeywordTok{which.min}\NormalTok{(reml)])}

\NormalTok{## Matern}
\NormalTok{gp2 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, effRange2)),}
           \DataTypeTok{data =} \NormalTok{braya,}
           \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{, }\DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
\NormalTok{## Power exponential}
\NormalTok{gp3 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, effRange3, }\DecValTok{1}\NormalTok{)),}
           \DataTypeTok{data =} \NormalTok{braya,}
           \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{, }\DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}

\NormalTok{newd <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                               \DataTypeTok{length.out =} \DecValTok{1000}\NormalTok{)))}
\NormalTok{p.gp2 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newd,}
                   \DataTypeTok{fitted =} \KeywordTok{predict}\NormalTok{(gp2, }\DataTypeTok{newdata =} \NormalTok{newd, }\DataTypeTok{type =} \StringTok{"response"}\NormalTok{),}
                   \DataTypeTok{effRange =} \KeywordTok{round}\NormalTok{(effRange2))}
\NormalTok{p.gp3 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newd,}
                   \DataTypeTok{fitted =} \KeywordTok{predict}\NormalTok{(gp3, }\DataTypeTok{newdata =} \NormalTok{newd, }\DataTypeTok{type =} \StringTok{"response"}\NormalTok{),}
                   \DataTypeTok{effRange =} \KeywordTok{round}\NormalTok{(effRange3))}
\NormalTok{## pred <- rbind(p.gp1, p.gp2, p.gp3)}
\NormalTok{pred <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(p.gp2, p.gp3)}
\NormalTok{pred <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(pred, }\DataTypeTok{effRange =} \KeywordTok{factor}\NormalTok{(effRange),}
                  \DataTypeTok{cor =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"Matérn"}\NormalTok{, }\StringTok{"Exponential"}\NormalTok{), }\DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(newd)))}
\end{Highlighting}
\end{Shaded}

The estimated trends are plotted using

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## plot at two values of h}
\NormalTok{gp.plt2 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(pred, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fitted, }\DataTypeTok{colour =} \NormalTok{cor)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}\StringTok{ }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"right"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37), }\DataTypeTok{data =} \NormalTok{braya, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{""}\NormalTok{, }\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#e66101"}\NormalTok{,}\StringTok{"#5e3c99"}\NormalTok{)) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

whilst Figure 12 in the manuscript was prepared using

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_grid}\NormalTok{(proflik.plt, gp.plt2, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \KeywordTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{,}\StringTok{"b"}\NormalTok{),}
          \DataTypeTok{align =} \StringTok{"hv"}\NormalTok{, }\DataTypeTok{axis =} \StringTok{"lr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure-12-1} \end{center}

\section{Adaptive smooths}\label{adaptive-smooths}

The adaptive smooth was fitted to the Braya-Sø data by adding
\texttt{bs\ =\ "ad"} to the \texttt{s()} term in the model formula. The
other aspects of the fit are as previously used for the other models,
REML smoothness selection and observational weights:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Adaptive spline, weights as sampleInterval}
\NormalTok{mod_ad <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"ad"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
              \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
              \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
\end{Highlighting}
\end{Shaded}

\section{Compare various trends}\label{compare-various-trends}

For the model comparison, I refitted all the models for consistency; the
code to fit each of the

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Thin plate regression spline,
\item
  Gaussian process spline (Matérn correlation functions), and
\item
  Adaptive smoother
\end{enumerate}

is shown below.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## model it using gam()}
\NormalTok{effRange <-}\StringTok{ }\NormalTok{effRange2}

\NormalTok{## TPRS, weights as sampleInterval, k = needs to be higher}
\NormalTok{mod_tprs <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"tp"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
                \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}

\NormalTok{## Gaussian process, Matern, kappa = 1.5, weights as sampleInterval}
\NormalTok{mod_gp <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, effRange)),}
              \DataTypeTok{data =} \NormalTok{braya,}
              \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
              \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}

\NormalTok{## Adaptive spline, weights as sampleInterval}
\NormalTok{mod_ad <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"ad"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
              \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
              \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
\end{Highlighting}
\end{Shaded}

We write a small function to predict from each model over the range of
\texttt{Year} and return the data in tidy format for plotting. The plot
produce reproduces figure 13 in the manuscript.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## wrap this in a function that will return all the plots & derived objects}
\NormalTok{processGAM <-}\StringTok{ }\NormalTok{function(mod) \{}
    \NormalTok{## Predict from model}
    \NormalTok{N <-}\StringTok{ }\DecValTok{500}
    \NormalTok{newYear <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya,}
                    \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                          \DataTypeTok{length.out =} \NormalTok{N)))}
    \NormalTok{newYear <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newYear,}
                     \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(mod, newYear, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}

    \NormalTok{out <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{objects =} \NormalTok{newYear)}
    \NormalTok{out}
\NormalTok{\}}

\NormalTok{plts_gp   <-}\StringTok{ }\KeywordTok{processGAM}\NormalTok{(}\DataTypeTok{mod =} \NormalTok{mod_gp) }\CommentTok{# Gaussian process smooth with weights}
\NormalTok{plts_ad   <-}\StringTok{ }\KeywordTok{processGAM}\NormalTok{(}\DataTypeTok{mod =} \NormalTok{mod_ad) }\CommentTok{# Adaptive smooth with weights}
\NormalTok{plts_tprs <-}\StringTok{ }\KeywordTok{processGAM}\NormalTok{(}\DataTypeTok{mod =} \NormalTok{mod_tprs) }\CommentTok{# TPRS with weights}

\NormalTok{pltData <-}\StringTok{ }\KeywordTok{do.call}\NormalTok{(}\StringTok{"rbind"}\NormalTok{, }\KeywordTok{lapply}\NormalTok{(}\KeywordTok{list}\NormalTok{(plts_gp, plts_ad, plts_tprs),}
                                   \StringTok{`}\DataTypeTok{[[}\StringTok{`}\NormalTok{, }\StringTok{"objects"}\NormalTok{))}
\NormalTok{pltData <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(pltData, }\DataTypeTok{Model =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GP"}\NormalTok{, }\StringTok{"Adaptive"}\NormalTok{, }\StringTok{"TPRS"}\NormalTok{),}
                              \DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(plts_gp$objects)))}

\NormalTok{allFits <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(pltData, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37), }\DataTypeTok{data =} \NormalTok{braya) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{colour =} \NormalTok{Model)) +}\StringTok{ }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year"}\NormalTok{) +}
\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"right"}\NormalTok{) +}
\StringTok{    }\KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{""}\NormalTok{,}
                        \DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#e66101"}\NormalTok{, }\StringTok{"#fdb863"}\NormalTok{, }\StringTok{"#5e3c99"}\NormalTok{))}
\NormalTok{allFits}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/process-models-1} \end{center}

\section{Accounting for age-model
uncertainty}\label{accounting-for-age-model-uncertainty}

The manuscript proposed to simulate from the posterior distribution of
the fitted age model as a way to account for age-model uncertainty. The
first step in the process is to fit the age model from which to simulate
new age models. This was done using the \emph{scam} package for a
\emph{shape-constrained GAM}, with the age-model spline constrained to
be monotonic decreasing (\texttt{bs\ =\ "mpd"}).

To make this section self-contained, I refitted the Small Water GAM plus
CAR(1) model

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{knots <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year), }\DataTypeTok{length =} \DecValTok{14}\NormalTok{)))}
\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year),}
            \DataTypeTok{knots =} \NormalTok{knots)}
\end{Highlighting}
\end{Shaded}

and then we load the \textsuperscript{210}Pb dating results for the
dated core sections.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{swAge <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"./data/small-water/small1-dating.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

before fitting the shape-constrained GAM. Currently, \emph{scam} can
only fit models using GCV smoothness selection. I used the
\texttt{gamma} argument here to add a larger penalty for more-complex
models. Each effective degree of freedom used by the spline is counted
as 1.4 degrees of freedom in the GCV score.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## monotonic spline age-depth model}
\NormalTok{swAge$Error[}\DecValTok{1}\NormalTok{] <-}\StringTok{ }\FloatTok{1.1}
\NormalTok{swAgeMod <-}\StringTok{ }\KeywordTok{scam}\NormalTok{(Date ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Depth, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"mpd"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{swAge,}
                 \DataTypeTok{weights =} \DecValTok{1} \NormalTok{/}\StringTok{ }\NormalTok{swAge$Error, }\DataTypeTok{gamma =} \FloatTok{1.4}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Note that I added a small amount of error to the surface sample age as
the model cannot be fitted if an observation has \texttt{0} weight.

Next, predict from the estimated age model, and draw 25 samples from the
posterior distribution using \texttt{simulate()}. The results are tidied
into a format suitable for further processing and plotting. Note that
the posterior samples here are only used for plotting.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## predict from the age model for a smooth set of points in `Depth`}
\NormalTok{newAge <-}\StringTok{ }\KeywordTok{with}\NormalTok{(swAge, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Depth =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Depth), }\KeywordTok{max}\NormalTok{(Depth),}
                                             \DataTypeTok{length.out =} \DecValTok{200}\NormalTok{)))}
\NormalTok{newAge <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newAge,}
                    \DataTypeTok{fitted =} \KeywordTok{predict}\NormalTok{(swAgeMod, }\DataTypeTok{newdata =} \NormalTok{newAge, }
                                     \DataTypeTok{type =} \StringTok{"response"}\NormalTok{))}
\NormalTok{newSims <-}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{simulate}\NormalTok{(swAgeMod, }\DataTypeTok{nsim =} \DecValTok{25}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newAge))}
\NormalTok{newSims <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(}\DataTypeTok{Depth =} \NormalTok{newAge$Depth, newSims)}
\NormalTok{newSims <-}\StringTok{ }\KeywordTok{gather}\NormalTok{(newSims, Simulation, Age, -Depth)}
\end{Highlighting}
\end{Shaded}

In the next code chunk, I draw 100 samples from the posterior
distribution of the age model, but notice that I pass in the
\texttt{small} data to \texttt{newdata} in the call to
\texttt{simulate()} as the locations I want new age estimates for are
the depths for which we have δ\textsuperscript{15}N values. A small
function (\texttt{fitSWModels}) is written to prepare each simulation
for fitting and then actually fit the GAM plus CAR(1) model using the
updated age information.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## simulate from age model; each column is a simulation}
\NormalTok{ageSims <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(swAgeMod, }\DataTypeTok{nsim =} \DecValTok{100}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{small, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{)}
\NormalTok{ageSims <-}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(ageSims)}

\NormalTok{fitSWModels <-}\StringTok{ }\NormalTok{function(x, y, knots) \{}
    \NormalTok{dat <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{d15N =} \NormalTok{y, }\DataTypeTok{Year =} \NormalTok{x)}
    \NormalTok{m <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{dat, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
              \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{knots =} \NormalTok{knots)}
\NormalTok{\}}

\NormalTok{## generate new trends using draws from age-model posterior}
\NormalTok{simTrendMods <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(ageSims, fitSWModels, }\DataTypeTok{y =} \NormalTok{small$d15N, }\DataTypeTok{knots =} \NormalTok{knots)}

\NormalTok{## function wrapper to predict new trends at locations over the}
\NormalTok{## range of `Year`}
\NormalTok{predSWModels <-}\StringTok{ }\NormalTok{function(mod, newdata) \{}
    \KeywordTok{predict}\NormalTok{(mod$gam, }\DataTypeTok{newdata =} \NormalTok{newdata, }\DataTypeTok{type =} \StringTok{"response"}\NormalTok{)}
\NormalTok{\}}

\NormalTok{## predict from fitted model to produce a smooth trend for each posterior}
\NormalTok{## sample}
\NormalTok{simTrends <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(simTrendMods, predSWModels, }\DataTypeTok{newdata =} \NormalTok{newYear)}

\NormalTok{## arrange in a tidy format form plottings}
\NormalTok{simTrends <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year  =} \KeywordTok{with}\NormalTok{(newYear, }\KeywordTok{rep}\NormalTok{(Year, }\KeywordTok{length}\NormalTok{(simTrends))),}
                        \DataTypeTok{Trend =} \KeywordTok{unlist}\NormalTok{(simTrends),}
                        \DataTypeTok{Group =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{seq_along}\NormalTok{(simTrends),}
                                    \DataTypeTok{times =} \KeywordTok{lengths}\NormalTok{(simTrends)))}
\end{Highlighting}
\end{Shaded}

The next chunk does the final step in the process. For each of the
models we just fitted to include age model uncertainty, we simulate 50
draws from the model posterior distribution. We start with a wrapper
function around the \texttt{simulate()} code we want to run on each
model, then do the actual posterior draws for each model using
\texttt{lapply()}. The final step just arranges data for plotting.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## wrapper to simulate from a fitted GAM with the arguments/settings}
\NormalTok{## I want}
\NormalTok{simulateSWModels <-}\StringTok{ }\NormalTok{function(mod, newdata, nsim, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{) \{}
    \NormalTok{sims <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(mod, }\DataTypeTok{nsim =} \NormalTok{nsim, }\DataTypeTok{newdata =} \NormalTok{newdata, }\DataTypeTok{seed =} \NormalTok{seed)}
    \KeywordTok{as.vector}\NormalTok{(sims)}
\NormalTok{\}}

\NormalTok{## now do the posterior simulation}
\NormalTok{NSIM <-}\StringTok{ }\DecValTok{50}     \CommentTok{# number of posterior samples *per* model}
\NormalTok{simSimulate <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(simTrendMods, simulateSWModels, }\DataTypeTok{newdata =} \NormalTok{newYear,}
                      \DataTypeTok{nsim =} \NormalTok{NSIM, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{)}

\NormalTok{## arrange in a tidy format}
\NormalTok{simSimulate <-}
\StringTok{  }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year  =} \KeywordTok{with}\NormalTok{(newYear,}
                          \KeywordTok{rep}\NormalTok{(Year, }\DataTypeTok{times =} \NormalTok{NSIM *}\StringTok{ }\KeywordTok{length}\NormalTok{(simSimulate))),}
             \DataTypeTok{Trend =} \KeywordTok{unlist}\NormalTok{(simSimulate),}
             \DataTypeTok{Group =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{seq_len}\NormalTok{(NSIM *}\StringTok{ }\KeywordTok{length}\NormalTok{(simSimulate)),}
                         \DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(newYear)))}
\end{Highlighting}
\end{Shaded}

Each of the steps is visualized using the plot code shown below.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{plt1 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(swAge, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{Date, }\DataTypeTok{x =} \NormalTok{Depth)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newSims,}
              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{Age, }\DataTypeTok{x =} \NormalTok{Depth, }\DataTypeTok{group =} \NormalTok{Simulation),}
              \DataTypeTok{alpha =} \DecValTok{1}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newAge, }\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{fitted, }\DataTypeTok{x =} \NormalTok{Depth)) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{size =} \FloatTok{1.5}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"red"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_errorbar}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{Date -}\StringTok{ }\NormalTok{Error, }\DataTypeTok{ymax =} \NormalTok{Date +}\StringTok{ }\NormalTok{Error, }\DataTypeTok{width =} \DecValTok{0}\NormalTok{),}
                  \DataTypeTok{colour =} \StringTok{"red"}\NormalTok{) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \StringTok{"Year CE"}\NormalTok{, }\DataTypeTok{x =} \StringTok{"Depth (cm)"}\NormalTok{)}

\NormalTok{plt2 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(simTrends, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{Trend, }\DataTypeTok{group =} \NormalTok{Group)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.1}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newYear,}
              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit), }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =} \NormalTok{small,}
               \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N),}
               \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{size =} \FloatTok{0.7}\NormalTok{) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{y =} \NormalTok{d15n_label)}

\NormalTok{plt3 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(simSimulate, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{Trend, }\DataTypeTok{group =} \NormalTok{Group)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =} \NormalTok{small,}
               \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N),}
               \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{,}
               \DataTypeTok{size =} \FloatTok{0.7}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newYear,}
              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit),}
              \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{y =} \NormalTok{d15n_label)}

\KeywordTok{plot_grid}\NormalTok{(plt1, plt2, plt3, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{, }\DataTypeTok{align =} \StringTok{"hv"}\NormalTok{,}
          \DataTypeTok{axis =} \StringTok{"lrtb"}\NormalTok{, }\DataTypeTok{rel_widths =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/small-scam-fit-plots-1} \end{center}

This reproduces figure 14 from the manuscript.

\section{Session information}\label{session-information}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{devtools::}\KeywordTok{session_info}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> Session info -------------------------------------------------------------
\end{verbatim}

\begin{verbatim}
#>  setting  value                                      
#>  version  R version 3.4.4 Patched (2018-03-17 r74422)
#>  system   x86_64, linux-gnu                          
#>  ui       X11                                        
#>  language (EN)                                       
#>  collate  en_CA.UTF-8                                
#>  tz       America/Regina                             
#>  date     2018-05-14
\end{verbatim}

\begin{verbatim}
#> Packages -----------------------------------------------------------------
\end{verbatim}

\begin{verbatim}
#>  package    * version date       source                                  
#>  backports    1.1.2   2017-12-13 CRAN (R 3.4.3)                          
#>  base       * 3.4.4   2018-03-18 local                                   
#>  codetools    0.2-15  2016-10-05 CRAN (R 3.4.4)                          
#>  colorspace   1.3-2   2016-12-14 CRAN (R 3.4.1)                          
#>  compiler     3.4.4   2018-03-18 local                                   
#>  cowplot    * 0.9.2   2017-12-17 CRAN (R 3.4.3)                          
#>  datasets   * 3.4.4   2018-03-18 local                                   
#>  devtools     1.13.5  2018-02-18 CRAN (R 3.4.4)                          
#>  digest       0.6.15  2018-01-28 CRAN (R 3.4.4)                          
#>  evaluate     0.10.1  2017-06-24 CRAN (R 3.4.1)                          
#>  ggplot2    * 2.2.1   2016-12-30 CRAN (R 3.4.1)                          
#>  glue         1.2.0   2017-10-29 CRAN (R 3.4.2)                          
#>  graphics   * 3.4.4   2018-03-18 local                                   
#>  grDevices  * 3.4.4   2018-03-18 local                                   
#>  grid         3.4.4   2018-03-18 local                                   
#>  gtable       0.2.0   2016-02-26 CRAN (R 3.4.1)                          
#>  htmltools    0.3.6   2017-04-28 CRAN (R 3.4.1)                          
#>  knitr        1.20    2018-02-20 CRAN (R 3.4.4)                          
#>  labeling     0.3     2014-08-23 CRAN (R 3.4.1)                          
#>  lattice      0.20-35 2017-03-25 CRAN (R 3.4.4)                          
#>  lazyeval     0.2.1   2017-10-29 CRAN (R 3.4.2)                          
#>  magrittr     1.5     2014-11-22 CRAN (R 3.4.1)                          
#>  MASS         7.3-49  2018-02-23 CRAN (R 3.4.4)                          
#>  Matrix       1.2-14  2018-04-09 CRAN (R 3.4.4)                          
#>  memoise      1.1.0   2017-04-21 CRAN (R 3.4.1)                          
#>  methods      3.4.4   2018-03-18 local                                   
#>  mgcv       * 1.8-23  2018-01-21 CRAN (R 3.4.4)                          
#>  munsell      0.4.3   2016-02-13 CRAN (R 3.4.1)                          
#>  nlme       * 3.1-137 2018-04-07 CRAN (R 3.4.4)                          
#>  pillar       1.2.1   2018-02-27 CRAN (R 3.4.4)                          
#>  plyr         1.8.4   2016-06-08 CRAN (R 3.4.1)                          
#>  purrr        0.2.4   2017-10-18 CRAN (R 3.4.2)                          
#>  Rcpp         0.12.16 2018-03-13 CRAN (R 3.4.4)                          
#>  rlang        0.2.0   2018-02-20 CRAN (R 3.4.4)                          
#>  rmarkdown  * 1.9     2018-03-01 CRAN (R 3.4.4)                          
#>  rprojroot    1.3-2   2018-01-03 CRAN (R 3.4.3)                          
#>  scales       0.5.0   2017-08-24 CRAN (R 3.4.1)                          
#>  scam       * 1.2-2   2017-09-24 CRAN (R 3.4.3)                          
#>  schoenberg * 0.0-6   2018-04-12 Github (gavinsimpson/schoenberg@f9b8a84)
#>  splines      3.4.4   2018-03-18 local                                   
#>  stats      * 3.4.4   2018-03-18 local                                   
#>  stringi      1.1.7   2018-03-12 CRAN (R 3.4.4)                          
#>  stringr      1.3.0   2018-02-19 CRAN (R 3.4.4)                          
#>  tibble       1.4.2   2018-01-22 CRAN (R 3.4.3)                          
#>  tidyr      * 0.8.0   2018-01-29 CRAN (R 3.4.4)                          
#>  tools        3.4.4   2018-03-18 local                                   
#>  utils      * 3.4.4   2018-03-18 local                                   
#>  withr        2.1.2   2018-03-15 CRAN (R 3.4.4)                          
#>  yaml         2.1.18  2018-03-08 CRAN (R 3.4.4)
\end{verbatim}

\end{document}

### 2018-06-25 manuscript.Rmd
fix figure 14 so it actually uses half a column for panel a

@@
 
-plot_grid(plt1, plt2, plt3, ncol = 1, labels = "auto", align = "hv", axis = "lrtb",
-          rel_widths = c(0.5, 1, 1))
+## align all plots vertically
+plots <- align_plots(plt1, plt2, plt3, align = 'v', axis = 'l')
+
+## create the two rows of figures, from `plots`
+top_row <- plot_grid(plots[[1]], NULL, ncol = 2, labels = "a")
+bot_row <- plot_grid(plots[[2]], plots[[3]], ncol = 1, labels = c("b", "c"))
+
+## combine the two rows, top row has 1 row of plots, bottom row has 2, hence
+## the rel_heights to even this out
+plot_grid(top_row, bot_row, ncol = 1, rel_heights = c(0.5, 1))
 ```

### 2018-06-25 manuscript.Rmd
bump date

@@
 mail: "gavin.simpson@uregina.ca"
-date: May 11, 2018
+date: June 25, 2018
 keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 

### 2018-08-15 supplementary-materials.Rmd
revise the materials

@@
 mail: "gavin.simpson@uregina.ca"
-date: May 13, 2018
+date: August 14, 2018
 keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
@@
   pdf_document:
+    highlight: tango
     number_sections: true
@@
     latex_engine: xelatex
-    #template: my.latex
     md_extensions: +header_attributes+superscript+subscript
@@
 
-The following packages are required: *mgcv*, *scam*, *ggplot2*, *cowplot*, and *tidyr*. Also, the *schoenberg* package is required; it is not on CRAN but can be installed from Github.
+The following packages are required: *mgcv*, *scam*, *ggplot2*, *cowplot*, and *tidyr*.
 
@@
 library("cowplot")
+library("tidyr")
+```
+In addition, the *schoenberg* package is required; this in-development package is not on CRAN but can be installed directly from GitHub using functions from the *devtools* package.
+To install the package, install *devtools* and then use `devtools::install_github()` to install *schoenberg*, as shown below:
 
+```{r install-schoenberg, eval = FALSE}
 ## schoenberg is not on CRAN, install from github:
-## install.packages("devtools")
-## devtools::install_github("gavinsimpson/schoenberg")
-library("schoenberg")
-library("tidyr")
+install.packages("devtools")
+devtools::install_github("gavinsimpson/schoenberg")
+```
+Assuming that the installation of *schoenberg* completes without error, the package can be loaded as usual
+```{r load-schoenberg, cache = FALSE}
+library("schoenberg") # need to change the name of the package
+```
 
+The final environment-preparation step is to set the default *ggplot* theme, which the loading of *cowplot* has over-ridden. Here I use the more-minimal black-and-white theme (`theme_bw()`)
+
+```{r ggplot-default-theme, cache = FALSE}
 ## Default ggplot theme
@@
 
-The example data sets are also stored on github; https://github.com/gavinsimpson/frontiers-palaeo-additive-modelling. Once downloaded the data are read in and processed a little
+# Load the data sets
+
+The example data sets are also stored on GitHub; https://github.com/gavinsimpson/frontiers-palaeo-additive-modelling. Once downloaded the data are read in and processed a little
 
@@
                     skip = 84)
+## clean up variable names
 names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung",
                   "YearOld", "UK37")
+## add a variable for the amount of time per sediment sample
 braya <- transform(braya, sampleInterval = YearYoung - YearOld)
@@
 
+Instead of following the structure of the paper exactly, the subsequent sections focus on the two example data sets in turn, beginning with Small Water.
+
+## Small Water
+
+A GAM is typically fitted using the `gam()` function from the *mgcv* package.
+The typical call includes a formula describing the response variable and the linear predictor separated by the `~` (tilde) symbol.
+The linear predictor part of the model contains the smooth function of the time variable in the data set, and is indicated using the `s()` function.
+Unless other options are needed, we only need specify where the variables in the formula can be found via the `data` argument, and that we wish to use REML smoothness selection.
+
+Putting this together we have the following function call to fit a simple GAM to the Small Water isotope data
+
+```{r fit-small-water-gam}
+m <- gam(d15N ~ s(Year, k = 15), data = small, method = "REML")
+```
+
+As discussed in the paper, this model assumes that residuals[^1] are independent.
+To account for residual temporal autocorrelation we can include a continuous-time first-order autoregressive (CAR(1)) process in the model residuals for GAMs fitted to data that are conditionally distributed Gaussian.
+
+[^1]: or equivalently that the observations, conditional upon the model, are independent.
+
 The GAM plus CAR(1) process is fitted to the Small Water data set using the `gamm()` function.
@@
 ```{r fit-small-water-gamm}
-## fit small water GAM usin gamm() with a CAR(1)
+## fit small water GAM using gamm() with a CAR(1)
 mod <- gamm(d15N ~ s(Year, k = 15), data = small,
@@
 
+The object returned by `gamm()` includes both the linear mixed model and GAM faces of the model, and as a result we need to access the separate elements (`lme` and `gam` respectively) when proceeding to explore the model fit.
 The model summary is prepared from the `$gam` component of the fitted model
@@
 
-The output shows the estimated complexity of the fitted smooth, in terms of the effective degrees of freedom of the spline.
-An associated $F$ statistic and test of the null hypothesis of no trend (effect).
+The output shows the estimated complexity of the fitted smooth, expressed in terms of the effective degrees of freedom of the spline.
+An associated $F$ statistic and test of the null hypothesis of no trend (effect) are also shown.
 Here the estimated trend provides strong evidence against this null.
@@
 
-```{r car1-plot}
+```{r car1-plot, dependson = 'extract-phi'}
 ## plot CAR(1) process
-maxS <- with(small, diff(range(Year))) ## too large, truncate to 50
 S <- seq(0, 50, length = 100)
-
 car1 <- setNames(as.data.frame(t(outer(smallPhi, S, FUN = `^`)[1, , ])),
@@
 
+The next code chunk prepares a plot of the fitted GAMS.
+The general idea is to predict from the fitted model for a fine grid of points over the range of the time variable.
+Below I plot the trend for Small Water with an approximate 95% point-wise confidence interval that assumes asymptotic normality
+
+```{r small-plot-fitted-models}
+N <- 300   # number of points at which to evaluate the smooth
+
+## create new data to predict at; 200 evenly-spaced values over `Year`
+newYear <- with(small, data.frame(Year = seq(min(Year), max(Year),
+                                             length.out = 200)))
+## Predict from the fitted model; note we predict from the $gam part
+newYear <- cbind(newYear,
+                 data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
+## Create the confidence interval
+crit.t <- qt(0.975, df = df.residual(mod$gam))
+newYear <- transform(newYear,
+                     upper = fit + (crit.t * se.fit),
+                     lower = fit - (crit.t * se.fit))
+
+## Plot estimated trend
+small_fitted <- ggplot(newYear, aes(x = Year, y = fit)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper, x = Year), alpha = 0.2,
+                inherit.aes = FALSE, fill = "black") +
+    geom_point(data = small, mapping = aes(x = Year, y = d15N),
+               inherit.aes = FALSE) +
+    geom_line() +
+    labs(y = d15n_label, x = "Year CE")
+small_fitted
+```
+
+## Braya-Sø
+
 The same model is fitted to the Braya-Sø data set.
-Note however that in order to even fit the model with both a smooth and the CAR(1) process, I have had to change the default optimiser used to fit the model, and reduce the basis dimension to a small number.
+Note however that in order to even fit the model with both a smooth and the CAR(1) process, I have had to change the default optimiser used to estimate the model parameters, and reduce the basis dimension to a small number.
 We also fit the model using GCV, which is the defaut, hence no `method` argument
@@
 
-The next couple of code chunks prepare plots of the fitted GAMS.
-The general idea is to predict fom the fitted model for a fine grid of points over the range of the time variable.
-First we plot the trend for Small Water with an approximate 95% confidence interval assuming asymptotic normality
-
-```{r small-plot-fitted-models}
-N <- 300   # number of points at which to evaluate the splines
-## Predict from the fitted model
-newYear <- with(small, data.frame(Year = seq(min(Year), max(Year),
-                                             length.out = 200)))
-newYear <- cbind(newYear,
-                 data.frame(predict(mod$gam, newYear, se.fit = TRUE)))
-newYear <- transform(newYear,
-                     upper = fit + (2 * se.fit),
-                     lower = fit - (2 * se.fit))
-
-## Plot simulated trends
-small_fitted <- ggplot(newYear, aes(x = Year, y = fit)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper, x = Year), alpha = 0.2,
-                inherit.aes = FALSE, fill = "black") +
-    geom_point(data = small, mapping = aes(x = Year, y = d15N),
-               inherit.aes = FALSE) +
-    geom_line() +
-    labs(y = d15n_label, x = "Year CE")
-small_fitted
-```
-
 For Braya-Sø, we repeat the process, but we do so for both models (GAMM + CAR(1) and GCV), and use a critical value from the $t$ distribution to form the confidence interval
@@
 ```{r braya-plot-fitted-models}
-## ggplot with data and fitted spline, then resids vs time in second panel
+N <- 300   # number of points at which to evaluate the smooth
+
+## data to predict at
 newBraya <- with(braya, data.frame(Year = seq(min(Year), max(Year),
                                               length.out = N)))
+## add predictions from GAMM + CAR(1) model
 newBraya <- cbind(newBraya,
@@
                       lower = fit - (crit.t * se.fit))
+
 ## add GAM GCV results
 fit_gcv <- predict(braya.gcv, newdata = newBraya, se.fit = TRUE)
-newBraya <- rbind(newBraya, newBraya) # extend newBraya to take GCV results
-newBraya[seq(N+1, length.out = N, by = 1), ]$fit <- fit_gcv$fit
-newBraya[seq(N+1, length.out = N, by = 1), ]$upper <-
-    fit_gcv$fit + (qt(0.975, df.residual(braya.gcv)) * fit_gcv$se.fit)
-newBraya[seq(N+1, length.out = N, by = 1), ]$lower <-
-    fit_gcv$fit - (qt(0.975, df.residual(braya.gcv)) * fit_gcv$se.fit)
+crit.t <- qt(0.975, df.residual(braya.gcv))
+newGCV <- data.frame(Year   = newBraya[["Year"]],
+                     fit    = fit_gcv$fit,
+                     se.fit = fit_gcv$se.fit)
+newGCV <- transform(newGCV,
+                    upper = fit + (crit.t * se.fit),
+                    lower = fit - (crit.t * se.fit))
+newBraya <- rbind(newBraya, newGCV)                # bind on GCV results
+## Add indicator variable for model
 newBraya <- transform(newBraya,
@@
 
-```{r manuscript-fig-2}
+```{r manuscript-fig-2, fig.show = "hide"}
 plot_grid(small_fitted, braya_fitted, ncol = 1, labels = "auto",
@@
 
-To proceed with the Braya-Sø example, we need to increase the basis dimension (`k = 45`), fit using `method = "REML"`, and use observational weights.
-Here I use the `sampleInterval` variable as the measure of lake years per sample, and to avoid changing the model likelihood, the weights are actually the values of `sampleInterval` divided by the mean of `sampleInterval`:
+### Checking if the size of the basis expansion is sufficient
+
+Clearly we are not able to identify both the wiggly trend and the CAR(1) process in a single model.
+We might proceed by fitting a simple GAM via `gam()` with REML smoothness selection and moderate number of basis functions --- here we set `k = 15` for illustration.
+
+```{r braya-fit-k-too-small}
+braya_low_k <- gam(UK37 ~ s(Year, k = 15), data = braya, method = "REML")
+```
+
+Before proceeding, we should perform a check to determine if the number of basis functions requested is sufficient to capture the wiggliness in the data. This test is provided by `gam.check()`
+
+```{r check-braya-fit-k-too-small, dependson = -1, fig.show = "hide"}
+gam.check(braya_low_k)
+```
+
+In the output, the column labelled `k'` is the size of the basis used (the number of basis functions) --- note that it is 14 and not the requested 15 functions as the constant basis function has been removed as it is confounded with the model constant term, the intercept.
+The column labelled `edf` indicates the *effective degrees of freedom* for the estimated smooth.
+The `k-index` column contains the value of a test statistic for a test of sufficient number of basis functions --- ideally, we'd like the value of `k-index` to be close to or greater than 1.
+The quoted *p* value measures the support for the null hypothesis that enough basis functions were available.
+In practial terms, the low `k-index` and very low *p* value suggest that the requested number of basis functions used to fit the model was too low.
+
+The solution is to increase `k`, say by doubling the original value, and refit the model and perform the basis dimension check using `gam.check()`, repeating this process until `k-index` is close to or greater than 1 and the *p* value is larger than a desired threshold.
+In my experience, it may not be possible with some data sets to increase `k` large enough to result in a `k-index` > 1 and a *p* value > 0.05, given the available.
+It is important to remember that the test provided by `gam.check()` is only an heuristic one and shouldn't be seen as infallible.
+If you find yourself increasing `k` to large values relative to the number of samples in your data set without achieving a `k-index` > 1 or *p* > 0.05, then note the value in the `edf` column as you increase `k`.
+If the `edf` of the model changes little as you continue to increase `k`, this may be evidence that the test is failing in this particular instance and that the estimated smooth is not dependent on the value of `k` chosen.
+
+## Accounting for heteroscedasticity due to time averaging
+
+To proceed with the Braya-Sø example, we need to increase the basis dimension (`k = 45`), fit using `method = "REML"`.
+As discussed in the manuscript, we should also account for the non-constant variance, or *heteroscedasticity*, that may arise due to variation in the amount of time (or "lake years") that each sample represents.
+All else equal, we expect that samples that average more time have smaller variance than those that average a smaller amount of time.
+To include information about the expected heteroscedasticity in the GAM we can use observational weights[^2].
+Here I use the `sampleInterval` variable that I created earlier as the measure of lake years per sample, and to avoid changing the model likelihood, the weights use are the values of `sampleInterval` divided by the mean of `sampleInterval`:
 
 ```{r final-braya-reml-fit}
-## TPRS, weights as sampleInterval
-braya_reml <- gam(UK37 ~ s(Year, k = 45, bs = "tp"), data = braya,
+braya_reml <- gam(UK37 ~ s(Year, k = 45), data = braya,
                   method = "REML",
                   weights = sampleInterval / mean(sampleInterval))
+summary(braya_reml)
 ```
+
+[^2]: This simple solution is for a Gaussian GAM.
+For models with other distributions, the exact interpretation of observational weights will vary.
+For such data, where the conditional mean and variance of the observations are linked, the use of location-scale, or distributional, models may be more apporpriate and informative.
+
+If we now check if 44 (`k = 45`) basis functions is sufficient
+```{r final-braya-gam-check, fig.show = "hide"}
+gam.check(braya_reml)
+```
+we note that the value of `k-index` is now greater than 1 and the *p* value suggests there is very little evidence against the null hypothesis.
+Also note that the `edf` of the estimated smooth is considerably below the maximum (`k'`).
+This is the effect of the smoothness penalty removing the excessive wiggliness possible with 44 basis functions.
+If the `edf` were close to `k'`, you might consider increasing `k` by a modest amount (in this instance by 5 or 10 additional basis functions) to assure yourself that you have sufficient basis functions.
+This has the additional advantage of allowing a richer set of smooths of a given complexity (`edf`) to be represented using the basis functions, which may improve the model fit to the data without changing the `edf` too much.
 
@@
 
-```{r figure-7}
+```{r figure-7, fig.show = "hide"}
 plot_grid(smallSim.plt, brayaSim.plt, ncol = 1, labels = "auto",
@@
 The type of interval required is given via the `type` argument with options `"confidence"` and `"simultaneous"`.
+Fro example, for Small Water we would use
 
 ```{r small-compare-intervals}
-## small water
 sw.cint <- confint(mod, parm = "Year", newdata = newYear,
@@
                    type = "simultaneous")
+head(sw.sint)
+```
 
+The `confint()` methods return data frames suitable for subsequent plotting with **ggplot**.
+The columns labelled `est` and `se` are the estimate values of the smooth and its standard error, respectively.
+The variables `lower` and `upper` contain the values of the lower and upper bounds on the requested interval.
+The intervals can be plotted as follows
+
+```{r plot-small-water-intervals}
 smallInt.plt <- ggplot(sw.cint, aes(x = Year, y = est)) +
@@
 
+The intervals for Braya-Sø are created
 ```{r braya-compare-intervals}
-## braya so
 bs.cint <- confint(braya_reml, parm = "Year", newdata = newBraya,
@@
                    type = "simultaneous")
-
+```
+and plotted
+```{r plot-braya-so-intervals}
 brayaInt.plt <- ggplot(bs.cint, aes(x = Year, y = est)) +
@@
 ```
+in the same way
 
 Figure 8 in the manuscript was prepared using
-
-```{r figure-8}
+```{r figure-8, fig.show = "hide"}
 plot_grid(smallInt.plt, brayaInt.plt, ncol = 1, labels = "auto",
@@
 Mat <- SEx <- numeric(length = nn)     # object to hold model fits
+
+## loop over the sequence of deparation distances and fit the models
 for (i in seq_along(dseq)) { 
@@
 ```{r fit-example-gp-smooths, dependsons = "profile-range-parameter-in-gp-smooth-model"}
-effRange1 <- 250      # sets effective range in years for Matérn correl
-## minima from profile likelihood
+## effective range minima from profile likelihood
 effRange2 <- with(subset(reml.scr, cor == "Matérn"), dseq[which.min(reml)])
@@
 
-## Matern
+## Refit these models: Matern
 gp2 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(3, effRange2)),
@@
            method = "REML", weights = sampleInterval / mean(sampleInterval))
-## Power exponential
+## Refit these models: Power exponential
 gp3 <- gam(UK37 ~ s(Year, k = 45, bs = "gp", m = c(2, effRange3, 1)),
@@
 
+## data to predict at
 newd <- with(braya, data.frame(Year = seq(min(Year), max(Year),
                                length.out = 1000)))
+## create predictions on response scale for both covariance functions
 p.gp2 <- transform(newd,
@@
                    effRange = round(effRange3))
-## pred <- rbind(p.gp1, p.gp2, p.gp3)
+## joint the two sets of predictions together
 pred <- rbind(p.gp2, p.gp3)
-pred <- transform(pred, effRange = factor(effRange),
+## add some categorical variables for plotting
+pred <- transform(pred,
+                  effRange = factor(effRange),
                   cor = rep(c("Matérn", "Exponential"), each = nrow(newd)))
@@
 
-The estimated trends are plotted using
+The estimated Gaussian process trends are plotted using
 
 ```{r plot-both-gp-smooth-trends}
-## plot at two values of h
 gp.plt2 <- ggplot(pred, aes(x = Year, y = fitted, colour = cor)) +
@@
     labs(y = braya_ylabel, x = "Year CE")
+gp.plt2
 ```
@@
 
-```{r figure-12}
+```{r figure-12, fig.show = "hide"}
 plot_grid(proflik.plt, gp.plt2, ncol = 1, labels = c("a","b"),
@@
 
-```{r adaptove-smooth}
+```{r adaptive-smooth}
 ## Adaptive spline, weights as sampleInterval
@@
 
-# Compare various trends
+# Comparing trends
 
@@
 
-1. Thin plate regression spline,
+1. Thin plate regression spline (TPRS),
 2. Gaussian process spline (Matérn correlation functions), and
@@
 ```{r braya-so-model-comparisons, dependson = "fit-example-gp-smooths"}
-## model it using gam()
+## effective range parameter from profile-likelihood of Matern model
 effRange <- effRange2
 
-## TPRS, weights as sampleInterval, k = needs to be higher
+## TPRS, weights as sampleInterval
 mod_tprs <- gam(UK37 ~ s(Year, k = 45, bs = "tp"), data = braya,
@@
 ```{r, dependson = -1}
-## wrapper to simulate from a fitted GAM with the arguments/settings
-## I want
+## wrapper to simulate from a fitted GAM with the required arguments
 simulateSWModels <- function(mod, newdata, nsim, seed = 42) {
@@
 
-plot_grid(plt1, plt2, plt3, ncol = 1, labels = "auto", align = "hv",
-          axis = "lrtb", rel_widths = c(0.5, 1, 1))
+## align all plots vertically
+plots <- align_plots(plt1, plt2, plt3, align = 'v', axis = 'l')
+
+## create the two rows of figures, from `plots`
+top_row <- plot_grid(plots[[1]], NULL, ncol = 2, labels = "a")
+bot_row <- plot_grid(plots[[2]], plots[[3]], ncol = 1, labels = c("b", "c"))
+
+## combine the two rows, top row has 1 plot row, bottom row has 2, hence
+## the rel_heights to even this out
+plot_grid(top_row, bot_row, ncol = 1, rel_heights = c(0.5, 1))
 ```

### 2018-08-15 supplementary-materials.tex
revise the materials

@@
 \setlength{\droptitle}{-2em}
+
   \title{Supplementary materials for: Modelling palaeoecological time series
@@
   \postdate{\par}
-  \date{May 13, 2018}
+    \date{August 14, 2018}
 
@@
 The following packages are required: \emph{mgcv}, \emph{scam},
-\emph{ggplot2}, \emph{cowplot}, and \emph{tidyr}. Also, the
-\emph{schoenberg} package is required; it is not on CRAN but can be
-installed from Github.
+\emph{ggplot2}, \emph{cowplot}, and \emph{tidyr}.
 
@@
 \begin{verbatim}
-#> This is mgcv 1.8-23. For overview type 'help("mgcv-package")'.
+#> This is mgcv 1.8-24. For overview type 'help("mgcv-package")'.
 \end{verbatim}
@@
 \begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(}\StringTok{"tidyr"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+In addition, the \emph{schoenberg} package is required; this
+in-development package is not on CRAN but can be installed directly from
+GitHub using functions from the \emph{devtools} package. To install the
+package, install \emph{devtools} and then use
+\texttt{devtools::install\_github()} to install \emph{schoenberg}, as
+shown below:
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{## schoenberg is not on CRAN, install from github:}
-\NormalTok{## install.packages("devtools")}
-\NormalTok{## devtools::install_github("gavinsimpson/schoenberg")}
-\KeywordTok{library}\NormalTok{(}\StringTok{"schoenberg"}\NormalTok{)}
-\KeywordTok{library}\NormalTok{(}\StringTok{"tidyr"}\NormalTok{)}
-
+\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
+\NormalTok{devtools::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"gavinsimpson/schoenberg"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Assuming that the installation of \emph{schoenberg} completes without
+error, the package can be loaded as usual
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(}\StringTok{"schoenberg"}\NormalTok{) }\CommentTok{# need to change the name of the package}
+\end{Highlighting}
+\end{Shaded}
+
+The final environment-preparation step is to set the default
+\emph{ggplot} theme, which the loading of \emph{cowplot} has
+over-ridden. Here I use the more-minimal black-and-white theme
+(\texttt{theme\_bw()})
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{## Default ggplot theme}
@@
 
-The example data sets are also stored on github;
+\section{Load the data sets}\label{load-the-data-sets}
+
+The example data sets are also stored on GitHub;
 \url{https://github.com/gavinsimpson/frontiers-palaeo-additive-modelling}.
@@
                     \DataTypeTok{skip =} \DecValTok{84}\NormalTok{)}
+\NormalTok{## clean up variable names}
 \KeywordTok{names}\NormalTok{(braya) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Depth"}\NormalTok{, }\StringTok{"DepthUpper"}\NormalTok{, }\StringTok{"DepthLower"}\NormalTok{, }\StringTok{"Year"}\NormalTok{, }\StringTok{"YearYoung"}\NormalTok{,}
                   \StringTok{"YearOld"}\NormalTok{, }\StringTok{"UK37"}\NormalTok{)}
+\NormalTok{## add a variable for the amount of time per sediment sample}
 \NormalTok{braya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(braya, }\DataTypeTok{sampleInterval =} \NormalTok{YearYoung -}\StringTok{ }\NormalTok{YearOld)}
@@
 
+Instead of following the structure of the paper exactly, the subsequent
+sections focus on the two example data sets in turn, beginning with
+Small Water.
+
+\subsection{Small Water}\label{small-water}
+
+A GAM is typically fitted using the \texttt{gam()} function from the
+\emph{mgcv} package. The typical call includes a formula describing the
+response variable and the linear predictor separated by the
+\texttt{\textasciitilde{}} (tilde) symbol. The linear predictor part of
+the model contains the smooth function of the time variable in the data
+set, and is indicated using the \texttt{s()} function. Unless other
+options are needed, we only need specify where the variables in the
+formula can be found via the \texttt{data} argument, and that we wish to
+use REML smoothness selection.
+
+Putting this together we have the following function call to fit a
+simple GAM to the Small Water isotope data
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{m <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+As discussed in the paper, this model assumes that residuals\footnote{or
+  equivalently that the observations, conditional upon the model, are
+  independent.} are independent. To account for residual temporal
+autocorrelation we can include a continuous-time first-order
+autoregressive (CAR(1)) process in the model residuals for GAMs fitted
+to data that are conditionally distributed Gaussian.
+
 The GAM plus CAR(1) process is fitted to the Small Water data set using
@@
 \begin{Highlighting}[]
-\NormalTok{## fit small water GAM usin gamm() with a CAR(1)}
+\NormalTok{## fit small water GAM using gamm() with a CAR(1)}
 \NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small,}
@@
 
-The model summary is prepared from the \texttt{\$gam} component of the
-fitted model
+The object returned by \texttt{gamm()} includes both the linear mixed
+model and GAM faces of the model, and as a result we need to access the
+separate elements (\texttt{lme} and \texttt{gam} respectively) when
+proceeding to explore the model fit. The model summary is prepared from
+the \texttt{\$gam} component of the fitted model
 
@@
 
-The output shows the estimated complexity of the fitted smooth, in terms
-of the effective degrees of freedom of the spline. An associated \(F\)
-statistic and test of the null hypothesis of no trend (effect). Here the
-estimated trend provides strong evidence against this null.
+The output shows the estimated complexity of the fitted smooth,
+expressed in terms of the effective degrees of freedom of the spline. An
+associated \(F\) statistic and test of the null hypothesis of no trend
+(effect) are also shown. Here the estimated trend provides strong
+evidence against this null.
 
@@
 \NormalTok{## plot CAR(1) process}
-\NormalTok{maxS <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(Year))) ## too large, truncate to 50}
 \NormalTok{S <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DataTypeTok{length =} \DecValTok{100}\NormalTok{)}
-
 \NormalTok{car1 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{t}\NormalTok{(}\KeywordTok{outer}\NormalTok{(smallPhi, S, }\DataTypeTok{FUN =} \StringTok{`}\DataTypeTok{^}\StringTok{`}\NormalTok{)[}\DecValTok{1}\NormalTok{, , ])),}
@@
 
-The same model is fitted to the Braya-Sø data set. Note however that in
-order to even fit the model with both a smooth and the CAR(1) process, I
-have had to change the default optimiser used to fit the model, and
-reduce the basis dimension to a small number. We also fit the model
-using GCV, which is the defaut, hence no \texttt{method} argument
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## fit the car(1) model --- needs optim as this is not a stable fit!}
-\NormalTok{## also needs k setting lower than default}
-\NormalTok{braya.car1 <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya, }
-                   \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year),}
-                   \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
-                   \DataTypeTok{control =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{niterEM =} \DecValTok{0}\NormalTok{, }\DataTypeTok{optimMethod =} \StringTok{"BFGS"}\NormalTok{, }
-                                  \DataTypeTok{opt =} \StringTok{"optim"}\NormalTok{))}
-
-\NormalTok{## fit model using GCV}
-\NormalTok{braya.gcv <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{30}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya)}
-
-\NormalTok{## estimate of phi and confidence interval}
-\NormalTok{brayaPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(braya.car1$lme)$corStruct}
-\NormalTok{brayaPhi}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-#>            lower      est. upper
-#> Phi 1.611049e-22 0.2000156     1
-#> attr(,"label")
-#> [1] "Correlation structure:"
-\end{verbatim}
-
-Note the wide confidence interval --- effectively 0--1 --- on \(\phi\).
-If you were to increase the value of \texttt{k} to be \texttt{k\ =\ 10}
-in the \texttt{s(Year)} above, the model will fit but a warning message
-will be emitted when trying to extract \(\phi\) due to a non-positive
-definite model covariance matrix, indicating problems with the model.
-
-The next couple of code chunks prepare plots of the fitted GAMS. The
-general idea is to predict fom the fitted model for a fine grid of
-points over the range of the time variable. First we plot the trend for
-Small Water with an approximate 95\% confidence interval assuming
+The next code chunk prepares a plot of the fitted GAMS. The general idea
+is to predict from the fitted model for a fine grid of points over the
+range of the time variable. Below I plot the trend for Small Water with
+an approximate 95\% point-wise confidence interval that assumes
 asymptotic normality
@@
 \begin{Highlighting}[]
-\NormalTok{N <-}\StringTok{ }\DecValTok{300}   \CommentTok{# number of points at which to evaluate the splines}
-\NormalTok{## Predict from the fitted model}
+\NormalTok{N <-}\StringTok{ }\DecValTok{300}   \CommentTok{# number of points at which to evaluate the smooth}
+
+\NormalTok{## create new data to predict at; 200 evenly-spaced values over `Year`}
 \NormalTok{newYear <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                              \DataTypeTok{length.out =} \DecValTok{200}\NormalTok{)))}
+\NormalTok{## Predict from the fitted model; note we predict from the $gam part}
 \NormalTok{newYear <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newYear,}
                  \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(mod$gam, newYear, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
+\NormalTok{## Create the confidence interval}
+\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\DataTypeTok{df =} \KeywordTok{df.residual}\NormalTok{(mod$gam))}
 \NormalTok{newYear <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newYear,}
-                     \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(}\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{se.fit),}
-                     \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(}\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{se.fit))}
-
-\NormalTok{## Plot simulated trends}
+                     \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
+                     \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}
+
+\NormalTok{## Plot estimated trend}
 \NormalTok{small_fitted <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(newYear, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
@@
 
+\subsection{Braya-Sø}\label{braya-s}
+
+The same model is fitted to the Braya-Sø data set. Note however that in
+order to even fit the model with both a smooth and the CAR(1) process, I
+have had to change the default optimiser used to estimate the model
+parameters, and reduce the basis dimension to a small number. We also
+fit the model using GCV, which is the defaut, hence no \texttt{method}
+argument
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{## fit the car(1) model --- needs optim as this is not a stable fit!}
+\NormalTok{## also needs k setting lower than default}
+\NormalTok{braya.car1 <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya, }
+                   \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year),}
+                   \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
+                   \DataTypeTok{control =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{niterEM =} \DecValTok{0}\NormalTok{, }\DataTypeTok{optimMethod =} \StringTok{"BFGS"}\NormalTok{, }
+                                  \DataTypeTok{opt =} \StringTok{"optim"}\NormalTok{))}
+
+\NormalTok{## fit model using GCV}
+\NormalTok{braya.gcv <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{30}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya)}
+
+\NormalTok{## estimate of phi and confidence interval}
+\NormalTok{brayaPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(braya.car1$lme)$corStruct}
+\NormalTok{brayaPhi}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#>            lower      est. upper
+#> Phi 1.611049e-22 0.2000156     1
+#> attr(,"label")
+#> [1] "Correlation structure:"
+\end{verbatim}
+
+Note the wide confidence interval --- effectively 0--1 --- on \(\phi\).
+If you were to increase the value of \texttt{k} to be \texttt{k\ =\ 10}
+in the \texttt{s(Year)} above, the model will fit but a warning message
+will be emitted when trying to extract \(\phi\) due to a non-positive
+definite model covariance matrix, indicating problems with the model.
+
 For Braya-Sø, we repeat the process, but we do so for both models (GAMM
@@
 \begin{Highlighting}[]
-\NormalTok{## ggplot with data and fitted spline, then resids vs time in second panel}
+\NormalTok{N <-}\StringTok{ }\DecValTok{300}   \CommentTok{# number of points at which to evaluate the smooth}
+
+\NormalTok{## data to predict at}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                               \DataTypeTok{length.out =} \NormalTok{N)))}
+\NormalTok{## add predictions from GAMM + CAR(1) model}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newBraya,}
@@
                       \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}
+
 \NormalTok{## add GAM GCV results}
 \NormalTok{fit_gcv <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(braya.gcv, }\DataTypeTok{newdata =} \NormalTok{newBraya, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)}
-\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(newBraya, newBraya) }\CommentTok{# extend newBraya to take GCV results}
-\NormalTok{newBraya[}\KeywordTok{seq}\NormalTok{(N}\DecValTok{+1}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{N, }\DataTypeTok{by =} \DecValTok{1}\NormalTok{), ]$fit <-}\StringTok{ }\NormalTok{fit_gcv$fit}
-\NormalTok{newBraya[}\KeywordTok{seq}\NormalTok{(N}\DecValTok{+1}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{N, }\DataTypeTok{by =} \DecValTok{1}\NormalTok{), ]$upper <-}
-\StringTok{    }\NormalTok{fit_gcv$fit +}\StringTok{ }\NormalTok{(}\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\KeywordTok{df.residual}\NormalTok{(braya.gcv)) *}\StringTok{ }\NormalTok{fit_gcv$se.fit)}
-\NormalTok{newBraya[}\KeywordTok{seq}\NormalTok{(N}\DecValTok{+1}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{N, }\DataTypeTok{by =} \DecValTok{1}\NormalTok{), ]$lower <-}
-\StringTok{    }\NormalTok{fit_gcv$fit -}\StringTok{ }\NormalTok{(}\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\KeywordTok{df.residual}\NormalTok{(braya.gcv)) *}\StringTok{ }\NormalTok{fit_gcv$se.fit)}
+\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\KeywordTok{df.residual}\NormalTok{(braya.gcv))}
+\NormalTok{newGCV <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year   =} \NormalTok{newBraya[[}\StringTok{"Year"}\NormalTok{]],}
+                     \DataTypeTok{fit    =} \NormalTok{fit_gcv$fit,}
+                     \DataTypeTok{se.fit =} \NormalTok{fit_gcv$se.fit)}
+\NormalTok{newGCV <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newGCV,}
+                    \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
+                    \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}
+\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(newBraya, newGCV)                }\CommentTok{# bind on GCV results}
+\NormalTok{## Add indicator variable for model}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newBraya,}
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/manuscript-fig-2-1} \end{center}
+\subsubsection{Checking if the size of the basis expansion is
+sufficient}\label{checking-if-the-size-of-the-basis-expansion-is-sufficient}
+
+Clearly we are not able to identify both the wiggly trend and the CAR(1)
+process in a single model. We might proceed by fitting a simple GAM via
+\texttt{gam()} with REML smoothness selection and moderate number of
+basis functions --- here we set \texttt{k\ =\ 15} for illustration.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{braya_low_k <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Before proceeding, we should perform a check to determine if the number
+of basis functions requested is sufficient to capture the wiggliness in
+the data. This test is provided by \texttt{gam.check()}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{gam.check}\NormalTok{(braya_low_k)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#> 
+#> Method: REML   Optimizer: outer newton
+#> full convergence after 6 iterations.
+#> Gradient range [-4.468978e-08,8.696688e-10]
+#> (score -187.2524 & scale 0.000689802).
+#> Hessian positive definite, eigenvalue range [0.5525272,43.5153].
+#> Model rank =  15 / 15 
+#> 
+#> Basis dimension (k) checking results. Low p-value (k-index<1) may
+#> indicate that k is too low, especially if edf is close to k'.
+#> 
+#>            k'   edf k-index p-value    
+#> s(Year) 14.00  2.62    0.56  <2e-16 ***
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{verbatim}
+
+In the output, the column labelled \texttt{k\textquotesingle{}} is the
+size of the basis used (the number of basis functions) --- note that it
+is 14 and not the requested 15 functions as the constant basis function
+has been removed as it is confounded with the model constant term, the
+intercept. The column labelled \texttt{edf} indicates the
+\emph{effective degrees of freedom} for the estimated smooth. The
+\texttt{k-index} column contains the value of a test statistic for a
+test of sufficient number of basis functions --- ideally, we'd like the
+value of \texttt{k-index} to be close to or greater than 1. The quoted
+\emph{p} value measures the support for the null hypothesis that enough
+basis functions were available. In practial terms, the low
+\texttt{k-index} and very low \emph{p} value suggest that the requested
+number of basis functions used to fit the model was too low.
+
+The solution is to increase \texttt{k}, say by doubling the original
+value, and refit the model and perform the basis dimension check using
+\texttt{gam.check()}, repeating this process until \texttt{k-index} is
+close to or greater than 1 and the \emph{p} value is larger than a
+desired threshold. In my experience, it may not be possible with some
+data sets to increase \texttt{k} large enough to result in a
+\texttt{k-index} \textgreater{} 1 and a \emph{p} value \textgreater{}
+0.05, given the available. It is important to remember that the test
+provided by \texttt{gam.check()} is only an heuristic one and shouldn't
+be seen as infallible. If you find yourself increasing \texttt{k} to
+large values relative to the number of samples in your data set without
+achieving a \texttt{k-index} \textgreater{} 1 or \emph{p} \textgreater{}
+0.05, then note the value in the \texttt{edf} column as you increase
+\texttt{k}. If the \texttt{edf} of the model changes little as you
+continue to increase \texttt{k}, this may be evidence that the test is
+failing in this particular instance and that the estimated smooth is not
+dependent on the value of \texttt{k} chosen.
+
+\subsection{Accounting for heteroscedasticity due to time
+averaging}\label{accounting-for-heteroscedasticity-due-to-time-averaging}
 
 To proceed with the Braya-Sø example, we need to increase the basis
-dimension (\texttt{k\ =\ 45}), fit using \texttt{method\ =\ "REML"}, and
-use observational weights. Here I use the \texttt{sampleInterval}
-variable as the measure of lake years per sample, and to avoid changing
-the model likelihood, the weights are actually the values of
-\texttt{sampleInterval} divided by the mean of \texttt{sampleInterval}:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## TPRS, weights as sampleInterval}
-\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"tp"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
+dimension (\texttt{k\ =\ 45}), fit using \texttt{method\ =\ "REML"}. As
+discussed in the manuscript, we should also account for the non-constant
+variance, or \emph{heteroscedasticity}, that may arise due to variation
+in the amount of time (or ``lake years'') that each sample represents.
+All else equal, we expect that samples that average more time have
+smaller variance than those that average a smaller amount of time. To
+include information about the expected heteroscedasticity in the GAM we
+can use observational weights\footnote{This simple solution is for a
+  Gaussian GAM. For models with other distributions, the exact
+  interpretation of observational weights will vary. For such data,
+  where the conditional mean and variance of the observations are
+  linked, the use of location-scale, or distributional, models may be
+  more apporpriate and informative.}. Here I use the
+\texttt{sampleInterval} variable that I created earlier as the measure
+of lake years per sample, and to avoid changing the model likelihood,
+the weights use are the values of \texttt{sampleInterval} divided by the
+mean of \texttt{sampleInterval}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
                   \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                   \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
-\end{Highlighting}
-\end{Shaded}
+\KeywordTok{summary}\NormalTok{(braya_reml)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#> 
+#> Family: gaussian 
+#> Link function: identity 
+#> 
+#> Formula:
+#> UK37 ~ s(Year, k = 45)
+#> 
+#> Parametric coefficients:
+#>              Estimate Std. Error t value Pr(>|t|)    
+#> (Intercept) -0.633734   0.001901  -333.3   <2e-16 ***
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> Approximate significance of smooth terms:
+#>           edf Ref.df     F p-value    
+#> s(Year) 28.49  33.99 7.755  <2e-16 ***
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> R-sq.(adj) =  0.751   Deviance explained = 83.2%
+#> -REML = -175.68  Scale est. = 0.00024036  n = 89
+\end{verbatim}
+
+If we now check if 44 (\texttt{k\ =\ 45}) basis functions is sufficient
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{gam.check}\NormalTok{(braya_reml)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#> 
+#> Method: REML   Optimizer: outer newton
+#> full convergence after 7 iterations.
+#> Gradient range [-2.210576e-06,-5.181457e-08]
+#> (score -175.684 & scale 0.0002403604).
+#> Hessian positive definite, eigenvalue range [2.524305,48.11146].
+#> Model rank =  45 / 45 
+#> 
+#> Basis dimension (k) checking results. Low p-value (k-index<1) may
+#> indicate that k is too low, especially if edf is close to k'.
+#> 
+#>           k'  edf k-index p-value
+#> s(Year) 44.0 28.5    1.27    0.99
+\end{verbatim}
+
+we note that the value of \texttt{k-index} is now greater than 1 and the
+\emph{p} value suggests there is very little evidence against the null
+hypothesis. Also note that the \texttt{edf} of the estimated smooth is
+considerably below the maximum (\texttt{k\textquotesingle{}}). This is
+the effect of the smoothness penalty removing the excessive wiggliness
+possible with 44 basis functions. If the \texttt{edf} were close to
+\texttt{k\textquotesingle{}}, you might consider increasing \texttt{k}
+by a modest amount (in this instance by 5 or 10 additional basis
+functions) to assure yourself that you have sufficient basis functions.
+This has the additional advantage of allowing a richer set of smooths of
+a given complexity (\texttt{edf}) to be represented using the basis
+functions, which may improve the model fit to the data without changing
+the \texttt{edf} too much.
 
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure-7-1} \end{center}
-
 \section{Confidence and simultaneous
@@
 given via the \texttt{type} argument with options \texttt{"confidence"}
-and \texttt{"simultaneous"}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## small water}
+and \texttt{"simultaneous"}. Fro example, for Small Water we would use
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{sw.cint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(mod, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newYear,}
@@
                    \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{)}
-
+\KeywordTok{head}\NormalTok{(sw.sint)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#>    smooth     Year      est        se     crit    lower    upper
+#> 1 s(Year) 1846.319 3.787215 0.1616144 3.087688 3.288200 4.286230
+#> 2 s(Year) 1847.128 3.789747 0.1509100 3.087688 3.323784 4.255710
+#> 3 s(Year) 1847.937 3.792304 0.1411439 3.087688 3.356496 4.228112
+#> 4 s(Year) 1848.745 3.794914 0.1324961 3.087688 3.385807 4.204020
+#> 5 s(Year) 1849.554 3.797602 0.1251070 3.087688 3.411310 4.183893
+#> 6 s(Year) 1850.363 3.800394 0.1190552 3.087688 3.432788 4.167999
+\end{verbatim}
+
+The \texttt{confint()} methods return data frames suitable for
+subsequent plotting with \textbf{ggplot}. The columns labelled
+\texttt{est} and \texttt{se} are the estimate values of the smooth and
+its standard error, respectively. The variables \texttt{lower} and
+\texttt{upper} contain the values of the lower and upper bounds on the
+requested interval. The intervals can be plotted as follows
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{smallInt.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(sw.cint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/small-compare-intervals-1} \end{center}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## braya so}
+\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/plot-small-water-intervals-1} \end{center}
+
+The intervals for Braya-Sø are created
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{bs.cint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(braya_reml, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newBraya,}
@@
                    \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{)}
-
+\end{Highlighting}
+\end{Shaded}
+
+and plotted
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{brayaInt.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(bs.cint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/braya-compare-intervals-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/plot-braya-so-intervals-1} \end{center}
+
+in the same way
 
@@
 \end{Shaded}
-
-\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure-8-1} \end{center}
 
@@
 \NormalTok{Mat <-}\StringTok{ }\NormalTok{SEx <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DataTypeTok{length =} \NormalTok{nn)     }\CommentTok{# object to hold model fits}
+
+\NormalTok{## loop over the sequence of deparation distances and fit the models}
 \NormalTok{for (i in }\KeywordTok{seq_along}\NormalTok{(dseq)) \{ }
@@
 \begin{Highlighting}[]
-\NormalTok{effRange1 <-}\StringTok{ }\DecValTok{250}      \CommentTok{# sets effective range in years for Matérn correl}
-\NormalTok{## minima from profile likelihood}
+\NormalTok{## effective range minima from profile likelihood}
 \NormalTok{effRange2 <-}\StringTok{ }\KeywordTok{with}\NormalTok{(}\KeywordTok{subset}\NormalTok{(reml.scr, cor ==}\StringTok{ "Matérn"}\NormalTok{), dseq[}\KeywordTok{which.min}\NormalTok{(reml)])}
@@
 
-\NormalTok{## Matern}
+\NormalTok{## Refit these models: Matern}
 \NormalTok{gp2 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, effRange2)),}
@@
            \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{, }\DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
-\NormalTok{## Power exponential}
+\NormalTok{## Refit these models: Power exponential}
 \NormalTok{gp3 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, effRange3, }\DecValTok{1}\NormalTok{)),}
@@
 
+\NormalTok{## data to predict at}
 \NormalTok{newd <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                \DataTypeTok{length.out =} \DecValTok{1000}\NormalTok{)))}
+\NormalTok{## create predictions on response scale for both covariance functions}
 \NormalTok{p.gp2 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newd,}
@@
                    \DataTypeTok{effRange =} \KeywordTok{round}\NormalTok{(effRange3))}
-\NormalTok{## pred <- rbind(p.gp1, p.gp2, p.gp3)}
+\NormalTok{## joint the two sets of predictions together}
 \NormalTok{pred <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(p.gp2, p.gp3)}
-\NormalTok{pred <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(pred, }\DataTypeTok{effRange =} \KeywordTok{factor}\NormalTok{(effRange),}
+\NormalTok{## add some categorical variables for plotting}
+\NormalTok{pred <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(pred,}
+                  \DataTypeTok{effRange =} \KeywordTok{factor}\NormalTok{(effRange),}
                   \DataTypeTok{cor =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"Matérn"}\NormalTok{, }\StringTok{"Exponential"}\NormalTok{), }\DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(newd)))}
@@
 
-The estimated trends are plotted using
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## plot at two values of h}
+The estimated Gaussian process trends are plotted using
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{gp.plt2 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(pred, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fitted, }\DataTypeTok{colour =} \NormalTok{cor)) +}
@@
 \StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+\NormalTok{gp.plt2}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/plot-both-gp-smooth-trends-1} \end{center}
 
@@
 \end{Shaded}
-
-\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure-12-1} \end{center}
 
@@
 
-\section{Compare various trends}\label{compare-various-trends}
+\section{Comparing trends}\label{comparing-trends}
 
@@
 \item
-  Thin plate regression spline,
+  Thin plate regression spline (TPRS),
 \item
@@
 \begin{Highlighting}[]
-\NormalTok{## model it using gam()}
+\NormalTok{## effective range parameter from profile-likelihood of Matern model}
 \NormalTok{effRange <-}\StringTok{ }\NormalTok{effRange2}
 
-\NormalTok{## TPRS, weights as sampleInterval, k = needs to be higher}
+\NormalTok{## TPRS, weights as sampleInterval}
 \NormalTok{mod_tprs <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"tp"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
@@
 \begin{Highlighting}[]
-\NormalTok{## wrapper to simulate from a fitted GAM with the arguments/settings}
-\NormalTok{## I want}
+\NormalTok{## wrapper to simulate from a fitted GAM with the required arguments}
 \NormalTok{simulateSWModels <-}\StringTok{ }\NormalTok{function(mod, newdata, nsim, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{) \{}
@@
 
-\KeywordTok{plot_grid}\NormalTok{(plt1, plt2, plt3, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{, }\DataTypeTok{align =} \StringTok{"hv"}\NormalTok{,}
-          \DataTypeTok{axis =} \StringTok{"lrtb"}\NormalTok{, }\DataTypeTok{rel_widths =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{))}
+\NormalTok{## align all plots vertically}
+\NormalTok{plots <-}\StringTok{ }\KeywordTok{align_plots}\NormalTok{(plt1, plt2, plt3, }\DataTypeTok{align =} \StringTok{'v'}\NormalTok{, }\DataTypeTok{axis =} \StringTok{'l'}\NormalTok{)}
+
+\NormalTok{## create the two rows of figures, from `plots`}
+\NormalTok{top_row <-}\StringTok{ }\KeywordTok{plot_grid}\NormalTok{(plots[[}\DecValTok{1}\NormalTok{]], }\OtherTok{NULL}\NormalTok{, }\DataTypeTok{ncol =} \DecValTok{2}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"a"}\NormalTok{)}
+\NormalTok{bot_row <-}\StringTok{ }\KeywordTok{plot_grid}\NormalTok{(plots[[}\DecValTok{2}\NormalTok{]], plots[[}\DecValTok{3}\NormalTok{]], }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \KeywordTok{c}\NormalTok{(}\StringTok{"b"}\NormalTok{, }\StringTok{"c"}\NormalTok{))}
+
+\NormalTok{## combine the two rows, top row has 1 plot row, bottom row has 2, hence}
+\NormalTok{## the rel_heights to even this out}
+\KeywordTok{plot_grid}\NormalTok{(top_row, bot_row, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{rel_heights =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\DecValTok{1}\NormalTok{))}
 \end{Highlighting}

### 2018-08-15 manuscript.Rmd
minor change  to manuscript

--- 2018-06-25 manuscript.Rmd
@@
 mail: "gavin.simpson@uregina.ca"
-date: May 11, 2018
+date: August 14, 2018
 keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 

### 2018-08-15 manuscript.Rmd
insert flow diagram; bump date

--- 2018-06-25 manuscript.Rmd
@@
 mail: "gavin.simpson@uregina.ca"
-date: May 11, 2018
+date: August 15, 2018
 keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
@@
 In the subsequent sections, I present an introduction to GAMs and discuss the issue of uncertainty in model-estimated trends, the topic of posterior simulation from a regression model and how to identify periods of significant environmental change using the first derivative of the estimated trend.
+The main steps in the analysis of palaeoenvironmental time series using GAMs are illustrated in Figure \ref{fig:flowchart}.
 Two non-standard types of spline --- adaptive smoothers and Gaussian process splines --- that are especially applicable to GAMs in the palaeoenvironmental setting are subsequently described, followed by an assessment of the the impact of age-model uncertainty on trend estimation via GAMs.
 Finally, I briefly discuss the application of GAM trend analysis to multivariate species abundance and compositional data.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.4\linewidth]{manual-figures/flow-diagram}
+
+}
+
+\caption{Flowchart showing the main steps in the analysis of time series using generalized additive models.}\label{fig:flowchart}
+\end{figure}
 

### 2018-08-15 supplementary-materials.Rmd
bump date

@@
 mail: "gavin.simpson@uregina.ca"
-date: August 14, 2018
+date: August 15, 2018
 keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 

### 2018-08-16 manuscript.Rmd
mass rebuild after switching to gratia package

@@
 mail: "gavin.simpson@uregina.ca"
-date: August 15, 2018
+date: August 16, 2018
 keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
@@
 library("grid")                         # for unit.pmax(), unit.list()
-library("schoenberg")
+library("gratia")
 library("tidyr")
@@
 
-Palaeoecology and palaeolimnology have moved away from being descriptive disciplines, rapidly adopting new statistical developments in the 1990s and beyond [@@Smol2012-xt].
+Palaeoecology and palaeolimnology have moved away from being descriptive disciplines, rapidly adopting new statistical developments in the 1990s and beyond [@Smol2012-xt].
 Less development has been observed in the area of trend estimation in palaeoenvironmental time series.
@@
 Irregular intervals between samples means that the time-series analysis methods of autoregressive or moving average processes, in the form of autoregressive integrated moving average (ARIMA) models, are practically impossible to apply because software typically requires even spacing of observations in time.
-@Dutilleul2012-su and @Birks2012-op, eschewing the term *time series*, prefer to call such data *temporal series* on account of the irregular spacing of samples, a distinction that I find unnecessary.
+@Dutilleul2012-su and @Birks2012-op, eschewing the term *time series*, prefer to call such data *temporal series* on account of the irregular spacing of samples, a distinction that I find unnecessary, however.
 
@@
 \loess{}, locally weighted scatterplot smoother, as it's name suggests, was developed to smooth x-y scatterplot data [@Cleveland1979-vf].
-The method fits a smooth line through data by fitting weighted least squares (WLS) models to observations within a particular, user-specified window of the focal point, whose width is typically expressed as a proportion $\alpha$ of the $n$ data points.
+The method fits a smooth line through data by fitting weighted least squares (WLS) models to observations within a user-specified window of the focal point, whose width is typically expressed as a proportion $\alpha$ of the $n$ data points.
 Weights are determined by how close (in the x-axis only) an observation in the window is to the focal point giving greatest weight given to points closest to the focal point.
@@
 Between them, the two examples, combine many of the features of interest to palaeoecologists that motivate the use of GAMs; non-linear trends and the question of when changes in the measured proxy occurred.
-The example analyses were all performed using the *mgcv* package [version 1.8.23; @Wood2017-qi] and R [version 3.4.4; @Rcore], and the supplementary material contains a fully annotated document showing the R code used to replicate all the analyses described in the remainder of the paper.
+The example analyses were all performed using the *mgcv* package [version 1.8.24; @Wood2017-qi] and R [version 3.4.4; @Rcore], and the supplementary material contains a fully annotated document showing the R code used to replicate all the analyses described in the remainder of the paper.
 
@@
 Alkenones are long-chained unsaturated organic compounds that are produced by a small number of planktonic organisms known as haptophytes.
-The \uk{} unsaturation index is [@Brassell1993-fw]
+The \uk{} unsaturation index [@Brassell1993-fw] is
 \begin{equation*}
@@
 
-It is clear from the plots of the data (Figure\ \ref{fig:data-figure}) that we require the fitted trends for the Small Water δ^15^N and Braya-Sø \uk{} time series to be non-linear functions, but it is less clear how to specify the actual shape require.
+It is clear from plots of the data (Figure\ \ref{fig:data-figure}) that we require the fitted trends for the Small Water δ^15^N and Braya-Sø \uk{} time series to be non-linear functions, but it is less clear how to specify the actual shape require.
 Ideally, we'd like to learn the shape of the trend from the data themselves.
@@
 Typically in this parametrization there are many fewer knots than unique data, with the knots distributed evenly over the range of $x_t$ or at the quantiles of $x_t$.
-Placing knots at the quantiles of $x_t$ has the effect of placing a greater number of knots where the data is densest.
+Placing knots at the quantiles of $x_t$ has the effect of placing a greater number of knots where the data is most dense.
 
@@
 Note that in order to estimate a coefficient for each basis function the model has to be fitted without an intercept term.
-In practice we would include an intercept term in the model and therefore the basis functions are modified via an identifiability constraint [@@Wood2017-qi].
+In practice we would include an intercept term in the model and therefore the basis functions are modified via an identifiability constraint [@Wood2017-qi].
 This has the effect of making the basis orthogonal to the intercept but results in more complicated basis functions than those shown in in Figure \ref{fig:basis-function-example-plot}a.
@@
 The scaled CRS basis functions for the Small Water time series are shown in Figure \ref{fig:basis-function-example-plot}b.
-The solid line passing through the data points is formed by summing up the values of the scaled basis functions ($b_j(x_t) \beta_j$) at any value of $x_t$ (time).
+The solid line passing through the data points is formed by summing up the values of the seven scaled basis functions ($b_j(x_t) \beta_j$) at any value of $x_t$ (time).
 
@@
 With practice and some knowledge of the system from which the observations arise, it can be relatively easy to put an upper limit on the expected complexity of the true trend in the data.
-Additionally, the number available data points places a constraint on the upper limit of the size of basis expansion that can be used.
-
-In practice, the size of the basis is an upper limit on the expected complexity of the trend, and a simple test to check if the basis used was sufficiently large [@Pya2016-rk].
+Additionally, the number of available data points places a constraint on the upper limit of the size of basis expansion that can be used.
+
+In practice, the size of the basis is an upper limit on the expected complexity of the trend, and a simple test can be used to check if the basis used was sufficiently large [@Pya2016-rk].
 This test is available via the `gam.check()` function in *mgcv* for example, which essentially looks at whether there is any additional nonlinearity or structure in the residuals that can be explained by a further smooth of $x_t$.
@@
 After standard modelling choices (size of basis to use, correct selection of response distribution and link function, etc.) are checked, it often pays to think carefully about the properties of the data and ensure that the assumptions of the model are met.
-Here, despite increasing the basis size, it was the failure to appreciate the magnitude of the effect on fitting of the non-constant variance that lead to the initially poor fit and the problems associated with the estimation of the CAR(1) process.
-I return to the issue of why the GAM plus CAR(1) model encountered problems during fitting later ([see Identifiability](#identifiability)).
+Here, despite increasing the basis size, it was the failure to appreciate the magnitude of the effect of the non-constant variance that lead to the initially poor fit and the problems associated with the estimation of the CAR(1) process.
+I return to the issue of why the GAM plus CAR(1) model encountered problems during fitting later ([see section Residual autocorrelation and model identification](#identifiability)).
 
@@
 
-If we want to ask whether either of the estimated trends is statistically interesting or proceed to identifying periods of significant change, we must address the issue of uncertainty in the estimated model.
+If we want to ask whether either of the estimated trends is statistically interesting or proceed to identify periods of significant change, we must address the issue of uncertainty in the estimated model.
 What uncertainty is associated with the trend estimates?
 One way to visualise this is through a 1 - $\alpha$ confidence interval around the fitted trend, where $\alpha$ is say 0.05 leading to a 95% interval.
-A 95% interval would be drawn at $\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))$, with $m_{1-\alpha} = 1.96$, the 0.95 probability quantile of a standard normal distribution, and $\text{SE}(\hat{y}_t)$ is the standard error of the estimated trend at time $x_t$.
+A 95% interval would be drawn at $\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))$, with $m_{1-\alpha} = 1.96$, the 0.95 probability quantile of a standard normal distribution[^tdistrib], and $\text{SE}(\hat{y}_t)$ is the standard error of the estimated trend at time $x_t$.
 This type of confidence interval would normally be described as *pointwise*; the coverage properties of the interval being correct for a single point on the fitted trend, but, if we were to consider additional points on the trend, the coverage would logically be lower than 1 - $\alpha$.
@@
 @Marra2012-bq recently explained Nychka's [-@Nychka1988-rz] surprising results and extended them to the case of generalized models (non-Gaussian responses).
+
+[^tdistrib]: The 0.95 probability quantile of the *t* distribution may be used instead, which will account for estimation of $\sigma$, the variance of the data.
+However, given the number of observations, and hence residual degrees of freedom, needed to motivate fitting GAMs, differences between intervals computed using extreme quantiles of the standard normal or the *t* distribution will be tiny.
 
@@
 
-Figure \ref{fig:posterior-simulation} shows `r take` random draws from the posterior distribution of the GAMs fitted to the Small Water and Braya-Sø data sets.
+Figure \ref{fig:posterior-simulation} shows `r take` random draws from the posterior distributions of the GAMs fitted to the Small Water and Braya-Sø data sets.
 In the early period of the δ^15^N time series many of the posterior simulations exhibit short periods of increasing and decreasing trend, balancing out to the relatively flat trend estimated by the GAM (Fig.\ \ref{fig:posterior-simulation}a).
@@
 Short runs would indicate weaker autocorrelation whilst longer runs are associated with stronger autocorrelation, and long runs of values greater (less) than the mean would be evident as non-linear trends in the time series.
-As a result, a wiggly trend and an autocorrelation function with large $\phi$ are two ways to describe the same pattern of values in a time series, and without any further information to constrain either the model is unable to distinguish both components uniquely.
+As a result, a wiggly trend and an autocorrelation function with large $\phi$ are two ways to describe the same pattern of values in a time series, and, without any further information to constrain either of these, the model is unable to distinguish both components uniquely.
 
@@
 A Gaussian process is a distribution over all possible smooth functions $f(x)$.
-In the field of spatial statistics, Gaussian processes are known by name *kriging*.
+In the field of spatial statistics, Gaussian processes are known by the name *kriging*.
 
@@
 
-Each of the spline types that I have discussed so far share a common feature; the degree of wiggliness over the time series is fixed due to the use of a single smoothness parameter, $\lambda$.
+Each of the spline types that I have discussed so far shares a common feature; the degree of wiggliness over the time series is fixed due to the use of a single smoothness parameter, $\lambda$.
 The definition of wiggliness, as the integrated squared second derivative of the spline, ensures that the fitted smoother does not jump about wildly.
@@
 Therefore, should we conclude that there is no trend in \uk{} and thence climate in this period?
-I believe that to be too-strong a statement; those oscillations in \uk{} may be real responses to climate forcing but may simply lack the statistical power to distinguish them from the null hypothesis of no trend through this period.
-The adaptive smoother is only adjusting to the data available to it; just because it does not detect a trend during this period does not lend itself to an interpretation of stable climate forcing or complacency in the lake's response to forcing.
-If there were particular interest in the climate of this particular period we might take from the Braya-Sø record that there is potential early variation in climate forcing, but that additional data from this or other sites is required before any definitive conclusion can be drawn.
+I believe that to be too-strong a statement; those oscillations in \uk{} may be real responses to climate forcing but we may simply lack the statistical power to distinguish them from the null hypothesis of no trend through this period.
+The adaptive smoother is only adjusting to the data available to it; just because it does not detect a trend during this period does not lend itself to an interpretation of stable climate forcing or complacency in the lake's response to forcing (although that is a justifiable interpretation of the result).
+If there were particular interest in the climate of this particular period we might take from the Braya-Sø record that there is potential early variation due to climate forcing, but that additional data from this or other sites is required before any definitive conclusion can be drawn.
 
@@
 Thus far, the trend models that I have described and illustrated assumed that the time covariate ($x_t$) was fixed and known.
-In both examples, and more generally for most palaeoecological records, this assumption is violated.
+In both examples, and generally for most palaeoecological records, this assumption is violated.
 Unless the record is annually laminated, assigning an age to a sediment interval requires the development of an age model from observations of the relationship between depth down the sediment core and estimates of the age of the sample arrived at using any of a number of techniques, for example ^210^Pb or ^14^C radiometric dating.
@@
 Formal statistical estimation of trends in palaeoenvironmental data has been hampered by the nature of the data that comprise the time series; the uneven spacing of samples in time makes it, if not impossible, difficult to fit classical statistical time series models like ARIMA.
-This has lead palaeoecologists and palaeolimnologists to fall back on basic statistical methods such as linear parametric and non-parametric correlations or simple linear regression models, where the assumptions of the method are often grossly violated by the dependencies inherent to time series data.
+This has led palaeoecologists and palaeolimnologists to either ignore statistical estimation of trends or fall back on basic statistical methods such as linear parametric and non-parametric correlations or simple linear regression models, where the assumptions of the method are often grossly violated by the dependencies inherent to time series data.
 GAMs, whilst similar to the popular \loess{} smoother, provide a superior alternative approach to trend estimation in palaeoenvironmental time series.
@@
 
-The ideas expressed in this paper are the result of many fruitful conversations with colleagues at the Environmental Change Research Centre, UCL, and the University of Regina.
+The ideas expressed in this paper are the result of many fruitful conversations with colleagues past and present at the Environmental Change Research Centre, UCL, and the University of Regina.
 In particular I am indebted to Helen Bennion, Rick Battarbee, and Peter Leavitt for their collaborations on projects over many years, and to David Miller, Eric Pedersen, and Noam Ross, my GAM workshop partners in crime.

### 2018-08-16 manuscript.tex
bump date
minor change  to manuscript
insert flow diagram; bump date
mass rebuild after switching to gratia package

@@
 
-\date{May 11, 2018}
+\date{August 16, 2018}
 
@@
 disciplines, rapidly adopting new statistical developments in the 1990s
-and beyond (@ Smol et al., 2012). Less development has been observed in
+and beyond (Smol et al., 2012). Less development has been observed in
 the area of trend estimation in palaeoenvironmental time series. The
@@
 \emph{temporal series} on account of the irregular spacing of samples, a
-distinction that I find unnecessary.
+distinction that I find unnecessary, however.
 
@@
 method fits a smooth line through data by fitting weighted least squares
-(WLS) models to observations within a particular, user-specified window
-of the focal point, whose width is typically expressed as a proportion
-\(\alpha\) of the \(n\) data points. Weights are determined by how close
-(in the x-axis only) an observation in the window is to the focal point
-giving greatest weight given to points closest to the focal point. The
-interim \loess{}-smoothed value for the focal point is the predicted
-value from the weighted regression at the focal point. The interim
-values are updated using weights based on how far in the y-axis
-direction the interim smoothed value lies from the observed value plus
-the x-axis distance weights; this has the effect of down-weighting
-outlier observations. The final \loess{} is obtained by joining the
-smoothed values. The user has to choose how large a window to use,
-whether to fit degree 1 (linear) or degree 2 (quadratic) polynomials in
-the WLS model, and how to weight points in the x-axis. When used in an
-exploratory mode, the user has considerable freedom to choose the detail
-of the \loess{} fit; the window width, for example, can be infinitely
-tweaked to give as close a fit to the data, as assessed by eye, as is
-desired. Using cross-validation (CV) to choose \(\alpha\) or the degree
-of polynomial in the WLS model is complicated for a number of reasons,
-not least because the CV scheme used must involve the time ordering of
-the data (e.g. Bergmeir et al., 2018). This subjectivity is problematic
+(WLS) models to observations within a user-specified window of the focal
+point, whose width is typically expressed as a proportion \(\alpha\) of
+the \(n\) data points. Weights are determined by how close (in the
+x-axis only) an observation in the window is to the focal point giving
+greatest weight given to points closest to the focal point. The interim
+\loess{}-smoothed value for the focal point is the predicted value from
+the weighted regression at the focal point. The interim values are
+updated using weights based on how far in the y-axis direction the
+interim smoothed value lies from the observed value plus the x-axis
+distance weights; this has the effect of down-weighting outlier
+observations. The final \loess{} is obtained by joining the smoothed
+values. The user has to choose how large a window to use, whether to fit
+degree 1 (linear) or degree 2 (quadratic) polynomials in the WLS model,
+and how to weight points in the x-axis. When used in an exploratory
+mode, the user has considerable freedom to choose the detail of the
+\loess{} fit; the window width, for example, can be infinitely tweaked
+to give as close a fit to the data, as assessed by eye, as is desired.
+Using cross-validation (CV) to choose \(\alpha\) or the degree of
+polynomial in the WLS model is complicated for a number of reasons, not
+least because the CV scheme used must involve the time ordering of the
+data (e.g. Bergmeir et al., 2018). This subjectivity is problematic
 however once we wish to move beyond exploratory analysis and
@@
 of significant environmental change using the first derivative of the
-estimated trend. Two non-standard types of spline --- adaptive smoothers
-and Gaussian process splines --- that are especially applicable to GAMs
-in the palaeoenvironmental setting are subsequently described, followed
-by an assessment of the the impact of age-model uncertainty on trend
+estimated trend. The main steps in the analysis of palaeoenvironmental
+time series using GAMs are illustrated in Figure \ref{fig:flowchart}.
+Two non-standard types of spline --- adaptive smoothers and Gaussian
+process splines --- that are especially applicable to GAMs in the
+palaeoenvironmental setting are subsequently described, followed by an
+assessment of the the impact of age-model uncertainty on trend
 estimation via GAMs. Finally, I briefly discuss the application of GAM
 trend analysis to multivariate species abundance and compositional data.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.4\linewidth]{manual-figures/flow-diagram}
+
+}
+
+\caption{Flowchart showing the main steps in the analysis of time series using generalized additive models.}\label{fig:flowchart}
+\end{figure}
 
@@
 proxy occurred. The example analyses were all performed using the
-\emph{mgcv} package (version 1.8.23; Wood, 2017) and R (version 3.4.4; R
+\emph{mgcv} package (version 1.8.24; Wood, 2017) and R (version 3.4.4; R
 Core Team, 2018), and the supplementary material contains a fully
@@
 compounds that are produced by a small number of planktonic organisms
-known as haptophytes. The \uk{} unsaturation index is (Brassell, 1993)
+known as haptophytes. The \uk{} unsaturation index (Brassell, 1993) is
 
@@
 
-It is clear from the plots of the data (Figure~\ref{fig:data-figure})
-that we require the fitted trends for the Small Water
-δ\textsuperscript{15}N and Braya-Sø \uk{} time series to be non-linear
-functions, but it is less clear how to specify the actual shape require.
-Ideally, we'd like to learn the shape of the trend from the data
-themselves. We will refer to these non-linear functions as \emph{smooth
-functions}, or just \emph{smooths} for short, and we will denote a
-smooth using \(f(x_t)\). Further, we would like to represent the smooths
-in a way that \eqref{eq:gam} is represented parametrically so that it
-can be estimate within the well-studied GLM framework. This is achieved
-by representing the smooth using a \emph{basis}. A basis is a set of
-functions that collectively span a space of smooths that, we hope,
-contains the true \(f(x_t)\) or a close approximation to it. The
-functions in the basis are known as \emph{basis functions}, and arise
-from a \emph{basis expansion} of a covariate. Writing \(b_j(x_t)\) as
-the \(j\)th basis function of \(x_t\), the smooth \(f(x_t)\) can be
-represented as a weighted sum of basis functions
+It is clear from plots of the data (Figure~\ref{fig:data-figure}) that
+we require the fitted trends for the Small Water δ\textsuperscript{15}N
+and Braya-Sø \uk{} time series to be non-linear functions, but it is
+less clear how to specify the actual shape require. Ideally, we'd like
+to learn the shape of the trend from the data themselves. We will refer
+to these non-linear functions as \emph{smooth functions}, or just
+\emph{smooths} for short, and we will denote a smooth using \(f(x_t)\).
+Further, we would like to represent the smooths in a way that
+\eqref{eq:gam} is represented parametrically so that it can be estimate
+within the well-studied GLM framework. This is achieved by representing
+the smooth using a \emph{basis}. A basis is a set of functions that
+collectively span a space of smooths that, we hope, contains the true
+\(f(x_t)\) or a close approximation to it. The functions in the basis
+are known as \emph{basis functions}, and arise from a \emph{basis
+expansion} of a covariate. Writing \(b_j(x_t)\) as the \(j\)th basis
+function of \(x_t\), the smooth \(f(x_t)\) can be represented as a
+weighted sum of basis functions
 
@@
 of \(x_t\). Placing knots at the quantiles of \(x_t\) has the effect of
-placing a greater number of knots where the data is densest.
+placing a greater number of knots where the data is most dense.
 
@@
 include an intercept term in the model and therefore the basis functions
-are modified via an identifiability constraint (@ Wood, 2017). This has
+are modified via an identifiability constraint (Wood, 2017). This has
 the effect of making the basis orthogonal to the intercept but results
@@
 in Figure \ref{fig:basis-function-example-plot}b. The solid line passing
-through the data points is formed by summing up the values of the scaled
-basis functions (\(b_j(x_t) \beta_j\)) at any value of \(x_t\) (time).
+through the data points is formed by summing up the values of the seven
+scaled basis functions (\(b_j(x_t) \beta_j\)) at any value of \(x_t\)
+(time).
 
@@
 to put an upper limit on the expected complexity of the true trend in
-the data. Additionally, the number available data points places a
+the data. Additionally, the number of available data points places a
 constraint on the upper limit of the size of basis expansion that can be
@@
 In practice, the size of the basis is an upper limit on the expected
-complexity of the trend, and a simple test to check if the basis used
-was sufficiently large (Pya and Wood, 2016). This test is available via
-the \texttt{gam.check()} function in \emph{mgcv} for example, which
-essentially looks at whether there is any additional nonlinearity or
-structure in the residuals that can be explained by a further smooth of
-\(x_t\). Should a smooth term in the fitted model fail this test the
-model can be refitted using a larger basis expansion, say by doubling
-the value of \emph{k} (the rank) used to fit the original. Note also
-that a smooth might fail this test whilst using fewer effective degrees
-of freedom than the maximum possible for the dimension of basis used.
-This may happen when the true function lies at the upper limit of the
-set of functions encompassed by the size of basis used. Additionally, a
-basis of size \(2k\) encompasses a richer space of functions of a given
-complexity than a basis of size \(k\) (Wood, 2017); increasing the basis
-dimension used to fit the model may unlock this additional function
-space resulting in a better fitting model whilst using a similar number
-of effective degrees of freedom.
+complexity of the trend, and a simple test can be used to check if the
+basis used was sufficiently large (Pya and Wood, 2016). This test is
+available via the \texttt{gam.check()} function in \emph{mgcv} for
+example, which essentially looks at whether there is any additional
+nonlinearity or structure in the residuals that can be explained by a
+further smooth of \(x_t\). Should a smooth term in the fitted model fail
+this test the model can be refitted using a larger basis expansion, say
+by doubling the value of \emph{k} (the rank) used to fit the original.
+Note also that a smooth might fail this test whilst using fewer
+effective degrees of freedom than the maximum possible for the dimension
+of basis used. This may happen when the true function lies at the upper
+limit of the set of functions encompassed by the size of basis used.
+Additionally, a basis of size \(2k\) encompasses a richer space of
+functions of a given complexity than a basis of size \(k\) (Wood, 2017);
+increasing the basis dimension used to fit the model may unlock this
+additional function space resulting in a better fitting model whilst
+using a similar number of effective degrees of freedom.
 
@@
 increasing the basis size, it was the failure to appreciate the
-magnitude of the effect on fitting of the non-constant variance that
-lead to the initially poor fit and the problems associated with the
-estimation of the CAR(1) process. I return to the issue of why the GAM
-plus CAR(1) model encountered problems during fitting later
-(\protect\hyperlink{identifiability}{see Identifiability}).
+magnitude of the effect of the non-constant variance that lead to the
+initially poor fit and the problems associated with the estimation of
+the CAR(1) process. I return to the issue of why the GAM plus CAR(1)
+model encountered problems during fitting later
+(\protect\hyperlink{identifiability}{see section Residual
+autocorrelation and model identification}).
 
@@
 If we want to ask whether either of the estimated trends is
-statistically interesting or proceed to identifying periods of
-significant change, we must address the issue of uncertainty in the
-estimated model. What uncertainty is associated with the trend
-estimates? One way to visualise this is through a 1 - \(\alpha\)
-confidence interval around the fitted trend, where \(\alpha\) is say
-0.05 leading to a 95\% interval. A 95\% interval would be drawn at
+statistically interesting or proceed to identify periods of significant
+change, we must address the issue of uncertainty in the estimated model.
+What uncertainty is associated with the trend estimates? One way to
+visualise this is through a 1 - \(\alpha\) confidence interval around
+the fitted trend, where \(\alpha\) is say 0.05 leading to a 95\%
+interval. A 95\% interval would be drawn at
 \(\hat{y}_t \pm (m_{1-\alpha} \times \text{SE}(\hat{y}_t))\), with
 \(m_{1-\alpha} = 1.96\), the 0.95 probability quantile of a standard
-normal distribution, and \(\text{SE}(\hat{y}_t)\) is the standard error
-of the estimated trend at time \(x_t\). This type of confidence interval
+normal distribution\footnote{The 0.95 probability quantile of the
+  \emph{t} distribution may be used instead, which will account for
+  estimation of \(\sigma\), the variance of the data. However, given the
+  number of observations, and hence residual degrees of freedom, needed
+  to motivate fitting GAMs, differences between intervals computed using
+  extreme quantiles of the standard normal or the \emph{t} distribution
+  will be tiny.}, and \(\text{SE}(\hat{y}_t)\) is the standard error of
+the estimated trend at time \(x_t\). This type of confidence interval
 would normally be described as \emph{pointwise}; the coverage properties
@@
 Figure \ref{fig:posterior-simulation} shows 20 random draws from the
-posterior distribution of the GAMs fitted to the Small Water and
+posterior distributions of the GAMs fitted to the Small Water and
 Braya-Sø data sets. In the early period of the δ\textsuperscript{15}N
@@
 critical value \(m_{1-\alpha}\) for the simultaneous interval of the
-estimated trend in δ\textsuperscript{15}N is 3.08, whilst the same value
-for the \uk{} series is 3.42, leading to intervals that are
+estimated trend in δ\textsuperscript{15}N is 3.07, whilst the same value
+for the \uk{} series is 3.43, leading to intervals that are
 approximately ±50\% and ±75\% wider than the equivalent
@@
 \(\phi\) are two ways to describe the same pattern of values in a time
-series, and without any further information to constrain either the
-model is unable to distinguish both components uniquely.
+series, and, without any further information to constrain either of
+these, the model is unable to distinguish both components uniquely.
 
@@
 distribution over all possible smooth functions \(f(x)\). In the field
-of spatial statistics, Gaussian processes are known by name
+of spatial statistics, Gaussian processes are known by the name
 \emph{kriging}.
@@
 
-Each of the spline types that I have discussed so far share a common
+Each of the spline types that I have discussed so far shares a common
 feature; the degree of wiggliness over the time series is fixed due to
@@
 to be too-strong a statement; those oscillations in \uk{} may be real
-responses to climate forcing but may simply lack the statistical power
-to distinguish them from the null hypothesis of no trend through this
-period. The adaptive smoother is only adjusting to the data available to
-it; just because it does not detect a trend during this period does not
-lend itself to an interpretation of stable climate forcing or
-complacency in the lake's response to forcing. If there were particular
+responses to climate forcing but we may simply lack the statistical
+power to distinguish them from the null hypothesis of no trend through
+this period. The adaptive smoother is only adjusting to the data
+available to it; just because it does not detect a trend during this
+period does not lend itself to an interpretation of stable climate
+forcing or complacency in the lake's response to forcing (although that
+is a justifiable interpretation of the result). If there were particular
 interest in the climate of this particular period we might take from the
-Braya-Sø record that there is potential early variation in climate
+Braya-Sø record that there is potential early variation due to climate
 forcing, but that additional data from this or other sites is required
@@
 that the time covariate (\(x_t\)) was fixed and known. In both examples,
-and more generally for most palaeoecological records, this assumption is
+and generally for most palaeoecological records, this assumption is
 violated. Unless the record is annually laminated, assigning an age to a
@@
 difficult to fit classical statistical time series models like ARIMA.
-This has lead palaeoecologists and palaeolimnologists to fall back on
-basic statistical methods such as linear parametric and non-parametric
-correlations or simple linear regression models, where the assumptions
-of the method are often grossly violated by the dependencies inherent to
-time series data. GAMs, whilst similar to the popular \loess{} smoother,
-provide a superior alternative approach to trend estimation in
-palaeoenvironmental time series. GAMs can estimate non-linear trends,
-provide estimates of the magnitude of change as well as allow the
-identification of periods of change, can account for the lack of
-independence (either via autocorrelation processes or via the fitting of
-a wiggly trend), and provide a formal framework for statistical
-inference on each of these features.
+This has led palaeoecologists and palaeolimnologists to either ignore
+statistical estimation of trends or fall back on basic statistical
+methods such as linear parametric and non-parametric correlations or
+simple linear regression models, where the assumptions of the method are
+often grossly violated by the dependencies inherent to time series data.
+GAMs, whilst similar to the popular \loess{} smoother, provide a
+superior alternative approach to trend estimation in palaeoenvironmental
+time series. GAMs can estimate non-linear trends, provide estimates of
+the magnitude of change as well as allow the identification of periods
+of change, can account for the lack of independence (either via
+autocorrelation processes or via the fitting of a wiggly trend), and
+provide a formal framework for statistical inference on each of these
+features.
 
@@
 The ideas expressed in this paper are the result of many fruitful
-conversations with colleagues at the Environmental Change Research
-Centre, UCL, and the University of Regina. In particular I am indebted
-to Helen Bennion, Rick Battarbee, and Peter Leavitt for their
-collaborations on projects over many years, and to David Miller, Eric
-Pedersen, and Noam Ross, my GAM workshop partners in crime. Without
+conversations with colleagues past and present at the Environmental
+Change Research Centre, UCL, and the University of Regina. In particular
+I am indebted to Helen Bennion, Rick Battarbee, and Peter Leavitt for
+their collaborations on projects over many years, and to David Miller,
+Eric Pedersen, and Noam Ross, my GAM workshop partners in crime. Without
 Simon Wood's \emph{mgcv} software and his research on GAMs, the

### 2018-08-16 supplementary-materials.Rmd
mass rebuild after switching to gratia package

@@
 ```
-In addition, the *schoenberg* package is required; this in-development package is not on CRAN but can be installed directly from GitHub using functions from the *devtools* package.
-To install the package, install *devtools* and then use `devtools::install_github()` to install *schoenberg*, as shown below:
+In addition, the *gratia* package is required; this in-development package is not on CRAN but can be installed directly from GitHub using functions from the *devtools* package.
+To install the package, install *devtools* and then use `devtools::install_github()` to install *gratia*, as shown below:
 
-```{r install-schoenberg, eval = FALSE}
-## schoenberg is not on CRAN, install from github:
+```{r install-gratia, eval = FALSE}
+## gratia is not on CRAN, install from github:
 install.packages("devtools")
-devtools::install_github("gavinsimpson/schoenberg")
+devtools::install_github("gavinsimpson/gratia")
 ```
-Assuming that the installation of *schoenberg* completes without error, the package can be loaded as usual
-```{r load-schoenberg, cache = FALSE}
-library("schoenberg") # need to change the name of the package
+Assuming that the installation of *gratia* completes without error, the package can be loaded as usual
+```{r load-gratia, cache = FALSE}
+library("gratia") # need to change the name of the package
 ```
@@
 
-## Generate a plot of the data
+## plot Braya-So data
 braya_plt <- ggplot(braya, aes(x = Year, y = UK37)) +
@@
 
-The CAR(1) process plotted in Figure 10 of the manuscript was prepared using
+The CAR(1) process plotted in Figure 11 of the manuscript was prepared using
 
@@
 
-The same model is fitted to the Braya-Sø data set.
-Note however that in order to even fit the model with both a smooth and the CAR(1) process, I have had to change the default optimiser used to estimate the model parameters, and reduce the basis dimension to a small number.
-We also fit the model using GCV, which is the defaut, hence no `method` argument
+The same model as used for Small Water was also initially attempted with the Braya-Sø data.
+However, in order to even fit the model with both a smooth and the CAR(1) process, I had to change the default optimizer used to estimate the model parameters, and reduce the basis dimension, `k`, to a small number.
+I also fit the model using GCV, which is, at the time of writing, the default in `gam()`, hence no `method` argument
 
@@
 
-For Braya-Sø, we repeat the process, but we do so for both models (GAMM + CAR(1) and GCV), and use a critical value from the $t$ distribution to form the confidence interval
+To plot the GAMs fitted to the Braya-Sø time series, we repeat the process used with the Small Water GAM, but we do this for both models (GAMM + CAR(1) and GCV), using a critical value from the $t$ distribution to form the confidence interval
 
@@
 
-Figure 5 in the manuscript was produced using:
+Figure 6 in the manuscript was produced using:
 
-```{r manuscript-fig-2, fig.show = "hide"}
+```{r manuscript-fig-6, fig.show = "hide"}
 plot_grid(small_fitted, braya_fitted, ncol = 1, labels = "auto",
@@
 
-Clearly we are not able to identify both the wiggly trend and the CAR(1) process in a single model.
-We might proceed by fitting a simple GAM via `gam()` with REML smoothness selection and moderate number of basis functions --- here we set `k = 15` for illustration.
+For the Braya-Sø data, clearly we are not able to identify both the wiggly trend and the CAR(1) process in a single model.
+We proceed by fitting a simple GAM via `gam()` with REML smoothness selection and a moderate number of basis functions --- here we set `k = 15` for illustration.
 
@@
 
-In the output, the column labelled `k'` is the size of the basis used (the number of basis functions) --- note that it is 14 and not the requested 15 functions as the constant basis function has been removed as it is confounded with the model constant term, the intercept.
-The column labelled `edf` indicates the *effective degrees of freedom* for the estimated smooth.
-The `k-index` column contains the value of a test statistic for a test of sufficient number of basis functions --- ideally, we'd like the value of `k-index` to be close to or greater than 1.
+The first few lines of output from `gam.check()` include information on the model fit, whether the algorithm converged, and some diagnostics from the optimization at convergence --- you are unlikely to need this information, but it can be useful in diagnosing problems with model fitting or when reporting problems to Simon Wood, the developer of the *mgcv* package.
+In the table at the bottom of the output from `gam.check()`, the column labelled `k'` is the size of the basis used to fit the model (the number of basis functions).
+Note that the stated value is 14 and not the requested 15 functions because the constant basis function has been removed due to it being confounded with the model constant term, the intercept.
+The column labelled `edf` indicates the *effective degrees of freedom* for the estimated smooth, which is a measure of the complexity or wiggliness of the final smooth.
+The `k-index` column contains the value of a test statistic for a test of a sufficient number of basis functions --- ideally, we'd like the value of `k-index` to be close to or greater than 1.
 The quoted *p* value measures the support for the null hypothesis that enough basis functions were available.
-In practial terms, the low `k-index` and very low *p* value suggest that the requested number of basis functions used to fit the model was too low.
+In practial terms, the low `k-index` and very low *p* value shown above strongly suggest that the requested number of basis functions used to fit the model was too low.
 
 The solution is to increase `k`, say by doubling the original value, and refit the model and perform the basis dimension check using `gam.check()`, repeating this process until `k-index` is close to or greater than 1 and the *p* value is larger than a desired threshold.
-In my experience, it may not be possible with some data sets to increase `k` large enough to result in a `k-index` > 1 and a *p* value > 0.05, given the available.
+In my experience fitting GAMs to palaeoenvironmental time series, it may not be possible with some data sets to increase `k` large enough to result in a `k-index` > 1 and a *p* value > 0.05, given the available number of samples.
 It is important to remember that the test provided by `gam.check()` is only an heuristic one and shouldn't be seen as infallible.
@@
 
-To proceed with the Braya-Sø example, we need to increase the basis dimension (`k = 45`), fit using `method = "REML"`.
+To proceed with the Braya-Sø example, we need to increase the basis dimension, fit using `method = "REML"`.
 As discussed in the manuscript, we should also account for the non-constant variance, or *heteroscedasticity*, that may arise due to variation in the amount of time (or "lake years") that each sample represents.
-All else equal, we expect that samples that average more time have smaller variance than those that average a smaller amount of time.
+All else equal, we expect that samples that average more time have lower variance than those that average a smaller amount of time.
 To include information about the expected heteroscedasticity in the GAM we can use observational weights[^2].
-Here I use the `sampleInterval` variable that I created earlier as the measure of lake years per sample, and to avoid changing the model likelihood, the weights use are the values of `sampleInterval` divided by the mean of `sampleInterval`:
+Here, I use the `sampleInterval` variable that I created earlier as the measure of lake years per sample, and, to avoid changing the model likelihood, the weights use are the values of `sampleInterval` divided by the mean of `sampleInterval`:
 
 ```{r final-braya-reml-fit}
-braya_reml <- gam(UK37 ~ s(Year, k = 45), data = braya,
+braya_reml <- gam(UK37 ~ s(Year, k = 40), data = braya,
                   method = "REML",
@@
 
-If we now check if 44 (`k = 45`) basis functions is sufficient
+If we now check if 39 (`k = 40`) basis functions is sufficient
 ```{r final-braya-gam-check, fig.show = "hide"}
@@
 we note that the value of `k-index` is now greater than 1 and the *p* value suggests there is very little evidence against the null hypothesis.
+This result strongly suggests that there were sufficient basis functions used to estimated the trend.
+
 Also note that the `edf` of the estimated smooth is considerably below the maximum (`k'`).
-This is the effect of the smoothness penalty removing the excessive wiggliness possible with 44 basis functions.
+This is the effect of the smoothness penalty removing the excessive wiggliness possible with 39 basis functions.
 If the `edf` were close to `k'`, you might consider increasing `k` by a modest amount (in this instance by 5 or 10 additional basis functions) to assure yourself that you have sufficient basis functions.
-This has the additional advantage of allowing a richer set of smooths of a given complexity (`edf`) to be represented using the basis functions, which may improve the model fit to the data without changing the `edf` too much.
+This has the additional advantage of allowing a richer set of smooths of a given complexity (`edf`) to be represented using this larger set of basis functions, which may improve the model fit to the data without changing the `edf` too much.
 
@@
 
-Samples from the posterior distribution of a GAM can be drawn using the `simulate()` methods from the *schoenberg* package.
+Samples from the posterior distribution of a GAM can be drawn using the `simulate()` methods from the *gratia* package.
 
@@
 ```{r braya-posterior-simulation}
-## posterior simulation
-## need to reset-up newBraya
+## data points to simulate at
 newBraya <- with(braya,
@@
                   unconditional = TRUE)
+
+## rearrange the output into a long/tidy format
 colnames(sims2) <- paste0("sim", seq_len(nsim))
@@
 
-Figure 7 in the manuscript was prepared using
+Figure 8 in the manuscript was prepared using
 
-```{r figure-7, fig.show = "hide"}
+```{r figure-8, fig.show = "hide"}
 plot_grid(smallSim.plt, brayaSim.plt, ncol = 1, labels = "auto",
@@
 The type of interval required is given via the `type` argument with options `"confidence"` and `"simultaneous"`.
-Fro example, for Small Water we would use
+For example, for Small Water we would use
 
@@
 
-Figure 8 in the manuscript was prepared using
-```{r figure-8, fig.show = "hide"}
+Figure 9 in the manuscript was prepared using
+```{r figure-9, fig.show = "hide"}
 plot_grid(smallInt.plt, brayaInt.plt, ncol = 1, labels = "auto",
@@
 There is also a `confint()` method for objects produced by `fderiv()`.
-The first derivatives and a 95% simultaneous confidence interval for the Small Water trend we computed and plotted using
+The first derivatives and a 95% simultaneous confidence interval for the Small Water trend were computed and plotted using
 
@@
 
-Figure 9 in the manuscript was prepared using
+Figure 10 in the manuscript was prepared using
 
-```{r figure 9}
+```{r figure-10}
 plot_grid(small_deriv_plt, braya_deriv_plt, ncol = 1, labels = "auto",
@@
 In the chunk below I fit 200 models with $\phi$ in the range 15--500.
-For value of $\phi$ I fit the required GAM and extract the REML score (from component `gcv.ubre`) and store it in the numeric vectors `Mat` or `SEx`, for Matérn and Squared Exponential correlation functions, respectively.
-The final line of the chunk prepares the REML scores for the two correlation functions in long or tidy format suitable for plotting with *ggplot2*.
+For each value of $\phi$ I fit the required GAM and extract the REML score (from component `gcv.ubre`) and store it in the numeric vectors `Mat` or `SEx`, for Matérn and Squared Exponential correlation functions, respectively.
+The final line of the chunk prepares the REML scores in long or tidy format suitable for plotting with *ggplot2*.
 
@@
 
-whilst Figure 12 in the manuscript was prepared using
+whilst Figure 13 in the manuscript was prepared using
 
-```{r figure-12, fig.show = "hide"}
+```{r figure-13, fig.show = "hide"}
 plot_grid(proflik.plt, gp.plt2, ncol = 1, labels = c("a","b"),
@@
 
-For the model comparison, I refitted all the models for consistency; the code to fit each of the
+For the model comparison I refitted all the models for consistency; the code to fit each of the
 
@@
 
-We write a small function to predict from each model over the range of `Year` and return the data in tidy format for plotting. The plot produce reproduces figure 13 in the manuscript.
+I wrote a small function to predict from each model over the range of `Year` and return the data in tidy format for plotting.
 
@@
                                    `[[`, "objects"))
-pltData <- transform(pltData, Model = rep(c("GP", "Adaptive", "TPRS"),
+pltData <- transform(pltData,
+                     Model = rep(c("GP", "Adaptive", "TPRS"),
                                  each = nrow(plts_gp$objects)))
@@
 ```
+The plot produced reproduces Figure 14 in the manuscript.
 
@@
 
-To make this section self-contained, I refitted the Small Water GAM plus CAR(1) model
+To make this section self-contained, I refitted the Small Water GAM plus CAR(1) model.
+Because we will be sampling ages for each observation from the posterior of the age-model GAM, the oldest sample will will be somewhat younger than the expected value from the model in some samples.
+This will cause problems when we come to simulate from the GAMs fitted to the resampled age-models as each GAM will cover a slightly different range of time.
+To avoid this problem I fix the knots at the extremes of the observed values of `Year` and spread the remaining 12 knots evenly inbetween.
+The knot locations are stored in the vector `knots`, which is passed to the argument `knots` when fitting the GAM
 
@@
 
-and then we load the ^210^Pb dating results for the dated core sections.
+Setting the knots like this doesn't change the model estimated above; I'm only repeating what happens internally if you don't supply `knots`.
+However, later we will need to specify this set of knots explicitly when accounting for age model uncertainty.
+
+Next we load the ^210^Pb dating results for the dated core sections.
 
@@
 The next chunk does the final step in the process.
-For each of the models we just fitted to include age model uncertainty, we simulate 50 draws from the model posterior distribution.
+For each of the models we just fitted, we simulate 50 draws from the model posterior distribution.
 We start with a wrapper function around the `simulate()` code we want to run on each model, then do the actual posterior draws for each model using `lapply()`.
@@
 ```{r small-scam-fit-plots, dependson = -1}
+## plot the estimated age model ad posterior simulations from it
 plt1 <- ggplot(swAge, aes(y = Date, x = Depth)) +
@@
 
+## plot the simulated trends showing the effect of age-model uncertainty
 plt2 <- ggplot(simTrends, aes(x = Year, y = Trend, group = Group)) +
@@
 
+## plot simulated trends showing the effect of age-model uncertainty and
+## the effect of uncertainty in the estimated trend itself
 plt3 <- ggplot(simSimulate, aes(x = Year, y = Trend, group = Group)) +
@@
 
-This reproduces figure 14 from the manuscript.
+This reproduces Figure 15 from the manuscript.
 

### 2018-08-16 supplementary-materials.tex
bump date
mass rebuild after switching to gratia package

@@
   \postdate{\par}
-    \date{August 14, 2018}
+    \date{August 16, 2018}
 
@@
 
-In addition, the \emph{schoenberg} package is required; this
-in-development package is not on CRAN but can be installed directly from
-GitHub using functions from the \emph{devtools} package. To install the
-package, install \emph{devtools} and then use
-\texttt{devtools::install\_github()} to install \emph{schoenberg}, as
-shown below:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## schoenberg is not on CRAN, install from github:}
+In addition, the \emph{gratia} package is required; this in-development
+package is not on CRAN but can be installed directly from GitHub using
+functions from the \emph{devtools} package. To install the package,
+install \emph{devtools} and then use
+\texttt{devtools::install\_github()} to install \emph{gratia}, as shown
+below:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{## gratia is not on CRAN, install from github:}
 \KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
-\NormalTok{devtools::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"gavinsimpson/schoenberg"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Assuming that the installation of \emph{schoenberg} completes without
-error, the package can be loaded as usual
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(}\StringTok{"schoenberg"}\NormalTok{) }\CommentTok{# need to change the name of the package}
+\NormalTok{devtools::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"gavinsimpson/gratia"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Assuming that the installation of \emph{gratia} completes without error,
+the package can be loaded as usual
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(}\StringTok{"gratia"}\NormalTok{) }\CommentTok{# need to change the name of the package}
 \end{Highlighting}
@@
 
-\NormalTok{## Generate a plot of the data}
+\NormalTok{## plot Braya-So data}
 \NormalTok{braya_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37)) +}
@@
 #>        lower      est.     upper
-#> Phi 0.2811107 0.6026967 0.8547543
+#> Phi 0.282388 0.6026967 0.8539689
 #> attr(,"label")
@@
 
-The CAR(1) process plotted in Figure 10 of the manuscript was prepared
+The CAR(1) process plotted in Figure 11 of the manuscript was prepared
 using
@@
 
-The same model is fitted to the Braya-Sø data set. Note however that in
-order to even fit the model with both a smooth and the CAR(1) process, I
-have had to change the default optimiser used to estimate the model
-parameters, and reduce the basis dimension to a small number. We also
-fit the model using GCV, which is the defaut, hence no \texttt{method}
-argument
+The same model as used for Small Water was also initially attempted with
+the Braya-Sø data. However, in order to even fit the model with both a
+smooth and the CAR(1) process, I had to change the default optimizer
+used to estimate the model parameters, and reduce the basis dimension,
+\texttt{k}, to a small number. I also fit the model using GCV, which is,
+at the time of writing, the default in \texttt{gam()}, hence no
+\texttt{method} argument
 
@@
 #>            lower      est. upper
-#> Phi 1.611049e-22 0.2000156     1
+#> Phi 2.680033e-31 0.2000156     1
 #> attr(,"label")
@@
 
-For Braya-Sø, we repeat the process, but we do so for both models (GAMM
-+ CAR(1) and GCV), and use a critical value from the \(t\) distribution
-to form the confidence interval
+To plot the GAMs fitted to the Braya-Sø time series, we repeat the
+process used with the Small Water GAM, but we do this for both models
+(GAMM + CAR(1) and GCV), using a critical value from the \(t\)
+distribution to form the confidence interval
 
@@
 
-Figure 5 in the manuscript was produced using:
+Figure 6 in the manuscript was produced using:
 
@@
 
-Clearly we are not able to identify both the wiggly trend and the CAR(1)
-process in a single model. We might proceed by fitting a simple GAM via
-\texttt{gam()} with REML smoothness selection and moderate number of
-basis functions --- here we set \texttt{k\ =\ 15} for illustration.
+For the Braya-Sø data, clearly we are not able to identify both the
+wiggly trend and the CAR(1) process in a single model. We proceed by
+fitting a simple GAM via \texttt{gam()} with REML smoothness selection
+and a moderate number of basis functions --- here we set
+\texttt{k\ =\ 15} for illustration.
 
@@
 #> full convergence after 6 iterations.
-#> Gradient range [-4.468978e-08,8.696688e-10]
+#> Gradient range [-4.468978e-08,8.696768e-10]
 #> (score -187.2524 & scale 0.000689802).
@@
 
-In the output, the column labelled \texttt{k\textquotesingle{}} is the
-size of the basis used (the number of basis functions) --- note that it
-is 14 and not the requested 15 functions as the constant basis function
-has been removed as it is confounded with the model constant term, the
-intercept. The column labelled \texttt{edf} indicates the
-\emph{effective degrees of freedom} for the estimated smooth. The
-\texttt{k-index} column contains the value of a test statistic for a
-test of sufficient number of basis functions --- ideally, we'd like the
-value of \texttt{k-index} to be close to or greater than 1. The quoted
-\emph{p} value measures the support for the null hypothesis that enough
-basis functions were available. In practial terms, the low
-\texttt{k-index} and very low \emph{p} value suggest that the requested
-number of basis functions used to fit the model was too low.
+The first few lines of output from \texttt{gam.check()} include
+information on the model fit, whether the algorithm converged, and some
+diagnostics from the optimization at convergence --- you are unlikely to
+need this information, but it can be useful in diagnosing problems with
+model fitting or when reporting problems to Simon Wood, the developer of
+the \emph{mgcv} package. In the table at the bottom of the output from
+\texttt{gam.check()}, the column labelled \texttt{k\textquotesingle{}}
+is the size of the basis used to fit the model (the number of basis
+functions). Note that the stated value is 14 and not the requested 15
+functions because the constant basis function has been removed due to it
+being confounded with the model constant term, the intercept. The column
+labelled \texttt{edf} indicates the \emph{effective degrees of freedom}
+for the estimated smooth, which is a measure of the complexity or
+wiggliness of the final smooth. The \texttt{k-index} column contains the
+value of a test statistic for a test of a sufficient number of basis
+functions --- ideally, we'd like the value of \texttt{k-index} to be
+close to or greater than 1. The quoted \emph{p} value measures the
+support for the null hypothesis that enough basis functions were
+available. In practial terms, the low \texttt{k-index} and very low
+\emph{p} value shown above strongly suggest that the requested number of
+basis functions used to fit the model was too low.
 
@@
 close to or greater than 1 and the \emph{p} value is larger than a
-desired threshold. In my experience, it may not be possible with some
-data sets to increase \texttt{k} large enough to result in a
-\texttt{k-index} \textgreater{} 1 and a \emph{p} value \textgreater{}
-0.05, given the available. It is important to remember that the test
-provided by \texttt{gam.check()} is only an heuristic one and shouldn't
-be seen as infallible. If you find yourself increasing \texttt{k} to
-large values relative to the number of samples in your data set without
-achieving a \texttt{k-index} \textgreater{} 1 or \emph{p} \textgreater{}
-0.05, then note the value in the \texttt{edf} column as you increase
-\texttt{k}. If the \texttt{edf} of the model changes little as you
-continue to increase \texttt{k}, this may be evidence that the test is
-failing in this particular instance and that the estimated smooth is not
-dependent on the value of \texttt{k} chosen.
+desired threshold. In my experience fitting GAMs to palaeoenvironmental
+time series, it may not be possible with some data sets to increase
+\texttt{k} large enough to result in a \texttt{k-index} \textgreater{} 1
+and a \emph{p} value \textgreater{} 0.05, given the available number of
+samples. It is important to remember that the test provided by
+\texttt{gam.check()} is only an heuristic one and shouldn't be seen as
+infallible. If you find yourself increasing \texttt{k} to large values
+relative to the number of samples in your data set without achieving a
+\texttt{k-index} \textgreater{} 1 or \emph{p} \textgreater{} 0.05, then
+note the value in the \texttt{edf} column as you increase \texttt{k}. If
+the \texttt{edf} of the model changes little as you continue to increase
+\texttt{k}, this may be evidence that the test is failing in this
+particular instance and that the estimated smooth is not dependent on
+the value of \texttt{k} chosen.
 
@@
 To proceed with the Braya-Sø example, we need to increase the basis
-dimension (\texttt{k\ =\ 45}), fit using \texttt{method\ =\ "REML"}. As
-discussed in the manuscript, we should also account for the non-constant
-variance, or \emph{heteroscedasticity}, that may arise due to variation
-in the amount of time (or ``lake years'') that each sample represents.
-All else equal, we expect that samples that average more time have
-smaller variance than those that average a smaller amount of time. To
-include information about the expected heteroscedasticity in the GAM we
-can use observational weights\footnote{This simple solution is for a
-  Gaussian GAM. For models with other distributions, the exact
-  interpretation of observational weights will vary. For such data,
-  where the conditional mean and variance of the observations are
-  linked, the use of location-scale, or distributional, models may be
-  more apporpriate and informative.}. Here I use the
-\texttt{sampleInterval} variable that I created earlier as the measure
-of lake years per sample, and to avoid changing the model likelihood,
-the weights use are the values of \texttt{sampleInterval} divided by the
-mean of \texttt{sampleInterval}:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
+dimension, fit using \texttt{method\ =\ "REML"}. As discussed in the
+manuscript, we should also account for the non-constant variance, or
+\emph{heteroscedasticity}, that may arise due to variation in the amount
+of time (or ``lake years'') that each sample represents. All else equal,
+we expect that samples that average more time have lower variance than
+those that average a smaller amount of time. To include information
+about the expected heteroscedasticity in the GAM we can use
+observational weights\footnote{This simple solution is for a Gaussian
+  GAM. For models with other distributions, the exact interpretation of
+  observational weights will vary. For such data, where the conditional
+  mean and variance of the observations are linked, the use of
+  location-scale, or distributional, models may be more apporpriate and
+  informative.}. Here, I use the \texttt{sampleInterval} variable that I
+created earlier as the measure of lake years per sample, and, to avoid
+changing the model likelihood, the weights use are the values of
+\texttt{sampleInterval} divided by the mean of \texttt{sampleInterval}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{40}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
                   \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
@@
 #> Formula:
-#> UK37 ~ s(Year, k = 45)
+#> UK37 ~ s(Year, k = 40)
 #> 
@@
 #>              Estimate Std. Error t value Pr(>|t|)    
-#> (Intercept) -0.633734   0.001901  -333.3   <2e-16 ***
+#> (Intercept) -0.633741   0.001929  -328.5   <2e-16 ***
 #> ---
@@
 #>           edf Ref.df     F p-value    
-#> s(Year) 28.49  33.99 7.755  <2e-16 ***
+#> s(Year) 27.45  32.36 7.782  <2e-16 ***
 #> ---
@@
 #> 
-#> R-sq.(adj) =  0.751   Deviance explained = 83.2%
-#> -REML = -175.68  Scale est. = 0.00024036  n = 89
-\end{verbatim}
-
-If we now check if 44 (\texttt{k\ =\ 45}) basis functions is sufficient
+#> R-sq.(adj) =  0.744   Deviance explained = 82.4%
+#> -REML = -175.61  Scale est. = 0.00024757  n = 89
+\end{verbatim}
+
+If we now check if 39 (\texttt{k\ =\ 40}) basis functions is sufficient
 
@@
 #> Method: REML   Optimizer: outer newton
-#> full convergence after 7 iterations.
-#> Gradient range [-2.210576e-06,-5.181457e-08]
-#> (score -175.684 & scale 0.0002403604).
-#> Hessian positive definite, eigenvalue range [2.524305,48.11146].
-#> Model rank =  45 / 45 
+#> full convergence after 9 iterations.
+#> Gradient range [-1.105422e-06,1.500959e-07]
+#> (score -175.6093 & scale 0.0002475669).
+#> Hessian positive definite, eigenvalue range [2.644434,47.78257].
+#> Model rank =  40 / 40 
 #> 
@@
 #>           k'  edf k-index p-value
-#> s(Year) 44.0 28.5    1.27    0.99
+#> s(Year) 39.0 27.5    1.27    0.99
 \end{verbatim}
@@
 \emph{p} value suggests there is very little evidence against the null
-hypothesis. Also note that the \texttt{edf} of the estimated smooth is
-considerably below the maximum (\texttt{k\textquotesingle{}}). This is
-the effect of the smoothness penalty removing the excessive wiggliness
-possible with 44 basis functions. If the \texttt{edf} were close to
+hypothesis. This result strongly suggests that there were sufficient
+basis functions used to estimated the trend.
+
+Also note that the \texttt{edf} of the estimated smooth is considerably
+below the maximum (\texttt{k\textquotesingle{}}). This is the effect of
+the smoothness penalty removing the excessive wiggliness possible with
+39 basis functions. If the \texttt{edf} were close to
 \texttt{k\textquotesingle{}}, you might consider increasing \texttt{k}
@@
 This has the additional advantage of allowing a richer set of smooths of
-a given complexity (\texttt{edf}) to be represented using the basis
-functions, which may improve the model fit to the data without changing
-the \texttt{edf} too much.
+a given complexity (\texttt{edf}) to be represented using this larger
+set of basis functions, which may improve the model fit to the data
+without changing the \texttt{edf} too much.
 
@@
 Samples from the posterior distribution of a GAM can be drawn using the
-\texttt{simulate()} methods from the \emph{schoenberg} package.
+\texttt{simulate()} methods from the \emph{gratia} package.
 
@@
 \begin{Highlighting}[]
-\NormalTok{## posterior simulation}
-\NormalTok{## need to reset-up newBraya}
+\NormalTok{## data points to simulate at}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya,}
@@
                   \DataTypeTok{unconditional =} \OtherTok{TRUE}\NormalTok{)}
+
+\NormalTok{## rearrange the output into a long/tidy format}
 \KeywordTok{colnames}\NormalTok{(sims2) <-}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"sim"}\NormalTok{, }\KeywordTok{seq_len}\NormalTok{(nsim))}
@@
 
-Figure 7 in the manuscript was prepared using
+Figure 8 in the manuscript was prepared using
 
@@
 given via the \texttt{type} argument with options \texttt{"confidence"}
-and \texttt{"simultaneous"}. Fro example, for Small Water we would use
+and \texttt{"simultaneous"}. For example, for Small Water we would use
 
@@
 #>    smooth     Year      est        se     crit    lower    upper
-#> 1 s(Year) 1846.319 3.787215 0.1616144 3.087688 3.288200 4.286230
-#> 2 s(Year) 1847.128 3.789747 0.1509100 3.087688 3.323784 4.255710
-#> 3 s(Year) 1847.937 3.792304 0.1411439 3.087688 3.356496 4.228112
-#> 4 s(Year) 1848.745 3.794914 0.1324961 3.087688 3.385807 4.204020
-#> 5 s(Year) 1849.554 3.797602 0.1251070 3.087688 3.411310 4.183893
-#> 6 s(Year) 1850.363 3.800394 0.1190552 3.087688 3.432788 4.167999
+#> 1 s(Year) 1846.319 3.787215 0.1616144 3.072403 3.290671 4.283760
+#> 2 s(Year) 1847.128 3.789747 0.1509100 3.072403 3.326090 4.253403
+#> 3 s(Year) 1847.937 3.792304 0.1411439 3.072403 3.358653 4.225955
+#> 4 s(Year) 1848.745 3.794914 0.1324961 3.072403 3.387832 4.201995
+#> 5 s(Year) 1849.554 3.797602 0.1251070 3.072403 3.413222 4.181981
+#> 6 s(Year) 1850.363 3.800394 0.1190552 3.072403 3.434608 4.166180
 \end{verbatim}
@@
 
-Figure 8 in the manuscript was prepared using
+Figure 9 in the manuscript was prepared using
 
@@
 first derivatives and a 95\% simultaneous confidence interval for the
-Small Water trend we computed and plotted using
+Small Water trend were computed and plotted using
 
@@
 
-Figure 9 in the manuscript was prepared using
+Figure 10 in the manuscript was prepared using
 
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure 9-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/figure-10-1} \end{center}
 
@@
 values for \(\phi\), which we do using a loop. In the chunk below I fit
-200 models with \(\phi\) in the range 15--500. For value of \(\phi\) I
-fit the required GAM and extract the REML score (from component
-\texttt{gcv.ubre}) and store it in the numeric vectors \texttt{Mat} or
-\texttt{SEx}, for Matérn and Squared Exponential correlation functions,
-respectively. The final line of the chunk prepares the REML scores for
-the two correlation functions in long or tidy format suitable for
-plotting with \emph{ggplot2}.
+200 models with \(\phi\) in the range 15--500. For each value of
+\(\phi\) I fit the required GAM and extract the REML score (from
+component \texttt{gcv.ubre}) and store it in the numeric vectors
+\texttt{Mat} or \texttt{SEx}, for Matérn and Squared Exponential
+correlation functions, respectively. The final line of the chunk
+prepares the REML scores in long or tidy format suitable for plotting
+with \emph{ggplot2}.
 
@@
 
-whilst Figure 12 in the manuscript was prepared using
+whilst Figure 13 in the manuscript was prepared using
 
@@
 
-For the model comparison, I refitted all the models for consistency; the
+For the model comparison I refitted all the models for consistency; the
 code to fit each of the
@@
 
-We write a small function to predict from each model over the range of
-\texttt{Year} and return the data in tidy format for plotting. The plot
-produce reproduces figure 13 in the manuscript.
+I wrote a small function to predict from each model over the range of
+\texttt{Year} and return the data in tidy format for plotting.
 
@@
                                    \StringTok{`}\DataTypeTok{[[}\StringTok{`}\NormalTok{, }\StringTok{"objects"}\NormalTok{))}
-\NormalTok{pltData <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(pltData, }\DataTypeTok{Model =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GP"}\NormalTok{, }\StringTok{"Adaptive"}\NormalTok{, }\StringTok{"TPRS"}\NormalTok{),}
+\NormalTok{pltData <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(pltData,}
+                     \DataTypeTok{Model =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GP"}\NormalTok{, }\StringTok{"Adaptive"}\NormalTok{, }\StringTok{"TPRS"}\NormalTok{),}
                                  \DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(plts_gp$objects)))}
@@
 
+The plot produced reproduces Figure 14 in the manuscript.
+
 \section{Accounting for age-model
@@
 To make this section self-contained, I refitted the Small Water GAM plus
-CAR(1) model
+CAR(1) model. Because we will be sampling ages for each observation from
+the posterior of the age-model GAM, the oldest sample will will be
+somewhat younger than the expected value from the model in some samples.
+This will cause problems when we come to simulate from the GAMs fitted
+to the resampled age-models as each GAM will cover a slightly different
+range of time. To avoid this problem I fix the knots at the extremes of
+the observed values of \texttt{Year} and spread the remaining 12 knots
+evenly inbetween. The knot locations are stored in the vector
+\texttt{knots}, which is passed to the argument \texttt{knots} when
+fitting the GAM
 
@@
 
-and then we load the \textsuperscript{210}Pb dating results for the
-dated core sections.
+Setting the knots like this doesn't change the model estimated above;
+I'm only repeating what happens internally if you don't supply
+\texttt{knots}. However, later we will need to specify this set of knots
+explicitly when accounting for age model uncertainty.
+
+Next we load the \textsuperscript{210}Pb dating results for the dated
+core sections.
 
@@
 The next chunk does the final step in the process. For each of the
-models we just fitted to include age model uncertainty, we simulate 50
-draws from the model posterior distribution. We start with a wrapper
-function around the \texttt{simulate()} code we want to run on each
-model, then do the actual posterior draws for each model using
-\texttt{lapply()}. The final step just arranges data for plotting.
+models we just fitted, we simulate 50 draws from the model posterior
+distribution. We start with a wrapper function around the
+\texttt{simulate()} code we want to run on each model, then do the
+actual posterior draws for each model using \texttt{lapply()}. The final
+step just arranges data for plotting.
 
@@
 \begin{Highlighting}[]
+\NormalTok{## plot the estimated age model ad posterior simulations from it}
 \NormalTok{plt1 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(swAge, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{Date, }\DataTypeTok{x =} \NormalTok{Depth)) +}
@@
 
+\NormalTok{## plot the simulated trends showing the effect of age-model uncertainty}
 \NormalTok{plt2 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(simTrends, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{Trend, }\DataTypeTok{group =} \NormalTok{Group)) +}
@@
 
+\NormalTok{## plot simulated trends showing the effect of age-model uncertainty and}
+\NormalTok{## the effect of uncertainty in the estimated trend itself}
 \NormalTok{plt3 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(simSimulate, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{Trend, }\DataTypeTok{group =} \NormalTok{Group)) +}
@@
 #>  setting  value                                      
-#>  version  R version 3.4.4 Patched (2018-03-17 r74422)
+#>  version  R version 3.4.3 Patched (2018-01-16 r74126)
 #>  system   x86_64, linux-gnu                          
@@
 #>  tz       America/Regina                             
-#>  date     2018-05-14
+#>  date     2018-08-16
 \end{verbatim}
@@
 #>  package    * version date       source        
+#>  assertthat   0.2.0   2017-04-11 CRAN (R 3.4.3)
 #>  backports    1.1.2   2017-12-13 CRAN (R 3.4.3)
-#>  base       * 3.4.4   2018-03-18 local                                   
-#>  codetools    0.2-15  2016-10-05 CRAN (R 3.4.4)                          
-#>  colorspace   1.3-2   2016-12-14 CRAN (R 3.4.1)                          
-#>  compiler     3.4.4   2018-03-18 local                                   
-#>  cowplot    * 0.9.2   2017-12-17 CRAN (R 3.4.3)                          
-#>  datasets   * 3.4.4   2018-03-18 local                                   
-#>  devtools     1.13.5  2018-02-18 CRAN (R 3.4.4)                          
-#>  digest       0.6.15  2018-01-28 CRAN (R 3.4.4)                          
-#>  evaluate     0.10.1  2017-06-24 CRAN (R 3.4.1)                          
-#>  ggplot2    * 2.2.1   2016-12-30 CRAN (R 3.4.1)                          
-#>  glue         1.2.0   2017-10-29 CRAN (R 3.4.2)                          
-#>  graphics   * 3.4.4   2018-03-18 local                                   
-#>  grDevices  * 3.4.4   2018-03-18 local                                   
-#>  grid         3.4.4   2018-03-18 local                                   
-#>  gtable       0.2.0   2016-02-26 CRAN (R 3.4.1)                          
-#>  htmltools    0.3.6   2017-04-28 CRAN (R 3.4.1)                          
-#>  knitr        1.20    2018-02-20 CRAN (R 3.4.4)                          
-#>  labeling     0.3     2014-08-23 CRAN (R 3.4.1)                          
-#>  lattice      0.20-35 2017-03-25 CRAN (R 3.4.4)                          
-#>  lazyeval     0.2.1   2017-10-29 CRAN (R 3.4.2)                          
-#>  magrittr     1.5     2014-11-22 CRAN (R 3.4.1)                          
-#>  MASS         7.3-49  2018-02-23 CRAN (R 3.4.4)                          
-#>  Matrix       1.2-14  2018-04-09 CRAN (R 3.4.4)                          
-#>  memoise      1.1.0   2017-04-21 CRAN (R 3.4.1)                          
-#>  methods      3.4.4   2018-03-18 local                                   
-#>  mgcv       * 1.8-23  2018-01-21 CRAN (R 3.4.4)                          
-#>  munsell      0.4.3   2016-02-13 CRAN (R 3.4.1)                          
-#>  nlme       * 3.1-137 2018-04-07 CRAN (R 3.4.4)                          
-#>  pillar       1.2.1   2018-02-27 CRAN (R 3.4.4)                          
-#>  plyr         1.8.4   2016-06-08 CRAN (R 3.4.1)                          
-#>  purrr        0.2.4   2017-10-18 CRAN (R 3.4.2)                          
-#>  Rcpp         0.12.16 2018-03-13 CRAN (R 3.4.4)                          
-#>  rlang        0.2.0   2018-02-20 CRAN (R 3.4.4)                          
-#>  rmarkdown  * 1.9     2018-03-01 CRAN (R 3.4.4)                          
+#>  base       * 3.4.3   2018-01-16 local         
+#>  bindr        0.1.1   2018-03-13 CRAN (R 3.4.3)
+#>  bindrcpp     0.2.2   2018-03-29 CRAN (R 3.4.3)
+#>  codetools    0.2-15  2016-10-05 CRAN (R 3.4.3)
+#>  colorspace   1.3-2   2016-12-14 CRAN (R 3.4.3)
+#>  compiler     3.4.3   2018-01-16 local         
+#>  cowplot    * 0.9.3   2018-07-15 CRAN (R 3.4.3)
+#>  crayon       1.3.4   2017-09-16 CRAN (R 3.4.3)
+#>  datasets   * 3.4.3   2018-01-16 local         
+#>  devtools     1.13.6  2018-06-27 CRAN (R 3.4.3)
+#>  digest       0.6.15  2018-01-28 CRAN (R 3.4.3)
+#>  dplyr        0.7.6   2018-06-29 CRAN (R 3.4.3)
+#>  evaluate     0.11    2018-07-17 CRAN (R 3.4.3)
+#>  ggplot2    * 3.0.0   2018-07-03 CRAN (R 3.4.3)
+#>  glue         1.3.0   2018-07-17 CRAN (R 3.4.3)
+#>  graphics   * 3.4.3   2018-01-16 local         
+#>  gratia     * 0.0-6   2018-08-16 local         
+#>  grDevices  * 3.4.3   2018-01-16 local         
+#>  grid         3.4.3   2018-01-16 local         
+#>  gtable       0.2.0   2016-02-26 CRAN (R 3.4.3)
+#>  htmltools    0.3.6   2017-04-28 CRAN (R 3.4.3)
+#>  knitr        1.20    2018-02-20 CRAN (R 3.4.3)
+#>  labeling     0.3     2014-08-23 CRAN (R 3.4.3)
+#>  lattice      0.20-35 2017-03-25 CRAN (R 3.4.3)
+#>  lazyeval     0.2.1   2017-10-29 CRAN (R 3.4.3)
+#>  magrittr     1.5     2014-11-22 CRAN (R 3.4.3)
+#>  MASS         7.3-50  2018-04-30 CRAN (R 3.4.3)
+#>  Matrix       1.2-14  2018-04-09 CRAN (R 3.4.3)
+#>  memoise      1.1.0   2017-04-21 CRAN (R 3.4.3)
+#>  methods      3.4.3   2018-01-16 local         
+#>  mgcv       * 1.8-24  2018-06-18 CRAN (R 3.4.3)
+#>  munsell      0.5.0   2018-06-12 CRAN (R 3.4.3)
+#>  nlme       * 3.1-137 2018-04-07 CRAN (R 3.4.3)
+#>  pillar       1.3.0   2018-07-14 CRAN (R 3.4.3)
+#>  pkgconfig    2.0.1   2017-03-21 CRAN (R 3.4.3)
+#>  plyr         1.8.4   2016-06-08 CRAN (R 3.4.3)
+#>  purrr        0.2.5   2018-05-29 CRAN (R 3.4.3)
+#>  R6           2.2.2   2017-06-17 CRAN (R 3.4.3)
+#>  Rcpp         0.12.18 2018-07-23 CRAN (R 3.4.3)
+#>  rlang        0.2.1   2018-05-30 CRAN (R 3.4.3)
+#>  rmarkdown  * 1.10    2018-06-11 CRAN (R 3.4.3)
 #>  rprojroot    1.3-2   2018-01-03 CRAN (R 3.4.3)
-#>  scales       0.5.0   2017-08-24 CRAN (R 3.4.1)                          
+#>  scales       0.5.0   2017-08-24 CRAN (R 3.4.3)
 #>  scam       * 1.2-2   2017-09-24 CRAN (R 3.4.3)
-#>  schoenberg * 0.0-6   2018-04-12 Github (gavinsimpson/schoenberg@f9b8a84)
-#>  splines      3.4.4   2018-03-18 local                                   
-#>  stats      * 3.4.4   2018-03-18 local                                   
-#>  stringi      1.1.7   2018-03-12 CRAN (R 3.4.4)                          
-#>  stringr      1.3.0   2018-02-19 CRAN (R 3.4.4)                          
+#>  splines      3.4.3   2018-01-16 local         
+#>  stats      * 3.4.3   2018-01-16 local         
+#>  stringi      1.2.4   2018-07-20 CRAN (R 3.4.3)
+#>  stringr      1.3.1   2018-05-10 CRAN (R 3.4.3)
 #>  tibble       1.4.2   2018-01-22 CRAN (R 3.4.3)
-#>  tidyr      * 0.8.0   2018-01-29 CRAN (R 3.4.4)                          
-#>  tools        3.4.4   2018-03-18 local                                   
-#>  utils      * 3.4.4   2018-03-18 local                                   
-#>  withr        2.1.2   2018-03-15 CRAN (R 3.4.4)                          
-#>  yaml         2.1.18  2018-03-08 CRAN (R 3.4.4)
+#>  tidyr      * 0.8.1   2018-05-18 CRAN (R 3.4.3)
+#>  tidyselect   0.2.4   2018-02-26 CRAN (R 3.4.3)
+#>  tools        3.4.3   2018-01-16 local         
+#>  utils      * 3.4.3   2018-01-16 local         
+#>  withr        2.1.2   2018-03-15 CRAN (R 3.4.3)
+#>  yaml         2.2.0   2018-07-25 CRAN (R 3.4.3)
 \end{verbatim}

### 2023-10-16 manuscript.Rmd
update supplementary materials to work with current gratia

@@
 mail: "gavin.simpson@uregina.ca"
-date: August 16, 2018
+date: September 19, 2018
 keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
@@
 
-\caption{Flowchart showing the main steps in the analysis of time series using generalized additive models.}\label{fig:flowchart}
+\caption{Flowchart showing the main steps in the analysis of time series using generalized additive models.
+The main R functions associated with each step or decision are shown in bold.}\label{fig:flowchart}
 \end{figure}
@@
 ```{r data-figure, fig = TRUE, fig.width = 8, fig.height = 5, fig.cap = "Example time series; a) Small Water bulk organic matter $\\delta^{15}\\text{N}$ time series on a ${}^{210}\\text{Pb}$ time scale, and b) Braya-Sø \\uk{} time series on a calibrated ${}^{14}\\text{C}$ time scale.
-The observations \\uk{} time series have been joined by lines purely as a visual aid to highlight potential trends."}
+The observations in the \\uk{} time series have been joined by lines purely as a visual aid to highlight potential trends."}
 ## Generate a plot of the data - used as a base later
@@
 
-In \eqref{eq:gam-ef}, we further assume that the observations are drawn from a member of the exponential family of distributions --- such as the Poisson for count data, the binomial for presence/absence or counts from a total --- with expected value $\mu_t$ and possibly some additional parameters $\Theta$ ($y_t \sim \text{EF}(\mu_t, \Theta)$).
+In \eqref{eq:gam-ef}, we further assume that the observations are drawn from a member of the exponential family of distributions --- such as the Poisson for count data, or the binomial for presence/absence or counts from a total --- with expected value $\mu_t$ and possibly some additional parameters $\Theta$ ($y_t \sim \text{EF}(\mu_t, \Theta)$).
 Additionally, many software implementations of the above model also allow for distributions that are not within the exponential family but which can be fitted using an algorithm superficially similar to the one used to fit GAMs to members of the exponential family [e.g.
@@
 As we have already seen, polynomial basis expansions do not necessarily lead to well-fitting models unless the true function $f$ is itself a polynomial.
-One of the primary criticisms is that polynomial basis functions are global [@Magee1998-oq]; the value of $f$ at time point $x_t$ affects the value of $f$ at time point $x_{t+s}$ even if the two time points are at opposite ends of the series.
+One of the primary criticisms is that polynomial basis functions are global [@Magee1998-oq]; the value of $f$ at time $x_t$ affects the value of $f$ at time point $x_{t+s}$ even if the two time points are at opposite ends of the series.
 There are many other bases we could use; here I discuss one such set of bases, that of splines.
@@
 ```{r basis-function-example-plot, fig.width = 8, fig.height = 5, fig.cap = "Basis functions for the time covariate and the Small Water $\\delta^{15}\\text{N}$ time series.
-A rank (size) 7 cubic regression spline (CRS) basis expansion is show in a), with knots, indicated by tick marks on the x-axis, spread evenly through the rang of the data.
+A rank (size) 7 cubic regression spline (CRS) basis expansion is show in a), with knots, indicated by tick marks on the x-axis, spread evenly through the range of the data.
 b) shows the same CRS basis functions weighted by the estimated coefficients $\\beta_j$, plus the resulting GAM trend line (black line drawn through the data).
@@
 A CRS basis expansion comprised of `r k` basis functions for the time covariate in the Small Water series, is shown in Figure \ref{fig:basis-function-example-plot}a.
-The tick marks on the x-axis show the locations of the knots, which are located at the ends of the series and evenly in between.
+The tick marks on the x-axis show the positions of the knots, which are located at the ends of the series and evenly in between.
 Notice that in this particular parametrization, the $j$th basis function takes a value of 1 at the $j$th knot and at all other knots a value of 0.
@@
 $\phi$ controls how quickly the correlation between any two residuals declines as a function of their separation in time and is an additional parameter that will be estimated during model fitting.
-The model in \eqref{eq:small-gam} was fitted using the `gamm()` function [@Wood2004-zv] in the *mgcv* package [@Wood2017-qi] for R [@R-core-team].
+The model in \eqref{eq:small-gam} was fitted using the `gamm()` function [@Wood2004-zv] in the *mgcv* package [@Wood2017-qi] for R [@Rcore].
 
@@
 The fitted trend is shown in Figure \ref{fig:plot-fitted-models}a, and well-captures the strong pattern in the data.
-The trend is statistically significant (estimated degrees of freedom = `r round(smallSumm[["s.table"]][1,2], 2)`; $F$ = `r round(smallSumm[["s.table"]][1,3], 2)`, approximate $p$ value = $\ll$ 0.0001).
+The trend is statistically significant (effective degrees of freedom = `r round(smallSumm[["s.table"]][1,2], 2)`; $F$ = `r round(smallSumm[["s.table"]][1,3], 2)`, approximate $p$ value = $\ll$ 0.0001).
 However further analysis of the fitted model is required to answer the other questions posed earlier about the timing of change and whether features in the trend can be distinguished from random noise.
@@
 
-The primary issue leading to poor fit is that neither model accounts for the different variance (known as (heteroscedasticity) of each observation in the \uk{} record.
+The primary issue leading to poor fit is that neither model accounts for the different variance (known as heteroscedasticity) of each observation in the \uk{} record.
 This seemingly isn't a problem for GCV which minimizes prediction error.
@@
 However, the GAM described here has a Bayesian interpretation [@Kimeldorf1970-cn; @Wahba1990-ag; @Wahba1983-eh; @Silverman1985-kw] and therefore the typical frequentist interpretation does not apply.
-@Nychka1988-rz investigated the properties of a confidence interval created as described above using standard errors derived from the Bayesian posterior covariance matrix for the estimated mode parameters.
+@Nychka1988-rz investigated the properties of a confidence interval created as described above using standard errors derived from the Bayesian posterior covariance matrix for the estimated model parameters.
 Such intervals have the interesting property that they have good *across-the-function* coverage when considered from a frequentist perspective.
@@
 The general idea is that if we want to create an interval that contains the whole of the true function with 1 - $\alpha$ probability, we need to increase the standard Bayesian credible interval by some amount.
-We could simulate a large number of functions from the posterior distribution of the model and then search for the value of $m_{1-\alpha}$ that when multiplied by $\text{SE}(\hat{f}(x_t))$ yielded an interval that contained the whole function for $(1-\alpha)$ 100% of the functions simulated.
+We could simulate a large number of functions from the posterior distribution of the model and then search for the value of $m_{1-\alpha}$ that when multiplied by $\text{SE}(\hat{f}(x_t))$ yielded an interval that contained the whole function for $(1-\alpha)$100% of the functions simulated.
 In practice, the simulation method of @Ruppert2003-pt does not involve a direct search, but yields the critical value $m_{1-\alpha}$ required.
@@
 In practice, the first derivative of the fitted trend is evaluated using finite differences at a large number of points in the time series.
-An approximate (1 - $\alpha$)100% pointwise confidence interval can be calculated for the derivative estimates using standard theory (i.e.\ $\pm 1.96 \times \text{SE}(\hat{y}_t)$ for a 85% interval) and the covariance matrix of the spline coefficients.
+An approximate (1 - $\alpha$)100% pointwise confidence interval can be calculated for the derivative estimates using standard theory (i.e.\ $\pm 1.96 \times \text{SE}(\hat{y}_t)$ for a 95% interval) and the covariance matrix of the spline coefficients.
 A (1 - $\alpha$)100% simultaneous interval for the derivatives can also be computed using the method described [earlier](#confints).
@@
 The estimated value of $\phi$ for the δ^15^N series is `r round(smallPhi[2], 2)` with 95% confidence interval `r round(smallPhi[1], 2)`--`r round(smallPhi[3], 2)`, indicating moderate to strong residual autocorrelation about the fitted trend.
-The correlation function is an exponentially decreasing function of temporal separation ($\Delta_t$), and whilst observations that are a few years apart are quite strongly dependent on one another, this dependence drops off rapidly as $\Delta_t$ increases and is effectively zero when samples are separated by a decade or more (Figure\ \ref{fig:car1-plot}).
+The correlation function is an exponentially decreasing function of temporal separation ($h$), and whilst observations that are a few years apart are quite strongly dependent on one another, this dependence drops off rapidly as $h$ increases and is effectively zero when samples are separated by a decade or more (Figure\ \ref{fig:car1-plot}).
 
@@
 ```{r car1-plot, fig.height = 2.5, fig.width = 4, out.width = "0.5\\linewidth", fig.cap = "Estimated CAR(1) process from the GAM fitted to the Small Water $\\delta^{15}\\text{N}$ time series.
-$h(\\Delta_t, \\phi)$ is the correlation between residuals separated by $\\Delta_t$ years, where $\\hat{\\phi} = \\text{0.6}$.
-The shaded band is a 95\\% pointwise confidence interval on the estimated correlation $h$."}
+$c(h, \\phi)$ is the correlation between residuals separated by $h$ years, where $\\hat{\\phi} = \\text{0.6}$.
+The shaded band is a 95\\% pointwise confidence interval on the estimated correlation $c$."}
 ## plot CAR(1) process
@@
     geom_line() +
-    ylab(expression(italic(h) * (list(Delta[t], varphi)))) +
-    xlab(expression(Delta[t] ~ (years)))
+    ylab(expression(italic(c) * (list(italic(h), phi)))) +
+    xlab(expression(italic(h) ~ (years)))
 car1Plt
@@
 (a) Estimated age model using a monotonically-constrained spline fitted to ${}^{210}\\text{Pb}$ inferred ages for selected depths in the sediment core (red points).
-The uncertainty in the ${}^{210}\\text{Pb}$ inferred age is show by the red vertical bars.
+The uncertainty in the ${}^{210}\\text{Pb}$ inferred age is shown by the red vertical bars.
 The fitted age model is illustrated by the solid black line.

### 2023-10-16 manuscript.tex
tweak to bib file
update supplementary materials to work with current gratia

@@
 
-\date{August 16, 2018}
+\date{September 19, 2018}
 
@@
 
-\usepackage{lineno}
-\linenumbers
+%\usepackage{lineno}
+%\linenumbers
 
@@
 \caption{Example time series; a) Small Water bulk organic matter $\delta^{15}\text{N}$ time series on a ${}^{210}\text{Pb}$ time scale, and b) Braya-Sø \uk{} time series on a calibrated ${}^{14}\text{C}$ time scale.
-The observations \uk{} time series have been joined by lines purely as a visual aid to highlight potential trends.}\label{fig:data-figure}
+The observations in the \uk{} time series have been joined by lines purely as a visual aid to highlight potential trends.}\label{fig:data-figure}
 \end{figure}
@@
 from a member of the exponential family of distributions --- such as the
-Poisson for count data, the binomial for presence/absence or counts from
-a total --- with expected value \(\mu_t\) and possibly some additional
-parameters \(\Theta\) (\(y_t \sim \text{EF}(\mu_t, \Theta)\)).
-Additionally, many software implementations of the above model also
-allow for distributions that are not within the exponential family but
-which can be fitted using an algorithm superficially similar to the one
-used to fit GAMs to members of the exponential family (e.g. Wood et al.,
-2016). Common examples of such extended families include the negative
-binomial distribution (for overdispersed counts) and the beta
-distribution (for true proportions or other interval-bounded data).
+Poisson for count data, or the binomial for presence/absence or counts
+from a total --- with expected value \(\mu_t\) and possibly some
+additional parameters \(\Theta\)
+(\(y_t \sim \text{EF}(\mu_t, \Theta)\)). Additionally, many software
+implementations of the above model also allow for distributions that are
+not within the exponential family but which can be fitted using an
+algorithm superficially similar to the one used to fit GAMs to members
+of the exponential family (e.g. Wood et al., 2016). Common examples of
+such extended families include the negative binomial distribution (for
+overdispersed counts) and the beta distribution (for true proportions or
+other interval-bounded data).
 
@@
 polynomial. One of the primary criticisms is that polynomial basis
-functions are global (Magee, 1998); the value of \(f\) at time point
-\(x_t\) affects the value of \(f\) at time point \(x_{t+s}\) even if the
-two time points are at opposite ends of the series. There are many other
+functions are global (Magee, 1998); the value of \(f\) at time \(x_t\)
+affects the value of \(f\) at time point \(x_{t+s}\) even if the two
+time points are at opposite ends of the series. There are many other
 bases we could use; here I discuss one such set of bases, that of
@@
 \caption{Basis functions for the time covariate and the Small Water $\delta^{15}\text{N}$ time series.
-A rank (size) 7 cubic regression spline (CRS) basis expansion is show in a), with knots, indicated by tick marks on the x-axis, spread evenly through the rang of the data.
+A rank (size) 7 cubic regression spline (CRS) basis expansion is show in a), with knots, indicated by tick marks on the x-axis, spread evenly through the range of the data.
 b) shows the same CRS basis functions weighted by the estimated coefficients $\beta_j$, plus the resulting GAM trend line (black line drawn through the data).
@@
 \ref{fig:basis-function-example-plot}a. The tick marks on the x-axis
-show the locations of the knots, which are located at the ends of the
+show the positions of the knots, which are located at the ends of the
 series and evenly in between. Notice that in this particular
@@
 the \texttt{gamm()} function (Wood, 2004) in the \emph{mgcv} package
-(Wood, 2017) for R (R Core Team, 2017).
+(Wood, 2017) for R (R Core Team, 2018).
 
@@
 well-captures the strong pattern in the data. The trend is statistically
-significant (estimated degrees of freedom = 7.95; \(F\) = 47.44,
+significant (effective degrees of freedom = 7.95; \(F\) = 47.44,
 approximate \(p\) value = \(\ll\) 0.0001). However further analysis of
@@
 The primary issue leading to poor fit is that neither model accounts for
-the different variance (known as (heteroscedasticity) of each
-observation in the \uk{} record. This seemingly isn't a problem for GCV
-which minimizes prediction error. The sediments in Braya-Sø are not
-annually laminated and therefore the core was sliced at regular depth
-intervals. Owing to compaction of older sediments and variation in
-accumulation rates over time, each sediment slice represents a different
-number of ``lake years''. We can think of older samples as representing
-some average of many years of sediment deposition, whilst younger
-samples are representative of fewer of these lake years. The average of
-a larger set of numbers is estimated more precisely than the average of
-a smaller set, all things equal. A direct result of this variable
-averaging of lake years it that some samples are more precise and
-therefore have lower variance than other samples and yet the model
-assumed that the variance was constant across samples.
+the different variance (known as heteroscedasticity) of each observation
+in the \uk{} record. This seemingly isn't a problem for GCV which
+minimizes prediction error. The sediments in Braya-Sø are not annually
+laminated and therefore the core was sliced at regular depth intervals.
+Owing to compaction of older sediments and variation in accumulation
+rates over time, each sediment slice represents a different number of
+``lake years''. We can think of older samples as representing some
+average of many years of sediment deposition, whilst younger samples are
+representative of fewer of these lake years. The average of a larger set
+of numbers is estimated more precisely than the average of a smaller
+set, all things equal. A direct result of this variable averaging of
+lake years it that some samples are more precise and therefore have
+lower variance than other samples and yet the model assumed that the
+variance was constant across samples.
 
@@
 standard errors derived from the Bayesian posterior covariance matrix
-for the estimated mode parameters. Such intervals have the interesting
+for the estimated model parameters. Such intervals have the interesting
 property that they have good \emph{across-the-function} coverage when
@@
 by \(\text{SE}(\hat{f}(x_t))\) yielded an interval that contained the
-whole function for \((1-\alpha)\) 100\% of the functions simulated. In
+whole function for \$(1-\alpha)\$100\% of the functions simulated. In
 practice, the simulation method of Ruppert et al. (2003) does not
@@
 derivative estimates using standard theory
-(i.e.~\(\pm 1.96 \times \text{SE}(\hat{y}_t)\) for a 85\% interval) and
+(i.e.~\(\pm 1.96 \times \text{SE}(\hat{y}_t)\) for a 95\% interval) and
 the covariance matrix of the spline coefficients. A (1 -
@@
 the fitted trend. The correlation function is an exponentially
-decreasing function of temporal separation (\(\Delta_t\)), and whilst
+decreasing function of temporal separation (\(h\)), and whilst
 observations that are a few years apart are quite strongly dependent on
-one another, this dependence drops off rapidly as \(\Delta_t\) increases
-and is effectively zero when samples are separated by a decade or more
+one another, this dependence drops off rapidly as \(h\) increases and is
+effectively zero when samples are separated by a decade or more
 (Figure~\ref{fig:car1-plot}).
@@
 \caption{Estimated CAR(1) process from the GAM fitted to the Small Water $\delta^{15}\text{N}$ time series.
-$h(\Delta_t, \phi)$ is the correlation between residuals separated by $\Delta_t$ years, where $\hat{\phi} = \text{0.6}$.
-The shaded band is a 95\% pointwise confidence interval on the estimated correlation $h$.}\label{fig:car1-plot}
+$c(h, \phi)$ is the correlation between residuals separated by $h$ years, where $\hat{\phi} = \text{0.6}$.
+The shaded band is a 95\% pointwise confidence interval on the estimated correlation $c$.}\label{fig:car1-plot}
 \end{figure}
@@
 (a) Estimated age model using a monotonically-constrained spline fitted to ${}^{210}\text{Pb}$ inferred ages for selected depths in the sediment core (red points).
-The uncertainty in the ${}^{210}\text{Pb}$ inferred age is show by the red vertical bars.
+The uncertainty in the ${}^{210}\text{Pb}$ inferred age is shown by the red vertical bars.
 The fitted age model is illustrated by the solid black line.
@@
 approach revisited with an application to surface temperature trends.
-\emph{Journal of applied statistics} 34, 963--972.
+\emph{Journal of Applied Statistics} 34, 963--972.
 doi:\href{https://doi.org/10.1080/02664760701590418}{10.1080/02664760701590418}.
@@
 generalized additive modelling. \emph{ArXiv e-prints}.
-
-\hypertarget{ref-R-core-team}{}
-R Core Team (2017). \emph{R: A language and environment for statistical
-computing}. Vienna, Austria: R Foundation for Statistical Computing
-Available at: \url{https://www.R-project.org/}.
 

### 2023-10-16 supplementary-materials.Rmd
update supplementary materials to work with current gratia

@@
 mail: "gavin.simpson@uregina.ca"
-date: August 16, 2018
+date: October 16, 2023
 keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
@@
 
-The following packages are required: *mgcv*, *scam*, *ggplot2*, *cowplot*, and *tidyr*.
+The following packages are required: *mgcv*, *scam*, *ggplot2*, *cowplot*, *dplyr*, and *tidyr*.
 
@@
 library("tidyr")
+library("dplyr")
 ```
-In addition, the *gratia* package is required; this in-development package is not on CRAN but can be installed directly from GitHub using functions from the *devtools* package.
-To install the package, install *devtools* and then use `devtools::install_github()` to install *gratia*, as shown below:
+In addition, the *gratia* package is required; install *gratia*, using:
 
 ```{r install-gratia, eval = FALSE}
-## gratia is not on CRAN, install from github:
-install.packages("devtools")
-devtools::install_github("gavinsimpson/gratia")
+## requires development version of gratia that is not on CRAN, install from github:
+install.packages("gratia", repos = c("https://gavinsimpson.r-universe.dev",
+    "https://cloud.r-project.org"))
+stopifnot(packageVersion("gratia") >= "0.8.1.42")
 ```
@@
     geom_line() +
-    ylab(expression(italic(h) * (list(Delta[t], varphi)))) +
-    xlab(expression(Delta[t] ~ (years)))
+    ylab(expression(italic(c) * (list(h, varphi)))) +
+    xlab(expression(h ~ (years)))
 car1Plt
@@
 The same model as used for Small Water was also initially attempted with the Braya-Sø data.
-However, in order to even fit the model with both a smooth and the CAR(1) process, I had to change the default optimizer used to estimate the model parameters, and reduce the basis dimension, `k`, to a small number.
 I also fit the model using GCV, which is, at the time of writing, the default in `gam()`, hence no `method` argument
 
-```{r fit-braya-so-car1-and-gcv-models}
+```{r fit-braya-so-car1-and-gcv-models, error = TRUE}
 ## fit the car(1) model --- needs optim as this is not a stable fit!
 ## also needs k setting lower than default
-braya.car1 <- gamm(UK37 ~ s(Year, k = 5), data = braya, 
+braya.car1 <- gamm(UK37 ~ s(Year, k = 10), data = braya,
                    correlation = corCAR1(form = ~ Year),
-                   method = "REML",
-		           control = list(niterEM = 0, optimMethod = "BFGS", 
-                                  opt = "optim"))
+                   method = "REML")
 
@@
 ## do the simulations
-sims <- simulate(mod, nsim = nsim, newdata = newYear, unconditional = TRUE)
-
-## rearrange the output into a long/tidy format
-colnames(sims) <- paste0("sim", seq_len(nsim))
-sims <- setNames(stack(as.data.frame(sims)), c("simulated", "run"))
-sims <- transform(sims, Year = rep(newYear$Year, nsim),
-                  simulated = simulated)
+sims <- smooth_samples(mod$gam, n = nsim, data = newYear,
+    unconditional = TRUE)
+sims <- mutate(sims, value = value + coef(mod$gam)[1L]) # add intercept
 
@@
     geom_line(data = sims,
-              mapping = aes(y = simulated, x = Year, group = run),
+              mapping = aes(y = value, x = Year, group = draw),
               colour = "grey80") +
@@
 set.seed(1)
-sims2 <- simulate(braya_reml, nsim = nsim, newdata = newBraya,
+sims2 <- smooth_samples(braya_reml, n = nsim, data = newBraya,
     unconditional = TRUE)
-
-## rearrange the output into a long/tidy format
-colnames(sims2) <- paste0("sim", seq_len(nsim))
-sims2 <- setNames(stack(as.data.frame(sims2)),
-                  c("simulated", "run"))
-sims2 <- transform(sims2, Year = rep(newBraya$Year, nsim),
-                   simulated = simulated)
+sims2 <- mutate(sims2, value = value + coef(braya_reml)[1L]) # add intercept
 
@@
     geom_line(data = sims2,
-              mapping = aes(y = simulated, x = Year, group = run),
+              mapping = aes(y = value, x = Year, group = draw),
               colour = "grey80") +
@@
 ```{r small-compare-intervals}
-sw.cint <- confint(mod, parm = "Year", newdata = newYear,
-                   type = "confidence")
-sw.sint <- confint(mod, parm = "Year", newdata = newYear,
-                   type = "simultaneous")
-head(sw.sint)
+sw.cint <- confint(mod, parm = "s(Year)", data = newYear, type = "confidence")
+sw.sint <- confint(mod, parm = "s(Year)", data = newYear, type = "simultaneous")
+sw.sint
 ```
@@
 ```{r plot-small-water-intervals}
-smallInt.plt <- ggplot(sw.cint, aes(x = Year, y = est)) +
+smallInt.plt <- ggplot(sw.cint, aes(x = Year, y = .estimate)) +
     geom_ribbon(data = sw.sint,
-                mapping = aes(ymin = lower, ymax = upper, x = Year),
+                mapping = aes(ymin = .lower_ci, ymax = .upper_ci, x = Year),
                 fill = "grey80", inherit.aes = FALSE) +
-    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
+    geom_ribbon(mapping = aes(ymin = .lower_ci, ymax = .upper_ci, x = Year),
                 fill = "grey60", inherit.aes = FALSE) +
@@
 ```{r braya-compare-intervals}
-bs.cint <- confint(braya_reml, parm = "Year", newdata = newBraya,
+bs.cint <- confint(braya_reml, parm = "s(Year)", data = newBraya,
                    type = "confidence")
-bs.sint <- confint(braya_reml, parm = "Year", newdata = newBraya,
+bs.sint <- confint(braya_reml, parm = "s(Year)", data = newBraya,
                    type = "simultaneous")
@@
 ```{r plot-braya-so-intervals}
-brayaInt.plt <- ggplot(bs.cint, aes(x = Year, y = est)) +
+brayaInt.plt <- ggplot(bs.cint, aes(x = Year, y = .estimate)) +
     geom_ribbon(data = bs.sint,
-                mapping = aes(ymin = lower, ymax = upper, x = Year),
+                mapping = aes(ymin = .lower_ci, ymax = .upper_ci, x = Year),
                 fill = "grey80", inherit.aes = FALSE) +
-    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
+    geom_ribbon(mapping = aes(ymin = .lower_ci, ymax = .upper_ci, x = Year),
                 fill = "grey60", inherit.aes = FALSE) +
@@
 ## monotonic spline age-depth model
-swAge$Error[1] <- 1.1
-swAgeMod <- scam(Date ~ s(Depth, k = 5, bs = "mpd"), data = swAge,
-                 weights = 1 / swAge$Error, gamma = 1.4)
+swAge$Error[1] <- 1
+swAgeMod <- scam(Date ~ s(Depth, k = 5, bs = "cv"), data = swAge,
+    weights = (1 / swAge$Error) / mean(1 / swAge$Error), gamma = 1)
 ```
@@
 
-Next, predict from the estimated age model, and draw 25 samples from the posterior distribution using `simulate()`.
-The results are tidied into a format suitable for further processing and plotting.
-Note that the posterior samples here are only used for plotting.
+Next, predict from the estimated age model, and draw 25 samples from the posterior distribution using `smooth_samples()`. Note that the posterior samples here are only used for plotting.
 
@@
 ## predict from the age model for a smooth set of points in `Depth`
-newAge <- with(swAge, data.frame(Depth = seq(min(Depth), max(Depth),
-                                             length.out = 200)))
-newAge <- transform(newAge,
-                    fitted = predict(swAgeMod, newdata = newAge, 
-                                     type = "response"))
-newSims <- as.data.frame(simulate(swAgeMod, nsim = 25, newdata = newAge))
-newSims <- cbind(Depth = newAge$Depth, newSims)
-newSims <- gather(newSims, Simulation, Age, -Depth)
+newAge <- data_slice(swAgeMod, Depth = evenly(Depth, n = 200))
+newAgeFit <- fitted_values(swAgeMod, data = newAge)
+newSims <- smooth_samples(swAgeMod, n = 25, data = newAge) %>%
+    mutate(value = value + coef(swAgeMod)[1L]) # %>%
+    #left_join(mutate(newAge, .row = row_number()), by = join_by(.row == .row))
 ```
 
-In the next code chunk, I draw 100 samples from the posterior distribution of the age model, but notice that I pass in the `small` data to `newdata` in the call to `simulate()` as the locations I want new age estimates for are the depths for which we have δ^15^N values.
+In the next code chunk, I draw 100 samples from the posterior distribution of the age model, but notice that I pass in the `small` data to `data` in the call to `smooth_samples()` as the locations I want new age estimates for the depths for which we have δ^15^N values.
 A small function (`fitSWModels`) is written to prepare each simulation for fitting and then actually fit the GAM plus CAR(1) model using the updated age information.
@@
 ## simulate from age model; each column is a simulation
-ageSims <- simulate(swAgeMod, nsim = 100, newdata = small, seed = 42)
-ageSims <- as.data.frame(ageSims)
+ageSims <- smooth_samples(swAgeMod, n = 100, data = small, seed = 42)
 
 fitSWModels <- function(x, y, knots) {
-    dat <- data.frame(d15N = y, Year = x)
+    dat <- data.frame(d15N = y,
+        Year = pull(x, value) + coef(swAgeMod, parameterized = TRUE)[1L])
     m <- gamm(d15N ~ s(Year, k = 15), data = dat, method = "REML",
@@
 ## generate new trends using draws from age-model posterior
-simTrendMods <- lapply(ageSims, fitSWModels, y = small$d15N, knots = knots)
+simTrendMods <- lapply(ageSims %>% split(ageSims$draw), fitSWModels,
+    y = small$d15N, knots = knots)
 
@@
 simulateSWModels <- function(mod, newdata, nsim, seed = 42) {
-    sims <- simulate(mod, nsim = nsim, newdata = newdata, seed = seed)
-    as.vector(sims)
+    sims <- fitted_samples(mod, n = nsim, data = newdata, seed = seed)
+    pull(sims, .fitted)
 }
@@
     geom_line(data = newSims,
-              mapping = aes(y = Age, x = Depth, group = Simulation),
+              mapping = aes(y = value, x = Depth, group = draw),
               alpha = 1, colour = "grey80") +
-    geom_line(data = newAge, mapping = aes(y = fitted, x = Depth)) +
+    geom_line(data = newAgeFit, mapping = aes(y = .fitted, x = Depth)) +
     geom_point(size = 1.5, colour = "red") +

### 2023-10-16 supplementary-materials.tex
update supplementary materials to work with current gratia

@@
+\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
+\PassOptionsToPackage{hyphens}{url}
+%
 \documentclass[12pt,]{article}
@@
   \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provides euro and other symbols
 \else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
+  \usepackage{unicode-math}
   \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
@@
 \IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\usepackage[]{microtype}
 \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
-\usepackage[left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm]{geometry}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \usepackage{hyperref}
-\hypersetup{unicode=true,
+\hypersetup{
             pdftitle={Supplementary materials for: Modelling palaeoecological time series using generalized additive models},
@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage[left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm]{geometry}
 \usepackage{color}
@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
-\newcommand{\ImportTok}[1]{{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
-\newcommand{\BuiltInTok}[1]{{#1}}
-\newcommand{\ExtensionTok}[1]{{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
-\newcommand{\RegionMarkerTok}[1]{{#1}}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
-\newcommand{\NormalTok}[1]{{#1}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx,grffile}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{Supplementary materials for: Modelling palaeoecological time series
-using generalized additive models}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-    \author{Gavin L. Simpson}
-    \preauthor{\centering\large\emph}
-  \postauthor{\par}
-      \predate{\centering\large\emph}
-  \postdate{\par}
-    \date{August 16, 2018}
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 
@@
 
-\usepackage{lineno}
-\linenumbers
+%\usepackage{lineno}
+%\linenumbers
+
+\title{Supplementary materials for: Modelling palaeoecological time series
+using generalized additive models}
+\author{Gavin L. Simpson}
+\date{October 16, 2023}
 
@@
 
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 The following packages are required: \emph{mgcv}, \emph{scam},
-\emph{ggplot2}, \emph{cowplot}, and \emph{tidyr}.
+\emph{ggplot2}, \emph{cowplot}, \emph{dplyr}, and \emph{tidyr}.
 
@@
 \begin{verbatim}
-#> This is mgcv 1.8-24. For overview type 'help("mgcv-package")'.
+#> This is mgcv 1.9-0. For overview type 'help("mgcv-package")'.
 \end{verbatim}
@@
 \begin{verbatim}
-#> This is scam 1.2-2.
+#> This is scam 1.2-14.
 \end{verbatim}
@@
 \KeywordTok{library}\NormalTok{(}\StringTok{"cowplot"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-#> 
-#> Attaching package: 'cowplot'
-\end{verbatim}
-
-\begin{verbatim}
-#> The following object is masked from 'package:ggplot2':
-#> 
-#>     ggsave
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \KeywordTok{library}\NormalTok{(}\StringTok{"tidyr"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-In addition, the \emph{gratia} package is required; this in-development
-package is not on CRAN but can be installed directly from GitHub using
-functions from the \emph{devtools} package. To install the package,
-install \emph{devtools} and then use
-\texttt{devtools::install\_github()} to install \emph{gratia}, as shown
-below:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## gratia is not on CRAN, install from github:}
-\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
-\NormalTok{devtools::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"gavinsimpson/gratia"}\NormalTok{)}
+\KeywordTok{library}\NormalTok{(}\StringTok{"dplyr"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#> 
+#> Attaching package: 'dplyr'
+\end{verbatim}
+
+\begin{verbatim}
+#> The following object is masked from 'package:nlme':
+#> 
+#>     collapse
+\end{verbatim}
+
+\begin{verbatim}
+#> The following objects are masked from 'package:stats':
+#> 
+#>     filter, lag
+\end{verbatim}
+
+\begin{verbatim}
+#> The following objects are masked from 'package:base':
+#> 
+#>     intersect, setdiff, setequal, union
+\end{verbatim}
+
+In addition, the \emph{gratia} package is required; install
+\emph{gratia}, using:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## requires development version of gratia that is not on CRAN, install from github:}
+\KeywordTok{install.packages}\NormalTok{(}\StringTok{"gratia"}\NormalTok{, }\DataTypeTok{repos =} \KeywordTok{c}\NormalTok{(}\StringTok{"https://gavinsimpson.r-universe.dev"}\NormalTok{,}
+    \StringTok{"https://cloud.r-project.org"}\NormalTok{))}
+\KeywordTok{stopifnot}\NormalTok{(}\KeywordTok{packageVersion}\NormalTok{(}\StringTok{"gratia"}\NormalTok{) }\OperatorTok{>=}\StringTok{ "0.8.1.42"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{## Default ggplot theme}
+\CommentTok{## Default ggplot theme}
 \KeywordTok{theme_set}\NormalTok{(}\KeywordTok{theme_bw}\NormalTok{())}
@@
 
-\section{Load the data sets}\label{load-the-data-sets}
+\hypertarget{load-the-data-sets}{%
+\section{Load the data sets}\label{load-the-data-sets}}
 
@@
 \begin{Highlighting}[]
-\NormalTok{## source Small Water data}
+\CommentTok{## source Small Water data}
 \NormalTok{small <-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"./data/small-water/small-water-isotope-data.rds"}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\NormalTok{## load braya so data set}
+\CommentTok{## load braya so data set}
 \NormalTok{braya <-}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"./data/braya-so/DAndrea.2011.Lake Braya So.txt"}\NormalTok{,}
                     \DataTypeTok{skip =} \DecValTok{84}\NormalTok{)}
-\NormalTok{## clean up variable names}
+\CommentTok{## clean up variable names}
 \KeywordTok{names}\NormalTok{(braya) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Depth"}\NormalTok{, }\StringTok{"DepthUpper"}\NormalTok{, }\StringTok{"DepthLower"}\NormalTok{, }\StringTok{"Year"}\NormalTok{, }\StringTok{"YearYoung"}\NormalTok{,}
                   \StringTok{"YearOld"}\NormalTok{, }\StringTok{"UK37"}\NormalTok{)}
-\NormalTok{## add a variable for the amount of time per sediment sample}
-\NormalTok{braya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(braya, }\DataTypeTok{sampleInterval =} \NormalTok{YearYoung -}\StringTok{ }\NormalTok{YearOld)}
+\CommentTok{## add a variable for the amount of time per sediment sample}
+\NormalTok{braya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(braya, }\DataTypeTok{sampleInterval =}\NormalTok{ YearYoung }\OperatorTok{-}\StringTok{ }\NormalTok{YearOld)}
 \KeywordTok{head}\NormalTok{(braya)}
@@
 \begin{verbatim}
-#>   Depth DepthUpper DepthLower     Year YearYoung YearOld   UK37
-#> 1  0.25        0.0        0.5 1999.125   2006.00 1992.25 -0.640
-#> 2  0.75        0.5        1.0 1985.375   1992.25 1978.50 -0.637
-#> 3  1.25        1.0        1.5 1971.525   1978.50 1964.55 -0.614
-#> 4  1.75        1.5        2.0 1957.575   1964.55 1950.60 -0.627
-#> 5  2.25        2.0        2.5 1943.150   1950.60 1935.70 -0.633
-#> 6  2.75        2.5        3.0 1928.250   1935.70 1920.80 -0.616
-#>   sampleInterval
-#> 1          13.75
-#> 2          13.75
-#> 3          13.95
-#> 4          13.95
-#> 5          14.90
-#> 6          14.90
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## plot labels}
-\NormalTok{d15n_label <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(delta^\{}\DecValTok{15}\NormalTok{\}*N)}
-\NormalTok{braya_ylabel <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(U)[}\DecValTok{37}\NormalTok{]^\{}\KeywordTok{italic}\NormalTok{(k)\})}
+#>   Depth DepthUpper DepthLower     Year YearYoung YearOld   UK37 sampleInterval
+#> 1  0.25        0.0        0.5 1999.125   2006.00 1992.25 -0.640          13.75
+#> 2  0.75        0.5        1.0 1985.375   1992.25 1978.50 -0.637          13.75
+#> 3  1.25        1.0        1.5 1971.525   1978.50 1964.55 -0.614          13.95
+#> 4  1.75        1.5        2.0 1957.575   1964.55 1950.60 -0.627          13.95
+#> 5  2.25        2.0        2.5 1943.150   1950.60 1935.70 -0.633          14.90
+#> 6  2.75        2.5        3.0 1928.250   1935.70 1920.80 -0.616          14.90
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## plot labels}
+\NormalTok{d15n_label <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(delta}\OperatorTok{^}\NormalTok{\{}\DecValTok{15}\NormalTok{\}}\OperatorTok{*}\NormalTok{N)}
+\NormalTok{braya_ylabel <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(U)[}\DecValTok{37}\NormalTok{]}\OperatorTok{^}\NormalTok{\{}\KeywordTok{italic}\NormalTok{(k)\})}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{## plot Small Water data}
-\NormalTok{small_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(small, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N)) +}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
-
-\NormalTok{## plot Braya-So data}
-\NormalTok{braya_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37)) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{colour =} \StringTok{"grey"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
-
-\NormalTok{## Recreate plot from manuscript}
+\CommentTok{## plot Small Water data}
+\NormalTok{small_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(small, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ d15N)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
+
+\CommentTok{## plot Braya-So data}
+\NormalTok{braya_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ UK37)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{colour =} \StringTok{"grey"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
+
+\CommentTok{## Recreate plot from manuscript}
 \KeywordTok{plot_grid}\NormalTok{(small_plt, braya_plt, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{, }\DataTypeTok{align =} \StringTok{"hv"}\NormalTok{,}
@@
 
-\section{Fitting GAMs}\label{fitting-gams}
+\hypertarget{fitting-gams}{%
+\section{Fitting GAMs}\label{fitting-gams}}
 
@@
 
-\subsection{Small Water}\label{small-water}
+\hypertarget{small-water}{%
+\subsection{Small Water}\label{small-water}}
 
@@
 \begin{Highlighting}[]
-\NormalTok{m <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
+\NormalTok{m <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(d15N }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ small, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{## fit small water GAM using gamm() with a CAR(1)}
-\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small,}
-            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
+\CommentTok{## fit small water GAM using gamm() with a CAR(1)}
+\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ small,}
+            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \OperatorTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{## estimate of phi and confidence interval}
-\NormalTok{smallPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(mod$lme, }\DataTypeTok{which =} \StringTok{"var-cov"}\NormalTok{)$corStruct}
+\CommentTok{## estimate of phi and confidence interval}
+\NormalTok{smallPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(mod}\OperatorTok{$}\NormalTok{lme, }\DataTypeTok{which =} \StringTok{"var-cov"}\NormalTok{)}\OperatorTok{$}\NormalTok{corStruct}
 \NormalTok{smallPhi}
@@
 #>         lower      est.     upper
-#> Phi 0.282388 0.6026967 0.8539689
+#> Phi 0.2811374 0.6026967 0.8547379
 #> attr(,"label")
@@
 \begin{Highlighting}[]
-\NormalTok{## summary object}
-\KeywordTok{summary}\NormalTok{(mod$gam)}
+\CommentTok{## summary object}
+\KeywordTok{summary}\NormalTok{(mod}\OperatorTok{$}\NormalTok{gam)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{## plot CAR(1) process}
+\CommentTok{## plot CAR(1) process}
 \NormalTok{S <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DataTypeTok{length =} \DecValTok{100}\NormalTok{)}
@@
                  \KeywordTok{c}\NormalTok{(}\StringTok{"Lower"}\NormalTok{,}\StringTok{"Correlation"}\NormalTok{,}\StringTok{"Upper"}\NormalTok{))}
-\NormalTok{car1 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(car1, }\DataTypeTok{S =} \NormalTok{S)}
-
-\NormalTok{car1Plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(car1, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{S, }\DataTypeTok{y =} \NormalTok{Correlation)) +}
-\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymax =} \NormalTok{Upper, }\DataTypeTok{ymin =} \NormalTok{Lower),}
-                \DataTypeTok{fill =} \StringTok{"black"}\NormalTok{, }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
-\StringTok{    }\KeywordTok{ylab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(h) *}\StringTok{ }\NormalTok{(}\KeywordTok{list}\NormalTok{(Delta[t], varphi)))) +}
-\StringTok{    }\KeywordTok{xlab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(Delta[t] ~}\StringTok{ }\NormalTok{(years)))}
+\NormalTok{car1 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(car1, }\DataTypeTok{S =}\NormalTok{ S)}
+
+\NormalTok{car1Plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(car1, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ S, }\DataTypeTok{y =}\NormalTok{ Correlation)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymax =}\NormalTok{ Upper, }\DataTypeTok{ymin =}\NormalTok{ Lower),}
+                \DataTypeTok{fill =} \StringTok{"black"}\NormalTok{, }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{() }\OperatorTok{+}
+\StringTok{    }\KeywordTok{ylab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(c) }\OperatorTok{*}\StringTok{ }\NormalTok{(}\KeywordTok{list}\NormalTok{(h, varphi)))) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{xlab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(h }\OperatorTok{~}\StringTok{ }\NormalTok{(years)))}
 \NormalTok{car1Plt}
@@
 
-\NormalTok{## create new data to predict at; 200 evenly-spaced values over `Year`}
+\CommentTok{## create new data to predict at; 200 evenly-spaced values over `Year`}
 \NormalTok{newYear <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                              \DataTypeTok{length.out =} \DecValTok{200}\NormalTok{)))}
-\NormalTok{## Predict from the fitted model; note we predict from the $gam part}
+\CommentTok{## Predict from the fitted model; note we predict from the $gam part}
 \NormalTok{newYear <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newYear,}
-                 \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(mod$gam, newYear, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
-\NormalTok{## Create the confidence interval}
-\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\DataTypeTok{df =} \KeywordTok{df.residual}\NormalTok{(mod$gam))}
+                 \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(mod}\OperatorTok{$}\NormalTok{gam, newYear, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
+\CommentTok{## Create the confidence interval}
+\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\DataTypeTok{df =} \KeywordTok{df.residual}\NormalTok{(mod}\OperatorTok{$}\NormalTok{gam))}
 \NormalTok{newYear <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newYear,}
-                     \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
-                     \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}
-
-\NormalTok{## Plot estimated trend}
-\NormalTok{small_fitted <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(newYear, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
-\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year), }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{,}
-                \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N),}
-               \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
+                     \DataTypeTok{upper =}\NormalTok{ fit }\OperatorTok{+}\StringTok{ }\NormalTok{(crit.t }\OperatorTok{*}\StringTok{ }\NormalTok{se.fit),}
+                     \DataTypeTok{lower =}\NormalTok{ fit }\OperatorTok{-}\StringTok{ }\NormalTok{(crit.t }\OperatorTok{*}\StringTok{ }\NormalTok{se.fit))}
+
+\CommentTok{## Plot estimated trend}
+\NormalTok{small_fitted <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(newYear, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ fit)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =}\NormalTok{ lower, }\DataTypeTok{ymax =}\NormalTok{ upper, }\DataTypeTok{x =}\NormalTok{ Year), }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{,}
+                \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ small, }\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ d15N),}
+               \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{() }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
 \NormalTok{small_fitted}
@@
 
-\subsection{Braya-Sø}\label{braya-s}
+\hypertarget{braya-suxf8}{%
+\subsection{Braya-Sø}\label{braya-suxf8}}
 
 The same model as used for Small Water was also initially attempted with
-the Braya-Sø data. However, in order to even fit the model with both a
-smooth and the CAR(1) process, I had to change the default optimizer
-used to estimate the model parameters, and reduce the basis dimension,
-\texttt{k}, to a small number. I also fit the model using GCV, which is,
-at the time of writing, the default in \texttt{gam()}, hence no
-\texttt{method} argument
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## fit the car(1) model --- needs optim as this is not a stable fit!}
-\NormalTok{## also needs k setting lower than default}
-\NormalTok{braya.car1 <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya, }
-                   \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year),}
-                   \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
-                   \DataTypeTok{control =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{niterEM =} \DecValTok{0}\NormalTok{, }\DataTypeTok{optimMethod =} \StringTok{"BFGS"}\NormalTok{, }
-                                  \DataTypeTok{opt =} \StringTok{"optim"}\NormalTok{))}
-
-\NormalTok{## fit model using GCV}
-\NormalTok{braya.gcv <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{30}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya)}
-
-\NormalTok{## estimate of phi and confidence interval}
-\NormalTok{brayaPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(braya.car1$lme)$corStruct}
+the Braya-Sø data. I also fit the model using GCV, which is, at the time
+of writing, the default in \texttt{gam()}, hence no \texttt{method}
+argument
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## fit the car(1) model --- needs optim as this is not a stable fit!}
+\CommentTok{## also needs k setting lower than default}
+\NormalTok{braya.car1 <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{10}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ braya,}
+                   \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \OperatorTok{~}\StringTok{ }\NormalTok{Year),}
+                   \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
+
+\CommentTok{## fit model using GCV}
+\NormalTok{braya.gcv <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{30}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ braya)}
+
+\CommentTok{## estimate of phi and confidence interval}
+\NormalTok{brayaPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(braya.car1}\OperatorTok{$}\NormalTok{lme)}\OperatorTok{$}\NormalTok{corStruct}
 \NormalTok{brayaPhi}
@@
 #>            lower      est. upper
-#> Phi 2.680033e-31 0.2000156     1
+#> Phi 1.239228e-18 0.2000199     1
 #> attr(,"label")
@@
 
-\NormalTok{## data to predict at}
+\CommentTok{## data to predict at}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
-                                              \DataTypeTok{length.out =} \NormalTok{N)))}
-\NormalTok{## add predictions from GAMM + CAR(1) model}
+                                              \DataTypeTok{length.out =}\NormalTok{ N)))}
+\CommentTok{## add predictions from GAMM + CAR(1) model}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newBraya,}
-                  \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(braya.car1$gam, newBraya,}
+                  \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(braya.car1}\OperatorTok{$}\NormalTok{gam, newBraya,}
                                      \DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
-\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\DataTypeTok{df =} \KeywordTok{df.residual}\NormalTok{(braya.car1$gam))}
+\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\DataTypeTok{df =} \KeywordTok{df.residual}\NormalTok{(braya.car1}\OperatorTok{$}\NormalTok{gam))}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newBraya,}
-                      \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
-                      \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}
-
-\NormalTok{## add GAM GCV results}
-\NormalTok{fit_gcv <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(braya.gcv, }\DataTypeTok{newdata =} \NormalTok{newBraya, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)}
+                      \DataTypeTok{upper =}\NormalTok{ fit }\OperatorTok{+}\StringTok{ }\NormalTok{(crit.t }\OperatorTok{*}\StringTok{ }\NormalTok{se.fit),}
+                      \DataTypeTok{lower =}\NormalTok{ fit }\OperatorTok{-}\StringTok{ }\NormalTok{(crit.t }\OperatorTok{*}\StringTok{ }\NormalTok{se.fit))}
+
+\CommentTok{## add GAM GCV results}
+\NormalTok{fit_gcv <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(braya.gcv, }\DataTypeTok{newdata =}\NormalTok{ newBraya, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)}
 \NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\KeywordTok{df.residual}\NormalTok{(braya.gcv))}
-\NormalTok{newGCV <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year   =} \NormalTok{newBraya[[}\StringTok{"Year"}\NormalTok{]],}
-                     \DataTypeTok{fit    =} \NormalTok{fit_gcv$fit,}
-                     \DataTypeTok{se.fit =} \NormalTok{fit_gcv$se.fit)}
+\NormalTok{newGCV <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year   =}\NormalTok{ newBraya[[}\StringTok{"Year"}\NormalTok{]],}
+                     \DataTypeTok{fit    =}\NormalTok{ fit_gcv}\OperatorTok{$}\NormalTok{fit,}
+                     \DataTypeTok{se.fit =}\NormalTok{ fit_gcv}\OperatorTok{$}\NormalTok{se.fit)}
 \NormalTok{newGCV <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newGCV,}
-                    \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
-                    \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}
+                    \DataTypeTok{upper =}\NormalTok{ fit }\OperatorTok{+}\StringTok{ }\NormalTok{(crit.t }\OperatorTok{*}\StringTok{ }\NormalTok{se.fit),}
+                    \DataTypeTok{lower =}\NormalTok{ fit }\OperatorTok{-}\StringTok{ }\NormalTok{(crit.t }\OperatorTok{*}\StringTok{ }\NormalTok{se.fit))}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(newBraya, newGCV)                }\CommentTok{# bind on GCV results}
-\NormalTok{## Add indicator variable for model}
+\CommentTok{## Add indicator variable for model}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newBraya,}
                       \DataTypeTok{Method =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GAMM (CAR(1))"}\NormalTok{, }\StringTok{"GAM (GCV)"}\NormalTok{), }
-                                   \DataTypeTok{each =} \NormalTok{N))}
-
-\NormalTok{## plot CAR(1) and GCV fits}
-\NormalTok{braya_fitted <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{UK37, }\DataTypeTok{x =} \NormalTok{Year)) +}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
-\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newBraya,}
-                \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{ymin =} \NormalTok{lower,}
-                              \DataTypeTok{fill =} \NormalTok{Method),}
-                \DataTypeTok{alpha =} \FloatTok{0.3}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newBraya,}
-              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{fit, }\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{colour =} \NormalTok{Method)) +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#5e3c99"}\NormalTok{, }\StringTok{"#e66101"}\NormalTok{)) +}
-\StringTok{    }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#5e3c99"}\NormalTok{, }\StringTok{"#e66101"}\NormalTok{)) +}
+                                   \DataTypeTok{each =}\NormalTok{ N))}
+
+\CommentTok{## plot CAR(1) and GCV fits}
+\NormalTok{braya_fitted <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ UK37, }\DataTypeTok{x =}\NormalTok{ Year)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ newBraya,}
+                \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{ymax =}\NormalTok{ upper, }\DataTypeTok{ymin =}\NormalTok{ lower,}
+                              \DataTypeTok{fill =}\NormalTok{ Method),}
+                \DataTypeTok{alpha =} \FloatTok{0.3}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ newBraya,}
+              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ fit, }\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{colour =}\NormalTok{ Method)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#5e3c99"}\NormalTok{, }\StringTok{"#e66101"}\NormalTok{)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#5e3c99"}\NormalTok{, }\StringTok{"#e66101"}\NormalTok{)) }\OperatorTok{+}
 \StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"right"}\NormalTok{)}
@@
 
+\hypertarget{checking-if-the-size-of-the-basis-expansion-is-sufficient}{%
 \subsubsection{Checking if the size of the basis expansion is
-sufficient}\label{checking-if-the-size-of-the-basis-expansion-is-sufficient}
+sufficient}\label{checking-if-the-size-of-the-basis-expansion-is-sufficient}}
 
@@
 \begin{Highlighting}[]
-\NormalTok{braya_low_k <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
+\NormalTok{braya_low_k <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
 \end{Highlighting}
@@
 #> full convergence after 6 iterations.
-#> Gradient range [-4.468978e-08,8.696768e-10]
+#> Gradient range [-4.468969e-08,8.696457e-10]
 #> (score -187.2524 & scale 0.000689802).
@@
 
+\hypertarget{accounting-for-heteroscedasticity-due-to-time-averaging}{%
 \subsection{Accounting for heteroscedasticity due to time
-averaging}\label{accounting-for-heteroscedasticity-due-to-time-averaging}
+averaging}\label{accounting-for-heteroscedasticity-due-to-time-averaging}}
 
@@
 \begin{Highlighting}[]
-\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{40}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
+\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{40}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ braya,}
                   \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
-                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
+                  \DataTypeTok{weights =}\NormalTok{ sampleInterval }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
 \KeywordTok{summary}\NormalTok{(braya_reml)}
@@
 #>           k'  edf k-index p-value
-#> s(Year) 39.0 27.5    1.27    0.99
+#> s(Year) 39.0 27.5    1.27       1
 \end{verbatim}
@@
 
-\section{Posterior simulation}\label{posterior-simulation}
+\hypertarget{posterior-simulation}{%
+\section{Posterior simulation}\label{posterior-simulation}}
 
@@
 
-\NormalTok{## do the simulations}
-\NormalTok{sims <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(mod, }\DataTypeTok{nsim =} \NormalTok{nsim, }\DataTypeTok{newdata =} \NormalTok{newYear, }\DataTypeTok{unconditional =} \OtherTok{TRUE}\NormalTok{)}
-
-\NormalTok{## rearrange the output into a long/tidy format}
-\KeywordTok{colnames}\NormalTok{(sims) <-}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"sim"}\NormalTok{, }\KeywordTok{seq_len}\NormalTok{(nsim))}
-\NormalTok{sims <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{stack}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(sims)), }\KeywordTok{c}\NormalTok{(}\StringTok{"simulated"}\NormalTok{, }\StringTok{"run"}\NormalTok{))}
-\NormalTok{sims <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(sims, }\DataTypeTok{Year =} \KeywordTok{rep}\NormalTok{(newYear$Year, nsim),}
-                  \DataTypeTok{simulated =} \NormalTok{simulated)}
-
-\NormalTok{## Plot simulated trends}
-\NormalTok{smallSim.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(newYear, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{sims,}
-              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{simulated, }\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{group =} \NormalTok{run),}
-              \DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
+\CommentTok{## do the simulations}
+\NormalTok{sims <-}\StringTok{ }\KeywordTok{smooth_samples}\NormalTok{(mod}\OperatorTok{$}\NormalTok{gam, }\DataTypeTok{n =}\NormalTok{ nsim, }\DataTypeTok{data =}\NormalTok{ newYear,}
+    \DataTypeTok{unconditional =} \OtherTok{TRUE}\NormalTok{)}
+\NormalTok{sims <-}\StringTok{ }\KeywordTok{mutate}\NormalTok{(sims, }\DataTypeTok{value =}\NormalTok{ value }\OperatorTok{+}\StringTok{ }\KeywordTok{coef}\NormalTok{(mod}\OperatorTok{$}\NormalTok{gam)[1L]) }\CommentTok{# add intercept}
+
+\CommentTok{## Plot simulated trends}
+\NormalTok{smallSim.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(newYear, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ fit)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sims,}
+              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ value, }\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{group =}\NormalTok{ draw),}
+              \DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
 \NormalTok{smallSim.plt}
@@
 \begin{Highlighting}[]
-\NormalTok{## data points to simulate at}
+\CommentTok{## data points to simulate at}
 \NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya,}
                  \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
-                                       \DataTypeTok{length.out =} \NormalTok{N)))}
+                                       \DataTypeTok{length.out =}\NormalTok{ N)))}
 \NormalTok{braya_pred <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newBraya,}
@@
 
-\NormalTok{## simulate}
+\CommentTok{## simulate}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(braya_reml, }\DataTypeTok{nsim =} \NormalTok{nsim, }\DataTypeTok{newdata =} \NormalTok{newBraya,}
+\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{smooth_samples}\NormalTok{(braya_reml, }\DataTypeTok{n =}\NormalTok{ nsim, }\DataTypeTok{data =}\NormalTok{ newBraya,}
     \DataTypeTok{unconditional =} \OtherTok{TRUE}\NormalTok{)}
-
-\NormalTok{## rearrange the output into a long/tidy format}
-\KeywordTok{colnames}\NormalTok{(sims2) <-}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"sim"}\NormalTok{, }\KeywordTok{seq_len}\NormalTok{(nsim))}
-\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{stack}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(sims2)),}
-                  \KeywordTok{c}\NormalTok{(}\StringTok{"simulated"}\NormalTok{, }\StringTok{"run"}\NormalTok{))}
-\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(sims2, }\DataTypeTok{Year =} \KeywordTok{rep}\NormalTok{(newBraya$Year, nsim),}
-                   \DataTypeTok{simulated =} \NormalTok{simulated)}
-
-\NormalTok{brayaSim.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya_pred, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{sims2,}
-              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{simulated, }\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{group =} \NormalTok{run),}
-              \DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
+\NormalTok{sims2 <-}\StringTok{ }\KeywordTok{mutate}\NormalTok{(sims2, }\DataTypeTok{value =}\NormalTok{ value }\OperatorTok{+}\StringTok{ }\KeywordTok{coef}\NormalTok{(braya_reml)[1L]) }\CommentTok{# add intercept}
+
+\NormalTok{brayaSim.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya_pred, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ fit)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sims2,}
+              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ value, }\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{group =}\NormalTok{ draw),}
+              \DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
 \NormalTok{brayaSim.plt}
@@
 
+\hypertarget{confidence-and-simultaneous-intervals}{%
 \section{Confidence and simultaneous
-intervals}\label{confidence-and-simultaneous-intervals}
+intervals}\label{confidence-and-simultaneous-intervals}}
 
@@
 \begin{Highlighting}[]
-\NormalTok{sw.cint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(mod, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newYear,}
-                   \DataTypeTok{type =} \StringTok{"confidence"}\NormalTok{)}
-\NormalTok{sw.sint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(mod, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newYear,}
-                   \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{)}
-\KeywordTok{head}\NormalTok{(sw.sint)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-#>    smooth     Year      est        se     crit    lower    upper
-#> 1 s(Year) 1846.319 3.787215 0.1616144 3.072403 3.290671 4.283760
-#> 2 s(Year) 1847.128 3.789747 0.1509100 3.072403 3.326090 4.253403
-#> 3 s(Year) 1847.937 3.792304 0.1411439 3.072403 3.358653 4.225955
-#> 4 s(Year) 1848.745 3.794914 0.1324961 3.072403 3.387832 4.201995
-#> 5 s(Year) 1849.554 3.797602 0.1251070 3.072403 3.413222 4.181981
-#> 6 s(Year) 1850.363 3.800394 0.1190552 3.072403 3.434608 4.166180
+\NormalTok{sw.cint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(mod, }\DataTypeTok{parm =} \StringTok{"s(Year)"}\NormalTok{, }\DataTypeTok{data =}\NormalTok{ newYear, }\DataTypeTok{type =} \StringTok{"confidence"}\NormalTok{)}
+\NormalTok{sw.sint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(mod, }\DataTypeTok{parm =} \StringTok{"s(Year)"}\NormalTok{, }\DataTypeTok{data =}\NormalTok{ newYear, }\DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{)}
+\NormalTok{sw.sint}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#> # A tibble: 200 x 9
+#>    smooth  type  by     Year .estimate   .se .crit .lower_ci .upper_ci
+#>    <chr>   <chr> <chr> <dbl>     <dbl> <dbl> <dbl>     <dbl>     <dbl>
+#>  1 s(Year) TPRS  <NA>  1846.     0.478 0.163  2.92   0.00222     0.954
+#>  2 s(Year) TPRS  <NA>  1847.     0.481 0.152  2.92   0.0357      0.926
+#>  3 s(Year) TPRS  <NA>  1848.     0.483 0.143  2.92   0.0665      0.900
+#>  4 s(Year) TPRS  <NA>  1849.     0.486 0.134  2.92   0.0940      0.878
+#>  5 s(Year) TPRS  <NA>  1850.     0.489 0.127  2.92   0.118       0.859
+#>  6 s(Year) TPRS  <NA>  1850.     0.491 0.121  2.92   0.138       0.845
+#>  7 s(Year) TPRS  <NA>  1851.     0.494 0.116  2.92   0.154       0.834
+#>  8 s(Year) TPRS  <NA>  1852.     0.497 0.113  2.92   0.167       0.827
+#>  9 s(Year) TPRS  <NA>  1853.     0.501 0.111  2.92   0.178       0.824
+#> 10 s(Year) TPRS  <NA>  1854.     0.504 0.109  2.92   0.186       0.823
+#> # i 190 more rows
 \end{verbatim}
@@
 \begin{Highlighting}[]
-\NormalTok{smallInt.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(sw.cint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
-\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{data =} \NormalTok{sw.sint,}
-                \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year),}
-                \DataTypeTok{fill =} \StringTok{"grey80"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year),}
-                \DataTypeTok{fill =} \StringTok{"grey60"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
+\NormalTok{smallInt.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(sw.cint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ .estimate)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sw.sint,}
+                \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =}\NormalTok{ .lower_ci, }\DataTypeTok{ymax =}\NormalTok{ .upper_ci, }\DataTypeTok{x =}\NormalTok{ Year),}
+                \DataTypeTok{fill =} \StringTok{"grey80"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =}\NormalTok{ .lower_ci, }\DataTypeTok{ymax =}\NormalTok{ .upper_ci, }\DataTypeTok{x =}\NormalTok{ Year),}
+                \DataTypeTok{fill =} \StringTok{"grey60"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
 \NormalTok{smallInt.plt}
@@
 \begin{Highlighting}[]
-\NormalTok{bs.cint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(braya_reml, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newBraya,}
+\NormalTok{bs.cint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(braya_reml, }\DataTypeTok{parm =} \StringTok{"s(Year)"}\NormalTok{, }\DataTypeTok{data =}\NormalTok{ newBraya,}
                    \DataTypeTok{type =} \StringTok{"confidence"}\NormalTok{)}
-\NormalTok{bs.sint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(braya_reml, }\DataTypeTok{parm =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newBraya,}
+\NormalTok{bs.sint <-}\StringTok{ }\KeywordTok{confint}\NormalTok{(braya_reml, }\DataTypeTok{parm =} \StringTok{"s(Year)"}\NormalTok{, }\DataTypeTok{data =}\NormalTok{ newBraya,}
                    \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\NormalTok{brayaInt.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(bs.cint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
-\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{data =} \NormalTok{bs.sint,}
-                \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year),}
-                \DataTypeTok{fill =} \StringTok{"grey80"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year),}
-                \DataTypeTok{fill =} \StringTok{"grey60"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
+\NormalTok{brayaInt.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(bs.cint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ .estimate)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ bs.sint,}
+                \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =}\NormalTok{ .lower_ci, }\DataTypeTok{ymax =}\NormalTok{ .upper_ci, }\DataTypeTok{x =}\NormalTok{ Year),}
+                \DataTypeTok{fill =} \StringTok{"grey80"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =}\NormalTok{ .lower_ci, }\DataTypeTok{ymax =}\NormalTok{ .upper_ci, }\DataTypeTok{x =}\NormalTok{ Year),}
+                \DataTypeTok{fill =} \StringTok{"grey60"}\NormalTok{, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{lwd =} \DecValTok{1}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
 \NormalTok{brayaInt.plt}
@@
 
+\hypertarget{derivatives-of-the-estimated-trend}{%
 \section{Derivatives of the estimated
-trend}\label{derivatives-of-the-estimated-trend}
+trend}\label{derivatives-of-the-estimated-trend}}
 
@@
 \begin{Highlighting}[]
-\NormalTok{small.d <-}\StringTok{ }\KeywordTok{fderiv}\NormalTok{(mod, }\DataTypeTok{newdata =} \NormalTok{newYear, }\DataTypeTok{n =} \NormalTok{N)}
+\NormalTok{small.d <-}\StringTok{ }\KeywordTok{fderiv}\NormalTok{(mod, }\DataTypeTok{newdata =}\NormalTok{ newYear, }\DataTypeTok{n =}\NormalTok{ N)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#> Warning: `fderiv()` was deprecated in gratia 0.7.0.
+#> i Please use `derivatives()` instead.
+#> This warning is displayed once every 8 hours.
+#> Call `lifecycle::last_lifecycle_warnings()` to see where this warning was generated.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{small.sint <-}\StringTok{ }\KeywordTok{with}\NormalTok{(newYear,}
-                   \KeywordTok{cbind}\NormalTok{(}\KeywordTok{confint}\NormalTok{(small.d, }\DataTypeTok{nsim =} \NormalTok{nsim,}
+                   \KeywordTok{cbind}\NormalTok{(}\KeywordTok{confint}\NormalTok{(small.d, }\DataTypeTok{nsim =}\NormalTok{ nsim,}
                                  \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{),}
-                         \DataTypeTok{Year =} \NormalTok{Year))}
-
-\NormalTok{small_deriv_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(small.sint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
-\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper), }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{,}
-                \DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
+                         \DataTypeTok{Year =}\NormalTok{ Year))}
+
+\NormalTok{small_deriv_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(small.sint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ est)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =}\NormalTok{ lower, }\DataTypeTok{ymax =}\NormalTok{ upper), }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{,}
+                \DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{() }\OperatorTok{+}
 \StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"First derivative"}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\NormalTok{braya.d <-}\StringTok{ }\KeywordTok{fderiv}\NormalTok{(braya_reml, }\DataTypeTok{newdata =} \NormalTok{newBraya, }\DataTypeTok{n =} \NormalTok{N)}
+\NormalTok{braya.d <-}\StringTok{ }\KeywordTok{fderiv}\NormalTok{(braya_reml, }\DataTypeTok{newdata =}\NormalTok{ newBraya, }\DataTypeTok{n =}\NormalTok{ N)}
 \NormalTok{braya.sint <-}\StringTok{ }\KeywordTok{with}\NormalTok{(newBraya,}
-                   \KeywordTok{cbind}\NormalTok{(}\KeywordTok{confint}\NormalTok{(braya.d, }\DataTypeTok{nsim =} \NormalTok{nsim,}
+                   \KeywordTok{cbind}\NormalTok{(}\KeywordTok{confint}\NormalTok{(braya.d, }\DataTypeTok{nsim =}\NormalTok{ nsim,}
                                  \DataTypeTok{type =} \StringTok{"simultaneous"}\NormalTok{),}
-                         \DataTypeTok{Year =} \NormalTok{Year))}
-
-\NormalTok{braya_deriv_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya.sint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{est)) +}
-\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper),}
-                \DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
+                         \DataTypeTok{Year =}\NormalTok{ Year))}
+
+\NormalTok{braya_deriv_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya.sint, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ est)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =}\NormalTok{ lower, }\DataTypeTok{ymax =}\NormalTok{ upper),}
+                \DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{() }\OperatorTok{+}
 \StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{, }\DataTypeTok{y =} \StringTok{"First derivative"}\NormalTok{)}
@@
 
-\section{Gaussian process smooths}\label{gaussian-process-smooths}
+\hypertarget{gaussian-process-smooths}{%
+\section{Gaussian process smooths}\label{gaussian-process-smooths}}
 
@@
 \NormalTok{nn <-}\StringTok{ }\DecValTok{200}    \CommentTok{# number of points at which to evaluate profile likelihood}
-\NormalTok{dseq <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{15}\NormalTok{, }\DecValTok{500}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{nn)  }\CommentTok{# effective ranges to fit at}
-\NormalTok{Mat <-}\StringTok{ }\NormalTok{SEx <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DataTypeTok{length =} \NormalTok{nn)     }\CommentTok{# object to hold model fits}
-
-\NormalTok{## loop over the sequence of deparation distances and fit the models}
-\NormalTok{for (i in }\KeywordTok{seq_along}\NormalTok{(dseq)) \{ }
-    \NormalTok{## iterate over dseq, fit GP GAM w Matérn covariance}
-    \NormalTok{Mat[i] <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, dseq[i])),}
-                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval),}
-                  \DataTypeTok{data =} \NormalTok{braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
+\NormalTok{dseq <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{15}\NormalTok{, }\DecValTok{500}\NormalTok{, }\DataTypeTok{length.out =}\NormalTok{ nn)  }\CommentTok{# effective ranges to fit at}
+\NormalTok{Mat <-}\StringTok{ }\NormalTok{SEx <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DataTypeTok{length =}\NormalTok{ nn)     }\CommentTok{# object to hold model fits}
+
+\CommentTok{## loop over the sequence of deparation distances and fit the models}
+\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \KeywordTok{seq_along}\NormalTok{(dseq)) \{ }
+    \CommentTok{## iterate over dseq, fit GP GAM w Matérn covariance}
+\NormalTok{    Mat[i] <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, dseq[i])),}
+                  \DataTypeTok{weights =}\NormalTok{ sampleInterval }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval),}
+                  \DataTypeTok{data =}\NormalTok{ braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                   \DataTypeTok{family =} \KeywordTok{gaussian}\NormalTok{())[[}\StringTok{"gcv.ubre"}\NormalTok{]]}
-    \NormalTok{## fit squared exponential}
-    \NormalTok{SEx[i] <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, dseq[i], }\DecValTok{1}\NormalTok{)),}
-                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval),}
-                  \DataTypeTok{data =} \NormalTok{braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
+    \CommentTok{## fit squared exponential}
+\NormalTok{    SEx[i] <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, dseq[i], }\DecValTok{1}\NormalTok{)),}
+                  \DataTypeTok{weights =}\NormalTok{ sampleInterval }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval),}
+                  \DataTypeTok{data =}\NormalTok{ braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                   \DataTypeTok{family =} \KeywordTok{gaussian}\NormalTok{())[[}\StringTok{"gcv.ubre"}\NormalTok{]]}
@@
 
-\NormalTok{## extract the REML score into ggplot-friendly object}
-\NormalTok{reml.scr <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{cor =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"Matérn"}\NormalTok{,}\StringTok{"Exponential"}\NormalTok{), }\DataTypeTok{each =} \NormalTok{nn),}
+\CommentTok{## extract the REML score into ggplot-friendly object}
+\NormalTok{reml.scr <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{cor =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"Matérn"}\NormalTok{,}\StringTok{"Exponential"}\NormalTok{), }\DataTypeTok{each =}\NormalTok{ nn),}
                        \DataTypeTok{effrange =} \KeywordTok{rep}\NormalTok{(dseq, }\DecValTok{2}\NormalTok{),}
@@
 \begin{Highlighting}[]
-\NormalTok{## profile-likelihood plot}
-\NormalTok{proflik.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(reml.scr, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{effrange, }\DataTypeTok{y =} \NormalTok{reml, }\DataTypeTok{colour =} \NormalTok{cor)) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
-\StringTok{    }\KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{""}\NormalTok{, }\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#e66101"}\NormalTok{,}\StringTok{"#5e3c99"}\NormalTok{)) +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \StringTok{"REML score"}\NormalTok{, }\DataTypeTok{x =} \KeywordTok{expression}\NormalTok{(Effective ~}\StringTok{ }\NormalTok{range ~}\StringTok{ }\NormalTok{(varphi)))}
+\CommentTok{## profile-likelihood plot}
+\NormalTok{proflik.plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(reml.scr, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ effrange, }\DataTypeTok{y =}\NormalTok{ reml, }\DataTypeTok{colour =}\NormalTok{ cor)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{() }\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{""}\NormalTok{, }\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#e66101"}\NormalTok{,}\StringTok{"#5e3c99"}\NormalTok{)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \StringTok{"REML score"}\NormalTok{, }\DataTypeTok{x =} \KeywordTok{expression}\NormalTok{(Effective }\OperatorTok{~}\StringTok{ }\NormalTok{range }\OperatorTok{~}\StringTok{ }\NormalTok{(varphi)))}
 \NormalTok{proflik.plt}
@@
 \begin{Highlighting}[]
-\NormalTok{## effective range minima from profile likelihood}
-\NormalTok{effRange2 <-}\StringTok{ }\KeywordTok{with}\NormalTok{(}\KeywordTok{subset}\NormalTok{(reml.scr, cor ==}\StringTok{ "Matérn"}\NormalTok{), dseq[}\KeywordTok{which.min}\NormalTok{(reml)])}
-\NormalTok{effRange3 <-}\StringTok{ }\KeywordTok{with}\NormalTok{(}\KeywordTok{subset}\NormalTok{(reml.scr, cor ==}\StringTok{ "Exponential"}\NormalTok{), dseq[}\KeywordTok{which.min}\NormalTok{(reml)])}
-
-\NormalTok{## Refit these models: Matern}
-\NormalTok{gp2 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, effRange2)),}
-           \DataTypeTok{data =} \NormalTok{braya,}
-           \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{, }\DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
-\NormalTok{## Refit these models: Power exponential}
-\NormalTok{gp3 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, effRange3, }\DecValTok{1}\NormalTok{)),}
-           \DataTypeTok{data =} \NormalTok{braya,}
-           \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{, }\DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
-
-\NormalTok{## data to predict at}
+\CommentTok{## effective range minima from profile likelihood}
+\NormalTok{effRange2 <-}\StringTok{ }\KeywordTok{with}\NormalTok{(}\KeywordTok{subset}\NormalTok{(reml.scr, cor }\OperatorTok{==}\StringTok{ "Matérn"}\NormalTok{), dseq[}\KeywordTok{which.min}\NormalTok{(reml)])}
+\NormalTok{effRange3 <-}\StringTok{ }\KeywordTok{with}\NormalTok{(}\KeywordTok{subset}\NormalTok{(reml.scr, cor }\OperatorTok{==}\StringTok{ "Exponential"}\NormalTok{), dseq[}\KeywordTok{which.min}\NormalTok{(reml)])}
+
+\CommentTok{## Refit these models: Matern}
+\NormalTok{gp2 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, effRange2)),}
+           \DataTypeTok{data =}\NormalTok{ braya,}
+           \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{, }\DataTypeTok{weights =}\NormalTok{ sampleInterval }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
+\CommentTok{## Refit these models: Power exponential}
+\NormalTok{gp3 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, effRange3, }\DecValTok{1}\NormalTok{)),}
+           \DataTypeTok{data =}\NormalTok{ braya,}
+           \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{, }\DataTypeTok{weights =}\NormalTok{ sampleInterval }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
+
+\CommentTok{## data to predict at}
 \NormalTok{newd <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                \DataTypeTok{length.out =} \DecValTok{1000}\NormalTok{)))}
-\NormalTok{## create predictions on response scale for both covariance functions}
+\CommentTok{## create predictions on response scale for both covariance functions}
 \NormalTok{p.gp2 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newd,}
-                   \DataTypeTok{fitted =} \KeywordTok{predict}\NormalTok{(gp2, }\DataTypeTok{newdata =} \NormalTok{newd, }\DataTypeTok{type =} \StringTok{"response"}\NormalTok{),}
+                   \DataTypeTok{fitted =} \KeywordTok{predict}\NormalTok{(gp2, }\DataTypeTok{newdata =}\NormalTok{ newd, }\DataTypeTok{type =} \StringTok{"response"}\NormalTok{),}
                    \DataTypeTok{effRange =} \KeywordTok{round}\NormalTok{(effRange2))}
 \NormalTok{p.gp3 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newd,}
-                   \DataTypeTok{fitted =} \KeywordTok{predict}\NormalTok{(gp3, }\DataTypeTok{newdata =} \NormalTok{newd, }\DataTypeTok{type =} \StringTok{"response"}\NormalTok{),}
+                   \DataTypeTok{fitted =} \KeywordTok{predict}\NormalTok{(gp3, }\DataTypeTok{newdata =}\NormalTok{ newd, }\DataTypeTok{type =} \StringTok{"response"}\NormalTok{),}
                    \DataTypeTok{effRange =} \KeywordTok{round}\NormalTok{(effRange3))}
-\NormalTok{## joint the two sets of predictions together}
+\CommentTok{## joint the two sets of predictions together}
 \NormalTok{pred <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(p.gp2, p.gp3)}
-\NormalTok{## add some categorical variables for plotting}
+\CommentTok{## add some categorical variables for plotting}
 \NormalTok{pred <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(pred,}
@@
 \begin{Highlighting}[]
-\NormalTok{gp.plt2 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(pred, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fitted, }\DataTypeTok{colour =} \NormalTok{cor)) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}\StringTok{ }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"right"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37), }\DataTypeTok{data =} \NormalTok{braya, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
-\StringTok{    }\KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{""}\NormalTok{, }\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#e66101"}\NormalTok{,}\StringTok{"#5e3c99"}\NormalTok{)) +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
+\NormalTok{gp.plt2 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(pred, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ fitted, }\DataTypeTok{colour =}\NormalTok{ cor)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{() }\OperatorTok{+}\StringTok{ }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"right"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ UK37), }\DataTypeTok{data =}\NormalTok{ braya, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{""}\NormalTok{, }\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"#e66101"}\NormalTok{,}\StringTok{"#5e3c99"}\NormalTok{)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
 \NormalTok{gp.plt2}
@@
 
-\section{Adaptive smooths}\label{adaptive-smooths}
+\hypertarget{adaptive-smooths}{%
+\section{Adaptive smooths}\label{adaptive-smooths}}
 
@@
 \begin{Highlighting}[]
-\NormalTok{## Adaptive spline, weights as sampleInterval}
-\NormalTok{mod_ad <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"ad"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
+\CommentTok{## Adaptive spline, weights as sampleInterval}
+\NormalTok{mod_ad <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"ad"}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ braya,}
               \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
-              \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
-\end{Highlighting}
-\end{Shaded}
-
-\section{Comparing trends}\label{comparing-trends}
+              \DataTypeTok{weights =}\NormalTok{ sampleInterval }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{comparing-trends}{%
+\section{Comparing trends}\label{comparing-trends}}
 
@@
 \begin{Highlighting}[]
-\NormalTok{## effective range parameter from profile-likelihood of Matern model}
+\CommentTok{## effective range parameter from profile-likelihood of Matern model}
 \NormalTok{effRange <-}\StringTok{ }\NormalTok{effRange2}
 
-\NormalTok{## TPRS, weights as sampleInterval}
-\NormalTok{mod_tprs <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"tp"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
+\CommentTok{## TPRS, weights as sampleInterval}
+\NormalTok{mod_tprs <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"tp"}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ braya,}
                 \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
-                \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
-
-\NormalTok{## Gaussian process, Matern, kappa = 1.5, weights as sampleInterval}
-\NormalTok{mod_gp <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, effRange)),}
-              \DataTypeTok{data =} \NormalTok{braya,}
+                \DataTypeTok{weights =}\NormalTok{ sampleInterval }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
+
+\CommentTok{## Gaussian process, Matern, kappa = 1.5, weights as sampleInterval}
+\NormalTok{mod_gp <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"gp"}\NormalTok{, }\DataTypeTok{m =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, effRange)),}
+              \DataTypeTok{data =}\NormalTok{ braya,}
               \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
-              \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
-
-\NormalTok{## Adaptive spline, weights as sampleInterval}
-\NormalTok{mod_ad <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"ad"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
+              \DataTypeTok{weights =}\NormalTok{ sampleInterval }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
+
+\CommentTok{## Adaptive spline, weights as sampleInterval}
+\NormalTok{mod_ad <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"ad"}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ braya,}
               \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
-              \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
+              \DataTypeTok{weights =}\NormalTok{ sampleInterval }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{## wrap this in a function that will return all the plots & derived objects}
-\NormalTok{processGAM <-}\StringTok{ }\NormalTok{function(mod) \{}
-    \NormalTok{## Predict from model}
-    \NormalTok{N <-}\StringTok{ }\DecValTok{500}
-    \NormalTok{newYear <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya,}
+\CommentTok{## wrap this in a function that will return all the plots & derived objects}
+\NormalTok{processGAM <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(mod) \{}
+    \CommentTok{## Predict from model}
+\NormalTok{    N <-}\StringTok{ }\DecValTok{500}
+\NormalTok{    newYear <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya,}
                     \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
-                                          \DataTypeTok{length.out =} \NormalTok{N)))}
-    \NormalTok{newYear <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newYear,}
+                                          \DataTypeTok{length.out =}\NormalTok{ N)))}
+\NormalTok{    newYear <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newYear,}
                      \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(mod, newYear, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
 
-    \NormalTok{out <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{objects =} \NormalTok{newYear)}
-    \NormalTok{out}
+\NormalTok{    out <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{objects =}\NormalTok{ newYear)}
+\NormalTok{    out}
 \NormalTok{\}}
 
-\NormalTok{plts_gp   <-}\StringTok{ }\KeywordTok{processGAM}\NormalTok{(}\DataTypeTok{mod =} \NormalTok{mod_gp) }\CommentTok{# Gaussian process smooth with weights}
-\NormalTok{plts_ad   <-}\StringTok{ }\KeywordTok{processGAM}\NormalTok{(}\DataTypeTok{mod =} \NormalTok{mod_ad) }\CommentTok{# Adaptive smooth with weights}
-\NormalTok{plts_tprs <-}\StringTok{ }\KeywordTok{processGAM}\NormalTok{(}\DataTypeTok{mod =} \NormalTok{mod_tprs) }\CommentTok{# TPRS with weights}
+\NormalTok{plts_gp   <-}\StringTok{ }\KeywordTok{processGAM}\NormalTok{(}\DataTypeTok{mod =}\NormalTok{ mod_gp) }\CommentTok{# Gaussian process smooth with weights}
+\NormalTok{plts_ad   <-}\StringTok{ }\KeywordTok{processGAM}\NormalTok{(}\DataTypeTok{mod =}\NormalTok{ mod_ad) }\CommentTok{# Adaptive smooth with weights}
+\NormalTok{plts_tprs <-}\StringTok{ }\KeywordTok{processGAM}\NormalTok{(}\DataTypeTok{mod =}\NormalTok{ mod_tprs) }\CommentTok{# TPRS with weights}
 
@@
                      \DataTypeTok{Model =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GP"}\NormalTok{, }\StringTok{"Adaptive"}\NormalTok{, }\StringTok{"TPRS"}\NormalTok{),}
-                                 \DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(plts_gp$objects)))}
-
-\NormalTok{allFits <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(pltData, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37), }\DataTypeTok{data =} \NormalTok{braya) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{colour =} \NormalTok{Model)) +}\StringTok{ }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"right"}\NormalTok{) +}
+                                 \DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(plts_gp}\OperatorTok{$}\NormalTok{objects)))}
+
+\NormalTok{allFits <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(pltData, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ fit)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ UK37), }\DataTypeTok{data =}\NormalTok{ braya) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{colour =}\NormalTok{ Model)) }\OperatorTok{+}\StringTok{ }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"right"}\NormalTok{) }\OperatorTok{+}
 \StringTok{    }\KeywordTok{scale_colour_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{""}\NormalTok{,}
@@
 
+\hypertarget{accounting-for-age-model-uncertainty}{%
 \section{Accounting for age-model
-uncertainty}\label{accounting-for-age-model-uncertainty}
+uncertainty}\label{accounting-for-age-model-uncertainty}}
 
@@
 \NormalTok{knots <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{list}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year), }\DataTypeTok{length =} \DecValTok{14}\NormalTok{)))}
-\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
-            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year),}
-            \DataTypeTok{knots =} \NormalTok{knots)}
+\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ small, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
+            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \OperatorTok{~}\StringTok{ }\NormalTok{Year),}
+            \DataTypeTok{knots =}\NormalTok{ knots)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{## monotonic spline age-depth model}
-\NormalTok{swAge$Error[}\DecValTok{1}\NormalTok{] <-}\StringTok{ }\FloatTok{1.1}
-\NormalTok{swAgeMod <-}\StringTok{ }\KeywordTok{scam}\NormalTok{(Date ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Depth, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"mpd"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{swAge,}
-                 \DataTypeTok{weights =} \DecValTok{1} \NormalTok{/}\StringTok{ }\NormalTok{swAge$Error, }\DataTypeTok{gamma =} \FloatTok{1.4}\NormalTok{)}
+\CommentTok{## monotonic spline age-depth model}
+\NormalTok{swAge}\OperatorTok{$}\NormalTok{Error[}\DecValTok{1}\NormalTok{] <-}\StringTok{ }\DecValTok{1}
+\NormalTok{swAgeMod <-}\StringTok{ }\KeywordTok{scam}\NormalTok{(Date }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Depth, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"cv"}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ swAge,}
+    \DataTypeTok{weights =}\NormalTok{ (}\DecValTok{1} \OperatorTok{/}\StringTok{ }\NormalTok{swAge}\OperatorTok{$}\NormalTok{Error) }\OperatorTok{/}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\DecValTok{1} \OperatorTok{/}\StringTok{ }\NormalTok{swAge}\OperatorTok{$}\NormalTok{Error), }\DataTypeTok{gamma =} \DecValTok{1}\NormalTok{)}
 \end{Highlighting}
@@
 Next, predict from the estimated age model, and draw 25 samples from the
-posterior distribution using \texttt{simulate()}. The results are tidied
-into a format suitable for further processing and plotting. Note that
-the posterior samples here are only used for plotting.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{## predict from the age model for a smooth set of points in `Depth`}
-\NormalTok{newAge <-}\StringTok{ }\KeywordTok{with}\NormalTok{(swAge, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Depth =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Depth), }\KeywordTok{max}\NormalTok{(Depth),}
-                                             \DataTypeTok{length.out =} \DecValTok{200}\NormalTok{)))}
-\NormalTok{newAge <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newAge,}
-                    \DataTypeTok{fitted =} \KeywordTok{predict}\NormalTok{(swAgeMod, }\DataTypeTok{newdata =} \NormalTok{newAge, }
-                                     \DataTypeTok{type =} \StringTok{"response"}\NormalTok{))}
-\NormalTok{newSims <-}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{simulate}\NormalTok{(swAgeMod, }\DataTypeTok{nsim =} \DecValTok{25}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{newAge))}
-\NormalTok{newSims <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(}\DataTypeTok{Depth =} \NormalTok{newAge$Depth, newSims)}
-\NormalTok{newSims <-}\StringTok{ }\KeywordTok{gather}\NormalTok{(newSims, Simulation, Age, -Depth)}
+posterior distribution using \texttt{smooth\_samples()}. Note that the
+posterior samples here are only used for plotting.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## predict from the age model for a smooth set of points in `Depth`}
+\NormalTok{newAge <-}\StringTok{ }\KeywordTok{data_slice}\NormalTok{(swAgeMod, }\DataTypeTok{Depth =} \KeywordTok{evenly}\NormalTok{(Depth, }\DataTypeTok{n =} \DecValTok{200}\NormalTok{))}
+\NormalTok{newAgeFit <-}\StringTok{ }\KeywordTok{fitted_values}\NormalTok{(swAgeMod, }\DataTypeTok{data =}\NormalTok{ newAge)}
+\NormalTok{newSims <-}\StringTok{ }\KeywordTok{smooth_samples}\NormalTok{(swAgeMod, }\DataTypeTok{n =} \DecValTok{25}\NormalTok{, }\DataTypeTok{data =}\NormalTok{ newAge) }\OperatorTok{%>%}
+\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{value =}\NormalTok{ value }\OperatorTok{+}\StringTok{ }\KeywordTok{coef}\NormalTok{(swAgeMod)[1L]) }\CommentTok{# %>%}
+    \CommentTok{#left_join(mutate(newAge, .row = row_number()), by = join_by(.row == .row))}
 \end{Highlighting}
@@
 distribution of the age model, but notice that I pass in the
-\texttt{small} data to \texttt{newdata} in the call to
-\texttt{simulate()} as the locations I want new age estimates for are
+\texttt{small} data to \texttt{data} in the call to
+\texttt{smooth\_samples()} as the locations I want new age estimates for
 the depths for which we have δ\textsuperscript{15}N values. A small
@@
 \begin{Highlighting}[]
-\NormalTok{## simulate from age model; each column is a simulation}
-\NormalTok{ageSims <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(swAgeMod, }\DataTypeTok{nsim =} \DecValTok{100}\NormalTok{, }\DataTypeTok{newdata =} \NormalTok{small, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{)}
-\NormalTok{ageSims <-}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(ageSims)}
-
-\NormalTok{fitSWModels <-}\StringTok{ }\NormalTok{function(x, y, knots) \{}
-    \NormalTok{dat <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{d15N =} \NormalTok{y, }\DataTypeTok{Year =} \NormalTok{x)}
-    \NormalTok{m <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{dat, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
-              \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{knots =} \NormalTok{knots)}
+\CommentTok{## simulate from age model; each column is a simulation}
+\NormalTok{ageSims <-}\StringTok{ }\KeywordTok{smooth_samples}\NormalTok{(swAgeMod, }\DataTypeTok{n =} \DecValTok{100}\NormalTok{, }\DataTypeTok{data =}\NormalTok{ small, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{)}
+
+\NormalTok{fitSWModels <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x, y, knots) \{}
+\NormalTok{    dat <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{d15N =}\NormalTok{ y,}
+        \DataTypeTok{Year =} \KeywordTok{pull}\NormalTok{(x, value) }\OperatorTok{+}\StringTok{ }\KeywordTok{coef}\NormalTok{(swAgeMod, }\DataTypeTok{parameterized =} \OtherTok{TRUE}\NormalTok{)[1L])}
+\NormalTok{    m <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ dat, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
+              \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \OperatorTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{knots =}\NormalTok{ knots)}
 \NormalTok{\}}
 
-\NormalTok{## generate new trends using draws from age-model posterior}
-\NormalTok{simTrendMods <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(ageSims, fitSWModels, }\DataTypeTok{y =} \NormalTok{small$d15N, }\DataTypeTok{knots =} \NormalTok{knots)}
-
-\NormalTok{## function wrapper to predict new trends at locations over the}
-\NormalTok{## range of `Year`}
-\NormalTok{predSWModels <-}\StringTok{ }\NormalTok{function(mod, newdata) \{}
-    \KeywordTok{predict}\NormalTok{(mod$gam, }\DataTypeTok{newdata =} \NormalTok{newdata, }\DataTypeTok{type =} \StringTok{"response"}\NormalTok{)}
+\CommentTok{## generate new trends using draws from age-model posterior}
+\NormalTok{simTrendMods <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(ageSims }\OperatorTok{%>%}\StringTok{ }\KeywordTok{split}\NormalTok{(ageSims}\OperatorTok{$}\NormalTok{draw), fitSWModels,}
+    \DataTypeTok{y =}\NormalTok{ small}\OperatorTok{$}\NormalTok{d15N, }\DataTypeTok{knots =}\NormalTok{ knots)}
+
+\CommentTok{## function wrapper to predict new trends at locations over the}
+\CommentTok{## range of `Year`}
+\NormalTok{predSWModels <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(mod, newdata) \{}
+    \KeywordTok{predict}\NormalTok{(mod}\OperatorTok{$}\NormalTok{gam, }\DataTypeTok{newdata =}\NormalTok{ newdata, }\DataTypeTok{type =} \StringTok{"response"}\NormalTok{)}
 \NormalTok{\}}
 
-\NormalTok{## predict from fitted model to produce a smooth trend for each posterior}
-\NormalTok{## sample}
-\NormalTok{simTrends <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(simTrendMods, predSWModels, }\DataTypeTok{newdata =} \NormalTok{newYear)}
-
-\NormalTok{## arrange in a tidy format form plottings}
+\CommentTok{## predict from fitted model to produce a smooth trend for each posterior}
+\CommentTok{## sample}
+\NormalTok{simTrends <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(simTrendMods, predSWModels, }\DataTypeTok{newdata =}\NormalTok{ newYear)}
+
+\CommentTok{## arrange in a tidy format form plottings}
 \NormalTok{simTrends <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year  =} \KeywordTok{with}\NormalTok{(newYear, }\KeywordTok{rep}\NormalTok{(Year, }\KeywordTok{length}\NormalTok{(simTrends))),}
@@
 \begin{Highlighting}[]
-\NormalTok{## wrapper to simulate from a fitted GAM with the required arguments}
-\NormalTok{simulateSWModels <-}\StringTok{ }\NormalTok{function(mod, newdata, nsim, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{) \{}
-    \NormalTok{sims <-}\StringTok{ }\KeywordTok{simulate}\NormalTok{(mod, }\DataTypeTok{nsim =} \NormalTok{nsim, }\DataTypeTok{newdata =} \NormalTok{newdata, }\DataTypeTok{seed =} \NormalTok{seed)}
-    \KeywordTok{as.vector}\NormalTok{(sims)}
+\CommentTok{## wrapper to simulate from a fitted GAM with the required arguments}
+\NormalTok{simulateSWModels <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(mod, newdata, nsim, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{) \{}
+\NormalTok{    sims <-}\StringTok{ }\KeywordTok{fitted_samples}\NormalTok{(mod, }\DataTypeTok{n =}\NormalTok{ nsim, }\DataTypeTok{data =}\NormalTok{ newdata, }\DataTypeTok{seed =}\NormalTok{ seed)}
+    \KeywordTok{pull}\NormalTok{(sims, .fitted)}
 \NormalTok{\}}
 
-\NormalTok{## now do the posterior simulation}
+\CommentTok{## now do the posterior simulation}
 \NormalTok{NSIM <-}\StringTok{ }\DecValTok{50}     \CommentTok{# number of posterior samples *per* model}
-\NormalTok{simSimulate <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(simTrendMods, simulateSWModels, }\DataTypeTok{newdata =} \NormalTok{newYear,}
-                      \DataTypeTok{nsim =} \NormalTok{NSIM, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{)}
-
-\NormalTok{## arrange in a tidy format}
+\NormalTok{simSimulate <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(simTrendMods, simulateSWModels, }\DataTypeTok{newdata =}\NormalTok{ newYear,}
+                      \DataTypeTok{nsim =}\NormalTok{ NSIM, }\DataTypeTok{seed =} \DecValTok{42}\NormalTok{)}
+
+\CommentTok{## arrange in a tidy format}
 \NormalTok{simSimulate <-}
 \StringTok{  }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year  =} \KeywordTok{with}\NormalTok{(newYear,}
-                          \KeywordTok{rep}\NormalTok{(Year, }\DataTypeTok{times =} \NormalTok{NSIM *}\StringTok{ }\KeywordTok{length}\NormalTok{(simSimulate))),}
+                          \KeywordTok{rep}\NormalTok{(Year, }\DataTypeTok{times =}\NormalTok{ NSIM }\OperatorTok{*}\StringTok{ }\KeywordTok{length}\NormalTok{(simSimulate))),}
              \DataTypeTok{Trend =} \KeywordTok{unlist}\NormalTok{(simSimulate),}
-             \DataTypeTok{Group =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{seq_len}\NormalTok{(NSIM *}\StringTok{ }\KeywordTok{length}\NormalTok{(simSimulate)),}
+             \DataTypeTok{Group =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{seq_len}\NormalTok{(NSIM }\OperatorTok{*}\StringTok{ }\KeywordTok{length}\NormalTok{(simSimulate)),}
                          \DataTypeTok{each =} \KeywordTok{nrow}\NormalTok{(newYear)))}
@@
 \begin{Highlighting}[]
-\NormalTok{## plot the estimated age model ad posterior simulations from it}
-\NormalTok{plt1 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(swAge, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{Date, }\DataTypeTok{x =} \NormalTok{Depth)) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newSims,}
-              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{Age, }\DataTypeTok{x =} \NormalTok{Depth, }\DataTypeTok{group =} \NormalTok{Simulation),}
-              \DataTypeTok{alpha =} \DecValTok{1}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newAge, }\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \NormalTok{fitted, }\DataTypeTok{x =} \NormalTok{Depth)) +}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{size =} \FloatTok{1.5}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"red"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_errorbar}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{Date -}\StringTok{ }\NormalTok{Error, }\DataTypeTok{ymax =} \NormalTok{Date +}\StringTok{ }\NormalTok{Error, }\DataTypeTok{width =} \DecValTok{0}\NormalTok{),}
-                  \DataTypeTok{colour =} \StringTok{"red"}\NormalTok{) +}
+\CommentTok{## plot the estimated age model ad posterior simulations from it}
+\NormalTok{plt1 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(swAge, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ Date, }\DataTypeTok{x =}\NormalTok{ Depth)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ newSims,}
+              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ value, }\DataTypeTok{x =}\NormalTok{ Depth, }\DataTypeTok{group =}\NormalTok{ draw),}
+              \DataTypeTok{alpha =} \DecValTok{1}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ newAgeFit, }\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ .fitted, }\DataTypeTok{x =}\NormalTok{ Depth)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{size =} \FloatTok{1.5}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"red"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_errorbar}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =}\NormalTok{ Date }\OperatorTok{-}\StringTok{ }\NormalTok{Error, }\DataTypeTok{ymax =}\NormalTok{ Date }\OperatorTok{+}\StringTok{ }\NormalTok{Error, }\DataTypeTok{width =} \DecValTok{0}\NormalTok{),}
+                  \DataTypeTok{colour =} \StringTok{"red"}\NormalTok{) }\OperatorTok{+}
 \StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \StringTok{"Year CE"}\NormalTok{, }\DataTypeTok{x =} \StringTok{"Depth (cm)"}\NormalTok{)}
 
-\NormalTok{## plot the simulated trends showing the effect of age-model uncertainty}
-\NormalTok{plt2 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(simTrends, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{Trend, }\DataTypeTok{group =} \NormalTok{Group)) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.1}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newYear,}
-              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit), }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =} \NormalTok{small,}
-               \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N),}
-               \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{size =} \FloatTok{0.7}\NormalTok{) +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{y =} \NormalTok{d15n_label)}
-
-\NormalTok{## plot simulated trends showing the effect of age-model uncertainty and}
-\NormalTok{## the effect of uncertainty in the estimated trend itself}
-\NormalTok{plt3 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(simSimulate, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{Trend, }\DataTypeTok{group =} \NormalTok{Group)) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =} \NormalTok{small,}
-               \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N),}
+\CommentTok{## plot the simulated trends showing the effect of age-model uncertainty}
+\NormalTok{plt2 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(simTrends, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ Trend, }\DataTypeTok{group =}\NormalTok{ Group)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.1}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ newYear,}
+              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ fit), }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ small,}
+               \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ d15N),}
+               \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{size =} \FloatTok{0.7}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{y =}\NormalTok{ d15n_label)}
+
+\CommentTok{## plot simulated trends showing the effect of age-model uncertainty and}
+\CommentTok{## the effect of uncertainty in the estimated trend itself}
+\NormalTok{plt3 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(simSimulate, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ Trend, }\DataTypeTok{group =}\NormalTok{ Group)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"grey80"}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ small,}
+               \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ d15N),}
                \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{,}
-               \DataTypeTok{size =} \FloatTok{0.7}\NormalTok{) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =} \NormalTok{newYear,}
-              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit),}
-              \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
-\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{y =} \NormalTok{d15n_label)}
-
-\NormalTok{## align all plots vertically}
+               \DataTypeTok{size =} \FloatTok{0.7}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ newYear,}
+              \DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ fit),}
+              \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x =} \StringTok{"Year"}\NormalTok{, }\DataTypeTok{y =}\NormalTok{ d15n_label)}
+
+\CommentTok{## align all plots vertically}
 \NormalTok{plots <-}\StringTok{ }\KeywordTok{align_plots}\NormalTok{(plt1, plt2, plt3, }\DataTypeTok{align =} \StringTok{'v'}\NormalTok{, }\DataTypeTok{axis =} \StringTok{'l'}\NormalTok{)}
 
-\NormalTok{## create the two rows of figures, from `plots`}
+\CommentTok{## create the two rows of figures, from `plots`}
 \NormalTok{top_row <-}\StringTok{ }\KeywordTok{plot_grid}\NormalTok{(plots[[}\DecValTok{1}\NormalTok{]], }\OtherTok{NULL}\NormalTok{, }\DataTypeTok{ncol =} \DecValTok{2}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"a"}\NormalTok{)}
@@
 
-\NormalTok{## combine the two rows, top row has 1 plot row, bottom row has 2, hence}
-\NormalTok{## the rel_heights to even this out}
+\CommentTok{## combine the two rows, top row has 1 plot row, bottom row has 2, hence}
+\CommentTok{## the rel_heights to even this out}
 \KeywordTok{plot_grid}\NormalTok{(top_row, bot_row, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{rel_heights =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\DecValTok{1}\NormalTok{))}
@@
 
-\section{Session information}\label{session-information}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{devtools::}\KeywordTok{session_info}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-#> Session info -------------------------------------------------------------
-\end{verbatim}
-
-\begin{verbatim}
+\hypertarget{session-information}{%
+\section{Session information}\label{session-information}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{devtools}\OperatorTok{::}\KeywordTok{session_info}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+#> - Session info -----------------------------------------------------------------------------------
 #>  setting  value
-#>  version  R version 3.4.3 Patched (2018-01-16 r74126)
+#>  version  R version 4.3.1 (2023-06-16)
+#>  os       Ubuntu 20.04.6 LTS
 #>  system   x86_64, linux-gnu
 #>  ui       X11
-#>  language (EN)                                       
-#>  collate  en_CA.UTF-8                                
-#>  tz       America/Regina                             
-#>  date     2018-08-16
-\end{verbatim}
-
-\begin{verbatim}
-#> Packages -----------------------------------------------------------------
-\end{verbatim}
-
-\begin{verbatim}
-#>  package    * version date       source        
-#>  assertthat   0.2.0   2017-04-11 CRAN (R 3.4.3)
-#>  backports    1.1.2   2017-12-13 CRAN (R 3.4.3)
-#>  base       * 3.4.3   2018-01-16 local         
-#>  bindr        0.1.1   2018-03-13 CRAN (R 3.4.3)
-#>  bindrcpp     0.2.2   2018-03-29 CRAN (R 3.4.3)
-#>  codetools    0.2-15  2016-10-05 CRAN (R 3.4.3)
-#>  colorspace   1.3-2   2016-12-14 CRAN (R 3.4.3)
-#>  compiler     3.4.3   2018-01-16 local         
-#>  cowplot    * 0.9.3   2018-07-15 CRAN (R 3.4.3)
-#>  crayon       1.3.4   2017-09-16 CRAN (R 3.4.3)
-#>  datasets   * 3.4.3   2018-01-16 local         
-#>  devtools     1.13.6  2018-06-27 CRAN (R 3.4.3)
-#>  digest       0.6.15  2018-01-28 CRAN (R 3.4.3)
-#>  dplyr        0.7.6   2018-06-29 CRAN (R 3.4.3)
-#>  evaluate     0.11    2018-07-17 CRAN (R 3.4.3)
-#>  ggplot2    * 3.0.0   2018-07-03 CRAN (R 3.4.3)
-#>  glue         1.3.0   2018-07-17 CRAN (R 3.4.3)
-#>  graphics   * 3.4.3   2018-01-16 local         
-#>  gratia     * 0.0-6   2018-08-16 local         
-#>  grDevices  * 3.4.3   2018-01-16 local         
-#>  grid         3.4.3   2018-01-16 local         
-#>  gtable       0.2.0   2016-02-26 CRAN (R 3.4.3)
-#>  htmltools    0.3.6   2017-04-28 CRAN (R 3.4.3)
-#>  knitr        1.20    2018-02-20 CRAN (R 3.4.3)
-#>  labeling     0.3     2014-08-23 CRAN (R 3.4.3)
-#>  lattice      0.20-35 2017-03-25 CRAN (R 3.4.3)
-#>  lazyeval     0.2.1   2017-10-29 CRAN (R 3.4.3)
-#>  magrittr     1.5     2014-11-22 CRAN (R 3.4.3)
-#>  MASS         7.3-50  2018-04-30 CRAN (R 3.4.3)
-#>  Matrix       1.2-14  2018-04-09 CRAN (R 3.4.3)
-#>  memoise      1.1.0   2017-04-21 CRAN (R 3.4.3)
-#>  methods      3.4.3   2018-01-16 local         
-#>  mgcv       * 1.8-24  2018-06-18 CRAN (R 3.4.3)
-#>  munsell      0.5.0   2018-06-12 CRAN (R 3.4.3)
-#>  nlme       * 3.1-137 2018-04-07 CRAN (R 3.4.3)
-#>  pillar       1.3.0   2018-07-14 CRAN (R 3.4.3)
-#>  pkgconfig    2.0.1   2017-03-21 CRAN (R 3.4.3)
-#>  plyr         1.8.4   2016-06-08 CRAN (R 3.4.3)
-#>  purrr        0.2.5   2018-05-29 CRAN (R 3.4.3)
-#>  R6           2.2.2   2017-06-17 CRAN (R 3.4.3)
-#>  Rcpp         0.12.18 2018-07-23 CRAN (R 3.4.3)
-#>  rlang        0.2.1   2018-05-30 CRAN (R 3.4.3)
-#>  rmarkdown  * 1.10    2018-06-11 CRAN (R 3.4.3)
-#>  rprojroot    1.3-2   2018-01-03 CRAN (R 3.4.3)
-#>  scales       0.5.0   2017-08-24 CRAN (R 3.4.3)
-#>  scam       * 1.2-2   2017-09-24 CRAN (R 3.4.3)
-#>  splines      3.4.3   2018-01-16 local         
-#>  stats      * 3.4.3   2018-01-16 local         
-#>  stringi      1.2.4   2018-07-20 CRAN (R 3.4.3)
-#>  stringr      1.3.1   2018-05-10 CRAN (R 3.4.3)
-#>  tibble       1.4.2   2018-01-22 CRAN (R 3.4.3)
-#>  tidyr      * 0.8.1   2018-05-18 CRAN (R 3.4.3)
-#>  tidyselect   0.2.4   2018-02-26 CRAN (R 3.4.3)
-#>  tools        3.4.3   2018-01-16 local         
-#>  utils      * 3.4.3   2018-01-16 local         
-#>  withr        2.1.2   2018-03-15 CRAN (R 3.4.3)
-#>  yaml         2.2.0   2018-07-25 CRAN (R 3.4.3)
+#>  language en_GB:en
+#>  collate  en_GB.UTF-8
+#>  ctype    en_GB.UTF-8
+#>  tz       Europe/Copenhagen
+#>  date     2023-10-16
+#>  pandoc   2.5 @ /usr/bin/ (via rmarkdown)
+#> 
+#> - Packages ---------------------------------------------------------------------------------------
+#>  package     * version  date (UTC) lib source
+#>  cachem        1.0.8    2023-05-01 [1] RSPM (R 4.3.0)
+#>  callr         3.7.3    2022-11-02 [1] RSPM (R 4.3.0)
+#>  cli           3.6.1    2023-03-23 [1] RSPM (R 4.3.0)
+#>  codetools     0.2-19   2023-02-01 [1] RSPM (R 4.3.0)
+#>  colorspace    2.1-0    2023-01-23 [1] RSPM (R 4.3.0)
+#>  cowplot     * 1.1.1    2020-12-30 [1] RSPM (R 4.3.0)
+#>  crayon        1.5.2    2022-09-29 [1] RSPM (R 4.3.0)
+#>  devtools      2.4.5    2022-10-11 [1] RSPM (R 4.3.0)
+#>  digest        0.6.33   2023-07-07 [1] RSPM (R 4.3.1)
+#>  dplyr       * 1.1.3    2023-09-03 [1] RSPM (R 4.3.1)
+#>  ellipsis      0.3.2    2021-04-29 [1] RSPM (R 4.3.0)
+#>  evaluate      0.21     2023-05-05 [1] RSPM (R 4.3.0)
+#>  fansi         1.0.5    2023-10-08 [1] CRAN (R 4.3.1)
+#>  farver        2.1.1    2022-07-06 [1] RSPM (R 4.3.0)
+#>  fastmap       1.1.1    2023-02-24 [1] RSPM (R 4.3.0)
+#>  fs            1.6.3    2023-07-20 [1] RSPM (R 4.3.1)
+#>  generics      0.1.3    2022-07-05 [1] RSPM (R 4.3.0)
+#>  ggokabeito    0.1.0    2021-10-18 [1] RSPM (R 4.3.0)
+#>  ggplot2     * 3.4.3    2023-08-14 [1] RSPM (R 4.3.1)
+#>  glue          1.6.2    2022-02-24 [1] RSPM (R 4.3.0)
+#>  gratia      * 0.8.1.41 2023-10-16 [1] local
+#>  gtable        0.3.4    2023-08-21 [1] RSPM (R 4.3.1)
+#>  htmltools     0.5.6    2023-08-10 [1] RSPM (R 4.3.1)
+#>  htmlwidgets   1.6.2    2023-03-17 [1] RSPM (R 4.3.0)
+#>  httpuv        1.6.11   2023-05-11 [1] RSPM (R 4.3.0)
+#>  knitr         1.44     2023-09-11 [1] RSPM (R 4.3.1)
+#>  labeling      0.4.3    2023-08-29 [1] RSPM (R 4.3.1)
+#>  later         1.3.1    2023-05-02 [1] RSPM (R 4.3.0)
+#>  lattice       0.21-8   2023-04-05 [1] RSPM (R 4.3.0)
+#>  lifecycle     1.0.3    2022-10-07 [1] RSPM (R 4.3.0)
+#>  magrittr      2.0.3    2022-03-30 [1] RSPM (R 4.3.0)
+#>  Matrix        1.6-1.1  2023-09-18 [1] RSPM (R 4.3.1)
+#>  memoise       2.0.1    2021-11-26 [1] RSPM (R 4.3.0)
+#>  mgcv        * 1.9-0    2023-07-11 [1] RSPM (R 4.3.1)
+#>  mime          0.12     2021-09-28 [1] RSPM (R 4.3.0)
+#>  miniUI        0.1.1.1  2018-05-18 [1] RSPM (R 4.3.0)
+#>  munsell       0.5.0    2018-06-12 [1] RSPM (R 4.3.0)
+#>  mvnfast       0.2.8    2023-02-23 [1] RSPM (R 4.3.0)
+#>  nlme        * 3.1-163  2023-08-09 [1] RSPM (R 4.3.1)
+#>  patchwork     1.1.3    2023-08-14 [1] RSPM (R 4.3.1)
+#>  pillar        1.9.0    2023-03-22 [1] RSPM (R 4.3.0)
+#>  pkgbuild      1.4.2    2023-06-26 [1] RSPM (R 4.3.1)
+#>  pkgconfig     2.0.3    2019-09-22 [1] RSPM (R 4.3.0)
+#>  pkgload       1.3.3    2023-09-22 [1] CRAN (R 4.3.1)
+#>  prettyunits   1.2.0    2023-09-24 [1] CRAN (R 4.3.1)
+#>  processx      3.8.2    2023-06-30 [1] RSPM (R 4.3.1)
+#>  profvis       0.3.8    2023-05-02 [1] RSPM (R 4.3.0)
+#>  promises      1.2.1    2023-08-10 [1] RSPM (R 4.3.1)
+#>  ps            1.7.5    2023-04-18 [1] RSPM (R 4.3.0)
+#>  purrr         1.0.2    2023-08-10 [1] RSPM (R 4.3.1)
+#>  R6            2.5.1    2021-08-19 [1] RSPM (R 4.3.0)
+#>  Rcpp          1.0.11   2023-07-06 [1] RSPM (R 4.3.1)
+#>  remotes       2.4.2.1  2023-07-18 [1] RSPM (R 4.3.1)
+#>  rlang         1.1.1    2023-04-28 [1] RSPM (R 4.3.0)
+#>  rmarkdown   * 2.25     2023-09-18 [1] RSPM (R 4.3.1)
+#>  scales        1.2.1    2022-08-20 [1] RSPM (R 4.3.0)
+#>  scam        * 1.2-14   2023-04-14 [1] RSPM (R 4.3.0)
+#>  sessioninfo   1.2.2    2021-12-06 [1] RSPM (R 4.3.0)
+#>  shiny         1.7.5    2023-08-12 [1] RSPM (R 4.3.1)
+#>  stringi       1.7.12   2023-01-11 [1] RSPM (R 4.3.0)
+#>  stringr       1.5.0    2022-12-02 [1] CRAN (R 4.3.0)
+#>  tibble        3.2.1    2023-03-20 [1] CRAN (R 4.3.0)
+#>  tidyr       * 1.3.0    2023-01-24 [1] RSPM (R 4.3.0)
+#>  tidyselect    1.2.0    2022-10-10 [1] RSPM (R 4.3.0)
+#>  urlchecker    1.0.1    2021-11-30 [1] RSPM (R 4.3.0)
+#>  usethis       2.2.2    2023-07-06 [1] RSPM (R 4.3.1)
+#>  utf8          1.2.3    2023-01-31 [1] RSPM (R 4.3.0)
+#>  vctrs         0.6.3    2023-06-14 [1] RSPM (R 4.3.0)
+#>  withr         2.5.1    2023-09-26 [1] RSPM (R 4.3.1)
+#>  xfun          0.40     2023-08-09 [1] RSPM (R 4.3.1)
+#>  xtable        1.8-4    2019-04-21 [1] RSPM (R 4.3.0)
+#>  yaml          2.3.7    2023-01-23 [1] RSPM (R 4.3.0)
+#> 
+#>  [1] /home/au690221/R/x86_64-pc-linux-gnu-library/4.3
+#>  [2] /usr/local/lib/R/site-library
+#>  [3] /usr/lib/R/site-library
+#>  [4] /usr/lib/R/library
+#> 
+#> --------------------------------------------------------------------------------------------------
 \end{verbatim}

### 2023-10-16 data/small-water/prepare-small.R

@@ -1,0 +1,12 @@
+library("readr")
+library("janitor")
+library("tibble")
+library("dplyr")
+
+small <- readr::read_rds("small-water-isotope-data.rds")
+small <- small |>
+  janitor::clean_names() |>
+  tibble::as_tibble()
+
+write_csv(small, "small-water-isotopes.csv")
+
